### 2014-07-01 interimreport1.Rnw
Maps and draft interim report code
Lots of stuff

\documentclass{article}

\usepackage{times}
\usepackage{hyperref}
\usepackage{tikz}
\usepackage{bibunits}
\usepackage{natbib}
\usepackage{graphics}
\usepackage{amsmath}
\usepackage{indentfirst}
\usepackage[utf8]{inputenc}

\DeclareMathOperator{\var}{var}
\DeclareMathOperator{\cov}{cov}

\begin{document}
\SweaveOpts{concordance=TRUE}

\title{Measles risk assessment and cost interim report\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Nigel P French}}
\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
Massey University,\\
Palmerston North 4442,\\
New Zealand\\
\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
\date{\today}  %\today is replaced with the current date
\maketitle

\section{Abstract}

Summary of findings

\section{Background}

As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committted to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002. The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved this.

The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997. Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011/2012 (mainly in Auckland region) and another significant outbreak is currently going on in Auckland region. The outbreak in 2011-2012 lasted for more than 12 months and the current 2013/2014 outbreak started at the end of December 2013 (as on \date{\today}). In 2013, prior to the 2013/2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) it can request verification of non-endemic status three years after the last case of the 2011-12 outbreak in June 2012.

Previous analyses, including two of measles in New Zealand by Prof. Roberts (INSERT REFS), have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles. Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, as described above, outbreaks still occur. Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are perhaps likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining. Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age. Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.

\section{Risk analysis review}

A measles risk assessment has been started by the Ministry of Health to better assess current and future population immunity and high risk groups. Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
\begin{itemize}
\item In this section we review the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}.
\end{itemize}

Overall, we thought the review was very thorough. The report included substantial background information on measles immunisation in New Zealand (\emph{section 1.3}), the epidemiology of measles in New Zealand (\emph{section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{section 3}), and the levels of population immunity against the virus (\emph{section 4}). Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{section 4.2}) and the sustainability of the national immunisation programme (\emph{section 5}).

Within the report there are many tables and figures which given substantial detail on the measles situation in New Zealand. Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.

Specific epidemiological details were provided for the 2011/2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month, 2011 and 2012, with additional breakdown by age group, 2011 and 2012 (\emph{Figure 5}) and per 100,000 population (e.g. \emph{Figures 6-8}). Similar presentation of the case data are provided for ethnicity (e.g. \emph{Figures 9-10}) and New Zealand Index of Deprivation (NZDep) (e.g. \emph{Figures 11-13}). Three figures,\emph{Figures 12, 13, and 28}, show that there is spatial clustering of cases.

[INSERT NEED FOR MULTIVARIATE ANALYSES HERE?]

The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ. We agree with this statement, but have included additional analyses on the current outbreak that [DEMONSTRATE/SUGGEST ...INSERT THE CONCLUSIONS RE R0]

Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007) and only Asian and Pacific ethnicities have consistently had MMR dose one coverage approching or over 95\% levels for cohorts from 2007 onwards. [COMMENT ON AGREEMENT HERE?] Furthermore, ... [COMMENT ON REFUSAL RATE?]

The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases. [COMMENT? Unsure, though don't want to be dismissive].

The development of a National Immunisation Register (NIR) that allows tracking of the vaccination status of children is very useful. [COMMENT ON NEED FOR FINER IMMUNISATION LEVEL DATA HERE?]. The report conludes that the surveillance performance indicators demonstrate a reasonable measles surveillance capability in New Zealand. We [COMMENT ON THIS, GIVEN JONATHAN'S FINDINGS]?.

We agree with the report's conclusions that testing for measles is performed appropriately within the required timeframe, but clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary.

\section{Additional risk analyses}

In this section we provide ... [].

\section{Cost analyses}

In this section we provide ... [].

<<frequentist>>=
library(mcmc)
data(logit)
out <- glm(y ~ x1 + x2 + x3 + x4, data = logit,
    family = binomial())
summary(out)
@

...text

<<log.unnormalized.posterior>>=
x <- logit
x$y <- NULL
x <- as.matrix(x)
x <- cbind(1, x)
dimnames(x) <- NULL

y <- logit$y

lupost <- function(beta, x, y) {
    eta <- x %*% beta
    p <- 1 / (1 + exp(- eta))   # note: works even when eta is Inf or -Inf
    logl <- sum(log(p[y == 1])) + sum(log(1 - p[y == 0]))
    return(logl + sum(dnorm(beta, 0, 2, log = TRUE)))
}
@

\section{Future analyses}

We received The Institute of Environmental Science and Research Ltd (ESR) EpiSurv measles case data on the 27 June 2014. Additional analyses we wish to do are:
\begin{itemize}
\item ...[]
\end{itemize}

\section{Future analyses}

We received The Institute of Environmental Science and Research Ltd (ESR) EpiSurv measles case data on the 27 June 2014. Additional analyses we wish to do are:
\begin{itemize}
\item ...[]
\end{itemize}

<<metropolis-try-1>>=
set.seed(42)    # to get reproducible results
beta.init <- as.numeric(coefficients(out))
out <- metrop(lupost, beta.init, 1e3, x = x, y = y)
names(out)
out$accept
@

...text

\begin{itemize}
\item ...text.
\end{itemize}

...text

\citet{geyer-temp} ...text

<<metropolis-try-2>>=
out <- metrop(out, scale = 0.1, x = x, y = y)
out$accept
out <- metrop(out, scale = 0.3, x = x, y = y)
out$accept
out <- metrop(out, scale = 0.5, x = x, y = y)
out$accept
out <- metrop(out, scale = 0.4, x = x, y = y)
out$accept
@

...text

<<label=metropolis-try-3>>=
out <- metrop(out, nbatch = 1e4, x = x, y = y)
out$accept
out$time
@

Figure~\ref{fig:fig1} (page~\pageref{fig:fig1})
shows the time series plot made by the R statement
<<label=fig1too,include=FALSE>>=
plot(ts(out$batch))
@
\begin{figure}
\begin{center}
<<label=fig1,fig=TRUE,echo=FALSE>>=
<<fig1too>>
@
\end{center}
\caption{Time series plot of MCMC output.}
\label{fig:fig1}
\end{figure}

Another way to look at the output is an autocorrelation plot.
Figure~\ref{fig:fig2} (page~\pageref{fig:fig2})
shows the time series plot made by the R statement
<<label=fig2too,include=FALSE>>=
acf(out$batch)
@
\begin{figure}
\begin{center}
<<label=fig2,fig=TRUE,echo=FALSE>>=
<<fig2too>>
@
\end{center}
\caption{Autocorrelation plot of MCMC output.}
\label{fig:fig2}
\end{figure}

...text
\begin{verbatim}
http://www.stat.umn.edu/~charlie/mcmc/diag.html
\end{verbatim}
says
\begin{quotation}
...text
\end{quotation}
...text
...text
\section{Monte Carlo Estimates and Standard Errors}

<<label=metropolis-try-4>>=
out <- metrop(out, nbatch = 1e2, blen = 100,
    outfun = function(z, ...) c(z, z^2), x = x, y = y)
out$accept
out$time
@

We have added an argument \verb@outfun@ ...text
We need to
use the identity
$$
   \var(X) = E(X^2) - E(X)^2
$$
...text
Hence our \verb@outfun@ returns
\verb@c(z, z^2)@ for an argument (the state vector) \verb@z@.

The \verb@...@ argument to \verb@outfun@ is required, since the
function is also passed the other arguments (here \verb@x@ and \verb@y@)
to \verb@metrop@.

\subsection{Simple Means}

The grand means (means of batch means) are
<<label=metropolis-batch>>=
apply(out$batch, 2, mean)
@
...text
<<label=metropolis-batch-too>>=
foo <- apply(out$batch, 2, mean)
mu <- foo[1:5]
sigmasq <- foo[6:10] - mu^2
mu
sigmasq
@

...text
<<label=metropolis-mcse-mu>>=
mu.mcse <- apply(out$batch[ , 1:5], 2, sd) / sqrt(out$nbatch)
mu.mcse
@
The extra factor \verb@sqrt(out$nbatch)@ arises because the batch means
have variance $\sigma^2 / b$ where $b$ is the batch length, which is
\verb@out$blen@,
whereas the overall means \verb@mu@ have variance $\sigma^2 / n$ where
$n$ is the total number of iterations, which is \verb@out$blen * out$nbatch@.

\subsection{Functions of Means}

To get the MCSE for the posterior variances we apply the delta method.
Let $u_i$ denote the sequence of batch means of the first kind for one
parameter and $\bar{u}$ the grand mean (the estimate of the posterior mean
of that parameter),
let $v_i$ denote the sequence of batch means of the second kind for the
same parameter and $\bar{v}$ the grand mean (the estimate of the posterior
second absolute moment of that parameter), and let $\mu = E(\bar{u})$ and
$\nu = E(\bar{v})$.  Then the delta method linearizes the nonlinear function
$$
   g(\mu, \nu) = \nu - \mu^2
$$
as
$$
   \Delta g(\mu, \nu) = \Delta \nu - 2 \mu \Delta \mu
$$
saying that
$$
   g(\bar{u}, \bar{v}) - g(\mu, \nu)
$$
has the same asymptotic normal distribution as
$$
   (\bar{v} - \nu) - 2 \mu (\bar{u} - \mu)
$$
which, of course, has variance \verb@1 / nbatch@ times that of
$$
   (v_i - \nu) - 2 \mu (u_i - \mu)
$$
and this variance is estimated by
$$
   \frac{1}{n_{\text{batch}}} \sum_{i = 1}^{n_{\text{batch}}}
   \bigl[ (v_i - \bar{v}) - 2 \bar{u} (u_i - \bar{u}) \bigr]^2
$$
So
<<label=metropolis-mcse-sigmasq>>=
u <- out$batch[ , 1:5]
v <- out$batch[ , 6:10]
ubar <- apply(u, 2, mean)
vbar <- apply(v, 2, mean)
deltau <- sweep(u, 2, ubar)
deltav <- sweep(v, 2, vbar)
foo <- sweep(deltau, 2, ubar, "*")
sigmasq.mcse <- sqrt(apply((deltav - 2 * foo)^2, 2, mean) / out$nbatch)
sigmasq.mcse
@
does the MCSE for the posterior variance.

Let's just check that this complicated \verb@sweep@ and \verb@apply@ stuff
does do the right thing.
<<label=metropolis-mcse-sigmasq-too>>=
sqrt(mean(((v[ , 2] - vbar[2]) - 2 * ubar[2] * (u[ , 2] - ubar[2]))^2) /
    out$nbatch)
@

\paragraph{Comment} Through version 0.5 of this vignette it contained
an incorrect procedure for calculating this MCSE, justified by a handwave.
Essentially, it said to use the standard deviation of the batch means called
\verb@v@ here, which appears to be very conservative.

\subsection{Functions of Functions of Means}

If we are also interested in the posterior standard deviation
(a natural question, although not asked on the exam problem),
the delta method gives its standard error in terms of that
for the variance
<<label=metropolis-mcse-sigma>>=
sigma <- sqrt(sigmasq)
sigma.mcse <- sigmasq.mcse / (2 * sigma)
sigma
sigma.mcse
@

\section{A Final Run}

...text...
(the exam problem directed ``use a long enough run of your Markov chain
sampler so that the MCSE are less than 0.01'')
<<label=metropolis-try-5>>=
out <- metrop(out, nbatch = 5e2, blen = 400, x = x, y = y)
out$accept
out$time
<<metropolis-batch-too>>
<<metropolis-mcse-mu>>
<<metropolis-mcse-sigmasq>>
<<metropolis-mcse-sigma>>
@
...text... \verb@xtable@ command in the \verb@xtable@ library.

First the posterior means,
\begin{table}[ht]
\caption{Posterior Means}
\label{tab:mu}
\begin{center}
<<label=tab1,echo=FALSE,results=tex>>=
foo <- rbind(mu, mu.mcse)
dimnames(foo) <- list(c("estimate", "MCSE"),
    c("constant", paste("$x_", 1:4, "$", sep = "")))
library(xtable)
print(xtable(foo, digits = rep(4, 6),
    align = c("l", rep("c", 5))), floating = FALSE,
    caption.placement = "top",
    sanitize.colnames.function = function(x) return(x))
@
\end{center}
\end{table}
then the posterior variances (table on page~\pageref{tab:sigmasq}),
\begin{table}[ht]
\caption{Posterior Variances}
\label{tab:sigmasq}
\begin{center}
<<label=tab1,echo=FALSE,results=tex>>=
foo <- rbind(sigmasq, sigmasq.mcse)
dimnames(foo) <- list(c("estimate", "MCSE"),
    c("constant", paste("$x_", 1:4, "$", sep = "")))
library(xtable)
print(xtable(foo, digits = rep(4, 6),
    align = c("l", rep("c", 5))), floating = FALSE,
    caption.placement = "top",
    sanitize.colnames.function = function(x) return(x))
@
\end{center}
\end{table}
and finally the posterior standard deviations
(table on page~\pageref{tab:sigma}).
\begin{table}[ht]
\caption{Posterior Standard Deviations}
\label{tab:sigma}
\begin{center}
<<label=tab1,echo=FALSE,results=tex>>=
foo <- rbind(sigma, sigma.mcse)
dimnames(foo) <- list(c("estimate", "MCSE"),
    c("constant", paste("$x_", 1:4, "$", sep = "")))
library(xtable)
print(xtable(foo, digits = rep(4, 6),
    align = c("l", rep("c", 5))), floating = FALSE,
    caption.placement = "top",
    sanitize.colnames.function = function(x) return(x))
@
\end{center}
\end{table}

Note for ...
<<label=time,echo=FALSE,results=tex>>=
cat(out$time[1], "\n")
@
...text...

\section{New Variance Estimation Functions}

A new function \texttt{initseq} estimates variances in the Markov chain
central limit theorem (CLT) following the methodology introduced by
\citet[Section~3.3]{practical}.  These methods only apply to scalar-valued
functionals of
reversible Markov chains, but the Markov chains produced by the \texttt{metrop}
function satisfy this condition, even, as we shall see below, when batching
is used.

...text.. AR(1) ...text... \texttt{initseq}.
<<x>>=
n <- 2e4
rho <- 0.99
x <- arima.sim(model = list(ar = rho), n = n)
@
...text...

Define
\begin{equation} \label{eq:little}
   \gamma_k = \cov(X_i, X_{i + k})
\end{equation}
where the covariances refer to the stationary Markov chain having the
same transition probabilities as \texttt{x}.  Then the variance in the CLT
is
$$
   \sigma^2 = \gamma_0 + 2 \sum_{k = 1}^\infty \gamma_k
$$
\citep[Theorem~2.1]{practical}, that is,
$$
   \bar{x}_n \approx \text{Normal}\left(\mu, \frac{\sigma^2}{n}\right),
$$
where $\mu = E(X_i)$ is the quantity being estimated by MCMC (in this
toy problem we know $\mu = 0$).

Naive estimates of $\sigma^2$ obtained by plugging in empirical
estimates of the gammas do not provide consistent estimation
\citep[Section~3.1]{practical}.  Thus the scheme implemented
by the R function \texttt{initseq}.  Define
\begin{equation} \label{eq:big}
   \Gamma_k = \gamma_{2 k} + \gamma_{2 k + 1}
\end{equation}
\citet[Theorem~3.1]{practical} says that $\Gamma_k$ ...text..
\citet[Section~3.3]{practical} ...text.. Theorem~3.2.

Figure~\ref{fig:gamma} (page~\pageref{fig:gamma})
shows the time series plot made by the R statement
<<label=figgamtoo,include=FALSE>>=
out <- initseq(x)
plot(seq(along = out$Gamma.pos) - 1, out$Gamma.pos,
        xlab = "k", ylab = expression(Gamma[k]), type = "l")
lines(seq(along = out$Gamma.dec) - 1, out$Gamma.dec, lty = "dotted")
lines(seq(along = out$Gamma.con) - 1, out$Gamma.con, lty = "dashed")
@
\begin{figure}
\begin{center}
<<label=figgam,fig=TRUE,echo=FALSE>>=
<<figgamtoo>>
@
\end{center}
\caption{Plot ``Big Gamma'' defined by \eqref{eq:little} and \eqref{eq:big}.
Solid line, initial positive sequence estimator.
Dotted line, initial monotone sequence estimator.
Dashed line, initial convex sequence estimator.}
\label{fig:gamma}
\end{figure}
...text... Figure~\ref{fig:gamma}, ...text... $\sigma^2$, .... 
<<assvar>>=
out$var.con
(1 + rho) / (1 - rho) * 1 / (1 - rho^2)
@
where for comparison we have given the exact theoretical value of $\sigma^2$,
...text...
<<batx>>=
blen <- 5
x.batch <- apply(matrix(x, nrow = blen), 2, mean)
bout <- initseq(x.batch)
@
Because the batch length is too short, the variance of the batch means
does not estimate $\sigma^2$.  We must account for the autocorrelation
of the batches, shown in Figure~\ref{fig:gambat}.
<<label=figgambattoo,include=FALSE>>=
plot(seq(along = bout$Gamma.con) - 1, bout$Gamma.con,
        xlab = "k", ylab = expression(Gamma[k]), type = "l")
@
\begin{figure}
\begin{center}
<<label=figgambat,fig=TRUE,echo=FALSE>>=
<<figgambattoo>>
@
\end{center}
\caption{Plot ``Big Gamma'' defined by \eqref{eq:little} and \eqref{eq:big}
for the sequence of batch means (batch length \Sexpr{blen}).
Only initial convex sequence estimator is shown.}
\label{fig:gambat}
\end{figure}
...text the $\sigma^2$ for the original series.
<<compvar>>=
out$var.con
bout$var.con * blen
@
Another way to look at this is that the MCMC estimator of $\mu$ is
either \texttt{mean(x)} or \texttt{mean(x.batch)}.  And the variance
must be divided by the sample size to give standard errors.  So either
<<ci-con>>=
mean(x) + c(-1, 1) * qnorm(0.975) * sqrt(out$var.con / length(x))
mean(x.batch) + c(-1, 1) * qnorm(0.975) * sqrt(bout$var.con / length(x.batch))
@
is an asymptotic 95\% confidence interval for $\mu$.  Just divide by
the relevant sample size.

\begin{thebibliography}{}

\bibitem[Gelman et al.(1996)Gelman, Roberts, and Gilks]{grg}
Gelman, A., G.~O. Roberts, and W.~R. Gilks (1996).
\newblock Efficient Metropolis jumping rules.
\newblock In \emph{Bayesian Statistics, 5 (Alicante, 1994)}, pp.~599--607.
  Oxford University Press.

\bibitem[Geyer(1992)]{practical}
Geyer, C.~J. (1992).
\newblock Practical Markov chain Monte Carlo (with discussion).
\newblock \emph{Statistical Science}, 7, 473--511.

\bibitem[Geyer and Thompson(1995)]{geyer-temp}
Geyer, C.~J. and E.~A. Thompson (1995).
\newblock Annealing Markov chain Monte Carlo with applications to
    ancestral inference.
\newblock \emph{Journal of the American Statistical Association}, 90, 909--920.

\end{thebibliography}

\end{document}

### 2014-07-01 maps/rworldmaps_measles.R

@@ -1,0 +1,571 @@
+library(maps)       # Provides functions that let us plot the maps
+library(mapdata)    # Contains the hi-resolution points that mark out the countries.
+library(rworldmap)
+library(adehabitat)
+library(rgdal)
+library(raster)
+library(maps)
+library(maptools)
+library(sp)
+#library(SDMTools)
+library(ggplot2)
+require(rgeos)
+library(ggmap)
+library(mapdata)
+gpclibPermit()
+library(mapproj)
+library(plyr)
+library(grid)
+library(gridExtra)
+##
+
+world_map <- map_data("world")
+world.ggmap <- fortify(world_map, region = "NAME")
+
+n <- length(unique(world.ggmap$region))
+df <- data.frame(id = unique(world.ggmap$region),
+                 growth = 4*runif(n),
+                 category = factor(sample(1:5, n, replace=T)))
+
+## noise
+df[c(sample(1:100,40)),c("growth", "category")] <- NA
+
+
+ggplot(df, aes(map_id = id)) +
+  geom_map(aes(fill = growth, color = category), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  scale_fill_gradient(low = "red", high = "blue", guide = "colorbar")
+
+map.world <- map_data(map = "world")
+# map = name of map provided by the maps package.
+# These include county, france, italy, nz, state, usa, world, world2.
+str(map.world)
+# how many regions
+length(unique(map.world$region))
+# how many group polygons (some regions have multiple parts)
+length(unique(map.world$group))
+p1 <- ggplot(map.world, aes(x = long, y = lat, group = group))
+p1 <- p1 + geom_polygon() # fill areas
+p1 <- p1 + labs(title = "World, plain")
+#print(p1)
+p2 <- ggplot(map.world, aes(x = long, y = lat, group = group, colour = region))
+p2 <- p2 + geom_polygon() # fill areas
+p2 <- p2 + theme(legend.position="none") # remove legend with fill colours
+p2 <- p2 + labs(title = "World, colour borders")
+#print(p2)
+p3 <- ggplot(map.world, aes(x = long, y = lat, group = group, fill = region))
+p3 <- p3 + geom_polygon() # fill areas
+p3 <- p3 + theme(legend.position="none") # remove legend with fill colours
+p3 <- p3 + labs(title = "World, filled regions")
+#print(p3)
+p4 <- ggplot(map.world, aes(x = long, y = lat, group = group, colour = region))
[…]

### 2014-07-01 multivariate_measles.R

@@ -100,7 +100,7 @@ hist(testtable$Cases,breaks=seq(0,max(testtable$Cases),by=5))
 ## glm with Poisson errors
 # start with full model
 model7<-glm(Cases~NZDep*Age*Ethnicity,data=testtable,family=poisson)
+
 summary(model7)
-summary.lm(model7)
-?summary.lm
+anova(model7,test="Chi")
 # why difference?

### 2014-07-01 rworldmaps_measles.R

@@ -1,571 +1,0 @@
-library(maps)       # Provides functions that let us plot the maps
-library(mapdata)    # Contains the hi-resolution points that mark out the countries.
-library(rworldmap)
-library(adehabitat)
-library(rgdal)
-library(raster)
-library(maps)
-library(maptools)
-library(sp)
-#library(SDMTools)
-library(ggplot2)
-require(rgeos)
-library(ggmap)
-library(mapdata)
-gpclibPermit()
-library(mapproj)
-library(plyr)
-library(grid)
-library(gridExtra)
-##
-
-world_map <- map_data("world")
-world.ggmap <- fortify(world_map, region = "NAME")
-
-n <- length(unique(world.ggmap$region))
-df <- data.frame(id = unique(world.ggmap$region),
-                 growth = 4*runif(n),
-                 category = factor(sample(1:5, n, replace=T)))
-
-## noise
-df[c(sample(1:100,40)),c("growth", "category")] <- NA
-
-
-ggplot(df, aes(map_id = id)) +
-  geom_map(aes(fill = growth, color = category), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  scale_fill_gradient(low = "red", high = "blue", guide = "colorbar")
-
-map.world <- map_data(map = "world")
-# map = name of map provided by the maps package.
-# These include county, france, italy, nz, state, usa, world, world2.
-str(map.world)
-# how many regions
-length(unique(map.world$region))
-# how many group polygons (some regions have multiple parts)
-length(unique(map.world$group))
-p1 <- ggplot(map.world, aes(x = long, y = lat, group = group))
-p1 <- p1 + geom_polygon() # fill areas
-p1 <- p1 + labs(title = "World, plain")
-#print(p1)
-p2 <- ggplot(map.world, aes(x = long, y = lat, group = group, colour = region))
-p2 <- p2 + geom_polygon() # fill areas
-p2 <- p2 + theme(legend.position="none") # remove legend with fill colours
-p2 <- p2 + labs(title = "World, colour borders")
-#print(p2)
-p3 <- ggplot(map.world, aes(x = long, y = lat, group = group, fill = region))
-p3 <- p3 + geom_polygon() # fill areas
-p3 <- p3 + theme(legend.position="none") # remove legend with fill colours
-p3 <- p3 + labs(title = "World, filled regions")
-#print(p3)
-p4 <- ggplot(map.world, aes(x = long, y = lat, group = group, colour = region))
[…]

### 2014-07-01 maps/rworldmaps_measles.R
adds a bit of code to get around the silly comma problem in the immigration data

@@ -312,6 +312,23 @@ head(immigration)
 class(immigration)
 
 ##
+
+## silly dataset contains weird comma's that R doesn't detect as the thousands separator.
+## This magic fixes it...
+
+# convert to characters
+num_clients_char <- as.character(immigration$Number.of.Clients)
+# split up by the silly comma
+num_clients_split <- strsplit(num_clients_char, ",")
+# now combine back
+combine_thousands <- function(x)
+{
+  x <- as.numeric(x)
+  thousand_power <- ((length(x):1)-1)*3
+  sum(10^thousand_power*x)
+}
+immigration$Number.of.Clients <- sapply(num_clients_split, combine_thousands)
+
 class(immigration$Month.of.Arrival)
 test<-aggregate( cbind(  Number.of.Clients ) ~ Nationality , data = immigration , sum )
 dim(test)

### 2014-07-01 estimatingR0/extract_data.R
update extraction of data to do various plots

@@ -58,31 +58,6 @@ all[!outbreak_rows,outbreak_code_col] <- 0
 notification_dates <- as.Date(all[,date_col], format=date_format)
 notification_weeks <- sapply(notification_dates, surv_week_start)
 
-# create output directory
-dir.create(file.path(output_dir), showWarnings = FALSE)
-
-# for each outbreak, compute the incidence
-outbreak_nos <- unique(all[outbreak_rows,outbreak_code_col])
-for (ob_num in outbreak_nos)
-{
-  # find the rows
-  ob_rows  <- all[,outbreak_code_col] == ob_num
-
-  # create a date range
-  ob_range <- range(as.Date(notification_weeks[ob_rows]))
-  ob_weeks <- rep(0, length=diff(ob_range)/7+1)
-  names(ob_weeks) <- seq(ob_range[1], ob_range[2], by=7)
-
-  # compute the incidence by tabling up
-  t <- table(notification_weeks[ob_rows])
-  ob_weeks[names(t)] <- t
-
-  # write out an outbreak file
-  outbreak_data <- data.frame(date=names(ob_weeks), week=1:length(ob_weeks), incidence=ob_weeks)
-  outbreak_file <- file.path(output_dir, sprintf("outbreak%02d.csv", ob_num))
-  write.csv(outbreak_data, outbreak_file, row.names=F)
-}
-
 # now the total incidence through time, with outbreak no's as well
 
 pdf("incidence.pdf", width=20, height=8)
@@ -116,5 +91,161 @@ dev.off()
 barplot(ob_weeks)
 dev.off()
 
+# plot finer-grained total cases per week since 2009
+rows <- notification_weeks > "2008-12-31"
 
+weeks_2007 <- notification_weeks[rows]
+outbreaks_2007 <- all[rows,outbreak_code_col]
+
+pdf("outbreaks_2009_2014.pdf", width=10, height=5)
+ob_range <- range(as.Date(weeks_2007))
+ob_weeks <- matrix(0, length(unique(outbreaks_2007)), diff(ob_range)/7+1)
+colnames(ob_weeks) <- as.character(seq(ob_range[1], ob_range[2], by=7))
+rownames(ob_weeks) <- sort(unique(outbreaks_2007))
+t <- table(outbreaks_2007, weeks_2007)
+ob_weeks[rownames(t), colnames(t)] <- t
[…]

### 2014-07-01 maps/rworldmaps_measles.R
Clean(er) map code with silly comma fix
Deleted large sections which did not map appropriate data

@@ -1,3 +1,5 @@
+rm(list=ls())
+
 library(maps)       # Provides functions that let us plot the maps
 library(mapdata)    # Contains the hi-resolution points that mark out the countries.
 library(rworldmap)
@@ -17,54 +19,6 @@ library(mapproj)
 library(plyr)
 library(grid)
 library(gridExtra)
-##
-
-world_map <- map_data("world")
-world.ggmap <- fortify(world_map, region = "NAME")
-
-n <- length(unique(world.ggmap$region))
-df <- data.frame(id = unique(world.ggmap$region),
-                 growth = 4*runif(n),
-                 category = factor(sample(1:5, n, replace=T)))
-
-## noise
-df[c(sample(1:100,40)),c("growth", "category")] <- NA
-
-
-ggplot(df, aes(map_id = id)) +
-  geom_map(aes(fill = growth, color = category), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  scale_fill_gradient(low = "red", high = "blue", guide = "colorbar")
-
-map.world <- map_data(map = "world")
-# map = name of map provided by the maps package.
-# These include county, france, italy, nz, state, usa, world, world2.
-str(map.world)
-# how many regions
-length(unique(map.world$region))
-# how many group polygons (some regions have multiple parts)
-length(unique(map.world$group))
-p1 <- ggplot(map.world, aes(x = long, y = lat, group = group))
-p1 <- p1 + geom_polygon() # fill areas
-p1 <- p1 + labs(title = "World, plain")
-#print(p1)
-p2 <- ggplot(map.world, aes(x = long, y = lat, group = group, colour = region))
-p2 <- p2 + geom_polygon() # fill areas
-p2 <- p2 + theme(legend.position="none") # remove legend with fill colours
-p2 <- p2 + labs(title = "World, colour borders")
-#print(p2)
-p3 <- ggplot(map.world, aes(x = long, y = lat, group = group, fill = region))
-p3 <- p3 + geom_polygon() # fill areas
-p3 <- p3 + theme(legend.position="none") # remove legend with fill colours
[…]

### 2014-07-02 interimreport1.Rnw
Revise interim report Rnw with mapdata
Some of the functions for creating maps were not working through Rnw
Seave, so I made separate results files to work from.

The interim report begins to include material from previous reports and
work on R0.

@@
 
+<<echo=FALSE>>=
+# haven't had time to sort out which packages I need
+library(maps)       # Provides functions that let us plot the maps
+library(mapdata)    # Contains the hi-resolution points that mark out the countries.
+library(rworldmap)
+library(adehabitat)
+library(rgdal)
+library(raster)
+library(maps)
+library(maptools)
+library(sp)
+#library(SDMTools)
+library(ggplot2)
+require(rgeos)
+library(ggmap)
+library(mapdata)
+#gpclibPermit()
+library(mapproj)
+library(plyr)
+library(grid)
+library(gridExtra)
+#gpclibPermit()
+library(utils)
+library(rgdal)
+#getwd()
+#setwd("~/data")
+#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
+load("world.ggmap.Rda")
+n <- length(unique(world.ggmap$id))
+id = unique(world.ggmap$id)
+id<-as.data.frame(id)
+dpdf<-read.csv("mapdata.csv",header=T)
+# output top countries
+
+topimmigration <- dpdf[order(-dpdf$immigration),] 
+topimmigration[1:10,c(1,4)]
+
+topincidence <- dpdf[order(-dpdf$incidence),] 
+topincidence[1:10,c(1,2)]
+
+topvaccine <- dpdf[order(dpdf$cover),] 
+topvaccine[1:10,c(1,3)]
+
+toprisk <- dpdf[order(-dpdf$risk),] 
+toprisk[1:10,c(1,5)]
+
+np1<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Immigration") +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+np2<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Incidence (per million)") +
+  geom_map(aes(fill = incidence), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+np3<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Vaccination cover (%)") +
+  geom_map(aes(fill = cover), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
+
+#grid.arrange(np1, np2, np3, ncol=3, main="Measles")
+
+np4<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Risk 2012") +
+  geom_map(aes(fill = risk), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
+#np4
+
+#grid.arrange(np4, np1, np2, np3, ncol=2, main="Measles (2012)")
+
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 4>>=
+print(grid.arrange(np1, np2, np3, ncol=1, main="Measles (2012)"))
+@
+\end{center}
+\caption{Global risk data: Immigration numbers (top); Measles incidence (per million) (middle); Vaccination coverage (bottom)}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 10>>=
+print(np4)
+@
+\end{center}
+\caption{Risk map from measles incidence and immigration rates}
+\end{figure}
+
 \section{Cost analyses}
@@
 
-<<frequentist>>=
-library(mcmc)
-data(logit)
-out <- glm(y ~ x1 + x2 + x3 + x4, data = logit,
-    family = binomial())
-summary(out)
-@
+\section{Modeling measles epidemics}
 
-...text
+A previously-published model of the dynamics of measles infections
+in New Zealand is used to evaluate the present vaccination strategy of
+MMR1 at 15 months and MMR2 before 5 years. The results show that
+achieving coverage of 90\% or better at both vaccination opportunities is
+necessary if future epidemics of measles are to be prevented.
 
-<<log.unnormalized.posterior>>=
-x <- logit
-x$y <- NULL
-x <- as.matrix(x)
-x <- cbind(1, x)
-dimnames(x) <- NULL
+The original mathematical model for the dynamics of measles in New Zealand
+was prepared in 1996 (Tobias & Roberts 1998). It successfully predicted the
+1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
+et al 1998; Roberts & Tobias 2000). A subsequent extension of this work in
+1998 showed that the then current schedule of MMR1 at 15 months and MMR2
+at 11 years was insufficient to prevent further epidemics. The schedule was
+changed in 2000 with MMR2 now being administered before 5 years (Anon.
+2002a).
 
-y <- logit$y
+A variety of similar models for measles vaccination strategies have been developed
+by other authors for various regions (Agur et al 1993; Babad et al 1995;
+Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001). These have invariably
+been based on sets of nonlinear differential equations, and the conclusions
+reached have been similar. The differences in the models have been in the details
+of the representation of the infectious period, and in the ways in which
+the age and contact structures of the population have been specified. As the
+time-course of a measles epidemic is short compared to that of demographic
+changes, details of the epidemic process are not relevant when examining population
+level changes over many years. In fact, it has been shown (Diekmann
+& Heesterbeek 2000) that the number infected in an epidemic depends only
+on the basic reproduction number (see below), and the numbers that are initially
+susceptible in each sub-population. It is therefore important to have a
+realistic representation of the population contact structure. The Royal Society
+(UK) report on the 2001 epidemic of Foot and Mouth Disease advocated that
+contact structures for infectious diseases should be described before epidemics
+occur, so that models may be speedily prepared. The report presented the
+New Zealand measles model as an example of the successful use of modelling
+in epidemiology (Anon. 2002b).
 
-lupost <- function(beta, x, y) {
-    eta <- x %*% beta
-    p <- 1 / (1 + exp(- eta))   # note: works even when eta is Inf or -Inf
-    logl <- sum(log(p[y == 1])) + sum(log(1 - p[y == 0]))
-    return(logl + sum(dnorm(beta, 0, 2, log = TRUE)))
-}
-@
+The quantity that determines whether an epidemic will occur is the basic
+reproduction number of the infection, R0. This is defined as the expected
+number of secondary infections that would arise from a single primary infection
+introduced into a fully susceptible population (Anderson & May 1991;
+Diekmann & Heesterbeek 2000). Clearly if R0 > 1 an epidemic will occur following
+an introduction of infection. The best estimate we had for measles in
+New Zealand was R0 = 12.8, the change in the birth rate (56780 p.a. in 2004,
+www.stats.govt.nz) could have reduced this slightly to R0 = 12.5 (see results
+section). Recall that R0 is calculated in the absence of control measures. We
+are interested in two related quantities:
+
+The basic reproduction number of the infection under vaccination, Rv,
+is the expected number of secondary infections that would arise from a
+single primary infection introduced into a vaccinated population at equilibrium.
+This is a robust indicator of the performance of a vaccination
+schedule. If Rv < 1 epidemics are prevented.
+
+The case reproduction number of the infection at time t, Rt, is the
+expected number of secondary infections that arise from a single infection
+at a particular time. This depends on the number in the population who
+are susceptible, either through prior infection or vaccination.
+
+The dynamics of the process may be summarised as follows. Assume that
+the population has Rt < 1 for measles. As the number of susceptibles in the
+population increases, through children that miss their scheduled immunisations,
+then Rt increases. An epidemic occurs when Rt has increased above
+one. During the epidemic the number of susceptibles and hence Rt decreases.
+
+In order to assess the efficacy of the vaccination campaign in the absence of
+precise information on levels of vaccination coverage, we assumed a range of
+plausible coverages at MMR1 and MMR2 and determined the corresponding
+value of Rv for each. For selected combinations we also numerically solved
+the model over a period of 150 years, not as a prediction of what will happen
+over that time period but in order to determine what the expected frequency
+of epidemics would be if this level of vaccination were to be in maintained. In
+all cases, if Rv < 1 epidemics are prevented.
+
+Method
+For this exercise we modelled the New Zealand population in the same way
+as in Roberts & Tobias (2000). We approximated the population structure by
+assuming that the birth rate was constant, and that deaths before the age of
+25 could be neglected. Hence the four age groups active in the epidemic have
+constant size. The age groups were taken as 6 - 15 months, 15 months - 5 years,
+5 years - 11 years and 11 years - 25 years. Hence MMR1 is applied as children
+pass from the first to the second group, and MMR2 from the second to the
+third. The same contact matrix was used as in Roberts & Tobias (2000). For
+a range of vaccination coverages for MMR1 and MMR2 (see Table 1) the basic
+reproduction number of the infection under vaccination, Rv, was calculated.
+In addition, for each combination of coverages the differential equation model
+was solved numerically until the epidemics settled into a regular pattern, and
+the inter-epidemic period was determined.
+
+Calculated values of Rv for the various coverage levels of MMR1 and MMR2
+that were considered are presented in Table 1.
+
+Table 1: The basic reproduction number of the infection under vaccination,
+Rv, for various coverage levels of MMR1 and MMR2. Superscripts refer to
+the figure number where numerical solutions are presented. If Rv < 1 then
+epidemics will not occur.
+
+When no vaccination is applied, the system settles down to an epidemic every
+two years. This corresponds to an estimated R0 = 12.5, similar to that which
+was observed in New Zealand in the 1960s (see Fig.1). The minimum level of
+vaccination investigated (MMR1 = 80\%, MMR2 = 60\%) resulted in Rv = 1.72
+and a pattern of epidemics settling down to one every 12 years (see Fig.2).
+The other combinations of MMR1 and MMR2 coverage levels that resulted
+in Rv > 1 all resulted in epidemics recurring at intervals of between 10 and
+20 years (Figs. 3-7). All other combinations of MMR1 and MMR2 resulted
+in Rv > 1 and the system settling down to a steady uninfected state with no
+epidemics, see Table 1 and Fig.8 for an example.
+
+Figures 1 & 2. The epidemic pattern with no vaccination (top) and with
+MMR1 = 80\%, MMR2 = 60\% (bottom). The continuous curves show the
+cumulative numbers of susceptibles in the population in the age groups 6
+months-15 months, 15 months - 5 years, 5 years - 11 years and 11 years -25
+years, as determined by the model. The spikes show the numbers infectious
+during epidemics (×10).
+
+Figures 3 & 4. The epidemic pattern with with MMR1 = 80\%, MMR2 = 70\%
+(top) and with MMR1 = 80\%, MMR2 = 80\% (bottom). The continuous
+curves show the cumulative numbers of susceptibles in the population in the
+age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
+11 years -25 years, as determined by the model. The spikes show the numbers
+infectious during epidemics (×10).
+
+Figures 5 & 6. The epidemic pattern with MMR1 = 85\%, MMR2 = 60\%
+(top) and with MMR1 = 85\%, MMR2 = 70\% (bottom). The continuous
+curves show the cumulative numbers of susceptibles in the population in the
+age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
+11 years -25 years, as determined by the model. The spikes show the numbers
+infectious during epidemics (×10).
+
+Figures 7 & 8. The epidemic pattern with MMR1 = 90\%, MMR2 = 60\%
+(top) and with MMR1 = 80\%, MMR2 = 85\% (bottom). The continuous
+curves show the cumulative numbers of susceptibles in the population in the
+age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
+11 years -25 years, as determined by the model. The spikes show the numbers
+infectious during epidemics (×10).
+
+The model developed by Roberts & Tobias (2000) supported the change in the
+immunisation schedule that took effect in January 2001, at which time MMR2
+was changed from delivery at 11 years to delivery before the age of five. These
+results were in line with those obtained by other authors, for example: Babad
+et al (1995) advocated a two-dose schedule for England and Wales, with the
+second vaccination given at age four; and Gay et al (1998) recommended a
+second vaccination at either 18 months or five years, to complement the first
+vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
+vaccinating 85\% of susceptible children aged one to seven years at five-yearly
+intervals would prevent epidemics in Israel. These authors all agree that two
+vaccinations at no less than five years apart are necessary to prevent measles
+epidemics.
+A different approach was taken by Wallinga et al (2001). These authors took
+existing policies in eight European countries and estimated the coverage rates
+required to reduce Rv below one. They found that results depended on the
+age at delivery, but no strategy succeeded if coverage rates were below approximately
+87\%. Our results presented in Table 1 are similar where coverage is
+assumed to be the same at MMR1 and MMR2. The results also show the absolute
+necessity of maintaining high coverage rates in order to prevent future
+epidemics. It is difficult to estimate the proportion of the school-age population
+that have been effectively immunised, but this is continuously being
+diluted by children who are not immunised. Wallinga et al (2000) noted that
+in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
+and that under most plausible assumptions for contact rates even
+100\% coverage would be insufficient to prevent epidemics.
+A mathematical model for measles vaccination 11
+Our final conclusion must depend on how much faith is placed in the model.
+The results are consistent with those of other authors, and from Table 1 it
+could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
+to prevent future measles epidemics. However, a study by Glass et al (2004)
+in the Netherlands showed that high overall levels of measles vaccination can
+obscure pockets of poor coverage, resulting in localised regions with increased
+risk of infection. Our results indicate that overall targets or 90% or more will
+need to be achieved to prevent future epidemics in New Zealand.
 
@@
 
-<<metropolis-try-1>>=
-set.seed(42)    # to get reproducible results
-beta.init <- as.numeric(coefficients(out))
-out <- metrop(lupost, beta.init, 1e3, x = x, y = y)
-names(out)
-out$accept
-@
-
-...text
-
-\begin{itemize}
-\item ...text.
-\end{itemize}
-
-...text
-
-\citet{geyer-temp} ...text
-
-<<metropolis-try-2>>=
-out <- metrop(out, scale = 0.1, x = x, y = y)
-out$accept
-out <- metrop(out, scale = 0.3, x = x, y = y)
-out$accept
-out <- metrop(out, scale = 0.5, x = x, y = y)
-out$accept
-out <- metrop(out, scale = 0.4, x = x, y = y)
-out$accept
-@
-
-...text
-
-<<label=metropolis-try-3>>=
-out <- metrop(out, nbatch = 1e4, x = x, y = y)
-out$accept
-out$time
-@
-
-Figure~\ref{fig:fig1} (page~\pageref{fig:fig1})
-shows the time series plot made by the R statement
-<<label=fig1too,include=FALSE>>=
-plot(ts(out$batch))
-@
-\begin{figure}
-\begin{center}
-<<label=fig1,fig=TRUE,echo=FALSE>>=
-<<fig1too>>
-@
-\end{center}
-\caption{Time series plot of MCMC output.}
-\label{fig:fig1}
-\end{figure}
-
-Another way to look at the output is an autocorrelation plot.
-Figure~\ref{fig:fig2} (page~\pageref{fig:fig2})
-shows the time series plot made by the R statement
-<<label=fig2too,include=FALSE>>=
-acf(out$batch)
-@
-\begin{figure}
-\begin{center}
-<<label=fig2,fig=TRUE,echo=FALSE>>=
-<<fig2too>>
-@
-\end{center}
-\caption{Autocorrelation plot of MCMC output.}
-\label{fig:fig2}
-\end{figure}
-
-...text
-\begin{verbatim}
-http://www.stat.umn.edu/~charlie/mcmc/diag.html
-\end{verbatim}
-says
-\begin{quotation}
-...text
-\end{quotation}
-...text
-...text
-\section{Monte Carlo Estimates and Standard Errors}
-
-<<label=metropolis-try-4>>=
-out <- metrop(out, nbatch = 1e2, blen = 100,
-    outfun = function(z, ...) c(z, z^2), x = x, y = y)
-out$accept
-out$time
-@
-
-We have added an argument \verb@outfun@ ...text
-We need to
-use the identity
-$$
-   \var(X) = E(X^2) - E(X)^2
-$$
-...text
-Hence our \verb@outfun@ returns
-\verb@c(z, z^2)@ for an argument (the state vector) \verb@z@.
-
-The \verb@...@ argument to \verb@outfun@ is required, since the
-function is also passed the other arguments (here \verb@x@ and \verb@y@)
-to \verb@metrop@.
-
-\subsection{Simple Means}
-
-The grand means (means of batch means) are
-<<label=metropolis-batch>>=
-apply(out$batch, 2, mean)
-@
-...text
-<<label=metropolis-batch-too>>=
-foo <- apply(out$batch, 2, mean)
-mu <- foo[1:5]
-sigmasq <- foo[6:10] - mu^2
-mu
-sigmasq
-@
-
-...text
-<<label=metropolis-mcse-mu>>=
-mu.mcse <- apply(out$batch[ , 1:5], 2, sd) / sqrt(out$nbatch)
-mu.mcse
-@
-The extra factor \verb@sqrt(out$nbatch)@ arises because the batch means
-have variance $\sigma^2 / b$ where $b$ is the batch length, which is
-\verb@out$blen@,
-whereas the overall means \verb@mu@ have variance $\sigma^2 / n$ where
-$n$ is the total number of iterations, which is \verb@out$blen * out$nbatch@.
-
-\subsection{Functions of Means}
-
-To get the MCSE for the posterior variances we apply the delta method.
-Let $u_i$ denote the sequence of batch means of the first kind for one
-parameter and $\bar{u}$ the grand mean (the estimate of the posterior mean
-of that parameter),
-let $v_i$ denote the sequence of batch means of the second kind for the
-same parameter and $\bar{v}$ the grand mean (the estimate of the posterior
-second absolute moment of that parameter), and let $\mu = E(\bar{u})$ and
-$\nu = E(\bar{v})$.  Then the delta method linearizes the nonlinear function
-$$
-   g(\mu, \nu) = \nu - \mu^2
-$$
-as
-$$
-   \Delta g(\mu, \nu) = \Delta \nu - 2 \mu \Delta \mu
-$$
-saying that
-$$
-   g(\bar{u}, \bar{v}) - g(\mu, \nu)
-$$
-has the same asymptotic normal distribution as
-$$
-   (\bar{v} - \nu) - 2 \mu (\bar{u} - \mu)
-$$
-which, of course, has variance \verb@1 / nbatch@ times that of
-$$
-   (v_i - \nu) - 2 \mu (u_i - \mu)
-$$
-and this variance is estimated by
-$$
-   \frac{1}{n_{\text{batch}}} \sum_{i = 1}^{n_{\text{batch}}}
-   \bigl[ (v_i - \bar{v}) - 2 \bar{u} (u_i - \bar{u}) \bigr]^2
-$$
-So
-<<label=metropolis-mcse-sigmasq>>=
-u <- out$batch[ , 1:5]
-v <- out$batch[ , 6:10]
-ubar <- apply(u, 2, mean)
-vbar <- apply(v, 2, mean)
-deltau <- sweep(u, 2, ubar)
-deltav <- sweep(v, 2, vbar)
-foo <- sweep(deltau, 2, ubar, "*")
-sigmasq.mcse <- sqrt(apply((deltav - 2 * foo)^2, 2, mean) / out$nbatch)
-sigmasq.mcse
-@
-does the MCSE for the posterior variance.
-
-Let's just check that this complicated \verb@sweep@ and \verb@apply@ stuff
-does do the right thing.
-<<label=metropolis-mcse-sigmasq-too>>=
-sqrt(mean(((v[ , 2] - vbar[2]) - 2 * ubar[2] * (u[ , 2] - ubar[2]))^2) /
-    out$nbatch)
-@
-
-\paragraph{Comment} Through version 0.5 of this vignette it contained
-an incorrect procedure for calculating this MCSE, justified by a handwave.
-Essentially, it said to use the standard deviation of the batch means called
-\verb@v@ here, which appears to be very conservative.
-
-\subsection{Functions of Functions of Means}
-
-If we are also interested in the posterior standard deviation
-(a natural question, although not asked on the exam problem),
-the delta method gives its standard error in terms of that
-for the variance
-<<label=metropolis-mcse-sigma>>=
-sigma <- sqrt(sigmasq)
-sigma.mcse <- sigmasq.mcse / (2 * sigma)
-sigma
-sigma.mcse
-@
-
-\section{A Final Run}
-
-...text...
-(the exam problem directed ``use a long enough run of your Markov chain
-sampler so that the MCSE are less than 0.01'')
-<<label=metropolis-try-5>>=
-out <- metrop(out, nbatch = 5e2, blen = 400, x = x, y = y)
-out$accept
-out$time
-<<metropolis-batch-too>>
-<<metropolis-mcse-mu>>
-<<metropolis-mcse-sigmasq>>
-<<metropolis-mcse-sigma>>
-@
-...text... \verb@xtable@ command in the \verb@xtable@ library.
-
-First the posterior means,
-\begin{table}[ht]
-\caption{Posterior Means}
-\label{tab:mu}
-\begin{center}
-<<label=tab1,echo=FALSE,results=tex>>=
-foo <- rbind(mu, mu.mcse)
-dimnames(foo) <- list(c("estimate", "MCSE"),
-    c("constant", paste("$x_", 1:4, "$", sep = "")))
-library(xtable)
-print(xtable(foo, digits = rep(4, 6),
-    align = c("l", rep("c", 5))), floating = FALSE,
-    caption.placement = "top",
-    sanitize.colnames.function = function(x) return(x))
-@
-\end{center}
-\end{table}
-then the posterior variances (table on page~\pageref{tab:sigmasq}),
-\begin{table}[ht]
-\caption{Posterior Variances}
-\label{tab:sigmasq}
-\begin{center}
-<<label=tab1,echo=FALSE,results=tex>>=
-foo <- rbind(sigmasq, sigmasq.mcse)
-dimnames(foo) <- list(c("estimate", "MCSE"),
-    c("constant", paste("$x_", 1:4, "$", sep = "")))
-library(xtable)
-print(xtable(foo, digits = rep(4, 6),
-    align = c("l", rep("c", 5))), floating = FALSE,
-    caption.placement = "top",
-    sanitize.colnames.function = function(x) return(x))
-@
-\end{center}
-\end{table}
-and finally the posterior standard deviations
-(table on page~\pageref{tab:sigma}).
-\begin{table}[ht]
-\caption{Posterior Standard Deviations}
-\label{tab:sigma}
-\begin{center}
-<<label=tab1,echo=FALSE,results=tex>>=
-foo <- rbind(sigma, sigma.mcse)
-dimnames(foo) <- list(c("estimate", "MCSE"),
-    c("constant", paste("$x_", 1:4, "$", sep = "")))
-library(xtable)
-print(xtable(foo, digits = rep(4, 6),
-    align = c("l", rep("c", 5))), floating = FALSE,
-    caption.placement = "top",
-    sanitize.colnames.function = function(x) return(x))
-@
-\end{center}
-\end{table}
-
-Note for ...
-<<label=time,echo=FALSE,results=tex>>=
-cat(out$time[1], "\n")
-@
-...text...
-
-\section{New Variance Estimation Functions}
-
-A new function \texttt{initseq} estimates variances in the Markov chain
-central limit theorem (CLT) following the methodology introduced by
-\citet[Section~3.3]{practical}.  These methods only apply to scalar-valued
-functionals of
-reversible Markov chains, but the Markov chains produced by the \texttt{metrop}
-function satisfy this condition, even, as we shall see below, when batching
-is used.
-
-...text.. AR(1) ...text... \texttt{initseq}.
-<<x>>=
-n <- 2e4
-rho <- 0.99
-x <- arima.sim(model = list(ar = rho), n = n)
-@
-...text...
-
-Define
-\begin{equation} \label{eq:little}
-   \gamma_k = \cov(X_i, X_{i + k})
-\end{equation}
-where the covariances refer to the stationary Markov chain having the
-same transition probabilities as \texttt{x}.  Then the variance in the CLT
-is
-$$
-   \sigma^2 = \gamma_0 + 2 \sum_{k = 1}^\infty \gamma_k
-$$
-\citep[Theorem~2.1]{practical}, that is,
-$$
-   \bar{x}_n \approx \text{Normal}\left(\mu, \frac{\sigma^2}{n}\right),
-$$
-where $\mu = E(X_i)$ is the quantity being estimated by MCMC (in this
-toy problem we know $\mu = 0$).
-
-Naive estimates of $\sigma^2$ obtained by plugging in empirical
-estimates of the gammas do not provide consistent estimation
-\citep[Section~3.1]{practical}.  Thus the scheme implemented
-by the R function \texttt{initseq}.  Define
-\begin{equation} \label{eq:big}
-   \Gamma_k = \gamma_{2 k} + \gamma_{2 k + 1}
-\end{equation}
-\citet[Theorem~3.1]{practical} says that $\Gamma_k$ ...text..
-\citet[Section~3.3]{practical} ...text.. Theorem~3.2.
-
-Figure~\ref{fig:gamma} (page~\pageref{fig:gamma})
-shows the time series plot made by the R statement
-<<label=figgamtoo,include=FALSE>>=
-out <- initseq(x)
-plot(seq(along = out$Gamma.pos) - 1, out$Gamma.pos,
-        xlab = "k", ylab = expression(Gamma[k]), type = "l")
-lines(seq(along = out$Gamma.dec) - 1, out$Gamma.dec, lty = "dotted")
-lines(seq(along = out$Gamma.con) - 1, out$Gamma.con, lty = "dashed")
-@
-\begin{figure}
-\begin{center}
-<<label=figgam,fig=TRUE,echo=FALSE>>=
-<<figgamtoo>>
-@
-\end{center}
-\caption{Plot ``Big Gamma'' defined by \eqref{eq:little} and \eqref{eq:big}.
-Solid line, initial positive sequence estimator.
-Dotted line, initial monotone sequence estimator.
-Dashed line, initial convex sequence estimator.}
-\label{fig:gamma}
-\end{figure}
-...text... Figure~\ref{fig:gamma}, ...text... $\sigma^2$, .... 
-<<assvar>>=
-out$var.con
-(1 + rho) / (1 - rho) * 1 / (1 - rho^2)
-@
-where for comparison we have given the exact theoretical value of $\sigma^2$,
-...text...
-<<batx>>=
-blen <- 5
-x.batch <- apply(matrix(x, nrow = blen), 2, mean)
-bout <- initseq(x.batch)
-@
-Because the batch length is too short, the variance of the batch means
-does not estimate $\sigma^2$.  We must account for the autocorrelation
-of the batches, shown in Figure~\ref{fig:gambat}.
-<<label=figgambattoo,include=FALSE>>=
-plot(seq(along = bout$Gamma.con) - 1, bout$Gamma.con,
-        xlab = "k", ylab = expression(Gamma[k]), type = "l")
-@
-\begin{figure}
-\begin{center}
-<<label=figgambat,fig=TRUE,echo=FALSE>>=
-<<figgambattoo>>
-@
-\end{center}
-\caption{Plot ``Big Gamma'' defined by \eqref{eq:little} and \eqref{eq:big}
-for the sequence of batch means (batch length \Sexpr{blen}).
-Only initial convex sequence estimator is shown.}
-\label{fig:gambat}
-\end{figure}
-...text the $\sigma^2$ for the original series.
-<<compvar>>=
-out$var.con
-bout$var.con * blen
-@
-Another way to look at this is that the MCMC estimator of $\mu$ is
-either \texttt{mean(x)} or \texttt{mean(x.batch)}.  And the variance
-must be divided by the sample size to give standard errors.  So either
-<<ci-con>>=
-mean(x) + c(-1, 1) * qnorm(0.975) * sqrt(out$var.con / length(x))
-mean(x.batch) + c(-1, 1) * qnorm(0.975) * sqrt(bout$var.con / length(x.batch))
-@
-is an asymptotic 95\% confidence interval for $\mu$.  Just divide by
-the relevant sample size.
+\section{Acknowledgments}
+The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR for help in obtaining the appropriate materials for analyses.
 

### 2014-07-02 maps/rworldmaps_measles.R

@@ -30,9 +30,9 @@ library(rgdal)
 getwd()
 setwd("~/data")
 #unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
-world.map <- readOGR(dsn="C:/Users/David Hayman/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
+world.map <- readOGR(dsn="C:/Users/dtshayma/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
 
-setwd("C:/Users/David Hayman/Dropbox/measles/data")
+setwd("C:/Users/dtshayma/Dropbox/measles/data")
 
 incidence<-read.csv("incidence_series.csv",header=T)
 head(incidence)
@@ -49,7 +49,8 @@ head(population)
 #######
 
 world.ggmap <- fortify(world.map, region = "NAME")
-
+save(world.ggmap,file="world.ggmap.Rda")
+load("world.ggmap.Rda")
 n <- length(unique(world.ggmap$id))
 
 id = unique(world.ggmap$id)
@@ -324,6 +325,8 @@ grid.arrange(np1, np2, np3, ncol=3, main="Measles")
 
 dpdf$risk <- with(dpdf, mp$X2012.x/mp$Numeric.Value*1000* mp$Immigration)
 head(dpdf)
+write.csv(dpdf, "mapdata.csv", row.names=FALSE,col.names=T)
+dpdf<-read.csv("mapdata.csv",header=T)
 
 np4<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Risk 2012") +
   geom_map(aes(fill = risk), map =world.ggmap) +

### 2014-07-02 estimatingR0/estR0.R
adds JM version of realtime R0 estimation

@@ -1,0 +1,141 @@
+jm_estR0 <- function (epid, GT, import = NULL, n.t0 = NULL, t = NULL, begin = NULL, 
+    end = NULL, date.first.obs = NULL, time.step = 1, q = c(0.025, 
+        0.975), correct = TRUE, nsim = 10000, checked = FALSE, 
+    ...) 
+{
+    if (nsim < 1000) 
+        warning("Accurate confidence interval for R(t) requires a large number of simulations. Consider increasing 'nsim'")
+    if (nsim > 100) 
+        warning("Simulations may take several minutes.")
+    DNAME = deparse(substitute(epid))
+    CALL = match.call()
+    if (checked == FALSE) {
+        parameters <- integrity.checks(epid, t, GT, begin, end, 
+            date.first.obs, time.step, AR = NULL, S0 = NULL, 
+            methods = "TD")
+        begin <- parameters$begin
+        end <- parameters$end
+    }
+    epid = check.incid(epid, t, date.first.obs, time.step)
+    begin.nb = which(epid$t == begin)
+    end.nb = which(epid$t == end)
+    epid.bak <- epid
+    t <- diff(c(FALSE, epid$incid == 0, FALSE), 1)
+    start <- which(t == 1)
+    end <- which(t == -1)
+    if (length(start) > 0 & length(end) > 0) {
+        longest <- max(end - start)
+        if (longest > length(GT$GT)) 
+            warning(paste("Gap in epidemic curve is longer than the generation interval. Consider using a different GT distribution (maybe with \"truncate=", 
+                longest, "\" (length of longest gap))."), sep = "")
+    }
+    if (is.null(import)) {
+        import <- rep(0, length(epid$incid))
+    }
+    if (!is.null(import) & (length(import) != length(epid$incid))) {
+        stop("Vector of imported cases should have the same length as 'epid' data.")
+    }
+    if (is.null(n.t0)) {
+        n.t0 = epid$incid[1]
+        warning("Using initial incidence as initial number of cases.")
+    }
+    if (n.t0 > epid$incid[1]) {
+        stop(paste("Provided initial number of cases (n.t0=", 
[…]

### 2014-07-02 estimatingR0/estimate_r0.R
adds graphing of R0 through time per epidemic, and average R0

@@ -9,25 +9,115 @@ mu_logn    <- log(mu) - log(1 + (sigma/mu)^2) / 2
 # then exp(rnorm(n, mu_logn, sigma_logn)) simulates from lognormal with the given mean and sd.
 
 # the R0 library can estimate our distribution from incidence data
-incidence <- read.csv("epidemic.csv")
-
 require(R0)
+source("estR0.R")
 
 # generation time in weeks
-gT <- generation.time(type="lognormal", val=c(12, 3.5)/7)
+genTime <- generation.time(type="lognormal", val=c(12, 3.5)/7)
 
-# convert our incidence data to something we can use
-counts <- incidence$Count
-names(counts) <- incidence$Date
+# read out outbreak folder
+outbreak_folder <- "outbreaks"
 
-counts <- counts[1:51]
-estR0<-estimate.R(counts, gT, t=1:51, methods=c("EG", "ML", "TD", "AR", "SB"), pop.size=1300000, nsim=100)
-estR0<-estimate.R(counts, gT, t=1:51, end=51, methods=c("EG"), pop.size=1300000, nsim=100)
-estR0<-estimate.R(counts, gT, t=1:51, end=51, methods=c("ML"), pop.size=1300000, nsim=100)
-estR0<-estimate.R(counts, gT, t=1:51, end=51, methods=c("TD"), pop.size=1300000, nsim=100)
-estR0<-estimate.R(counts, gT, t=1:51, end=51, methods=c("SB"), pop.size=1300000, nsim=100)
-estR0<-estimate.R(counts, gT, t=1:51, end=51, methods=c("TD"), pop.size=1300000, nsim=1000)
-pdf("rt.pdf", width=10, height=6)
-plot(estR0$estimates$TD, TD.split=T)
+# read in and reorder by date
+outbreak_files <- list.files(path=outbreak_folder, pattern="*.csv")
+
+max_dates <- rep("", length(outbreak_files))
+for (i in 1:length(outbreak_files))
+{
+  incidence <- read.csv(file.path(outbreak_folder, outbreak_files[i]), stringsAsFactors=F)
+  max_dates[i] <- max(incidence$date)
+}
+
+outbreak_files <- outbreak_files[order(max_dates)]
+
+# function for modifying colours to make them transparent
+alpha <- function(col, a)
+{
+  if (nchar(col) == 9)
[…]

### 2014-07-02 interimreport1.Rnw
Revised Sweave report for work at home
Better plot dimensions too. Will include JM's work later tonight.

@@
 
-\title{Measles risk assessment and cost interim report\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Nigel P French}}
+\title{Measles risk assessment and cost interim report\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
 \author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
@@
 The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011/2012 (mainly in Auckland region) and another significant outbreak is currently going on in Auckland region.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011/2012 (mainly in Auckland region) and another significant outbreak is currently going on in the Auckland and Waikato regions.
 The outbreak in 2011-2012 lasted for more than 12 months and the current 2013/2014 outbreak started at the end of December 2013 (as on \date{\today}).
@@
 \emph{Figures 11-13}).
-Three figures,\emph{Figures 12, 13, and 28}, show that there is spatial clustering of cases.
-
-[INSERT NEED FOR MULTIVARIATE ANALYSES HERE?]
+Three figures,\emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
 
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with this statement, but have included additional analyses on the current outbreak that [DEMONSTRATE/SUGGEST ...INSERT THE CONCLUSIONS RE R0]
+We agree with the statement that measles is not endemic and provide some preliminary analyses on the outbreaks since endemic measles 'elimination' (see section X) that gives information regarding the likelihood of this occuring.
 
-Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007) and only Asian and Pacific ethnicities have consistently had MMR dose one coverage approching or over 95\% levels for cohorts from 2007 onwards.
-[COMMENT ON AGREEMENT HERE?] Furthermore, ...
-[COMMENT ON REFUSAL RATE?]
+We agree with the report's conclusions that testing for measles is performed appropriately within the required timeframe, but clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions.
 
-The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases.
-[COMMENT?
-Unsure, though don't want to be dismissive].
+Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
+However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approching or over 95\% levels for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
+This is particularly in light of our modeling results (section X).
 
-The development of a National Immunisation Register (NIR) that allows tracking of the vaccination status of children is very useful.
-[COMMENT ON NEED FOR FINER IMMUNISATION LEVEL DATA HERE?].
-The report conludes that the surveillance performance indicators demonstrate a reasonable measles surveillance capability in New Zealand.
-We [COMMENT ON THIS, GIVEN JONATHAN'S FINDINGS]?.
-
-We agree with the report's conclusions that testing for measles is performed appropriately within the required timeframe, but clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary.
+The analyses we thought could further inform the understanding of risk from measles infection were:
+\begin{itemize}
+\item Multivariate modelling to account for confounding within the univariate analyses.
+\item Inclusion of multivariate model results with local level immunity data to develop a risk map for NZ. 
+\item Analyses of changing risk factors through time during outbreaks.
+\item Modelling transmission chains in the population to understand effective reproduction number.
+\item Development of risk maps to understand measles importation.
+\end{itemize}
 
@@
 
-In this section we provide ... [].
+In this section we provide a brief description of the work we are performing that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
+These analyses are intended to build on the excellent analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
+
+We received the raw case data from The Institute of Environmental Science and Research Ltd (ESR) EpiSurv on 27 June 2014.
+Initial analyses of those data suggest that we require denominator data to perform multivariate analyses.
+Specifically for the multivariate analyses we wish to perform we require \emph{Age * Prioritized Ethnicity * NZDep} data for New Zealand to see if the interactions among case covariates give us any additional information on risk from the univariate analyses already performed in the report reviewed above.
+
+These data are currently unavailable in the form we require and after contacting ESR and NZ Statistics we and NZ statistics have contacted the University of Otago to see if we can obtain these denominator data.
+Additional data we believe would enable us and the Ministry to understand measles risk better is fine scale (lower than DHB) immunization coverage data.
+
+We have reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help better understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers, we sought to understand better the risk of measles importation.
+For our initial analyses, we used the arrivals data from \href{http://www.immigration.govt.nz/}{www.immigration.govt.nz} as an estimate of human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO \href{http://www.who.int/research/en/}{www.who.int/research/en/}.
+
+<<echo=FALSE>>=
+library(xtable)
+dpdf<-read.csv("mapdata.csv",header=T)
+# output top countries
+topimmigration <- dpdf[order(-dpdf$immigration),] 
+immigration<-topimmigration[1:10,c(1,4)]
+topincidence <- dpdf[order(-dpdf$incidence),] 
+incidence<-topincidence[1:10,c(1,2)]
+incidence[,2]<-round(incidence[,2],0)
+topvaccine <- dpdf[order(dpdf$cover),] 
+vaccinecover<-topvaccine[1:10,c(1,3)]
+toprisk <- dpdf[order(-dpdf$risk),] 
+risk<-(toprisk[1:10,c(1,5)])
+risk[,2]<-round(risk[,2],0)
+@
+
+\begin{table}
+\caption{Immigration to New Zealand (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(immigration, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+\begin{table}
+\caption{Measles Incidence per million (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(incidence, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(risk, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+The preliminary analyses of 2012 data suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table 1.
+However, vaccination coverage is lowest and measles incidence highest in less developed nations (Tables 2 and 3).
+Though the interactions between these different risk factors are unknown, simple products of measles incidence in 2012 and immigration numbers in 2012 suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some countries also poses a risk.
+These data are shown in Table 4.
+The data for all the variables for each nation state and territories for 2012 are plotted in Figure 1 and the risk map for measles incidence and immigration in Figure 2.
 
@@
 
-topimmigration <- dpdf[order(-dpdf$immigration),] 
-topimmigration[1:10,c(1,4)]
-
-topincidence <- dpdf[order(-dpdf$incidence),] 
-topincidence[1:10,c(1,2)]
-
-topvaccine <- dpdf[order(dpdf$cover),] 
-topvaccine[1:10,c(1,3)]
-
-toprisk <- dpdf[order(-dpdf$risk),] 
-toprisk[1:10,c(1,5)]
-
 np1<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Immigration") +
@@
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
   scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
@@
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
   scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
@@
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
   scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
 
-#grid.arrange(np1, np2, np3, ncol=3, main="Measles")
+#grid.arrange(np1, np2, np3, ncol=3, main="")
 
-np4<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Risk 2012") +
+np4<-ggplot(dpdf, aes(map_id = id)) +
   geom_map(aes(fill = risk), map =world.ggmap) +
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
   scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
 #np4
-
-#grid.arrange(np4, np1, np2, np3, ncol=2, main="Measles (2012)")
-
 @
@@
 \begin{figure}[h!]
+\caption{Global risk data 2012: Immigration numbers to New Zealand (top); Measles incidence (per million) (middle); Vaccination coverage (bottom)}
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 4>>=
-print(grid.arrange(np1, np2, np3, ncol=1, main="Measles (2012)"))
+<<fig=TRUE,echo=FALSE,width = 5>>=
+print(grid.arrange(np1, np2, np3, ncol=1, main="2012"))
 @
 \end{center}
-\caption{Global risk data: Immigration numbers (top); Measles incidence (per million) (middle); Vaccination coverage (bottom)}
 \end{figure}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 10>>=
+<<fig=TRUE,echo=FALSE,width = 11>>=
 print(np4)
@@
 
+\section{Future risk analyses}
+We received ESR EpiSurv measles case data on the 27 June 2014. The following are analyses are still in progress and aim to include them in later reports:
+\begin{itemize}
+\item Multivariate regression analyses to look for interactions between risk factors that may confound the univariate analyses.
+\item Update of the importation risk analyses using more than the 2012 data.
+\end{itemize}
+The data gaps that we have that will hinder us in completing all the analyses are:
+\begin{itemize}
+\item A lack of meshblock level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
+We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
+\emph{meshblock}) level would allow better understanding of risk and resource allocation.
+\item A lack of data at any level that combines NZDep with age, prioritized ethnicity, and other variables, that allow better use of data collected by the Ministry, ESR, and NZ statistics for multivariate analyses.
+\end{itemize}
+
 \section{Cost analyses}
 
-In this section we provide ... [].
+In this section we provide an analysis of the costs involved with this current measles outbreak.
+
+... [TEXT HERE] ...
 
 \section{Modeling measles epidemics}
+
+...The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases.
+[COMMENT?
+Unsure, though don't want to be dismissive].
+....
 
@@
 
-We received The Institute of Environmental Science and Research Ltd (ESR) EpiSurv measles case data on the 27 June 2014. Additional analyses we wish to do are:
-\begin{itemize}
-\item ...[]
-\end{itemize}
-
-\section{Future analyses}
-
-We received The Institute of Environmental Science and Research Ltd (ESR) EpiSurv measles case data on the 27 June 2014. Additional analyses we wish to do are:
-\begin{itemize}
-\item ...[]
-\end{itemize}
-
 \section{Acknowledgments}

### 2014-07-02 interimreport1.Rnw
Summary data, figures, and revised report code
Include summary data and figures in PDF form to work into the report
without having to reload the data.

@@
 \usepackage[utf8]{inputenc}
+\usepackage{graphicx}
 
@@
 
+\section{Future cost-benefit analyses}
+Using the results above we aim to:
+\begin{itemize}
+\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date and in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified if the multivariate analyses can be performed once appropriate denominator data are obtained.
+\end{itemize}
+
 \section{Modeling measles epidemics}
 
-...The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases.
-[COMMENT?
-Unsure, though don't want to be dismissive].
-....
-
 A previously-published model of the dynamics of measles infections
-in New Zealand is used to evaluate the present vaccination strategy of
+in New Zealand has been used to evaluate the vaccination strategy in New Zealand of
 MMR1 at 15 months and MMR2 before 5 years. The results show that
-achieving coverage of 90\% or better at both vaccination opportunities is
-necessary if future epidemics of measles are to be prevented.
+achieving coverage of >90\% at both vaccination opportunities is
+necessary if future epidemics of measles are to be prevented. [AND LIKELY MORE?!]
 
@@
 of the representation of the infectious period, and in the ways in which
-the age and contact structures of the population have been specified. As the
-time-course of a measles epidemic is short compared to that of demographic
-changes, details of the epidemic process are not relevant when examining population
-level changes over many years. In fact, it has been shown (Diekmann
-& Heesterbeek 2000) that the number infected in an epidemic depends only
-on the basic reproduction number (see below), and the numbers that are initially
-susceptible in each sub-population. It is therefore important to have a
-realistic representation of the population contact structure. The Royal Society
-(UK) report on the 2001 epidemic of Foot and Mouth Disease advocated that
-contact structures for infectious diseases should be described before epidemics
-occur, so that models may be speedily prepared. The report presented the
-New Zealand measles model as an example of the successful use of modelling
-in epidemiology (Anon. 2002b).
+the age and contact structures of the population have been specified.
+
+The model developed by Roberts & Tobias (2000) supported the change in the
+immunisation schedule that took effect in January 2001, at which time MMR2
+was changed from delivery at 11 years to delivery before the age of five. These
+results were in line with those obtained by other authors, for example: Babad
+et al (1995) advocated a two-dose schedule for England and Wales, with the
+second vaccination given at age four; and Gay et al (1998) recommended a
+second vaccination at either 18 months or five years, to complement the first
+vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
+vaccinating 85\% of susceptible children aged one to seven years at five-yearly
+intervals would prevent epidemics in Israel. These authors all agree that two
+vaccinations at no less than five years apart are necessary to prevent measles
+epidemics.
+A different approach was taken by Wallinga et al (2001). These authors took
+existing policies in eight European countries and estimated the coverage rates
+required to reduce Rv below one. They found that results depended on the
+age at delivery, but no strategy succeeded if coverage rates were below approximately
+87\%. Our results presented in Table 1 are similar where coverage is
+assumed to be the same at MMR1 and MMR2. The results also show the absolute
+necessity of maintaining high coverage rates in order to prevent future
+epidemics. It is difficult to estimate the proportion of the school-age population
+that have been effectively immunised, but this is continuously being
+diluted by children who are not immunised. Wallinga et al (2000) noted that
+in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
+and that under most plausible assumptions for contact rates even
+100\% coverage would be insufficient to prevent epidemics.
+Our final conclusion must depend on how much faith is placed in the model.
+The results are consistent with those of other authors, and from Table 1 it
+could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
+to prevent future measles epidemics. However, a study by Glass et al (2004)
+in the Netherlands showed that high overall levels of measles vaccination can
+obscure pockets of poor coverage, resulting in localised regions with increased
+risk of infection. Our results indicate that overall targets or 90\% or more will
+need to be achieved to prevent future epidemics in New Zealand.
 
@@
 
-In order to assess the efficacy of the vaccination campaign in the absence of
-precise information on levels of vaccination coverage, we assumed a range of
-plausible coverages at MMR1 and MMR2 and determined the corresponding
-value of Rv for each. For selected combinations we also numerically solved
-the model over a period of 150 years, not as a prediction of what will happen
-over that time period but in order to determine what the expected frequency
-of epidemics would be if this level of vaccination were to be in maintained. In
-all cases, if Rv < 1 epidemics are prevented.
-
-Method
-For this exercise we modelled the New Zealand population in the same way
-as in Roberts & Tobias (2000). We approximated the population structure by
-assuming that the birth rate was constant, and that deaths before the age of
-25 could be neglected. Hence the four age groups active in the epidemic have
-constant size. The age groups were taken as 6 - 15 months, 15 months - 5 years,
-5 years - 11 years and 11 years - 25 years. Hence MMR1 is applied as children
-pass from the first to the second group, and MMR2 from the second to the
-third. The same contact matrix was used as in Roberts & Tobias (2000). For
-a range of vaccination coverages for MMR1 and MMR2 (see Table 1) the basic
-reproduction number of the infection under vaccination, Rv, was calculated.
-In addition, for each combination of coverages the differential equation model
-was solved numerically until the epidemics settled into a regular pattern, and
-the inter-epidemic period was determined.
-
 Calculated values of Rv for the various coverage levels of MMR1 and MMR2
@@
 
-Table 1: The basic reproduction number of the infection under vaccination,
-Rv, for various coverage levels of MMR1 and MMR2. Superscripts refer to
-the figure number where numerical solutions are presented. If Rv < 1 then
-epidemics will not occur.
+\begin{figure}[p]
+     \centering
+     \includegraphics[width=0.8\textwidth]{averageR0.pdf}
+     \caption{Test image}
+     \label{fig:test image}
+\end{figure}
 
-When no vaccination is applied, the system settles down to an epidemic every
-two years. This corresponds to an estimated R0 = 12.5, similar to that which
-was observed in New Zealand in the 1960s (see Fig.1). The minimum level of
-vaccination investigated (MMR1 = 80\%, MMR2 = 60\%) resulted in Rv = 1.72
-and a pattern of epidemics settling down to one every 12 years (see Fig.2).
-The other combinations of MMR1 and MMR2 coverage levels that resulted
-in Rv > 1 all resulted in epidemics recurring at intervals of between 10 and
-20 years (Figs. 3-7). All other combinations of MMR1 and MMR2 resulted
-in Rv > 1 and the system settling down to a steady uninfected state with no
-epidemics, see Table 1 and Fig.8 for an example.
-
-Figures 1 & 2. The epidemic pattern with no vaccination (top) and with
-MMR1 = 80\%, MMR2 = 60\% (bottom). The continuous curves show the
-cumulative numbers of susceptibles in the population in the age groups 6
-months-15 months, 15 months - 5 years, 5 years - 11 years and 11 years -25
-years, as determined by the model. The spikes show the numbers infectious
-during epidemics (×10).
-
-Figures 3 & 4. The epidemic pattern with with MMR1 = 80\%, MMR2 = 70\%
-(top) and with MMR1 = 80\%, MMR2 = 80\% (bottom). The continuous
-curves show the cumulative numbers of susceptibles in the population in the
-age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
-11 years -25 years, as determined by the model. The spikes show the numbers
-infectious during epidemics (×10).
-
-Figures 5 & 6. The epidemic pattern with MMR1 = 85\%, MMR2 = 60\%
-(top) and with MMR1 = 85\%, MMR2 = 70\% (bottom). The continuous
-curves show the cumulative numbers of susceptibles in the population in the
-age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
-11 years -25 years, as determined by the model. The spikes show the numbers
-infectious during epidemics (×10).
-
-Figures 7 & 8. The epidemic pattern with MMR1 = 90\%, MMR2 = 60\%
-(top) and with MMR1 = 80\%, MMR2 = 85\% (bottom). The continuous
-curves show the cumulative numbers of susceptibles in the population in the
-age groups 6 months-15 months, 15 months - 5 years, 5 years - 11 years and
-11 years -25 years, as determined by the model. The spikes show the numbers
-infectious during epidemics (×10).
-
-The model developed by Roberts & Tobias (2000) supported the change in the
-immunisation schedule that took effect in January 2001, at which time MMR2
-was changed from delivery at 11 years to delivery before the age of five. These
-results were in line with those obtained by other authors, for example: Babad
-et al (1995) advocated a two-dose schedule for England and Wales, with the
-second vaccination given at age four; and Gay et al (1998) recommended a
-second vaccination at either 18 months or five years, to complement the first
-vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
-vaccinating 85\% of susceptible children aged one to seven years at five-yearly
-intervals would prevent epidemics in Israel. These authors all agree that two
-vaccinations at no less than five years apart are necessary to prevent measles
-epidemics.
-A different approach was taken by Wallinga et al (2001). These authors took
-existing policies in eight European countries and estimated the coverage rates
-required to reduce Rv below one. They found that results depended on the
-age at delivery, but no strategy succeeded if coverage rates were below approximately
-87\%. Our results presented in Table 1 are similar where coverage is
-assumed to be the same at MMR1 and MMR2. The results also show the absolute
-necessity of maintaining high coverage rates in order to prevent future
-epidemics. It is difficult to estimate the proportion of the school-age population
-that have been effectively immunised, but this is continuously being
-diluted by children who are not immunised. Wallinga et al (2000) noted that
-in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
-and that under most plausible assumptions for contact rates even
-100\% coverage would be insufficient to prevent epidemics.
-A mathematical model for measles vaccination 11
-Our final conclusion must depend on how much faith is placed in the model.
-The results are consistent with those of other authors, and from Table 1 it
-could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
-to prevent future measles epidemics. However, a study by Glass et al (2004)
-in the Netherlands showed that high overall levels of measles vaccination can
-obscure pockets of poor coverage, resulting in localised regions with increased
-risk of infection. Our results indicate that overall targets or 90% or more will
-need to be achieved to prevent future epidemics in New Zealand.
+...The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases.
+[COMMENT?
+Unsure, though don't want to be dismissive].
+....
 
 \section{Future analyses}
+
+\section{Future cost-benefit analyses}
+Using the results above we aim to:
+\begin{itemize}
+\item Update previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
+\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
+\end{itemize}
 

### 2014-07-02 interimreport1.tex
R0 figures included
Included R0 estimation and figures in the report text.

--- 2014-07-02 interimreport1.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{interimreport1-concordance}
 
@@
 
-<<echo=FALSE>>=
-library(xtable)
-dpdf<-read.csv("mapdata.csv",header=T)
-# output top countries
-topimmigration <- dpdf[order(-dpdf$immigration),] 
-immigration<-topimmigration[1:10,c(1,4)]
-topincidence <- dpdf[order(-dpdf$incidence),] 
-incidence<-topincidence[1:10,c(1,2)]
-incidence[,2]<-round(incidence[,2],0)
-topvaccine <- dpdf[order(dpdf$cover),] 
-vaccinecover<-topvaccine[1:10,c(1,3)]
-toprisk <- dpdf[order(-dpdf$risk),] 
-risk<-(toprisk[1:10,c(1,5)])
-risk[,2]<-round(risk[,2],0)
-@
-
 \begin{table}
 \caption{Immigration to New Zealand (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(immigration, file="", table.env=FALSE,rowname=NULL)
-@
-\end{table}
+%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{id}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$809775$\tabularnewline
+United Kingdom&$306177$\tabularnewline
+China&$256036$\tabularnewline
+United States&$194438$\tabularnewline
+Japan&$ 86676$\tabularnewline
+Germany&$ 83608$\tabularnewline
+Korea, Republic of&$ 73459$\tabularnewline
+France&$ 71448$\tabularnewline
+India&$ 69038$\tabularnewline
+Canada&$ 54981$\tabularnewline
+\hline
+\end{tabular}\end{center}\end{table}
 
@@
 \caption{Lowest national measles vaccine cover (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
-@
-\end{table}
+%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{id}&\multicolumn{1}{c}{cover}\tabularnewline
+\hline
+Equatorial Guinea&$34$\tabularnewline
+Somalia&$49$\tabularnewline
+Lesotho&$60$\tabularnewline
+Central African Republic&$65$\tabularnewline
+Papua New Guinea&$67$\tabularnewline
+Chad&$69$\tabularnewline
+Haiti&$69$\tabularnewline
+South Sudan&$70$\tabularnewline
+Gabon&$71$\tabularnewline
+Yemen&$71$\tabularnewline
+\hline
+\end{tabular}\end{center}\end{table}
 
@@
 \caption{Measles Incidence per million (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(incidence, file="", table.env=FALSE,rowname=NULL)
-@
-\end{table}
+%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{id}&\multicolumn{1}{c}{incidence}\tabularnewline
+\hline
+Equatorial Guinea&$1617$\tabularnewline
+Nauru&$1100$\tabularnewline
+Democratic Republic of the Congo&$1096$\tabularnewline
+Somalia&$ 979$\tabularnewline
+Djibouti&$ 824$\tabularnewline
+Sudan&$ 786$\tabularnewline
+Burkina Faso&$ 447$\tabularnewline
+Romania&$ 342$\tabularnewline
+Ukraine&$ 280$\tabularnewline
+Sudan&$ 229$\tabularnewline
+\hline
+\end{tabular}\end{center}\end{table}
 
@@
 \caption{Risk of measles importation to New Zealand in 2012}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(risk, file="", table.env=FALSE,rowname=NULL)
-@
-\end{table}
+%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{id}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+United Kingdom&$10202161$\tabularnewline
+Australia&$ 6991116$\tabularnewline
+Malaysia&$ 3226580$\tabularnewline
+Thailand&$ 1576414$\tabularnewline
+China&$ 1138900$\tabularnewline
+India&$ 1039356$\tabularnewline
+Indonesia&$  984022$\tabularnewline
+Ukraine&$  664315$\tabularnewline
+Ireland&$  585413$\tabularnewline
+Romania&$  490731$\tabularnewline
+\hline
+\end{tabular}\end{center}\end{table}
 
@@
 
-<<echo=FALSE>>=
-# haven't had time to sort out which packages I need
-library(maps)       # Provides functions that let us plot the maps
-library(mapdata)    # Contains the hi-resolution points that mark out the countries.
-library(rworldmap)
-library(adehabitat)
-library(rgdal)
-library(raster)
-library(maps)
-library(maptools)
-library(sp)
-#library(SDMTools)
-library(ggplot2)
-require(rgeos)
-library(ggmap)
-library(mapdata)
-#gpclibPermit()
-library(mapproj)
-library(plyr)
-library(grid)
-library(gridExtra)
-#gpclibPermit()
-library(utils)
-library(rgdal)
-#getwd()
-#setwd("~/data")
-#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
-load("world.ggmap.Rda")
-n <- length(unique(world.ggmap$id))
-id = unique(world.ggmap$id)
-id<-as.data.frame(id)
-dpdf<-read.csv("mapdata.csv",header=T)
-# output top countries
-
-np1<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Immigration") +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np2<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Incidence (per million)") +
-  geom_map(aes(fill = incidence), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np3<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Vaccination cover (%)") +
-  geom_map(aes(fill = cover), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
-
-#grid.arrange(np1, np2, np3, ncol=3, main="")
-
-np4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-@
-
 \begin{figure}[h!]
-\caption{Global risk data 2012: Immigration numbers to New Zealand (top); Measles incidence (per million) (middle); Vaccination coverage (bottom)}
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 5>>=
-print(grid.arrange(np1, np2, np3, ncol=1, main="2012"))
-@
+\includegraphics{interimreport1-007}
 \end{center}
+\caption{Immigration to New Zealand 2012}
 \end{figure}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np4)
-@
+\includegraphics{interimreport1-008}
+\end{center}
+\caption{Measles incidence per million 2012}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport1-009}
+\end{center}
+\caption{Measles vaccination cover (\%) 2012 }
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport1-010}
 \end{center}
@@
 
-\begin{figure}[p]
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
+     \caption{Measles incidence from 1997 to 2014}
+     %\label{fig:test image}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{cases_by_dhb_2009_2014.pdf}
+     \caption{Measles cases by DHB from 2009 to 2014}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
+     \caption{Measles data classified as outbreaks for R0 estimation}
+     \label{fig:outbreaks}
+\end{figure}
+
+\begin{figure}
      \centering
      \includegraphics[width=0.8\textwidth]{averageR0.pdf}
-     \caption{Test image}
-     \label{fig:test image}
+     \caption{Estimates of R0 for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
 \end{figure}

### 2014-07-02 interimreport1.Rnw
Times series immigration to NZ and global measles
Minor revisions to some other code and analyses of the global change in
measles incidence and changes in NZ immigration

@@
 
-np1<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Immigration") +
+np1<-ggplot(dpdf, aes(map_id = id))  +
   geom_map(aes(fill = immigration), map =world.ggmap) +
@@
   ylab("latitude") +
+   theme(text = element_text(size=30))+
   scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
 
-np2<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Incidence (per million)") +
+np2<-ggplot(dpdf, aes(map_id = id))  +
   geom_map(aes(fill = incidence), map =world.ggmap) +
@@
   ylab("latitude") +
+   theme(text = element_text(size=30))+
   scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
 
-np3<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Vaccination cover (%)") +
+np3<-ggplot(dpdf, aes(map_id = id))  +
   geom_map(aes(fill = cover), map =world.ggmap) +
@@
   ylab("latitude") +
+  theme(text = element_text(size=30))+
   scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
@@
 \begin{figure}[h!]
-\caption{Global risk data 2012: Immigration numbers to New Zealand (top); Measles incidence (per million) (middle); Vaccination coverage (bottom)}
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 5>>=
-print(grid.arrange(np1, np2, np3, ncol=1, main="2012"))
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np1)
 @
 \end{center}
+\caption{Immigration to New Zealand 2012}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np2)
+@
+\end{center}
+\caption{Measles incidence per million 2012}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np3)
+@
+\end{center}
+\caption{Measles vaccination cover (\%) 2012 }
 \end{figure}
@@
 \end{figure}
+
+Though [SOMETHING ON MEASLES], immigration rates to New Zealand are increasing. [INSERT FIGURE]
+
+<<fig=TRUE,echo=FALSE>>=
+ttimes<-read.csv("times.csv",header=T)
+lmres<-lm(res~time,data=ttimes)
+summary(lmres)
+plot(ttimes$res,xaxt="n",type="l",xlab="Date",ylab="")
+axis(1, at=1:71, labels=ttimes$mydate) 
+abline(lmres,col="red",
+       lty=2)
+acf(ttimes$res,lag.max = length(ttimes$res),plot=T,main="")
+@
 

### 2014-07-02 maps/rworldmaps_measles.R

@@ -30,9 +30,9 @@ library(rgdal)
 getwd()
 setwd("~/data")
 #unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
-world.map <- readOGR(dsn="C:/Users/dtshayma/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
+world.map <- readOGR(dsn="C:/Users/David Hayman/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
 
-setwd("C:/Users/dtshayma/Dropbox/measles/data")
+setwd("C:/Users/David Hayman/Dropbox/measles/data")
 
 incidence<-read.csv("incidence_series.csv",header=T)
 head(incidence)

### 2014-07-02 timeseries_immigration.R

@@ -1,0 +1,74 @@
+setwd("C:/Users/David Hayman/Dropbox/measles/data")
+
+immigration<-read.csv("immigration.csv",header=T)
+head(immigration)
+
+summary(immigration$Nationality,maxsum=Inf)
+## match immigration names to IDs from maps
+levels(immigration$Nationality)[match(c("Antigua & Barbuda","Bosnia & Herzegovina","Myanmar","Ivory Coast",
+                                        "Faeroe Islands","Guinea - Bissau","Vatican City","Iran","North Korea",
+                                        "South Korea","Laos","Libya","Federated States of Micronesia","Moldova",
+                                        "St Kitts - Nevis","St Lucia","St Vincent and the Grenadines","Sao Tome & Principe",
+                                        "Serbia & Montenegro","South Sudan","Syria","Timor Leste","Turkemenistan","Great Britain",
+                                        "United States of America","Virgin Islands, USA","Vietnam"),levels(immigration$Nationality))] <-
+  c("Antigua and Barbuda","Bosnia and Herzegovina","Burma","Cote d'Ivoire",
+    "Faeroe Islands","Guinea-Bissau","Holy See (Vatican City)","Iran (Islamic Republic of)","Korea, Democratic People's Republic of",
+    "Korea, Republic of","Lao People's Democratic Republic","Libyan Arab Jamahiriya","Micronesia, Federated States of","Republic of Moldova",
+    "Saint Kitts and Nevis","Saint Lucia","Saint Vincent and the Grenadines","Sao Tome and Principe","Serbia","Sudan","Syrian Arab Republic",
+    "Timor-Leste","Turkmenistan","United Kingdom","United States","United States Virgin Islands","Viet Nam")
+## to use the travel data...
+names(immigration)
+head(immigration)
+class(immigration)
+
+## silly dataset contains weird comma's that R doesn't detect as the thousands separator.
+## This magic fixes it...
+
+# convert to characters
+num_clients_char <- as.character(immigration$Number.of.Clients)
+# split up by the silly comma
+num_clients_split <- strsplit(num_clients_char, ",")
+# now combine back
+combine_thousands <- function(x)
+{
+  x <- as.numeric(x)
+  thousand_power <- ((length(x):1)-1)*3
[…]

### 2014-07-02 timesseries_measles.R

@@ -1,0 +1,176 @@
+rm(list=ls())
+
+library(maps)       # Provides functions that let us plot the maps
+library(mapdata)    # Contains the hi-resolution points that mark out the countries.
+library(rworldmap)
+library(adehabitat)
+library(rgdal)
+library(raster)
+library(maps)
+library(maptools)
+library(sp)
+#library(SDMTools)
+library(ggplot2)
+require(rgeos)
+library(ggmap)
+library(mapdata)
+gpclibPermit()
+library(mapproj)
+library(plyr)
+library(grid)
+library(gridExtra)
+
+# Data from http://thematicmapping.org/downloads/world_borders.php.
+# Direct link: http://thematicmapping.org/downloads/TM_WORLD_BORDERS_SIMPL-0.3.zip
+# Unpack and put the files in a dir 'data'
+
+gpclibPermit()
+library(utils)
+library(rgdal)
+getwd()
+setwd("~/data")
+#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
+world.map <- readOGR(dsn="C:/Users/David Hayman/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
+
+setwd("C:/Users/David Hayman/Dropbox/measles/data")
+
+incidence<-read.csv("incidence_series.csv",header=T)
+head(incidence)
+
+population<-read.csv("population.csv", header=T)
+head(population)
+
+#######
+
+world.ggmap <- fortify(world.map, region = "NAME")
+save(world.ggmap,file="world.ggmap.Rda")
+load("world.ggmap.Rda")
+n <- length(unique(world.ggmap$id))
+
+id = unique(world.ggmap$id)
+id<-as.data.frame(id)
+
+colnames(incidence)[3]<-"id"
+levels(incidence$id)[match(c("Bahamas (the)",
+                             "Bolivia (Plurinational State of)",
+                             "Central African Republic (the)",
+                             "Comoros (the)",
+                             "Congo (the)",
+                             "Côte d'Ivoire",
+                             "Czech Republic (the)",
+                             "Democratic People's Republic of Korea (the)",
+                             "Democratic Republic of the Congo (the)",
+                             "Dominican Republic (the)",
+                             "Gambia (the)",
+                             "Lao People's Democratic Republic (the)",
+                             "Libya",
[…]

### 2014-07-03 interimreport1.Rnw
Cost analyses from Tim included
Edited interim report with additional tables and figures.

@@
 
-\title{Measles risk assessment and cost interim report\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
+\title{Interim report\\ Measles risk assessment, modeling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
 \author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
@@
 The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011/2012 (mainly in Auckland region) and another significant outbreak is currently going on in the Auckland and Waikato regions.
-The outbreak in 2011-2012 lasted for more than 12 months and the current 2013/2014 outbreak started at the end of December 2013 (as on \date{\today}).
-In 2013, prior to the 2013/2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) it can request verification of non-endemic status three years after the last case of the 2011-12 outbreak in June 2012.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011-2012 (mainly in Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
+The outbreak in 2011-2012 lasted for more than 12 months and the current 2013-2014 outbreak started at the end of December 2013 (as on \date{\today}).
+In 2013, prior to the 2013-2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) it can request verification of non-endemic status three years after the last case of the 2011-12 outbreak in June 2012.
 
-Previous analyses, including two of measles in New Zealand by Prof.
+Previous analyses of measles, including two in New Zealand by Prof.
 Roberts (INSERT REFS), have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, as described above, outbreaks still occur.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
 Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are perhaps likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
@@
 Overall, we thought the review was very thorough.
-The report included substantial background information on measles immunisation in New Zealand (\emph{section 1.3}), the epidemiology of measles in New Zealand (\emph{section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{section 3}), and the levels of population immunity against the virus (\emph{section 4}).
-Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{section 4.2}) and the sustainability of the national immunisation programme (\emph{section 5}).
+The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
+Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).
 
-Within the report there are many tables and figures which given substantial detail on the measles situation in New Zealand.
+Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand.
 Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
 
-Specific epidemiological details were provided for the 2011/2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month, 2011 and 2012, with additional breakdown by age group, 2011 and 2012 (\emph{Figure 5}) and per 100,000 population (e.g.
-\emph{Figures 6-8}).
-Similar presentation of the case data are provided for ethnicity (e.g.
-\emph{Figures 9-10}) and New Zealand Index of Deprivation (NZDep) (e.g.
-\emph{Figures 11-13}).
-Three figures,\emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
+Specific epidemiological details were provided for the 2011-2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6-8}).
+Similar presentation of the case data are provided for ethnicity (\emph{Figures 9-10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11-13}).
+Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
 
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles is not endemic and provide some preliminary analyses on the outbreaks since endemic measles 'elimination' (see section X) that gives information regarding the likelihood of this occuring.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section X) that gives information regarding the likelihood of measles persisting within the population and becoming endemic.
 
-We agree with the report's conclusions that testing for measles is performed appropriately within the required timeframe, but clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions.
+We agree with the report's conclusions that testing for measles is performed appropriately within the required timeframe, but clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
 
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approching or over 95\% levels for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modeling results (section X).
+This is particularly in light of our preliminary modeling and risk analyses results (sections X and Y).
 
-The analyses we thought could further inform the understanding of risk from measles infection were:
+The analyses we believe could further inform the understanding of risk from measles infection are:
 \begin{itemize}
@@
 
-We received the raw case data from The Institute of Environmental Science and Research Ltd (ESR) EpiSurv on 27 June 2014.
-Initial analyses of those data suggest that we require denominator data to perform multivariate analyses.
-Specifically for the multivariate analyses we wish to perform we require \emph{Age * Prioritized Ethnicity * NZDep} data for New Zealand to see if the interactions among case covariates give us any additional information on risk from the univariate analyses already performed in the report reviewed above.
+We received the raw EpiSurv case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
+Initial analyses of those data suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of indenpendence among risk factors.
+Specifically for the multivariate analyses we wish to perform that detect interactions we require \emph{Age * Prioritized Ethnicity * NZDep} data for New Zealand to see if the interactions among case covariates give us any additional information on risk from the univariate analyses already performed in the report reviewed above.
 
 These data are currently unavailable in the form we require and after contacting ESR and NZ Statistics we and NZ statistics have contacted the University of Otago to see if we can obtain these denominator data.
-Additional data we believe would enable us and the Ministry to understand measles risk better is fine scale (lower than DHB) immunization coverage data.
+Additional data we believe would enable us and the Ministry to understand measles risk better is fine scale (lower than District Health Board (DHB)) immunization coverage data.
 
-We have reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help better understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers, we sought to understand better the risk of measles importation.
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
+     \caption{Measles incidence from 1997 to 2014}
+     %\label{fig:test image}
+\end{figure}
+
+In light of the apparent increasing trend in measles incidence in the last few years [Fig], we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers, we sought to use semi-quantitative methods to evaluate the risk of measles importation.
 For our initial analyses, we used the arrivals data from \href{http://www.immigration.govt.nz/}{www.immigration.govt.nz} as an estimate of human movement to and from New Zealand.
@@
 However, vaccination coverage is lowest and measles incidence highest in less developed nations (Tables 2 and 3).
-Though the interactions between these different risk factors are unknown, simple products of measles incidence in 2012 and immigration numbers in 2012 suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some countries also poses a risk.
+Though the precise interactions between these different risk factors are unknown, the simple product of measles incidence in 2012 and immigration numbers in 2012 suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a risk.
 These data are shown in Table 4.
@@
 
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE>>=
+change<-read.csv("change.csv",header=T)
+plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
+     bty="n")
+ylim <- range(c(0,max(change[,4])))
+xlim <- range(change[,1])
+par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
+plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
+     ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
+@
+\end{center}
+\caption{Trend in global per capita measles incidence}
+\end{figure}
+
 Though [SOMETHING ON MEASLES], immigration rates to New Zealand are increasing. [INSERT FIGURE]
 
+\begin{figure}[h!]
+\begin{center}
 <<fig=TRUE,echo=FALSE>>=
@@
 @
+\end{center}
+\caption{Trend in total immigrations into New Zealand}
+\end{figure}
+
+\section{Section summary}
+\begin{itemize}
+\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
+\end{itemize}
 
@@
 \item Multivariate regression analyses to look for interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using more than the 2012 data.
+\item Update of the importation risk analyses using more than the 2012 data, including modeling the trend in importations.
 \end{itemize}
@@
 \item A lack of meshblock level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level. 
+\end{itemize}
+
 We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
 \emph{meshblock}) level would allow better understanding of risk and resource allocation.
-\item A lack of data at any level that combines NZDep with age, prioritized ethnicity, and other variables, that allow better use of data collected by the Ministry, ESR, and NZ statistics for multivariate analyses.
-\end{itemize}
 
@@
 
-... [TEXT HERE] ...
+Approximately 50 years ago, approximately 135 million cases and 7-8 million deaths were believed to occur in the world (Clements and Hussey, 2004).
+Thirty years later, it was estimated there were still approximately 45 million cases of measles offering annual, including 6 million measles-related fatalities.
+Globally, measles remains a problem.
+Wolfson et al.
+(2007) estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
+Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
+They estimated approximately 7.5 million deaths from measles were avoided from 2000-05 due vaccination.
+The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 (WHO, 2013).
+In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country  of origin or in the previously measles-free country.
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million (Carabin et al., 2003).
+The estimated cost for a case ranged from US\$ 189-344 (Carabin et al., 2003); however, the average estimated cost of a typical hospital case ranges from US\$ 967-1,755 (Carabin et al., 2002).
+Stack et al (2011) estimated the economic benefits from cases averted due to measles vaccination.
+They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$ 10 billion of costs averted between 2011 and 2020.
+Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
+Italy has the highest reported annual cost of measles among industrialised countries (Carabin et al., 2003).
+In 2001, it reported losses related to measles of approximately US\$ 50 million.
+The economic impact of a large measles outbreak in Italy, 2002-03 examined the costs associated with 5,154 hospitalizations where measles was the main discharge diagnosis.
+The mean length of hospital stay was 5.2 days (median = 4 days and range = 1-303 days).
+The total cost of these hospitalizations amounted to ???
+8.83 million (???
+1 ~ NZ\$ 2.0 in 2002-03), or approximately ???
+1,700 per case.
+The average cost per non-complicated measles was ???
+1,429, while the mean cost of a case with complicated measles was ???
+2,721.
+The average daily cost of a hospital stay was ???
+327.
+An outbreak of measles occurred in Sydney, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases (Flego et al., 2013).
+Seven (27\%) of the cases required hospitalization for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
+During this outbreak, a total of 1,395 contacts were identified and managed by a public health unit in western Sydney.
+The mean number of contacts per case was 54 (median = 28, maximum = 206).
+The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$ 48,000, with 90\% of this being associated with staff time.
+Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in fecent years.
+In 2001 more than 6000 cases were reported in Germany but by 2004 this number fell to 122 (Wichmann et al., 2009).
+However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children (Siedler et al, 2006).
+An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
+In that study, they estimated the health-care provider costs to be approximately ???
+229,000, or ???
+373/case.
+Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalized.
+The mean costs of the hospitalized patients was ???
+1,877, including one patient with encephalitis at a cost of ???
+35,623.
+In addition to the health-care provider costs, additional costs of ???
+89,400 were incurred by the district public health office, the majority (???
+85,000, 95.1\%) for personnel, ???
+2,300 (2.6\%) for vaccination, and ???
+2,100 (2.3\%) for serologic testing.
+Therefore the combined direct costs of these 612 cases amounted to ???
+318,400, or ???
+520 per case.
+In addition, to determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
+Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world (Parker et al., 2006).
+Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.
+Ortega-Sanchez et al.
+(2014) estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
+The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$ 2.7 and 5.3 million.
+Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$ 298 per contact.
+Dayan et al.
+(2005) calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
+They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination.
+In total, it was estimated the direct costs associated with three cases of measles was US\$ 142,452, or nearly US\$ 50,000 per case.
+
+Parker et al.
+(2006) reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
+A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak.
+Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak.
+Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overhead.
+This amounted to a direct cost of \$4,932 per measles case.
+These costs did not include either patient care or indirect costs, which would have made the total and per case cost higher.
+
+Coleman et al.
+(2012) estimated the direct medical and public health costs in response to a single case of refugee-imported case measles.
+Costs included labor, translation, benefits for public health workers.
+In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalization, transportation and diagnosis.
+In total, 387 hours were associated with this single case, resulting in a cost of US\$ 11,881.
+In addition, per-contact costs amounted to US\$ 264.
+The cost of hospitalization for the 3-day stay by the index case was US \$931.
+Additional costs were associated with physician visits ((US\$ 294), vaccine and immunoglobulin (US\$ 1,765), mileage (US\$ 205) and immunologic screening tests for the parents' exposed to measles (US\$ 240) for a total of US\$ 23,816.
+Economic analyses of measles control programs have shown them to be financially effective.
+In the Republic of Korea, the economics of alternative measles vaccination programs were compared (.
+All of the alternatives were found to be economically efficient (B/C > 1.0), with the alternative using two doses of the measles, mumps and rubella program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27).
+The purpose of the current study is to estimate the cost of the current measles outbreak in New Zealand.
+Using this information, we will then evaluate the economics of alternative measles control strategies in order to provide additional information to public health officials and decision makers.
+
+\section{Materials and methods}
+Costs were evaluated as either direct or indirect.
+Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
+Direct costs can be divided into medical and non-medical (Saha and Gerdtham, 2013).
+Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
+Personnel time (investigation and emergency response), materials (phone calls, vaccine, personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
+Direct non-medical costs include transportation to and from health care providers.
+Indirect costs are productivity losses for the case and/or health care provider, e.g.
+parent of a school child.
+Indirect costs included work loss for cases and caregivers.
+This could also include the economic value of premature life lost, costs associated with permanent disability, e.g.
+deafness and mental retardation.
+Commonly the human value approach (HVA) has been used to estimate economic impact of life.
+The HVA measures the potential future earnings of an individual and discounts it into a present value.
+Typically this is 3\% but 5\% has also been used in a sensitivity analysis, which is more compared to non-human life calculations and will tend to reduce the present value of the future earnings (saved by avoiding a case).
+Data for the current measles outbreak were obtained from the New Zealand Ministry of Health, from 2008 through June 2014.
+Data included information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, case weight and associated cost.
+Cost of the Auckland Regional Public Health Service (ARPHS) for measles response were obtained from the Ministry of Health.
+Data, for the period January 1 - March 9, 2014, reported salaries for people involved with the measles outbreak management medical team.
+The costs were reported as direct, additional (above normal budgeting) costs required to enable the management of measles.
+It includes a breakdown by individual performing the work and whether it was during the normal work schedule (M-F) or weekends.
+Normal work was calculated as 1.2*FTE* number of days worked.
+Overtime was calculated as 1.6*FTE (M-F) and 2.0*FTE (weekend).
+A full day was considered as 8 hours worked.
+Salary (hourly) rates were calculated for the following: public health nurse (\$36), public health assistant (\$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), IMT , which had the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and SSO (\$26).
+In addition, measles operations personnel were calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
+Mean wages for New Zealand workers, by age and gender were obtained for the period, 2008-2013 from the New Zealand Income Survey (Statistics New Zealand, 2013).
+Measles cases were assumed to not work for a period of 5 days.
+Similarly, a care taker was assumed to not work for 5 days if the case were less than 20 years of age.
+In order to calculate the wage loss associated with the care taker, it was assumed that the person was a female between the ages of 35-39.
+Age and gender information for the 192 publicly funded hospital discharges with a measles primary diagnosis from 2008-2013 were matched to the New Zealand wage file to calculate lost wages due to measles.
+A regression analysis was performed to test for significant associations between hospital cost and the following explanatory variables: case age at discharge, gender, length of stay (days) and year of case.
+
+\section{Results}
+Direct costs for measles management in New Zealand for the 10-week period, January 1 - March 9, 2014 are shown in table 1.
+The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in table 2.
+
+<<echo=FALSE>>=
+costtable1<-read.csv("costtable1.csv",header=T)
+@
+
+\begin{table}
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 - March 9, 2014}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable1, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was $289,694 (Table 2).
+The mean cost per case was $1,509.
+The mean cost per day of stay in the hospital was $1,154.
+
+<<echo=FALSE>>=
+costtable2<-read.csv("costtable2.csv",header=T)
+@
+
+\begin{table}
+\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008-2013}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable2, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 - 19 June 2014).
+The number of cases by age group is shown in Table 3.
+Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 0.47 and 0.33, respectively.
+
+<<echo=FALSE>>=
+costtable3<-read.csv("costtable3.csv",header=T)
+@
+
+\begin{table}
+\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 - 19 June, 2014}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable3, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+The majority (191/192) of the measles cases stayed in the hospital from 0 to 7 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
+Eight-three of the cases stayed for 0 days and the remaining cases stayed a mean of 2.30 (median 2) days.
+The number of cases, length of hospital stay, cost, cost/case and cost/day for patients with measles as the primary diagnosis, by year and gender for 2008-13 appear in Table 4.
+The cost for these 192 hospitalizations was \$289,964.
+Of the 292 cases, 87 were female and 105 were male.
+The number of days spent in the hospital was 251, making the mean daily cost of \$1,509.
+Of the 251 days, 114 were spent by females and 137 by males.
+The mean hospital stay per case was 1.31 and 1.30 days for females and males, respectively.
+The mean cost per case was \$1,573 and \$1,456 for females and males, respectively.
+
+<<echo=FALSE>>=
+costtable4<-read.csv("costtable4.csv",header=T)
+@
+
+\begin{table}
+\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008-2013}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable4, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+The cost per measles case, by age and gender appear in Figure 1.
+
+[No figure]
+
+Regression analyses showed statistically significant (P < 0.05) associations between cost of hospitalization and three variables, length of hospitalization, case age and year of case, and marginally significant (P = 0.08) association with case gender.
+Results showed an increase of approximately \$358 per day of hospitalization, \$139 for a female case, and \$1 per year of case diagnosis.
+The cost was found to decrease by approximately \$5 per year of case age.
+
+<<echo=FALSE>>=
+costtable5<-read.csv("costtable5.csv",header=T)
+@
+
+\begin{table}
+\caption{Regression results (R$^{2}$ adj = 0.8969, P<0.001) for measles hospitalization cost based on length of stay (days), gender, case age and year of case (n=192) in New Zealand, 2008 - 2013}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable5, file="", table.env=FALSE,rowname=NULL)
+@
+\end{table}
+
+Wages lost due to measles were calculated based on the assumptions that 5 days of work were lost for each case; however, individuals between 0-15 years of age were not assumed to be employed and therefore did not suffer an income loss.
+If the case were < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
+Total wage lost for the 192 cases and care givers was estimated to be \$166,638.
+This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases <15 years of age.
+
+\section{Discussion}
+The results presented here are based on available data.
+While some of the data are complete and detailed, this is not true of all the data.
+In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
+For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
+In addition to this information, similar data would be needed for cases occurring outside the period 2011-2013 at publicly funded hospitals.
+Similar data would be needed for non-publicly funded hospitals, e.g.
+private clinics.
+Detailed measles outbreak management costs were provided for the period of January 1 - March 9, 2014.
+Similar data are needed for the period preceding 2014.
+If detailed data, such as that provided for early 2014, aggregated data would be acceptable.
+However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
+As other studies have demonstrated, direct costs required to manage measles are not linear.
+
+Additional clarification of data is needed regarding what hospital costs refer to.
+In addition, if a hospital day were 0 does that mean the case stayed in the hospital but not overnight?
+Or, does it mean the case stayed for less than 24 hours?
+Once more complete data are available, comparisons of these results will be made to other published studies. Examples of findings from some of these studies appear below.
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
+In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
+They estimated direct costs per case to be less than US\$500.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at \$31,059 and \$78,448, respectively (Prouty and Lakin, 2000)
+In 2000 expenditures for care in large state MR/DD facilities continued to increase and reached a national annual average of \$113,864 per person.
+In 2000 the average annual expenditures for care in large state (mental retardation/developmental disabilities) MR/DD facilities were \$113,864.
+The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalization) to \$29,556 (encephalitis and hospitalization for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) (Zhou et al., 2004).
 
@@
 
-The original mathematical model for the dynamics of measles in New Zealand
-was prepared in 1996 (Tobias & Roberts 1998). It successfully predicted the
-1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
-et al 1998; Roberts & Tobias 2000). A subsequent extension of this work in
-1998 showed that the then current schedule of MMR1 at 15 months and MMR2
-at 11 years was insufficient to prevent further epidemics. The schedule was
-changed in 2000 with MMR2 now being administered before 5 years (Anon.
-2002a).
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 (Tobias & Roberts 1998) successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
+et al 1998; Roberts & Tobias 2000).
+Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
+The model developed by Roberts & Tobias (2000) supported the change in the
+immunisation schedule that took effect in January 2001, at which time MMR2
+was changed from delivery at 11 years to delivery before the age of five. The schedule was changed in 2000 with MMR2 now being administered before 5 years (Anon.
+2002a) and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasized that it is necessary to maintaining high coverage rates in order to prevent future
+epidemics.
 
@@
 
-The model developed by Roberts & Tobias (2000) supported the change in the
-immunisation schedule that took effect in January 2001, at which time MMR2
-was changed from delivery at 11 years to delivery before the age of five. These
-results were in line with those obtained by other authors, for example: Babad
+These results were in line with those obtained by other authors, for example: Babad
 et al (1995) advocated a two-dose schedule for England and Wales, with the
@@
 age at delivery, but no strategy succeeded if coverage rates were below approximately
-87\%. Our results presented in Table 1 are similar where coverage is
-assumed to be the same at MMR1 and MMR2. The results also show the absolute
-necessity of maintaining high coverage rates in order to prevent future
-epidemics. It is difficult to estimate the proportion of the school-age population
+87\%.
+
+It is difficult to estimate the proportion of the school-age population
 that have been effectively immunised, but this is continuously being
-diluted by children who are not immunised. Wallinga et al (2000) noted that
-in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
-and that under most plausible assumptions for contact rates even
-100\% coverage would be insufficient to prevent epidemics.
-Our final conclusion must depend on how much faith is placed in the model.
-The results are consistent with those of other authors, and from Table 1 it
-could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
-to prevent future measles epidemics. However, a study by Glass et al (2004)
+diluted by children who are not immunised. While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, Glass et al (2004)
 in the Netherlands showed that high overall levels of measles vaccination can
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
-     \caption{Measles incidence from 1997 to 2014}
-     %\label{fig:test image}
-\end{figure}
+To estimate Rv or R0 we were required to compute the generation time for measles. We used a lognormal distribution with mean 12.0 and sd 3.5 from
+%# http://www.sciencedirect.com/science/article/pii/S0022519311003146
+We then estimated R0 using the methods adapted from REF from the incidence data for each outbreak, which we defined from the dataset [FIG] given their temporal and geographic correlations.
+The outbreaks are shown in Fig [x].
 
@@
 
-...The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases.
-[COMMENT?
-Unsure, though don't want to be dismissive].
-....
+The estimated R0 for each outbreak is shown in figure [y].
+The probability density of the R0 estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an R0 above one and thus we may expect this outbreak to persist if conditions remain the same.
+
+These analyses also impy that the regule (~yearly)
+
+\section{Summary of modeling}
+
+\begin{itemize}
+\item blah.
+\end{itemize}
 
@@
 
+\section {Summary of key findings}
+\begin{itemize}
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles eslewhere in the globe.
+\item Cost
+\item Analyses of outbreak data suggest that measles Rv values often include 1 and in this year, 2014, well above one.
+This analyses suggest improved vaccination is a requesit to prevent measles becoming endemic again.
+\end{itemize}
+
 \section{Acknowledgments}
-The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR for help in obtaining the appropriate materials for analyses.
+The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR, and June Atkinson from University of Otago for help in obtaining the appropriate materials for analyses.
+
+The R0 package: a toolbox to estimate reproduction numbers for epidemic outbreaks
+Thomas Obadia, Romana Haneef and Pierre-Yves Boëlle, BMC Medical Informatics and Decision Making 2012, 12:147  doi:10.1186/1472-6947-12-147
+
+Bae, G-R, Choe, YJ, Go, UY, Kim, Y-I and Lee, J-K, 2013. Economic analysis of measles elimination program in the Republic of Korea, 2001: A cost benefit analysis study, Vaccine, 31:2661-2666.
+
+Carabin, H, Edmunds, WJ, Kou, U, van den Hof, S and Nguyen, VH, 2002. Measles in industrialized countries: a review of the average costs of adverse events and measles cases, BMC Public Health, 2:22.
+
+Carabin, H, Edmunds, WJ, Gyldmark, M, Beutels, P, Lévy-Bruhl, D, Salo, H and Griffiths, UK, 2003. The cost of measles in industrialised countries, Vaccine, 21:4167-4177.
+
+Clements, CJ and Hussey, GD, 2004. Chapter 4: Measles, IN: The Global Epidemiology of Infectious Diseases, Murray, CJL, Lopez, AD and Mathers, CD (eds.), World Health Organization, Geneva, 391 p.
+
+Coleman, MS, Garbat-Welch, L, Burke, H, Weinberg, M, Humbaugh, K, Tindall, A and Cambron, J, 2012. Direct costs of a single case of refugee-imported measles in Kentucky, Vaccine, 30:317-321.
+
+Dayan, GH, Ortega-Sánchez, IR, LeBaron, CW, Quinlisk, MP and the Iowa Measles Response Team, 2005.
+The cost of containing one case of measles: the economic impact on the public health infrastructure - Iowa, 2004, Pediatrics, 116:e1; DOI:10/1542/peds.2004-2512.
+
+Filia, A, Brenna, A, Paná, A, Cavallaro, GM, Massari, M and Atti, MLCd, 2007. Health burden and economic impact of measles-related hospitalization in Italy, 2002-2003.
+
+Flego, KL, Belshaw, DA, Sheppeard, V and Weston, KM, 2013. Impacts of a measles outbreak in western Sydney on public health resources, Communicable Diseases Intelligence Quarterly Report, 37:E240-245.
+
+Koopmanschap, MA, 1998. Cost-of-illness studies: useful for health policy?, Pharmacoeonomics, 14:143-148.
+Larg, A and Moss, JR, 2011. Cost-of-illness studies: a guide to critical evaluation, Pharmacoeconomics, 29:653-671.
+
+Ortega-Sanchez, IR, Vijayaraghavan, M, Barskey, AE and Wallace, GS, 2014. The economic burden of sixteen measles outbreaks on United States public health departments in 2011, Vaccine, 32:1311-1317.
+
+Parker, AA, Staggs, W, Dayan, GH, Ortega-Sánchez, IR, Rota, PA, Lowe, L, Boardman, P, Teclaw, R.
+Graves, C and LeBaron, CW.
+2006.
+Implications of a 2005 measles outbreak in Indiana for sustained elimination of measles in the United States, The New England Journal of Medicine, 355:447-455.
+
+Prouty, RW, Smith, G and Lakin, KC, 2001.
+Residential services for persons with developmental disabilities: status and trends through 2000.
+Minneapolis: Institute on Community Integration, University of Minnesota, 179 pages, rtc.umn.edu/risp00.
+
+Saha, S and Gerdtham, UG, 2013. Cost of illness studies on reproductive, maternal, newborn, and child health: a systematic literature review, Health Economics Review, doi:10.1186/2191-1991-3-24.
+
+Siedler, A.
+Tischer, A, Mankertz, A and Santibanez, S, 2006.
+Two outbreaks of measles in Germany 2005, Eurosurveillance 2006:11(4) article 5, \href{http://www.eurosurveillance.org/ViewArticle.aspx?ArticleId=615}, accessed 14 June 2014.
+
+Stack, ML, Ozawa, S, Bishai, DM, Mirelman, A, Tam, Y, Niessen, L, Walker, DG and Levine, OS, 2011.
+Estimated economic benefits during the 'decade of vaccine' include treatment savings, gains in labor productivity, Health Affairs, 30:1021-1028.
+
+Statistics New Zealand, 2014. http://nzdotstat.stats.govt.nz/, accessed 17 June 2014,
+
+Wichmann, O, Siedler, A, Sagebiel, D, Hellenbrand, W, Santibanez, S, Mankertz, A, Vogt, G, van Treeck, U and Krause, G, 2009.
+Further efforts needed to achieve measles elimination in Germany: results of an outbreak investigation, Bulletin of the World Health Organization, 87:108-115.
+
+Wolfson, LJ, Strebel, PM, Gacic-Dobo, M, Hoekstra, EJ, McFarland, JW and Hersh, BS, 2007.
+Has the 2005 measles mortality reduction goal been achieved?
+A natural history modelling study, Lancet, 369:191-200.
+
+WHO measles media centre. Geneva: World Health Organization, January 2013. \href{\http://www.who.int/mediacentre/news/notes/2013/measles_20130117/en/}, accessed July 1, 2014.
+
+Zhou, F, Reef, S, Massoudi, M, Papania, MJ, Yusuf, HR, Bardenheier, B, Zimmerman, L and McCauley, MM, 2004.
+An economic analysis of the current universal 2-dose measles-mumps-rubella vaccination program in the United States, Journal of Infectious Diseases, 189:S131-45.
 

### 2014-07-03 interimreport1.tex
Draft Interim Report
Draft with no references in it. I am working on them in a text file and
will paste in later.

@@
 
-Summary of findings
+New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
+In this interim report we reivewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
+We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013-2014 outbreak and began modeling measles outbreaks to understand how they may be controled and the possibility that measles may become endemic.
+Our key findings were:
+\begin{itemize}
+\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles eslewhere in the globe.
+\item The cost of the current 2013-2014 measles outbreak is estimated to be at least \$$\frac{3}{4}$ million.
+\item Analyses of outbreak data suggest that measles basic reproductive number (Rv, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
+This analyses suggest improved vaccination is a requesite to prevent measles becoming endemic again.
+\end{itemize}
 
@@
 As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committted to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
-The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved this.
+The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.
 
 The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011/2012 (mainly in Auckland region) and another significant outbreak is currently going on in the Auckland and Waikato regions.
-The outbreak in 2011-2012 lasted for more than 12 months and the current 2013/2014 outbreak started at the end of December 2013 (as on \date{\today}).
-In 2013, prior to the 2013/2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) it can request verification of non-endemic status three years after the last case of the 2011-12 outbreak in June 2012.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011-2012 (mainly in Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
+The outbreak in 2011-2012 lasted for more than 12 months and the current 2013-2014 outbreak started at the end of December 2013 (as on \date{\today}).
+In 2013, prior to the 2013-2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) it can request verification of non-endemic status three years after the last case of the 2011-12 outbreak in June 2012.
 
-Previous analyses, including two of measles in New Zealand by Prof.
+Previous analyses of measles, including two in New Zealand by Prof.
 Roberts (INSERT REFS), have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, as described above, outbreaks still occur.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
 Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are perhaps likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
@@
 Overall, we thought the review was very thorough.
-The report included substantial background information on measles immunisation in New Zealand (\emph{section 1.3}), the epidemiology of measles in New Zealand (\emph{section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{section 3}), and the levels of population immunity against the virus (\emph{section 4}).
-Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{section 4.2}) and the sustainability of the national immunisation programme (\emph{section 5}).
+The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
+Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).
 
-Within the report there are many tables and figures which given substantial detail on the measles situation in New Zealand.
+Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand.
 Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
 
-Specific epidemiological details were provided for the 2011/2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month, 2011 and 2012, with additional breakdown by age group, 2011 and 2012 (\emph{Figure 5}) and per 100,000 population (e.g.
-\emph{Figures 6-8}).
-Similar presentation of the case data are provided for ethnicity (e.g.
-\emph{Figures 9-10}) and New Zealand Index of Deprivation (NZDep) (e.g.
-\emph{Figures 11-13}).
-Three figures,\emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
+Specific epidemiological details were provided for the 2011-2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6-8}).
+Similar presentation of the case data are provided for ethnicity (\emph{Figures 9-10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11-13}).
+Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
 
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles is not endemic and provide some preliminary analyses on the outbreaks since endemic measles 'elimination' (see section X) that gives information regarding the likelihood of this occuring.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section X) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
 
-We agree with the report's conclusions that testing for measles is performed appropriately within the required timeframe, but clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions.
-
+We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe.
+Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
 Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approching or over 95\% levels for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modeling results (section X).
+This is particularly in light of our preliminary modeling and risk analyses results (sections X and Y).
 
-The analyses we thought could further inform the understanding of risk from measles infection were:
+The analyses we believe could further inform the understanding of risk from measles infection are:
 \begin{itemize}
 \item Multivariate modelling to account for confounding within the univariate analyses.
-\item Inclusion of multivariate model results with local level immunity data to develop a risk map for NZ. 
+\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
 \item Analyses of changing risk factors through time during outbreaks.
@@
 
-\section{Additional risk analyses}
+\subsection{Additional risk analyses}
 
 In this section we provide a brief description of the work we are performing that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
-These analyses are intended to build on the excellent analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
+These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
 
-We received the raw case data from The Institute of Environmental Science and Research Ltd (ESR) EpiSurv on 27 June 2014.
-Initial analyses of those data suggest that we require denominator data to perform multivariate analyses.
-Specifically for the multivariate analyses we wish to perform we require \emph{Age * Prioritized Ethnicity * NZDep} data for New Zealand to see if the interactions among case covariates give us any additional information on risk from the univariate analyses already performed in the report reviewed above.
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
+     \caption{Measles incidence from 1997 to 2014}
+     \label{fig:incidence1997}
+\end{figure}
 
-These data are currently unavailable in the form we require and after contacting ESR and NZ Statistics we and NZ statistics have contacted the University of Otago to see if we can obtain these denominator data.
-Additional data we believe would enable us and the Ministry to understand measles risk better is fine scale (lower than DHB) immunization coverage data.
+In light of the apparent increasing trend in measles incidence in the last few years Figure~\ref{fig:incidence1997}, we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
 
-We have reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help better understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers, we sought to understand better the risk of measles importation.
-For our initial analyses, we used the arrivals data from \href{http://www.immigration.govt.nz/}{www.immigration.govt.nz} as an estimate of human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO \href{http://www.who.int/research/en/}{www.who.int/research/en/}.
+For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}.
 
 \begin{table}
-\caption{Immigration to New Zealand (2012)}
+\caption{Immigration numbers to New Zealand (2012)}
 %latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\end{table}
+\end{tabular}\end{center}\label{table:immigration12}
+\end{table}
 
@@
 \hline
-\end{tabular}\end{center}\end{table}
+\end{tabular}\end{center}\label{table:cover12}
+\end{table}
 
 \begin{table}
-\caption{Measles Incidence per million (2012)}
+\caption{Measles incidence per million (2012)}
 %latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\end{table}
+\end{tabular}\end{center}\label{table:incidence12}
+\end{table}
 
@@
 \hline
-\end{tabular}\end{center}\end{table}
+\end{tabular}\end{center}\label{table:risk12}
+\end{table}
 
-The preliminary analyses of 2012 data suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table 1.
-However, vaccination coverage is lowest and measles incidence highest in less developed nations (Tables 2 and 3).
-Though the interactions between these different risk factors are unknown, simple products of measles incidence in 2012 and immigration numbers in 2012 suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some countries also poses a risk.
-These data are shown in Table 4.
-The data for all the variables for each nation state and territories for 2012 are plotted in Figure 1 and the risk map for measles incidence and immigration in Figure 2.
+The preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
+However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:incidence12} and Table \ref{table:cover12}).
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importantion to New Zealand.
+These data are shown in Table \ref{table:risk12}.
+The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
 
@@
 \caption{Immigration to New Zealand 2012}
+\label{fig:immigration12}
 \end{figure}
@@
 \caption{Measles incidence per million 2012}
+\label{fig:incidence12}
 \end{figure}
@@
 \caption{Measles vaccination cover (\%) 2012 }
+\label{fig:cover12}
 \end{figure}
@@
 \caption{Risk map from measles incidence and immigration rates}
+\label{fig:risk12}
 \end{figure}
 
-\section{Future risk analyses}
-We received ESR EpiSurv measles case data on the 27 June 2014. The following are analyses are still in progress and aim to include them in later reports:
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport1-011}
+\end{center}
+\caption{Trend in global per capita measles incidence}
+\label{fig:trendincidence}
+\end{figure}
+
+Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
+This suggests that the risk of measles important could increase, though further analyses are require to understand these interatction between these.
+Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).
+This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin or travel to will be an important factor in determining the risk of measles importation.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport1-012}
+\end{center}
+\caption{Trend in total immigrations into New Zealand}
+\label{fig:trendimmigration}
+\end{figure}
+
+\subsection{Risk analysis summary}
+\begin{itemize}
+\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
+\item There may be seasonal changes n risk of measles importation, though further analyses are needed.
+\end{itemize}
+
+\subsection{Future risk analyses}
+We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
+Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of indenpendence among risk factors.
+Specifically for the multivariate analyses we wish to perform that detect interactions we require \emph{Age * Prioritized Ethnicity * NZDep} data for New Zealand to see if the interactions among case covariates give us any additional information on risk from the univariate analyses already performed in the report reviewed above.
+These data only became available to us on 3 July 2014 after because ESR and NZ Statistics did not have those data and we had to obtain these from the University of Otago.
+
+The following are analyses are still in progress and aim to include them in later reports:
 \begin{itemize}
 \item Multivariate regression analyses to look for interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using more than the 2012 data.
+\item Update of the importation risk analyses using more than the 2012 data, including modeling the trend in importations.
 \end{itemize}
-The data gaps that we have that will hinder us in completing all the analyses are:
+Additional data we believe would enable us and the Ministry to understand measles risk better is fine scale (lower than District Health Board (DHB)) immunization coverage data.
+Thus the data gap that we have that will hinder us providing fine scale risk maps are:
 \begin{itemize}
 \item A lack of meshblock level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
+\item The number of New Zealanders arriving from abroad each year.
+\end{itemize}
+
 We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
-\emph{meshblock}) level would allow better understanding of risk and resource allocation.
-\item A lack of data at any level that combines NZDep with age, prioritized ethnicity, and other variables, that allow better use of data collected by the Ministry, ESR, and NZ statistics for multivariate analyses.
-\end{itemize}
+\emph{meshblock}) level would allow better understanding of risk of measles infection and resource allocation because they may allow targetted immunisation programmes.
 
@@
 
-In this section we provide an analysis of the costs involved with this current measles outbreak.
+In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with this current measles outbreak.
 
-... [TEXT HERE] ...
+Approximately 50 years ago, approximately 135 million cases and 7-8 million deaths were believed to occur in the world (Clements and Hussey, 2004).
+Thirty years later, it was estimated there were still approximately 45 million cases of measles offering annual, including 6 million measles-related fatalities.
+Wolfson et al.
+(2007) estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
+Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
+They estimated approximately 7.5 million deaths from measles were avoided from 2000-05 due vaccination.
+The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 (WHO, 2013).
+In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country  of origin or in the previously measles-free country.
 
-\section{Future cost-benefit analyses}
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million (Carabin et al., 2003).
+The estimated cost for a case ranged from US\$ 189-344 (Carabin et al., 2003); however, the average estimated cost of a typical hospital case ranges from US\$ 967-1,755 (Carabin et al., 2002).
+Stack et al (2011) estimated the economic benefits from cases averted due to measles vaccination.
+They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$ 10 billion of costs averted between 2011 and 2020.
+Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
+
+Italy has the highest reported annual cost of measles among industrialised countries (Carabin et al., 2003).
+In 2001, it reported losses related to measles of approximately US\$ 50 million.
+The economic impact of a large measles outbreak in Italy, 2002-03 examined the costs associated with 5,154 hospitalizations where measles was the main discharge diagnosis.
+The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
+The total cost of these hospitalizations amounted to \euro 8.83 million (\euro 1 ~ NZ\$ 2.0 in 2002-03), or approximately \euro 1,700 per case.
+The average cost per non-complicated measles was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
+The average daily cost of a hospital stay was  \euro 327.
+
+An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases (Flego et al., 2013).
+Seven (27\%) of the cases required hospitalization for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
+During this outbreak, a total of 1,395 contacts were identified and managed by a public health unit in western Sydney.
+The mean number of contacts per case was 54 (median = 28, maximum = 206).
+The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$ 48,000, with 90\% of this being associated with staff time.
+
+Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years.
+In 2001 more than 6000 cases were reported in Germany but by 2004 this number fell to 122 (Wichmann et al., 2009).
+However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children (Siedler et al, 2006).
+An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
+In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case.
+Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalized.
+The mean costs of the hospitalized patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623.
+In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing.
+Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case.
+In addition, to determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
+
+Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world (Parker et al., 2006).
+Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.
+Ortega-Sanchez et al.
+(2014) estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
+The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$ 2.7 and 5.3 million.
+Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$ 298 per contact.
+Dayan et al.
+(2005) calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
+They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination.
+In total, it was estimated the direct costs associated with three cases of measles was US\$ 142,452, or nearly US\$ 50,000 per case.
+
+Parker et al.
+(2006) reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
+A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak.
+Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak.
+Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overhead.
+This amounted to a direct cost of \$4,932 per measles case.
+These costs did not include either patient care or indirect costs, which would have made the total and per case cost higher.
+
+Coleman et al.
+(2012) estimated the direct medical and public health costs in response to a single case of refugee-imported case measles.
+Costs included labor, translation, benefits for public health workers.
+In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalization, transportation and diagnosis.
+In total, 387 hours were associated with this single case, resulting in a cost of US\$ 11,881.
+In addition, per-contact costs amounted to US\$ 264.
+The cost of hospitalization for the 3-day stay by the index case was US\$ 931.
+Additional costs were associated with physician visits (US\$ 294), vaccine and immunoglobulin (US\$ 1,765), mileage (US\$ 205) and immunologic screening tests for the parents' exposed to measles (US\$ 240) for a total of US\$ 23,816.
+Economic analyses of measles control programs have shown them to be financially effective.
+In the Republic of Korea, the economics of alternative measles vaccination programs were compared.
+All of the alternatives were found to be economically efficient (benefit/cost ration (B/C) > 1.0), with the alternative using two doses of the MMR program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27).
+The purpose of the current study is to estimate the cost of the current measles outbreak in New Zealand.
+Using this information, we will then evaluate the economics of alternative measles control strategies in order to provide additional information to public health officials and decision makers.
+
+\subsection{Methods}
+Costs were evaluated as either direct or indirect.
+Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
+Direct costs can be divided into medical and non-medical (Saha and Gerdtham, 2013).
+Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
+Personnel time (investigation and emergency response), materials (phone calls, vaccine, personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
+Direct non-medical costs include transportation to and from health care providers.
+Indirect costs are productivity losses for the case and/or health care provider, e.g.
+parent of a school child.
+Indirect costs included work loss for cases and caregivers.
+This could also include the economic value of premature life lost, costs associated with permanent disability, e.g.
+deafness and mental retardation.
+Commonly the human value approach (HVA) has been used to estimate economic impact of life.
+The HVA measures the potential future earnings of an individual and discounts it into a present value.
+Typically this is 3\% but 5\% has also been used in a sensitivity analysis, which is more compared to non-human life calculations and will tend to reduce the present value of the future earnings (saved by avoiding a case).
+Data for the current measles outbreak were obtained from the New Zealand Ministry of Health, from 2008 through June 2014.
+Data included information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, case weight and associated cost.
+
+Cost of the Auckland Regional Public Health Service (ARPHS) for measles response were obtained from the Ministry of Health.
+Data, for the period January 1 - March 9, 2014, reported salaries for people involved with the measles outbreak management medical team.
+The costs were reported as direct, additional (above normal budgeting) costs required to enable the management of measles.
+It includes a breakdown by individual performing the work and whether it was during the normal work schedule (Monday to Friday, M-F) or weekends.
+Normal work was calculated as 1.2*full time equivalent (FTE)*number of days worked.
+Overtime was calculated as 1.6*FTE (M-F) and 2.0*FTE (weekend).
+A full day was considered as 8 hours worked.
+Salary (hourly) rates were calculated for the following: public health nurse (PHN, \$36), public health assistant (PHA, \$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), incident management team (IMT), which had the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and safety/security offier (SSO) (\$26).
+In addition, measles operations personnel were calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
+
+Mean wages for New Zealand workers, by age and gender were obtained for the period, 2008-2013 from the New Zealand Income Survey (Statistics New Zealand, 2013).
+Measles cases were assumed to not work for a period of 5 days.
+Similarly, a care taker was assumed to not work for 5 days if the case were less than 20 years of age.
+In order to calculate the wage loss associated with the care taker, it was assumed that the person was a female between the ages of 35-39.
+Age and gender information for the 192 publicly funded hospital discharges with a measles primary diagnosis from 2008-2013 were matched to the New Zealand wage file to calculate lost wages due to measles.
+
+A regression analysis was performed to test for significant associations between hospital cost and the following explanatory variables: case age at discharge, gender, length of stay (days) and year of case.
+
+\subsection{Results}
+Direct costs for measles management in New Zealand for the 10-week period, January 1 - March 9, 2014 are shown in  Table \ref{table:direct}.
+The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table \ref{table:hosp}.
+
+\begin{table}
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 - March 9, 2014 (see text for abbreviations)}
+%latex.default(costtable1, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllll}
+\hline\hline
+\multicolumn{1}{c}{Category}&\multicolumn{1}{c}{January}&\multicolumn{1}{c}{February}&\multicolumn{1}{c}{March}&\multicolumn{1}{c}{Total}\tabularnewline
+\hline
+PHN&55,296&71,175&24,087&150,558\tabularnewline
+PHA&0&0&2,656&2,656\tabularnewline
+Data support&0&7,752&4,552&12,304\tabularnewline
+Supervisors&10,656&10,464&3,232&24,352\tabularnewline
+IMT&32,918&28,624&7,156&68,698\tabularnewline
+SSO&0&2,746&1,186&3,932\tabularnewline
+Measles operations&1,800&10,326&6,678&18,804\tabularnewline
+Operations partner&2,187&14,580&7,290&24,057\tabularnewline
+IMT controller partner&2,916&14,580&7,290&24,786\tabularnewline
+Total&105,773&160,247&64,127&330,147\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:direct}
+\end{table}
+
+The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was \$289,694 (Table \ref{table:hosp}).
+The mean cost per case was \$1,509.
+The mean cost per day of stay in the hospital was \$1,154.
+
+\begin{table}
+\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008-2013}
+%latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrlll}
+\hline\hline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Days}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Per.case}&\multicolumn{1}{c}{Per.day}\tabularnewline
+\hline
+2008&$  3$&$  1$&3,038&1,013&3,038\tabularnewline
+2009&$ 29$&$ 38$&40,782&1,406&1,073\tabularnewline
+2010&$  5$&$  5$&6,701&1,340&1,340\tabularnewline
+2011&$132$&$189$&205,303&1,555&1,086\tabularnewline
+2012&$ 19$&$ 12$&28,540&1,502&2,378\tabularnewline
+2013&$  4$&$  6$&5,330&1,333&888\tabularnewline
+Total&$192$&$251$&289,694&1,509&1,154\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:hosp}
+\end{table}
+
+From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 - 19 June 2014).
+The number of cases by age group is shown in Table \ref{table:freq}.
+Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 0.47 and 0.33, respectively.
+
+\begin{table}
+\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 - 19 June, 2014}
+%latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrr}
+\hline\hline
+\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Admitted}&\multicolumn{1}{c}{Proportion}\tabularnewline
+\hline
+\textless  15 months&$ 21$&$10$&$0.47$\tabularnewline
+15 months – 3 years&$  7$&$ 1$&$0.14$\tabularnewline
+4 – 9 years&$  8$&$ 0$&$0.00$\tabularnewline
+10 -1 19 years&$132$&$12$&$0.09$\tabularnewline
+\textgreater  19 years&$ 33$&$11$&$0.33$\tabularnewline
+Total&$201$&$34$&$0.17$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:freq}
+\end{table}
+
+The majority (191/192) of the measles cases stayed in the hospital from 0 to 7 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
+Eighty-three of the cases stayed for 0 days and the remaining cases stayed a mean of 2.30 (median 2) days.
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2008-2013 appear in Table \ref{table:cases}.
+The cost for these 192 hospitalizations was \$289,964.
+Of the 292 cases, 87 were female and 105 were male.
+The number of days spent in the hospital was 251, making the mean daily cost of \$1,509.
+Of the 251 days, 114 were spent by females and 137 by males.
+The mean hospital stay per case was 1.31 and 1.30 days for females and males, respectively.
+The mean cost per case was \$1,573 and \$1,456 for females and males, respectively.
+
+\begin{table}
+\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008-2013}
+%latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllrrl}
+\hline\hline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Gender}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Days}&\multicolumn{1}{c}{Per.case}\tabularnewline
+\hline
+2008&F&714&$  1$&$  0$&714\tabularnewline
+&M&2,324&$  2$&$  1$&1,162\tabularnewline
+                    &Total&3,038&$  3$&$  1$&1,013\tabularnewline
+2009&F&11,953&$  7$&$ 15$&1,708\tabularnewline
+&M&28,830&$ 22$&$ 23$&1,310\tabularnewline
+                    &Total&40,782&$ 29$&$ 38$&1,406\tabularnewline
+2010&F&5,884&$  4$&$  5$&1,471\tabularnewline
+&M&817&$  1$&$  0$&817\tabularnewline
+                    &Total&6,701&$  5$&$  5$&1,340\tabularnewline
+2011&F&103,460&$ 66$&$ 86$&1,568\tabularnewline
+&M&101,842&$ 66$&$103$&1,543\tabularnewline
+                    &Total&205,303&$132$&$189$&1,555\tabularnewline
+2012&F&13,054&$  8$&$  6$&1,632\tabularnewline
+&M&15,486&$ 11$&$  6$&1,408\tabularnewline
+                    &Total&28,540&$ 19$&$ 12$&1,502\tabularnewline
+2013&F&1,800&$  1$&$  2$&1,800\tabularnewline
+&M&3,530&$  3$&$  4$&1,177\tabularnewline
+                    &Total&5,330&$  4$&$  6$&1,333\tabularnewline
+Total&&289,694&$192$&$251$&1,509\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cases}
+\end{table}
+
+Regression analyses showed statistically significant associations between cost of hospitalization and three variables, length of hospitalization, case age and year of case, and a less strong association with case gender (Table \ref{table:regression}).
+Results showed an increase of approximately \$358 per day of hospitalization, \$139 for a female case, and \$1 per year of case diagnosis.
+The cost was found to decrease by approximately \$5 per year of case age.
+
+\begin{table}
+\caption{Regression results (R$^{2}$ adj = 0.9, p-value <0.001) for measles hospitalization cost based on length of stay (days), gender, case age and year of case (n=192) in New Zealand, 2008 - 2013}
+%latex.default(costtable5, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrl}
+\hline\hline
+\multicolumn{1}{c}{Variable}&\multicolumn{1}{c}{Coefficient}&\multicolumn{1}{c}{P.value}\tabularnewline
+\hline
+Length of stay (days)&$ 357.71$&\textless  0.001\tabularnewline
+Gender (0 = F, 1 = M)&$-138.55$&0.084\tabularnewline
+Case age (years)&$  -5.23$&0.032\tabularnewline
+Year of case&$   0.60$&\textless  0.001\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:regression}
+\end{table}
+
+Wages lost due to measles were calculated based on the assumptions that 5 days of work were lost for each case; however, individuals between 0-15 years of age were not assumed to be employed and therefore did not suffer an income loss.
+If the case were < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
+Total wage lost for the 192 cases and care givers was estimated to be \$166,638.
+This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases <15 years of age.
+
+This final figure brings an approximate estimate of \$786,479 for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
+
+\subsection{Cost analysis discussion}
+The results presented here are based on available data.
+While some of the data are complete and detailed, this is not true of all the data.
+In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
+For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
+In addition to this information, similar data would be needed for cases occurring outside the period 2011-2013 at publicly funded hospitals.
+Similar data would be needed for non-publicly funded hospitals, e.g.
+private clinics.
+Detailed measles outbreak management costs were provided for the period of January 1 - March 9, 2014.
+Similar data are needed for the period preceding 2014.
+If detailed data, such as that provided for early 2014, aggregated data would be acceptable.
+However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
+As other studies have demonstrated, direct costs required to manage measles are not linear.
+
+In other outbreaks, the average cost per measles case was estimated to be US\$ 254, US\$ 276, and US\$ 307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$ 142,542.
+In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
+They estimated direct costs per case to be less than US\$ 500.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively (Prouty and Lakin, 2000).
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$ 113,864.
+The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalization) to \$ 29,556 (encephalitis and hospitalization for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) (Zhou et al., 2004).
+
+\subsection{Cost analysis summary}
+\begin{itemize}
+\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$ $\frac{3}{4}$ million.
+\end{itemize}
+
+\subsection{Future cost-benefit analyses}
 Using the results above we aim to:
 \begin{itemize}
-\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date and in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified if the multivariate analyses can be performed once appropriate denominator data are obtained.
+\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modeling.
 \end{itemize}
+
+We also require additional clarifications of the data, regarding what hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight?
+Or, does it mean the case stayed for less than 24 hours?
+Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
 
@@
 
-A previously-published model of the dynamics of measles infections
-in New Zealand has been used to evaluate the vaccination strategy in New Zealand of
-MMR1 at 15 months and MMR2 before 5 years. The results show that
-achieving coverage of >90\% at both vaccination opportunities is
-necessary if future epidemics of measles are to be prevented. [AND LIKELY MORE?!]
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years.
+The results show that achieving coverage of >90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
 
-The original mathematical model for the dynamics of measles in New Zealand
-was prepared in 1996 (Tobias & Roberts 1998). It successfully predicted the
-1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
-et al 1998; Roberts & Tobias 2000). A subsequent extension of this work in
-1998 showed that the then current schedule of MMR1 at 15 months and MMR2
-at 11 years was insufficient to prevent further epidemics. The schedule was
-changed in 2000 with MMR2 now being administered before 5 years (Anon.
-2002a).
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 (Tobias & Roberts 1998) successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
+et al 1998; Roberts & Tobias 2000).
+Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
+The model developed by Roberts & Tobias (2000) supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years (Anon.
+2002a) and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasized that it is necessary to maintaining high coverage rates in order to prevent future epidemics.
 
-A variety of similar models for measles vaccination strategies have been developed
-by other authors for various regions (Agur et al 1993; Babad et al 1995;
-Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001). These have invariably
-been based on sets of nonlinear differential equations, and the conclusions
-reached have been similar. The differences in the models have been in the details
-of the representation of the infectious period, and in the ways in which
-the age and contact structures of the population have been specified.
+These results were in line with those obtained by other authors, for example: Babad et al (1995) advocated a two-dose schedule for England and Wales, with the second vaccination given at age four; and Gay et al (1998) recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, Agur et al (1993) found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+These authors all agree that two
+vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+A different approach was taken by Wallinga et al (2001).
+These authors took existing policies in eight European countries and estimated the coverage rates required to reduce Rv below one.
+They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately
+87\%.
 
-The model developed by Roberts & Tobias (2000) supported the change in the
-immunisation schedule that took effect in January 2001, at which time MMR2
-was changed from delivery at 11 years to delivery before the age of five. These
-results were in line with those obtained by other authors, for example: Babad
-et al (1995) advocated a two-dose schedule for England and Wales, with the
-second vaccination given at age four; and Gay et al (1998) recommended a
-second vaccination at either 18 months or five years, to complement the first
-vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
-vaccinating 85\% of susceptible children aged one to seven years at five-yearly
-intervals would prevent epidemics in Israel. These authors all agree that two
-vaccinations at no less than five years apart are necessary to prevent measles
-epidemics.
-A different approach was taken by Wallinga et al (2001). These authors took
-existing policies in eight European countries and estimated the coverage rates
-required to reduce Rv below one. They found that results depended on the
-age at delivery, but no strategy succeeded if coverage rates were below approximately
-87\%. Our results presented in Table 1 are similar where coverage is
-assumed to be the same at MMR1 and MMR2. The results also show the absolute
-necessity of maintaining high coverage rates in order to prevent future
-epidemics. It is difficult to estimate the proportion of the school-age population
-that have been effectively immunised, but this is continuously being
-diluted by children who are not immunised. Wallinga et al (2000) noted that
-in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
-and that under most plausible assumptions for contact rates even
-100\% coverage would be insufficient to prevent epidemics.
-Our final conclusion must depend on how much faith is placed in the model.
-The results are consistent with those of other authors, and from Table 1 it
-could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
-to prevent future measles epidemics. However, a study by Glass et al (2004)
-in the Netherlands showed that high overall levels of measles vaccination can
-obscure pockets of poor coverage, resulting in localised regions with increased
-risk of infection. Our results indicate that overall targets or 90\% or more will
-need to be achieved to prevent future epidemics in New Zealand.
+A variety of similar models for measles vaccination strategies have been developed by other authors for various regions (Agur et al 1993; Babad et al 1995; Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001) based on sets of nonlinear differential equations (ODE), and the conclusions reached have been similar.
+The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
 
-The quantity that determines whether an epidemic will occur is the basic
-reproduction number of the infection, R0. This is defined as the expected
-number of secondary infections that would arise from a single primary infection
-introduced into a fully susceptible population (Anderson & May 1991;
-Diekmann & Heesterbeek 2000). Clearly if R0 > 1 an epidemic will occur following
-an introduction of infection. The best estimate we had for measles in
-New Zealand was R0 = 12.8, the change in the birth rate (56780 p.a. in 2004,
-www.stats.govt.nz) could have reduced this slightly to R0 = 12.5 (see results
-section). Recall that R0 is calculated in the absence of control measures. We
-are interested in two related quantities:
+It is difficult to estimate the proportion of the school-age population that have been effectively immunised, but this is continuously being diluted by children who are not immunised.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, Glass et al (2004) in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection.
 
-The basic reproduction number of the infection under vaccination, Rv,
-is the expected number of secondary infections that would arise from a
-single primary infection introduced into a vaccinated population at equilibrium.
-This is a robust indicator of the performance of a vaccination
-schedule. If Rv < 1 epidemics are prevented.
+The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, R0.
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population (Anderson & May 1991; Diekmann & Heesterbeek 2000).
+If R0 > 1 an epidemic will occur following an introduction of infection.
+The best estimate we had for measles in New Zealand was R0 = 12.8 - 12.5 (Roberts, see results section).
+The basic reproduction number of the infection under vaccination, Rv, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
+If Rv < 1 epidemics are prevented.
+The case reproduction number of the infection at time t, Rt, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
 
-The case reproduction number of the infection at time t, Rt, is the
-expected number of secondary infections that arise from a single infection
-at a particular time. This depends on the number in the population who
-are susceptible, either through prior infection or vaccination.
+To understand the level of immunity in the population, how likely an outbreak was of becoming endemic, we estimated Rv from all the outbreaks since 2009.
+To do this we estimated Rt, following an adaptation of the methods in [REF], and used these to estimate the Rv of measles in the recent outbreaks, which helps understand the transmission dynamics of measles in the partially immune population.
 
-The dynamics of the process may be summarised as follows. Assume that
-the population has Rt < 1 for measles. As the number of susceptibles in the
-population increases, through children that miss their scheduled immunisations,
-then Rt increases. An epidemic occurs when Rt has increased above
-one. During the epidemic the number of susceptibles and hence Rt decreases.
-
-Calculated values of Rv for the various coverage levels of MMR1 and MMR2
-that were considered are presented in Table 1.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
-     \caption{Measles incidence from 1997 to 2014}
-     %\label{fig:test image}
-\end{figure}
+To estimate Rt we were required to compute the generation time for measles.
+We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from %# http://www.sciencedirect.com/science/article/pii/S0022519311003146
+We then estimated Rt using the methods adapted from [REF] from the incidence data for each outbreak.
+We defined outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
+The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.
 
@@
      \includegraphics[width=0.8\textwidth]{cases_by_dhb_2009_2014.pdf}
-     \caption{Measles cases by DHB from 2009 to 2014}
+     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
+     \label{fig:dhbcases}
 \end{figure}
@@
      \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
-     \caption{Measles data classified as outbreaks for R0 estimation}
+     \caption{Measles data classified as outbreaks for reproductive number of the infection (Rv) estimation}
      \label{fig:outbreaks}
@@
      \includegraphics[width=0.8\textwidth]{averageR0.pdf}
-     \caption{Estimates of R0 for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \caption{Estimates of Rv (R0) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \label{fig:r0}
 \end{figure}
 
-...The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases.
-[COMMENT?
-Unsure, though don't want to be dismissive].
-....
+The estimated Rv for each outbreak is shown in (Figure~\ref{fig:r0}).
+The probability density of the Rv estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an Rv well above one and thus we may expect this outbreak to persist if conditions remain the same.
 
-\section{Future analyses}
+These analyses also imply that the reguler (approximately yearly) importation of measles is an ongoing process.
+Given the risk of importation of measles, as highlighted in the \emph {risk analysis} section is likely to continue these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+The measles outbreak in 2011-2012 had an Rv of just greater than one, and yet it persisted for over 12 months.
+This implies that the current outbreak may persist within the population for a substantial period, given it's Rv is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other Rv estimates is that the 2013-2014 may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+However, sub-clinical and underreporting may lower the estimate.
+The relative contributions of both to our estimates are currnelty unknown.
 
-\section{Future cost-benefit analyses}
-Using the results above we aim to:
+\subsection{Summary of modeling}
+
 \begin{itemize}
-\item Update previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
+\item Regular introductons of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section [X]).
+\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
+ \item The reproduction number for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to become endemic.
+\end{itemize}
+
+\section{Future modeling analyses}
+Future modeling analyses we aim to perform are:
+\begin{itemize}
+\item An update of previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
 \item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
@@
 
+\section {Summary of key findings}
+\begin{itemize}
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles eslewhere in the globe.
+\item The cost of the current measles outbreak is estimated to be at least \$$\frac{3}{4}$ million.
+\item Analyses of outbreak data suggest that measles Rv values often include 1 and in this year, 2014, well above one.
+This analyses suggest improved vaccination is a requesite to prevent measles becoming endemic again.
+\end{itemize}
+
 \section{Acknowledgments}
-The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR for help in obtaining the appropriate materials for analyses.
+The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR, and June Atkinson from University of Otago for help in obtaining the appropriate materials for analyses.
 

### 2014-07-03 interimreport1.Rnw
a bunch of edits (not run through TeX, so it may not build...), most of which are minor.  There's some labels included for section referencing

--- 2014-07-02 interimreport1.Rnw
@@
 \usepackage{graphicx}
+\usepackage{todonotes}
 
@@
 
-Summary of findings
+New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
+In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
+We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modeling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more.
+Our key findings were:
+\begin{itemize}
+\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
+\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
+\item Analyses of outbreak data suggest that measles basic reproductive number (Rv, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
+This analyses suggest improved vaccination is a requisite to prevent measles becoming endemic again.
+\end{itemize}
 
@@
 
-As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committted to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
-The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved this.
+As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
+The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.
 
 The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011/2012 (mainly in Auckland region) and another significant outbreak is currently going on in the Auckland and Waikato regions.
-The outbreak in 2011-2012 lasted for more than 12 months and the current 2013/2014 outbreak started at the end of December 2013 (as on \date{\today}).
-In 2013, prior to the 2013/2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) it can request verification of non-endemic status three years after the last case of the 2011-12 outbreak in June 2012.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
+The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{\today}).
+In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
 
-Previous analyses, including two of measles in New Zealand by Prof.
+Previous analyses of measles, including two in New Zealand by Prof.
 Roberts (INSERT REFS), have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, as described above, outbreaks still occur.
-Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are perhaps likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
+Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 
-A measles risk assessment has been started by the Ministry of Health to better assess current and future population immunity and high risk groups.
+A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups.
 Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
@@
 
-Overall, we thought the review was very thorough.
-The report included substantial background information on measles immunisation in New Zealand (\emph{section 1.3}), the epidemiology of measles in New Zealand (\emph{section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{section 3}), and the levels of population immunity against the virus (\emph{section 4}).
-Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{section 4.2}) and the sustainability of the national immunisation programme (\emph{section 5}).
+Overall, the review was very thorough.
+The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
+Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).
 
-Within the report there are many tables and figures which given substantial detail on the measles situation in New Zealand.
+Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand.
 Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
 
-Specific epidemiological details were provided for the 2011/2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month, 2011 and 2012, with additional breakdown by age group, 2011 and 2012 (\emph{Figure 5}) and per 100,000 population (e.g.
-\emph{Figures 6-8}).
-Similar presentation of the case data are provided for ethnicity (e.g.
-\emph{Figures 9-10}) and New Zealand Index of Deprivation (NZDep) (e.g.
-\emph{Figures 11-13}).
-Three figures,\emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
+Specific epidemiological details were provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}).
+Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}).
+Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
 
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles is not endemic and provide some preliminary analyses on the outbreaks since endemic measles 'elimination' (see section X) that gives information regarding the likelihood of this occuring.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modeling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
 
-We agree with the report's conclusions that testing for measles is performed appropriately within the required timeframe, but clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions.
+We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe.
+Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
+Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
+However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
+This is particularly in light of our preliminary modeling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modeling}).
 
-Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
-However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approching or over 95\% levels for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modeling results (section X).
-
-The analyses we thought could further inform the understanding of risk from measles infection were:
+The analyses we believe could further inform the understanding of risk from measles infection are:
 \begin{itemize}
-\item Multivariate modelling to account for confounding within the univariate analyses.
-\item Inclusion of multivariate model results with local level immunity data to develop a risk map for NZ. 
+\item Multivariate modelling to account for confounding within the univariate analyses.\todo{one L or two in modelling?}
+\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
 \item Analyses of changing risk factors through time during outbreaks.
@@
 
-\section{Additional risk analyses}
+\subsection{Additional risk analyses}\label{sec:risk_analyses}
 
-In this section we provide a brief description of the work we are performing that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
-These analyses are intended to build on the excellent analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
+In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
+These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
 
-We received the raw case data from The Institute of Environmental Science and Research Ltd (ESR) EpiSurv on 27 June 2014.
-Initial analyses of those data suggest that we require denominator data to perform multivariate analyses.
-Specifically for the multivariate analyses we wish to perform we require \emph{Age * Prioritized Ethnicity * NZDep} data for New Zealand to see if the interactions among case covariates give us any additional information on risk from the univariate analyses already performed in the report reviewed above.
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
+     \caption{Measles incidence from 1997 to 2014}
+     \label{fig:incidence1997}
+\end{figure}
 
-These data are currently unavailable in the form we require and after contacting ESR and NZ Statistics we and NZ statistics have contacted the University of Otago to see if we can obtain these denominator data.
-Additional data we believe would enable us and the Ministry to understand measles risk better is fine scale (lower than DHB) immunization coverage data.
+In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
 
-We have reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help better understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers, we sought to understand better the risk of measles importation.
-For our initial analyses, we used the arrivals data from \href{http://www.immigration.govt.nz/}{www.immigration.govt.nz} as an estimate of human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO \href{http://www.who.int/research/en/}{www.who.int/research/en/}.
+For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).\todo{this sentence seems to need a bit more after it?
+Something like "to produce pretty maps"}
 
@@
 \begin{table}
-\caption{Immigration to New Zealand (2012)}
+\caption{Immigration numbers to New Zealand (2012)}
 <<results=tex,echo=FALSE>>=
@@
 @
+\label{table:immigration12}
 \end{table}
@@
 @
+\label{table:cover12}
 \end{table}
@@
 \begin{table}
-\caption{Measles Incidence per million (2012)}
+\caption{Measles incidence per million (2012)}
 <<results=tex,echo=FALSE>>=
@@
 @
+\label{table:incidence12}
 \end{table}
@@
 @
-\end{table}
+\label{table:risk12}
+\end{table}\todo{these tables need country rather than id as the first column name}
 
-The preliminary analyses of 2012 data suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table 1.
-However, vaccination coverage is lowest and measles incidence highest in less developed nations (Tables 2 and 3).
-Though the interactions between these different risk factors are unknown, simple products of measles incidence in 2012 and immigration numbers in 2012 suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some countries also poses a risk.
-These data are shown in Table 4.
-The data for all the variables for each nation state and territories for 2012 are plotted in Figure 1 and the risk map for measles incidence and immigration in Figure 2.
+Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
+However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:incidence12} and Table \ref{table:cover12}).
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
+These data are shown in Table \ref{table:risk12}.
+The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
 
@@
 \caption{Risk map from measles incidence and immigration rates}
+\label{fig:risk12}
 \end{figure}
 
-Though [SOMETHING ON MEASLES], immigration rates to New Zealand are increasing. [INSERT FIGURE]
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE>>=
+change<-read.csv("change.csv",header=T)
+plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
+     bty="n")
+ylim <- range(c(0,max(change[,4])))
+xlim <- range(change[,1])
+par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
+plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
+     ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
+@
+\end{center}
+\caption{Trend in global per capita measles incidence}
+\label{fig:trendincidence}
+\end{figure}
 
+Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
+This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
+Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).
+This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
+
+\begin{figure}[h!]
+\begin{center}
 <<fig=TRUE,echo=FALSE>>=
@@
 lmres<-lm(res~time,data=ttimes)
-summary(lmres)
+#summary(lmres)
 plot(ttimes$res,xaxt="n",type="l",xlab="Date",ylab="")
@@
 @
+\end{center}
+\caption{Trend in total immigrations into New Zealand}
+\label{fig:trendimmigration}
+\end{figure}
 
-\section{Future risk analyses}
-We received ESR EpiSurv measles case data on the 27 June 2014. The following are analyses are still in progress and aim to include them in later reports:
+\subsection{Risk analysis summary}
 \begin{itemize}
-\item Multivariate regression analyses to look for interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using more than the 2012 data.
+\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
+\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
 \end{itemize}
-The data gaps that we have that will hinder us in completing all the analyses are:
+
+\subsection{Future risk analyses}\sec{risk_analyses}
+We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
+Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
+Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above.
+These data only became available to us on 3 July 2014, provided by the University of Otago.
+
+The following analyses are in progress, for inclusion in later reports:
 \begin{itemize}
-\item A lack of meshblock level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
+\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
+\item Update of the importation risk analyses using a broader range of years, including modeling the trend in importations.
+\end{itemize}
+Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
+Thus the data gap that we have that will hinder us providing fine scale risk maps are:
+\begin{itemize}
+\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel. \todo{added a bit here you may want to check}
+\end{itemize}
+
 We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
-\emph{meshblock}) level would allow better understanding of risk and resource allocation.
-\item A lack of data at any level that combines NZDep with age, prioritized ethnicity, and other variables, that allow better use of data collected by the Ministry, ESR, and NZ statistics for multivariate analyses.
-\end{itemize}
+meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
 
@@
 
-In this section we provide an analysis of the costs involved with this current measles outbreak.
+In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current measles outbreak.
 
-... [TEXT HERE] ...
+Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles (Clements and Hussey, 2004).
+Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities.
+Wolfson et al.
+(2007) estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
+Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
+They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination.
+The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 (WHO, 2013).
+In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.
 
-\section{Future cost-benefit analyses}
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million (Carabin et al., 2003).
+The estimated cost for a case ranged from US\$ 189--344 (Carabin et al., 2003); however, the average estimated cost of a typical hospital case ranges from US\$ 967--1,755 (Carabin et al., 2002).
+Stack et al (2011) estimated the economic benefits from cases averted due to measles vaccination.
+They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$ 10 billion of costs averted between 2011 and 2020.
+Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
+
+Italy has the highest reported annual cost of measles among industrialised countries (Carabin et al., 2003).
+In 2001, it reported losses related to measles of approximately US\$ 50 million.
+The economic impact of a large measles outbreak in Italy, 2002--03 examined the costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis.
+The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
+The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 ~ NZ\$ 2.0 in 2002-03), or approximately \euro 1,700 per case.
+The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
+The average daily cost of a hospital stay was  \euro 327.
+
+An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases (Flego et al., 2013).
+Seven (27\%) of the cases required hospitalisation for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
+During this outbreak, a total of 1,395 contacts were identified and managed by a public health unit in western Sydney.
+The mean number of contacts per case was 54 (median = 28, maximum = 206).
+The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$ 48,000, with 90\% of this being associated with staff time.
+
+Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years.
+In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 (Wichmann et al., 2009).
+However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children (Siedler et al, 2006).
+An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
+In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case.
+Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised.
+The mean costs of the hospitalised patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623.
+In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing.
+Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case.
+In addition, to determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
+
+Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world (Parker et al., 2006).
+Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.
+Ortega-Sanchez et al.
+(2014) estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
+The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$ 2.7 and 5.3 million.
+Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$ 298 per contact.
+Dayan et al.
+(2005) calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
+They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination.
+In total, it was estimated the direct costs associated with three cases of measles was US\$ 142,452, or nearly US\$ 50,000 per case.
+
+Parker et al.
+(2006) reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
+A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak.
+Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak.
+Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overhead.
+This amounted to a direct cost of \$4,932 per measles case.
+These costs did not include either patient care or indirect costs, which would have made the total and per case cost higher.
+
+Coleman et al.
+(2012) estimated the direct medical and public health costs in response to a single case of refugee-imported measles.
+Costs included labor, translation and benefits for public health workers.
+In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis.
+In total, 387 hours were associated with this single case, resulting in a cost of US\$ 11,881.
+In addition, per-contact costs amounted to US\$ 264.
+The cost of hospitalisation for the 3-day stay by the index case was US\$ 931.
+Additional costs were associated with physician visits (US\$ 294), vaccine and immunoglobulin (US\$ 1,765), mileage (US\$ 205) and immunologic screening tests for the parents' exposed to measles (US\$ 240) for a total of US\$ 23,816.
+
+Economic analyses of measles control programs have shown them to be financially effective.
+In the Republic of Korea, the economics of alternative measles vaccination programs were compared.
+All of the alternatives were found to be economically efficient (benefit/cost ratio (B/C) > 1.0), with the alternative using two doses of the MMR program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27).
+
+The purpose of the current study is to estimate the cost of the current measles outbreak in New Zealand.
+Using this information, we will then evaluate the economics of alternative measles control strategies in order to provide additional information to public health officials and decision makers.
+
+\subsection{Methods}
+Costs were evaluated as either direct or indirect.
+Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
+Direct costs can be divided into medical and non-medical (Saha and Gerdtham, 2013).
+Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
+Personnel time (investigation and emergency response), materials (phone calls, vaccine), personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
+Direct non-medical costs include transportation to and from health care providers.
+
+Indirect costs are productivity losses for the case and/or health care provider, e.g.
+parent of a school child.
+Indirect costs included work loss for cases and caregivers.
+This could also include the economic value of premature life lost, costs associated with permanent disability, e.g.
+deafness and mental retardation.
+Commonly the human value approach (HVA) has been used to estimate economic impact of life.
+The HVA measures the potential future earnings of an individual and discounts it into a present value.
+Typically this is 3\% but 5\% has also been used in a sensitivity analysis, which is more compared to non-human life calculations and will tend to reduce the present value of the future earnings (saved by avoiding a case).
+
+Data for the current measles outbreak were obtained from the New Zealand Ministry of Health, from 2008 through June 2014.
+Data included information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, case weight and associated cost.
+
+Cost of the Auckland Regional Public Health Service (ARPHS) for measles response were obtained from the Ministry of Health.
+Data, for the period January 1 - March 9, 2014, reported salaries for people involved with the measles outbreak management medical team.
+The costs were reported as direct, additional (above normal budgeting) costs required to enable the management of measles.
+It includes a breakdown by individual performing the work and whether it was during the normal work schedule (Monday to Friday, M-F) or weekends.
+Normal work was calculated as $1.2 \times \textsf{full time equivalent (FTE)} \times \textsf{number of days worked}$.
+Overtime was calculated as $1.6\times\textsf{FTE}$ (M-F) and $2.0 \times \textsf{FTE}$ (weekend).
+A full day was considered as 8 hours worked.
+Salary (hourly) rates were calculated for the following: public health nurse (PHN, \$36), public health assistant (PHA, \$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), incident management team (IMT), which had the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and safety/security officer (SSO) (\$26).
+In addition, measles operations personnel were calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
+
+Mean wages for New Zealand workers, by age and gender were obtained for the period, 2008--2013 from the New Zealand Income Survey (Statistics New Zealand, 2013).
+Measles cases were assumed to not work for a period of 5 days.
+Similarly, a care taker was assumed to not work for 5 days if the case were less than 20 years of age.
+In order to calculate the wage loss associated with the care taker, it was assumed that the person was a female between the ages of 35-39.
+Age and gender information for the 192 publicly funded hospital discharges with a measles primary diagnosis from 2008--2013 were matched to the New Zealand wage file to calculate lost wages due to measles.
+
+A regression analysis was performed to test for significant associations between hospital cost and the following explanatory variables: case age at discharge, gender, length of stay (days) and year of case.
+
+\subsection{Results}
+Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
+The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\ref{table:hosp}.
+
+<<echo=FALSE>>=
+costtable1<-read.csv("costtable1.csv",header=T)
+@
+
+\begin{table}
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable1, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:direct}
+\end{table}
+
+The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was \$289,694 (Table~\ref{table:hosp}).
+The mean cost per case was \$1,509.
+The mean cost per day of stay in the hospital was \$1,154.
+
+<<echo=FALSE>>=
+costtable2<-read.csv("costtable2.csv",header=T)
+@
+
+\begin{table}
+\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008--2013}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable2, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:hosp}
+\end{table}
+
+From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
+The number of cases by age group is shown in Table~\ref{table:freq}.
+Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 0.47 and 0.33, respectively.\todo{make these percentages?}
+
+<<echo=FALSE>>=
+costtable3<-read.csv("costtable3.csv",header=T)
+@
+
+\begin{table}
+\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable3, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:freq}
+\end{table}
+
+The majority (191/192) of the measles cases stayed in the hospital from 0 to 7 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
+Eighty-three of the cases stayed for 0 days and the remaining cases stayed a mean of 2.30 (median 2) days.
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2008--2013 appear in Table~\ref{table:cases}.
+The cost for these 192 hospitalisations was \$289,964.
+Of the 292 cases, 87 were female and 105 were male.
+The number of days spent in the hospital was 251, making the mean daily cost \$1,509.
+Of the 251 days, 114 were spent by females and 137 by males.
+The mean hospital stay per case was 1.31 and 1.30 days for females and males, respectively.
+The mean cost per case was \$1,573 and \$1,456 for females and males, respectively.
+
+<<echo=FALSE>>=
+costtable4<-read.csv("costtable4.csv",header=T)
+@
+
+\begin{table}
+\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008--2013}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable4, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cases}
+\end{table}
+
+Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
+Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis.
+The cost was found to decrease by approximately \$5 per year of case age.\todo{later on we could assess whether or not a linear term for case age is appropriate, or what if any interaction there is between age and length of stay?}
+
+<<echo=FALSE>>=
+costtable5<-read.csv("costtable5.csv",header=T)
+@
+
+\begin{table}
+\caption{Regression results ($R^{2}_\textsf{adj} = 0.9$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=192$) in New Zealand, 2008 -- 2013}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable5, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:regression}
+\end{table}
+
+Wages lost due to measles were calculated based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
+If the case were less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
+Total wage lost for the 192 cases and care givers was estimated to be \$166,638.
+This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases under 15 years of age.
+
+This final figure brings an approximate estimate of \$786,479 for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
+
+\subsection{Cost analysis discussion}
+The results presented here are based on available data.
+While some of the data are complete and detailed, this is not true of all the data.
+In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
+For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
+In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
+In addition, similar data would be needed for non-publicly funded hospitals, e.g.
+private clinics.
+
+Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
+Similar data are needed for the period preceding 2014.
+If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
+However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
+As other studies have demonstrated, direct costs required to manage measles are not linear.
+
+In other outbreaks, the average cost per measles case was estimated to be US\$ 254, US\$ 276, and US\$ 307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$ 142,542.
+In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
+They estimated direct costs per case to be less than US\$ 500.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively (Prouty and Lakin, 2000).
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$ 113,864.
+The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalisation) to \$ 29,556 (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) (Zhou et al., 2004).
+
+\subsection{Cost analysis summary}
+\begin{itemize}
+\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$ 750,000.
+\end{itemize}
+
+\subsection{Future cost-benefit analyses}
 Using the results above we aim to:
 \begin{itemize}
-\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date and in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified if the multivariate analyses can be performed once appropriate denominator data are obtained.
+\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modeling.
 \end{itemize}
 
-\section{Modeling measles epidemics}
+We also require additional clarifications of the data, regarding what hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight?
+Or, does it mean the case stayed for less than 24 hours?
+Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
 
-A previously-published model of the dynamics of measles infections
-in New Zealand has been used to evaluate the vaccination strategy in New Zealand of
-MMR1 at 15 months and MMR2 before 5 years. The results show that
-achieving coverage of >90\% at both vaccination opportunities is
-necessary if future epidemics of measles are to be prevented. [AND LIKELY MORE?!]
+\section{Modeling measles epidemics}\label{sec:epidemic_modeling}
 
-The original mathematical model for the dynamics of measles in New Zealand
-was prepared in 1996 (Tobias & Roberts 1998). It successfully predicted the
-1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
-et al 1998; Roberts & Tobias 2000). A subsequent extension of this work in
-1998 showed that the then current schedule of MMR1 at 15 months and MMR2
-at 11 years was insufficient to prevent further epidemics. The schedule was
-changed in 2000 with MMR2 now being administered before 5 years (Anon.
-2002a).
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years.
+The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
 
-A variety of similar models for measles vaccination strategies have been developed
-by other authors for various regions (Agur et al 1993; Babad et al 1995;
-Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001). These have invariably
-been based on sets of nonlinear differential equations, and the conclusions
-reached have been similar. The differences in the models have been in the details
-of the representation of the infectious period, and in the ways in which
-the age and contact structures of the population have been specified.
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 (Tobias & Roberts 1998) successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
+et al 1998; Roberts & Tobias 2000).
+Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
+The model developed by Roberts & Tobias (2000) supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years (Anon.
+2002a) and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics.
 
-The model developed by Roberts & Tobias (2000) supported the change in the
-immunisation schedule that took effect in January 2001, at which time MMR2
-was changed from delivery at 11 years to delivery before the age of five. These
-results were in line with those obtained by other authors, for example: Babad
-et al (1995) advocated a two-dose schedule for England and Wales, with the
-second vaccination given at age four; and Gay et al (1998) recommended a
-second vaccination at either 18 months or five years, to complement the first
-vaccination at 12 months in Canada. In addition, Agur et al (1993) found that
-vaccinating 85\% of susceptible children aged one to seven years at five-yearly
-intervals would prevent epidemics in Israel. These authors all agree that two
-vaccinations at no less than five years apart are necessary to prevent measles
-epidemics.
-A different approach was taken by Wallinga et al (2001). These authors took
-existing policies in eight European countries and estimated the coverage rates
-required to reduce Rv below one. They found that results depended on the
-age at delivery, but no strategy succeeded if coverage rates were below approximately
-87\%. Our results presented in Table 1 are similar where coverage is
-assumed to be the same at MMR1 and MMR2. The results also show the absolute
-necessity of maintaining high coverage rates in order to prevent future
-epidemics. It is difficult to estimate the proportion of the school-age population
-that have been effectively immunised, but this is continuously being
-diluted by children who are not immunised. Wallinga et al (2000) noted that
-in Italy only MMR1 was offered at 18 months with no second vaccination opportunity,
-and that under most plausible assumptions for contact rates even
-100\% coverage would be insufficient to prevent epidemics.
-Our final conclusion must depend on how much faith is placed in the model.
-The results are consistent with those of other authors, and from Table 1 it
-could be deduced that 85\% coverage at MMR1 and MMR2 could be sufficient
-to prevent future measles epidemics. However, a study by Glass et al (2004)
-in the Netherlands showed that high overall levels of measles vaccination can
-obscure pockets of poor coverage, resulting in localised regions with increased
-risk of infection. Our results indicate that overall targets or 90\% or more will
-need to be achieved to prevent future epidemics in New Zealand.
+These results were in line with those obtained by other authors, for example: Babad et al (1995) advocated a two-dose schedule for England and Wales, with the second vaccination given at age four; and Gay et al (1998) recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, Agur et al (1993) found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+These authors all agree that two
+vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+A different approach was taken by Wallinga et al (2001).
+These authors took existing policies in eight European countries and estimated the coverage rates required to reduce Rv below one.
+They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately
+87\%.
 
-The quantity that determines whether an epidemic will occur is the basic
-reproduction number of the infection, R0. This is defined as the expected
-number of secondary infections that would arise from a single primary infection
-introduced into a fully susceptible population (Anderson & May 1991;
-Diekmann & Heesterbeek 2000). Clearly if R0 > 1 an epidemic will occur following
-an introduction of infection. The best estimate we had for measles in
-New Zealand was R0 = 12.8, the change in the birth rate (56780 p.a. in 2004,
-www.stats.govt.nz) could have reduced this slightly to R0 = 12.5 (see results
-section). Recall that R0 is calculated in the absence of control measures. We
-are interested in two related quantities:
+A variety of similar models for measles vaccination strategies have been developed by other authors for various regions (Agur et al 1993; Babad et al 1995; Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001) based on sets of nonlinear differential equations (ODE), and the conclusions reached have been similar.
+The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
 
-The basic reproduction number of the infection under vaccination, Rv,
-is the expected number of secondary infections that would arise from a
-single primary infection introduced into a vaccinated population at equilibrium.
-This is a robust indicator of the performance of a vaccination
-schedule. If Rv < 1 epidemics are prevented.
+It is difficult to estimate the proportion of the school-age population that have been effectively immunised, but this is continuously being diluted by children who are not immunised.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, Glass et al (2004) in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection.
 
-The case reproduction number of the infection at time t, Rt, is the
-expected number of secondary infections that arise from a single infection
-at a particular time. This depends on the number in the population who
-are susceptible, either through prior infection or vaccination.
+The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$\todo{not sure if you want it like this throughout?}.
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population (Anderson & May 1991; Diekmann & Heesterbeek 2000).
+If $R_0 > 1$ an epidemic will occur following an introduction of infection.
+The best estimate we had for measles in New Zealand was R0 = 12.8 -- 12.5\todo{this seems an odd range?} (Roberts, see results section).
+The basic reproduction number of the infection under vaccination, Rv\todo{or $R_v$ ?}, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
+If $R_v < 1$ epidemics are prevented.
+The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
 
-The dynamics of the process may be summarised as follows. Assume that
-the population has Rt < 1 for measles. As the number of susceptibles in the
-population increases, through children that miss their scheduled immunisations,
-then Rt increases. An epidemic occurs when Rt has increased above
-one. During the epidemic the number of susceptibles and hence Rt decreases.
+To understand the level of immunity in the population and how likely an outbreak was of becoming endemic, we estimated Rv from all the outbreaks since 2009.
+To do this we estimated Rt, following an adaptation of the methods in [REF]\todo{there's 2 references in \texttt{?est.R0.TD}}, and used these to estimate the Rv of measles in the recent outbreaks, which helps understand the transmission dynamics of measles in the partially immune population.
 
-Calculated values of Rv for the various coverage levels of MMR1 and MMR2
-that were considered are presented in Table 1.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
-     \caption{Measles incidence from 1997 to 2014}
-     %\label{fig:test image}
-\end{figure}
+To estimate Rt we were required to compute the generation time\todo{what is this?
+define?} for measles.
+We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from %# http://www.sciencedirect.com/science/article/pii/S0022519311003146
+We then estimated Rt using the methods adapted from [REF] from the incidence data for each outbreak.
+We defined outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
+The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.\todo{this repeats a bit of the previous paragraph - could do with rewrite}
 
@@
      \includegraphics[width=0.8\textwidth]{averageR0.pdf}
-     \caption{Estimates of R0 for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \caption{Estimates of Rv (R0) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \label{fig:r0}
 \end{figure}
 
-...The report also concludes that increased immunisation uptake during the 2011-2012 outbreak, especially in the Auckland region, likely contributed to minimise the number of cases.
-[COMMENT?
-Unsure, though don't want to be dismissive].
-....
+The estimated Rv for each outbreak is shown in Figure~\ref{fig:r0}.
+The probability density of the Rv estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an Rv well above one and thus we may expect this outbreak to persist if conditions remain the same.\todo{to be fair, this is because it's ongoing - as we haven't seen a decline yet, $R_0$ is necessarily over one.}
 
-\section{Future analyses}
+These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in section~\ref{sec:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+The measles outbreak in 2011--2012 had an Rv of just greater than one, and yet it persisted for over 12 months.
+This implies that the current outbreak may persist within the population for a substantial period, given it's Rv is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other Rv estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+However, sub-clinical and underreporting may lower the estimate.
+The relative contributions of both to our estimates are currently unknown.
 
-\section{Future cost-benefit analyses}
-Using the results above we aim to:
+\subsection{Summary of modeling}
+
 \begin{itemize}
-\item Update previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sec:risk_analyses}).
+\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
+ \item The reproduction number for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to become endemic.\todo{this seems a bit extreme -- caveat?}
+\end{itemize}
+
+\section{Future modeling}
+Future modeling we aim to perform are:
+\begin{itemize}
+\item An update of previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
 \item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
@@
 
+\section {Summary of key findings}
+\begin{itemize}
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
+\item The cost of the current measles outbreak is estimated to be at least \$750,000.
+\item Analyses of outbreak data suggest that measles Rv values often include 1 and in this year, 2014, are well above one.
+This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
+\end{itemize}
+
 \section{Acknowledgments}
-The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR for help in obtaining the appropriate materials for analyses.
+The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR, and June Atkinson from University of Otago for help in obtaining the appropriate materials for analyses.
 

### 2014-07-03 interimreport1.Rnw
Bibliography inserted and citations included
Not more to say...

@@
 Previous analyses of measles, including two in New Zealand by Prof.
-Roberts (INSERT REFS), have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
+Roberts \cite{roberts0,roberts4}, have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
@@
 
-Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles (Clements and Hussey, 2004).
+Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \cite{clements4}.
 Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities.
-Wolfson et al.
-(2007) estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
+\cite{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
 Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
 They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination.
-The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 (WHO, 2013).
+The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \cite{who13}.
 In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.
 
-The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million (Carabin et al., 2003).
-The estimated cost for a case ranged from US\$ 189--344 (Carabin et al., 2003); however, the average estimated cost of a typical hospital case ranges from US\$ 967--1,755 (Carabin et al., 2002).
-Stack et al (2011) estimated the economic benefits from cases averted due to measles vaccination.
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million \cite{carabin3].
+The estimated cost for a case ranged from US\$ 189--344 \cite{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$ 967--1,755 \cite{carabin2}.
+\cite{stack11} estimated the economic benefits from cases averted due to measles vaccination.
 They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$ 10 billion of costs averted between 2011 and 2020.
@@
 
-Italy has the highest reported annual cost of measles among industrialised countries (Carabin et al., 2003).
+Italy has the highest reported annual cost of measles among industrialised countries \cite{carabin3}.
 In 2001, it reported losses related to measles of approximately US\$ 50 million.
@@
 
-An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases (Flego et al., 2013).
+An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases \cite{flego13}.
 Seven (27\%) of the cases required hospitalisation for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
@@
 Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years.
-In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 (Wichmann et al., 2009).
+In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \cite{wichmann9}.
 However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children (Siedler et al, 2006).
@@
 
-Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world (Parker et al., 2006).
+Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world \cite{parker6}.
 Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.
-Ortega-Sanchez et al.
-(2014) estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
+\cite{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
 The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$ 2.7 and 5.3 million.
 Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$ 298 per contact.
-Dayan et al.
-(2005) calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
+\cite{dayan5} calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
 They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination.
@@
 
-Parker et al.
-(2006) reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
+\cite{parker6} reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
 A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak.
@@
 
-Coleman et al.
-(2012) estimated the direct medical and public health costs in response to a single case of refugee-imported measles.
+\cite{coleman12} estimated the direct medical and public health costs in response to a single case of refugee-imported measles.
 Costs included labor, translation and benefits for public health workers.
@@
 Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
-Direct costs can be divided into medical and non-medical (Saha and Gerdtham, 2013).
+Direct costs can be divided into medical and non-medical \cite{saha13}.
 Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
@@
 They estimated direct costs per case to be less than US\$ 500.
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively (Prouty and Lakin, 2000).
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty0}.
 In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person.
@@
 The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalisation) to \$ 29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) (Zhou et al., 2004).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \cite{zhou4}.
 
@@
 
-A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years.
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \cite{roberts0,roberts4,tobias98}.
 The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
 
-The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 (Tobias & Roberts 1998) successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
-et al 1998; Roberts & Tobias 2000).
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \cite{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \cite{mansoor98,roberts0}.
 Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
-The model developed by Roberts & Tobias (2000) supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
-The schedule was changed in 2000 with MMR2 now being administered before 5 years (Anon.
-2002a) and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics.
+The model developed by \cite{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years \cite{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics.
 
-These results were in line with those obtained by other authors, for example: Babad et al (1995) advocated a two-dose schedule for England and Wales, with the second vaccination given at age four; and Gay et al (1998) recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
-In addition, Agur et al (1993) found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
-These authors all agree that two
-vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-A different approach was taken by Wallinga et al (2001).
-These authors took existing policies in eight European countries and estimated the coverage rates required to reduce Rv below one.
-They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately
-87\%.
+These results were in line with those obtained by other authors, for example: \cite{babad95} advocated a two-dose schedule for England and Wales, with the second vaccination given at age four; and \cite{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, \cite{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+These authors all agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+A different approach was taken by \cite{wallinga1} who took existing policies in eight European countries and estimated the coverage rates required to reduce Rv below one.
+They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
 
-A variety of similar models for measles vaccination strategies have been developed by other authors for various regions (Agur et al 1993; Babad et al 1995; Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001) based on sets of nonlinear differential equations (ODE), and the conclusions reached have been similar.
+A variety of similar models for measles vaccination strategies have been developed by other authors for various regions \cite{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equations (ODE), and the conclusions reached have been similar.
 The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
@@
 It is difficult to estimate the proportion of the school-age population that have been effectively immunised, but this is continuously being diluted by children who are not immunised.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, Glass et al (2004) in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \cite{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection.
 
 The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$\todo{not sure if you want it like this throughout?}.
-This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population (Anderson & May 1991; Diekmann & Heesterbeek 2000).
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \cite{anderson91, diekmann0}.
 If $R_0 > 1$ an epidemic will occur following an introduction of infection.
-The best estimate we had for measles in New Zealand was R0 = 12.8 -- 12.5\todo{this seems an odd range?} (Roberts, see results section).
+The best estimate we had for measles in New Zealand was R0 = 12.8 \cite{roberts4}.
 The basic reproduction number of the infection under vaccination, Rv\todo{or $R_v$ ?}, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
@@
 To understand the level of immunity in the population and how likely an outbreak was of becoming endemic, we estimated Rv from all the outbreaks since 2009.
-To do this we estimated Rt, following an adaptation of the methods in [REF]\todo{there's 2 references in \texttt{?est.R0.TD}}, and used these to estimate the Rv of measles in the recent outbreaks, which helps understand the transmission dynamics of measles in the partially immune population.
+To do this we estimated Rt, following an adaptation of the methods in \cite{obidia12}\todo{there's 2 references in \texttt{?est.R0.TD}}, and used these to estimate the Rv of measles in the recent outbreaks, which helps understand the transmission dynamics of measles in the partially immune population.
 
@@
 define?} for measles.
-We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from %# http://www.sciencedirect.com/science/article/pii/S0022519311003146
-We then estimated Rt using the methods adapted from [REF] from the incidence data for each outbreak.
+We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \cite{klinkenberg11}.
+We then estimated Rt using the methods adapted from \cite{obidia12} from the incidence data for each outbreak.
 We defined outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
@@
 
-\bibitem[Gelman et al.(1996)Gelman, Roberts, and Gilks]{grg}
-Gelman, A., G.~O. Roberts, and W.~R. Gilks (1996).
-\newblock Efficient Metropolis jumping rules.
-\newblock In \emph{Bayesian Statistics, 5 (Alicante, 1994)}, pp.~599--607.
-  Oxford University Press.
+\bibitem[Agur et al.(1993)]{agur93}
+Agur, Z., L. Cojocaru, G. Mazor, R.~M. Anderson and Y.~L. Danon (1993).
+\newblock Pulse mass measles vaccination across age cohorts.
+\newblock \emph{Proceedings of the National Academy of Sciences USA}, 90, 11698--11702.
 
-\bibitem[Geyer(1992)]{practical}
-Geyer, C.~J. (1992).
-\newblock Practical Markov chain Monte Carlo (with discussion).
-\newblock \emph{Statistical Science}, 7, 473--511.
+\bibitem[Anderson and May(1991)]{anderson91}
+Anderson, R.~M. and R.~M. May (1991).
+\newblock \emph{Infectious diseases of humans: dynamics and control}. Oxford: Oxford University Press.
 
-\bibitem[Geyer and Thompson(1995)]{geyer-temp}
-Geyer, C.~J. and E.~A. Thompson (1995).
-\newblock Annealing Markov chain Monte Carlo with applications to
-    ancestral inference.
-\newblock \emph{Journal of the American Statistical Association}, 90, 909--920.
+\bibitem[Anon. (2002a)]{anon2a}
+Anon. (2002a).
+\newblock \emph{Immunisation handbook}
+\newblock Wellington: Ministry of Health. pp.~131--146.
+
+\bibitem[Anon. (2002b)]{anon2b}
+Anon. (2002b).
+\newblock \emph{Infectious diseases in livestock}
+\newblock The Royal Society. pp.~68.
+
+\bibitem[Babad et al.(1995)]{babad95}
+Babad, H.~R., D.~J. Nokes, N.~J. Gay, E. Miller, P. Morgan-Capner, and R.~M. Anderson (1995).
+\newblock Predicting the impact of measles vaccination in England and Wales: model validation and analysis of policy options.
+\newblock \emph{Epidemiology and Infection}, 114, 319--344.
+
+\bibitem[Bae et al. (2013)]{bae13}
+Bae, G.~R, Y.~J. Choe, U.~Y. Go, Y.~I. Kim, and J.~K. Lee (2013). 
+\newblock Economic analysis of measles elimination program in the Republic of Korea, 2001: A cost benefit analysis study.
+\newblock \emph {Vaccine}, 31, 2661--2666.
+
+\bibitem[Carabin et al. (2002)]{carabin2}
+Carabin, H., W.~J. Edmunds, U. Kou, S. van den Hof, and V.~H. Nguyen (2002). 
+\newblock Measles in industrialized countries: a review of the average costs of adverse events and measles cases.
+\newblock \emph{BMC Public Health}, 2, 22.
+
+\bibitem[Carabin et al. (2003)]{carabin3}
+Carabin, H., W.~J. Edmunds, M. Gyldmark, P. Beutels, D. Lévy-Bruhl, H. Salo, U.~K. and Griffiths (2003).
+\newblock The cost of measles in industrialised countries.
+\newblock \emph{Vaccine}, 21,4167--4177.
+
+\bibitem[Clements and Hussey (2004]{clements4}
+Clements, C.~J. and G.~D. Hussey (2004).
+\newblock Chapter 4: Measles.
+\newblock In \emph{The Global Epidemiology of Infectious Diseases},  Murray, C., A.~D. Lopez, and C.~D. Mathers, (eds.), Geneva.
+  World Health Organization, pp.~391.
+
+\bibitem[Coleman et al.(2012)]{coleman12}
+Coleman, M.~S., L. Garbat-Welch, H. Burke, M. Weinberg, K. Humbaugh, A. Tindall, and J. Cambron (2012).
+\newblock Direct costs of a single case of refugee-imported measles in Kentucky.
+\newblock \emph{Vaccine}, 30,317--321.
+
+\bibitem[Dayan et al.(2005)]{dayan5}
+Dayan, G.~H, I.~R. Ortega-Sánchez, C.~W. LeBaron, M.~P. Quinlisk, and the Iowa Measles Response Team (2005).
+\newblock The cost of containing one case of measles: the economic impact on the public health infrastructure - Iowa, 2004.
+\newblock \emph{Pediatrics), 116:e1; DOI:10/1542/peds.2004-2512.
+
+\bibitem[Diekmann et al.(2000)]{diekmann0}
+Diekmann, O. &  J.~A.~P. Heesterbeek (2000).
+\newblock \emph{Mathematical epidemiology of infectious diseases: model building, analysis and interpretation}.
+Chichester: Wiley.
+
+\bibitem[Edmunds et al.(2000)]{edmunds0}
+Edmunds, W.~J., N.~J. Gay, M. Kretzschmar, R.~G. Pebody and H. Wachman (2000).
+\newblock \The pre-vaccination epidemiology of measles, mumps and rubella in Europe: implications for modelling studies.
+\newblock \emph{Epidemiology and Infection}, 125, 635--650.
+
+\bibitem[Filia et al.(2007)]{filia7}
+Filia, A., A. Brenna, A. Paná, G.~M. Cavallaro, M. Massari, and M. Atti, (2007).
+\newblock Health burden and economic impact of measles-related hospitalization in Italy, 2002-2003.
+\newblock \emph{BMC Public Health}, 7,169
+
+\bibitem[Flego et al.(2013)]{flego13}
+Flego, K.~L., D.~A. Belshaw, V. Sheppeard, and K.~M. Weston (2013).
+\newblock Impacts of a measles outbreak in western Sydney on public health resources.
+\newblock \emph{Communicable Diseases Intelligence Quarterly Report}, 37, E240--245.
+
+\bibitem[Geyer and Thompson(1995)]{gay98}
+Gay, N.~J., L. Pelletier, and P. Duclos (1998).
+\newblock Modelling the incidence of measles in Canada: an assessment of the options for vaccination policy.
+\newblock \emph{Vaccine}, 16, 794--801.
+
+\bibitem[Glass et al.(2004)]{glass4}
+Glass, K., J. Kappey, and B.~T. Grenfell (2004).
+\newblock The effect of heterogeneity in measles vaccination population immunity.
+\newblock \emph{Epidemiology and Infection}, 132, 675--683.
+
+\bibitem[Klinkenberg et al.(2011)]{klinkenberg11}
+Klinkenberg, D. and H. Nishiuraa (2011).
+\newblock The correlation between infectivity and incubation period of measles, estimated from households with two cases.
+\newblock \emph{Journal of Theoretical Biology},284, 52--60
+
+\bibitem[Koopmanschap(1998)][koopmanschap98}
+Koopmanschap, M.~A. (1998).
+\newblock Cost-of-illness studies: useful for health policy?
+\newblock \emph{Pharmacoeonomics}, 14, 143--148.
+
+\bibitem[Larg and Moss(2011)]{larg11}
+Larg, A. and J.~R. Moss (2011).
+\newblock Cost-of-illness studies: a guide to critical evaluation.
+\newblock \emph{Pharmacoeconomics}, 29,653--671.
+
+\bibitem[Mansoor et al.(1998)]{mansoor98}
+Mansoor, O., A. Blakely, M. Baker, M. Tobias, and A. Bloomfield (1998).
+\newblock A measles epidemic controlled by immunisation. 
+\newblock \emph{New Zealand Medical Journal}, 111, 467--471.
+
+\bibitem[Ortega-Sanchez et al.(2014)]{ortegasanchez14}
+Ortega-Sanchez, I.~R., M. Vijayaraghavan, A.~E. Barskey, and G.~S. Wallace (2014).
+\newblock The economic burden of sixteen measles outbreaks on United States public health departments in 2011.
+\newblock \emph{Vaccine}, 32,1311--1317.
+
+\bibitem[Obidia et al.(2012)]{obidia12}
+Obadia, T., R. Haneef and P.~Y., Boëlle (2012).
+\newblock The R0 package: a toolbox to estimate reproduction numbers for epidemic outbreaks.
+\newblock \emph{BMC Medical Informatics and Decision Making}, 2012, 12--147.
+
+\bibitem[Parker et al.(2006)]{parker6}
+Parker, A.~A., W. Staggs, G.~H. Dayan, T.~R. Ortega-Sánchez, P.~A. Rota, L. Lowe, P. Boardman, R. Teclaw, C. Graves, and C.~W. LeBaron (2006).
+\newblock Implications of a 2005 measles outbreak in Indiana for sustained elimination of measles in the United States.
+\newblock \emph{The New England Journal of Medicine}, 355, 447--455.
+
+\bibitem[Prouty et al. (2001)]{prouty1}
+Prouty, R.W., G. Smith and K.~C. Lakin (2001).
+\newblock Residential services for persons with developmental disabilities: status and trends through 2000.
+\newblock \emph{Minneapolis: Institute on Community Integration}, University of Minnesota, pp.~179, rtc.umn.edu/risp00.
+
+\bibitem[Roberts(2004)]{roberts4}
+Roberts, M. (2004).
+\newblock A mathematical model for measles vaccination.
+\newblock Wellington: Ministry of Health.
+
+\bibitem[Roberts and Tobias(2000)]{roberts0}
+Roberts, M.~G. and M.~I. Tobias (2000).
+\newblock Predicting and preventing measles epidemics in New Zealand: Application of a mathematical model. 
+\newblock \emph{Epidemiology and Infection}, 124, 279--287.
+
+\bibitem[Saha and Gerdtham(2013)]{saha13}
+Saha, S. and U.~G. Gerdtham (2013).
+\newblock Cost of illness studies on reproductive, maternal, newborn, and child health: a systematic literature review.
+\newblock \emph{Health Economics Review}, doi:10.1186/2191-1991-3-24.
+
+\bibitem[Siedler et al.(2006)]{siedler6}
+Siedler, A., A. Tischer, A. Mankertz, and S. Santibanez (2006).
+\newblock Two outbreaks of measles in Germany 2005.
+\newblock \emph{Eurosurveillance} 2006:11(4) article 5, http://www.eurosurveillance.org/ViewArticle.aspx?ArticleId=615, accessed 14 June 2014.
+
+\bibitem[Stack et al.(2011)]{stack11}
+Stack, M.~L., S. Ozawa, D.~M. Bishai, A. Mirelman, Y. Tam, L. Niessen, D.~G. Walker, and O.S. Levine (2011).
+\newblock Estimated economic benefits during the 'decade of vaccine' include treatment savings, gains in labor productivity.
+\newblock \emph{Health Affairs}, 30,1021--1028.
+
+\bibitem[Statistics New Zealand)]{stats14}
+\newblock \emph{Statistics New Zealand} (2014).
+http://nzdotstat.stats.govt.nz/, accessed 17 June 2014.
+
+\bibitem[Tobias and Roberts(1998)]{tobias98}
+Tobias, M.~I. and M.~G. Roberts (1998).
+\newblock Predicting and preventing measles epidemics in New Zealand: Application of a mathematical model.
+\newblock Wellington: Ministry of Health.
+
+\bibitem[Wallinga et al.(2001)]{wallinga1}
+Wallinga, J., D. Levy-Bruhl, N.~J. Gay, and C.~H. Wachman (2001).
+\newblock Estimation of measles reproduction ratios and prospects for elimination of measles by vaccination in some Western European countries.
+\newblock \emph{Epidemiology and Infection}, 127, 281--295.
+
+\bibitem[Wichmann et al.(2009)]{wichmann9}
+Wichmann, O., A. Siedler, D. Sagebiel, W. Hellenbrand, S. Santibanez, A. Mankertz, G. Vogt, U. van Treeck, and G. Krause (2009).
+\newblock Further efforts needed to achieve measles elimination in Germany: results of an outbreak investigation.
+\newblock \emph{Bulletin of the World Health Organization}, 87, 108--115.
+
+\bibitem[Wolfson et al(2007)]{wolfson7}
+Wolfson, L.~J., P.~M. Strebel, M. Gacic-Dobo, E.~J. Hoekstra, J.~W. McFarland, and B.~S. Hersh (2007).
+\newblock Has the 2005 measles mortality reduction goal been achieved? A natural history modelling study.
+\newblock \emph{Lancet}, 369, 191--200.
+
+\bibitem[WHO (2013)]{who13}
+World Health Organisation measles media centre, January (2013)
+\newblock Geneva: World Health Organization.
+http://www.who.int/mediacentre/news/notes/2013/measles_20130117/en/, accessed July 1, 2014.
+
+\bibitem[Zhou et al.(2004)]{zhou4}
+Zhou, F, S. Reef, M. Massoudi, M.~J. Papania, H.~R. Yusuf, B. Bardenheier, L. Zimmerman, and M.~M. McCauley (2004).
+\newblock An economic analysis of the current universal 2-dose measles-mumps-rubella vaccination program in the United States.
+\newblock \emph{Journal of Infectious Diseases}, 189, S131--45.
 

### 2014-07-04 interimreport1.Rnw
Draft final interim report
Bibliography complete, compiles, etc.

@@
 \usepackage{bibunits}
-\usepackage{natbib}
+%\usepackage{natbib}
 \usepackage{graphics}
@@
 \usepackage{graphicx}
+\usepackage{eurosym}
 \usepackage{todonotes}
@@
 
-\title{Interim report\\ Measles risk assessment, modeling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
+\title{Interim report\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
 \author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
@@
 In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
-We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modeling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more.
+We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modelling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more.
 Our key findings were:
@@
 \item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
-\item Analyses of outbreak data suggest that measles basic reproductive number (Rv, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
 This analyses suggest improved vaccination is a requisite to prevent measles becoming endemic again.
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modeling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
 
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our preliminary modeling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modeling}).
+This is particularly in light of our preliminary modelling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modelling}).
 
@@
 \begin{itemize}
-\item Multivariate modelling to account for confounding within the univariate analyses.\todo{one L or two in modelling?}
+\item Multivariate modelling to account for confounding within the univariate analyses.
 \item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
@@
 For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).\todo{this sentence seems to need a bit more after it?
-Something like "to produce pretty maps"}
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Note this uses all immigration of foreign nationals, coming for whatever purpose.
+We used the WHO data to determine per capita measles cases for each year and used this and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
 
@@
 # output top countries
+colnames(dpdf)[1]<-c("country")
 topimmigration <- dpdf[order(-dpdf$immigration),] 
@@
 \label{table:risk12}
-\end{table}\todo{these tables need country rather than id as the first column name}
+\end{table}
 
@@
 
-\subsection{Future risk analyses}\sec{risk_analyses}
+\subsection{Future risk analyses}\label{sec:risk_analyses}
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
@@
 \item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using a broader range of years, including modeling the trend in importations.
+\item Update of the importation risk analyses using a broader range of years, including modelling the trend in importations.
 \end{itemize}
@@
 \item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel. \todo{added a bit here you may want to check}
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
 \end{itemize}
@@
 The number of cases by age group is shown in Table~\ref{table:freq}.
-Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 0.47 and 0.33, respectively.\todo{make these percentages?}
+Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.
 
@@
 Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis.
-The cost was found to decrease by approximately \$5 per year of case age.\todo{later on we could assess whether or not a linear term for case age is appropriate, or what if any interaction there is between age and length of stay?}
+The cost was found to decrease by approximately \$5 per year of case age.
 
@@
 private clinics.
+Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
 
@@
 They estimated direct costs per case to be less than US\$ 500.
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty0}.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty1}.
 In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person.
@@
 \begin{itemize}
-\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modeling.
+\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modelling.
 \end{itemize}
@@
 
-\section{Modeling measles epidemics}\label{sec:epidemic_modeling}
+\section{Modelling measles epidemics}\label{sec:epidemic_modelling}
 
@@
 
-The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$\todo{not sure if you want it like this throughout?}.
+The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
 This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \cite{anderson91, diekmann0}.
 If $R_0 > 1$ an epidemic will occur following an introduction of infection.
-The best estimate we had for measles in New Zealand was R0 = 12.8 \cite{roberts4}.
-The basic reproduction number of the infection under vaccination, Rv\todo{or $R_v$ ?}, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
+The best estimate we had for measles in New Zealand was $R_0$ = 12.8 \cite{roberts4}.
+The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
 If $R_v < 1$ epidemics are prevented.
@@
 
-To understand the level of immunity in the population and how likely an outbreak was of becoming endemic, we estimated Rv from all the outbreaks since 2009.
-To do this we estimated Rt, following an adaptation of the methods in \cite{obidia12}\todo{there's 2 references in \texttt{?est.R0.TD}}, and used these to estimate the Rv of measles in the recent outbreaks, which helps understand the transmission dynamics of measles in the partially immune population.
-
-To estimate Rt we were required to compute the generation time\todo{what is this?
-define?} for measles.
+To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks since 2009.
+To do this we estimated $R_t$, following an adaptation of the methods in \cite{obidia12,wallinga4}.
+We were required to compute the generation time for measles to do so.
+The generation time is the average time an index case infects others after becoming infected.
 We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \cite{klinkenberg11}.
-We then estimated Rt using the methods adapted from \cite{obidia12} from the incidence data for each outbreak.
-We defined outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
-The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.\todo{this repeats a bit of the previous paragraph - could do with rewrite}
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.
 
@@
      \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
-     \caption{Measles data classified as outbreaks for reproductive number of the infection (Rv) estimation}
+     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
      \label{fig:outbreaks}
@@
      \includegraphics[width=0.8\textwidth]{averageR0.pdf}
-     \caption{Estimates of Rv (R0) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
      \label{fig:r0}
@@
 
-The estimated Rv for each outbreak is shown in Figure~\ref{fig:r0}.
-The probability density of the Rv estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak, which has an Rv well above one and thus we may expect this outbreak to persist if conditions remain the same.\todo{to be fair, this is because it's ongoing - as we haven't seen a decline yet, $R_0$ is necessarily over one.}
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The probability density of the $R_v$ estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.An important caveat to this analysis is that because this is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one.
 
@@
 Given the risk of importation of measles as highlighted in section~\ref{sec:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
-The measles outbreak in 2011--2012 had an Rv of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given it's Rv is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other Rv estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
 However, sub-clinical and underreporting may lower the estimate.
@@
 
-\subsection{Summary of modeling}
+\subsection{Summary of modelling}
 
@@
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to become endemic.\todo{this seems a bit extreme -- caveat?}
+ \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
 \end{itemize}
 
-\section{Future modeling}
-Future modeling we aim to perform are:
+\section{Future modelling}
+Future modelling we aim to perform are:
 \begin{itemize}
@@
 \item The cost of the current measles outbreak is estimated to be at least \$750,000.
-\item Analyses of outbreak data suggest that measles Rv values often include 1 and in this year, 2014, are well above one.
+\item Analyses of outbreak data suggest that measles $R_v$ values often include 1 and in this year, 2014, are well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
@@
 \bibitem[Carabin et al. (2003)]{carabin3}
-Carabin, H., W.~J. Edmunds, M. Gyldmark, P. Beutels, D. Lévy-Bruhl, H. Salo, U.~K. and Griffiths (2003).
+NA
 \newblock The cost of measles in industrialised countries.
@@
 \bibitem[Dayan et al.(2005)]{dayan5}
-Dayan, G.~H, I.~R. Ortega-Sánchez, C.~W. LeBaron, M.~P. Quinlisk, and the Iowa Measles Response Team (2005).
+NA
 \newblock The cost of containing one case of measles: the economic impact on the public health infrastructure - Iowa, 2004.
-\newblock \emph{Pediatrics), 116:e1; DOI:10/1542/peds.2004-2512.
+\newblock \emph{Pediatrics}, 116:e1; DOI:10/1542/peds.2004-2512.
 
 \bibitem[Diekmann et al.(2000)]{diekmann0}
-Diekmann, O. &  J.~A.~P. Heesterbeek (2000).
+Diekmann, O. and  J.~A.~P. Heesterbeek (2000).
 \newblock \emph{Mathematical epidemiology of infectious diseases: model building, analysis and interpretation}.
@@
 Edmunds, W.~J., N.~J. Gay, M. Kretzschmar, R.~G. Pebody and H. Wachman (2000).
-\newblock \The pre-vaccination epidemiology of measles, mumps and rubella in Europe: implications for modelling studies.
+\newblock The pre-vaccination epidemiology of measles, mumps and rubella in Europe: implications for modelling studies.
 \newblock \emph{Epidemiology and Infection}, 125, 635--650.
@@
 \bibitem[Filia et al.(2007)]{filia7}
-Filia, A., A. Brenna, A. Paná, G.~M. Cavallaro, M. Massari, and M. Atti, (2007).
+NA
 \newblock Health burden and economic impact of measles-related hospitalization in Italy, 2002-2003.
@@
 
-\bibitem[Geyer and Thompson(1995)]{gay98}
+\bibitem[Gay et al.(1998)]{gay98}
 Gay, N.~J., L. Pelletier, and P. Duclos (1998).
@@
 
-\bibitem[Koopmanschap(1998)][koopmanschap98}
+\bibitem[Koopmanschap(1998)]{koopmanschap98}
 Koopmanschap, M.~A. (1998).
@@
 \bibitem[Obidia et al.(2012)]{obidia12}
-Obadia, T., R. Haneef and P.~Y., Boëlle (2012).
+NA
 \newblock The R0 package: a toolbox to estimate reproduction numbers for epidemic outbreaks.
@@
 \bibitem[Parker et al.(2006)]{parker6}
-Parker, A.~A., W. Staggs, G.~H. Dayan, T.~R. Ortega-Sánchez, P.~A. Rota, L. Lowe, P. Boardman, R. Teclaw, C. Graves, and C.~W. LeBaron (2006).
+NA
 \newblock Implications of a 2005 measles outbreak in Indiana for sustained elimination of measles in the United States.
@@
 
+\bibitem[Wallinga and Teunis(2004)]{wallinga4}
+Wallinga, J., and P. Teunis (2004).
+\newblock Different Epidemic Curves for Severe Acute Respiratory Syndrome Reveal Similar Impacts of Control Measures.
+\newblock \emph{American Journal of Epidemiology}, 160, 509.
+
 \bibitem[Wichmann et al.(2009)]{wichmann9}
@@
 \newblock Geneva: World Health Organization.
-http://www.who.int/mediacentre/news/notes/2013/measles_20130117/en/, accessed July 1, 2014.
+\href{http://www.who.int/mediacentre/news/notes/2013/measles_20130117/en/}{www.who.int}, accessed July 1, 2014.
 

### 2014-07-04 interimreport1.tex
Draft final interim report
Bibliography complete, compiles, etc.

@@
 \usepackage{graphicx}
+\usepackage{eurosym}
+\usepackage{todonotes}
 
@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we reivewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
-We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013-2014 outbreak and began modeling measles outbreaks to understand how they may be controled and the possibility that measles may become endemic.
+In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
+We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modelling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more.
 Our key findings were:
@@
 \item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
-\item New Zealand is at risk of frequent measles importation due to travel and endemic measles eslewhere in the globe.
-\item The cost of the current 2013-2014 measles outbreak is estimated to be at least \$$\frac{3}{4}$ million.
-\item Analyses of outbreak data suggest that measles basic reproductive number (Rv, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
-This analyses suggest improved vaccination is a requesite to prevent measles becoming endemic again.
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
+\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
+This analyses suggest improved vaccination is a requisite to prevent measles becoming endemic again.
 \end{itemize}
@@
 
-As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committted to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
+As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
 The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.
@@
 The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011-2012 (mainly in Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
-The outbreak in 2011-2012 lasted for more than 12 months and the current 2013-2014 outbreak started at the end of December 2013 (as on \date{\today}).
-In 2013, prior to the 2013-2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) it can request verification of non-endemic status three years after the last case of the 2011-12 outbreak in June 2012.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
+The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{\today}).
+In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
 
 Previous analyses of measles, including two in New Zealand by Prof.
-Roberts (INSERT REFS), have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
+Roberts \cite{roberts0,roberts4}, have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are perhaps likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 
-Overall, we thought the review was very thorough.
+Overall, the review was very thorough.
 The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
@@
 
-Specific epidemiological details were provided for the 2011-2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6-8}).
-Similar presentation of the case data are provided for ethnicity (\emph{Figures 9-10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11-13}).
+Specific epidemiological details were provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}).
+Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}).
 Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section X) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
 
@@
 Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
-However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approching or over 95\% levels for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our preliminary modeling and risk analyses results (sections X and Y).
+However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
+This is particularly in light of our preliminary modelling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modelling}).
 
@@
 
-\subsection{Additional risk analyses}
+\subsection{Additional risk analyses}\label{sec:risk_analyses}
 
-In this section we provide a brief description of the work we are performing that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
+In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
 These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
@@
 
-In light of the apparent increasing trend in measles incidence in the last few years Figure~\ref{fig:incidence1997}, we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
 To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
@@
 For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Note this uses all immigration of foreign nationals, coming for whatever purpose.
+We used the WHO data to determine per capita measles cases for each year and used this and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
 
@@
 \hline\hline
-\multicolumn{1}{c}{id}&\multicolumn{1}{c}{immigration}\tabularnewline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
 \hline
@@
 \hline\hline
-\multicolumn{1}{c}{id}&\multicolumn{1}{c}{cover}\tabularnewline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
 \hline
@@
 \hline\hline
-\multicolumn{1}{c}{id}&\multicolumn{1}{c}{incidence}\tabularnewline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
 \hline
@@
 \hline\hline
-\multicolumn{1}{c}{id}&\multicolumn{1}{c}{risk}\tabularnewline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
 \hline
@@
 
-The preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
+Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
 However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:incidence12} and Table \ref{table:cover12}).
-Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importantion to New Zealand.
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
 These data are shown in Table \ref{table:risk12}.
@@
 Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
-This suggests that the risk of measles important could increase, though further analyses are require to understand these interatction between these.
+This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
 Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).
-This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin or travel to will be an important factor in determining the risk of measles importation.
+This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
 
@@
 \item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
-\item There may be seasonal changes n risk of measles importation, though further analyses are needed.
+\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
 \end{itemize}
 
-\subsection{Future risk analyses}
+\subsection{Future risk analyses}\label{sec:risk_analyses}
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of indenpendence among risk factors.
-Specifically for the multivariate analyses we wish to perform that detect interactions we require \emph{Age * Prioritized Ethnicity * NZDep} data for New Zealand to see if the interactions among case covariates give us any additional information on risk from the univariate analyses already performed in the report reviewed above.
-These data only became available to us on 3 July 2014 after because ESR and NZ Statistics did not have those data and we had to obtain these from the University of Otago.
+Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
+Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above.
+These data only became available to us on 3 July 2014, provided by the University of Otago.
 
-The following are analyses are still in progress and aim to include them in later reports:
+The following analyses are in progress, for inclusion in later reports:
 \begin{itemize}
-\item Multivariate regression analyses to look for interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using more than the 2012 data, including modeling the trend in importations.
+\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
+\item Update of the importation risk analyses using a broader range of years, including modelling the trend in importations.
 \end{itemize}
-Additional data we believe would enable us and the Ministry to understand measles risk better is fine scale (lower than District Health Board (DHB)) immunization coverage data.
+Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
 Thus the data gap that we have that will hinder us providing fine scale risk maps are:
 \begin{itemize}
-\item A lack of meshblock level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
-\item The number of New Zealanders arriving from abroad each year.
+\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
 \end{itemize}
@@
 We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
-\emph{meshblock}) level would allow better understanding of risk of measles infection and resource allocation because they may allow targetted immunisation programmes.
+meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
 
@@
 
-In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with this current measles outbreak.
+In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current measles outbreak.
 
-Approximately 50 years ago, approximately 135 million cases and 7-8 million deaths were believed to occur in the world (Clements and Hussey, 2004).
-Thirty years later, it was estimated there were still approximately 45 million cases of measles offering annual, including 6 million measles-related fatalities.
-Wolfson et al.
-(2007) estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
+Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \cite{clements4}.
+Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities.
+\cite{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
 Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
-They estimated approximately 7.5 million deaths from measles were avoided from 2000-05 due vaccination.
-The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 (WHO, 2013).
+They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination.
+The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \cite{who13}.
 In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.
 
-The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million (Carabin et al., 2003).
-The estimated cost for a case ranged from US\$ 189-344 (Carabin et al., 2003); however, the average estimated cost of a typical hospital case ranges from US\$ 967-1,755 (Carabin et al., 2002).
-Stack et al (2011) estimated the economic benefits from cases averted due to measles vaccination.
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million \cite{carabin3].
+The estimated cost for a case ranged from US\$ 189--344 \cite{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$ 967--1,755 \cite{carabin2}.
+\cite{stack11} estimated the economic benefits from cases averted due to measles vaccination.
 They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$ 10 billion of costs averted between 2011 and 2020.
@@
 
-Italy has the highest reported annual cost of measles among industrialised countries (Carabin et al., 2003).
+Italy has the highest reported annual cost of measles among industrialised countries \cite{carabin3}.
 In 2001, it reported losses related to measles of approximately US\$ 50 million.
-The economic impact of a large measles outbreak in Italy, 2002-03 examined the costs associated with 5,154 hospitalizations where measles was the main discharge diagnosis.
+The economic impact of a large measles outbreak in Italy, 2002--03 examined the costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis.
 The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
-The total cost of these hospitalizations amounted to \euro 8.83 million (\euro 1 ~ NZ\$ 2.0 in 2002-03), or approximately \euro 1,700 per case.
-The average cost per non-complicated measles was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
+The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 ~ NZ\$ 2.0 in 2002-03), or approximately \euro 1,700 per case.
+The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
 The average daily cost of a hospital stay was  \euro 327.
 
-An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases (Flego et al., 2013).
-Seven (27\%) of the cases required hospitalization for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
+An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases \cite{flego13}.
+Seven (27\%) of the cases required hospitalisation for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
 During this outbreak, a total of 1,395 contacts were identified and managed by a public health unit in western Sydney.
@@
 Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years.
-In 2001 more than 6000 cases were reported in Germany but by 2004 this number fell to 122 (Wichmann et al., 2009).
+In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \cite{wichmann9}.
 However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children (Siedler et al, 2006).
@@
 In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case.
-Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalized.
-The mean costs of the hospitalized patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623.
+Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised.
+The mean costs of the hospitalised patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623.
 In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing.
@@
 
-Coleman et al.
-(2012) estimated the direct medical and public health costs in response to a single case of refugee-imported case measles.
-Costs included labor, translation, benefits for public health workers.
-In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalization, transportation and diagnosis.
+\cite{coleman12} estimated the direct medical and public health costs in response to a single case of refugee-imported measles.
+Costs included labor, translation and benefits for public health workers.
+In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis.
 In total, 387 hours were associated with this single case, resulting in a cost of US\$ 11,881.
 In addition, per-contact costs amounted to US\$ 264.
-The cost of hospitalization for the 3-day stay by the index case was US\$ 931.
+The cost of hospitalisation for the 3-day stay by the index case was US\$ 931.
 Additional costs were associated with physician visits (US\$ 294), vaccine and immunoglobulin (US\$ 1,765), mileage (US\$ 205) and immunologic screening tests for the parents' exposed to measles (US\$ 240) for a total of US\$ 23,816.
+
 Economic analyses of measles control programs have shown them to be financially effective.
 In the Republic of Korea, the economics of alternative measles vaccination programs were compared.
-All of the alternatives were found to be economically efficient (benefit/cost ration (B/C) > 1.0), with the alternative using two doses of the MMR program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27).
+All of the alternatives were found to be economically efficient (benefit/cost ratio (B/C) > 1.0), with the alternative using two doses of the MMR program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27).
+
 The purpose of the current study is to estimate the cost of the current measles outbreak in New Zealand.
@@
 Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
-Direct costs can be divided into medical and non-medical (Saha and Gerdtham, 2013).
+Direct costs can be divided into medical and non-medical \cite{saha13}.
 Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
-Personnel time (investigation and emergency response), materials (phone calls, vaccine, personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
+Personnel time (investigation and emergency response), materials (phone calls, vaccine), personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
 Direct non-medical costs include transportation to and from health care providers.
+
 Indirect costs are productivity losses for the case and/or health care provider, e.g.
@@
 Typically this is 3\% but 5\% has also been used in a sensitivity analysis, which is more compared to non-human life calculations and will tend to reduce the present value of the future earnings (saved by avoiding a case).
+
 Data for the current measles outbreak were obtained from the New Zealand Ministry of Health, from 2008 through June 2014.
@@
 It includes a breakdown by individual performing the work and whether it was during the normal work schedule (Monday to Friday, M-F) or weekends.
-Normal work was calculated as 1.2*full time equivalent (FTE)*number of days worked.
-Overtime was calculated as 1.6*FTE (M-F) and 2.0*FTE (weekend).
+Normal work was calculated as $1.2 \times \textsf{full time equivalent (FTE)} \times \textsf{number of days worked}$.
+Overtime was calculated as $1.6\times\textsf{FTE}$ (M-F) and $2.0 \times \textsf{FTE}$ (weekend).
 A full day was considered as 8 hours worked.
-Salary (hourly) rates were calculated for the following: public health nurse (PHN, \$36), public health assistant (PHA, \$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), incident management team (IMT), which had the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and safety/security offier (SSO) (\$26).
+Salary (hourly) rates were calculated for the following: public health nurse (PHN, \$36), public health assistant (PHA, \$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), incident management team (IMT), which had the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and safety/security officer (SSO) (\$26).
 In addition, measles operations personnel were calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
 
-Mean wages for New Zealand workers, by age and gender were obtained for the period, 2008-2013 from the New Zealand Income Survey (Statistics New Zealand, 2013).
+Mean wages for New Zealand workers, by age and gender were obtained for the period, 2008--2013 from the New Zealand Income Survey (Statistics New Zealand, 2013).
 Measles cases were assumed to not work for a period of 5 days.
@@
 In order to calculate the wage loss associated with the care taker, it was assumed that the person was a female between the ages of 35-39.
-Age and gender information for the 192 publicly funded hospital discharges with a measles primary diagnosis from 2008-2013 were matched to the New Zealand wage file to calculate lost wages due to measles.
+Age and gender information for the 192 publicly funded hospital discharges with a measles primary diagnosis from 2008--2013 were matched to the New Zealand wage file to calculate lost wages due to measles.
 
@@
 \subsection{Results}
-Direct costs for measles management in New Zealand for the 10-week period, January 1 - March 9, 2014 are shown in  Table \ref{table:direct}.
-The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table \ref{table:hosp}.
+Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
+The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\ref{table:hosp}.
 
 \begin{table}
-\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 - March 9, 2014 (see text for abbreviations)}
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
 %latex.default(costtable1, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was \$289,694 (Table \ref{table:hosp}).
+The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was \$289,694 (Table~\ref{table:hosp}).
 The mean cost per case was \$1,509.
@@
 \begin{table}
-\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008-2013}
+\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008--2013}
 %latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 - 19 June 2014).
-The number of cases by age group is shown in Table \ref{table:freq}.
-Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 0.47 and 0.33, respectively.
+From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
+The number of cases by age group is shown in Table~\ref{table:freq}.
+Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.
 
 \begin{table}
-\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 - 19 June, 2014}
+\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
 %latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
@@
 Eighty-three of the cases stayed for 0 days and the remaining cases stayed a mean of 2.30 (median 2) days.
-The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2008-2013 appear in Table \ref{table:cases}.
-The cost for these 192 hospitalizations was \$289,964.
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2008--2013 appear in Table~\ref{table:cases}.
+The cost for these 192 hospitalisations was \$289,964.
 Of the 292 cases, 87 were female and 105 were male.
-The number of days spent in the hospital was 251, making the mean daily cost of \$1,509.
+The number of days spent in the hospital was 251, making the mean daily cost \$1,509.
 Of the 251 days, 114 were spent by females and 137 by males.
@@
 \begin{table}
-\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008-2013}
+\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008--2013}
 %latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-Regression analyses showed statistically significant associations between cost of hospitalization and three variables, length of hospitalization, case age and year of case, and a less strong association with case gender (Table \ref{table:regression}).
-Results showed an increase of approximately \$358 per day of hospitalization, \$139 for a female case, and \$1 per year of case diagnosis.
+Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
+Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis.
 The cost was found to decrease by approximately \$5 per year of case age.
@@
 \begin{table}
-\caption{Regression results (R$^{2}$ adj = 0.9, p-value <0.001) for measles hospitalization cost based on length of stay (days), gender, case age and year of case (n=192) in New Zealand, 2008 - 2013}
+\caption{Regression results ($R^{2}_\textsf{adj} = 0.9$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=192$) in New Zealand, 2008 -- 2013}
 %latex.default(costtable5, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-Wages lost due to measles were calculated based on the assumptions that 5 days of work were lost for each case; however, individuals between 0-15 years of age were not assumed to be employed and therefore did not suffer an income loss.
-If the case were < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
+Wages lost due to measles were calculated based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
+If the case were less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
 Total wage lost for the 192 cases and care givers was estimated to be \$166,638.
-This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases <15 years of age.
+This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases under 15 years of age.
 
@@
 For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
-In addition to this information, similar data would be needed for cases occurring outside the period 2011-2013 at publicly funded hospitals.
-Similar data would be needed for non-publicly funded hospitals, e.g.
+In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
+In addition, similar data would be needed for non-publicly funded hospitals, e.g.
 private clinics.
-Detailed measles outbreak management costs were provided for the period of January 1 - March 9, 2014.
+Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
+
+Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
 Similar data are needed for the period preceding 2014.
-If detailed data, such as that provided for early 2014, aggregated data would be acceptable.
+If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
 However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
@@
 They estimated direct costs per case to be less than US\$ 500.
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively (Prouty and Lakin, 2000).
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty1}.
 In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person.
 In 2000 the average annual expenditures for care in large state MR/DD facilities were \$ 113,864.
-The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalization) to \$ 29,556 (encephalitis and hospitalization for 8.7 days).
-They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) (Zhou et al., 2004).
+The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalisation) to \$ 29,556 (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \cite{zhou4}.
 
@@
 \begin{itemize}
-\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$ $\frac{3}{4}$ million.
+\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$ 750,000.
 \end{itemize}
@@
 
-\section{Modeling measles epidemics}
+\section{Modelling measles epidemics}\label{sec:epidemic_modelling}
 
-A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years.
-The results show that achieving coverage of >90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \cite{roberts0,roberts4,tobias98}.
+The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
 
-The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 (Tobias & Roberts 1998) successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign (Mansoor
-et al 1998; Roberts & Tobias 2000).
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \cite{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \cite{mansoor98,roberts0}.
 Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
-The model developed by Roberts & Tobias (2000) supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
-The schedule was changed in 2000 with MMR2 now being administered before 5 years (Anon.
-2002a) and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasized that it is necessary to maintaining high coverage rates in order to prevent future epidemics.
+The model developed by \cite{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years \cite{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics.
 
-These results were in line with those obtained by other authors, for example: Babad et al (1995) advocated a two-dose schedule for England and Wales, with the second vaccination given at age four; and Gay et al (1998) recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
-In addition, Agur et al (1993) found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
-These authors all agree that two
-vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-A different approach was taken by Wallinga et al (2001).
-These authors took existing policies in eight European countries and estimated the coverage rates required to reduce Rv below one.
-They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately
-87\%.
+These results were in line with those obtained by other authors, for example: \cite{babad95} advocated a two-dose schedule for England and Wales, with the second vaccination given at age four; and \cite{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, \cite{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+These authors all agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+A different approach was taken by \cite{wallinga1} who took existing policies in eight European countries and estimated the coverage rates required to reduce Rv below one.
+They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
 
-A variety of similar models for measles vaccination strategies have been developed by other authors for various regions (Agur et al 1993; Babad et al 1995; Edmunds et al 2000; Gay et al 1998; Wallinga et al 2001) based on sets of nonlinear differential equations (ODE), and the conclusions reached have been similar.
+A variety of similar models for measles vaccination strategies have been developed by other authors for various regions \cite{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equations (ODE), and the conclusions reached have been similar.
 The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
@@
 It is difficult to estimate the proportion of the school-age population that have been effectively immunised, but this is continuously being diluted by children who are not immunised.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, Glass et al (2004) in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \cite{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection.
 
-The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, R0.
-This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population (Anderson & May 1991; Diekmann & Heesterbeek 2000).
-If R0 > 1 an epidemic will occur following an introduction of infection.
-The best estimate we had for measles in New Zealand was R0 = 12.8 - 12.5 (Roberts, see results section).
-The basic reproduction number of the infection under vaccination, Rv, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
-If Rv < 1 epidemics are prevented.
-The case reproduction number of the infection at time t, Rt, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
+The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \cite{anderson91, diekmann0}.
+If $R_0 > 1$ an epidemic will occur following an introduction of infection.
+The best estimate we had for measles in New Zealand was $R_0$ = 12.8 \cite{roberts4}.
+The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
+If $R_v < 1$ epidemics are prevented.
+The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
 
-To understand the level of immunity in the population, how likely an outbreak was of becoming endemic, we estimated Rv from all the outbreaks since 2009.
-To do this we estimated Rt, following an adaptation of the methods in [REF], and used these to estimate the Rv of measles in the recent outbreaks, which helps understand the transmission dynamics of measles in the partially immune population.
-
-To estimate Rt we were required to compute the generation time for measles.
-We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from %# http://www.sciencedirect.com/science/article/pii/S0022519311003146
-We then estimated Rt using the methods adapted from [REF] from the incidence data for each outbreak.
-We defined outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
-The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.
+To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks since 2009.
+To do this we estimated $R_t$, following an adaptation of the methods in \cite{obidia12,wallinga4}.
+We were required to compute the generation time for measles to do so.
+The generation time is the average time an index case infects others after becoming infected.
+We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \cite{klinkenberg11}.
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.
 
@@
 
-The estimated Rv for each outbreak is shown in (Figure~\ref{fig:r0}).
-The probability density of the Rv estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak, which has an Rv well above one and thus we may expect this outbreak to persist if conditions remain the same.
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The probability density of the $R_v$ estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.An important caveat to this analysis is that because this is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one.
 
-These analyses also imply that the reguler (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles, as highlighted in the \emph {risk analysis} section is likely to continue these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
-The measles outbreak in 2011-2012 had an Rv of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given it's Rv is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other Rv estimates is that the 2013-2014 may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in section~\ref{sec:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
 However, sub-clinical and underreporting may lower the estimate.
-The relative contributions of both to our estimates are currnelty unknown.
+The relative contributions of both to our estimates are currently unknown.
 
-\subsection{Summary of modeling}
+\subsection{Summary of modelling}
 
 \begin{itemize}
-\item Regular introductons of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section [X]).
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sec:risk_analyses}).
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to become endemic.
+ \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
 \end{itemize}
 
-\section{Future modeling analyses}
-Future modeling analyses we aim to perform are:
+\section{Future modelling}
+Future modelling we aim to perform are:
 \begin{itemize}
@@
 \begin{itemize}
-\item New Zealand is at risk of frequent measles importation due to travel and endemic measles eslewhere in the globe.
-\item The cost of the current measles outbreak is estimated to be at least \$$\frac{3}{4}$ million.
-\item Analyses of outbreak data suggest that measles Rv values often include 1 and in this year, 2014, well above one.
-This analyses suggest improved vaccination is a requesite to prevent measles becoming endemic again.
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
+\item The cost of the current measles outbreak is estimated to be at least \$750,000.
+\item Analyses of outbreak data suggest that measles $R_v$ values often include 1 and in this year, 2014, are well above one.
+This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
 \end{itemize}
@@
 
-\bibitem[Gelman et al.(1996)Gelman, Roberts, and Gilks]{grg}
-Gelman, A., G.~O. Roberts, and W.~R. Gilks (1996).
-\newblock Efficient Metropolis jumping rules.
-\newblock In \emph{Bayesian Statistics, 5 (Alicante, 1994)}, pp.~599--607.
-  Oxford University Press.
+\bibitem[Agur et al.(1993)]{agur93}
+Agur, Z., L. Cojocaru, G. Mazor, R.~M. Anderson and Y.~L. Danon (1993).
+\newblock Pulse mass measles vaccination across age cohorts.
+\newblock \emph{Proceedings of the National Academy of Sciences USA}, 90, 11698--11702.
 
-\bibitem[Geyer(1992)]{practical}
-Geyer, C.~J. (1992).
-\newblock Practical Markov chain Monte Carlo (with discussion).
-\newblock \emph{Statistical Science}, 7, 473--511.
+\bibitem[Anderson and May(1991)]{anderson91}
+Anderson, R.~M. and R.~M. May (1991).
+\newblock \emph{Infectious diseases of humans: dynamics and control}. Oxford: Oxford University Press.
 
-\bibitem[Geyer and Thompson(1995)]{geyer-temp}
-Geyer, C.~J. and E.~A. Thompson (1995).
-\newblock Annealing Markov chain Monte Carlo with applications to
-    ancestral inference.
-\newblock \emph{Journal of the American Statistical Association}, 90, 909--920.
+\bibitem[Anon. (2002a)]{anon2a}
+Anon. (2002a).
+\newblock \emph{Immunisation handbook}
+\newblock Wellington: Ministry of Health. pp.~131--146.
+
+\bibitem[Anon. (2002b)]{anon2b}
+Anon. (2002b).
+\newblock \emph{Infectious diseases in livestock}
+\newblock The Royal Society. pp.~68.
+
+\bibitem[Babad et al.(1995)]{babad95}
+Babad, H.~R., D.~J. Nokes, N.~J. Gay, E. Miller, P. Morgan-Capner, and R.~M. Anderson (1995).
+\newblock Predicting the impact of measles vaccination in England and Wales: model validation and analysis of policy options.
+\newblock \emph{Epidemiology and Infection}, 114, 319--344.
+
+\bibitem[Bae et al. (2013)]{bae13}
+Bae, G.~R, Y.~J. Choe, U.~Y. Go, Y.~I. Kim, and J.~K. Lee (2013). 
+\newblock Economic analysis of measles elimination program in the Republic of Korea, 2001: A cost benefit analysis study.
+\newblock \emph {Vaccine}, 31, 2661--2666.
+
+\bibitem[Carabin et al. (2002)]{carabin2}
+Carabin, H., W.~J. Edmunds, U. Kou, S. van den Hof, and V.~H. Nguyen (2002). 
+\newblock Measles in industrialized countries: a review of the average costs of adverse events and measles cases.
+\newblock \emph{BMC Public Health}, 2, 22.
+
+\bibitem[Carabin et al. (2003)]{carabin3}
+NA
+\newblock The cost of measles in industrialised countries.
+\newblock \emph{Vaccine}, 21,4167--4177.
+
+\bibitem[Clements and Hussey (2004]{clements4}
+Clements, C.~J. and G.~D. Hussey (2004).
+\newblock Chapter 4: Measles.
+\newblock In \emph{The Global Epidemiology of Infectious Diseases},  Murray, C., A.~D. Lopez, and C.~D. Mathers, (eds.), Geneva.
+  World Health Organization, pp.~391.
+
+\bibitem[Coleman et al.(2012)]{coleman12}
+Coleman, M.~S., L. Garbat-Welch, H. Burke, M. Weinberg, K. Humbaugh, A. Tindall, and J. Cambron (2012).
+\newblock Direct costs of a single case of refugee-imported measles in Kentucky.
+\newblock \emph{Vaccine}, 30,317--321.
+
+\bibitem[Dayan et al.(2005)]{dayan5}
+NA
+\newblock The cost of containing one case of measles: the economic impact on the public health infrastructure - Iowa, 2004.
+\newblock \emph{Pediatrics}, 116:e1; DOI:10/1542/peds.2004-2512.
+
+\bibitem[Diekmann et al.(2000)]{diekmann0}
+Diekmann, O. and  J.~A.~P. Heesterbeek (2000).
+\newblock \emph{Mathematical epidemiology of infectious diseases: model building, analysis and interpretation}.
+Chichester: Wiley.
+
+\bibitem[Edmunds et al.(2000)]{edmunds0}
+Edmunds, W.~J., N.~J. Gay, M. Kretzschmar, R.~G. Pebody and H. Wachman (2000).
+\newblock The pre-vaccination epidemiology of measles, mumps and rubella in Europe: implications for modelling studies.
+\newblock \emph{Epidemiology and Infection}, 125, 635--650.
+
+\bibitem[Filia et al.(2007)]{filia7}
+NA
+\newblock Health burden and economic impact of measles-related hospitalization in Italy, 2002-2003.
+\newblock \emph{BMC Public Health}, 7,169
+
+\bibitem[Flego et al.(2013)]{flego13}
+Flego, K.~L., D.~A. Belshaw, V. Sheppeard, and K.~M. Weston (2013).
+\newblock Impacts of a measles outbreak in western Sydney on public health resources.
+\newblock \emph{Communicable Diseases Intelligence Quarterly Report}, 37, E240--245.
+
+\bibitem[Gay et al.(1998)]{gay98}
+Gay, N.~J., L. Pelletier, and P. Duclos (1998).
+\newblock Modelling the incidence of measles in Canada: an assessment of the options for vaccination policy.
+\newblock \emph{Vaccine}, 16, 794--801.
+
+\bibitem[Glass et al.(2004)]{glass4}
+Glass, K., J. Kappey, and B.~T. Grenfell (2004).
+\newblock The effect of heterogeneity in measles vaccination population immunity.
+\newblock \emph{Epidemiology and Infection}, 132, 675--683.
+
+\bibitem[Klinkenberg et al.(2011)]{klinkenberg11}
+Klinkenberg, D. and H. Nishiuraa (2011).
+\newblock The correlation between infectivity and incubation period of measles, estimated from households with two cases.
+\newblock \emph{Journal of Theoretical Biology},284, 52--60
+
+\bibitem[Koopmanschap(1998)]{koopmanschap98}
+Koopmanschap, M.~A. (1998).
+\newblock Cost-of-illness studies: useful for health policy?
+\newblock \emph{Pharmacoeonomics}, 14, 143--148.
+
+\bibitem[Larg and Moss(2011)]{larg11}
+Larg, A. and J.~R. Moss (2011).
+\newblock Cost-of-illness studies: a guide to critical evaluation.
+\newblock \emph{Pharmacoeconomics}, 29,653--671.
+
+\bibitem[Mansoor et al.(1998)]{mansoor98}
+Mansoor, O., A. Blakely, M. Baker, M. Tobias, and A. Bloomfield (1998).
+\newblock A measles epidemic controlled by immunisation. 
+\newblock \emph{New Zealand Medical Journal}, 111, 467--471.
+
+\bibitem[Ortega-Sanchez et al.(2014)]{ortegasanchez14}
+Ortega-Sanchez, I.~R., M. Vijayaraghavan, A.~E. Barskey, and G.~S. Wallace (2014).
+\newblock The economic burden of sixteen measles outbreaks on United States public health departments in 2011.
+\newblock \emph{Vaccine}, 32,1311--1317.
+
+\bibitem[Obidia et al.(2012)]{obidia12}
+NA
+\newblock The R0 package: a toolbox to estimate reproduction numbers for epidemic outbreaks.
+\newblock \emph{BMC Medical Informatics and Decision Making}, 2012, 12--147.
+
+\bibitem[Parker et al.(2006)]{parker6}
+NA
+\newblock Implications of a 2005 measles outbreak in Indiana for sustained elimination of measles in the United States.
+\newblock \emph{The New England Journal of Medicine}, 355, 447--455.
+
+\bibitem[Prouty et al. (2001)]{prouty1}
+Prouty, R.W., G. Smith and K.~C. Lakin (2001).
+\newblock Residential services for persons with developmental disabilities: status and trends through 2000.
+\newblock \emph{Minneapolis: Institute on Community Integration}, University of Minnesota, pp.~179, rtc.umn.edu/risp00.
+
+\bibitem[Roberts(2004)]{roberts4}
+Roberts, M. (2004).
+\newblock A mathematical model for measles vaccination.
+\newblock Wellington: Ministry of Health.
+
+\bibitem[Roberts and Tobias(2000)]{roberts0}
+Roberts, M.~G. and M.~I. Tobias (2000).
+\newblock Predicting and preventing measles epidemics in New Zealand: Application of a mathematical model. 
+\newblock \emph{Epidemiology and Infection}, 124, 279--287.
+
+\bibitem[Saha and Gerdtham(2013)]{saha13}
+Saha, S. and U.~G. Gerdtham (2013).
+\newblock Cost of illness studies on reproductive, maternal, newborn, and child health: a systematic literature review.
+\newblock \emph{Health Economics Review}, doi:10.1186/2191-1991-3-24.
+
+\bibitem[Siedler et al.(2006)]{siedler6}
+Siedler, A., A. Tischer, A. Mankertz, and S. Santibanez (2006).
+\newblock Two outbreaks of measles in Germany 2005.
+\newblock \emph{Eurosurveillance} 2006:11(4) article 5, http://www.eurosurveillance.org/ViewArticle.aspx?ArticleId=615, accessed 14 June 2014.
+
+\bibitem[Stack et al.(2011)]{stack11}
+Stack, M.~L., S. Ozawa, D.~M. Bishai, A. Mirelman, Y. Tam, L. Niessen, D.~G. Walker, and O.S. Levine (2011).
+\newblock Estimated economic benefits during the 'decade of vaccine' include treatment savings, gains in labor productivity.
+\newblock \emph{Health Affairs}, 30,1021--1028.
+
+\bibitem[Statistics New Zealand)]{stats14}
+\newblock \emph{Statistics New Zealand} (2014).
+http://nzdotstat.stats.govt.nz/, accessed 17 June 2014.
+
+\bibitem[Tobias and Roberts(1998)]{tobias98}
+Tobias, M.~I. and M.~G. Roberts (1998).
+\newblock Predicting and preventing measles epidemics in New Zealand: Application of a mathematical model.
+\newblock Wellington: Ministry of Health.
+
+\bibitem[Wallinga et al.(2001)]{wallinga1}
+Wallinga, J., D. Levy-Bruhl, N.~J. Gay, and C.~H. Wachman (2001).
+\newblock Estimation of measles reproduction ratios and prospects for elimination of measles by vaccination in some Western European countries.
+\newblock \emph{Epidemiology and Infection}, 127, 281--295.
+
+\bibitem[Wallinga and Teunis(2004)]{wallinga4}
+Wallinga, J., and P. Teunis (2004).
+\newblock Different Epidemic Curves for Severe Acute Respiratory Syndrome Reveal Similar Impacts of Control Measures.
+\newblock \emph{American Journal of Epidemiology}, 160, 509.
+
+\bibitem[Wichmann et al.(2009)]{wichmann9}
+Wichmann, O., A. Siedler, D. Sagebiel, W. Hellenbrand, S. Santibanez, A. Mankertz, G. Vogt, U. van Treeck, and G. Krause (2009).
+\newblock Further efforts needed to achieve measles elimination in Germany: results of an outbreak investigation.
+\newblock \emph{Bulletin of the World Health Organization}, 87, 108--115.
+
+\bibitem[Wolfson et al(2007)]{wolfson7}
+Wolfson, L.~J., P.~M. Strebel, M. Gacic-Dobo, E.~J. Hoekstra, J.~W. McFarland, and B.~S. Hersh (2007).
+\newblock Has the 2005 measles mortality reduction goal been achieved? A natural history modelling study.
+\newblock \emph{Lancet}, 369, 191--200.
+
+\bibitem[WHO (2013)]{who13}
+World Health Organisation measles media centre, January (2013)
+\newblock Geneva: World Health Organization.
+\href{http://www.who.int/mediacentre/news/notes/2013/measles_20130117/en/}{www.who.int}, accessed July 1, 2014.
+
+\bibitem[Zhou et al.(2004)]{zhou4}
+Zhou, F, S. Reef, M. Massoudi, M.~J. Papania, H.~R. Yusuf, B. Bardenheier, L. Zimmerman, and M.~M. McCauley (2004).
+\newblock An economic analysis of the current universal 2-dose measles-mumps-rubella vaccination program in the United States.
+\newblock \emph{Journal of Infectious Diseases}, 189, S131--45.
 

### 2014-07-04 interimreport1.Rnw
Interim report

@@
 \usepackage{tikz}
-\usepackage{bibunits}
-%\usepackage{natbib}
+%\usepackage{bibunits}
+\usepackage{natbib}
 \usepackage{graphics}
@@
 \href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
-\date{\today}  %\today is replaced with the current date
+\date{3 July 2014}  %\today is replaced with the current date
 \maketitle
@@
 \item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
-This analyses suggest improved vaccination is a requisite to prevent measles becoming endemic again.
+This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
 \end{itemize}
@@
 Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
-The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{\today}).
+The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}).
 In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
 
-Previous analyses of measles, including two in New Zealand by Prof.
-Roberts \cite{roberts0,roberts4}, have estimated that the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles.
+Previous measles analyses, including two in New Zealand by Prof.
+Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section \ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
 
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our preliminary modelling and risk analyses results (sections \ref{sec:risk_analyses} and \ref{sec:epidemic_modelling}).
+This is particularly in light of our preliminary modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).
 
@@
 
-\subsection{Additional risk analyses}\label{sec:risk_analyses}
+\subsection{Additional risk analyses}
+\label{sub:risk_analyses}
 
@@
 Note this uses all immigration of foreign nationals, coming for whatever purpose.
-We used the WHO data to determine per capita measles cases for each year and used this and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
+We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
 
@@
 \begin{table}
-\caption{Lowest national measles vaccine cover (2012)}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
 <<results=tex,echo=FALSE>>=
@@
 Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
-However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:incidence12} and Table \ref{table:cover12}).
+However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}).
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
@@
 
-\subsection{Future risk analyses}\label{sec:risk_analyses}
+\subsection{Future risk analyses}
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
@@
 Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above.
-These data only became available to us on 3 July 2014, provided by the University of Otago.
+These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data only became available to us on 3 July 2014, provided by the University of Otago.
 
@@
 
-Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \cite{clements4}.
+Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \citep{clements4}.
 Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities.
-\cite{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
+\citep{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
 Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
 They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination.
-The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \cite{who13}.
+The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \citep{who13}.
 In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.
 
-The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$ 150 million \cite{carabin3].
-The estimated cost for a case ranged from US\$ 189--344 \cite{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$ 967--1,755 \cite{carabin2}.
-\cite{stack11} estimated the economic benefits from cases averted due to measles vaccination.
-They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$ 10 billion of costs averted between 2011 and 2020.
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3].
+The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}.
+\citep{stack11} estimated the economic benefits from cases averted due to measles vaccination.
+They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion of costs averted between 2011 and 2020.
 Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
 
-Italy has the highest reported annual cost of measles among industrialised countries \cite{carabin3}.
-In 2001, it reported losses related to measles of approximately US\$ 50 million.
+Italy has the highest reported annual cost of measles among industrialised countries \citep{carabin3}.
+In 2001, it reported losses related to measles of approximately US\$50 million.
 The economic impact of a large measles outbreak in Italy, 2002--03 examined the costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis.
 The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
-The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 ~ NZ\$ 2.0 in 2002-03), or approximately \euro 1,700 per case.
+The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 ~ NZ\$2.0 in 2002-03), or approximately \euro 1,700 per case.
 The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
@@
 
-An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases \cite{flego13}.
+An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases \citep{flego13}.
 Seven (27\%) of the cases required hospitalisation for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
@@
 The mean number of contacts per case was 54 (median = 28, maximum = 206).
-The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$ 48,000, with 90\% of this being associated with staff time.
+The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$48,000, with 90\% of this being associated with staff time.
 
 Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years.
-In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \cite{wichmann9}.
-However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children (Siedler et al, 2006).
+In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \citep{wichmann9}.
+However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children \citep{siedler6}.
 An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
@@
 
-Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world \cite{parker6}.
+Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world \citep{parker6}.
 Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.
-\cite{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
-The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$ 2.7 and 5.3 million.
-Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$ 298 per contact.
-\cite{dayan5} calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
+\citep{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
+The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$2.7 and 5.3 million.
+Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$298 per contact.
+\citep{dayan5} calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
 They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination.
-In total, it was estimated the direct costs associated with three cases of measles was US\$ 142,452, or nearly US\$ 50,000 per case.
+In total, it was estimated the direct costs associated with three cases of measles was US\$142,452, or nearly US\$50,000 per case.
 
-\cite{parker6} reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
+\citep{parker6} reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
 A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak.
@@
 
-\cite{coleman12} estimated the direct medical and public health costs in response to a single case of refugee-imported measles.
+\citep{coleman12} estimated the direct medical and public health costs in response to a single case of refugee-imported measles.
 Costs included labor, translation and benefits for public health workers.
 In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis.
-In total, 387 hours were associated with this single case, resulting in a cost of US\$ 11,881.
-In addition, per-contact costs amounted to US\$ 264.
-The cost of hospitalisation for the 3-day stay by the index case was US\$ 931.
-Additional costs were associated with physician visits (US\$ 294), vaccine and immunoglobulin (US\$ 1,765), mileage (US\$ 205) and immunologic screening tests for the parents' exposed to measles (US\$ 240) for a total of US\$ 23,816.
+In total, 387 hours were associated with this single case, resulting in a cost of US\$11,881.
+In addition, per-contact costs amounted to US\$264.
+The cost of hospitalisation for the 3-day stay by the index case was US\$931.
+Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents' exposed to measles (US\$240) for a total of US\$23,816.
 
@@
 Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
-Direct costs can be divided into medical and non-medical \cite{saha13}.
+Direct costs can be divided into medical and non-medical \citep{saha13}.
 Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
@@
 
-In other outbreaks, the average cost per measles case was estimated to be US\$ 254, US\$ 276, and US\$ 307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
 This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
-The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$ 142,542.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
 In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
-They estimated direct costs per case to be less than US\$ 500.
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$ 31,059 and US\$ 78,448, respectively \cite{prouty1}.
-In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$ 113,864 per person.
-In 2000 the average annual expenditures for care in large state MR/DD facilities were \$ 113,864.
-The cost of a case of measles was estimated to range from \$ 71 (no complications and no hospitalisation) to \$ 29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \cite{zhou4}.
+They estimated direct costs per case to be less than US\$500.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
+The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
 
@@
 \begin{itemize}
-\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$ 750,000.
+\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$750,000.
 \end{itemize}
@@
 \begin{itemize}
-\item Estimate the costs for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modelling.
+\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modelling.
 \end{itemize}
 
-We also require additional clarifications of the data, regarding what hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight?
-Or, does it mean the case stayed for less than 24 hours?
+We also require additional clarifications of the data, regarding:
+\begin {itemize}
+\item What hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight? Or, does it mean the case stayed for less than 24 hours?
+\end {itemize}
 Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
 
-\section{Modelling measles epidemics}\label{sec:epidemic_modelling}
+\section{Modelling measles epidemics}
+\label{sec:epidemic_modelling}
 
-A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \cite{roberts0,roberts4,tobias98}.
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
 The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
 
-The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \cite{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \cite{mansoor98,roberts0}.
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
 Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
-The model developed by \cite{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
-The schedule was changed in 2000 with MMR2 now being administered before 5 years \cite{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics.
+The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
 
-These results were in line with those obtained by other authors, for example: \cite{babad95} advocated a two-dose schedule for England and Wales, with the second vaccination given at age four; and \cite{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
-In addition, \cite{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
-These authors all agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-A different approach was taken by \cite{wallinga1} who took existing policies in eight European countries and estimated the coverage rates required to reduce Rv below one.
+These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
 They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
 
-A variety of similar models for measles vaccination strategies have been developed by other authors for various regions \cite{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equations (ODE), and the conclusions reached have been similar.
+Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
 The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
-
-It is difficult to estimate the proportion of the school-age population that have been effectively immunised, but this is continuously being diluted by children who are not immunised.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \cite{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
 
 The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
-This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \cite{anderson91, diekmann0}.
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}.
 If $R_0 > 1$ an epidemic will occur following an introduction of infection.
-The best estimate we had for measles in New Zealand was $R_0$ = 12.8 \cite{roberts4}.
+The best estimate for measles in New Zealand was $R_0$ = 12.8 \citep{roberts4}.
 The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
@@
 
-To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks since 2009.
-To do this we estimated $R_t$, following an adaptation of the methods in \cite{obidia12,wallinga4}.
+To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
+To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
 We were required to compute the generation time for measles to do so.
 The generation time is the average time an index case infects others after becoming infected.
-We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \cite{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).The outbreaks we used in out analyses are shown in Figure~\ref{fig:outbreaks}.
+We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
 
@@
 The probability density of the $R_v$ estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.An important caveat to this analysis is that because this is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one.
+Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
+An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
 
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sec:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
 The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
@@
 \subsection{Summary of modelling}
-
 \begin{itemize}
-\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sec:risk_analyses}).
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
@@
 \begin{itemize}
-\item An update of previous ODE models of measles outbreak risk in the overall population according the differing vaccine coverage scenarios.
+\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
 \item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
@@
 \begin{thebibliography}{}
+\bibliographystyle{plain}
+\bibliography{sample}
 
@@
 
-\bibitem[Anon. (2002a)]{anon2a}
+\bibitem[Anon.(2002a)]{anon2a}
 Anon. (2002a).
@@
 
-\bibitem[Anon. (2002b)]{anon2b}
+\bibitem[Anon.(2002b)]{anon2b}
 Anon. (2002b).
@@
 
-\bibitem[Bae et al. (2013)]{bae13}
+\bibitem[Bae et al.(2013)]{bae13}
 Bae, G.~R, Y.~J. Choe, U.~Y. Go, Y.~I. Kim, and J.~K. Lee (2013). 
@@
 
-\bibitem[Carabin et al. (2002)]{carabin2}
+\bibitem[Carabin et al.(2002)]{carabin2}
 Carabin, H., W.~J. Edmunds, U. Kou, S. van den Hof, and V.~H. Nguyen (2002). 
@@
 
-\bibitem[Carabin et al. (2003)]{carabin3}
-NA
+\bibitem[Carabin et al.(2003)]{carabin3}
+Carabin, H., W.~J. Edmunds, M. Gyldmark, P. Beutels, D. Lévy-Bruhl, H. Salo, U.~K. and Griffiths (2003)
 \newblock The cost of measles in industrialised countries.
@@
 
-\bibitem[Clements and Hussey (2004]{clements4}
+\bibitem[Clements and Hussey(2004]{clements4}
 Clements, C.~J. and G.~D. Hussey (2004).
@@
 
-\bibitem[Prouty et al. (2001)]{prouty1}
+\bibitem[Prouty et al.(2001)]{prouty1}
 Prouty, R.W., G. Smith and K.~C. Lakin (2001).
@@
 
-\bibitem[Wolfson et al(2007)]{wolfson7}
+\bibitem[Wolfson et al.(2007)]{wolfson7}
 Wolfson, L.~J., P.~M. Strebel, M. Gacic-Dobo, E.~J. Hoekstra, J.~W. McFarland, and B.~S. Hersh (2007).
@@
 
-\bibitem[WHO (2013)]{who13}
+\bibitem[WHO(2013)]{who13}
 World Health Organisation measles media centre, January (2013)

### 2014-07-04 interimreport1.tex
Interim report

@@
 \begin{table}
-\caption{Lowest national measles vaccine cover (2012)}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
 %latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-\bibitem[Carabin et al. (2003)]{carabin3}
+\bibitem[Carabin et al.(2003)]{carabin3}
 NA

### 2014-07-04 interimreport1.Rnw
Minor edits

@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that gives information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
 
@@
 To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
+
+\subsection{Additional risk analyses methods}
 
@@
 \end{table}
+
+\subsection{Additional risk analyses results}
 
@@
 Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
-Thus the data gap that we have that will hinder us providing fine scale risk maps are:
+We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
+meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
+Thus the data gap that we have that will hinder us providing fine scale risk maps is:
 \begin{itemize}
 \item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
+\end {itemize}
+An additional data set that would enable us to develop the understanding of measles importation risk is:
+\begin {itemize}
 \item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
 \end{itemize}
-
-We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
-meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
 
@@
 The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
-The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 ~ NZ\$2.0 in 2002-03), or approximately \euro 1,700 per case.
+The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 $\approx$ NZ\$2.0 in 2002-03), or approximately \euro 1,700 per case.
 The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
@@
 The mean number of contacts per case was 54 (median = 28, maximum = 206).
-The estimated cost to the public health unit for contact management for the epidemic was in excess of A\$48,000, with 90\% of this being associated with staff time.
+The estimated cost to the public health unit for contact management for the epidemic was in excess of AUS\$48,000, with 90\% of this being associated with staff time.
 
@@
 
-\subsection{Methods}
+\subsection{Cost analyses methods}
 Costs were evaluated as either direct or indirect.
@@
 
-\subsection{Results}
+\subsection{Cost analyses results}
 Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
@@
 \begin{itemize}
-\item Our initial estimates suggest the ongoing 2014 measles outbreak has cost New Zealand over \$750,000.
+\item Our initial estimates suggest the ongoing 2013--2014 measles outbreak has cost New Zealand over \$750,000.
 \end{itemize}
@@
 \begin{itemize}
-\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses or modelling.
+\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses (section~\ref{sub:risk_analyses}) or modelling (section~\ref{sec:epidemic_modelling}).
 \end{itemize}
@@
 
+\subsection{Modelling methods}
+
 To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
@@
 We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
+The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
 
@@
 
+\subsection{Modelling results and discussion}
+
 The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
@@
 
-\section{Future modelling}
+\subsection{Future modelling}
 Future modelling we aim to perform are:
@@
 \bibliographystyle{plain}
-\bibliography{sample}
 

### 2014-07-04 interimreport1.tex
Minor edits

@@
 
+\subsection{Additional risk analyses results}
+
 Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.

### 2014-07-07 interimreport1.Rnw
Updated bibliography
No other changes at all

@@
 \bibitem[Carabin et al.(2003)]{carabin3}
-Carabin, H., W.~J. Edmunds, M. Gyldmark, P. Beutels, D. Lévy-Bruhl, H. Salo, U.~K. and Griffiths (2003)
+Carabin, H., W.~J. Edmunds, M. Gyldmark, P. Beutels, D. Levy-Bruhl, H. Salo, U.~K. and Griffiths (2003)
 \newblock The cost of measles in industrialised countries.
@@
 \bibitem[Dayan et al.(2005)]{dayan5}
-NA
+G.~H. Dayan, I.~R. Ortega-Sanchez, C.~W. LeBaron, M.~P. Quinlisk, and the Iowa Measles Response Team (2005).
 \newblock The cost of containing one case of measles: the economic impact on the public health infrastructure - Iowa, 2004.
@@
 \bibitem[Filia et al.(2007)]{filia7}
-NA
+Filia, A., A. Brenna, A. Pana, G.~M. Cavallaro, M. Massari and M.~L.C. degli Atti (2007).
 \newblock Health burden and economic impact of measles-related hospitalization in Italy, 2002-2003.
@@
 \bibitem[Obidia et al.(2012)]{obidia12}
-NA
+Obadia, T., R. Haneef and P--Y. Boelle
 \newblock The R0 package: a toolbox to estimate reproduction numbers for epidemic outbreaks.
@@
 \bibitem[Parker et al.(2006)]{parker6}
-NA
+Parker, A.~A., W. Staggs, G.~H. Dayan, I.~R. Ortega-Sanchez, P.~A. Rota, L. Lowe, P. Boardman, R. Teclaw, C. Graves, and C.~W. LeBaron (2006).
 \newblock Implications of a 2005 measles outbreak in Indiana for sustained elimination of measles in the United States.
@@
 \newblock Two outbreaks of measles in Germany 2005.
-\newblock \emph{Eurosurveillance} 2006:11(4) article 5, http://www.eurosurveillance.org/ViewArticle.aspx?ArticleId=615, accessed 14 June 2014.
+\newblock \emph{Eurosurveillance} 2006:11(4) article 5, \href{http://www.eurosurveillance.org/ViewArticle.aspx?ArticleId=615}{www.eurosurveillance.org}, accessed 14 June 2014.
 
@@
 
-\bibitem[Statistics New Zealand)]{stats14}
+\bibitem[Statistics New Zealand(2014))]{stats14}
 \newblock \emph{Statistics New Zealand} (2014).

### 2014-07-07 interimreport1.tex
Updated bibliography
No other changes at all

@@
 \bibitem[Carabin et al.(2003)]{carabin3}
-NA
+Carabin, H., W.~J. Edmunds, M. Gyldmark, P. Beutels, D. Levy-Bruhl, H. Salo, U.~K. and Griffiths (2003)
 \newblock The cost of measles in industrialised countries.

### 2014-08-21 interimreport2.Rnw
interim report 2 draft to work on
last report as template to be edited at home with cost benefit analyses
table

--- 2014-07-07 interimreport1.Rnw
@@
 
-\title{Interim report\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
+\title{Interim report 2\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
 \author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\

### 2014-08-21 interimreport2.Rnw
major remodel
more to come...

@@
 \href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
-\date{3 July 2014}  %\today is replaced with the current date
+\today  %\today is replaced with the current date \date{....}
 \maketitle
@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
-We identified additional analyses that may help understand risk of infection in New Zealand, began analyses of the cost of the current 2013--2014 outbreak and began modelling measles outbreaks to understand how they may be controlled, and to evaluate the likelihood that measles may become endemic once more.
+In this interim report we report the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated analyses of the cost of previous and the current 2013--2014 outbreak and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
+We provide some preliminary cost-benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
 Our key findings were:
 \begin{itemize}
-\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
-\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
-\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
-\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
-This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
+\item text.
+\item text.
+\item text.
 \end{itemize}
@@
 As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
-The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.
-
-The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
-The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}).
+A brief review of the history of measles in New Zealand was provided in our previous interim report.
+However, the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \today).
 In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
@@
 Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
-Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.
 
-\section{Risk analysis review}
+\section{Risk analysis update}
+\label{sub:risk_analyses}
 
@@
 Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-\begin{itemize}
-\item In this section we review the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}.
-\end{itemize}
-
-Overall, the review was very thorough.
-The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
-Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).
-
-Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand.
-Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
-
-Specific epidemiological details were provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}).
-Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}).
-Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
-
-The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
-
-We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe.
-Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
-Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
-However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our preliminary modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).
-
-The analyses we believe could further inform the understanding of risk from measles infection are:
+In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection.
+These additional analyses were:
 \begin{itemize}
@@
 
-\subsection{Additional risk analyses}
-\label{sub:risk_analyses}
+<<echo=FALSE>>=
+## regression analyses - measles
+## get data
+# set wd
+ setwd("~/Massey 2014/DHayman_20140627")
+# read data
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
 
-In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
-These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+                 data = data , FUN=sum)
+library(reshape)
+library(reshape2)
+library(plyr)
+library(pscl)
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
-     \caption{Measles incidence from 1997 to 2014}
-     \label{fig:incidence1997}
-\end{figure}
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+head(test)
+dim(test)
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+                   data = test , FUN=sum)
+caseyr
+test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
+test$AgeInYears<-as.factor(test$AgeInYears)
+tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
+test$AgeInYears<-tage
+#summary(test$EthnicityPrioritised)
+teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
+                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
 
-In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
+test$EthnicityPrioritised<-teth
+head(test)
+test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
+test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
 
-\subsection{Additional risk analyses methods}
+# want to set NZ Deprivation to the appropriate year
 
-For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
-Note this uses all immigration of foreign nationals, coming for whatever purpose.
-We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
+testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
+                              ifelse (test$RptYear == "2008",test$NZDep06,
+                              ifelse (test$RptYear == "2009",test$NZDep06,
+                              ifelse (test$RptYear == "2010",test$NZDep06,
+                              ifelse (test$RptYear == "2011",test$NZDep06,
+                              ifelse (test$RptYear == "2012",test$NZDep06,
+                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
+head(testt)
+tail(testt)
+str(testt)
+testt$NZDep<-as.factor(testt$NZDep-1)
 
-<<echo=FALSE>>=
-library(xtable)
-dpdf<-read.csv("mapdata.csv",header=T)
-# output top countries
-colnames(dpdf)[1]<-c("country")
-topimmigration <- dpdf[order(-dpdf$immigration),] 
-immigration<-topimmigration[1:10,c(1,4)]
-topincidence <- dpdf[order(-dpdf$incidence),] 
-incidence<-topincidence[1:10,c(1,2)]
-incidence[,2]<-round(incidence[,2],0)
-topvaccine <- dpdf[order(dpdf$cover),] 
-vaccinecover<-topvaccine[1:10,c(1,3)]
-toprisk <- dpdf[order(-dpdf$risk),] 
-risk<-(toprisk[1:10,c(1,5)])
-risk[,2]<-round(risk[,2],0)
+testtable<-aggregate( cbind( DiseaseName + as.numeric(SurvWeek)) ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
+                 data = testt , FUN=sum)
+dim(testtable)
+colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
+head(testtable)
+
+tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
+                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
+                                 "9"="6-10","10"="6-10"));
+
+testtable$NZDep<-tnzd
+head(testtable)
+tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
+                 data = testtable , FUN=sum)
+
+### denominator data
+ setwd("~/Massey_2014/measles/data")
+denom<-read.csv("NZDep2006Denominators.csv",header=T)
+head(denom)
+denom<-denom[,-c(9)]
+summary(denom)
+dim(denom)
+denom<-denom[,1:18]
+row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
+sum(row.has.na)
+denomt <- denom[!row.has.na,]
+head(denomt)
+denomt<-denomt[denomt$Sex_code==99,]
+denomt$Eth_Level<-as.factor(denomt$Eth_Level)
+denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
+dm<-cbind(denomt[,c(3,6,9:18)])
+head(dm)
+dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
+dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
+dm$Age_Label<-dage
+colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
+head(dm)
+datadm<-dm[dm$Age=="<6",]
+dataA<-(rbind(datadm,datadm)/2)
+dataA<-round(dataA[,3:12])
+dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
+colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
+ddm<-dm[!dm$Age=="<6",]
+dm<-rbind(dataA,ddm)
+dm$Ethnicity <- factor(dm$Ethnicity)
+deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
+                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
+dm$Ethnicity<-deth
+head(dm)
+
+popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
+colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
+popn
+pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
+                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
+                                 "Dep9"="6-10","Dep10"="6-10"));
+
+popn$NZDep<-pnzd
+head(popn)
+tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
+               data = popn , FUN=sum)
+
+tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
+testtable$merge <- paste(testtable$NZDep, testtable$Age, testtable$Ethnicity)
+testtable <- testtable[testtable$Ethnicity!="None",]
+
+tp$cases <- 0
+cases <- matrix(0, length(tp$merge),1)
+rownames(cases) <- tp$merge
+cases[testtable$merge,] <- testtable$Cases
+tp$cases <- cases
+
+## plot this figure?
+hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+tp$Ethnicity<- relevel(tp$Ethnicity, "European")
+model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
+summary(model)
+
+model1<-update(model,~.-Age:Ethnicity:NZDep)
+summary(model1)
+anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
+
+model2<-update(model1,~.-Age:NZDep)
+summary(model2)
+anova(model2,test="F")
+
+par(mfrow=c(2,2))
+hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+res<-predict(model2)
+plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
+cor(exp(res),tp$cases)
+cor.test(exp(res),tp$cases)
+hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
+par(mfrow=c(1,1))
+plot(tp$cases/tp$Popn,main="Cases per capita",ylab="Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
+     xlab="Category")
+legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
+       col=c(1:4,6),pch=16,bty="n")
+
+## drop MLA
+tpsub<-tp[!(tp$Ethnicity=="MLA"),]
+
+hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
+
+model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
+summary(model)
+
+model1<-update(model,~.-Age:Ethnicity:NZDep)
+summary(model1)
+anova(model1,test="F")
+
+model2<-update(model1,~.-Ethnicity:NZDep)
+summary(model2)
+anova(model2,test="F")
+
+model3<-update(model2,~.-Age:NZDep)
+summary(model3)
+anova(model3,test="F")
+
+##
+
+par(mfrow=c(2,2))
+hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+res<-predict(model3)
+plot(exp(res),tpsub$cases,xlab="predictions",ylab='cases',main="Fit",pch=16,col="darkgrey")
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+#abline(lm(exp(res)~tpsub$cases))
+hist(model3$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
+
+##
 @
-
-\begin{table}
-\caption{Immigration numbers to New Zealand (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(immigration, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:immigration12}
-\end{table}
-
-\begin{table}
-\caption{Lowest national measles vaccine cover (\%, 2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cover12}
-\end{table}
-
-\begin{table}
-\caption{Measles incidence per million (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(incidence, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:incidence12}
-\end{table}
-
-\begin{table}
-\caption{Risk of measles importation to New Zealand in 2012}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(risk, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:risk12}
-\end{table}
-
-\subsection{Additional risk analyses results}
-
-Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
-However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}).
-Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
-These data are shown in Table \ref{table:risk12}.
-The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
-
-<<echo=FALSE>>=
-# haven't had time to sort out which packages I need
-library(maps)       # Provides functions that let us plot the maps
-library(mapdata)    # Contains the hi-resolution points that mark out the countries.
-library(rworldmap)
-library(adehabitat)
-library(rgdal)
-library(raster)
-library(maps)
-library(maptools)
-library(sp)
-#library(SDMTools)
-library(ggplot2)
-require(rgeos)
-library(ggmap)
-library(mapdata)
-#gpclibPermit()
-library(mapproj)
-library(plyr)
-library(grid)
-library(gridExtra)
-#gpclibPermit()
-library(utils)
-library(rgdal)
-#getwd()
-#setwd("~/data")
-#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
-load("world.ggmap.Rda")
-n <- length(unique(world.ggmap$id))
-id = unique(world.ggmap$id)
-id<-as.data.frame(id)
-dpdf<-read.csv("mapdata.csv",header=T)
-# output top countries
-
-np1<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np2<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = incidence), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np3<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = cover), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
-
-#grid.arrange(np1, np2, np3, ncol=3, main="")
-
-np4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-@
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np1)
-@
-\end{center}
-\caption{Immigration to New Zealand 2012}
-\label{fig:immigration12}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np2)
-@
-\end{center}
-\caption{Measles incidence per million 2012}
-\label{fig:incidence12}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np3)
-@
-\end{center}
-\caption{Measles vaccination cover (\%) 2012 }
-\label{fig:cover12}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np4)
-@
-\end{center}
-\caption{Risk map from measles incidence and immigration rates}
-\label{fig:risk12}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE>>=
-change<-read.csv("change.csv",header=T)
-plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
-     bty="n")
-ylim <- range(c(0,max(change[,4])))
-xlim <- range(change[,1])
-par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
-plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
-     ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
-@
-\end{center}
-\caption{Trend in global per capita measles incidence}
-\label{fig:trendincidence}
-\end{figure}
-
-Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
-This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
-Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).
-This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE>>=
-ttimes<-read.csv("times.csv",header=T)
-lmres<-lm(res~time,data=ttimes)
-#summary(lmres)
-plot(ttimes$res,xaxt="n",type="l",xlab="Date",ylab="")
-axis(1, at=1:71, labels=ttimes$mydate) 
-abline(lmres,col="red",
-       lty=2)
-acf(ttimes$res,lag.max = length(ttimes$res),plot=T,main="")
-@
-\end{center}
-\caption{Trend in total immigrations into New Zealand}
-\label{fig:trendimmigration}
-\end{figure}
 
@@
 Using this information, we will then evaluate the economics of alternative measles control strategies in order to provide additional information to public health officials and decision makers.
+
+.....Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
+Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.....
 

### 2014-08-22 interimreport2.tex
major revisions

--- 2014-08-21 interimreport2.Rnw
@@
 \usepackage{todonotes}
+\usepackage{pdflscape}
+\usepackage{booktabs}
+\usepackage{array}
+\usepackage{rotating}
+\usepackage{threeparttable}
+\usepackage{blindtext}
+\usepackage{dcolumn}
+\usepackage{tabularx}
 
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
+\input{interimreport2-concordance}
 
@@
 \href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
-\today  %\today is replaced with the current date \date{....}
 \maketitle
@@
 In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection.
-These additional analyses were:
+The additional analyses included in this section are:
 \begin{itemize}
 \item Multivariate modelling to account for confounding within the univariate analyses.
-\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
 \item Analyses of changing risk factors through time during outbreaks.
-\item Modelling transmission chains in the population to understand effective reproduction number.
-\item Development of risk maps to understand measles importation.
 \end{itemize}
 
-<<echo=FALSE>>=
-## regression analyses - measles
-## get data
-# set wd
- setwd("~/Massey 2014/DHayman_20140627")
-# read data
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+\begin{Schunk}
+\begin{Sinput}
+> summary(model3)
+\end{Sinput}
+\begin{Soutput}
+Call:
+glm(formula = cases ~ Age + Ethnicity + NZDep + Age:Ethnicity + 
+    offset(log(Popn)), family = "quasipoisson", data = tpsub)
 
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-                 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
+Deviance Residuals: 
+    Min       1Q   Median       3Q      Max  
+-6.1174  -1.1562   0.0000   0.8599   6.7625
 
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-                   data = test , FUN=sum)
-caseyr
-test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
-test$AgeInYears<-as.factor(test$AgeInYears)
-tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
-test$AgeInYears<-tage
-#summary(test$EthnicityPrioritised)
-teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
-                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
+Coefficients:
+                            Estimate Std. Error t value Pr(>|t|)    
+(Intercept)               -8.268e+00  8.641e-01  -9.569 1.06e-08 ***
+Age3-5                     6.931e-01  1.053e+00   0.658 0.518322    
+Age6-17                    7.176e-01  9.064e-01   0.792 0.438300    
+Age18-24                  -5.001e-01  1.149e+00  -0.435 0.668367    
+Age25+                    -2.620e+00  1.158e+00  -2.262 0.035602 *  
+EthnicityAsian             2.167e+00  1.203e+00   1.801 0.087576 .  
+EthnicityMaori             1.772e+00  1.061e+00   1.671 0.111064    
+EthnicityPacific           3.871e+00  9.405e-01   4.117 0.000587 ***
+NZDep6-10                 -4.878e-01  2.608e-01  -1.870 0.076939 .  
+Age3-5:EthnicityAsian     -2.431e-15  1.473e+00   0.000 1.000000    
+Age6-17:EthnicityAsian    -2.473e+00  1.530e+00  -1.616 0.122649    
+Age18-24:EthnicityAsian   -7.910e-01  1.622e+00  -0.488 0.631293    
+Age25+:EthnicityAsian     -6.025e-01  1.792e+00  -0.336 0.740386    
+Age3-5:EthnicityMaori     -2.845e+00  2.177e+00  -1.307 0.206941    
+Age6-17:EthnicityMaori    -2.102e-01  1.121e+00  -0.187 0.853294    
+Age18-24:EthnicityMaori   -8.881e-01  1.630e+00  -0.545 0.592102    
+Age25+:EthnicityMaori     -1.090e+00  2.011e+00  -0.542 0.594088    
+Age3-5:EthnicityPacific   -4.423e+00  2.581e+00  -1.714 0.102866    
+Age6-17:EthnicityPacific  -2.253e+00  1.051e+00  -2.143 0.045264 *  
+Age18-24:EthnicityPacific -8.321e-01  1.318e+00  -0.631 0.535272    
+Age25+:EthnicityPacific   -2.718e-01  1.338e+00  -0.203 0.841210    
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
 
-test$EthnicityPrioritised<-teth
-head(test)
-test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
-test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
+(Dispersion parameter for quasipoisson family taken to be 16.26672)
 
-# want to set NZ Deprivation to the appropriate year
+    Null deviance: 3097.24  on 39  degrees of freedom
+Residual deviance:  291.05  on 19  degrees of freedom
+AIC: NA
 
-testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
-                              ifelse (test$RptYear == "2008",test$NZDep06,
-                              ifelse (test$RptYear == "2009",test$NZDep06,
-                              ifelse (test$RptYear == "2010",test$NZDep06,
-                              ifelse (test$RptYear == "2011",test$NZDep06,
-                              ifelse (test$RptYear == "2012",test$NZDep06,
-                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-head(testt)
-tail(testt)
-str(testt)
-testt$NZDep<-as.factor(testt$NZDep-1)
+Number of Fisher Scoring iterations: 6
+\end{Soutput}
+\end{Schunk}
 
-testtable<-aggregate( cbind( DiseaseName + as.numeric(SurvWeek)) ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
-                 data = testt , FUN=sum)
-dim(testtable)
-colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
-head(testtable)
+\begin{Schunk}
+\begin{Sinput}
+> anova(model3,test="F")
+\end{Sinput}
+\begin{Soutput}
+Analysis of Deviance Table
 
-tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-                                 "9"="6-10","10"="6-10"));
+Model: quasipoisson, link: log
 
-testtable$NZDep<-tnzd
-head(testtable)
-tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
-                 data = testtable , FUN=sum)
+Response: cases
 
-### denominator data
- setwd("~/Massey_2014/measles/data")
-denom<-read.csv("NZDep2006Denominators.csv",header=T)
-head(denom)
-denom<-denom[,-c(9)]
-summary(denom)
-dim(denom)
-denom<-denom[,1:18]
-row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
-sum(row.has.na)
-denomt <- denom[!row.has.na,]
-head(denomt)
-denomt<-denomt[denomt$Sex_code==99,]
-denomt$Eth_Level<-as.factor(denomt$Eth_Level)
-denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
-dm<-cbind(denomt[,c(3,6,9:18)])
-head(dm)
-dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
-dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
-dm$Age_Label<-dage
-colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-head(dm)
-datadm<-dm[dm$Age=="<6",]
-dataA<-(rbind(datadm,datadm)/2)
-dataA<-round(dataA[,3:12])
-dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
-colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-ddm<-dm[!dm$Age=="<6",]
-dm<-rbind(dataA,ddm)
-dm$Ethnicity <- factor(dm$Ethnicity)
-deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
-                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
-dm$Ethnicity<-deth
-head(dm)
+Terms added sequentially (first to last)
 
-popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
-colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
-popn
-pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
-                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
-                                 "Dep9"="6-10","Dep10"="6-10"));
-
-popn$NZDep<-pnzd
-head(popn)
-tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
-               data = popn , FUN=sum)
-
-tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
-testtable$merge <- paste(testtable$NZDep, testtable$Age, testtable$Ethnicity)
-testtable <- testtable[testtable$Ethnicity!="None",]
-
-tp$cases <- 0
-cases <- matrix(0, length(tp$merge),1)
-rownames(cases) <- tp$merge
-cases[testtable$merge,] <- testtable$Cases
-tp$cases <- cases
-
-## plot this figure?
-hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-tp$Ethnicity<- relevel(tp$Ethnicity, "European")
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
-
-model2<-update(model1,~.-Age:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-par(mfrow=c(2,2))
-hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
-res<-predict(model2)
-plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
-cor(exp(res),tp$cases)
-cor.test(exp(res),tp$cases)
-hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-par(mfrow=c(1,1))
-plot(tp$cases/tp$Popn,main="Cases per capita",ylab="Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="Category")
-legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
-       col=c(1:4,6),pch=16,bty="n")
-
-## drop MLA
-tpsub<-tp[!(tp$Ethnicity=="MLA"),]
-
-hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
-
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F")
-
-model2<-update(model1,~.-Ethnicity:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-model3<-update(model2,~.-Age:NZDep)
-summary(model3)
-anova(model3,test="F")
-
-##
-
-par(mfrow=c(2,2))
-hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
-res<-predict(model3)
-plot(exp(res),tpsub$cases,xlab="predictions",ylab='cases',main="Fit",pch=16,col="darkgrey")
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-#abline(lm(exp(res)~tpsub$cases))
-hist(model3$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-
-##
-@
+              Df Deviance Resid. Df Resid. Dev       F    Pr(>F)    
+NULL                             39    3097.24                      
+Age            4  1573.25        35    1523.98 24.1790 3.142e-07 ***
+Ethnicity      3   738.20        32     785.78 15.1271 2.853e-05 ***
+NZDep          1    55.39        31     730.39  3.4051   0.08064 .  
+Age:Ethnicity 12   439.34        19     291.05  2.2507   0.05519 .  
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{Soutput}
+\end{Schunk}
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
+     \caption{test}
+     \label{fig:regn}
+\end{figure}
 
@@
 
+\section{Modelling measles epidemics}
+\label{sec:epidemic_modelling}
+
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
+The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
+
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
+Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
+The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
+
+These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
+They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
+
+Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
+The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+
+The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}.
+If $R_0 > 1$ an epidemic will occur following an introduction of infection.
+The best estimate for measles in New Zealand was $R_0$ = 12.8 \citep{roberts4}.
+The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
+If $R_v < 1$ epidemics are prevented.
+The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
+
+\subsection{Modelling methods}
+
+To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
+To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
+We were required to compute the generation time for measles to do so.
+The generation time is the average time an index case infects others after becoming infected.
+We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
+The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{cases_by_dhb_2009_2014.pdf}
+     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
+     \label{fig:dhbcases}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
+     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
+     \label{fig:outbreaks}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
+     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \label{fig:r0}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
+     \caption{test}
+     \label{fig:percap}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
+     \caption{test}
+     \label{fig:percap}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
+     \caption{test}
+     \label{fig:percap}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
+     \caption{test}
+     \label{fig:agecases}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
+     \caption{test}
+     \label{fig:ageofvacc}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
+     \caption{test}
+     \label{fig:numvacc}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
+     \caption{test}
+     \label{fig:naive}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{simulations.pdf}
+     \caption{test}
+     \label{fig:sim}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
+     \caption{test}
+     \label{fig:vaccstat}
+\end{figure}
+
+\subsection{Modelling results and discussion}
+
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The probability density of the $R_v$ estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
+An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+
+These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+However, sub-clinical and underreporting may lower the estimate.
+The relative contributions of both to our estimates are currently unknown.
+
+\subsection{Summary of modelling}
+\begin{itemize}
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
+\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
+ \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
+\end{itemize}
+
+\subsection{Future modelling}
+Future modelling we aim to perform are:
+\begin{itemize}
+\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
+\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
+\end{itemize}
+
 \section{Cost analyses}
@@
 
-The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3].
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3}.
 The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}.
@@
 They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion of costs averted between 2011 and 2020.
-Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining \<1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
+Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining <1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
 
@@
 
-<<echo=FALSE>>=
-costtable1<-read.csv("costtable1.csv",header=T)
-@
+\begin{table}
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
+%latex.default(costtable1, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllll}
+\hline\hline
+\multicolumn{1}{c}{Category}&\multicolumn{1}{c}{January}&\multicolumn{1}{c}{February}&\multicolumn{1}{c}{March}&\multicolumn{1}{c}{Total}\tabularnewline
+\hline
+PHN&55,296&71,175&24,087&150,558\tabularnewline
+PHA&0&0&2,656&2,656\tabularnewline
+Data support&0&7,752&4,552&12,304\tabularnewline
+Supervisors&10,656&10,464&3,232&24,352\tabularnewline
+IMT&32,918&28,624&7,156&68,698\tabularnewline
+SSO&0&2,746&1,186&3,932\tabularnewline
+Measles operations&1,800&10,326&6,678&18,804\tabularnewline
+Operations partner&2,187&14,580&7,290&24,057\tabularnewline
+IMT controller partner&2,916&14,580&7,290&24,786\tabularnewline
+Total&105,773&160,247&64,127&330,147\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:direct}
+\end{table}
+
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hosptial was \$550,024 (Table~\ref{table:hosp}).
+The mean cost per case was \$1,877.
+The mean cost per day of stay in the hospital was \$1,170.
 
 \begin{table}
-\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable1, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:direct}
+\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
+%latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrlll}
+\hline\hline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Days}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Per.case}&\multicolumn{1}{c}{Per.day}\tabularnewline
+\hline
+2000&$  6$&$ 13$&8,850&1,475&681\tabularnewline
+2001&$ 13$&$ 18$&11,267&867&626\tabularnewline
+2002&$  5$&$  2$&3,869&774&1,934\tabularnewline
+2003&$  9$&$ 12$&10,241&1,138&853\tabularnewline
+2004&$  4$&$  5$&4,765&1,191&953\tabularnewline
+2005&$  3$&$ 11$&5,111&1,704&465\tabularnewline
+2006&$  1$&$  0$&602&602&NC\tabularnewline
+2007&$  5$&$ 25$&82,977&16,595&3,319\tabularnewline
+2008&$  3$&$  1$&3,038&1,013&3,038\tabularnewline
+2009&$ 29$&$ 38$&40,782&1,406&1,073\tabularnewline
+2010&$  5$&$  5$&6,701&1,340&1,340\tabularnewline
+2011&$132$&$189$&205,303&1,555&1,086\tabularnewline
+2012&$ 19$&$ 12$&28,540&1,502&2,378\tabularnewline
+2013&$  4$&$  6$&5,330&1,333&888\tabularnewline
+2014&$ 55$&$133$&132,648&2,412&997\tabularnewline
+TOTAL&$293$&$470$&550,024&1,877&1,170\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:hosp}
+ \centering
+ \begin{tablenotes}
+      \small
+      \item As of 11 July, 2014. NC - not calculated.
+    \end{tablenotes}
 \end{table}
 
-The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was \$289,694 (Table~\ref{table:hosp}).
-The mean cost per case was \$1,509.
-The mean cost per day of stay in the hospital was \$1,154.
-
-<<echo=FALSE>>=
-costtable2<-read.csv("costtable2.csv",header=T)
-@
-
-\begin{table}
-\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008--2013}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable2, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:hosp}
-\end{table}
-
-From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
+From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before 1 January 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
 The number of cases by age group is shown in Table~\ref{table:freq}.
@@
 
-<<echo=FALSE>>=
-costtable3<-read.csv("costtable3.csv",header=T)
-@
+\begin{table}
+\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
+%latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrr}
+\hline\hline
+\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Admitted}&\multicolumn{1}{c}{Proportion}\tabularnewline
+\hline
+\textless  15 months&$ 21$&$10$&$0.47$\tabularnewline
+15 months – 3 years&$  7$&$ 1$&$0.14$\tabularnewline
+4 – 9 years&$  8$&$ 0$&$0.00$\tabularnewline
+10 -1 19 years&$132$&$12$&$0.09$\tabularnewline
+\textgreater  19 years&$ 33$&$11$&$0.33$\tabularnewline
+Total&$201$&$34$&$0.17$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:freq}
+\end{table}
+
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 7 19 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
+Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital.
+Only eight cases spent a week or more in the hospital.
+Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days.
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
 
 \begin{table}
-\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable3, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:freq}
-\end{table}
-
-The majority (191/192) of the measles cases stayed in the hospital from 0 to 7 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
-Eighty-three of the cases stayed for 0 days and the remaining cases stayed a mean of 2.30 (median 2) days.
-The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2008--2013 appear in Table~\ref{table:cases}.
-The cost for these 192 hospitalisations was \$289,964.
-Of the 292 cases, 87 were female and 105 were male.
-The number of days spent in the hospital was 251, making the mean daily cost \$1,509.
-Of the 251 days, 114 were spent by females and 137 by males.
-The mean hospital stay per case was 1.31 and 1.30 days for females and males, respectively.
-The mean cost per case was \$1,573 and \$1,456 for females and males, respectively.
-
-<<echo=FALSE>>=
-costtable4<-read.csv("costtable4.csv",header=T)
-@
-
-\begin{table}
-\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008--2013}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable4, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cases}
+\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
+%latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllrrl}
+\hline\hline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Gender}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Length.of.stay}&\multicolumn{1}{c}{Cost.per.case}\tabularnewline
+\hline
+2000&F&4,296&$  2$&$  4$&2,148\tabularnewline
+&M&4,554&$  4$&$  9$&1,139\tabularnewline
+&Total&8,850&$  6$&$ 13$&1,475\tabularnewline
+2001&F&3,740&$  5$&$  5$&748\tabularnewline
+&M&7,527&$  8$&$ 13$&941\tabularnewline
+&Total&11,267&$ 13$&$ 18$&867\tabularnewline
+2002&F&924&$  2$&$  0$&462\tabularnewline
+&M&2,945&$  3$&$  2$&982\tabularnewline
+&Total&3,869&$  5$&$  2$&774\tabularnewline
+2003&F&9,766&$  8$&$ 12$&1,221\tabularnewline
+&M&475&$  1$&$  0$&475\tabularnewline
+&Total&10,241&$  9$&$ 12$&1,138\tabularnewline
+2004&F&1,437&$  1$&$  2$&1,437\tabularnewline
+&M&3,328&$  3$&$  3$&1,109\tabularnewline
+&Total&4,765&$  4$&$  5$&1,191\tabularnewline
+2005&F&0&$  0$&$  0$&0\tabularnewline
+&M&5,111&$  3$&$ 11$&1,704\tabularnewline
+&Total&5,111&$  3$&$ 11$&1,704\tabularnewline
+2006&F&0&$  0$&$  0$&0\tabularnewline
+&M&602&$  1$&$  0$&602\tabularnewline
+&Total&602&$  1$&$  0$&602\tabularnewline
+2007&F&1,930&$  1$&$  3$&1,930\tabularnewline
+&M&81,046&$  4$&$ 22$&20,262\tabularnewline
+&Total&82,977&$  5$&$ 25$&16,595\tabularnewline
+2008&F&714&$  1$&$  0$&714\tabularnewline
+&M&2,324&$  2$&$  1$&1,162\tabularnewline
+                    &Total&3,038&$  3$&$  1$&1,013\tabularnewline
+2009&F&11,953&$  7$&$ 15$&1,708\tabularnewline
+&M&28,830&$ 22$&$ 23$&1,310\tabularnewline
+                    &Total&40,782&$ 29$&$ 38$&1,406\tabularnewline
+2010&F&5,884&$  4$&$  5$&1,471\tabularnewline
+&M&817&$  1$&$  0$&817\tabularnewline
+                    &Total&6,701&$  5$&$  5$&1,340\tabularnewline
+2011&F&103,460&$ 66$&$ 86$&1,568\tabularnewline
+&M&101,842&$ 66$&$103$&1,543\tabularnewline
+                    &Total&205,303&$132$&$189$&1,555\tabularnewline
+2012&F&13,054&$  8$&$  6$&1,632\tabularnewline
+&M&15,486&$ 11$&$  6$&1,408\tabularnewline
+                    &Total&28,540&$ 19$&$ 12$&1,502\tabularnewline
+2013&F&1,800&$  1$&$  2$&1,800\tabularnewline
+&M&3,530&$  3$&$  4$&1,177\tabularnewline
+                    &Total&5,330&$  4$&$  6$&1,333\tabularnewline
+2014&F&55,633&$ 21$&$ 46$&2,649\tabularnewline
+&M&77,014&$ 34$&$ 87$&2,265\tabularnewline
+&Total&132,647&$ 55$&$133$&2,412\tabularnewline
+2000-2014&      F&335,431&$166$&$284$&2,021\tabularnewline
+&     M&214,591&$127$&$186$&1,690\tabularnewline
+&   TOTAL&550,022&$293$&$470$&1,877\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cases}
+ \centering
+ \begin{tablenotes}
+      \small
+      \item As of 7 August, 2014.
+    \end{tablenotes}
 \end{table}
@@
 Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
-Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis.
-The cost was found to decrease by approximately \$5 per year of case age.
-
-<<echo=FALSE>>=
-costtable5<-read.csv("costtable5.csv",header=T)
-@
+Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.
+The cost was \$256 less if the case were a male.
+It increased of approximately \$406 per night of hospitalisationand \$64 per year over the time period of 2000 - 2014.
+The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.
 
 \begin{table}
-\caption{Regression results ($R^{2}_\textsf{adj} = 0.9$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=192$) in New Zealand, 2008 -- 2013}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable5, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:regression}
+\caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
+%latex.default(costtable5, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrl}
+\hline\hline
+\multicolumn{1}{c}{Variable}&\multicolumn{1}{c}{Coefficient}&\multicolumn{1}{c}{P.value}\tabularnewline
+\hline
+Intercept&$ 581.39$&\textless  0.001\tabularnewline
+Length of stay (nights)&$ 406.07$&\textless  0.001\tabularnewline
+Gender (0 = F, 1 = M)&$-255.98$&0.006\tabularnewline
+Case age (years)&$  -8.23$&0.007\tabularnewline
+Year of case (vs. 2000)&$  64.35$&\textless  0.001\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:regression}
 \end{table}
 
-Wages lost due to measles were calculated based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
+Wages lost due to measles were calculated for the period January 2008 - August 2014.
+Calculations were based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
 If the case were less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
-Total wage lost for the 192 cases and care givers was estimated to be \$166,638.
-This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases under 15 years of age.
+Total wage lost for the 247 cases and care givers was estimated to be \$210,436.
+This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
+Overall, the cost per case from 2008 - 2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
 
-This final figure brings an approximate estimate of \$786,479 for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
+\begin{sidewaystable}[htbp]  
+  \centering  
+  \caption{Cost benefit analyses using simulated epidemic sizes}  
+   \begin{threeparttable}[b] \tiny
+    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
+    \toprule  
+        Years R0 estimated from & R0 range	& R0  for simulations	& Mean # of cases in the population (1000 simulations)	& Vaccines required to reduce R0 to 1	& Costs per vaccine	& Total vaccine costs	& Total hospitalised cases	\sym{*}& Total costs for cases \sym{**}	& R0  for simulations after response	& Mean number of cases in the population after action (1000 simulations)	& Total cases over 10 years after action \sym{***}	& Total hospitalised cases after action	& Total costs for cases after action	& Cases reduced due to action	& Savings	& Benefit--cost ratio\\
+    \midrule 2009-2014 & 0.92-1.19 & 0.92	& 13	& 0	& 0 & 2	& 14855	& 0.92	& 13	& 130	& 22	& 148551	& 	& 	&  \\
+2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 20	& 1600000	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1999725	& 127151	& 145295448	& 40.36 \\
+2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 50	& 4000000	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1999725	& 127151	& 145295448	& 24.22 \\
+2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 20	& 4163060	& 57805	& 388554566	& 1	& 116 &	1160	& 197	& 1325532 & 338872	& 387229034	& 70.55 \\
+2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 50	& 10407650	& 57805	& 388554566	& 1	& 116	& 1160	& 197	& 1325532	& 338872	& 387229034	& 33 \\
+2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 20	& 5051220	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 1325532	& 380914	& 435270428	& 68.26 \\
+2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 50	& 12628050	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 1325532	& 380914	& 435270428	& 31.19 \\
+\bottomrule  
+    \end{tabular}%  
+    \begin{tablenotes}\footnotesize  
+        \item \sym{*} Proportion of cases hospitalised 0.17
+        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
+        \item \sym{***} Based on 10 introductions of measles, one per year
+         \end{tablenotes}  
+    \end{threeparttable}  
+  \label{tab:xx}
+\end{sidewaystable}
+
+\normalsize
+
+\begin{sidewaystable}[htbp]  
+  \centering  
+  \begin{raggedright}
+  \caption{Cost benefit analyses using simulated epidemic sizes}  
+   \begin{threeparttable}[b] \tiny
+    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
+    \toprule  
+        Years R0 estimated from &R0 range	&R0 for simulations	&Vaccines required to reduce R0 to 1	&Costs per vaccine	&Total vaccine costs	&Expected number of cases expected over 10 years based on average since 1997	&Total hospitalised cases	\sym{*}&Total costs for cases \sym{**}	&R0  for simulations after response	&Mean number of cases in the population after action (1000 simulations)	&Total cases over 10 years after action \sym{***}	&Total hospitalised cases after action	&Total costs for cases after action	&Cases reduced due to action	&Savings	&Benefit--cost ratio\\
+    \midrule 2009-2014 &0.92-1.19	&0.92	&0	&	&0	&2200	&374	&2513940	&0.92	&13	&130	&22	&148551	&	&	& \\
+2009-2014	&0.92-1.19	&1.19	&80000	&20	&1600000	&2200	&374	&2513940	&1	&175	&1750	&298	&1999725	&450	&514215	&0.14\\
+2009-2014	&0.92-1.19	&1.19	&80000	&50	&4000000	&2200	&374	&2513940	&1	&175	&1750	&298	&1999725	&450	&514215	&0.09\\
+2013-2014	&1.82-2.13	&1.82	&208153	&20	&4163060	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.22\\
+2013-2014	&1.82-2.13	&1.82	&208153	&50	&10407650	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.1\\
+2013-2014	&1.82-2.13	&2.13	&252561	&20	&5051220	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.19\\
+2013-2014	&1.82-2.13	&2.13	&252561	&50	&12628050	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.09\\
+\end{raggedright}
+\bottomrule  
+    \end{tabular}%  
+    \begin{tablenotes}\footnotesize  
+        \item \sym{*} Proportion of cases hospitalised 0.17
+        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
+        \item \sym{***} Based on 10 introductions of measles, one per year
+         \end{tablenotes}  
+    \end{threeparttable}  
+  \label{tab:xx}
+\end{sidewaystable}
+
+\normalsize
 
 \subsection{Cost analysis discussion}
-The results presented here are based on available data.
+The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
 While some of the data are complete and detailed, this is not true of all the data.
@@
 
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin02}.
 This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
@@
 Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
-
-\section{Modelling measles epidemics}
-\label{sec:epidemic_modelling}
-
-A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
-The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
-
-The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
-Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
-The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
-The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
-
-These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
-In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
-All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
-They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
-
-Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
-The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
-
-The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
-This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}.
-If $R_0 > 1$ an epidemic will occur following an introduction of infection.
-The best estimate for measles in New Zealand was $R_0$ = 12.8 \citep{roberts4}.
-The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
-If $R_v < 1$ epidemics are prevented.
-The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
-
-\subsection{Modelling methods}
-
-To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
-To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
-We were required to compute the generation time for measles to do so.
-The generation time is the average time an index case infects others after becoming infected.
-We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
-The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{cases_by_dhb_2009_2014.pdf}
-     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
-     \label{fig:dhbcases}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
-     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
-     \label{fig:outbreaks}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
-     \label{fig:r0}
-\end{figure}
-
-\subsection{Modelling results and discussion}
-
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
-The probability density of the $R_v$ estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
-An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
-
-These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
-The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
-However, sub-clinical and underreporting may lower the estimate.
-The relative contributions of both to our estimates are currently unknown.
-
-\subsection{Summary of modelling}
-\begin{itemize}
-\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
-\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
-\end{itemize}
-
-\subsection{Future modelling}
-Future modelling we aim to perform are:
-\begin{itemize}
-\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
-\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
-\end{itemize}
 

### 2014-08-22 interimreport2.Rnw
removed duplicate figures

@@
 In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection.
-These additional analyses were:
+The additional analyses included in this section are:
 \begin{itemize}
 \item Multivariate modelling to account for confounding within the univariate analyses.
-\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
 \item Analyses of changing risk factors through time during outbreaks.
-\item Modelling transmission chains in the population to understand effective reproduction number.
-\item Development of risk maps to understand measles importation.
 \end{itemize}
 
-<<echo=FALSE>>=
+<<echo=FALSE,include=FALSE,results=hide>>=
 ## regression analyses - measles
@@
 # set wd
- setwd("~/Massey 2014/DHayman_20140627")
+# setwd("~/Massey 2014/DHayman_20140627")
 # read data
@@
 ### denominator data
- setwd("~/Massey_2014/measles/data")
+# setwd("~/Massey_2014/measles/data")
 denom<-read.csv("NZDep2006Denominators.csv",header=T)
@@
 anova(model3,test="F")
+@
 
-##
+<<label=regmodel,include=FALSE>>=
+summary(model3)
+@
 
-par(mfrow=c(2,2))
-hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
-res<-predict(model3)
-plot(exp(res),tpsub$cases,xlab="predictions",ylab='cases',main="Fit",pch=16,col="darkgrey")
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-#abline(lm(exp(res)~tpsub$cases))
-hist(model3$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-
-##
+<<label=regmodelanova,include=FALSE>>=
+anova(model3,test="F")
 @
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
+     \caption{test}
+     \label{fig:regn}
+\end{figure}
 
@@
 
+\section{Modelling measles epidemics}
+\label{sec:epidemic_modelling}
+
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
+The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
+
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
+Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
+The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
+
+These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
+They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
+
+Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
+The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+
+The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}.
+If $R_0 > 1$ an epidemic will occur following an introduction of infection.
+The best estimate for measles in New Zealand was $R_0$ = 12.8 \citep{roberts4}.
+The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
+If $R_v < 1$ epidemics are prevented.
+The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
+
+\subsection{Modelling methods}
+
+To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
+To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
+We were required to compute the generation time for measles to do so.
+The generation time is the average time an index case infects others after becoming infected.
+We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
+The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{cases_by_dhb_2009_2014.pdf}
+     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
+     \label{fig:dhbcases}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
+     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
+     \label{fig:outbreaks}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
+     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \label{fig:r0}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
+     \caption{test}
+     \label{fig:percap}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
+     \caption{test}
+     \label{fig:agecases}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
+     \caption{test}
+     \label{fig:ageofvacc}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
+     \caption{test}
+     \label{fig:numvacc}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
+     \caption{test}
+     \label{fig:naive}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{simulations.pdf}
+     \caption{test}
+     \label{fig:sim}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
+     \caption{test}
+     \label{fig:vaccstat}
+\end{figure}
+
+\subsection{Modelling results and discussion}
+
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The probability density of the $R_v$ estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
+An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+
+These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+However, sub-clinical and underreporting may lower the estimate.
+The relative contributions of both to our estimates are currently unknown.
+
+\subsection{Summary of modelling}
+\begin{itemize}
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
+\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
+ \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
+\end{itemize}
+
+\subsection{Future modelling}
+Future modelling we aim to perform are:
+\begin{itemize}
+\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
+\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
+\end{itemize}
+
 \section{Cost analyses}
@@
 
-The total cost for the 192 publicly funded hospital discharges with a measles primary diagnosis was \$289,694 (Table~\ref{table:hosp}).
-The mean cost per case was \$1,509.
-The mean cost per day of stay in the hospital was \$1,154.
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hosptial was \$550,024 (Table~\ref{table:hosp}).
+The mean cost per case was \$1,877.
+The mean cost per day of stay in the hospital was \$1,170.
 
 <<echo=FALSE>>=
-costtable2<-read.csv("costtable2.csv",header=T)
+costtable2<-read.csv("costtable2v2.csv",header=T)
 @
@@
 \begin{table}
-\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2008--2013}
+\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
 <<results=tex,echo=FALSE>>=
@@
 \label{table:hosp}
+ \centering
+ \begin{tablenotes}
+      \small
+      \item As of 11 July, 2014. NC - not calculated.
+    \end{tablenotes}
 \end{table}
 
-From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before Jan 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
+From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before 1 January 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
 The number of cases by age group is shown in Table~\ref{table:freq}.
@@
 
-The majority (191/192) of the measles cases stayed in the hospital from 0 to 7 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
-Eighty-three of the cases stayed for 0 days and the remaining cases stayed a mean of 2.30 (median 2) days.
-The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2008--2013 appear in Table~\ref{table:cases}.
-The cost for these 192 hospitalisations was \$289,964.
-Of the 292 cases, 87 were female and 105 were male.
-The number of days spent in the hospital was 251, making the mean daily cost \$1,509.
-Of the 251 days, 114 were spent by females and 137 by males.
-The mean hospital stay per case was 1.31 and 1.30 days for females and males, respectively.
-The mean cost per case was \$1,573 and \$1,456 for females and males, respectively.
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 7 19 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
+Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital.
+Only eight cases spent a week or more in the hospital.
+Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days.
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
 
 <<echo=FALSE>>=
-costtable4<-read.csv("costtable4.csv",header=T)
+costtable4<-read.csv("costtable4v2.csv",header=T)
 @
@@
 \begin{table}
-\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2008--2013}
+\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
 <<results=tex,echo=FALSE>>=
@@
 \label{table:cases}
+ \centering
+ \begin{tablenotes}
+      \small
+      \item As of 7 August, 2014.
+    \end{tablenotes}
 \end{table}
@@
 Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
-Results showed an increase of approximately \$358 per day of hospitalisation, \$139 for a female case, and \$1 per year of case diagnosis.
-The cost was found to decrease by approximately \$5 per year of case age.
+Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.
+The cost was \$256 less if the case were a male.
+It increased of approximately \$406 per night of hospitalisationand \$64 per year over the time period of 2000 - 2014.
+The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.
 
 <<echo=FALSE>>=
-costtable5<-read.csv("costtable5.csv",header=T)
+costtable5<-read.csv("costtable5v2.csv",header=T)
 @
@@
 \begin{table}
-\caption{Regression results ($R^{2}_\textsf{adj} = 0.9$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=192$) in New Zealand, 2008 -- 2013}
+\caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
 <<results=tex,echo=FALSE>>=
@@
 
-Wages lost due to measles were calculated based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
+Wages lost due to measles were calculated for the period January 2008 - August 2014.
+Calculations were based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
 If the case were less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
-Total wage lost for the 192 cases and care givers was estimated to be \$166,638.
-This consisted of \$90,620 for the cases and \$76,018 for the care giver, but did not include wage losses for cases under 15 years of age.
+Total wage lost for the 247 cases and care givers was estimated to be \$210,436.
+This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
+Overall, the cost per case from 2008 - 2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
 
-This final figure brings an approximate estimate of \$786,479 for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
+\begin{sidewaystable}[htbp]  
+  \centering  
+  \caption{Cost benefit analyses using simulated epidemic sizes}  
+   \begin{threeparttable}[b] \tiny
+    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
+    \toprule  
+        Years R0 estimated from & R0 range	& R0  for simulations	& Mean # of cases in the population (1000 simulations)	& Vaccines required to reduce R0 to 1	& Costs per vaccine	& Total vaccine costs	& Total hospitalised cases	\sym{*}& Total costs for cases \sym{**}	& R0  for simulations after response	& Mean number of cases in the population after action (1000 simulations)	& Total cases over 10 years after action \sym{***}	& Total hospitalised cases after action	& Total costs for cases after action	& Cases reduced due to action	& Savings	& Benefit--cost ratio\\
+    \midrule 2009-2014 & 0.92-1.19 & 0.92	& 13	& 0	& 0 & 2	& 14855	& 0.92	& 13	& 130	& 22	& 148551	& 	& 	&  \\
+2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 20	& 1600000	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1999725	& 127151	& 145295448	& 40.36 \\
+2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 50	& 4000000	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1999725	& 127151	& 145295448	& 24.22 \\
+2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 20	& 4163060	& 57805	& 388554566	& 1	& 116 &	1160	& 197	& 1325532 & 338872	& 387229034	& 70.55 \\
+2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 50	& 10407650	& 57805	& 388554566	& 1	& 116	& 1160	& 197	& 1325532	& 338872	& 387229034	& 33 \\
+2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 20	& 5051220	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 1325532	& 380914	& 435270428	& 68.26 \\
+2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 50	& 12628050	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 1325532	& 380914	& 435270428	& 31.19 \\
+\bottomrule  
+    \end{tabular}%  
+    \begin{tablenotes}\footnotesize  
+        \item \sym{*} Proportion of cases hospitalised 0.17
+        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
+        \item \sym{***} Based on 10 introductions of measles, one per year
+         \end{tablenotes}  
+    \end{threeparttable}  
+  \label{tab:xx}
+\end{sidewaystable}
+
+\normalsize
+
+\begin{sidewaystable}[htbp]  
+  \centering  
+  \begin{raggedright}
+  \caption{Cost benefit analyses using simulated epidemic sizes}  
+   \begin{threeparttable}[b] \tiny
+    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
+    \toprule  
+        Years R0 estimated from &R0 range	&R0 for simulations	&Vaccines required to reduce R0 to 1	&Costs per vaccine	&Total vaccine costs	&Expected number of cases expected over 10 years based on average since 1997	&Total hospitalised cases	\sym{*}&Total costs for cases \sym{**}	&R0  for simulations after response	&Mean number of cases in the population after action (1000 simulations)	&Total cases over 10 years after action \sym{***}	&Total hospitalised cases after action	&Total costs for cases after action	&Cases reduced due to action	&Savings	&Benefit--cost ratio\\
+    \midrule 2009-2014 &0.92-1.19	&0.92	&0	&	&0	&2200	&374	&2513940	&0.92	&13	&130	&22	&148551	&	&	& \\
+2009-2014	&0.92-1.19	&1.19	&80000	&20	&1600000	&2200	&374	&2513940	&1	&175	&1750	&298	&1999725	&450	&514215	&0.14\\
+2009-2014	&0.92-1.19	&1.19	&80000	&50	&4000000	&2200	&374	&2513940	&1	&175	&1750	&298	&1999725	&450	&514215	&0.09\\
+2013-2014	&1.82-2.13	&1.82	&208153	&20	&4163060	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.22\\
+2013-2014	&1.82-2.13	&1.82	&208153	&50	&10407650	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.1\\
+2013-2014	&1.82-2.13	&2.13	&252561	&20	&5051220	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.19\\
+2013-2014	&1.82-2.13	&2.13	&252561	&50	&12628050	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.09\\
+\end{raggedright}
+\bottomrule  
+    \end{tabular}%  
+    \begin{tablenotes}\footnotesize  
+        \item \sym{*} Proportion of cases hospitalised 0.17
+        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
+        \item \sym{***} Based on 10 introductions of measles, one per year
+         \end{tablenotes}  
+    \end{threeparttable}  
+  \label{tab:xx}
+\end{sidewaystable}
+
+\normalsize
 
 \subsection{Cost analysis discussion}
-The results presented here are based on available data.
+The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
 While some of the data are complete and detailed, this is not true of all the data.

### 2014-08-24 interimreport2.Rnw
Updated measles with hosp costs
Major editing and inc of hosp figures

@@
 A brief review of the history of measles in New Zealand was provided in our previous interim report.
-However, the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \today).
+The current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \today).
 In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
@@
 \end{itemize}
+
+\subsection{Risk analyses methods}
+We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
+Initial analyses of those data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
+Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
+These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
+We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist.
+
+The Otago denominator data provided were not to the same detail as the ESR case data.
+Notably, the denominator age data was categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
+The denominator ethnicity was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided.
+After discussions with Otago we have provided results based on the best available data, though for smaller groups, some results may be unreliable as these are discussed below.
+
+The prioritized ethnicities, therefore, are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African; Other Ethnicity; Residual Categories, though for this report only the first five are used, as these categories covered all the cases.
+Denominator data were classified according to the recommendations of Otago.
+Thus, with the 10 NZDep classes and the age classes above, this led us to have 250 categories.
+Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:agecases}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics (REF)).
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
+     \caption{test}
+     \label{fig:agecases}
+\end{figure}
+
+This large number of categories lead to both zeroinflation and over-dispersion as there were many categories with zero cases in, particularly in the adult age classes.
+Furthermore, initial preliminary analyses, including multi and univariate analyses (not shown) suggested little effect of NZDep and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two, 1--5 and 6--10.
+We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
+By doing so, we reduce the zeroinflation present in the data.
+Subsequent analyses also suggested that the Middle Eastern/Latin American/African category was over-represented (Figure~\ref{fig:percap}).
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
+     \caption{Cases per capita for Prioritized Ethnicities.
+Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
+Points are coloured by Ethnicity.}
+     \label{fig:percap}
+\end{figure}
+
+However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, which have small population sizes, and which there were numberous issues with estimating the denominator data for.
+This category was also an outlier (Figure~\ref{fig:percap}), and thus reducing our ability to make inferences regarding the other classifications and risk factors.
+Thus, we removed the MELAA category for the below analyses.
+In the below analyses we used a quasipoission regression model to account for overdispersion in the data.
+We also account for differences in population sizes by using an offset term, the log(population size).
+Thus, the model structure we used is:
+\begin{equation} \label{eq:quasi}
+   y ~ quasipossion [\lambda]
+ \end{equation}
+
+\begin{equation} \label{eq:reg}
+\lambda = exp[Y(x) + log(P) + \pi(x)]
+  \pi(x) ~ N[0,\sigma ^2]
+  \end{equation}
+
+Where:
+\begin{equation} \label{eq:reg}
+ Y(x) = \beta _1 (x_1)+ \beta _2 (x_2) ...
+  \end{equation}
 
@@
 
+\section{Regression analyses results}
+\label{sub:regression_results}
+The results are:
+
 <<label=regmodel,include=FALSE>>=
@@
 @
+
+And plotted in (Figure~\ref{fig:regn}).
 \begin{figure}
@@
      \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
-     \caption{test}
+     \caption{Top left: distribution of the cases per category; Top right: Cases per Category; Bottom left: Model predictions (x) versus case data (y); and Bottom right: the residuals of the model}
      \label{fig:regn}
@@
 
-\subsection{Risk analysis summary}
-\begin{itemize}
-\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
-\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
-\end{itemize}
+The results of the model suggest that, apart from over representation of MELAA, older people over 25 years of age are significantly less likely to be infected with measles than the very young (0--2 years old), but Pacific peoples are at greater risk than Europeans and others (as can also be seen in Figure 1), but not the 6--17 years among them.
 
-\subsection{Future risk analyses}
-We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above.
-These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data only became available to us on 3 July 2014, provided by the University of Otago.
+...
 
-The following analyses are in progress, for inclusion in later reports:
-\begin{itemize}
-\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using a broader range of years, including modelling the trend in importations.
-\end{itemize}
 Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
@@
 \item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
+\end{itemize}
+
+.
+.
+.
+.
+.
+
+\subsection{Risk analysis summary}
+\begin{itemize}
+\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
+\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
 \end{itemize}
@@
      \label{fig:r0}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
-     \caption{test}
-     \label{fig:percap}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
-     \caption{test}
-     \label{fig:agecases}
 \end{figure}
@@
 
-The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 7 19 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (Figure~\ref{fig:hosp}).
+
+<<echo=FALSE>>=
+hosp<-read.csv("hospital.csv",header=T)
+library(ggplot2)
+p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(p)
+@
+\end{center}
+\caption{Number of cases attending hospital and stay duration}
+\label{fig:hosp}
+\end{figure}
+
 Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital.

### 2014-08-24 hosp.R

@@ -1,0 +1,4 @@
+hosp<-read.csv("hospital.csv",header=T)
+str(hosp)
+library(ggplot2)
+ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")

### 2014-08-25 interimreport2.Rnw
Major revision of interim report
include data, figures, debugged code for mx variate analyses, etc.

@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we report the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated analyses of the cost of previous and the current 2013--2014 outbreak and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
-We provide some preliminary cost-benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
+In this interim report we report the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
+We provide preliminary cost-benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
 Our key findings were:
@@
 A brief review of the history of measles in New Zealand was provided in our previous interim report.
-The current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \today).
-In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
-
-Previous measles analyses, including two in New Zealand by Prof.
-Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.
@@
 In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection.
-The additional analyses included in this section are:
-\begin{itemize}
-\item Multivariate modelling to account for confounding within the univariate analyses.
-\item Analyses of changing risk factors through time during outbreaks.
-\end{itemize}
+The additional analyses included in this section are multivariate modelling to account for confounding within the univariate analyses.
 
@@
 
+<<echo=FALSE,include=FALSE,results=hide>>=
+## get data
+# set wd
+# setwd("~/Massey 2014/DHayman_20140627")
+# read data
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+                 data = data , FUN=sum)
+library(reshape)
+library(reshape2)
+library(plyr)
+library(pscl)
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+head(test)
+dim(test)
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mfrow=c(1,1))
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="Age of cases",breaks=90,include.lowest=TRUE,right=F)
+hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+## plot cases / age class
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+                   data = test , FUN=sum)
+#caseyr
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="Age of vaccination of vaccinated cases")
+abline(v=12,col="blue")
+hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+abline(v=60,col="red")
+legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n")
+legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+require(reshape2)  # this is the library that lets one flatten out data
+require(ggplot2)   # plotting library
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+@
+
+Previous measles analyses, including two in New Zealand by Prof.
+Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+The overall population immunity rate is approximately 89\% and the breakdown of this by age category can be seen in Figure N.
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+plot(pop,xlab="Age",ylab="Population")
+points(naive,pch=16)
+legend("topright",c("Population","Naïve"),pch=c(1,16),bty="n")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+## match cases per age
+AgeInYears<-0:100
+naive<-cbind(AgeInYears,naive)
+colnames(naive)<-c("AgeInYears","Naive")
+naive<-merge(naive,caseyr,by="AgeInYears",all=T)
+colnames(naive)<-c("AgeInYears","Naive","Cases")
+
+dose1=as.factor(round(vac$Dose1Mths/12,0));
+dose1<-summary(dose1)
+dose1<-as.data.frame(dose1)
+dose1$AgeInYears<-rownames(dose1)
+dose2=as.factor(round(na.omit(vac$Dose2Mths/12,0)));
+dose2<-summary(dose2)
+dose2<-as.data.frame(dose2)
+dose2$AgeInYears<-rownames(dose2)
+
+## need to merge vaccination data with data
+## to get years the cases were from
+datav<-merge(vac,data,by="CaseCode",all=T)
+##
+datav$Dose1Mths[which(is.na(datav$Dose1Mths))] <- 9999
+datav$Dose2Mths[which(is.na(datav$Dose2Mths))] <- 9999
+#########################
+datav$RptYear<-as.factor(datav$RptYear)
+datav$SurvWeek<-as.factor(datav$SurvWeek)
+datav$NZDep01<-as.factor(datav$NZDep01)
+datav$NZDep06<-as.factor(datav$NZDep06)
+datav$NZDep13<-as.factor(datav$NZDep13)
+datav$Dose1Mths<-as.factor(datav$Dose1Mths)
+datav$Dose2Mths<-as.factor(datav$Dose2Mths)
+
+#data$AgeInYears<-as.factor(data$AgeInYears)
+datav$EthnicityPrioritised<-as.factor(datav$EthnicityPrioritised)
+
+timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
+                 data = datav , FUN=sum,na.rm=F,na.action=na.pass)
+
+testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+library(plyr)
+testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
+testv$Dose2Mths<-revalue(testv$Dose2Mths, c("9999"=NA));
+testv$Dose1Mths<-as.numeric(testv$Dose1Mths)
+testv$Dose2Mths<-as.numeric(testv$Dose2Mths)
+testv$D2vac<-ifelse(testv$Dose1Mths > -1 & testv$Dose2Mths >= testv$Dose1Mths,testv$Dose2Mths,NA)
+testv$D1vac<-ifelse(testv$Dose1Mths > -1 & is.na(testv$D2vac) == T,testv$Dose1Mths,NA)
+testv$Unvac<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,testv$AgeInYears,NA)
+
+testv$VC<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,0,
+                 ifelse(testv$Dose1Mths >= 0 & is.na(testv$Dose2Mths) == T,1,2))
+AgeVac<-table(testv$VC,testv$AgeInYears)
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+barplot(AgeVac,xlab="Age",col=c("darkgrey","red","orange"),main="Age and vaccination status of cases")
+legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"))
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+AgeInYears<-(as.numeric(rownames(AgeVac)))
+AgeVac<-cbind(AgeVac,AgeInYears)       
+AgeInYears<-(as.numeric(rownames(pop)))
+pop<-cbind(pop,AgeInYears)
+colnames(pop)<-c("Population","AgeInYears")
+AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
+head(AgeV)
+#AgeVac<-t(AgeVac)
+@
 \begin{figure}
@@
 We also account for differences in population sizes by using an offset term, the log(population size).
-Thus, the model structure we used is:
-\begin{equation} \label{eq:quasi}
-   y ~ quasipossion [\lambda]
- \end{equation}
+Thus, we used a quasipoisson error structure and the following linear predictor:
 
 \begin{equation} \label{eq:reg}
-\lambda = exp[Y(x) + log(P) + \pi(x)]
-  \pi(x) ~ N[0,\sigma ^2]
-  \end{equation}
-
-Where:
-\begin{equation} \label{eq:reg}
- Y(x) = \beta _1 (x_1)+ \beta _2 (x_2) ...
+ log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _N (x_N) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
   \end{equation}
@@
 
-testtable<-aggregate( cbind( DiseaseName + as.numeric(SurvWeek)) ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
+testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
+                      ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
                  data = testt , FUN=sum)
@@
 
+ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
+                 data = testtable , FUN=sum)
+@
+
+The numbers of cases per category can be seen in Table N.
+\begin{table}\tiny
+\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(ttyr, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cases}
+\end{table}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
 ### denominator data
@@
 tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
-testtable$merge <- paste(testtable$NZDep, testtable$Age, testtable$Ethnicity)
-testtable <- testtable[testtable$Ethnicity!="None",]
+ttyr$merge <- paste(ttyr$NZDep, ttyr$Age, ttyr$Ethnicity)
+ttyr <- ttyr[ttyr$Ethnicity!="None",]
 
@@
 rownames(cases) <- tp$merge
-cases[testtable$merge,] <- testtable$Cases
+cases[ttyr$merge,] <- ttyr$Cases
 tp$cases <- cases
-
+PerCapita<-(tp$cases/tp$Popn)
+PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita,4))
+colnames(PerCap)[5]<-"Cases"
+colnames(PerCap)[6]<-"Per Capita"
+@
+
+The numbers of cases per category can be seen in Table N.
+\begin{table}\tiny
+\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(PerCap, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:percap}
+\end{table}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
 ## plot this figure?
-hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+#hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
 tp$Ethnicity<- relevel(tp$Ethnicity, "European")
@@
 anova(model2,test="F")
-
-par(mfrow=c(2,2))
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
 hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
 plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
 res<-predict(model2)
 plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
 cor(exp(res),tp$cases)
 cor.test(exp(res),tp$cases)
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
 hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-par(mfrow=c(1,1))
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
 plot(tp$cases/tp$Popn,main="Cases per capita",ylab="Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
@@
        col=c(1:4,6),pch=16,bty="n")
-
-## drop MLA
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\section{Regression analyses results}
+\label{sub:regression_results}
+The results are:
+
+<<label=regmodel,include=FALSE>>=
+summary(model2)
+@
+
+<<label=regmodelanova,include=FALSE>>=
+anova(model2,test="F")
+@
+
+<<echo=FALSE,include=FALSE,results=hide>>=
 tpsub<-tp[!(tp$Ethnicity=="MLA"),]
@@
 
+The results are:
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+res<-predict(model3)
+plot(exp(res),tpsub$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+hist(model3$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+plot(tp$cases/tp$Popn,main="Cases per capita",ylab="Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
+     xlab="Category")
+legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
+       col=c(1:4,6),pch=16,bty="n")
+@
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
 \section{Regression analyses results}
@@
 
-...
+In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected.
+The very young are still most likely to be infections, but older teenagers are more likely to be repreested (Figure N).
+This pattern suggests that imporving vaccination coverage in the young is reducing the burden of measles in those age categories.
+See discussion regarding vaccination protocols.
 
@@
 
-.
-.
-.
-.
-.
-
 \subsection{Risk analysis summary}
@@
 The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
+The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
+     \caption{test}
+     \label{fig:naive}
+\end{figure}
 
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
-     \caption{test}
-     \label{fig:ageofvacc}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
-     \caption{test}
-     \label{fig:numvacc}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
-     \caption{test}
-     \label{fig:naive}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{simulations.pdf}
-     \caption{test}
-     \label{fig:sim}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
-     \caption{test}
-     \label{fig:vaccstat}
-\end{figure}
-
-\subsection{Modelling results and discussion}
+To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we make several simplifying assumptions and the simple relationship between the proportion requiring vaccination, $p_c$, and $R_0$.
+Specifically, we use the simple relationship:
+
+\begin{equation} \label{eq:pc}
+p_c = 1-(1/R0)
+  \end{equation}
+
+We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population, estimated from the national level vaccination coverage data provided by the Ministry of Health for this work.
+Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one.
+
+And discuss the implications for this simplification in the discussion. However, briefly, there are several caveats to our use of this equation:
+\begin {itemize}
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
+\end{itemize}
+
+discussion of simplifications and use of other av:
+\begin {itemize}
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
+\end{itemize}
+
+We then used the proportions of the population that were estimated be require vaccination as estimated above, and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+
+\subsection{Modelling results}
 
@@
 
+To use the results from our modeling exercise to help inform the appropriate measles vaccination coverage.
+Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 to 53\%.
+These figures can be reached in a number of different ways, and these are discussed in the cost-benefit section.
+
 \subsection{Summary of modelling}
@@
 In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current measles outbreak.
+For completeness, we include much of the introduction from the previous interim report, but we have revised some of the figures.
 
@@
 <<echo=FALSE>>=
+#setwd("C:\Users\dtshayma\Documents\GitHub\measles")
 costtable1<-read.csv("costtable1.csv",header=T)
@@
 
+To estimate the benefits from additional vaccinations, as estimated from the above modeling section, we did several things.
+Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ in the estimated naive population from the modeling section, assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
+We used these values of numbers of predicted cases and the cost figures above to estimate the cost of not vaccinating additional populations.
+We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would now be one, though where our estimates for $R_v$ are < 1, we use this.
+We can therefore estimate the number of cases prevents, and the savings made from the additional vaccinations.
+We then use the costs of the catch up vaccination schemes and the costs of the expected measles-related costs due to constant introduction of measles and the savings from reduced measles cases to work out the benefit to cost ratio.
+For our benefit to cost ratio to be beneficial, this must be greater than 1.
+A value less than one, suggests the costs are higher than the economic benefits.
+Lastly, because there is a discounting effect with time, we use an 8\% discount per year for the costs, as is common for health care discounting, and we calculate these over a 10 year time frame, using a discounting such that:
+\begin{equation} \label{eq:reg}
+discount per year = cost * 1/1.08^(year-1))
+  \end{equation}
+
+Our estimates for the numbers of measles cases are highly variable, though typically in our model, if $R_v$ were greater than one, either the measles outbreak would take off and all the New Zealand population would become infected, or not, and we used the average of 1000 simulations to estimate the mean of this.
+However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes.
+We therefore also compared our results to those using the mean number of measles cases seen in New Zealand per yea, since 1997.
+This was 220 cases per year.
+
+The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
+     \caption{test}
+     \label{fig:numvacc}
+\end{figure}
+
+The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure N.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{simulations.pdf}
+     \caption{test}
+     \label{fig:sim}
+\end{figure}
+
+The benefit-cost results are in Table N and N.
+
 \begin{sidewaystable}[htbp]  
@@
         Years R0 estimated from & R0 range	& R0  for simulations	& Mean # of cases in the population (1000 simulations)	& Vaccines required to reduce R0 to 1	& Costs per vaccine	& Total vaccine costs	& Total hospitalised cases	\sym{*}& Total costs for cases \sym{**}	& R0  for simulations after response	& Mean number of cases in the population after action (1000 simulations)	& Total cases over 10 years after action \sym{***}	& Total hospitalised cases after action	& Total costs for cases after action	& Cases reduced due to action	& Savings	& Benefit--cost ratio\\
-    \midrule 2009-2014 & 0.92-1.19 & 0.92	& 13	& 0	& 0 & 2	& 14855	& 0.92	& 13	& 130	& 22	& 148551	& 	& 	&  \\
-2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 20	& 1600000	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1999725	& 127151	& 145295448	& 40.36 \\
-2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 50	& 4000000	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1999725	& 127151	& 145295448	& 24.22 \\
-2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 20	& 4163060	& 57805	& 388554566	& 1	& 116 &	1160	& 197	& 1325532 & 338872	& 387229034	& 70.55 \\
-2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 50	& 10407650	& 57805	& 388554566	& 1	& 116	& 1160	& 197	& 1325532	& 338872	& 387229034	& 33 \\
-2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 20	& 5051220	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 1325532	& 380914	& 435270428	& 68.26 \\
-2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 50	& 12628050	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 1325532	& 380914	& 435270428	& 31.19 \\
+    \midrule 
+    2009-2014 & 0.92-1.19 & 0.92	& 13	& 0	& 0 & 2	& 10765	& 0.92	& 13	& 130	& 22	& 107653	& 	& 	&  \\
+2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 20	& 1159502	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1449178& 127151	& 105293982	& 40.36 \\
+2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 50	& 2898755	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1449178	& 127151	& 105293982	& 24.22 \\
+2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 20	& 3016923	& 57805	& 388554566	& 1	& 116 &	1160	& 197	& 960598 & 338872	& 280620541	& 70.55 \\
+2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 50	& 7542307	& 57805	& 388554566	& 1	& 116	& 1160	& 197	& 960598	& 338872	& 280620541	& 33 \\
+2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 20	& 3660563	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 960598	& 380914	& 315435600 & 68.26 \\
+2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 50	& 9151406	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 960598	& 380914	& 315435600	& 31.19 \\
 \bottomrule  
@@
         Years R0 estimated from &R0 range	&R0 for simulations	&Vaccines required to reduce R0 to 1	&Costs per vaccine	&Total vaccine costs	&Expected number of cases expected over 10 years based on average since 1997	&Total hospitalised cases	\sym{*}&Total costs for cases \sym{**}	&R0  for simulations after response	&Mean number of cases in the population after action (1000 simulations)	&Total cases over 10 years after action \sym{***}	&Total hospitalised cases after action	&Total costs for cases after action	&Cases reduced due to action	&Savings	&Benefit--cost ratio\\
-    \midrule 2009-2014 &0.92-1.19	&0.92	&0	&	&0	&2200	&374	&2513940	&0.92	&13	&130	&22	&148551	&	&	& \\
-2009-2014	&0.92-1.19	&1.19	&80000	&20	&1600000	&2200	&374	&2513940	&1	&175	&1750	&298	&1999725	&450	&514215	&0.14\\
-2009-2014	&0.92-1.19	&1.19	&80000	&50	&4000000	&2200	&374	&2513940	&1	&175	&1750	&298	&1999725	&450	&514215	&0.09\\
-2013-2014	&1.82-2.13	&1.82	&208153	&20	&4163060	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.22\\
-2013-2014	&1.82-2.13	&1.82	&208153	&50	&10407650	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.1\\
-2013-2014	&1.82-2.13	&2.13	&252561	&20	&5051220	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.19\\
-2013-2014	&1.82-2.13	&2.13	&252561	&50	&12628050	&2200	&374	&2513940	&1	&116	&1160	&197	&1325532	&1040	&1188408	&0.09\\
+    \midrule 
+2009-2014 &0.92-1.19	&0.92	&0	&	&0	&2200	&374	&1821824	&0.92	&13	&130	&22	&107653	&	&	& \\
+2009-2014	&0.92-1.19	&1.19	&80000	&20	& 1159502	&2200	&374	&2513940	&1	&175	&1750	&298	&1449178	&450	&372646	&0.14\\
+2009-2014	&0.92-1.19	&1.19	&80000	&50	& 2898755  &2200	&374	&2513940	&1	&175	&1750	&298	&1449178	&450	&372646 &0.09\\
+2013-2014	&1.82-2.13	&1.82	&208153	&20	& 3016923	&2200	&374	&2513940	&1	&116	&1160	&197	&960598	&1040	&861226	&0.22\\
+2013-2014	&1.82-2.13	&1.82	&208153	&50	& 7542307 &2200	&374	&2513940	&1	&116	&1160	&197	&960598 &1040	&861226  &0.1\\
+2013-2014	&1.82-2.13	&2.13	&252561	&20	& 3660563	&2200	&374	&2513940	&1	&116	&1160	&197	&960598	&1040	&861226  &0.19\\
+2013-2014	&1.82-2.13	&2.13	&252561	&50	& 9151406	&2200	&374	&2513940	&1	&116	&1160	&197	&960598	&1040	&861226  &0.09\\
 \end{raggedright}
@@
 \normalsize
+and the proportions of each age category shown in N.
+
+However, it is worth noting that vaccination strategies that target the very young (<1) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure ) were vaccinates with a signle vaccine at a very young age (Figure N).
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
+     \caption{test}
+     \label{fig:vaccstat}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
+     \caption{test}
+     \label{fig:ageofvacc}
+\end{figure}
 

### 2014-08-25 interimreport2.tex
removed duplicate figures
Major revision of interim report
include data, figures, debugged code for mx variate analyses, etc.

@@
 A brief review of the history of measles in New Zealand was provided in our previous interim report.
-However, the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \today).
-In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
+In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.
+
+\section{Risk analysis update}
+\label{sub:risk_analyses}
+
+A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups.
+Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
+In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection.
+The additional analyses included in this section are multivariate modelling to account for confounding within the univariate analyses.
+
+\subsection{Risk analyses methods}
+We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
+Initial analyses of those data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
+Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
+These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
+We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist.
+
+The Otago denominator data provided were not to the same detail as the ESR case data.
+Notably, the denominator age data was categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
+The denominator ethnicity was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided.
+After discussions with Otago we have provided results based on the best available data, though for smaller groups, some results may be unreliable as these are discussed below.
+
+The prioritized ethnicities, therefore, are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African; Other Ethnicity; Residual Categories, though for this report only the first five are used, as these categories covered all the cases.
+Denominator data were classified according to the recommendations of Otago.
+Thus, with the 10 NZDep classes and the age classes above, this led us to have 250 categories.
+Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:agecases}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics (REF)).
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-002}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-004}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
 
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.
-
-\section{Risk analysis update}
-\label{sub:risk_analyses}
-
-A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups.
-Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection.
-The additional analyses included in this section are:
-\begin{itemize}
-\item Multivariate modelling to account for confounding within the univariate analyses.
-\item Analyses of changing risk factors through time during outbreaks.
-\end{itemize}
+The overall population immunity rate is approximately 89\% and the breakdown of this by age category can be seen in Figure N.
+
+\begin{figure}[h!]
+\begin{center}
+\begin{Schunk}
+\begin{Soutput}
+[1] NA
+\end{Soutput}
+\end{Schunk}
+\includegraphics{interimreport2-006}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-008}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
+     \caption{test}
+     \label{fig:agecases}
+\end{figure}
+
+This large number of categories lead to both zeroinflation and over-dispersion as there were many categories with zero cases in, particularly in the adult age classes.
+Furthermore, initial preliminary analyses, including multi and univariate analyses (not shown) suggested little effect of NZDep and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two, 1--5 and 6--10.
+We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
+By doing so, we reduce the zeroinflation present in the data.
+Subsequent analyses also suggested that the Middle Eastern/Latin American/African category was over-represented (Figure~\ref{fig:percap}).
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
+     \caption{Cases per capita for Prioritized Ethnicities.
+Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
+Points are coloured by Ethnicity.}
+     \label{fig:percap}
+\end{figure}
+
+However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, which have small population sizes, and which there were numberous issues with estimating the denominator data for.
+This category was also an outlier (Figure~\ref{fig:percap}), and thus reducing our ability to make inferences regarding the other classifications and risk factors.
+Thus, we removed the MELAA category for the below analyses.
+In the below analyses we used a quasipoission regression model to account for overdispersion in the data.
+We also account for differences in population sizes by using an offset term, the log(population size).
+Thus, we used a quasipoisson error structure and the following linear predictor:
+
+\begin{equation} \label{eq:reg}
+ log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _N (x_N) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
+  \end{equation}
+
+The numbers of cases per category can be seen in Table N.
+\begin{table}\tiny
+\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllr}
+\hline\hline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
+\hline
+1-5&0-2&Asian&$ 11$\tabularnewline
+6-10&0-2&Asian&$  8$\tabularnewline
+1-5&3-5&Asian&$  1$\tabularnewline
+1-5&6-17&Asian&$ 11$\tabularnewline
+6-10&6-17&Asian&$  5$\tabularnewline
+1-5&18-24&Asian&$  3$\tabularnewline
+6-10&18-24&Asian&$  5$\tabularnewline
+1-5&25+&Asian&$ 10$\tabularnewline
+6-10&25+&Asian&$ 13$\tabularnewline
+1-5&0-2&European&$ 83$\tabularnewline
+6-10&0-2&European&$ 64$\tabularnewline
+1-5&3-5&European&$ 42$\tabularnewline
+6-10&3-5&European&$ 17$\tabularnewline
+1-5&6-17&European&$219$\tabularnewline
+6-10&6-17&European&$ 80$\tabularnewline
+1-5&18-24&European&$ 34$\tabularnewline
+6-10&18-24&European&$ 36$\tabularnewline
+1-5&25+&European&$ 78$\tabularnewline
+6-10&25+&European&$ 51$\tabularnewline
+1-5&0-2&Maori&$ 18$\tabularnewline
+6-10&0-2&Maori&$ 48$\tabularnewline
+1-5&3-5&Maori&$  7$\tabularnewline
+6-10&3-5&Maori&$ 11$\tabularnewline
+1-5&6-17&Maori&$ 19$\tabularnewline
+6-10&6-17&Maori&$ 92$\tabularnewline
+1-5&18-24&Maori&$  5$\tabularnewline
+6-10&18-24&Maori&$  8$\tabularnewline
+1-5&25+&Maori&$  2$\tabularnewline
+6-10&25+&Maori&$  6$\tabularnewline
+6-10&0-2&MLA&$  3$\tabularnewline
+6-10&3-5&MLA&$  1$\tabularnewline
+6-10&6-17&MLA&$  1$\tabularnewline
+6-10&18-24&MLA&$  2$\tabularnewline
+6-10&25+&MLA&$  6$\tabularnewline
+1-5&0-2&Pacific&$  5$\tabularnewline
+6-10&0-2&Pacific&$ 58$\tabularnewline
+6-10&3-5&Pacific&$  3$\tabularnewline
+1-5&6-17&Pacific&$  5$\tabularnewline
+6-10&6-17&Pacific&$ 22$\tabularnewline
+1-5&18-24&Pacific&$  1$\tabularnewline
+6-10&18-24&Pacific&$  8$\tabularnewline
+1-5&25+&Pacific&$  2$\tabularnewline
+6-10&25+&Pacific&$ 11$\tabularnewline
+1-5&0-2&None&$  3$\tabularnewline
+1-5&3-5&None&$  1$\tabularnewline
+1-5&6-17&None&$  3$\tabularnewline
+6-10&6-17&None&$  4$\tabularnewline
+1-5&18-24&None&$  2$\tabularnewline
+6-10&18-24&None&$  1$\tabularnewline
+1-5&25+&None&$  5$\tabularnewline
+6-10&25+&None&$  3$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cases}
+\end{table}
+
+The numbers of cases per category can be seen in Table N.
+\begin{table}\tiny
+\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllrrr}
+\hline\hline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Popn}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per Capita}\tabularnewline
+\hline
+1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
+6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
+1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
+6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
+1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
+6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
+1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
+6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
+1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
+6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
+1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
+6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
+1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
+6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
+1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
+6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
+1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
+6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
+1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
+6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
+1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
+6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
+1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
+6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
+1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
+6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
+1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
+6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
+1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
+6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
+1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
+6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
+1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
+6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
+1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
+6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
+1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
+6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
+1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
+6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
+1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
+6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
+1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
+6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
+1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
+6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
+1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
+6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
+1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
+6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:percap}
+\end{table}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-015}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-016}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-017}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-019}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-020}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\section{Regression analyses results}
+\label{sub:regression_results}
+The results are:
+
+\begin{Schunk}
+\begin{Sinput}
+> summary(model2)
+\end{Sinput}
+\begin{Soutput}
+Call:
+glm(formula = cases ~ Age + Ethnicity + NZDep + Age:Ethnicity + 
+    Ethnicity:NZDep + offset(log(Popn)), family = "quasipoisson", 
+    data = tp)
+
+Deviance Residuals: 
+     Min        1Q    Median        3Q       Max  
+-1.98437  -0.58185  -0.00005   0.45759   1.47521
+
+Coefficients:
+                             Estimate Std. Error t value Pr(>|t|)    
+(Intercept)                -6.422e+00  1.016e-01 -63.219  < 2e-16 ***
+Age3-5                     -9.129e-01  1.794e-01  -5.088 5.62e-05 ***
+Age6-17                    -7.653e-01  1.173e-01  -6.524 2.33e-06 ***
+Age18-24                   -1.495e+00  1.692e-01  -8.836 2.43e-08 ***
+Age25+                     -2.953e+00  1.405e-01 -21.024 4.16e-15 ***
+EthnicityAsian              2.147e-03  3.167e-01   0.007  0.99466    
+EthnicityMaori             -6.061e-02  2.267e-01  -0.267  0.79191    
+EthnicityMLA               -1.820e+01  2.572e+03  -0.007  0.99442    
+EthnicityPacific            6.950e-01  3.541e-01   1.963  0.06374 .  
+NZDep6-10                  -3.339e-01  9.205e-02  -3.627  0.00168 ** 
+Age3-5:EthnicityAsian      -2.032e+00  1.208e+00  -1.682  0.10813    
+Age6-17:EthnicityAsian     -1.003e+00  4.125e-01  -2.432  0.02455 *  
+Age18-24:EthnicityAsian    -8.417e-01  5.194e-01  -1.620  0.12079    
+Age25+:EthnicityAsian       4.237e-01  3.873e-01   1.094  0.28707    
+Age3-5:EthnicityMaori      -3.864e-01  3.578e-01  -1.080  0.29305    
+Age6-17:EthnicityMaori     -7.940e-02  2.157e-01  -0.368  0.71660    
+Age18-24:EthnicityMaori    -5.954e-01  3.917e-01  -1.520  0.14417    
+Age25+:EthnicityMaori      -1.039e+00  4.579e-01  -2.270  0.03443 *  
+Age3-5:EthnicityMLA        -1.857e-01  1.356e+00  -0.137  0.89245    
+Age6-17:EthnicityMLA       -1.591e+00  1.349e+00  -1.179  0.25212    
+Age18-24:EthnicityMLA       2.201e-01  1.076e+00   0.205  0.84001    
+Age25+:EthnicityMLA         1.565e+00  8.351e-01   1.874  0.07556 .  
+Age3-5:EthnicityPacific    -2.132e+00  7.110e-01  -2.998  0.00711 ** 
+Age6-17:EthnicityPacific   -1.448e+00  2.924e-01  -4.953 7.66e-05 ***
+Age18-24:EthnicityPacific  -9.854e-01  4.481e-01  -2.199  0.03979 *  
+Age25+:EthnicityPacific    -5.958e-01  3.815e-01  -1.562  0.13406    
+EthnicityAsian:NZDep6-10    1.330e-01  3.006e-01   0.442  0.66291    
+EthnicityMaori:NZDep6-10    4.296e-01  2.080e-01   2.065  0.05210 .  
+EthnicityMLA:NZDep6-10      1.890e+01  2.572e+03   0.007  0.99421    
+EthnicityPacific:NZDep6-10  6.070e-01  3.551e-01   1.709  0.10283    
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for quasipoisson family taken to be 1.355392)
+
+    Null deviance: 1473.124  on 49  degrees of freedom
+Residual deviance:   27.468  on 20  degrees of freedom
+AIC: NA
+
+Number of Fisher Scoring iterations: 16
+\end{Soutput}
+\end{Schunk}
+
+\begin{Schunk}
+\begin{Sinput}
+> anova(model2,test="F")
+\end{Sinput}
+\begin{Soutput}
+Analysis of Deviance Table
+
+Model: quasipoisson, link: log
+
+Response: cases
+
+Terms added sequentially (first to last)
+
+                Df Deviance Resid. Df Resid. Dev        F    Pr(>F)    
+NULL                               49    1473.12                       
+Age              4  1297.52        45     175.60 239.3263 < 2.2e-16 ***
+Ethnicity        4    20.21        41     155.39   3.7285  0.020056 *  
+NZDep            1     8.62        40     146.77   6.3573  0.020285 *  
+Age:Ethnicity   16    94.49        24      52.28   4.3569  0.001223 ** 
+Ethnicity:NZDep  4    24.82        20      27.47   4.5771  0.008694 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{Soutput}
+\end{Schunk}
+
+The results are:
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-024}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-025}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-026}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-028}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-029}
+\end{center}
+\caption{test}
+\label{fig:test}
+\end{figure}
+
+\section{Regression analyses results}
+\label{sub:regression_results}
+The results are:
 
@@
      Min        1Q    Median        3Q       Max  
--6.1174  -1.1562   0.0000   0.8599   6.7625
+-2.76316  -0.52409   0.04736   0.49883   1.92918
 
@@
                           Estimate Std. Error t value Pr(>|t|)    
-(Intercept)               -8.268e+00  8.641e-01  -9.569 1.06e-08 ***
-Age3-5                     6.931e-01  1.053e+00   0.658 0.518322    
-Age6-17                    7.176e-01  9.064e-01   0.792 0.438300    
-Age18-24                  -5.001e-01  1.149e+00  -0.435 0.668367    
-Age25+                    -2.620e+00  1.158e+00  -2.262 0.035602 *  
-EthnicityAsian             2.167e+00  1.203e+00   1.801 0.087576 .  
-EthnicityMaori             1.772e+00  1.061e+00   1.671 0.111064    
-EthnicityPacific           3.871e+00  9.405e-01   4.117 0.000587 ***
-NZDep6-10                 -4.878e-01  2.608e-01  -1.870 0.076939 .  
-Age3-5:EthnicityAsian     -2.431e-15  1.473e+00   0.000 1.000000    
-Age6-17:EthnicityAsian    -2.473e+00  1.530e+00  -1.616 0.122649    
-Age18-24:EthnicityAsian   -7.910e-01  1.622e+00  -0.488 0.631293    
-Age25+:EthnicityAsian     -6.025e-01  1.792e+00  -0.336 0.740386    
-Age3-5:EthnicityMaori     -2.845e+00  2.177e+00  -1.307 0.206941    
-Age6-17:EthnicityMaori    -2.102e-01  1.121e+00  -0.187 0.853294    
-Age18-24:EthnicityMaori   -8.881e-01  1.630e+00  -0.545 0.592102    
-Age25+:EthnicityMaori     -1.090e+00  2.011e+00  -0.542 0.594088    
-Age3-5:EthnicityPacific   -4.423e+00  2.581e+00  -1.714 0.102866    
-Age6-17:EthnicityPacific  -2.253e+00  1.051e+00  -2.143 0.045264 *  
-Age18-24:EthnicityPacific -8.321e-01  1.318e+00  -0.631 0.535272    
-Age25+:EthnicityPacific   -2.718e-01  1.338e+00  -0.203 0.841210    
+(Intercept)               -6.46738    0.11483 -56.324  < 2e-16 ***
+Age3-5                    -0.91290    0.20539  -4.445 0.000278 ***
+Age6-17                   -0.76175    0.13427  -5.673 1.81e-05 ***
+Age18-24                  -1.50501    0.19366  -7.772 2.57e-07 ***
+Age25+                    -2.95073    0.16079 -18.352 1.51e-13 ***
+EthnicityAsian             0.05304    0.32498   0.163 0.872072    
+EthnicityMaori             0.21245    0.19935   1.066 0.299919    
+EthnicityPacific           1.16032    0.20429   5.680 1.78e-05 ***
+NZDep6-10                 -0.21192    0.08548  -2.479 0.022713 *  
+Age3-5:EthnicityAsian     -2.03154    1.38265  -1.469 0.158112    
+Age6-17:EthnicityAsian    -1.00742    0.47171  -2.136 0.045940 *  
+Age18-24:EthnicityAsian   -0.83115    0.59409  -1.399 0.177917    
+Age25+:EthnicityAsian      0.42108    0.44334   0.950 0.354143    
+Age3-5:EthnicityMaori     -0.38639    0.40959  -0.943 0.357344    
+Age6-17:EthnicityMaori    -0.08552    0.24685  -0.346 0.732812    
+Age18-24:EthnicityMaori   -0.58278    0.44835  -1.300 0.209209    
+Age25+:EthnicityMaori     -1.04821    0.52418  -2.000 0.060036 .  
+Age3-5:EthnicityPacific   -2.13163    0.81388  -2.619 0.016882 *  
+Age6-17:EthnicityPacific  -1.45411    0.33466  -4.345 0.000349 ***
+Age18-24:EthnicityPacific -0.98266    0.51289  -1.916 0.070546 .  
+Age25+:EthnicityPacific   -0.61270    0.43665  -1.403 0.176692    
 ---
@@
 
-(Dispersion parameter for quasipoisson family taken to be 16.26672)
-
-    Null deviance: 3097.24  on 39  degrees of freedom
-Residual deviance:  291.05  on 19  degrees of freedom
+(Dispersion parameter for quasipoisson family taken to be 1.776063)
+
+    Null deviance: 1453.62  on 39  degrees of freedom
+Residual deviance:   36.83  on 19  degrees of freedom
 AIC: NA
 
-Number of Fisher Scoring iterations: 6
+Number of Fisher Scoring iterations: 5
 \end{Soutput}
@@
               Df Deviance Resid. Df Resid. Dev        F   Pr(>F)    
-NULL                             39    3097.24                      
-Age            4  1573.25        35    1523.98 24.1790 3.142e-07 ***
-Ethnicity      3   738.20        32     785.78 15.1271 2.853e-05 ***
-NZDep          1    55.39        31     730.39  3.4051   0.08064 .  
-Age:Ethnicity 12   439.34        19     291.05  2.2507   0.05519 .  
+NULL                             39    1453.62                      
+Age            4  1304.20        35     149.43 183.5795 6.73e-15 ***
+Ethnicity      3    20.00        32     129.43   3.7529 0.028500 *  
+NZDep          1    10.70        31     118.74   6.0236 0.023932 *  
+Age:Ethnicity 12    81.91        19      36.83   3.8430 0.004459 ** 
 ---
@@
 \end{Schunk}
+
+And plotted in (Figure~\ref{fig:regn}).
 \begin{figure}
@@
 
-\subsection{Risk analysis summary}
-\begin{itemize}
-\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
-\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
-\end{itemize}
-
-\subsection{Future risk analyses}
-We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above.
-These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data only became available to us on 3 July 2014, provided by the University of Otago.
-
-The following analyses are in progress, for inclusion in later reports:
-\begin{itemize}
-\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using a broader range of years, including modelling the trend in importations.
-\end{itemize}
+The results of the model suggest that, apart from over representation of MELAA, older people over 25 years of age are significantly less likely to be infected with measles than the very young (0--2 years old), but Pacific peoples are at greater risk than Europeans and others (as can also be seen in Figure 1), but not the 6--17 years among them.
+
+In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected.
+The very young are still most likely to be infections, but older teenagers are more likely to be repreested (Figure N).
+This pattern suggests that imporving vaccination coverage in the young is reducing the burden of measles in those age categories.
+See discussion regarding vaccination protocols.
+
 Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
@@
 
+\subsection{Risk analysis summary}
+\begin{itemize}
+\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
+\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
+\end{itemize}
+
 \section{Modelling measles epidemics}
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
-     \caption{test}
-     \label{fig:percap}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
-     \caption{test}
-     \label{fig:percap}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
-     \caption{test}
-     \label{fig:percap}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
-     \caption{test}
-     \label{fig:agecases}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
-     \caption{test}
-     \label{fig:ageofvacc}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
-     \caption{test}
-     \label{fig:numvacc}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
-     \caption{test}
-     \label{fig:naive}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{simulations.pdf}
-     \caption{test}
-     \label{fig:sim}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
-     \caption{test}
-     \label{fig:vaccstat}
-\end{figure}
-
-\subsection{Modelling results and discussion}
+To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we make several simplifying assumptions and the simple relationship between the proportion requiring vaccination, $p_c$, and $R_0$.
+Specifically, we use the simple relationship:
+
+\begin{equation} \label{eq:pc}
+p_c = 1-(1/R0)
+  \end{equation}
+
+We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population, estimated from the national level vaccination coverage data provided by the Ministry of Health for this work.
+Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one.
+
+And discuss the implications for this simplification in the discussion. However, briefly, there are several caveats to our use of this equation:
+\begin {itemize}
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
+\end{itemize}
+
+discussion of simplifications and use of other av:
+\begin {itemize}
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
+\end{itemize}
+
+We then used the proportions of the population that were estimated be require vaccination as estimated above, and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+
+\subsection{Modelling results}
 
@@
 
-The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 7 19 days, with the exception of a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213.
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (Figure~\ref{fig:hosp}).
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-039}
+\end{center}
+\caption{Number of cases attending hospital and stay duration}
+\label{fig:hosp}
+\end{figure}
+
 Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital.

### 2014-08-25 naive_cor_cases.R

@@ -1,0 +1,17 @@
+ttage<-aggregate( cbind(Cases) ~ Age, 
+               data = testtable , FUN=sum)
+
+ttage<-aggregate( cbind( DiseaseName ) ~ AgeInYears,# + Dose1Mths + Dose2Mths, 
+                 data = data , FUN=sum)
+ttage
+naive
+colnames(naive)<-"naive"
+aget<-1:100
+aget
+aget<-as.data.frame(aget)
+colnames(aget)<-"AgeInYears"
+caset<-merge(aget,ttage,by="AgeInYears",all="T")
+caset[is.na(caset)] <- 0
+caset<-cbind(caset,naive)
+cor(caset$DiseaseName,caset$naive)
+cor(caset$DiseaseName[2:100],caset$naive[2:100])

### 2014-08-25 interimreport2.Rnw
major revisions of interim report

@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we report the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
-We provide preliminary cost-benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
+In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
+We provide preliminary cost--benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
 Our key findings were:
@@
 
-As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
+As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission.
 A brief review of the history of measles in New Zealand was provided in our previous interim report.
-In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.
+In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the cost--benefit analyses for vaccination dependent on differing scenarios.
 
@@
 Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we believed additional analyses could further inform the understanding of risk from measles infection.
+In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
 The additional analyses included in this section are multivariate modelling to account for confounding within the univariate analyses.
@@
 
-The Otago denominator data provided were not to the same detail as the ESR case data.
+The University of Otago denominator data provided were not to the same detail as the ESR case data.
 Notably, the denominator age data was categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
-The denominator ethnicity was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided.
-After discussions with Otago we have provided results based on the best available data, though for smaller groups, some results may be unreliable as these are discussed below.
-
-The prioritized ethnicities, therefore, are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African; Other Ethnicity; Residual Categories, though for this report only the first five are used, as these categories covered all the cases.
-Denominator data were classified according to the recommendations of Otago.
-Thus, with the 10 NZDep classes and the age classes above, this led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:agecases}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics (REF)).
+The ethnicity denominator data was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes.
+After discussions with the University Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.
+
+With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, this led us to have 250 categories.
+Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
 
@@
 par(mfrow=c(1,1))
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="Age of cases",breaks=90,include.lowest=TRUE,right=F)
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F)
 hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
@@
 \end{center}
-\caption{test}
-\label{fig:test}
+\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+\label{fig:ageinyears}
 \end{figure}
@@
 @
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="Age of vaccination of vaccinated cases")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n")
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
 
@@
 @
-
-Previous measles analyses, including two in New Zealand by Prof.
-Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-The overall population immunity rate is approximately 89\% and the breakdown of this by age category can be seen in Figure N.
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-plot(pop,xlab="Age",ylab="Population")
-points(naive,pch=16)
-legend("topright",c("Population","Naïve"),pch=c(1,16),bty="n")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
 
@@
 
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-barplot(AgeVac,xlab="Age",col=c("darkgrey","red","orange"),main="Age and vaccination status of cases")
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"))
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
 AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
-head(AgeV)
+# head(AgeV)
 #AgeVac<-t(AgeVac)
 @
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
-     \caption{test}
-     \label{fig:agecases}
-\end{figure}
-
-This large number of categories lead to both zeroinflation and over-dispersion as there were many categories with zero cases in, particularly in the adult age classes.
+
+This large number of categories, some with population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
 Furthermore, initial preliminary analyses, including multi and univariate analyses (not shown) suggested little effect of NZDep and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two, 1--5 and 6--10.
@@
 By doing so, we reduce the zeroinflation present in the data.
-Subsequent analyses also suggested that the Middle Eastern/Latin American/African category was over-represented (Figure~\ref{fig:percap}).
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
-     \caption{Cases per capita for Prioritized Ethnicities.
-Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
-Points are coloured by Ethnicity.}
-     \label{fig:percap}
-\end{figure}
-
-However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, which have small population sizes, and which there were numberous issues with estimating the denominator data for.
-This category was also an outlier (Figure~\ref{fig:percap}), and thus reducing our ability to make inferences regarding the other classifications and risk factors.
-Thus, we removed the MELAA category for the below analyses.
-In the below analyses we used a quasipoission regression model to account for overdispersion in the data.
-We also account for differences in population sizes by using an offset term, the log(population size).
-Thus, we used a quasipoisson error structure and the following linear predictor:
-
-\begin{equation} \label{eq:reg}
- log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _N (x_N) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
-  \end{equation}
-
+
+The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:cases}).
 <<echo=FALSE,include=FALSE,results=hide>>=
 ## regression analyses - measles
-## get data
-# set wd
 # setwd("~/Massey 2014/DHayman_20140627")
@@
 testtable$NZDep<-tnzd
-head(testtable)
+# head(testtable)
 tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
@@
 
-The numbers of cases per category can be seen in Table N.
+The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:cases}.
+
 \begin{table}\tiny
@@
 PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita,4))
+colnames(PerCap)[4]<-"Population"
 colnames(PerCap)[5]<-"Cases"
@@
 
-The numbers of cases per category can be seen in Table N.
+Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over- or under-represented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{fig:}), leading to very large standard error in regression analyses.
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
+plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
+     xlab="Category")
+legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
+       col=c(1:4,6),pch=16,bty="n")
+@
+\end{center}
+\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
+\label{fig:percap}
+\end{figure}
+
 \begin{table}\tiny
@@
 
+However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes, and there several issues with estimating the denominator data for these group (University of Otago, personal communication).
+Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007.
+
+For all our statistical analyses (including those above not shown) we used a Poisson error strutuce, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model.
+We also account for differences in population sizes by using an offset term, the log(population size).
+We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms.
+Thus, the final model that remained with all significat interaction terms had the following linear predictor:
+
+\begin{equation} \label{eq:reg}
+ log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
+  \end{equation}
+Where $y$ = cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.
+
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
 @
-
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+%@
+%\end{center}
+%\caption{test}
+%\label{fig:test}
+%\end{figure}
+
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+%@
+%\end{center}
+%\caption{Cases per capita for Prioritized Ethnicities.
+Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
+Points are coloured by Ethnicity.}
+%\label{fig:test}
+%\end{figure}
+%
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%res<-predict(model2)
+%plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
+%@
+%\end{center}
+%\caption{test}
+%\label{fig:test}
+%\end{figure}
+
+%<<echo=FALSE,include=FALSE,results=hide>>=
+%cor(exp(res),tp$cases)
+%cor.test(exp(res),tp$cases)
+%@
+%
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
+%@
+%\end{center}
+%\caption{test}
+%\label{fig:test}
+%\end{figure}
+
+%\section{Regression analyses results}
+%\label{sub:regression_results}
+%The results are:
+
+%<<label=regmodel,include=FALSE>>=
+%summary(model2)
+%@
+
+%<<label=regmodelanova,include=FALSE>>=
+%anova(model2,test="F")
+%@
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+tpsub<-tp[!(tp$Ethnicity=="MLA"),]
+
+#hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
+
+model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
+summary(model)
+
+model1<-update(model,~.-Age:Ethnicity:NZDep)
+summary(model1)
+anova(model1,test="F")
+
+model2<-update(model1,~.-Ethnicity:NZDep)
+summary(model2)
+anova(model2,test="F")
+
+model3<-update(model2,~.-Age:NZDep)
+summary(model3)
+anova(model3,test="F")
+@
+
+\subsection{Regression analyses results}
+\label{sub:regression_results}
+
+The distribution of the cases per category used are in Figure~\fig{histcasecat}.
 \begin{figure}[h!]
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20)
 @
 \end{center}
-\caption{test}
-\label{fig:test}
+\caption{Histogram of cases per category used in the final regression model}
+\label{fig:histcasecat}
 \end{figure}
+
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+%@
+%\end{center}
+%\caption{test}
+%\label{fig:test}
+%\end{figure}
+The predicted values plotted against the cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width =11, height=11>>=
+res<-predict(model3)
+plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey")
+@
+\end{center}
+\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
+\label{fig:predict}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+@
 
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+hist(model3$residuals,main="",xlab="residuals",col="grey")
 @
 \end{center}
-\caption{test}
-\label{fig:test}
+\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
+\label{fig:resid}
 \end{figure}
+The significance of the different variables can be seen here, in the ANOVA results:
+
+<<label=regmodelanova,include=FALSE>>=
+anova(model3,test="F")
+@
+
+A summary of the model with the individual effects and the statistical support for the estimated coefficients can be seen here:
+
+<<label=regmodel,include=FALSE>>=
+summary(model3)
+@
+
+The results of the model suggest that, apart from over representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age.
+This can be seen in Figure~\ref{fig:ageinyears}.
+
+People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+
+In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected.
+The very young are still most likely to be infections, but older teenagers are more likely to be represented (Figure~\ref{fig:ageinyears}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly these results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+
+\subsection{Risk analysis discussion}
+
+The analyses suggest that age is a particularly strong risk factor for measles.
+This comes as no surpise to epidemiologists or health care providers.
+However, our analyses also highlight other groups that are at risk.
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5, and European 6--17 year old children.
+The results of these analyses are discussed in the cost--benefit section below (Section\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
+A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.
+This work is ongoing.
+
+\subsection{Risk analysis summary}
+\begin{itemize}
+\item Risk of measles infection decreases significantly with age
+\item Pacific people are more at risk on a per capita basis
+\item There is some support for those living in better socio-economic situations being at greater risk of measles
+\item There is support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
+\end{itemize}
+
+\section{Modelling measles epidemics}
+\label{sec:epidemic_modelling}
+
+In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the cost--benefit analysis (Section~\ref{sub:cost_benefit}).
+
+A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
+The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
 
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-res<-predict(model2)
-plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
+plot(pop,xlab="Age",ylab="Population")
+points(naive,pch=16)
+legend("topright",c("Population","Naive"),pch=c(1,16),bty="n")
 @
 \end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tp$cases)
-cor.test(exp(res),tp$cases)
-@
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-plot(tp$cases/tp$Popn,main="Cases per capita",ylab="Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="Category")
-legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
-       col=c(1:4,6),pch=16,bty="n")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\section{Regression analyses results}
-\label{sub:regression_results}
-The results are:
-
-<<label=regmodel,include=FALSE>>=
-summary(model2)
-@
-
-<<label=regmodelanova,include=FALSE>>=
-anova(model2,test="F")
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-tpsub<-tp[!(tp$Ethnicity=="MLA"),]
-
-hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
-
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F")
-
-model2<-update(model1,~.-Ethnicity:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-model3<-update(model2,~.-Age:NZDep)
-summary(model3)
-anova(model3,test="F")
-@
-
-The results are:
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-res<-predict(model3)
-plot(exp(res),tpsub$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-hist(model3$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-plot(tp$cases/tp$Popn,main="Cases per capita",ylab="Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="Category")
-legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
-       col=c(1:4,6),pch=16,bty="n")
-@
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\section{Regression analyses results}
-\label{sub:regression_results}
-The results are:
-
-<<label=regmodel,include=FALSE>>=
-summary(model3)
-@
-
-<<label=regmodelanova,include=FALSE>>=
-anova(model3,test="F")
-@
-
-And plotted in (Figure~\ref{fig:regn}).
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
-     \caption{Top left: distribution of the cases per category; Top right: Cases per Category; Bottom left: Model predictions (x) versus case data (y); and Bottom right: the residuals of the model}
-     \label{fig:regn}
-\end{figure}
-
-The results of the model suggest that, apart from over representation of MELAA, older people over 25 years of age are significantly less likely to be infected with measles than the very young (0--2 years old), but Pacific peoples are at greater risk than Europeans and others (as can also be seen in Figure 1), but not the 6--17 years among them.
-
-In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected.
-The very young are still most likely to be infections, but older teenagers are more likely to be repreested (Figure N).
-This pattern suggests that imporving vaccination coverage in the young is reducing the burden of measles in those age categories.
-See discussion regarding vaccination protocols.
-
-Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
-We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
-meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
-Thus the data gap that we have that will hinder us providing fine scale risk maps is:
-\begin{itemize}
-\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
-\end {itemize}
-An additional data set that would enable us to develop the understanding of measles importation risk is:
-\begin {itemize}
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
-\end{itemize}
-
-\subsection{Risk analysis summary}
-\begin{itemize}
-\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
-\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
-\end{itemize}
-
-\section{Modelling measles epidemics}
-\label{sec:epidemic_modelling}
-
-A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
-The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
-The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
-     \caption{test}
+\caption{New Zealand population by age and estimates numbers of naive people in each age class}
 \label{fig:naive}
 \end{figure}
-
-The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
-Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
-The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
-The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
-
-These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
-In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
-All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
-They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
-
-Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
-The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
 
@@
 
-To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we make several simplifying assumptions and the simple relationship between the proportion requiring vaccination, $p_c$, and $R_0$.
+To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we make several simplifying assumptions regarding the relationship between the proportion requiring vaccination, $p_c$, and $R_0$.
 Specifically, we use the simple relationship:
@@
 \begin{equation} \label{eq:pc}
-p_c = 1-(1/R0)
+p_c = 1-(1/R_0)
   \end{equation}
 
-We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population, estimated from the national level vaccination coverage data provided by the Ministry of Health for this work.
-Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one.
-
-And discuss the implications for this simplification in the discussion. However, briefly, there are several caveats to our use of this equation:
-\begin {itemize}
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
-\end{itemize}
-
-discussion of simplifications and use of other av:
-\begin {itemize}
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
-\end{itemize}
-
-We then used the proportions of the population that were estimated be require vaccination as estimated above, and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand.
+We estimated this from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
+Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
+
+We then used the proportions of the population that were estimated to require vaccination (estimated above), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the cost--benefit analyses section below (Section~\ref{sub:cost_benefit}).
 
@@
 
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
-The probability density of the $R_v$ estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
-An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report.
+The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one.
+As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+However, the results of these differences are presented in the cost--benefit analysis section (Section~\ref{sub:cost_benefit}).
 
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
-The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic and we explore the cost--benefit effects of this in Section~\ref{sub:cost_benefit}.
+
+Again The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
 This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
@@
 To use the results from our modeling exercise to help inform the appropriate measles vaccination coverage.
-Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 to 53\%.
-These figures can be reached in a number of different ways, and these are discussed in the cost-benefit section.
+Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
+These figures can be reached in a number of different ways, and these are discussed in the cost--benefit section (Section~\ref{sub:cost_benefit}).
 
 \subsection{Summary of modelling}
+These results suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
+However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
+In Ireasl researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
+Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+In summary:
 \begin{itemize}
-\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
-\end{itemize}
-
-\subsection{Future modelling}
-Future modelling we aim to perform are:
-\begin{itemize}
-\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
-\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
+ \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
+ \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0 to 53\%, depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
 \end{itemize}
@@
 Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$298 per contact.
-\citep{dayan5} calculated the cost of containing a single case of measles that occurred in Iowa in 2004.
-They estimated that for the one week that the Iowa Department of Public Health (DPH) investigated the case, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination.
+It was estimated that for the one week that the Iowa Department of Public Health (DPH) investigated a case in 2004, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination \citep{dayan5}.
 In total, it was estimated the direct costs associated with three cases of measles was US\$142,452, or nearly US\$50,000 per case.
 
-\citep{parker6} reported the impact of a large measles outbreak due to a non-autochthonous case in Indiana.
-A total of 34 cases, 94\% of which were not vaccinated against measles were reported in the outbreak.
+The impact of a measles outbreak due to a non-autochthonous case in Indiana was also reported \citep{parker6}, and a total of 34 cases, 94\% of which were not vaccinated against measles, were reported in the outbreak.
 Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak.
@@
 
-\citep{coleman12} estimated the direct medical and public health costs in response to a single case of refugee-imported measles.
+The direct medical and public health costs in response to a single case of refugee-imported measles has been reported \citep{coleman12}.
 Costs included labor, translation and benefits for public health workers.
@@
 Using this information, we will then evaluate the economics of alternative measles control strategies in order to provide additional information to public health officials and decision makers.
-
-.....Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
-Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.....
 
@@
 \subsection{Cost analyses results}
+\label{sub:cost_results}
 Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
@@
 
-The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hosptial was \$550,024 (Table~\ref{table:hosp}).
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (Table~\ref{table:hosp}).
 The mean cost per case was \$1,877.
@@
 
-Wages lost due to measles were calculated for the period January 2008 - August 2014.
+Wages lost due to measles were calculated for the period January 2008 -- August 2014.
 Calculations were based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
@@
 This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008 - 2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
-
+Overall, the cost per case from 2008--2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
+
+\subsection{Benefit--cost analyses methods}
 To estimate the benefits from additional vaccinations, as estimated from the above modeling section, we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ in the estimated naive population from the modeling section, assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
+Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated naive population of New Zealand (Figure~\ref{fir:naive}), assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
 We used these values of numbers of predicted cases and the cost figures above to estimate the cost of not vaccinating additional populations.
-We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would now be one, though where our estimates for $R_v$ are < 1, we use this.
+We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would now be one, though where our estimates for $R_v$ are already < 1, we use this.
 We can therefore estimate the number of cases prevents, and the savings made from the additional vaccinations.
@@
 \begin{equation} \label{eq:reg}
-discount per year = cost * 1/1.08^(year-1))
+discount/year = cost * 1/1.08^{year-1}
   \end{equation}
 
-Our estimates for the numbers of measles cases are highly variable, though typically in our model, if $R_v$ were greater than one, either the measles outbreak would take off and all the New Zealand population would become infected, or not, and we used the average of 1000 simulations to estimate the mean of this.
+Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the New Zealand population would become infected, or it would stutter after a few infects and fail to take off due to stochasticity.
+We used the average of 1000 simulations to estimate the mean of these simulations for our benefit--cost analysis.
 However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes.
-We therefore also compared our results to those using the mean number of measles cases seen in New Zealand per yea, since 1997.
+We therefore also compared our results to those using the mean number of measles cases seen in New Zealand per year, since 1997.
 This was 220 cases per year.
 
-The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure
+The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure~\ref{numvacc}.
 
@@
 
-The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure N.
+The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim}.
 
@@
 
-The benefit-cost results are in Table N and N.
+For the cost parts we used the figures from the above cost section (Section~\ref{sub:cost_results}).
+Specifically, we used a 17\% proportion of cases to be hospitalised and the average cost of a case for the analyses were \$852 lost in wages and \$1710 in hospitalisation costs.
+We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
+We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US litarature (REF).
+
+\subsection{Benefit--cost analyses results}
+The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_mean}.
 
@@
     \toprule  
-        Years R0 estimated from & R0 range	& R0  for simulations	& Mean # of cases in the population (1000 simulations)	& Vaccines required to reduce R0 to 1	& Costs per vaccine	& Total vaccine costs	& Total hospitalised cases	\sym{*}& Total costs for cases \sym{**}	& R0  for simulations after response	& Mean number of cases in the population after action (1000 simulations)	& Total cases over 10 years after action \sym{***}	& Total hospitalised cases after action	& Total costs for cases after action	& Cases reduced due to action	& Savings	& Benefit--cost ratio\\
+        Years R0 estimated from & R0 range estimated	& R0  for simulations	& Mean # of cases in the population \dagger	& Vaccines required to reduce R0 to 1	& Costs per vaccine	& Total vaccine costs (\$)	& Total hospitalised cases	(\$)\sym{*}& Total costs for cases (\$)\sym{**}	& R0  for simulations after response	& Mean cases in the population after catch up \dagger	& Total cases over 10 years after action \sym{***}	& Total hospitalised cases after action	& Total costs for cases after action (\$)	& Cases reduced due to action	& Savings (\$)	& Benefit--cost ratio\\
     \midrule 
@@
     \begin{tablenotes}\footnotesize  
+        \item \dagger 1000 simulations
         \item \sym{*} Proportion of cases hospitalised 0.17
@@
     \end{threeparttable}  
-  \label{tab:xx}
+  \label{tab:cost_sim}
 \end{sidewaystable}
@@
     \toprule  
-        Years R0 estimated from &R0 range	&R0 for simulations	&Vaccines required to reduce R0 to 1	&Costs per vaccine	&Total vaccine costs	&Expected number of cases expected over 10 years based on average since 1997	&Total hospitalised cases	\sym{*}&Total costs for cases \sym{**}	&R0  for simulations after response	&Mean number of cases in the population after action (1000 simulations)	&Total cases over 10 years after action \sym{***}	&Total hospitalised cases after action	&Total costs for cases after action	&Cases reduced due to action	&Savings	&Benefit--cost ratio\\
+        Years R0 estimated from &R0 range estimated	&R0 for simulations	&Vaccines required to reduce R0 to 1	&Costs per vaccine (\$)	&Total vaccine costs (\$)	&Expected number of cases expected based on average since 1997 \sym{***}	&Total hospitalised cases	\sym{*}&Total costs for cases (\$)\sym{**}	&R0  for simulations after catch up	&Mean number of cases in the population after action (1000 simulations)	&Total cases over 10 years after action \sym{***}	&Total hospitalised cases after action	&Total costs for cases after action (\$)	&Cases reduced due to action	&Savings (\$)	&Benefit--cost ratio\\
     \midrule 
@@
     \end{threeparttable}  
-  \label{tab:xx}
+  \label{tab:cost_mean}
 \end{sidewaystable}
@@
 \normalsize
-and the proportions of each age category shown in N.
-
-However, it is worth noting that vaccination strategies that target the very young (<1) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure ) were vaccinates with a signle vaccine at a very young age (Figure N).
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
+
+The results in the two tables show two alternative trends.
+In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}.
+However, if $R_v$ is less than one and the previous recent history of cases since 1997 is the 'status quo' then the additional effort to vaccinate the naive population based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r_0} is not a cost effective exercise (Table~\ref{tab:cost_mean}.
+
+It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) and that they were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+barplot(AgeVac,xlab="Age",col=c("darkgrey","red","orange"),main="Age and vaccination status of cases")
+legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"))
+@
+\end{center}
 \caption{test}
+\label{fig:ageandvac}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="Age of vaccination of vaccinated cases")
+abline(v=12,col="blue")
+hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+abline(v=60,col="red")
+legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n")
+legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n")
+@
+\end{center}
+\caption{Age of vaccination of vaccinated measles cases}
 \label{fig:vaccstat}
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
-     \caption{test}
-     \label{fig:ageofvacc}
-\end{figure}
-
 \subsection{Cost analysis discussion}
 The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
-While some of the data are complete and detailed, this is not true of all the data.
-In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
-For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
-In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
-In addition, similar data would be needed for non-publicly funded hospitals, e.g.
-private clinics.
-Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
-
-Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
-Similar data are needed for the period preceding 2014.
-If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
-However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
-As other studies have demonstrated, direct costs required to manage measles are not linear.
+The majority of the cost analyses were discussed in our previous interim report and though some details have changed, we will focus here on the benefit--cost analyses.
+However, for all analyses more detailed costs of all aspects of hopsitalisation would be beneficial.
+So too would information of the cost of measles vaccination in New Zealand.
+
+We have used standard (REF) discounting from healthcare of 8\% per annum, but again there may be official New Zealand discounting rates that we are unaware of.
+For our cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 and are included in our cost analsyes.
+It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
 
 In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin02}.
-This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
-The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
-In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
-They estimated direct costs per case to be less than US\$500.
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
-In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
-In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
-The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
-
-\subsection{Cost analysis summary}
+In our analyses if we include loss of earnings and hospitalisation costs, our estimates are higher, around US\$959 per case.
+However, costs can vary widely, with the containment of a single case and 2 secondary cases of measles in 2004 in Iowa, USA, estimated to cost US\$142,542 alone \citep{dayan5}.
+
+Our estimates fo the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the naive New Zealand population until overyone is infected.
+Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
+However, if we presume that in a naive population $R_0$ for measles is around 12, and sometimes estimates to be more than 18 (i.e.
+1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r_0}) and the benefits of catch up vaccination are clear (Table~\ref{tab:cost_sim}).
+If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r_0}), then there is little benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_mean}).
+
+Our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios.
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions (see previous report) occuring on an annual basis may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim}).
+
+\subsection{Benefit--cost analysis summary}
 \begin{itemize}
-\item Our initial estimates suggest the ongoing 2013--2014 measles outbreak has cost New Zealand over \$750,000.
+\item Wage losses per measles case is estiamted to be \$852
+\item Cost per hospitalised measles case is estimated to be \$1710
+\item Approximately 17\% of cases are hopsitalised
+\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear.
+\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective.
+\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many hundreds, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out with certainty.
 \end{itemize}
-
-\subsection{Future cost-benefit analyses}
-Using the results above we aim to:
-\begin{itemize}
-\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses (section~\ref{sub:risk_analyses}) or modelling (section~\ref{sec:epidemic_modelling}).
-\end{itemize}
-
-We also require additional clarifications of the data, regarding:
-\begin {itemize}
-\item What hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight? Or, does it mean the case stayed for less than 24 hours?
-\end {itemize}
-Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
 
@@
 \begin{itemize}
-\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
-\item The cost of the current measles outbreak is estimated to be at least \$750,000.
-\item Analyses of outbreak data suggest that measles $R_v$ values often include 1 and in this year, 2014, are well above one.
-This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
+\item Age is the best predictor of risk, though some additional factors ...
+\item Modeling and vaccination rates....
+\item benefit--cost ......
 \end{itemize}

### 2014-09-01 interimreport2.Rnw
Figures and cost benefit changes
Large changes to sweave code, especially relating to figures, and
altered c/b ratio analyses after going through with Tim

@@
 
-\title{Interim report 2\\ Measles risk assessment, modelling and cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
+\title{Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
 \author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
@@
 In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
-We provide preliminary cost--benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
+We provide preliminary benefit--cost analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
 Our key findings were:
 \begin{itemize}
-\item text.
-\item text.
-\item text.
+\item Age is the best predictor of risk of measles infection in multi-variate regression analyses, though some groups such as Pacific people, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
+\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
+\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is very beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
 \end{itemize}
@@
 A brief review of the history of measles in New Zealand was provided in our previous interim report.
-In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the cost--benefit analyses for vaccination dependent on differing scenarios.
+In this report were present regression analyses to determine which populations are most at risk, the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the benefit--cost analyses for vaccination dependent on differing scenarios.
 
@@
 These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
-We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist.
+We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist among cases allowing better understanding of risk of measles infection.
 
 The University of Otago denominator data provided were not to the same detail as the ESR case data.
-Notably, the denominator age data was categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
-The ethnicity denominator data was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes.
+Notably, the denominator age data were categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
+The ethnicity denominator data were not Prioritized Ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes.
 After discussions with the University Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.
@@
 par(mfrow=c(1,1))
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F)
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="",cex.lab=2)
 hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n")
+legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1.8)
 @
@@
 
-This large number of categories, some with population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi and univariate analyses (not shown) suggested little effect of NZDep and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two, 1--5 and 6--10.
+This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: 1--5 and 6--10.
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
@@
 
-The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:cases}).
+The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:case}).
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
 
-The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:cases}.
-
 \begin{table}\tiny
@@
 @
-\label{table:cases}
+\label{table:case}
 \end{table}
@@
 
-Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over- or under-represented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{fig:}), leading to very large standard error in regression analyses.
+The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
 
@@
 <<fig=TRUE,echo=FALSE,width = 11,height=9>>=
-plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="Category")
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
+     xlab="Category",cex.lab=2)
 legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
-       col=c(1:4,6),pch=16,bty="n")
+       col=c(1:4,6),pch=16,bty="n",cex=1.8)
 @
@@
 
-However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes, and there several issues with estimating the denominator data for these group (University of Otago, personal communication).
+However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007.
-
-For all our statistical analyses (including those above not shown) we used a Poisson error strutuce, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).
+
+For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model.
 We also account for differences in population sizes by using an offset term, the log(population size).
 We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms.
-Thus, the final model that remained with all significat interaction terms had the following linear predictor:
+Thus, the final model that remained with all significant interaction terms had the following linear predictor:
 
@@
   \end{equation}
-Where $y$ = cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.
+Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.
 
@@
 
-The distribution of the cases per category used are in Figure~\fig{histcasecat}.
+The distribution of the cases per category used are in Figure~\fig{fig:histcasecat}.
 \begin{figure}[h!]
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20)
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
 @
@@
 
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%plot(tpsub$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
-%@
-%\end{center}
-%\caption{test}
-%\label{fig:test}
-%\end{figure}
 The predicted values plotted against the cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
@@
 res<-predict(model3)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey")
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
 @
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-hist(model3$residuals,main="",xlab="residuals",col="grey")
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
 @
@@
 \end{figure}
-The significance of the different variables can be seen here, in the ANOVA results:
+The significance of the different predictor variables can be seen here, in the ANOVA results:
 
@@
 
-The results of the model suggest that, apart from over representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age.
-This can be seen in Figure~\ref{fig:ageinyears}.
-
-People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-
-In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected.
-The very young are still most likely to be infections, but older teenagers are more likely to be represented (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly these results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+The results of the model suggest that, apart from over-representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
+
+People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+
+In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young are still most likely to be infections, but older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
 
@@
 However, our analyses also highlight other groups that are at risk.
-In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5, and European 6--17 year old children.
-The results of these analyses are discussed in the cost--benefit section below (Section\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European 6--17 year old children.
+The results of these analyses are discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
 A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.
-This work is ongoing.
 
@@
 \item Risk of measles infection decreases significantly with age
-\item Pacific people are more at risk on a per capita basis
-\item There is some support for those living in better socio-economic situations being at greater risk of measles
-\item There is support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
+\item Pacific people are statistically more at risk on a per capita basis
+\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
+\item There is some statistical support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
 \end{itemize}
@@
 
-In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the cost--benefit analysis (Section~\ref{sub:cost_benefit}).
+In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the benefit--cost analysis (Section~\ref{sub:cost_benefit}).
 
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-plot(pop,xlab="Age",ylab="Population")
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+#length(pop)
+naive<-round(pop-(pop*impop))
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+plot(pop,xlab="Age",ylab="Population",cex.lab=2)
 points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n")
+legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
 @
@@
      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks}}
      \label{fig:r0}
@@
 We then used the proportions of the population that were estimated to require vaccination (estimated above), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
-These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the cost--benefit analyses section below (Section~\ref{sub:cost_benefit}).
+These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).
 
@@
 As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
-However, the results of these differences are presented in the cost--benefit analysis section (Section~\ref{sub:cost_benefit}).
+However, the results of these differences are presented in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).
 
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic and we explore the cost--benefit effects of this in Section~\ref{sub:cost_benefit}.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, and we explore the benefit--cost effects of this in Section~\ref{sub:cost_benefit}.
 
@@
 
-To use the results from our modeling exercise to help inform the appropriate measles vaccination coverage.
+To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage.
 Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
-These figures can be reached in a number of different ways, and these are discussed in the cost--benefit section (Section~\ref{sub:cost_benefit}).
+These figures can be reached in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
 
@@
 However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
-In Ireasl researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
+In Isreal researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
 Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
@@
  \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
- \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0 to 53\%, depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
+ \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
 \end{itemize}
@@
 \subsection{Cost analyses results}
-\label{sub:cost_results}
+\label{sub:cost_benefit}
 Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
@@
 library(ggplot2)
-p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")
+p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")+theme(axis.title.x = element_text(size=25),axis.title.y = element_text(size=25),axis.text.x  = element_text(size=16),axis.text.y  = element_text(size=16))
 @
@@
 \subsection{Benefit--cost analyses methods}
-To estimate the benefits from additional vaccinations, as estimated from the above modeling section, we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated naive population of New Zealand (Figure~\ref{fir:naive}), assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section, we did several things.
+Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated naive population of New Zealand (Figure~\ref{fig:naive}), assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
 We used these values of numbers of predicted cases and the cost figures above to estimate the cost of not vaccinating additional populations.
-We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would now be one, though where our estimates for $R_v$ are already < 1, we use this.
+We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this.
 We can therefore estimate the number of cases prevents, and the savings made from the additional vaccinations.
@@
 Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the New Zealand population would become infected, or it would stutter after a few infects and fail to take off due to stochasticity.
-We used the average of 1000 simulations to estimate the mean of these simulations for our benefit--cost analysis.
+We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
 However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes.
@@
 
-The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure~\ref{numvacc}.
+The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure~\ref{fig:numvacc}.
 
@@
      \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
-     \caption{test}
+     \caption{Numbers of the currently naive New Zealand population to ensure $R_v$ is less than one, and thus all measles introductions will fail to persist within the population.
+$R_v$ estimates are from the outbreaks in Figure ~\ref{fig:outbreaks} and shown in Figure ~\ref{fig:r0}}
      \label{fig:numvacc}
@@
 
-The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim}.
+<<echo=FALSE,include=FALSE,results=hide>>=
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+## match cases per age
+AgeInYears<-0:100
+naive<-cbind(AgeInYears,naive)
+colnames(naive)<-c("AgeInYears","Naive")
+naive<-merge(naive,caseyr,by="AgeInYears",all=T)
+#naive[is.na(naive)] <- 0
+colnames(naive)<-c("AgeInYears","Naive","Cases")
+@
 
@@
      \centering
-     \includegraphics[width=0.9\textwidth]{simulations.pdf}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+##
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+plot(round(0.53*naive$Naive) ,col="grey",bg="red",pch=21,ylab="Number to vaccinate",xlab="Age in Years",
+     main="",cex.lab=2)
+points(naive$Naive[0:21],col="grey",bg="orange",pch=22) # 
+points(naive$Naive[3:30],col="grey",bg="yellow",pch=23) 
+points(jitter(round(0.16*naive$Naive)),col="grey",bg="lightblue",pch=21) 
+points(jitter(naive$Naive[5:13]),col="grey",pch=22,bg="skyblue") # 
+points(naive$Naive[1:5],col="grey",pch=23,bg="navyblue") #
+legend("topright",c("53% naive, all ages","53% naive, all 0-21 yr olds","53% naive, all 3-30 yr olds",
+                    "16% naive, all ages","16% naive, all 5-13 yr olds","16% naive, all 1-5 yr olds"),
+       bty="n",pch=c(rep(21:23,2)),col=rep("grey",6),pt.bg=c("red","orange","yellow","lightblue","skyblue","navyblue"),cex=2)
+@
      \caption{test}
@@
 
-For the cost parts we used the figures from the above cost section (Section~\ref{sub:cost_results}).
-Specifically, we used a 17\% proportion of cases to be hospitalised and the average cost of a case for the analyses were \$852 lost in wages and \$1710 in hospitalisation costs.
+The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim}.
+These simulations show that even in scenarios when $R_v$ is one and thus stochastically should fail to persist for long, large outbreaks can occur.
+In this modelling exercise, the median value of the 1000 simulations was 2 cases, but the mean 175 and the maximum nearly 25,000 cases.
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+## simulate measles outbreaks/epidemics using GT and R0 from data.
+# http://www.sciencedirect.com/science/article/pii/S0022519311003146
+mu    <- 12
+sigma <- 3.5
+sigma_logn <- sqrt(log(1 + (sigma/mu)^2))
+mu_logn    <- log(mu) - log(1 + (sigma/mu)^2) / 2
+
+# then exp(rnorm(n, mu_logn, sigma_logn)) simulates from lognormal with the given mean and sd.
+
+# the R0 library 
+require(R0)
+#source("estR0.R")
+
+# generation time 
+genTime <- generation.time(type="lognormal", val=c(12, 3.5))
+
+jm.epid <- function (epid.nb, GT, R0, epid.length, family, negbin.size = NULL, 
+          peak.value = 300000, popn = 300000) 
+{
+  if (class(GT) != "R0.GT") {
+    stop("GT object must be of class R0.GT.")
+  }
+  if (family == "negbin" & is.null(negbin.size)) {
+    negbin.size <- R0/4
+  }
+  GT <- GT$GT
+  epidemics <- matrix(data = 0, nrow = epid.length, ncol = epid.nb)
+  for (n in 1:epid.nb) {
+    sim.epid = c(1, rep(0, epid.length - 1))
+    for (t in 1:epid.length) {
+      if (family == "poisson") {
+        susc <- max(0,popn - sum(sim.epid, na.rm=T))
+    #    peak.value[t]<-susc
+        new <- rbinom(sim.epid[t], susc, min(1, R0/susc))
+        if (sum(new) > susc) # arggrggg
+        {
+    #      cat("doing the new thing", susc, sim.epid[t], sum(sim.epid[1:t]),"\n")
+          new <- rep(0, sim.epid[t])
+          for (i in 1:sim.epid[t]) {
+   #         cat("taking a random number, susc=", susc, "\n")
+            new[i] <- rbinom(1, susc, min(1, R0/susc))
+            susc <- susc - new[i]
+            if (susc == 0)
+            {
+ #             cat("got 'em all", sum(new), "\n")
+              break;
+            }
+          }
+        }
+#        cat("generating", new[1], "individuals from first, susc=", susc, "\n")
+      }
+      else if (family == "negbin") {
+        new <- rnbinom(sim.epid[t], size = negbin.size, 
+                       mu = R0)
+      }
+if (is.na(sum(new)))
+  break
+
+      newd <- rmultinom(1, sum(new), GT)[, 1]
+      sim.epid[t:(t + length(GT) - 1)] <- sim.epid[t:(t + 
+                                                        length(GT) - 1)] + newd
+      if (sim.epid[t + 1] > peak.value & t < (epid.length - 
+                                                1)) {
+        sim.epid[(t + 2):epid.length] <- 0
+        break
+      }
+    }
+    sim.epid <- sim.epid[!is.na(sim.epid)]
+    epidemics[, n] <- sim.epid
+  }
+  return(epidemics)
+}
+
+## if R 0 == 2
+set.seed(1) # all infected
+res<-jm.epid(epid.nb=1000,GT=genTime,R0=1,epid.length=365*5,popn=4.2*1e6 * 0.11
+             ,family="poisson",peak.value=4.2*1e6 * 0.11)
+sizes<-colSums(res)
+sizes
+legMd<-as.factor(paste("median =",c(median(sizes))))
+legMn<-as.factor(paste("mean =",c(round(mean(sizes)))))
+@
+
+\begin{figure}
+     \centering
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(sizes,xlim=c(0,200),col="grey",breaks=3000,xlab="Outbreak size",main="",cex.lab=2)
+#summary(sizes)
+abline(v=median(sizes),col="orange",lty=3,lwd=10)
+abline(v=mean(sizes),col="red",lty=3,lwd=10)
+#length(which(sizes>5))/1000
+legend('top',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
+@
+     \caption{test}
+     \label{fig:sim}
+\end{figure}
+
+\begin{figure}
+     \centering
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(sizes,breaks=1000,xlab="Outbreak size",main="",cex.lab=2)
+abline(v=median(sizes),col="orange",lty=3,lwd=10)
+abline(v=mean(sizes),col="red",lty=3,lwd=10)
+legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
+@
+     \caption{test}
+     \label{fig:sim}
+\end{figure}
+
+For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
+Specifically, used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
 We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
-We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US litarature (REF).
+We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
 
 \subsection{Benefit--cost analyses results}
-The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_mean}.
 
@@
     \begin{tablenotes}\footnotesize  
-        \item \dagger 1000 simulations
+        \item $\dagger$ {1000 simulations}
         \item \sym{*} Proportion of cases hospitalised 0.17
@@
   \caption{Cost benefit analyses using simulated epidemic sizes}  
+  \label{tab:cost_ave}
    \begin{threeparttable}[b] \tiny
@@
     \end{threeparttable}  
-  \label{tab:cost_mean}
 \end{sidewaystable}
@@
 
+The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}.
 The results in the two tables show two alternative trends.
-In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}.
-However, if $R_v$ is less than one and the previous recent history of cases since 1997 is the 'status quo' then the additional effort to vaccinate the naive population based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r_0} is not a cost effective exercise (Table~\ref{tab:cost_mean}.
+In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
+However, if the previous recent history of cases since 1997 is the 'status quo' and what we may expect in the future then the \emph{additional} effort to vaccinate the naive population of New Zealand based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r0}) is not a cost effective exercise (Table~\ref{tab:cost_ave}).
 
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-barplot(AgeVac,xlab="Age",col=c("darkgrey","red","orange"),main="Age and vaccination status of cases")
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"))
+AgeVac<-table(testv$VC,testv$AgeInYears)
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
+legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
 @
 \end{center}
-\caption{test}
+\caption{Age and vaccination status of cases}
 \label{fig:ageandvac}
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="Age of vaccination of vaccinated cases")
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
 abline(v=12,col="blue")
@@
 abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n")
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n")
+legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
+legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
 @
@@
 
-\subsection{Cost analysis discussion}
+\subsection{Benefit--cost analysis discussion}
 The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
@@
 
-We have used standard (REF) discounting from healthcare of 8\% per annum, but again there may be official New Zealand discounting rates that we are unaware of.
-For our cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 and are included in our cost analsyes.
+We have used standard discounting from healthcare of 8\% per annum, but again there may be official New Zealand discounting rates that we are unaware of.
+For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 and are included in our benefit--cost analsyes.
 It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
 
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin02}.
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin2}.
 In our analyses if we include loss of earnings and hospitalisation costs, our estimates are higher, around US\$959 per case.
-However, costs can vary widely, with the containment of a single case and 2 secondary cases of measles in 2004 in Iowa, USA, estimated to cost US\$142,542 alone \citep{dayan5}.
-
-Our estimates fo the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the naive New Zealand population until overyone is infected.
+This figure is from NZ\$ 852 for loss of wages and 0.17 * NZ \$ 1710 for costs related to attending hospital.
+However, costs can vary widely, with the containment of a single case and two secondary cases of measles in 2004 in Iowa, USA, estimated to cost US\$142,542 alone \citep{dayan5}.
+
+Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the naive New Zealand population until overyone is infected.
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
-However, if we presume that in a naive population $R_0$ for measles is around 12, and sometimes estimates to be more than 18 (i.e.
-1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r_0}) and the benefits of catch up vaccination are clear (Table~\ref{tab:cost_sim}).
-If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r_0}), then there is little benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_mean}).
-
-Our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios.
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions (see previous report) occuring on an annual basis may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim}).
+However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimates to be more than 18 (i.e.
+1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}) and the benefits of catch up vaccination are clear (Table~\ref{tab:cost_sim}).
+If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r0}), then there is little financial benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}), though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
+
+Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios, such as the size of the local naive population.
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim}).
 
@@
 \begin{itemize}
-\item Wage losses per measles case is estiamted to be \$852
-\item Cost per hospitalised measles case is estimated to be \$1710
-\item Approximately 17\% of cases are hopsitalised
-\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear.
-\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective.
-\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many hundreds, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out with certainty.
+\item The mean wage losses per measles case is estimated to be \$852
+\item The mean cost per measles case attending hospital is estimated to be \$1710
+\item Approximately 17\% of measles cases attend hopsital
+\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
+\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
+\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
 \end{itemize}
@@
 \begin{itemize}
-\item Age is the best predictor of risk, though some additional factors ...
-\item Modeling and vaccination rates....
-\item benefit--cost ......
+\item Risk of measles infection decreases significantly with age
+\item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
+\item Pacific and Asian children in the 6--17 year age caterories have been at lower risk than European or Maori children of the same age
+\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
+\item The mean wage losses per measles case is estimated to be \$852
+\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hopsital
+\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
+\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
+\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
 \end{itemize}

### 2014-09-01 interimreport2.tex
Figures and cost benefit changes
Large changes to sweave code, especially relating to figures, and
altered c/b ratio analyses after going through with Tim

@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we report the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
-We provide preliminary cost-benefit analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
+In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
+We provide preliminary benefit--cost analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
 Our key findings were:
 \begin{itemize}
-\item text.
-\item text.
-\item text.
+\item Age is the best predictor of risk of measles infection in multi-variate regression analyses, though some groups such as Pacific people, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
+\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
+\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is very beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
 \end{itemize}
@@
 
-As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
+As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission.
 A brief review of the history of measles in New Zealand was provided in our previous interim report.
-In this report were report some regression analyses to determine which populations are most at risk, and the likely outcomes of measles infections based on a number of assumptions, and the cost benefit analyses for vaccination dependent on differing scenarios.
+In this report were present regression analyses to determine which populations are most at risk, the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the benefit--cost analyses for vaccination dependent on differing scenarios.
 
@@
 These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
-We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist.
-
-The Otago denominator data provided were not to the same detail as the ESR case data.
-Notably, the denominator age data was categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
-The denominator ethnicity was not prioritized ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided.
-After discussions with Otago we have provided results based on the best available data, though for smaller groups, some results may be unreliable as these are discussed below.
-
-The prioritized ethnicities, therefore, are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African; Other Ethnicity; Residual Categories, though for this report only the first five are used, as these categories covered all the cases.
-Denominator data were classified according to the recommendations of Otago.
-Thus, with the 10 NZDep classes and the age classes above, this led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:agecases}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics (REF)).
+We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist among cases allowing better understanding of risk of measles infection.
+
+The University of Otago denominator data provided were not to the same detail as the ESR case data.
+Notably, the denominator age data were categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
+The ethnicity denominator data were not Prioritized Ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes.
+After discussions with the University Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.
+
+With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, this led us to have 250 categories.
+Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
 
@@
 \end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-004}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-Previous measles analyses, including two in New Zealand by Prof.
-Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-The overall population immunity rate is approximately 89\% and the breakdown of this by age category can be seen in Figure N.
-
-\begin{figure}[h!]
-\begin{center}
-\begin{Schunk}
-\begin{Soutput}
-[1] NA
-\end{Soutput}
-\end{Schunk}
-\includegraphics{interimreport2-006}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-008}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeCases.pdf}
-     \caption{test}
-     \label{fig:agecases}
-\end{figure}
-
-This large number of categories lead to both zeroinflation and over-dispersion as there were many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi and univariate analyses (not shown) suggested little effect of NZDep and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two, 1--5 and 6--10.
+\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+\label{fig:ageinyears}
+\end{figure}
+
+This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: 1--5 and 6--10.
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
 By doing so, we reduce the zeroinflation present in the data.
-Subsequent analyses also suggested that the Middle Eastern/Latin American/African category was over-represented (Figure~\ref{fig:percap}).
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PerCapitaCases.pdf}
-     \caption{Cases per capita for Prioritized Ethnicities.
-Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
-Points are coloured by Ethnicity.}
-     \label{fig:percap}
-\end{figure}
-
-However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, which have small population sizes, and which there were numberous issues with estimating the denominator data for.
-This category was also an outlier (Figure~\ref{fig:percap}), and thus reducing our ability to make inferences regarding the other classifications and risk factors.
-Thus, we removed the MELAA category for the below analyses.
-In the below analyses we used a quasipoission regression model to account for overdispersion in the data.
-We also account for differences in population sizes by using an offset term, the log(population size).
-Thus, we used a quasipoisson error structure and the following linear predictor:
-
-\begin{equation} \label{eq:reg}
- log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _N (x_N) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
-  \end{equation}
-
-The numbers of cases per category can be seen in Table N.
+
+The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:case}).
+
 \begin{table}\tiny
@@
 \hline
-\end{tabular}\end{center}\label{table:cases}
+\end{tabular}\end{center}\label{table:case}
 \end{table}
 
-The numbers of cases per category can be seen in Table N.
+The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-010}
+\end{center}
+\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
+\label{fig:percap}
+\end{figure}
+
 \begin{table}\tiny
@@
 \hline\hline
-\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Popn}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per Capita}\tabularnewline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per Capita}\tabularnewline
 \hline
@@
 
+However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
+Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).
+
+For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model.
+We also account for differences in population sizes by using an offset term, the log(population size).
+We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms.
+Thus, the final model that remained with all significant interaction terms had the following linear predictor:
+
+\begin{equation} \label{eq:reg}
+ log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
+  \end{equation}
+Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.
+
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+%@
+%\end{center}
+%\caption{test}
+%\label{fig:test}
+%\end{figure}
+
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
+%@
+%\end{center}
+%\caption{Cases per capita for Prioritized Ethnicities.
+Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
+Points are coloured by Ethnicity.}
+%\label{fig:test}
+%\end{figure}
+%
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%res<-predict(model2)
+%plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
+%@
+%\end{center}
+%\caption{test}
+%\label{fig:test}
+%\end{figure}
+
+%<<echo=FALSE,include=FALSE,results=hide>>=
+%cor(exp(res),tp$cases)
+%cor.test(exp(res),tp$cases)
+%@
+%
+%\begin{figure}[h!]
+%\begin{center}
+%<<fig=TRUE,echo=FALSE,width = 11>>=
+%hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
+%@
+%\end{center}
+%\caption{test}
+%\label{fig:test}
+%\end{figure}
+
+%\section{Regression analyses results}
+%\label{sub:regression_results}
+%The results are:
+
+%<<label=regmodel,include=FALSE>>=
+%summary(model2)
+%@
+
+%<<label=regmodelanova,include=FALSE>>=
+%anova(model2,test="F")
+%@
+
+\subsection{Regression analyses results}
+\label{sub:regression_results}
+
+The distribution of the cases per category used are in Figure~\fig{fig:histcasecat}.
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-014}
+\end{center}
+\caption{Histogram of cases per category used in the final regression model}
+\label{fig:histcasecat}
+\end{figure}
+
+The predicted values plotted against the cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
 \begin{figure}[h!]
@@
 \end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-016}
-\end{center}
-\caption{test}
-\label{fig:test}
+\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
+\label{fig:predict}
 \end{figure}
@@
 \end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-019}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-020}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\section{Regression analyses results}
-\label{sub:regression_results}
-The results are:
+\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
+\label{fig:resid}
+\end{figure}
+The significance of the different predictor variables can be seen here, in the ANOVA results:
 
@@
 \begin{Sinput}
-> summary(model2)
-\end{Sinput}
-\begin{Soutput}
-Call:
-glm(formula = cases ~ Age + Ethnicity + NZDep + Age:Ethnicity + 
-    Ethnicity:NZDep + offset(log(Popn)), family = "quasipoisson", 
-    data = tp)
-
-Deviance Residuals: 
-     Min        1Q    Median        3Q       Max  
--1.98437  -0.58185  -0.00005   0.45759   1.47521
-
-Coefficients:
-                             Estimate Std. Error t value Pr(>|t|)    
-(Intercept)                -6.422e+00  1.016e-01 -63.219  < 2e-16 ***
-Age3-5                     -9.129e-01  1.794e-01  -5.088 5.62e-05 ***
-Age6-17                    -7.653e-01  1.173e-01  -6.524 2.33e-06 ***
-Age18-24                   -1.495e+00  1.692e-01  -8.836 2.43e-08 ***
-Age25+                     -2.953e+00  1.405e-01 -21.024 4.16e-15 ***
-EthnicityAsian              2.147e-03  3.167e-01   0.007  0.99466    
-EthnicityMaori             -6.061e-02  2.267e-01  -0.267  0.79191    
-EthnicityMLA               -1.820e+01  2.572e+03  -0.007  0.99442    
-EthnicityPacific            6.950e-01  3.541e-01   1.963  0.06374 .  
-NZDep6-10                  -3.339e-01  9.205e-02  -3.627  0.00168 ** 
-Age3-5:EthnicityAsian      -2.032e+00  1.208e+00  -1.682  0.10813    
-Age6-17:EthnicityAsian     -1.003e+00  4.125e-01  -2.432  0.02455 *  
-Age18-24:EthnicityAsian    -8.417e-01  5.194e-01  -1.620  0.12079    
-Age25+:EthnicityAsian       4.237e-01  3.873e-01   1.094  0.28707    
-Age3-5:EthnicityMaori      -3.864e-01  3.578e-01  -1.080  0.29305    
-Age6-17:EthnicityMaori     -7.940e-02  2.157e-01  -0.368  0.71660    
-Age18-24:EthnicityMaori    -5.954e-01  3.917e-01  -1.520  0.14417    
-Age25+:EthnicityMaori      -1.039e+00  4.579e-01  -2.270  0.03443 *  
-Age3-5:EthnicityMLA        -1.857e-01  1.356e+00  -0.137  0.89245    
-Age6-17:EthnicityMLA       -1.591e+00  1.349e+00  -1.179  0.25212    
-Age18-24:EthnicityMLA       2.201e-01  1.076e+00   0.205  0.84001    
-Age25+:EthnicityMLA         1.565e+00  8.351e-01   1.874  0.07556 .  
-Age3-5:EthnicityPacific    -2.132e+00  7.110e-01  -2.998  0.00711 ** 
-Age6-17:EthnicityPacific   -1.448e+00  2.924e-01  -4.953 7.66e-05 ***
-Age18-24:EthnicityPacific  -9.854e-01  4.481e-01  -2.199  0.03979 *  
-Age25+:EthnicityPacific    -5.958e-01  3.815e-01  -1.562  0.13406    
-EthnicityAsian:NZDep6-10    1.330e-01  3.006e-01   0.442  0.66291    
-EthnicityMaori:NZDep6-10    4.296e-01  2.080e-01   2.065  0.05210 .  
-EthnicityMLA:NZDep6-10      1.890e+01  2.572e+03   0.007  0.99421    
-EthnicityPacific:NZDep6-10  6.070e-01  3.551e-01   1.709  0.10283    
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-
-(Dispersion parameter for quasipoisson family taken to be 1.355392)
-
-    Null deviance: 1473.124  on 49  degrees of freedom
-Residual deviance:   27.468  on 20  degrees of freedom
-AIC: NA
-
-Number of Fisher Scoring iterations: 16
-\end{Soutput}
-\end{Schunk}
-
-\begin{Schunk}
-\begin{Sinput}
-> anova(model2,test="F")
+> anova(model3,test="F")
 \end{Sinput}
@@
               Df Deviance Resid. Df Resid. Dev        F   Pr(>F)    
-NULL                               49    1473.12                       
-Age              4  1297.52        45     175.60 239.3263 < 2.2e-16 ***
-Ethnicity        4    20.21        41     155.39   3.7285  0.020056 *  
-NZDep            1     8.62        40     146.77   6.3573  0.020285 *  
-Age:Ethnicity   16    94.49        24      52.28   4.3569  0.001223 ** 
-Ethnicity:NZDep  4    24.82        20      27.47   4.5771  0.008694 ** 
+NULL                             39    1453.62                      
+Age            4  1304.20        35     149.43 183.5795 6.73e-15 ***
+Ethnicity      3    20.00        32     129.43   3.7529 0.028500 *  
+NZDep          1    10.70        31     118.74   6.0236 0.023932 *  
+Age:Ethnicity 12    81.91        19      36.83   3.8430 0.004459 ** 
 ---
@@
 
-The results are:
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-024}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-025}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-026}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-028}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-029}
-\end{center}
-\caption{test}
-\label{fig:test}
-\end{figure}
-
-\section{Regression analyses results}
-\label{sub:regression_results}
-The results are:
+A summary of the model with the individual effects and the statistical support for the estimated coefficients can be seen here:
 
@@
 
-\begin{Schunk}
-\begin{Sinput}
-> anova(model3,test="F")
-\end{Sinput}
-\begin{Soutput}
-Analysis of Deviance Table
-
-Model: quasipoisson, link: log
-
-Response: cases
-
-Terms added sequentially (first to last)
-
-              Df Deviance Resid. Df Resid. Dev        F   Pr(>F)    
-NULL                             39    1453.62                      
-Age            4  1304.20        35     149.43 183.5795 6.73e-15 ***
-Ethnicity      3    20.00        32     129.43   3.7529 0.028500 *  
-NZDep          1    10.70        31     118.74   6.0236 0.023932 *  
-Age:Ethnicity 12    81.91        19      36.83   3.8430 0.004459 ** 
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{Soutput}
-\end{Schunk}
-
-And plotted in (Figure~\ref{fig:regn}).
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{Rplot_quasipoisson_noMLA.pdf}
-     \caption{Top left: distribution of the cases per category; Top right: Cases per Category; Bottom left: Model predictions (x) versus case data (y); and Bottom right: the residuals of the model}
-     \label{fig:regn}
-\end{figure}
-
-The results of the model suggest that, apart from over representation of MELAA, older people over 25 years of age are significantly less likely to be infected with measles than the very young (0--2 years old), but Pacific peoples are at greater risk than Europeans and others (as can also be seen in Figure 1), but not the 6--17 years among them.
-
-In later outbreaks (since 2007) there has also been a shift in the distribution of ages infected.
-The very young are still most likely to be infections, but older teenagers are more likely to be repreested (Figure N).
-This pattern suggests that imporving vaccination coverage in the young is reducing the burden of measles in those age categories.
-See discussion regarding vaccination protocols.
-
-Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
-We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
-meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
-Thus the data gap that we have that will hinder us providing fine scale risk maps is:
-\begin{itemize}
-\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
-\end {itemize}
-An additional data set that would enable us to develop the understanding of measles importation risk is:
-\begin {itemize}
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
-\end{itemize}
+The results of the model suggest that, apart from over-representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
+
+People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+
+In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young are still most likely to be infections, but older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+
+\subsection{Risk analysis discussion}
+
+The analyses suggest that age is a particularly strong risk factor for measles.
+This comes as no surpise to epidemiologists or health care providers.
+However, our analyses also highlight other groups that are at risk.
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European 6--17 year old children.
+The results of these analyses are discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
+A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.
 
@@
 \begin{itemize}
-\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
-\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
+\item Risk of measles infection decreases significantly with age
+\item Pacific people are statistically more at risk on a per capita basis
+\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
+\item There is some statistical support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
 \end{itemize}
@@
 
+In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the benefit--cost analysis (Section~\ref{sub:cost_benefit}).
+
 A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
-The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
-The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{PopulationNaive.pdf}
-     \caption{test}
+The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-020}
+\end{center}
+\caption{New Zealand population by age and estimates numbers of naive people in each age class}
 \label{fig:naive}
 \end{figure}
-
-The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
-Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
-The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
-The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
-
-These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
-In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
-All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
-They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
-
-Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
-The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
 
@@
 \begin{equation} \label{eq:pc}
-p_c = 1-(1/R0)
+p_c = 1-(1/R_0)
   \end{equation}
 
-We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population, estimated from the national level vaccination coverage data provided by the Ministry of Health for this work.
-Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one.
-
-And discuss the implications for this simplification in the discussion. However, briefly, there are several caveats to our use of this equation:
-\begin {itemize}
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
-\end{itemize}
-
-discussion of simplifications and use of other av:
-\begin {itemize}
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
-\end{itemize}
-
-We then used the proportions of the population that were estimated be require vaccination as estimated above, and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand.
+We estimated this from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
+Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
+
+We then used the proportions of the population that were estimated to require vaccination (estimated above), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).
 
@@
 
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
-The probability density of the $R_v$ estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
-An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report.
+The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one.
+As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+However, the results of these differences are presented in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).
 
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
-The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, and we explore the benefit--cost effects of this in Section~\ref{sub:cost_benefit}.
+
+Again The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
 This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
@@
 
-To use the results from our modeling exercise to help inform the appropriate measles vaccination coverage.
-Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 to 53\%.
-These figures can be reached in a number of different ways, and these are discussed in the cost-benefit section.
+To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage.
+Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
+These figures can be reached in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
 
 \subsection{Summary of modelling}
+These results suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
+However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
+In Isreal researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
+Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+In summary:
 \begin{itemize}
-\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
-\end{itemize}
-
-\subsection{Future modelling}
-Future modelling we aim to perform are:
-\begin{itemize}
-\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
-\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
+ \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
+ \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
 \end{itemize}
@@
 \subsection{Cost analyses results}
+\label{sub:cost_benefit}
 Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
@@
 \begin{center}
-\includegraphics{interimreport2-039}
+\includegraphics{interimreport2-028}
 \end{center}
@@
 This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008 - 2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
-
-To estimate the benefits from additional vaccinations, as estimated from the above modeling section, we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ in the estimated naive population from the modeling section, assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
+Overall, the cost per case from 2008--2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
+
+\subsection{Benefit--cost analyses methods}
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section, we did several things.
+Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated naive population of New Zealand (Figure~\ref{fig:naive}), assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
 We used these values of numbers of predicted cases and the cost figures above to estimate the cost of not vaccinating additional populations.
-We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would now be one, though where our estimates for $R_v$ are < 1, we use this.
+We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this.
 We can therefore estimate the number of cases prevents, and the savings made from the additional vaccinations.
@@
 \begin{equation} \label{eq:reg}
-discount per year = cost * 1/1.08^(year-1))
+discount/year = cost * 1/1.08^{year-1}
   \end{equation}
 
-Our estimates for the numbers of measles cases are highly variable, though typically in our model, if $R_v$ were greater than one, either the measles outbreak would take off and all the New Zealand population would become infected, or not, and we used the average of 1000 simulations to estimate the mean of this.
+Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the New Zealand population would become infected, or it would stutter after a few infects and fail to take off due to stochasticity.
+We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
 However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes.
-We therefore also compared our results to those using the mean number of measles cases seen in New Zealand per yea, since 1997.
+We therefore also compared our results to those using the mean number of measles cases seen in New Zealand per year, since 1997.
 This was 220 cases per year.
 
-The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure
+The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure~\ref{fig:numvacc}.
 
@@
      \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
-     \caption{test}
+     \caption{Numbers of the currently naive New Zealand population to ensure $R_v$ is less than one, and thus all measles introductions will fail to persist within the population.
+$R_v$ estimates are from the outbreaks in Figure ~\ref{fig:outbreaks} and shown in Figure ~\ref{fig:r0}}
      \label{fig:numvacc}
 \end{figure}
-
-The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure N.
 
@@
      \centering
-     \includegraphics[width=0.9\textwidth]{simulations.pdf}
+\includegraphics{interimreport2-034}
      \caption{test}
@@
 
-The benefit-cost results are in Table N and N.
+The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim}.
+These simulations show that even in scenarios when $R_v$ is one and thus stochastically should fail to persist for long, large outbreaks can occur.
+In this modelling exercise, the median value of the 1000 simulations was 2 cases, but the mean 175 and the maximum nearly 25,000 cases.
+
+\begin{figure}
+     \centering
+\includegraphics{interimreport2-036}
+     \caption{test}
+     \label{fig:sim}
+\end{figure}
+
+\begin{figure}
+     \centering
+\includegraphics{interimreport2-037}
+     \caption{test}
+     \label{fig:sim}
+\end{figure}
+
+For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
+Specifically, used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
+We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
+We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
+
+\subsection{Benefit--cost analyses results}
 
@@
     \begin{tablenotes}\footnotesize  
+        \item $\dagger$ {1000 simulations}
         \item \sym{*} Proportion of cases hospitalised 0.17
@@
     \end{threeparttable}  
-  \label{tab:xx}
 \end{sidewaystable}
@@
 \normalsize
-and the proportions of each age category shown in N.
-
-However, it is worth noting that vaccination strategies that target the very young (<1) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure ) were vaccinates with a signle vaccine at a very young age (Figure N).
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{vaccinationstatusage.pdf}
-     \caption{test}
+
+The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}.
+The results in the two tables show two alternative trends.
+In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
+However, if the previous recent history of cases since 1997 is the 'status quo' and what we may expect in the future then the \emph{additional} effort to vaccinate the naive population of New Zealand based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r0}) is not a cost effective exercise (Table~\ref{tab:cost_ave}).
+
+It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) and that they were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-038}
+\end{center}
+\caption{Age and vaccination status of cases}
+\label{fig:ageandvac}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{interimreport2-039}
+\end{center}
+\caption{Age of vaccination of vaccinated measles cases}
 \label{fig:vaccstat}
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{AgeOfVaccinationVaccinated.pdf}
-     \caption{test}
-     \label{fig:ageofvacc}
-\end{figure}
-
-\subsection{Cost analysis discussion}
+\subsection{Benefit--cost analysis discussion}
 The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
-While some of the data are complete and detailed, this is not true of all the data.
-In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
-For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
-In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
-In addition, similar data would be needed for non-publicly funded hospitals, e.g.
-private clinics.
-Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
-
-Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
-Similar data are needed for the period preceding 2014.
-If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
-However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
-As other studies have demonstrated, direct costs required to manage measles are not linear.
-
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin02}.
-This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
-The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
-In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
-They estimated direct costs per case to be less than US\$500.
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
-In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
-In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
-The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
-
-\subsection{Cost analysis summary}
+The majority of the cost analyses were discussed in our previous interim report and though some details have changed, we will focus here on the benefit--cost analyses.
+However, for all analyses more detailed costs of all aspects of hopsitalisation would be beneficial.
+So too would information of the cost of measles vaccination in New Zealand.
+
+We have used standard discounting from healthcare of 8\% per annum, but again there may be official New Zealand discounting rates that we are unaware of.
+For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 and are included in our benefit--cost analsyes.
+It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
+
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin2}.
+In our analyses if we include loss of earnings and hospitalisation costs, our estimates are higher, around US\$959 per case.
+This figure is from NZ\$ 852 for loss of wages and 0.17 * NZ \$ 1710 for costs related to attending hospital.
+However, costs can vary widely, with the containment of a single case and two secondary cases of measles in 2004 in Iowa, USA, estimated to cost US\$142,542 alone \citep{dayan5}.
+
+Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the naive New Zealand population until overyone is infected.
+Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
+However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimates to be more than 18 (i.e.
+1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}) and the benefits of catch up vaccination are clear (Table~\ref{tab:cost_sim}).
+If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r0}), then there is little financial benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}), though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
+
+Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios, such as the size of the local naive population.
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim}).
+
+\subsection{Benefit--cost analysis summary}
 \begin{itemize}
-\item Our initial estimates suggest the ongoing 2013--2014 measles outbreak has cost New Zealand over \$750,000.
+\item The mean wage losses per measles case is estimated to be \$852
+\item The mean cost per measles case attending hospital is estimated to be \$1710
+\item Approximately 17\% of measles cases attend hopsital
+\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
+\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
+\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
 \end{itemize}
-
-\subsection{Future cost-benefit analyses}
-Using the results above we aim to:
-\begin{itemize}
-\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses (section~\ref{sub:risk_analyses}) or modelling (section~\ref{sec:epidemic_modelling}).
-\end{itemize}
-
-We also require additional clarifications of the data, regarding:
-\begin {itemize}
-\item What hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight? Or, does it mean the case stayed for less than 24 hours?
-\end {itemize}
-Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
 
@@
 \begin{itemize}
-\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
-\item The cost of the current measles outbreak is estimated to be at least \$750,000.
-\item Analyses of outbreak data suggest that measles $R_v$ values often include 1 and in this year, 2014, are well above one.
-This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
+\item Risk of measles infection decreases significantly with age
+\item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
+\item Pacific and Asian children in the 6--17 year age caterories have been at lower risk than European or Maori children of the same age
+\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
+\item The mean wage losses per measles case is estimated to be \$852
+\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hopsital
+\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
+\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
+\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
 \end{itemize}

### 2014-09-01 estimatingR0/estimate_r0.R

@@ -29,7 +29,7 @@ for (i in 1:length(outbreak_files))
 }
 
 outbreak_files <- outbreak_files[order(max_dates)]
-
+cols <- c("#FF0000FF","#CCFF00","#66FF33","#0066FF","#9900CC")
 # function for modifying colours to make them transparent
 alpha <- function(col, a)
 {
@@ -38,7 +38,6 @@ alpha <- function(col, a)
   else
     rgb(t(col2rgb(col)/255), alpha=a)
 }
-cols <- rainbow(length(average_R0))
 
 # do the separate analyses
 average_R0 <- list()
@@ -97,7 +96,6 @@ for (i in 1:length(outbreak_files))
 
   average_R0[[length(average_R0)+1]] <- estR0$R0
 }
-
 
 # plot R0 averages
 

### 2014-09-01 interimreport2.Rnw
Draft final interim report 2

@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses measles outbreaks, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
-We provide preliminary benefit--cost analyses using the results from those simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
+In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
+We provide preliminary benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
 Our key findings were:
 \begin{itemize}
-\item Age is the best predictor of risk of measles infection in multi-variate regression analyses, though some groups such as Pacific people, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
+\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
 \item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
-\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is very beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
+\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
 \end{itemize}
@@
 Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, found the report to be very thorough, however, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
-The additional analyses included in this section are multivariate modelling to account for confounding within the univariate analyses.
+In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, we found the report to be very thorough, however, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
+The additional analyses included in this section are multivariate regression modelling to account for confounding within the univariate analyses.
 
@@
 This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: 1--5 and 6--10.
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: NZDep 1--5 and NZDep 6--10.
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
@@
 
-The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:case}).
+The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see "None", Table~\ref{table:case}).
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
      xlab="Category",cex.lab=2)
-legend("topleft",c("Asian","European","Maori","MELAA","Pacific"),
+legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
        col=c(1:4,6),pch=16,bty="n",cex=1.8)
@@
 
-However, there are several issues with the data for Middle Eastern, Latin American and African (MELAA) ethnicities, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
+However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
@@
 
-For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used, and present, a quasipoission regression model.
+For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model.
 We also account for differences in population sizes by using an offset term, the log(population size).
@@
 @
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-%@
-%\end{center}
-%\caption{test}
-%\label{fig:test}
-%\end{figure}
-
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
-%@
-%\end{center}
-%\caption{Cases per capita for Prioritized Ethnicities.
-Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
-Points are coloured by Ethnicity.}
-%\label{fig:test}
-%\end{figure}
-%
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%res<-predict(model2)
-%plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
-%@
-%\end{center}
-%\caption{test}
-%\label{fig:test}
-%\end{figure}
-
-%<<echo=FALSE,include=FALSE,results=hide>>=
-%cor(exp(res),tp$cases)
-%cor.test(exp(res),tp$cases)
-%@
-%
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-%@
-%\end{center}
-%\caption{test}
-%\label{fig:test}
-%\end{figure}
-
-%\section{Regression analyses results}
-%\label{sub:regression_results}
-%The results are:
-
-%<<label=regmodel,include=FALSE>>=
-%summary(model2)
-%@
-
-%<<label=regmodelanova,include=FALSE>>=
-%anova(model2,test="F")
-%@
 
@@
 
-The distribution of the cases per category used are in Figure~\fig{fig:histcasecat}.
+The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
 \begin{figure}[h!]
@@
 
-The predicted values plotted against the cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
+The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
 \begin{figure}[h!]
@@
 \end{figure}
-The significance of the different predictor variables can be seen here, in the ANOVA results:
+The significance of the different predictor variables can be seen in the ANOVA results:
 
@@
 
-The results of the model suggest that, apart from over-representation of some MELAA categories discussed above and not included here, age is a strong predictor of being a measles case.
+The results of the model suggest that, apart from over-representation of some MLA categories discussed above and not included here, age is a strong predictor of being a measles case.
 Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
@@
 In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infections, but older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories, but interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
+Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
 
@@
 
-The analyses suggest that age is a particularly strong risk factor for measles.
+The regression analyses suggest that age is a particularly strong risk factor for measles.
 This comes as no surpise to epidemiologists or health care providers.
 However, our analyses also highlight other groups that are at risk.
-In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European 6--17 year old children.
-The results of these analyses are discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit}) Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
+Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses.
+Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
 A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.
+
+The results of these regression analyses are also discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit})
 
@@
 The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles.
+Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
+The figure shown (Figure~\ref{fig:naive}) and for all subsequent analyses assume a currently susceptible population in New Zealand of approximately 11\% of the total population.
 
@@
 The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
-If $R_v < 1$ epidemics are prevented.
+If $R_v < 1$ epidemics are predicted to be prevented (but see below) and infection persistence is prevented.
 The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
@@
 
-To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
+To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we previously estimated $R_v$ from all the outbreaks in New Zealand since 2009.
 To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
@@
 
-We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand.
-We estimated this from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
+We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
+We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
 Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
 
-We then used the proportions of the population that were estimated to require vaccination (estimated above), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+We then used the proportions of the population that were estimated to require vaccination (Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
 These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).
 
-\subsection{Modelling results}
+finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of suscpetible people in New Zealand was now reduced by the numbers vaccinated.
+Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+
+\subsection{Modelling results and discussion}
 
@@
 The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one.
-As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
-However, the results of these differences are presented in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).
+As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
+However, the results of these differences are presented and discussed in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).
 
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, and we explore the benefit--cost effects of this in Section~\ref{sub:cost_benefit}.
-
-Again The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we explore the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+
+The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
 This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
@@
 
-To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage.
+To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:pc}.
 Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
@@
 
+The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
+However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
+Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
+Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+
 \subsection{Summary of modelling}
-These results suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
-However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
-In Isreal researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
-Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
-In summary:
 \begin{itemize}
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
+ \item The reproduction number, $R_v$, for measles in the current 2013--2014 outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
  \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
@@
 The cost was \$256 less if the case were a male.
-It increased of approximately \$406 per night of hospitalisationand \$64 per year over the time period of 2000 - 2014.
+It increased of approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014.
 The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.
@@
 \subsection{Benefit--cost analyses methods}
-To estimate the benefits from additional vaccinations, as estimated from the above modelling section, we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated naive population of New Zealand (Figure~\ref{fig:naive}), assuming recovery from infection lead to immunity and thus constantly reducing the population size by the number recovered.
-We used these values of numbers of predicted cases and the cost figures above to estimate the cost of not vaccinating additional populations.
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.
+Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}), assuming recovery from infection led to immunity and thus constantly reducing the population size by the number recovered.
+We used these values of numbers of predicted cases and the cost figures above to estimate what the cost of not vaccinating additional populations would be.
 We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this.
-We can therefore estimate the number of cases prevents, and the savings made from the additional vaccinations.
-We then use the costs of the catch up vaccination schemes and the costs of the expected measles-related costs due to constant introduction of measles and the savings from reduced measles cases to work out the benefit to cost ratio.
-For our benefit to cost ratio to be beneficial, this must be greater than 1.
-A value less than one, suggests the costs are higher than the economic benefits.
-Lastly, because there is a discounting effect with time, we use an 8\% discount per year for the costs, as is common for health care discounting, and we calculate these over a 10 year time frame, using a discounting such that:
-\begin{equation} \label{eq:reg}
-discount/year = cost * 1/1.08^{year-1}
+We can therefore estimate the number of cases prevented, and the savings made from the additional vaccinations.
+We then use the costs of the catch up vaccination schemes, the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity, and the savings from reduced measles cases to work out the benefit--cost ratio ((B/C), where a B/C ratio > 1 means that the program benefits exceed their costs.
+A B/C value less than one suggests the costs are higher than the economic benefits.
+Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}:
+
+\begin{equation} \label{eq:disc}
+\textit{present value} = \sum \textit{future value}_y_r * (1/(1+\textit{discount rate})^{yr}
   \end{equation}
-
-Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the New Zealand population would become infected, or it would stutter after a few infects and fail to take off due to stochasticity.
+Where $_y_r$ is the year 0 (current) through to 9 (ten years into the future).
+
+Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the susceptible New Zealand population would become infected during the outbreak, or the epidemic it would stutter after a few infects and fail to cause a large epidemic due to stochasticity.
 We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
-However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes.
-We therefore also compared our results to those using the mean number of measles cases seen in New Zealand per year, since 1997.
+However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
+However, to address this, we not only use $R_v$ values of 1, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year, since 1997.
 This was 220 cases per year.
 
-The numbers to vaccinate in New Zealand assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates are shown in Figure~\ref{fig:numvacc}.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{NumbersToVaccinate.pdf}
-     \caption{Numbers of the currently naive New Zealand population to ensure $R_v$ is less than one, and thus all measles introductions will fail to persist within the population.
-$R_v$ estimates are from the outbreaks in Figure ~\ref{fig:outbreaks} and shown in Figure ~\ref{fig:r0}}
-     \label{fig:numvacc}
-\end{figure}
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates (Section~\ref{sec:epidemic_modelling}) are shown in Figure~\ref{fig:numvacc}.
 
@@
 @
-     \caption{test}
-     \label{fig:sim}
+     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling})}
+     \label{fig:numvac}
 \end{figure}
 
-The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim}.
-These simulations show that even in scenarios when $R_v$ is one and thus stochastically should fail to persist for long, large outbreaks can occur.
-In this modelling exercise, the median value of the 1000 simulations was 2 cases, but the mean 175 and the maximum nearly 25,000 cases.
+The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
+These simulations show that even in scenarios when $R_v$ is one, and thus stochastically should fail to persist for long, large outbreaks can occur due to stochastic processes.
+Though the median value from 1000 simulations is low (2 cases), and thus most measles introductions will be single cases or lead to minor outbreaks, in this modelling exercise the mean value was 151 cases and the maximum nearly 25,000 cases.
 
@@
 @
-     \caption{test}
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
+The full distribution of results can be seen in Figure~\ref{fig:sim1}}
      \label{fig:sim}
@@
 @
-     \caption{test}
-     \label{fig:sim}
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
+A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
+     \label{fig:sim1}
 \end{figure}
@@
 For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
-Specifically, used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
+Specifically, we used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
 We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
@@
   \centering  
+  \begin{raggedleft}
   \caption{Cost benefit analyses using simulated epidemic sizes}  
+  \label{tab:cost_sim}
    \begin{threeparttable}[b] \tiny
@@
     \toprule
-        Years R0 estimated from & R0 range estimated	& R0  for simulations	& Mean # of cases in the population \dagger	& Vaccines required to reduce R0 to 1	& Costs per vaccine	& Total vaccine costs (\$)	& Total hospitalised cases	(\$)\sym{*}& Total costs for cases (\$)\sym{**}	& R0  for simulations after response	& Mean cases in the population after catch up \dagger	& Total cases over 10 years after action \sym{***}	& Total hospitalised cases after action	& Total costs for cases after action (\$)	& Cases reduced due to action	& Savings (\$)	& Benefit--cost ratio\\
+
+Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
 \midrule
-    2009-2014 & 0.92-1.19 & 0.92	& 13	& 0	& 0 & 2	& 10765	& 0.92	& 13	& 130	& 22	& 107653	& 	& 	&  \\
-2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 20	& 1159502	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1449178& 127151	& 105293982	& 40.36 \\
-2009-2014	& 0.92-1.19	& 1.19	& 128901	& 80000	& 50	& 2898755	& 21913	& 147295173	& 1	& 175	& 1750	& 298	& 1449178	& 127151	& 105293982	& 24.22 \\
-2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 20	& 3016923	& 57805	& 388554566	& 1	& 116 &	1160	& 197	& 960598 & 338872	& 280620541	& 70.55 \\
-2013-2014	& 1.82-2.13	& 1.82	& 340032	& 208153	& 50	& 7542307	& 57805	& 388554566	& 1	& 116	& 1160	& 197	& 960598	& 338872	& 280620541	& 33 \\
-2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 20	& 3660563	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 960598	& 380914	& 315435600 & 68.26 \\
-2013-2014	& 1.82-2.13	& 2.13	& 382074	& 252561	& 50	& 9151406	& 64953	& 436595960	& 1	& 116	& 1160	& 197	& 960598	& 380914	& 315435600	& 31.19 \\
+2009-2014 &	0.92-1.19	& 0.92	& 0	    & na	  & 0	      & 130	    & 22  	& 148551	  & 130519	  & 13	& 130	  & 22	& 148551	& 0	      & 0	        & 0	        & na \\
+2009-2014	& 0.92-1.19	& 1.19	& 80000	& 20	  & 1600000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 126359360	& 79.97 \\
+2009-2014	& 0.92-1.19	& 1.19	& 80000	& 50	  & 4000000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 123959360	& 31.99 \\
+2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 336060587	& 81.72 \\
+2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 329815997	& 32.69 \\
+2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 377382119	& 75.71 \\
+2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 369805289	& 30.28 \\
+
+\end{raggedleft}
 \bottomrule  
@@
     \end{threeparttable}  
-  \label{tab:cost_sim}
 \end{sidewaystable}
@@
   \begin{raggedright}
-  \caption{Cost benefit analyses using simulated epidemic sizes}  
+  \caption{Cost benefit analyses using historic epidemic sizes since 1997}  
   \label{tab:cost_ave}
@@
     \toprule  
-        Years R0 estimated from &R0 range estimated	&R0 for simulations	&Vaccines required to reduce R0 to 1	&Costs per vaccine (\$)	&Total vaccine costs (\$)	&Expected number of cases expected based on average since 1997 \sym{***}	&Total hospitalised cases	\sym{*}&Total costs for cases (\$)\sym{**}	&R0  for simulations after catch up	&Mean number of cases in the population after action (1000 simulations)	&Total cases over 10 years after action \sym{***}	&Total hospitalised cases after action	&Total costs for cases after action (\$)	&Cases reduced due to action	&Savings (\$)	&Benefit--cost ratio\\
+Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
 \midrule
-2009-2014 &0.92-1.19	&0.92	&0	&	&0	&2200	&374	&1821824	&0.92	&13	&130	&22	&107653	&	&	& \\
-2009-2014	&0.92-1.19	&1.19	&80000	&20	& 1159502	&2200	&374	&2513940	&1	&175	&1750	&298	&1449178	&450	&372646	&0.14\\
-2009-2014	&0.92-1.19	&1.19	&80000	&50	& 2898755  &2200	&374	&2513940	&1	&175	&1750	&298	&1449178	&450	&372646 &0.09\\
-2013-2014	&1.82-2.13	&1.82	&208153	&20	& 3016923	&2200	&374	&2513940	&1	&116	&1160	&197	&960598	&1040	&861226	&0.22\\
-2013-2014	&1.82-2.13	&1.82	&208153	&50	& 7542307 &2200	&374	&2513940	&1	&116	&1160	&197	&960598 &1040	&861226  &0.1\\
-2013-2014	&1.82-2.13	&2.13	&252561	&20	& 3660563	&2200	&374	&2513940	&1	&116	&1160	&197	&960598	&1040	&861226  &0.19\\
-2013-2014	&1.82-2.13	&2.13	&252561	&50	& 9151406	&2200	&374	&2513940	&1	&116	&1160	&197	&960598	&1040	&861226  &0.09\\
+
+2009-2014	& 0.92-1.19	& 0.92	& 0	      & na	& 0	      & 130	  & 22	& 148551	  & 130519	& 13	& 130	  & 22	& 148551	& 0	  & 0	      & 0	        & na \\
+2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 20	& 1600000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -847009	  & 0.47 \\
+2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 50	& 4000000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -3247009	& 0.19 \\ 
+2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -3118912	& 0.25 \\
+2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -9363502	& 0.10 \\
+2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -4007072	& 0.21 \\
+2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -11583902	& 0.08 \\
 \end{raggedright}
@@
 In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
-However, if the previous recent history of cases since 1997 is the 'status quo' and what we may expect in the future then the \emph{additional} effort to vaccinate the naive population of New Zealand based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r0}) is not a cost effective exercise (Table~\ref{tab:cost_ave}).
-
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) and that they were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
+However, if the previous recent history of cases since 1997 is the 'status quo' and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r0}) is not a cost effective exercise (Table~\ref{tab:cost_ave}).
+
+It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
 
@@
 
-We have used standard discounting from healthcare of 8\% per annum, but again there may be official New Zealand discounting rates that we are unaware of.
-For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 and are included in our benefit--cost analsyes.
+We have used standard discounting from healthcare of 3\% per annum \citep{honeycutt6}, but again there may be official New Zealand discounting rates that we are unaware of.
+For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014.
+Including these could be done, but we would like some further data of these before including them in our benefit--cost analsyes.
 It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
@@
 
-Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the naive New Zealand population until overyone is infected.
+Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the presently naive and susceptible New Zealand population until overyone has been infected.
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
-However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimates to be more than 18 (i.e.
-1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}) and the benefits of catch up vaccination are clear (Table~\ref{tab:cost_sim}).
+However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimated to be more than 18 (i.e.
+1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}).
+The benefits of catch up vaccination are clear once $R_v$ is greater than one (Table~\ref{tab:cost_sim}).
 If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r0}), then there is little financial benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}), though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
@@
 Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios, such as the size of the local naive population.
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim}).
+The spatial effects of measles transmission may have affected both our multivariate regression analyses (\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises.
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
 
@@
 \item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
-\item Pacific and Asian children in the 6--17 year age caterories have been at lower risk than European or Maori children of the same age
+\item Pacific and Asian children in the 6--17 year age caterories have been at lower risk of measles than European or Maori children of the same age
 \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
@@
 
+\bibitem[Honeycutt et al.(2006)]{honeycutt6}
+Honeycutt, A. A., L. Clayton, O. Khavjou, E.~A. Finkelstein, M. Prabhu, J.~L. Blitstein, W. Dougles Evans, and J.~M. Renaud (2006).
+\newblock Guide to Analyzing the Cost-Effectiveness of Community Public Health Prevention Approaches.
+http://aspe.hhs.gov/health/reports/06/cphpa/report.pdf
+
 \bibitem[Klinkenberg et al.(2011)]{klinkenberg11}

### 2014-09-01 interimreport2.tex
Draft final interim report 2

@@
 
-The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MELAA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (Table~\ref{table:case}).
+The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see "None", Table~\ref{table:case}).
 
@@
 
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-%@
-%\end{center}
-%\caption{test}
-%\label{fig:test}
-%\end{figure}
-
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%plot(tp$cases,main="Cases per category",ylab="Count",pch=16,col="darkgrey")
-%@
-%\end{center}
-%\caption{Cases per capita for Prioritized Ethnicities.
-Note categories are not given, but include Age (0-2, 3-5, 6-17, 18-24, and 25+ in years), NZ deprivation index (1-5, 6-10), and Ethnicity.
-Points are coloured by Ethnicity.}
-%\label{fig:test}
-%\end{figure}
-%
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%res<-predict(model2)
-%plot(exp(res),tp$cases,xlab="results",ylab='predictions',main="Fit",pch=16,col="darkgrey")
-%@
-%\end{center}
-%\caption{test}
-%\label{fig:test}
-%\end{figure}
-
-%<<echo=FALSE,include=FALSE,results=hide>>=
-%cor(exp(res),tp$cases)
-%cor.test(exp(res),tp$cases)
-%@
-%
-%\begin{figure}[h!]
-%\begin{center}
-%<<fig=TRUE,echo=FALSE,width = 11>>=
-%hist(model2$residuals,main="Histogram of residuals",xlab="residuals",col="grey")
-%@
-%\end{center}
-%\caption{test}
-%\label{fig:test}
-%\end{figure}
-
-%\section{Regression analyses results}
-%\label{sub:regression_results}
-%The results are:
-
-%<<label=regmodel,include=FALSE>>=
-%summary(model2)
-%@
-
-%<<label=regmodelanova,include=FALSE>>=
-%anova(model2,test="F")
-%@
-
 \subsection{Regression analyses results}
@@
 \includegraphics{interimreport2-034}
-     \caption{test}
-     \label{fig:sim}
-\end{figure}
-
-The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim}.
-These simulations show that even in scenarios when $R_v$ is one and thus stochastically should fail to persist for long, large outbreaks can occur.
-In this modelling exercise, the median value of the 1000 simulations was 2 cases, but the mean 175 and the maximum nearly 25,000 cases.
+     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling})}
+     \label{fig:numvac}
+\end{figure}
+
+The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
+These simulations show that even in scenarios when $R_v$ is one, and thus stochastically should fail to persist for long, large outbreaks can occur due to stochastic processes.
+Though the median value from 1000 simulations is low (2 cases), and thus most measles introductions will be single cases or lead to minor outbreaks, in this modelling exercise the mean value was 151 cases and the maximum nearly 25,000 cases.
 
@@
 \includegraphics{interimreport2-036}
-     \caption{test}
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
+The full distribution of results can be seen in Figure~\ref{fig:sim1}}
      \label{fig:sim}
@@
 \includegraphics{interimreport2-037}
-     \caption{test}
-     \label{fig:sim}
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
+A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
+     \label{fig:sim1}
 \end{figure}

### 2014-09-01 interimreport2.Rnw
Draft final revised

@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
+In this interim report we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations.
 We provide preliminary benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
@@
 \begin{itemize}
-\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
+\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
 \item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
-\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
+\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
 \end{itemize}
@@
 Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, we found the report to be very thorough, however, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
-The additional analyses included in this section are multivariate regression modelling to account for confounding within the univariate analyses.
+In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, we found the report to be very thorough.
+However, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
+The additional analyses included in this section are multivariate regression modelling to account for confounding within the univariate analyses, particularly age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep).
 
@@
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
+Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
 Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
 These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
-We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist among cases allowing better understanding of risk of measles infection.
+We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.
 
@@
 The ethnicity denominator data were not Prioritized Ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes.
-After discussions with the University Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.
-
-With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, this led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
+After discussions with the University of Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.
+
+With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories.
+Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago demoninator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
 
@@
 This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: NZDep 1--5 and NZDep 6--10.
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 and NZDep 6--10.
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
@@
 
-The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see "None", Table~\ref{table:case}).
+The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
 PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita,4))
-colnames(PerCap)[4]<-"Population"
-colnames(PerCap)[5]<-"Cases"
-colnames(PerCap)[6]<-"Per Capita"
-@
-
-The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
+colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
+@
+
+The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
 Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
+Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).
 
@@
 \end{center}
-\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
+\caption{Per capita rates of measles infections broken down by ethnicity, see Table~\ref{table:percap} for details}
 \label{fig:percap}
@@
 \begin{table}\tiny
-\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
 <<results=tex,echo=FALSE>>=
@@
 
-However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
-Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).
-
 For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model.
 We also account for differences in population sizes by using an offset term, the log(population size).
-We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms.
+We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms.
 Thus, the final model that remained with all significant interaction terms had the following linear predictor:
@@
 \begin{equation} \label{eq:reg}
- log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
+ log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a * x_e)+ log(population)  + \epsilon
   \end{equation}
@@
 \end{center}
-\caption{Histogram of cases per category used in the final regression model}
+\caption{Distribution of cases per category used in the final regression model}
 \label{fig:histcasecat}
@@
 \end{center}
-\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
+\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
 \label{fig:predict}
@@
 
-A summary of the model with the individual effects and the statistical support for the estimated coefficients can be seen here:
+A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
 
@@
 
-The results of the model suggest that, apart from over-representation of some MLA categories discussed above and not included here, age is a strong predictor of being a measles case.
+Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
 Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
@@
 People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-
 In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
@@
 This comes as no surpise to epidemiologists or health care providers.
-However, our analyses also highlight other groups that are at risk.
+However, our analyses also highlight other groups that are at greater risk of measles.
 In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
@@
 A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.
-
+These analyses are possible future directions for this work.
 The results of these regression analyses are also discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit})
@@
 \item Risk of measles infection decreases significantly with age
-\item Pacific people are statistically more at risk on a per capita basis
+\item Pacific people are statistically more at risk of measles infection
 \item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
@@
 Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
-The figure shown (Figure~\ref{fig:naive}) and for all subsequent analyses assume a currently susceptible population in New Zealand of approximately 11\% of the total population.
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-#length(pop)
-naive<-round(pop-(pop*impop))
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
-\end{center}
-\caption{New Zealand population by age and estimates numbers of naive people in each age class}
-\label{fig:naive}
-\end{figure}
+The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, estimated using the NZ census data \citep{stats14} and the proportion of each age class immune based on data provided to us by the Ministry of Health on commencing this project.
 
@@
      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks}}
+     \caption{Estimates of $R_v$ (Average $R_0$) for the outbreaks each year, as classified by outbreaks in Figure~\ref{fig:outbreaks}}
      \label{fig:r0}
@@
   \end{equation}
-
-We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
+To acocunt for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.
+
+We then assume that the measles epidemics are occuring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
 We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
-Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
-
-We then used the proportions of the population that were estimated to require vaccination (Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+require(reshape2)  # this is the library that lets one flatten out data
+require(ggplot2)   # plotting library
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+@
+
+\begin{figure}[ht!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+plot(pop,xlab="Age",ylab="Population",cex.lab=2)
+points(naive,pch=16)
+legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
+@
+\end{center}
+\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\label{fig:naive}
+\end{figure}
+
+We then used the proportions of the population that were estimated to require vaccination ($p_c$, Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
 These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).
 
-finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of suscpetible people in New Zealand was now reduced by the numbers vaccinated.
+Finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of susceptible people in New Zealand was now reduced by the numbers vaccinated.
 Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).
@@
 However, the results of these differences are presented and discussed in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).
-
-These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
+The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current outbreak 1.82--2.13.
+
+The regularity of these outbreaks also imply that the regular (approximately yearly) importation of measles is an ongoing process.
 Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we explore the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
@@
 To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:pc}.
-Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
-These figures can be reached in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, lower 95\% confidence limit) to an upper 95\% confidence limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
+These additional vaccination numbers can be made up in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
 
@@
 \end{center}
-\caption{Number of cases attending hospital and stay duration}
+\caption{Number of measles cases attending hospital and stay duration from 2000--2014}
 \label{fig:hosp}
@@
 
-\begin{table}
+\begin{table}\small
 \caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
@@
 To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}), assuming recovery from infection led to immunity and thus constantly reducing the population size by the number recovered.
+Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ distribution from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}).
+We assumed recovery from infection led to immunity and thus constantly reduced the population size by the number recovered.
 We used these values of numbers of predicted cases and the cost figures above to estimate what the cost of not vaccinating additional populations would be.
@@
 However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
-However, to address this, we not only use $R_v$ values of 1, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year, since 1997.
+However, to address this, we not only use $R_v$ values of 1 in simulation models to estimate predicted outbreak sizes, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year since 1997.
 This was 220 cases per year.
 
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates (Section~\ref{sec:epidemic_modelling}) are shown in Figure~\ref{fig:numvacc}.
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population and using the upper 95\% confidence limits of our $R_v$ estimates (Section~\ref{sec:epidemic_modelling}) are shown in Figure~\ref{fig:numvac}.
 
@@
 @
-     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling})}
+     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave}}
      \label{fig:numvac}
@@
 The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
-These simulations show that even in scenarios when $R_v$ is one, and thus stochastically should fail to persist for long, large outbreaks can occur due to stochastic processes.
+These simulations show that even in scenarios when $R_v$ is one, and thus stochastically should fail to persist (i.e.
+become endemic), large outbreaks can occur due to stochastic processes.
 Though the median value from 1000 simulations is low (2 cases), and thus most measles introductions will be single cases or lead to minor outbreaks, in this modelling exercise the mean value was 151 cases and the maximum nearly 25,000 cases.
@@
   \centering  
-  \begin{raggedleft}
+  \begin{raggedright}
   \caption{Cost benefit analyses using simulated epidemic sizes}  
@@
 
-Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
+Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \dagger&	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
 \midrule
@@
 
-\end{raggedleft}
+\end{raggedright}
 \bottomrule  
@@
         \item \sym{*} Proportion of cases hospitalised 0.17
-        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
+        \item \sym{**} Wage losses per case = \$852 and cost per hospitalised case = \$1710
         \item \sym{***} Based on 10 introductions of measles, one per year
@@
     \toprule  
-Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
+Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \ddagger &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
 \midrule
@@
     \begin{tablenotes}\footnotesize  
+        \item $\ddagger$ Based on average number of cases per year in New Zealand from 1997 onward
         \item $\dagger$ {1000 simulations}
@@
 In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
-However, if the previous recent history of cases since 1997 is the 'status quo' and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r0}) is not a cost effective exercise (Table~\ref{tab:cost_ave}).
+However, if the previous recent history of cases since 1997 is the \emph{status quo} and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one is not a cost effective exercise (Table~\ref{tab:cost_ave}).
 
@@
 We have used standard discounting from healthcare of 3\% per annum \citep{honeycutt6}, but again there may be official New Zealand discounting rates that we are unaware of.
-For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014.
+For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 (Table ~\ref{table:direct}).
 Including these could be done, but we would like some further data of these before including them in our benefit--cost analsyes.
@@
 
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin2}.
+In other nations' outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin2}.
 In our analyses if we include loss of earnings and hospitalisation costs, our estimates are higher, around US\$959 per case.
@@
 Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios, such as the size of the local naive population.
-The spatial effects of measles transmission may have affected both our multivariate regression analyses (\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises.
+The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{sec:epidemic_modelling}).
 Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
@@
 \item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
-\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
+\item Large outbreaks, with a mean size of approximately 151 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
 \end{itemize}
@@
 \item Pacific and Asian children in the 6--17 year age caterories have been at lower risk of measles than European or Maori children of the same age
-\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
+\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, upper 95\% confidencelimit 2013--2014 outbreak, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
 \item The mean wage losses per measles case is estimated to be \$852

### 2014-09-01 interimreport2.tex
Draft final revised

@@
 
-The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see "None", Table~\ref{table:case}).
+The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).
 
@@
 
-The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
+The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
 Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
+Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).
 
@@
 \begin{table}\tiny
-\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
 %latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per Capita}\tabularnewline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
 \hline
@@
 Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
-The figure shown (Figure~\ref{fig:naive}) and for all subsequent analyses assume a currently susceptible population in New Zealand of approximately 11\% of the total population.
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{interimreport2-020}
-\end{center}
-\caption{New Zealand population by age and estimates numbers of naive people in each age class}
-\label{fig:naive}
-\end{figure}
+The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, estimated using the NZ census data \citep{stats14} and the proportion of each age class immune based on data provided to us by the Ministry of Health on commencing this project.
 
@@
   \end{equation}
-
-We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
+To acocunt for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.
+
+We then assume that the measles epidemics are occuring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
 We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
-Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
-
-We then used the proportions of the population that were estimated to require vaccination (Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
+Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
+
+\begin{figure}[ht!]
+\begin{center}
+\includegraphics{interimreport2-021}
+\end{center}
+\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\label{fig:naive}
+\end{figure}
+
+We then used the proportions of the population that were estimated to require vaccination ($p_c$, Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
 These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).
 
-finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of suscpetible people in New Zealand was now reduced by the numbers vaccinated.
+Finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of susceptible people in New Zealand was now reduced by the numbers vaccinated.
 Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).
@@
 4 – 9 years&$  8$&$ 0$&$0.00$\tabularnewline
-10 -1 19 years&$132$&$12$&$0.09$\tabularnewline
+10 – 19 years&$132$&$12$&$0.09$\tabularnewline
 \textgreater  19 years&$ 33$&$11$&$0.33$\tabularnewline
@@
 \begin{center}
-\includegraphics{interimreport2-028}
+\includegraphics{interimreport2-029}
 \end{center}
-\caption{Number of cases attending hospital and stay duration}
+\caption{Number of measles cases attending hospital and stay duration from 2000--2014}
 \label{fig:hosp}
@@
      \centering
-\includegraphics{interimreport2-034}
-     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling})}
+\includegraphics{interimreport2-035}
+     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave}}
      \label{fig:numvac}
@@
      \centering
-\includegraphics{interimreport2-036}
+\includegraphics{interimreport2-037}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
@@
      \centering
-\includegraphics{interimreport2-037}
+\includegraphics{interimreport2-038}
      \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
@@
 \begin{center}
-\includegraphics{interimreport2-038}
+\includegraphics{interimreport2-039}
 \end{center}
@@
 \begin{center}
-\includegraphics{interimreport2-039}
+\includegraphics{interimreport2-040}
 \end{center}

### 2014-09-01 interimreport2.Rnw
minor revisions

@@
 \item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
-\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
+\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
 \end{itemize}
@@
 With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago demoninator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
+Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
 
@@
 \item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
-\item There is some statistical support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
+\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk than European or Maori children
 \end{itemize}
@@
   \end{equation}
-To acocunt for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.
-
-We then assume that the measles epidemics are occuring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
-We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
+To account for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.
+
+We then assume that the measles epidemics are occurring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
+We estimated the susceptible population in New Zealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
 Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
@@
 However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
-Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
+Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
 Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
@@
 The direct medical and public health costs in response to a single case of refugee-imported measles has been reported \citep{coleman12}.
-Costs included labor, translation and benefits for public health workers.
+Costs included labour, translation and benefits for public health workers.
 In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis.
@@
 
-Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \dagger&	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
+Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \dagger&	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
 \midrule
@@
     \toprule  
-Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \ddagger &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
+Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \ddagger &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
 \midrule
@@
 The majority of the cost analyses were discussed in our previous interim report and though some details have changed, we will focus here on the benefit--cost analyses.
-However, for all analyses more detailed costs of all aspects of hopsitalisation would be beneficial.
+However, for all analyses more detailed costs of all aspects of hospitalisation would be beneficial.
 So too would information of the cost of measles vaccination in New Zealand.
@@
 For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 (Table ~\ref{table:direct}).
-Including these could be done, but we would like some further data of these before including them in our benefit--cost analsyes.
+Including these could be done, but we would like some further data of these before including them in our benefit--cost analyses.
 It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
@@
 
-Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the presently naive and susceptible New Zealand population until overyone has been infected.
+Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate among the presently naive and susceptible New Zealand population until everyone has been infected.
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
@@
 
-Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios, such as the size of the local naive population.
+Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local naive population.
 The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{sec:epidemic_modelling}).
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
 
@@
 \item The mean cost per measles case attending hospital is estimated to be \$1710
-\item Approximately 17\% of measles cases attend hopsital
+\item Approximately 17\% of measles cases attend hospital
 \item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
@@
 \item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
-\item Pacific and Asian children in the 6--17 year age caterories have been at lower risk of measles than European or Maori children of the same age
+\item Pacific and Asian children in the 6--17 year age categories have been at lower risk of measles than European or Maori children of the same age
 \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, upper 95\% confidencelimit 2013--2014 outbreak, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
 \item The mean wage losses per measles case is estimated to be \$852
-\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hopsital
+\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
 \item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).

### 2014-10-21 timesseries_measles.R
add New Zealander holidays

@@ -174,3 +174,64 @@ xlim <- range(change[,1])
 par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
 plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
      ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
+
+## set wd to get data
+nzers<-read.csv("monthtravel.csv",header=T)
+head(nzers)
+names(nzers)
+dim(nzers)
+
+plot(as.numeric(nzers[247,2:79]),type="l",ylim=c(min(as.numeric(nzers[247,2:79])),max(ttimes$res)),
+     )
+points(ttimes$res,add=T,type="l",col="red")
+
+par(mar = c(7, 4, 4, 2) + 0.1)
+plot(as.numeric(nzers[247,8:79]),type="l",xaxt="n",xlab="",ylab="Numbers",
+     ,ylim=c(min(as.numeric(nzers[247,2:79])),max(ttimes$res)))
+points(ttimes$res,type="l",col="red")
+# Set up x axis with tick marks alone
+# Create arbitrary text
+labels <- paste(ttimes$mydate, sep = " ")
+axis(side=1, at=1:71, labels=labels,las=2)
+legend("topleft",c("New Zealanders","Non-New Zealanders"),
+       bty="n",col=c("black","red"),lty=1)
+# check same length
+length(nzers[247,8:78])
+length(ttimes$res)
+
+nz<-nzers[247,8:78]
+im<-ttimes$res
+newd<-rbind(nz,im)
+rownames(newd)<-c("nz","im")
+colnames(newd)<-ttimes$mydate
+sum<-colSums(newd)
+newd<-rbind(sum,newd)
+rownames(newd)<-c("total","nz","im")
+
+pdf(paste("nzers.pdf"), width=7, height=5)
+plot(as.numeric(newd[1,]),type="l",xaxt="n",xlab="",ylab="Numbers",
+     ,ylim=c(min(newd),max(newd)))
+points(as.numeric(newd[2,]),type="l",col="red")
+points(as.numeric(newd[3,]),type="l",col="black",lty=2)
+# Set up x axis with tick marks alone
+# Create arbitrary text
+labels <- paste(colnames(newd), sep = " ")
+axis(side=1, at=seq(from=1,to=71,by=2), labels=labels[seq(from=1,to=71,by=2)],las=2)
[…]

### 2014-10-21 FinalSize.tex
Mick final size analyses and text

\documentclass[11pt,a4paper]{article}
\usepackage{amssymb,amsmath, graphicx, caption2}
\usepackage{fancyhdr}
\setlength{\textheight}{20cm}
\setlength{\topmargin}{1.5cm}
\setlength{\textwidth}{13cm}
\addtolength{\hoffset}{-1cm}
\setlength{\parindent}{0pt}
\setlength{\parskip}{10pt}
% The next 3 commands force floats to the pagetop
\makeatletter
\setlength{\@fptop}{0pt}
\makeatother
%\pagestyle {fancyplain}
\pagestyle {myheadings}
\markright{Measles vaccination by DHB}
\newcommand{\vecA}{\mathbf{A}}
\newcommand{\vecB}{\mathbf{B}}
\newcommand{\vecC}{\mathbf{C}}
\newcommand{\vecD}{\mathbf{D}}
\newcommand{\vecE}{\mathbf{E}}
\newcommand{\vecF}{\mathbf{F}}
\newcommand{\vecH}{\mathbf{H}}
\newcommand{\vecI}{\mathbf{I}}
\newcommand{\vecJ}{\mathbf{J}}
\newcommand{\vecK}{\mathbf{K}}
\newcommand{\vecM}{\mathbf{M}}
\newcommand{\vecN}{\mathbf{N}}
\newcommand{\vecP}{\mathbf{P}}
\newcommand{\vecQ}{\mathbf{Q}}
\newcommand{\vecS}{\mathbf{S}}
\newcommand{\vecT}{\mathbf{T}}
\newcommand{\vecW}{\mathbf{W}}
\newcommand{\vecZ}{\mathbf{Z}}
\newcommand{\vecb}{\mathbf{b}}
\newcommand{\vecc}{\mathbf{c}}
\newcommand{\vece}{\mathbf{e}}
\newcommand{\vecf}{\mathbf{f}}
\newcommand{\vecg}{\mathbf{g}}
\newcommand{\vech}{\mathbf{h}}
\newcommand{\veci}{\mathbf{i}}
\newcommand{\vecr}{\mathbf{r}}
\newcommand{\vecu}{\mathbf{u}}
\newcommand{\vecv}{\mathbf{v}}
\newcommand{\vecw}{\mathbf{w}}
\newcommand{\vecx}{\mathbf{x}}
\newcommand{\vecy}{\mathbf{y}}
\newcommand{\vecz}{\mathbf{z}}
\newcommand{\Ach}{\mathcal{H}}
\newcommand{\Ell}{\mathcal{L}}
\newcommand{\En}{\mathcal{N}}
\newcommand{\Pe}{\mathcal{P}}
\newcommand{\Ro}{\mathcal{R}_0}
\newcommand{\Rc}{\mathcal{R}_c}
\newcommand{\Rr}{\mathcal{R}}
\newcommand{\Qu}{\mathcal{Q}}
\newcommand{\Te}{\mathcal{T}}
\newcommand{\Ve}{\mathcal{V}}
\newcommand{\Dub}{\mathcal{W}}
\newcommand{\Zed}{\mathcal{Z}}
\newcommand{\Exp}{\mathbb{E}}
\newcommand{\de}{\mbox{d}}
\newcommand{\wrt}{\,\mbox{d}}
\newcommand{\veczero}{\mathbf{0}}
\newcommand{\vecone}{\mathbf{1}}
\newcommand{\vecalpha}{\mbox{\boldmath$\alpha$}}
\newcommand{\vectheta}{\mbox{\boldmath$\theta$}}
\newcommand{\vecphi}{\mbox{\boldmath$\phi$}}
\newcommand{\vecPhi}{\mbox{\boldmath$\Phi$}}
\newcommand{\vecSigma}{\mbox{\boldmath$\Sigma$}}
\newcommand{\vecimath}{\mbox{\boldmath$\imath$}}
\newcommand{\cosech}{\mbox{cosech}\,}
\newcommand{\sign}{\mbox{sign}\,}
\newcommand{\trace}{\mbox{trace}\,}
\newcommand{\Inf}{\mbox{Inf}\,}
\rhead{\thepage}
\cfoot{}
\renewcommand{\baselinestretch}{1.5}
\begin{document}

\title{A review of measles vaccination status in New Zealand District Health Boards\footnote{FinalSize.tex \today}}
\author{M.G. Roberts \\\\
{\small Institute of Natural \& Mathematical Sciences,}\\
{\small   New Zealand Institute for Advanced Study and the Infectious Disease Research Centre,}\\
{\small   Massey University, Private Bag 102 904, North Shore Mail Centre, Auckland, New Zealand. } \\ 
Email: \texttt{m.g.roberts@massey.ac.nz}}
\date{}
\maketitle

\thispagestyle{empty}

\begin{abstract}

No abstract yet.

\end{abstract}

\newpage

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population is \cite{DHB}
$$\log\left(1-\Pe\right)+\Ro\Pe=0$$
where $\Ro$ is the basic reproduction number and $\Pe$ is the proportion of the population infected over the course of the outbreak.
If a proportion $x_0$ of the population is susceptible following vaccination, then the  reproduction number under vaccination is $\Rr_V=x_0\Ro$, and the final size equation becomes
$$\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0$$
Hence the relationship between the proportion initially susceptible and the proportion infected in an epidemic is
$$x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}$$
In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks is $x_0-1/\Ro$.

These formulae were applied at a District Health Board (DHB) level, assuming no mixing between DHBs.

\begin{table}[htdp]

\begin{center}

\begin{tabular}{lrrrr}
\hline
DHB	& Size	& Na\"{\i}ve    & 	Attack	& Vacc		\\
\hline
Auckland	&	436350	&	52010	&	31159	&	17920	\\
Bay of Plenty	&	206000	&	20679	&	8437	&	4585	\\
Canterbury	&	482180	&	51357	&	24695	&	13687	\\
Capital and Coast	&	283700	&	32625	&	18403	&	10461	\\
Counties Manukau	&	469300	&	55544	&	32903	&	18880	\\
Hawke's Bay	&	151700	&	15602	&	6846	&	3751	\\
Hutt Valley	&	138380	&	15198	&	7836	&	4388	\\
Lakes	&	98196	&	10558	&	5192	&	2886	\\
MidCentral	&	162560	&	17328	&	8348	&	4628	\\
Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	\\
Northland	&	151690	&	14921	&	5688	&	3071	\\
South Canterbury	&	55620	&	5238	&	1678	&	893	\\
Southern	&	297420	&	31607	&	15115	&	8371	\\
Tairawhiti &	43650	&	4769	&	2431	&	1359	\\
Taranaki	&	109750	&	11473	&	5262	&	2899	\\
Waikato	&	359310	&	39402	&	20248	&	11331	\\
Wairarapa	&	41112	&	3932	&	1346	&	720	\\
Waitemata	&	525550	&	58350	&	30774	&	17291	\\
West Coast	&	32151	&	3197	&	1265	&	685	\\
Whanganui	&	60120	&	6075	&	2530	&	1378	\\
\hline			
TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 \\
\hline
\end{tabular}
\end{center}
\caption{Size: DHB Population, Statistics NZ 2013;		
Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$.			}
\label{Table1}
\end{table}%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\clearpage 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{thebibliography}{11}
\bibitem{DHB} Diekmann, O., Heesterbeek, J.A.P. \& T. Britton (2013) Mathematical tools for understanding infectious disease dynamics. Princeton University Press, Princeton.
\end{thebibliography}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\end{document}

### 2014-10-21 maps/rworldmaps_measles_travel.R
Matched new NZ travel

@@ -26,7 +26,7 @@ library(utils)
 getwd()
 setwd("~/data")
 #unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
-world.map <- readOGR(dsn="C:/Users/David Hayman/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
+# world.map <- readOGR(dsn="C:/Users/David Hayman/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
 
 setwd("C:/Users/David Hayman/Documents/GitHub/measles/maps")
 
@@ -50,8 +50,8 @@ head(monthtravel)
 
 #######
 
-world.ggmap <- fortify(world.map, region = "NAME")
-save(world.ggmap,file="world.ggmap.Rda")
+# world.ggmap <- fortify(world.map, region = "NAME")
+# save(world.ggmap,file="world.ggmap.Rda")
 load("world.ggmap.Rda")
 n <- length(unique(world.ggmap$id))
 
@@ -215,211 +215,133 @@ levels(population$id)[match(c(
     "Venezuela")
 
 ####
+colnames(annualtravel)[1]<-"id"
+levels(annualtravel$id)[match(c(
+  'Samoa, American',
+  'Virgin Islands, British',
+  "Burma (Myanmar)",
+  "China, People's Republic of",
+  "Congo, the Democratic Republic of the",
+  'Falkland Islands',
+  'Faeroe Islands',
+  'French Southern Territories',
+  'Vatican City State',
+  'Hong Kong (Special Administrative Region)',
+  'Iran',
+  'Laos',
+  'Libya',
+  'Macau (Special Administrative Region)',
+  'Gaza Strip/Palestine/West Bank',
+  'Moldova',
+  'St Helena',
+  'St Kitts and Nevis',
+  'St Lucia',
+  'St Maarten (Dutch Part)',
+  'St Pierre and Miquelon',
+  'St Vincent and the Grenadines',
+  'South Georgia and the South Sandwich Islands',
+  'Syria',
+  'Former Yugoslav Republic of Macedonia (FYROM)',
+  'Tanzania',
+  'United States of America',
+  'Virgin Islands, United States',
+  'Wallis and Futuna'),levels(annualtravel$id))] <-
+  c(
+    "American Samoa",
+    "British Virgin Islands",
+    "Burma",
+    "China",
+      "Democratic Republic of the Congo",
[…]

### 2014-10-22 maps/rworldmaps_measles_travel.R
all travellers mapped
Need to make minor revisions - e.g. Greenland missing, but close

@@ -426,8 +426,6 @@ head(newpop2)
 
 mp <- merge(m4,newpop2, by.x = "id",by.y="id",all=T)
 
-dim(mp)
-head(mp)
 dpdf <- data.frame(id = mp$id,
                   incidence = mp$X2012.x/mp$Numeric.Value*1000, ## note 2012 best data
                   cover = mp$X2012.y,## 2012 best data
@@ -479,3 +477,106 @@ topvaccine[1:10,c(1,3)]
 toprisk <- dpdf[order(-dpdf$risk),] 
 toprisk[1:10,c(1,5)]
 
+##
+
+colnames(annualtravel)[2:7]<-c("nzX2008","nzX2009","nzX2010","nzX2011","nzX2012","nzX2013")
+
+mpNZ <- merge(mp,annualtravel, by.x = "id",by.y="id",all=T)
+
+dim(mpNZ)
+
+
+dpdfNZ <- data.frame(id = mpNZ$id,
+                   incidence = mpNZ$X2012.x/mpNZ$Numeric.Value*1000, ## note 2012 best data
+                   cover = mpNZ$X2012.y,## 2012 best data
+                   immigration = mpNZ$nzX2012) 
+
+dpdfNZ$risk <- with(dpdfNZ, mpNZ$X2012.x/mpNZ$Numeric.Value*1000* mpNZ$nzX2012)
+head(dpdfNZ)
+
+head(dpdfNZ)
+plot(dpdfNZ$id,dpdfNZ$immigration)
+
+dpdfNZ <- subset(dpdfNZ,id != "Total")
+dpdfNZ <- subset(dpdfNZ,id != "ASIA")
+
+nznp1<-ggplot(dpdfNZ, aes(map_id = id)) +ggtitle("Travel") +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+nznp2<-ggplot(dpdfNZ, aes(map_id = id)) +ggtitle("Incidence (per million)") +
+  geom_map(aes(fill = incidence), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+nznp3<-ggplot(dpdfNZ, aes(map_id = id)) +ggtitle("Vaccination cover (%)") +
[…]

### 2014-10-22 maps/rworldmaps_measles_travel.R
revised maps and pdfs for NF

@@ -573,10 +573,98 @@ grid.arrange(totnp4, totnp1, totnp2, totnp3, ncol=2, main="Measles (2012)")
 
 ##
 
+pdf(paste("risk_measles_base.pdf"))
 grid.arrange(np2,np3, ncol=1, main="Measles incidence and vaccination cover (2012)")
+dev.off()
 
+pdf(paste("risk_measles_foreign.pdf"))
 grid.arrange(np1, np4, ncol=1, main="Foreign travellers to New Zealand and risk (Travellers * Country incidence)")
+dev.off()
 
+pdf(paste("risk_measles_nzers.pdf"))
 grid.arrange(nznp1, nznp4, ncol=1, main="New Zealand travellers and risk (Travellers * Country incidence)")
+dev.off()
 
+pdf(paste("risk_measles_tot.pdf"))
 grid.arrange(totnp1, totnp4, main="Total travellers and risk (Travellers * Country incidence)")
+dev.off()
+
+
+##
+ttimes<-read.csv("times.csv",header=T)
+lmres<-lm(res~time,data=ttimes)
+summary(lmres)
+plot(ttimes$res,xaxt="n",type="l",xlab="Date",ylab="")
+axis(1, at=1:71, labels=ttimes$mydate) 
+abline(lmres,col="red",
+       lty=2)
+acf(ttimes$res,lag.max = length(ttimes$res),plot=T,main="")
+change<-read.csv("change.csv",header=T)
+
+plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
+     bty="n")
+ylim <- range(c(0,max(change[,4])))
+xlim <- range(change[,1])
+par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
+plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
+     ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
+
+## set wd to get data
+nzers<-read.csv("monthtravel.csv",header=T)
+head(nzers)
+names(nzers)
+dim(nzers)
+
+plot(as.numeric(nzers[247,2:79]),type="l",ylim=c(min(as.numeric(nzers[247,2:79])),max(ttimes$res)),
+)
+points(ttimes$res,add=T,type="l",col="red")
+
+par(mar = c(7, 4, 4, 2) + 0.1)
[…]

### 2014-10-28 draftfinalreport.Rnw
New benefit cost and final report draft
Using Mick's values and new simulations, added new benefit-cost
analyses, incorporating the management costs.

--- 2014-09-01 interimreport2.Rnw
@@
 \usepackage{tabularx}
+\usepackage{amssymb}
+\usepackage{amsmath}
+\usepackage{graphicx}
+\usepackage{caption2}
+\usepackage{fancyhdr}
+\newcommand{\vecA}{\mathbf{A}}
+\newcommand{\vecB}{\mathbf{B}}
+\newcommand{\vecC}{\mathbf{C}}
+\newcommand{\vecD}{\mathbf{D}}
+\newcommand{\vecE}{\mathbf{E}}
+\newcommand{\vecF}{\mathbf{F}}
+\newcommand{\vecH}{\mathbf{H}}
+\newcommand{\vecI}{\mathbf{I}}
+\newcommand{\vecJ}{\mathbf{J}}
+\newcommand{\vecK}{\mathbf{K}}
+\newcommand{\vecM}{\mathbf{M}}
+\newcommand{\vecN}{\mathbf{N}}
+\newcommand{\vecP}{\mathbf{P}}
+\newcommand{\vecQ}{\mathbf{Q}}
+\newcommand{\vecS}{\mathbf{S}}
+\newcommand{\vecT}{\mathbf{T}}
+\newcommand{\vecW}{\mathbf{W}}
+\newcommand{\vecZ}{\mathbf{Z}}
+\newcommand{\vecb}{\mathbf{b}}
+\newcommand{\vecc}{\mathbf{c}}
+\newcommand{\vece}{\mathbf{e}}
+\newcommand{\vecf}{\mathbf{f}}
+\newcommand{\vecg}{\mathbf{g}}
+\newcommand{\vech}{\mathbf{h}}
+\newcommand{\veci}{\mathbf{i}}
+\newcommand{\vecr}{\mathbf{r}}
+\newcommand{\vecu}{\mathbf{u}}
+\newcommand{\vecv}{\mathbf{v}}
+\newcommand{\vecw}{\mathbf{w}}
+\newcommand{\vecx}{\mathbf{x}}
+\newcommand{\vecy}{\mathbf{y}}
+\newcommand{\vecz}{\mathbf{z}}
+\newcommand{\Ach}{\mathcal{H}}
+\newcommand{\Ell}{\mathcal{L}}
+\newcommand{\En}{\mathcal{N}}
+\newcommand{\Pe}{\mathcal{P}}
+\newcommand{\Ro}{\mathcal{R}_0}
+\newcommand{\Rc}{\mathcal{R}_c}
+\newcommand{\Rr}{\mathcal{R}}
+\newcommand{\Qu}{\mathcal{Q}}
+\newcommand{\Te}{\mathcal{T}}
+\newcommand{\Ve}{\mathcal{V}}
+\newcommand{\Dub}{\mathcal{W}}
+\newcommand{\Zed}{\mathcal{Z}}
+\newcommand{\Exp}{\mathbb{E}}
+\newcommand{\de}{\mbox{d}}
+\newcommand{\wrt}{\,\mbox{d}}
+\newcommand{\veczero}{\mathbf{0}}
+\newcommand{\vecone}{\mathbf{1}}
+\newcommand{\vecalpha}{\mbox{\boldmath$\alpha$}}
+\newcommand{\vectheta}{\mbox{\boldmath$\theta$}}
+\newcommand{\vecphi}{\mbox{\boldmath$\phi$}}
+\newcommand{\vecPhi}{\mbox{\boldmath$\Phi$}}
+\newcommand{\vecSigma}{\mbox{\boldmath$\Sigma$}}
+\newcommand{\vecimath}{\mbox{\boldmath$\imath$}}
+\newcommand{\cosech}{\mbox{cosech}\,}
+\newcommand{\sign}{\mbox{sign}\,}
+\newcommand{\trace}{\mbox{trace}\,}
+\newcommand{\Inf}{\mbox{Inf}\,}
 
@@
 
-\section{Abstract}
+\section{Introduction}
 
-New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we present the results of statistical analyses of risk factors for measles cases since 2007 in New Zealand during outbreaks, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks pre- and post different vaccination scenarios, based on alternative situations.
-We provide preliminary benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
-Our key findings were:
-\begin{itemize}
-\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be cases in outbreaks since 2007.
-\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
-\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if measles in New Zealand continues to infect the same proportion of secondary cases as in the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreaks remained the size of the average outbreaks since 1997, there is no benefit to additional vaccination campaigns.
-\end{itemize}
+The well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population is \citep{diekmann13}
+$$\log\left(1-\Pe\right)+\Ro\Pe=0$$
+where $\Ro$ is the basic reproduction number and $\Pe$ is the proportion of the population infected over the course of the outbreak.
+If a proportion $x_0$ of the population is susceptible following vaccination, then the  reproduction number under vaccination is $\Rr_V=x_0\Ro$, and the final size equation becomes
+$$\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0$$
+Hence the relationship between the proportion initially susceptible and the proportion infected in an epidemic is
+$$x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}$$
+In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks is $x_0-1/\Ro$.
 
-\section{Background}
+These formulae were applied at a District Health Board (DHB) level, assuming no mixing between DHBs.
 
-As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission.
-A brief review of the history of measles in New Zealand was provided in our previous interim report.
-In this report were present regression analyses to determine which populations are most at risk, the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the benefit--cost analyses for vaccination dependent on differing scenarios.
+\begin{table}[htdp]
 
-\section{Risk analysis update}
-\label{sub:risk_analyses}
+\begin{center}
 
-A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups.
-Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, we found the report to be very thorough, however, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
-The additional analyses included in this section are multivariate regression modelling to account for confounding within the univariate analyses.
-
-\subsection{Risk analyses methods}
-We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
-These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
-We used these denominator data to determine if there were interactions among specific age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep) that might exist among cases allowing better understanding of risk of measles infection.
-
-The University of Otago denominator data provided were not to the same detail as the ESR case data.
-Notably, the denominator age data were categorised into several classes: 0--5, 6--17, 18--24, 25--64, and 65+ year categories.
-The ethnicity denominator data were not Prioritized Ethnicity at the Level 1 Ethnic Group Codes, but at the Level 2 Ethnic Group Codes, though with some alternative codes provided that did not match the Level 2 Ethnic codes.
-After discussions with the University Otago we have provided results based on the best available data, though for smaller ethnic group categories, some results may be unreliable and these are discussed below.
-
-With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, this led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## get data
-# set wd
-# setwd("~/Massey 2014/DHayman_20140627")
-# read data
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-                 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-@
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mfrow=c(1,1))
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="",cex.lab=2)
-hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1.8)
-@
+\begin{tabular}{lrrrr}
+\hline
+DHB  & Size	& Na\"{\i}ve    & 	Attack	& Vacc		\\
+\hline
+Auckland	&	436350	&	52010	&	31159	&	17920	\\
+Bay of Plenty	&	206000	&	20679	&	8437	&	4585	\\
+Canterbury	&	482180	&	51357	&	24695	&	13687	\\
+Capital and Coast	&	283700	&	32625	&	18403	&	10461	\\
+Counties Manukau	&	469300	&	55544	&	32903	&	18880	\\
+Hawke's Bay	&	151700	&	15602	&	6846	&	3751	\\
+Hutt Valley	&	138380	&	15198	&	7836	&	4388	\\
+Lakes	&	98196	&	10558	&	5192	&	2886	\\
+MidCentral	&	162560	&	17328	&	8348	&	4628	\\
+Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	\\
+Northland	&	151690	&	14921	&	5688	&	3071	\\
+South Canterbury	&	55620	&	5238	&	1678	&	893	\\
+Southern	&	297420	&	31607	&	15115	&	8371	\\
+Tairawhiti &	43650	&	4769	&	2431	&	1359	\\
+Taranaki	&	109750	&	11473	&	5262	&	2899	\\
+Waikato	&	359310	&	39402	&	20248	&	11331	\\
+Wairarapa	&	41112	&	3932	&	1346	&	720	\\
+Waitemata	&	525550	&	58350	&	30774	&	17291	\\
+West Coast	&	32151	&	3197	&	1265	&	685	\\
+Whanganui	&	60120	&	6075	&	2530	&	1378	\\
+\hline			
+TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 \\
+\hline
+\end{tabular}
 \end{center}
-\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
-\label{fig:ageinyears}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## plot cases / age class
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-                   data = test , FUN=sum)
-#caseyr
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-colnames(naive)<-c("AgeInYears","Naive","Cases")
-
-dose1=as.factor(round(vac$Dose1Mths/12,0));
-dose1<-summary(dose1)
-dose1<-as.data.frame(dose1)
-dose1$AgeInYears<-rownames(dose1)
-dose2=as.factor(round(na.omit(vac$Dose2Mths/12,0)));
-dose2<-summary(dose2)
-dose2<-as.data.frame(dose2)
-dose2$AgeInYears<-rownames(dose2)
-
-## need to merge vaccination data with data
-## to get years the cases were from
-datav<-merge(vac,data,by="CaseCode",all=T)
-##
-datav$Dose1Mths[which(is.na(datav$Dose1Mths))] <- 9999
-datav$Dose2Mths[which(is.na(datav$Dose2Mths))] <- 9999
-#########################
-datav$RptYear<-as.factor(datav$RptYear)
-datav$SurvWeek<-as.factor(datav$SurvWeek)
-datav$NZDep01<-as.factor(datav$NZDep01)
-datav$NZDep06<-as.factor(datav$NZDep06)
-datav$NZDep13<-as.factor(datav$NZDep13)
-datav$Dose1Mths<-as.factor(datav$Dose1Mths)
-datav$Dose2Mths<-as.factor(datav$Dose2Mths)
-
-#data$AgeInYears<-as.factor(data$AgeInYears)
-datav$EthnicityPrioritised<-as.factor(datav$EthnicityPrioritised)
-
-timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
-                 data = datav , FUN=sum,na.rm=F,na.action=na.pass)
-
-testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-library(plyr)
-testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
-testv$Dose2Mths<-revalue(testv$Dose2Mths, c("9999"=NA));
-testv$Dose1Mths<-as.numeric(testv$Dose1Mths)
-testv$Dose2Mths<-as.numeric(testv$Dose2Mths)
-testv$D2vac<-ifelse(testv$Dose1Mths > -1 & testv$Dose2Mths >= testv$Dose1Mths,testv$Dose2Mths,NA)
-testv$D1vac<-ifelse(testv$Dose1Mths > -1 & is.na(testv$D2vac) == T,testv$Dose1Mths,NA)
-testv$Unvac<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,testv$AgeInYears,NA)
-
-testv$VC<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,0,
-                 ifelse(testv$Dose1Mths >= 0 & is.na(testv$Dose2Mths) == T,1,2))
-AgeVac<-table(testv$VC,testv$AgeInYears)
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-AgeInYears<-(as.numeric(rownames(AgeVac)))
-AgeVac<-cbind(AgeVac,AgeInYears)       
-AgeInYears<-(as.numeric(rownames(pop)))
-pop<-cbind(pop,AgeInYears)
-colnames(pop)<-c("Population","AgeInYears")
-AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
-# head(AgeV)
-#AgeVac<-t(AgeVac)
-@
-
-This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from 1 to 10 to two: NZDep 1--5 and NZDep 6--10.
-We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
-By doing so, we reduce the zeroinflation present in the data.
-
-The prioritized ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories, though for the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see "None", Table~\ref{table:case}).
-<<echo=FALSE,include=FALSE,results=hide>>=
-## regression analyses - measles
-# setwd("~/Massey 2014/DHayman_20140627")
-# read data
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-                 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
-
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-                   data = test , FUN=sum)
-caseyr
-test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
-test$AgeInYears<-as.factor(test$AgeInYears)
-tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
-test$AgeInYears<-tage
-#summary(test$EthnicityPrioritised)
-teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
-                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
-
-test$EthnicityPrioritised<-teth
-head(test)
-test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
-test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
-
-# want to set NZ Deprivation to the appropriate year
-
-testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
-                              ifelse (test$RptYear == "2008",test$NZDep06,
-                              ifelse (test$RptYear == "2009",test$NZDep06,
-                              ifelse (test$RptYear == "2010",test$NZDep06,
-                              ifelse (test$RptYear == "2011",test$NZDep06,
-                              ifelse (test$RptYear == "2012",test$NZDep06,
-                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-head(testt)
-tail(testt)
-str(testt)
-testt$NZDep<-as.factor(testt$NZDep-1)
-
-testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
-                      ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
-                 data = testt , FUN=sum)
-dim(testtable)
-colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
-head(testtable)
-
-tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-                                 "9"="6-10","10"="6-10"));
-
-testtable$NZDep<-tnzd
-# head(testtable)
-tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
-                 data = testtable , FUN=sum)
-
-ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
-                 data = testtable , FUN=sum)
-@
-
-\begin{table}\tiny
-\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(ttyr, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:case}
-\end{table}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-### denominator data
-# setwd("~/Massey_2014/measles/data")
-denom<-read.csv("NZDep2006Denominators.csv",header=T)
-head(denom)
-denom<-denom[,-c(9)]
-summary(denom)
-dim(denom)
-denom<-denom[,1:18]
-row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
-sum(row.has.na)
-denomt <- denom[!row.has.na,]
-head(denomt)
-denomt<-denomt[denomt$Sex_code==99,]
-denomt$Eth_Level<-as.factor(denomt$Eth_Level)
-denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
-dm<-cbind(denomt[,c(3,6,9:18)])
-head(dm)
-dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
-dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
-dm$Age_Label<-dage
-colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-head(dm)
-datadm<-dm[dm$Age=="<6",]
-dataA<-(rbind(datadm,datadm)/2)
-dataA<-round(dataA[,3:12])
-dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
-colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-ddm<-dm[!dm$Age=="<6",]
-dm<-rbind(dataA,ddm)
-dm$Ethnicity <- factor(dm$Ethnicity)
-deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
-                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
-dm$Ethnicity<-deth
-head(dm)
-
-popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
-colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
-popn
-pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
-                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
-                                 "Dep9"="6-10","Dep10"="6-10"));
-
-popn$NZDep<-pnzd
-head(popn)
-tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
-               data = popn , FUN=sum)
-
-tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
-ttyr$merge <- paste(ttyr$NZDep, ttyr$Age, ttyr$Ethnicity)
-ttyr <- ttyr[ttyr$Ethnicity!="None",]
-
-tp$cases <- 0
-cases <- matrix(0, length(tp$merge),1)
-rownames(cases) <- tp$merge
-cases[ttyr$merge,] <- ttyr$Cases
-tp$cases <- cases
-PerCapita<-(tp$cases/tp$Popn)
-PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita,4))
-colnames(PerCap)[4]<-"Population"
-colnames(PerCap)[5]<-"Cases"
-colnames(PerCap)[6]<-"Per Capita"
-@
-
-The numbers of cases per category for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="Category",cex.lab=2)
-legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
-       col=c(1:4,6),pch=16,bty="n",cex=1.8)
-@
-\end{center}
-\caption{Per capita rates of measles infections broken down by ethicity, see Table~\ref{table:percap} for details}
-\label{fig:percap}
-\end{figure}
-
-\begin{table}\tiny
-\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:percap}
-\end{table}
-
-However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
-Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table\ref{table:percap}).
-
-For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model.
-We also account for differences in population sizes by using an offset term, the log(population size).
-We used a model simplification approach, by beginning our analyses with all terms and all interactions, and simplifying the models through removal of non-significant higher order interaction terms.
-Thus, the final model that remained with all significant interaction terms had the following linear predictor:
-
-\begin{equation} \label{eq:reg}
- log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a_e(x_a * x_e)+ log(population)  + \epsilon
-  \end{equation}
-Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## plot this figure?
-#hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-tp$Ethnicity<- relevel(tp$Ethnicity, "European")
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
-
-model2<-update(model1,~.-Age:NZDep)
-summary(model2)
-anova(model2,test="F")
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-tpsub<-tp[!(tp$Ethnicity=="MLA"),]
-
-#hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
-
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F")
-
-model2<-update(model1,~.-Ethnicity:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-model3<-update(model2,~.-Age:NZDep)
-summary(model3)
-anova(model3,test="F")
-@
-
-\subsection{Regression analyses results}
-\label{sub:regression_results}
-
-The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
-@
-\end{center}
-\caption{Histogram of cases per category used in the final regression model}
-\label{fig:histcasecat}
-\end{figure}
-
-The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width =11, height=11>>=
-res<-predict(model3)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-@
-\end{center}
-\caption{Regression model predictions plotted against the cases (Equation~\ref{eq:reg})}
-\label{fig:predict}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-@
-\end{center}
-\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
-\label{fig:resid}
-\end{figure}
-The significance of the different predictor variables can be seen in the ANOVA results:
-
-<<label=regmodelanova,include=FALSE>>=
-anova(model3,test="F")
-@
-
-A summary of the model with the individual effects and the statistical support for the estimated coefficients can be seen here:
-
-<<label=regmodel,include=FALSE>>=
-summary(model3)
-@
-
-The results of the model suggest that, apart from over-representation of some MLA categories discussed above and not included here, age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
-
-People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-
-In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
-Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
-
-\subsection{Risk analysis discussion}
-
-The regression analyses suggest that age is a particularly strong risk factor for measles.
-This comes as no surpise to epidemiologists or health care providers.
-However, our analyses also highlight other groups that are at risk.
-In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
-Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses.
-Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
-A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.
-
-The results of these regression analyses are also discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit})
-
-\subsection{Risk analysis summary}
-\begin{itemize}
-\item Risk of measles infection decreases significantly with age
-\item Pacific people are statistically more at risk on a per capita basis
-\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
-\item There is some statistical support for Pacific and Asian children in the 6--17 year age caterories being at lower risk than European or Maori children
-\end{itemize}
-
-\section{Modelling measles epidemics}
-\label{sec:epidemic_modelling}
-
-In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the benefit--cost analysis (Section~\ref{sub:cost_benefit}).
-
-A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
-The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles.
-Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, as etimated using the NZ census data \citep{stats14} and the proportion of each age class immune, based on data provided to us by the Ministry of Health on commencing this project.
-The figure shown (Figure~\ref{fig:naive}) and for all subsequent analyses assume a currently susceptible population in New Zealand of approximately 11\% of the total population.
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-#length(pop)
-naive<-round(pop-(pop*impop))
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
-\end{center}
-\caption{New Zealand population by age and estimates numbers of naive people in each age class}
-\label{fig:naive}
-\end{figure}
-
-The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
-This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}.
-If $R_0 > 1$ an epidemic will occur following an introduction of infection.
-The best estimate for measles in New Zealand was $R_0$ = 12.8 \citep{roberts4}.
-The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
-If $R_v < 1$ epidemics are predicted to be prevented (but see below) and infection persistence is prevented.
-The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
-
-\subsection{Modelling methods}
-
-To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we previously estimated $R_v$ from all the outbreaks in New Zealand since 2009.
-To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
-We were required to compute the generation time for measles to do so.
-The generation time is the average time an index case infects others after becoming infected.
-We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
-The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{cases_by_dhb_2009_2014.pdf}
-     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
-     \label{fig:dhbcases}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
-     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
-     \label{fig:outbreaks}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks}}
-     \label{fig:r0}
-\end{figure}
-
-To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we make several simplifying assumptions regarding the relationship between the proportion requiring vaccination, $p_c$, and $R_0$.
-Specifically, we use the simple relationship:
-
-\begin{equation} \label{eq:pc}
-p_c = 1-(1/R_0)
-  \end{equation}
-
-We then assume that the measles epidemics are occuring in a naive population that is the size of the naive population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
-We estimated the susceptible population in NewZealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
-Thus, the $p_c$ comes from the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
-
-We then used the proportions of the population that were estimated to require vaccination (Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
-These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).
-
-finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of suscpetible people in New Zealand was now reduced by the numbers vaccinated.
-Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).
-
-\subsection{Modelling results and discussion}
-
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report.
-The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one.
-As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
-However, the results of these differences are presented and discussed in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).
-
-These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we explore the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
-
-The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
-However, sub-clinical and underreporting may lower the estimate.
-The relative contributions of both to our estimates are currently unknown.
-
-To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:pc}.
-Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, Figure~\ref{fig:r0}) to an upper limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
-These figures can be reached in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
-
-The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
-However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
-Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
-Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
-
-\subsection{Summary of modelling}
-\begin{itemize}
-\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number, $R_v$, for measles in the current 2013--2014 outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
- \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
-\end{itemize}
-
-\section{Cost analyses}
-
-In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current measles outbreak.
-For completeness, we include much of the introduction from the previous interim report, but we have revised some of the figures.
-
-Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \citep{clements4}.
-Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities.
-\citep{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
-Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
-They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination.
-The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \citep{who13}.
-In addition to the substantial losses occurring in measles-endemic countries, a significant impact is felt in heavily measles-vaccinated countries, which may be considered measles-free, due to contact with cases either in the country of origin or in the previously measles-free country.
-
-The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3}.
-The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}.
-\citep{stack11} estimated the economic benefits from cases averted due to measles vaccination.
-They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion of costs averted between 2011 and 2020.
-Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining <1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
-
-Italy has the highest reported annual cost of measles among industrialised countries \citep{carabin3}.
-In 2001, it reported losses related to measles of approximately US\$50 million.
-The economic impact of a large measles outbreak in Italy, 2002--03 examined the costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis.
-The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
-The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 $\approx$ NZ\$2.0 in 2002-03), or approximately \euro 1,700 per case.
-The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
-The average daily cost of a hospital stay was  \euro 327.
-
-An outbreak of measles occurred in Sydney, Australia, lasting nearly 2 months in 2011 and resulted in 26 confirmed cases \citep{flego13}.
-Seven (27\%) of the cases required hospitalisation for more than 1 day and 10 (38\%) resulted in management within a hospital emergency department.
-During this outbreak, a total of 1,395 contacts were identified and managed by a public health unit in western Sydney.
-The mean number of contacts per case was 54 (median = 28, maximum = 206).
-The estimated cost to the public health unit for contact management for the epidemic was in excess of AUS\$48,000, with 90\% of this being associated with staff time.
-
-Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years.
-In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \citep{wichmann9}.
-However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children \citep{siedler6}.
-An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
-In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case.
-Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised.
-The mean costs of the hospitalised patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623.
-In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing.
-Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case.
-In addition, to determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
-
-Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world \citep{parker6}.
-Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.
-\citep{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
-The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$2.7 and 5.3 million.
-Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$298 per contact.
-It was estimated that for the one week that the Iowa Department of Public Health (DPH) investigated a case in 2004, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination \citep{dayan5}.
-In total, it was estimated the direct costs associated with three cases of measles was US\$142,452, or nearly US\$50,000 per case.
-
-The impact of a measles outbreak due to a non-autochthonous case in Indiana was also reported \citep{parker6}, and a total of 34 cases, 94\% of which were not vaccinated against measles, were reported in the outbreak.
-Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak.
-Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overhead.
-This amounted to a direct cost of \$4,932 per measles case.
-These costs did not include either patient care or indirect costs, which would have made the total and per case cost higher.
-
-The direct medical and public health costs in response to a single case of refugee-imported measles has been reported \citep{coleman12}.
-Costs included labor, translation and benefits for public health workers.
-In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis.
-In total, 387 hours were associated with this single case, resulting in a cost of US\$11,881.
-In addition, per-contact costs amounted to US\$264.
-The cost of hospitalisation for the 3-day stay by the index case was US\$931.
-Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents' exposed to measles (US\$240) for a total of US\$23,816.
-
-Economic analyses of measles control programs have shown them to be financially effective.
-In the Republic of Korea, the economics of alternative measles vaccination programs were compared.
-All of the alternatives were found to be economically efficient (benefit/cost ratio (B/C) > 1.0), with the alternative using two doses of the MMR program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27).
-
-The purpose of the current study is to estimate the cost of the current measles outbreak in New Zealand.
-Using this information, we will then evaluate the economics of alternative measles control strategies in order to provide additional information to public health officials and decision makers.
-
-\subsection{Cost analyses methods}
-Costs were evaluated as either direct or indirect.
-Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
-Direct costs can be divided into medical and non-medical \citep{saha13}.
-Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
-Personnel time (investigation and emergency response), materials (phone calls, vaccine), personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
-Direct non-medical costs include transportation to and from health care providers.
-
-Indirect costs are productivity losses for the case and/or health care provider, e.g.
-parent of a school child.
-Indirect costs included work loss for cases and caregivers.
-This could also include the economic value of premature life lost, costs associated with permanent disability, e.g.
-deafness and mental retardation.
-Commonly the human value approach (HVA) has been used to estimate economic impact of life.
-The HVA measures the potential future earnings of an individual and discounts it into a present value.
-Typically this is 3\% but 5\% has also been used in a sensitivity analysis, which is more compared to non-human life calculations and will tend to reduce the present value of the future earnings (saved by avoiding a case).
-
-Data for the current measles outbreak were obtained from the New Zealand Ministry of Health, from 2008 through June 2014.
-Data included information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, case weight and associated cost.
-
-Cost of the Auckland Regional Public Health Service (ARPHS) for measles response were obtained from the Ministry of Health.
-Data, for the period January 1 - March 9, 2014, reported salaries for people involved with the measles outbreak management medical team.
-The costs were reported as direct, additional (above normal budgeting) costs required to enable the management of measles.
-It includes a breakdown by individual performing the work and whether it was during the normal work schedule (Monday to Friday, M-F) or weekends.
-Normal work was calculated as $1.2 \times \textsf{full time equivalent (FTE)} \times \textsf{number of days worked}$.
-Overtime was calculated as $1.6\times\textsf{FTE}$ (M-F) and $2.0 \times \textsf{FTE}$ (weekend).
-A full day was considered as 8 hours worked.
-Salary (hourly) rates were calculated for the following: public health nurse (PHN, \$36), public health assistant (PHA, \$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), incident management team (IMT), which had the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and safety/security officer (SSO) (\$26).
-In addition, measles operations personnel were calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
-
-Mean wages for New Zealand workers, by age and gender were obtained for the period, 2008--2013 from the New Zealand Income Survey (Statistics New Zealand, 2013).
-Measles cases were assumed to not work for a period of 5 days.
-Similarly, a care taker was assumed to not work for 5 days if the case were less than 20 years of age.
-In order to calculate the wage loss associated with the care taker, it was assumed that the person was a female between the ages of 35-39.
-Age and gender information for the 192 publicly funded hospital discharges with a measles primary diagnosis from 2008--2013 were matched to the New Zealand wage file to calculate lost wages due to measles.
-
-A regression analysis was performed to test for significant associations between hospital cost and the following explanatory variables: case age at discharge, gender, length of stay (days) and year of case.
-
-\subsection{Cost analyses results}
-\label{sub:cost_benefit}
-Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
-The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\ref{table:hosp}.
-
-<<echo=FALSE>>=
-#setwd("C:\Users\dtshayma\Documents\GitHub\measles")
-costtable1<-read.csv("costtable1.csv",header=T)
-@
-
-\begin{table}
-\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable1, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:direct}
-\end{table}
-
-The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (Table~\ref{table:hosp}).
-The mean cost per case was \$1,877.
-The mean cost per day of stay in the hospital was \$1,170.
-
-<<echo=FALSE>>=
-costtable2<-read.csv("costtable2v2.csv",header=T)
-@
-
-\begin{table}
-\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable2, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:hosp}
- \centering
- \begin{tablenotes}
-      \small
-      \item As of 11 July, 2014. NC - not calculated.
-    \end{tablenotes}
-\end{table}
-
-From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before 1 January 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
-The number of cases by age group is shown in Table~\ref{table:freq}.
-Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.
-
-<<echo=FALSE>>=
-costtable3<-read.csv("costtable3.csv",header=T)
-@
-
-\begin{table}
-\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable3, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:freq}
-\end{table}
-
-The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (Figure~\ref{fig:hosp}).
-
-<<echo=FALSE>>=
-hosp<-read.csv("hospital.csv",header=T)
-library(ggplot2)
-p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")+theme(axis.title.x = element_text(size=25),axis.title.y = element_text(size=25),axis.text.x  = element_text(size=16),axis.text.y  = element_text(size=16))
-@
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(p)
-@
-\end{center}
-\caption{Number of cases attending hospital and stay duration}
-\label{fig:hosp}
-\end{figure}
-
-Nearly 40\% (114/293) of the cases did not spend a night in the hospital, while approximately one-quarter (69/293) spent 1 night and more than three-quarters (222/293) spent less than three nights in the hospital.
-Only eight cases spent a week or more in the hospital.
-Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days.
-The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
-
-<<echo=FALSE>>=
-costtable4<-read.csv("costtable4v2.csv",header=T)
-@
-
-\begin{table}
-\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable4, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cases}
- \centering
- \begin{tablenotes}
-      \small
-      \item As of 7 August, 2014.
-    \end{tablenotes}
-\end{table}
-
-Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
-Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.
-The cost was \$256 less if the case were a male.
-It increased of approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014.
-The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.
-
-<<echo=FALSE>>=
-costtable5<-read.csv("costtable5v2.csv",header=T)
-@
-
-\begin{table}
-\caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable5, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:regression}
-\end{table}
-
-Wages lost due to measles were calculated for the period January 2008 -- August 2014.
-Calculations were based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
-If the case were less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
-Total wage lost for the 247 cases and care givers was estimated to be \$210,436.
-This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008--2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
-
-\subsection{Benefit--cost analyses methods}
-To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}), assuming recovery from infection led to immunity and thus constantly reducing the population size by the number recovered.
-We used these values of numbers of predicted cases and the cost figures above to estimate what the cost of not vaccinating additional populations would be.
-We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this.
-We can therefore estimate the number of cases prevented, and the savings made from the additional vaccinations.
-We then use the costs of the catch up vaccination schemes, the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity, and the savings from reduced measles cases to work out the benefit--cost ratio ((B/C), where a B/C ratio > 1 means that the program benefits exceed their costs.
-A B/C value less than one suggests the costs are higher than the economic benefits.
-Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}:
-
-\begin{equation} \label{eq:disc}
-\textit{present value} = \sum \textit{future value}_y_r * (1/(1+\textit{discount rate})^{yr}
-  \end{equation}
-Where $_y_r$ is the year 0 (current) through to 9 (ten years into the future).
-
-Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the susceptible New Zealand population would become infected during the outbreak, or the epidemic it would stutter after a few infects and fail to cause a large epidemic due to stochasticity.
-We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
-However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
-However, to address this, we not only use $R_v$ values of 1, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year, since 1997.
-This was 220 cases per year.
-
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population and using the upper limits of our $R_v$ estimates (Section~\ref{sec:epidemic_modelling}) are shown in Figure~\ref{fig:numvacc}.
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-#naive[is.na(naive)] <- 0
-colnames(naive)<-c("AgeInYears","Naive","Cases")
-@
-
-\begin{figure}
-     \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-##
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(round(0.53*naive$Naive) ,col="grey",bg="red",pch=21,ylab="Number to vaccinate",xlab="Age in Years",
-     main="",cex.lab=2)
-points(naive$Naive[0:21],col="grey",bg="orange",pch=22) # 
-points(naive$Naive[3:30],col="grey",bg="yellow",pch=23) 
-points(jitter(round(0.16*naive$Naive)),col="grey",bg="lightblue",pch=21) 
-points(jitter(naive$Naive[5:13]),col="grey",pch=22,bg="skyblue") # 
-points(naive$Naive[1:5],col="grey",pch=23,bg="navyblue") #
-legend("topright",c("53% naive, all ages","53% naive, all 0-21 yr olds","53% naive, all 3-30 yr olds",
-                    "16% naive, all ages","16% naive, all 5-13 yr olds","16% naive, all 1-5 yr olds"),
-       bty="n",pch=c(rep(21:23,2)),col=rep("grey",6),pt.bg=c("red","orange","yellow","lightblue","skyblue","navyblue"),cex=2)
-@
-     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling})}
-     \label{fig:numvac}
-\end{figure}
-
-The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
-These simulations show that even in scenarios when $R_v$ is one, and thus stochastically should fail to persist for long, large outbreaks can occur due to stochastic processes.
-Though the median value from 1000 simulations is low (2 cases), and thus most measles introductions will be single cases or lead to minor outbreaks, in this modelling exercise the mean value was 151 cases and the maximum nearly 25,000 cases.
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## simulate measles outbreaks/epidemics using GT and R0 from data.
-# http://www.sciencedirect.com/science/article/pii/S0022519311003146
-mu    <- 12
-sigma <- 3.5
-sigma_logn <- sqrt(log(1 + (sigma/mu)^2))
-mu_logn    <- log(mu) - log(1 + (sigma/mu)^2) / 2
-
-# then exp(rnorm(n, mu_logn, sigma_logn)) simulates from lognormal with the given mean and sd.
-
-# the R0 library 
-require(R0)
-#source("estR0.R")
-
-# generation time 
-genTime <- generation.time(type="lognormal", val=c(12, 3.5))
-
-jm.epid <- function (epid.nb, GT, R0, epid.length, family, negbin.size = NULL, 
-          peak.value = 300000, popn = 300000) 
-{
-  if (class(GT) != "R0.GT") {
-    stop("GT object must be of class R0.GT.")
-  }
-  if (family == "negbin" & is.null(negbin.size)) {
-    negbin.size <- R0/4
-  }
-  GT <- GT$GT
-  epidemics <- matrix(data = 0, nrow = epid.length, ncol = epid.nb)
-  for (n in 1:epid.nb) {
-    sim.epid = c(1, rep(0, epid.length - 1))
-    for (t in 1:epid.length) {
-      if (family == "poisson") {
-        susc <- max(0,popn - sum(sim.epid, na.rm=T))
-    #    peak.value[t]<-susc
-        new <- rbinom(sim.epid[t], susc, min(1, R0/susc))
-        if (sum(new) > susc) # arggrggg
-        {
-    #      cat("doing the new thing", susc, sim.epid[t], sum(sim.epid[1:t]),"\n")
-          new <- rep(0, sim.epid[t])
-          for (i in 1:sim.epid[t]) {
-   #         cat("taking a random number, susc=", susc, "\n")
-            new[i] <- rbinom(1, susc, min(1, R0/susc))
-            susc <- susc - new[i]
-            if (susc == 0)
-            {
- #             cat("got 'em all", sum(new), "\n")
-              break;
-            }
-          }
-        }
-#        cat("generating", new[1], "individuals from first, susc=", susc, "\n")
-      }
-      else if (family == "negbin") {
-        new <- rnbinom(sim.epid[t], size = negbin.size, 
-                       mu = R0)
-      }
-if (is.na(sum(new)))
-  break
-
-      newd <- rmultinom(1, sum(new), GT)[, 1]
-      sim.epid[t:(t + length(GT) - 1)] <- sim.epid[t:(t + 
-                                                        length(GT) - 1)] + newd
-      if (sim.epid[t + 1] > peak.value & t < (epid.length - 
-                                                1)) {
-        sim.epid[(t + 2):epid.length] <- 0
-        break
-      }
-    }
-    sim.epid <- sim.epid[!is.na(sim.epid)]
-    epidemics[, n] <- sim.epid
-  }
-  return(epidemics)
-}
-
-## if R 0 == 2
-set.seed(1) # all infected
-res<-jm.epid(epid.nb=1000,GT=genTime,R0=1,epid.length=365*5,popn=4.2*1e6 * 0.11
-             ,family="poisson",peak.value=4.2*1e6 * 0.11)
-sizes<-colSums(res)
-sizes
-legMd<-as.factor(paste("median =",c(median(sizes))))
-legMn<-as.factor(paste("mean =",c(round(mean(sizes)))))
-@
-
-\begin{figure}
-     \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(sizes,xlim=c(0,200),col="grey",breaks=3000,xlab="Outbreak size",main="",cex.lab=2)
-#summary(sizes)
-abline(v=median(sizes),col="orange",lty=3,lwd=10)
-abline(v=mean(sizes),col="red",lty=3,lwd=10)
-#length(which(sizes>5))/1000
-legend('top',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
-@
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
-The full distribution of results can be seen in Figure~\ref{fig:sim1}}
-     \label{fig:sim}
-\end{figure}
-
-\begin{figure}
-     \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(sizes,breaks=1000,xlab="Outbreak size",main="",cex.lab=2)
-abline(v=median(sizes),col="orange",lty=3,lwd=10)
-abline(v=mean(sizes),col="red",lty=3,lwd=10)
-legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
-@
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
-A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
-     \label{fig:sim1}
-\end{figure}
-
-For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
-We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
-We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
-
-\subsection{Benefit--cost analyses results}
-
-\begin{sidewaystable}[htbp]  
-  \centering  
-  \begin{raggedleft}
-  \caption{Cost benefit analyses using simulated epidemic sizes}  
-  \label{tab:cost_sim}
-   \begin{threeparttable}[b] \tiny
-    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
-    \toprule
-
-Years $R_v$ estimated from & $R_v$ range &	$R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
-\midrule
-2009-2014 &	0.92-1.19	& 0.92	& 0	    & na	  & 0	      & 130	    & 22  	& 148551	  & 130519	  & 13	& 130	  & 22	& 148551	& 0	      & 0	        & 0	        & na \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	& 20	  & 1600000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 126359360	& 79.97 \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	& 50	  & 4000000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 123959360	& 31.99 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 336060587	& 81.72 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 329815997	& 32.69 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 377382119	& 75.71 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 369805289	& 30.28 \\
-
-\end{raggedleft}
-\bottomrule  
-    \end{tabular}%  
-    \begin{tablenotes}\footnotesize  
-        \item $\dagger$ {1000 simulations}
-        \item \sym{*} Proportion of cases hospitalised 0.17
-        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
-        \item \sym{***} Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
-    \end{threeparttable}  
-\end{sidewaystable}
-
-\normalsize
-
-\begin{sidewaystable}[htbp]  
-  \centering  
-  \begin{raggedright}
-  \caption{Cost benefit analyses using historic epidemic sizes since 1997}  
-  \label{tab:cost_ave}
-   \begin{threeparttable}[b] \tiny
-    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
-    \toprule  
-Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following addtional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
-\midrule
-
-2009-2014	& 0.92-1.19	& 0.92	& 0	      & na	& 0	      & 130	  & 22	& 148551	  & 130519	& 13	& 130	  & 22	& 148551	& 0	  & 0	      & 0	        & na \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 20	& 1600000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -847009	  & 0.47 \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 50	& 4000000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -3247009	& 0.19 \\ 
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -3118912	& 0.25 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -9363502	& 0.10 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -4007072	& 0.21 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -11583902	& 0.08 \\
-\end{raggedright}
-\bottomrule  
-    \end{tabular}%  
-    \begin{tablenotes}\footnotesize  
-        \item $\dagger$ {1000 simulations}
-        \item \sym{*} Proportion of cases hospitalised 0.17
-        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
-        \item \sym{***} Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
-    \end{threeparttable}  
-\end{sidewaystable}
-
-\normalsize
-
-The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}.
-The results in the two tables show two alternative trends.
-In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
-However, if the previous recent history of cases since 1997 is the 'status quo' and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one (Figure~\ref{fig:r0}) is not a cost effective exercise (Table~\ref{tab:cost_ave}).
-
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-AgeVac<-table(testv$VC,testv$AgeInYears)
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-@
-\end{center}
-\caption{Age and vaccination status of cases}
-\label{fig:ageandvac}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-@
-\end{center}
-\caption{Age of vaccination of vaccinated measles cases}
-\label{fig:vaccstat}
-\end{figure}
-
-\subsection{Benefit--cost analysis discussion}
-The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
-The majority of the cost analyses were discussed in our previous interim report and though some details have changed, we will focus here on the benefit--cost analyses.
-However, for all analyses more detailed costs of all aspects of hopsitalisation would be beneficial.
-So too would information of the cost of measles vaccination in New Zealand.
-
-We have used standard discounting from healthcare of 3\% per annum \citep{honeycutt6}, but again there may be official New Zealand discounting rates that we are unaware of.
-For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014.
-Including these could be done, but we would like some further data of these before including them in our benefit--cost analsyes.
-It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
-
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin2}.
-In our analyses if we include loss of earnings and hospitalisation costs, our estimates are higher, around US\$959 per case.
-This figure is from NZ\$ 852 for loss of wages and 0.17 * NZ \$ 1710 for costs related to attending hospital.
-However, costs can vary widely, with the containment of a single case and two secondary cases of measles in 2004 in Iowa, USA, estimated to cost US\$142,542 alone \citep{dayan5}.
-
-Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate amont the presently naive and susceptible New Zealand population until overyone has been infected.
-Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
-However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimated to be more than 18 (i.e.
-1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}).
-The benefits of catch up vaccination are clear once $R_v$ is greater than one (Table~\ref{tab:cost_sim}).
-If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r0}), then there is little financial benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}), though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
-
-Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outreaks depending on contact structure and other scenarios, such as the size of the local naive population.
-The spatial effects of measles transmission may have affected both our multivariate regression analyses (\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises.
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occuring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
-
-\subsection{Benefit--cost analysis summary}
-\begin{itemize}
-\item The mean wage losses per measles case is estimated to be \$852
-\item The mean cost per measles case attending hospital is estimated to be \$1710
-\item Approximately 17\% of measles cases attend hopsital
-\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
-\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
-\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
-\end{itemize}
-
-\section {Summary of key findings}
-\begin{itemize}
-\item Risk of measles infection decreases significantly with age
-\item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
-\item Pacific and Asian children in the 6--17 year age caterories have been at lower risk of measles than European or Maori children of the same age
-\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
-\item The mean wage losses per measles case is estimated to be \$852
-\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hopsital
-\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
-\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
-\item Large outbreaks, with a mean size of approximately 175 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
-\end{itemize}
-
-\section{Acknowledgments}
-The authors wish to thank Tomasz Kiedrzynski, Lisa Oakley and Nic Aagaard from the Ministry of Health, Ruth Pirie and colleagues from ESR, and June Atkinson from University of Otago for help in obtaining the appropriate materials for analyses.
+\caption{Size: DHB Population, Statistics NZ 2013;		
+Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
+Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
+Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$.			}
+\label{Table1}
+\end{table}%
 
@@
 Chichester: Wiley.
+
+\bibitem[Diekmann et al.(2013)]{diekmann13}
+Diekmann, O.,  J.~A.~P. Heesterbeek, and T. Britton (2013).
+\newblock \emph{Mathematical tools for understanding infectious disease dynamics}.
+Princeton: Princeton University Press.
 

### 2014-10-28 new_outbreaks.R

@@ -1,0 +1,158 @@
+## simulate measles outbreaks/epidemics using GT and R0 from data.
+# http://www.sciencedirect.com/science/article/pii/S0022519311003146
+mu    <- 12
+sigma <- 3.5
+sigma_logn <- sqrt(log(1 + (sigma/mu)^2))
+mu_logn    <- log(mu) - log(1 + (sigma/mu)^2) / 2
+
+# then exp(rnorm(n, mu_logn, sigma_logn)) simulates from lognormal with the given mean and sd.
+
+# the R0 library 
+require(R0)
+#source("estR0.R")
+
+# generation time 
+genTime <- generation.time(type="lognormal", val=c(12, 3.5))
+class(genTime)
+
+res<-sim.epid(epid.nb=10,GT=genTime,R0=2,epid.length=365,family="poisson",peak.value=100000)
+#matplot(res,pch=16,col="grey")
+#head(res)
+#colSums(res)
+sizes<-colSums(res)
+#sizes<-apply(res, 2, max)
+hist(sizes,breaks=1000)
+summary(sizes)
+
+jm.epid <- function (epid.nb, GT, R0, epid.length, family, negbin.size = NULL, 
+                     peak.value = 300000, popn = 300000) 
+{
+  if (class(GT) != "R0.GT") {
+    stop("GT object must be of class R0.GT.")
+  }
+  if (family == "negbin" & is.null(negbin.size)) {
+    negbin.size <- R0/4
+  }
+  GT <- GT$GT
+  epidemics <- matrix(data = 0, nrow = epid.length, ncol = epid.nb)
+  for (n in 1:epid.nb) {
+    sim.epid = c(1, rep(0, epid.length - 1))
+    for (t in 1:epid.length) {
+      if (family == "poisson") {
+        susc <- max(0,popn - sum(sim.epid, na.rm=T))
+        #    peak.value[t]<-susc
+        new <- rbinom(sim.epid[t], susc, min(1, R0/susc))
+        if (sum(new) > susc) # arggrggg
+        {
+          #      cat("doing the new thing", susc, sim.epid[t], sum(sim.epid[1:t]),"\n")
+          new <- rep(0, sim.epid[t])
+          for (i in 1:sim.epid[t]) {
+            #         cat("taking a random number, susc=", susc, "\n")
+            new[i] <- rbinom(1, susc, min(1, R0/susc))
+            susc <- susc - new[i]
+            if (susc == 0)
+            {
+              #             cat("got 'em all", sum(new), "\n")
+              break;
[…]

### 2014-10-28 maps/rworldmaps_measles_travel.R
Revised risk maps for report
Saved data for tables too

@@ -436,30 +436,55 @@ head(dpdf)
 np1<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Immigration") +
   geom_map(aes(fill = immigration), map =world.ggmap) +
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
+ # theme(text = element_text(size=30))+
   scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+pdf(paste("np1.pdf"), width=10, height=5)
+np1
+dev.off()
 
 np2<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Incidence (per million)") +
   geom_map(aes(fill = incidence), map =world.ggmap) +
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
+ # theme(text = element_text(size=30))+
   scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+pdf(paste("np2.pdf"), width=10, height=5)
+np2
+dev.off()
 
 np3<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Vaccination cover (%)") +
   geom_map(aes(fill = cover), map =world.ggmap) +
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
+  # theme(text = element_text(size=30))+
   scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
+pdf(paste("np3.pdf"), width=10, height=5)
+np3
+dev.off()
 
 grid.arrange(np1, np2, np3, ncol=3, main="Measles")
 
 dpdf$risk <- with(dpdf, mp$X2012.x/mp$Numeric.Value*1000* mp$Immigration)
 head(dpdf)
 write.csv(dpdf, "mapdata.csv", row.names=FALSE,col.names=T)
+write.csv(dpdf, "mapdata_immigration.csv", row.names=FALSE,col.names=T)
 dpdf<-read.csv("mapdata.csv",header=T)
 
-np4<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Risk 2012") +
+np4<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Risk  (Immigrant travel)") +
[…]

### 2014-10-28 draftfinalreport.Rnw
Massive revision of final draft report
Moved some files from maps folder - must tidy up...

--- 2014-09-01 interimreport2.Rnw
@@
 
-\title{Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
+\title{Draft final report\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
 \author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this interim report we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include further modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations.
-We provide preliminary benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
+In this draft final report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
+We identified additional analyses that may help understand risk of infection in New Zealand, and present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations.
+We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
 Our key findings were:
+
 \begin{itemize}
+\item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
+\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
+This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
 \item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
@@
 
-As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission.
-A brief review of the history of measles in New Zealand was provided in our previous interim report.
-In this report were present regression analyses to determine which populations are most at risk, the likely outcomes of measles infections within New Zealand based on a number of assumptions, and the benefit--cost analyses for vaccination dependent on differing scenarios.
-
-\section{Risk analysis update}
-\label{sub:risk_analyses}
+As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
+The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.
+
+The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
+The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}).
+In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
+
+Previous measles analyses, including two in New Zealand by Prof.
+Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
+Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
+Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
+Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
+
+\section{Risk analysis review}
 
@@
 Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, we found the report to be very thorough.
-However, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
-The additional analyses included in this section are multivariate regression modelling to account for confounding within the univariate analyses, particularly age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep).
-
+\begin{itemize}
+\item In this section we review the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}.
+\end{itemize}
+
+Overall, the review was very thorough.
+The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
+Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).
+
+Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand.
+Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
+
+Specific epidemiological details were provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}).
+Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}).
+Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
+
+The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
+
+We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe.
+Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
+Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
+However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
+This is particularly in light of our preliminary modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).
+
+The analyses we believe could further inform the understanding of risk from measles infection are:
+\begin{itemize}
+\item Multivariate modelling to account for confounding within the univariate analyses.
+\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
+\item Analyses of changing risk factors through time during outbreaks.
+\item Modelling transmission chains in the population to understand effective reproduction number.
+\item Development of risk maps to understand measles importation.
+\end{itemize}
+
+\subsection{Additional risk analyses}
+\label{sub:risk_analyses}
+
+In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
+These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
+     \caption{Measles incidence from 1997 to 2014}
+     \label{fig:incidence1997}
+\end{figure}
+
+In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
+
+\subsection{Additional risk analyses methods}
+
+For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Note this uses all immigration of foreign nationals, coming for whatever purpose.
+We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
+
+<<echo=FALSE>>=
+library(xtable)
+dpdf<-read.csv("mapdata.csv",header=T)
+# output top countries
+colnames(dpdf)[1]<-c("country")
+topimmigration <- dpdf[order(-dpdf$immigration),] 
+immigration<-topimmigration[1:10,c(1,4)]
+topincidence <- dpdf[order(-dpdf$incidence),] 
+incidence<-topincidence[1:10,c(1,2)]
+incidence[,2]<-round(incidence[,2],0)
+topvaccine <- dpdf[order(dpdf$cover),] 
+vaccinecover<-topvaccine[1:10,c(1,3)]
+toprisk <- dpdf[order(-dpdf$risk),] 
+risk<-(toprisk[1:10,c(1,5)])
+risk[,2]<-round(risk[,2],0)
+@
+
+\begin{table}
+\caption{Immigration numbers to New Zealand (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(immigration, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:immigration12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT NZ Travelers here
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT Total Travelers here
+%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cover12}
+\end{table}
+
+\begin{table}
+\caption{Measles incidence per million (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(incidence, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:incidence12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(risk, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:risk12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT NZ risk here
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT Total risk here
+%%%%%%%%%%%%%%%%%%
 \subsection{Risk analyses methods}
@@
 
+\subsection{Additional risk analyses results}
+
+Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
+However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}).
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
+These data are shown in Table \ref{table:risk12}.
+The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
+
+<<echo=FALSE>>=
+# haven't had time to sort out which packages I need
+library(maps)       # Provides functions that let us plot the maps
+library(mapdata)    # Contains the hi-resolution points that mark out the countries.
+library(rworldmap)
+library(adehabitat)
+library(rgdal)
+library(raster)
+library(maps)
+library(maptools)
+library(sp)
+#library(SDMTools)
+library(ggplot2)
+require(rgeos)
+library(ggmap)
+library(mapdata)
+#gpclibPermit()
+library(mapproj)
+library(plyr)
+library(grid)
+library(gridExtra)
+#gpclibPermit()
+library(utils)
+library(rgdal)
+#getwd()
+#setwd("~/data")
+#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
+load("world.ggmap.Rda")
+n <- length(unique(world.ggmap$id))
+id = unique(world.ggmap$id)
+id<-as.data.frame(id)
+dpdf<-read.csv("mapdata.csv",header=T)
+# output top countries
+
+np1<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+np2<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = incidence), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+np3<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = cover), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
+
+#grid.arrange(np1, np2, np3, ncol=3, main="")
+
+np4<-ggplot(dpdf, aes(map_id = id)) +
+  geom_map(aes(fill = risk), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
+#np4
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np1)
+@
+\end{center}
+\caption{Immigration to New Zealand 2012}
+\label{fig:immigration12}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT NZ Travellers here
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT Total Travelers here
+%%%%%%%%%%%%%%%%%%
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np2)
+@
+\end{center}
+\caption{Measles incidence per million 2012}
+\label{fig:incidence12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np3)
+@
+\end{center}
+\caption{Measles vaccination cover (\%) 2012 }
+\label{fig:cover12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np4)
+@
+\end{center}
+\caption{Risk map from measles incidence and immigration rates}
+\label{fig:risk12}
+\end{figure}
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT NZ risk here
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%
+% INSERT Total risk here
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE>>=
+change<-read.csv("change.csv",header=T)
+plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
+     bty="n")
+ylim <- range(c(0,max(change[,4])))
+xlim <- range(change[,1])
+par(fig = c(.6, 1, 0.6, 1), mar=c(0,0,0,0), new=TRUE)
+plot(change[,1],change[,4],type="l",xlab="Year",ylab="",
+     ylim=c(0,max(change[,4])),bg="grey",add=T,bty="n")
+@
+\end{center}
+\caption{Trend in global per capita measles incidence}
+\label{fig:trendincidence}
+\end{figure}
+
+Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
+This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
+Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).
+This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE>>=
+ttimes<-read.csv("times.csv",header=T)
+lmres<-lm(res~time,data=ttimes)
+#summary(lmres)
+plot(ttimes$res,xaxt="n",type="l",xlab="Date",ylab="")
+axis(1, at=1:71, labels=ttimes$mydate) 
+abline(lmres,col="red",
+       lty=2)
+acf(ttimes$res,lag.max = length(ttimes$res),plot=T,main="")
+@
+\end{center}
+\caption{Trend in total immigrations into New Zealand}
+\label{fig:trendimmigration}
+\end{figure}
+
+%%%%%%%%% ADD NAIVE ## BY DHB HERE
+
+%%%%%%%%%%%%%%%%%%%
+% ADD REPORT # 2 RISK ANAYSIS HERE
+%%%%%%%%%%%%%%%%%%
 \subsection{Regression analyses results}
@@
 \begin{itemize}
+\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
+\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
 \item Risk of measles infection decreases significantly with age
@@
 
+\subsection{Future risk analyses}
+We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
+Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
+Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above.
+These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data only became available to us on 3 July 2014, provided by the University of Otago.
+
+The following analyses are in progress, for inclusion in later reports:
+\begin{itemize}
+\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
+\item Update of the importation risk analyses using a broader range of years, including modelling the trend in importations.
+\end{itemize}
+Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
+We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
+meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
+Thus the data gap that we have that will hinder us providing fine scale risk maps is:
+\begin{itemize}
+\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
+\end {itemize}
+An additional data set that would enable us to develop the understanding of measles importation risk is:
+\begin {itemize}
+\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
+\end{itemize}
+
 \section{Modelling measles epidemics}
@@
 
-In this section we briefly review the parts of our previous interim report that are pertinent to this study and include details and new work that is relevant for the benefit--cost analysis (Section~\ref{sub:cost_benefit}).
-
 A previously-published model of the dynamics of measles infections in New Zealand has been used to evaluate the vaccination strategy in New Zealand of MMR1 at 15 months and MMR2 before 5 years \citep{roberts0,roberts4,tobias98}.
-The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented, though herd immunity of approximately 95 percent a homogeneously mixed immune population is recommended to interupt measles \citep{roberts0,roberts4}.
-Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur and the current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles.
-Thus, the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-The distribution of the naive population for each age class can be seen in Figure~\ref{fig:naive}, estimated using the NZ census data \citep{stats14} and the proportion of each age class immune based on data provided to us by the Ministry of Health on commencing this project.
+The results show that achieving coverage of greater than 90\% at both vaccination opportunities is necessary if future epidemics of measles are to be prevented.
+
+The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
+Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
+The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
+
+These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
+In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
+They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
+
+Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
+The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
 
@@
 The basic reproduction number of the infection under vaccination, $R_v$, is the expected number of secondary infections that would arise from a single primary infection introduced into a vaccinated population at equilibrium and is a robust indicator of the performance of a vaccination schedule.
-If $R_v < 1$ epidemics are predicted to be prevented (but see below) and infection persistence is prevented.
+If $R_v < 1$ epidemics are prevented.
 The case reproduction number of the infection at time $t$, $R_t$, is the expected number of secondary infections that arise from a single infection at a particular time and depends on the number in the population who are susceptible.
@@
 
-To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we previously estimated $R_v$ from all the outbreaks in New Zealand since 2009.
+To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
 To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
@@
 
+\section{Mick Introduction}
+
+The well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population is \citep{diekmann13}
+$$\log\left(1-\Pe\right)+\Ro\Pe=0$$
+where $\Ro$ is the basic reproduction number and $\Pe$ is the proportion of the population infected over the course of the outbreak.
+If a proportion $x_0$ of the population is susceptible following vaccination, then the  reproduction number under vaccination is $\Rr_V=x_0\Ro$, and the final size equation becomes
+$$\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0$$
+Hence the relationship between the proportion initially susceptible and the proportion infected in an epidemic is
+$$x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}$$
+In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks is $x_0-1/\Ro$.
+
+These formulae were applied at a District Health Board (DHB) level, assuming no mixing between DHBs.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{cases_by_dhb_2009_2014.pdf}
+     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
+     \label{fig:dhbcases}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
+     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
+     \label{fig:outbreaks}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{averageR0.pdf}
+     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
+     \label{fig:r0}
+\end{figure}
+
 \subsection{Modelling results and discussion}
+
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The probability density of the $R_v$ estimates for each outbreak all include one.
+Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
+An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+
+These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+However, sub-clinical and underreporting may lower the estimate.
+The relative contributions of both to our estimates are currently unknown.
 
@@
 
+\begin{table}[htdp]
+
+\begin{center}
+
+\begin{tabular}{lrrrr}
+\hline
+DHB  & Size  & Na\"{\i}ve    &   Attack	& Vacc		\\
+\hline
+Auckland	&	436350	&	52010	&	31159	&	17920	\\
+Bay of Plenty	&	206000	&	20679	&	8437	&	4585	\\
+Canterbury	&	482180	&	51357	&	24695	&	13687	\\
+Capital and Coast	&	283700	&	32625	&	18403	&	10461	\\
+Counties Manukau	&	469300	&	55544	&	32903	&	18880	\\
+Hawke's Bay	&	151700	&	15602	&	6846	&	3751	\\
+Hutt Valley	&	138380	&	15198	&	7836	&	4388	\\
+Lakes	&	98196	&	10558	&	5192	&	2886	\\
+MidCentral	&	162560	&	17328	&	8348	&	4628	\\
+Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	\\
+Northland	&	151690	&	14921	&	5688	&	3071	\\
+South Canterbury	&	55620	&	5238	&	1678	&	893	\\
+Southern	&	297420	&	31607	&	15115	&	8371	\\
+Tairawhiti &	43650	&	4769	&	2431	&	1359	\\
+Taranaki	&	109750	&	11473	&	5262	&	2899	\\
+Waikato	&	359310	&	39402	&	20248	&	11331	\\
+Wairarapa	&	41112	&	3932	&	1346	&	720	\\
+Waitemata	&	525550	&	58350	&	30774	&	17291	\\
+West Coast	&	32151	&	3197	&	1265	&	685	\\
+Whanganui	&	60120	&	6075	&	2530	&	1378	\\
+\hline			
+TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Size: DHB Population, Statistics NZ 2013;		
+Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
+Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
+Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$.			}
+\label{Table1}
+\end{table}%
+
 \subsection{Summary of modelling}
 \begin{itemize}
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
+\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
+ \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
@@
 
+\subsection{Future modelling}
+Future modelling we aim to perform are:
+\begin{itemize}
+\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
+\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
+\end{itemize}
+
 \section{Cost analyses}
@@
 In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current measles outbreak.
-For completeness, we include much of the introduction from the previous interim report, but we have revised some of the figures.
 
@@
 \label{sub:cost_benefit}
+
 Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
@@
 Overall, the cost per case from 2008--2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
+
+%%%%%%%% REVISE
+
+This final figure brings an approximate estimate of \$786,479 for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
 
@@
 
+\subsection{Cost analysis discussion}
+The results presented here are based on available data.
+While some of the data are complete and detailed, this is not true of all the data.
+In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
+For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
+In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
+In addition, similar data would be needed for non-publicly funded hospitals, e.g.
+private clinics.
+Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
+
+Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
+Similar data are needed for the period preceding 2014.
+If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
+However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
+As other studies have demonstrated, direct costs required to manage measles are not linear.
+
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
+In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
+They estimated direct costs per case to be less than US\$500.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
+The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
+
 \subsection{Benefit--cost analyses results}
@@
 
+\subsection{Cost analysis summary}
+
+\begin{itemize}
+\item Our initial estimates suggest the ongoing 2013--2014 measles outbreak has cost New Zealand over \$750,000.
+\end{itemize}
+
 \subsection{Benefit--cost analysis summary}
@@
 
+\subsection{Future cost-benefit analyses}
+Using the results above we aim to:
+\begin{itemize}
+\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses (section~\ref{sub:risk_analyses}) or modelling (section~\ref{sec:epidemic_modelling}).
+\end{itemize}
+
+We also require additional clarifications of the data, regarding:
+\begin {itemize}
+\item What hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight? Or, does it mean the case stayed for less than 24 hours?
+\end {itemize}
+Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
+
 \section {Summary of key findings}
 \begin{itemize}
+\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
+\item The cost of the current measles outbreak is estimated to be at least \$750,000.
+\item Analyses of outbreak data suggest that measles $R_v$ values often include 1 and in this year, 2014, are well above one.
+This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
 \item Risk of measles infection decreases significantly with age
@@
 
+\bibitem[Diekmann et al.(2013)]{diekmann13}
+Diekmann, O.,  J.~A.~P. Heesterbeek, and T. Britton (2013).
+\newblock \emph{Mathematical tools for understanding infectious disease dynamics}.
+Princeton: Princeton University Press.
+
 \bibitem[Edmunds et al.(2000)]{edmunds0}

### 2014-10-28 draftfinalreport.Rnw
Inserted figures

@@
 
-%%%%%%%%%%%%%%%%%%%
-% INSERT NZ Travelers here
-%%%%%%%%%%%%%%%%%%
-
+%%%%%%%%%%%%%%%%%%%%%
+% INSERT NZ Travlers here
 %%%%%%%%%%%%%%%%%%%
@@
 
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np1)
-@
-\end{center}
-\caption{Immigration to New Zealand 2012}
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{np1.pdf}
+     \caption{Non-New Zealander travel}
      \label{fig:immigration12}
@@
 
-%%%%%%%%%%%%%%%%%%%
-% INSERT NZ Travellers here
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total Travelers here
-%%%%%%%%%%%%%%%%%%
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{nznp1.pdf}
+     \caption{New Zealand travellers 2012}
+     \label{fig:nztravellers}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{totnp1.pdf}
+     \caption{Total travellers 2012}
+     \label{fig:tottravellers}
+\end{figure}
 
@@
 \begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np4)
-@
-\end{center}
+ \centering
+ \includegraphics[width=0.8\textwidth]{np4.pdf}
 \caption{Risk map from measles incidence and immigration rates}
@@
 
-%%%%%%%%%%%%%%%%%%%
-% INSERT NZ risk here
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total risk here
-%%%%%%%%%%%%%%%%%%
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{nznp4.pdf}
+     \caption{Risk map from measles incidence and New Zealander travel rates}
+     \label{fig:nzrisk12}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{totnp4.pdf}
+     \caption{Risk map from measles incidence and total travel rates}
+     \label{fig:totrisk12}
+\end{figure}
 
@@
 
-%%%%%%%%% ADD NAIVE ## BY DHB HERE
-
-%%%%%%%%%%%%%%%%%%%
-% ADD REPORT # 2 RISK ANAYSIS HERE
-%%%%%%%%%%%%%%%%%%
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{naive_map.pdf}
+     \caption{Numbers of naive individuals per District Health Board}
+     \label{fig:naive_map}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb1.pdf}
+     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+     \label{fig:Northland}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb2.pdf}
+     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+     \label{fig:Waitemata}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb3.pdf}
+     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+     \label{fig:Auckland}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb4.pdf}
+     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+     \label{fig:Counties_Manukau}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb5.pdf}
+     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+     \label{fig:Waikato}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb6.pdf}
+     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+     \label{fig:Lakes}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb7.pdf}
+     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+     \label{fig:BayofPlenty}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb8.pdf}
+     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+     \label{fig:Tairawhiti}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb9.pdf}
+     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+     \label{fig:Taranaki}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb10.pdf}
+     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+     \label{fig:HawkesBay}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb11.pdf}
+     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+     \label{fig:Whanganui}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb12.pdf}
+     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+     \label{fig:Midcentral}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb13.pdf}
+     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+     \label{fig:Hutt}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb14.pdf}
+     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+     \label{fig:CapitalandCoast}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb15.pdf}
+     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+     \label{fig:Wairarapa}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb16.pdf}
+     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+     \label{fig:NelsonMarlborough}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb16.pdf}
+     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+     \label{fig:WestCoast}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb17.pdf}
+     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
+     \label{fig:Canterbury}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb18.pdf}
+     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
+     \label{fig:SouthCanterbury}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{dhb19.pdf}
+     \caption{Numbers of naive individuals per age class, Southern District Health Board}
+     \label{fig:Southern}
+\end{figure}
+
 \subsection{Regression analyses results}

### 2014-10-28 draftfinalreport.Rnw
Revisions throughout

@@
      \centering
-     \includegraphics[width=0.8\textwidth]{incidence_1997_2014.pdf}
+     \includegraphics[width=1\textwidth]{incidence_1997_2014.pdf}
      \caption{Measles incidence from 1997 to 2014}
@@
 We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
+
+%%%%%%%%%%%%%%%%%%
+<<echo=FALSE>>=
+library(xtable)
+totdpdf<-read.csv("mapdata_tot.csv",header=T)
+# output top countries
+colnames(totdpdf)[1]<-c("country")
+toptot <- totdpdf[order(-totdpdf$immigration),] 
+immigration<-toptot[1:10,c(1,4)]
+topincidence <- totdpdf[order(-totdpdf$incidence),] 
+incidence<-topincidence[1:10,c(1,2)]
+incidence[,2]<-round(incidence[,2],0)
+topvaccine <- totdpdf[order(totdpdf$cover),] 
+vaccinecover<-topvaccine[1:10,c(1,3)]
+toprisk <- totdpdf[order(-totdpdf$risk),] 
+risk<-(toprisk[1:10,c(1,5)])
+risk[,2]<-round(risk[,2],0)
+@
+
+\begin{table}
+\caption{Total travel numbers (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(immigration, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:immigration12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cover12}
+\end{table}
+
+\begin{table}
+\caption{Measles incidence per million (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(incidence, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:incidence12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(risk, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:risk12}
+\end{table}
 
@@
 %%%%%%%%%%%%%%%%%%%%%
-% INSERT NZ Travlers here
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total Travelers here
-%%%%%%%%%%%%%%%%%%
 
@@
 \end{table}
-
 %%%%%%%%%%%%%%%%%%%
-% INSERT NZ risk here
+<<echo=FALSE>>=
+library(xtable)
+nzdpdf<-read.csv("mapdata_nzers.csv",header=T)
+# output top countries
+colnames(nzdpdf)[1]<-c("country")
+topnz <- nzdpdf[order(-nzdpdf$immigration),] 
+nz<-topnz[1:10,c(1,4)]
+topnzrisk <- nzdpdf[order(-nzdpdf$risk),] 
+nzrisk<-(topnzrisk[1:10,c(1,5)])
+nzrisk[,2]<-round(nzrisk[,2],0)
+@
+
+\begin{table}
+\caption{New Zealander travel numbers (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(nz, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:nztravel12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(nzrisk, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:nzrisk12}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%
 
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total risk here
 %%%%%%%%%%%%%%%%%%
@@
 
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=1\textwidth]{nzers.pdf}
+     \end{center}
+     \caption{Trends in travel}
+     \label{fig:travel}
+\end{figure}
+
 Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{naive_map.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{naive_map.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb1.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Northland District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb2.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb3.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Auckland District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb4.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb5.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Waikato District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb6.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Lakes District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb7.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb8.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb9.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb9.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb10.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb11.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb12.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb13.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Hutt District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb14.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb15.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb16.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb16.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, West Coast District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb17.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb18.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
@@
 \begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{dhb19.pdf}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
+     \end{center}
      \caption{Numbers of naive individuals per age class, Southern District Health Board}
@@
 
+<<echo=FALSE>>=
+costtable20<-read.csv("cost_benefit_20.csv",header=T)
+@
+
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 20 dollars per vaccine}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable20, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cost20}
+\end{sidewaystable}
+
+<<echo=FALSE>>=
+costtable50<-read.csv("cost_benefit_50.csv",header=T)
+@
+
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 50 dollars per vaccine}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable50, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cost50}
+\end{sidewaystable}
+
 The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}.

### 2014-10-28 draftfinalreport.tex
New benefit cost and final report draft
Using Mick's values and new simulations, added new benefit-cost
analyses, incorporating the management costs.
Revisions throughout

--- 2014-09-01 interimreport2.Rnw
@@
 
+\usepackage{Sweave}
 \begin{document}
-\SweaveOpts{concordance=TRUE}
-
-\title{Interim report 2\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
+\input{draftfinalreport-concordance}
+
+\title{Draft final report\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
 \author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
@@
 Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
-In our review of the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}, we found the report to be very thorough.
-However, we identified additional analyses that might further inform the understanding of risk from measles infection within New Zealand.
-The additional analyses included in this section are multivariate regression modelling to account for confounding within the univariate analyses, particularly age categories, prioritized ethnicities, and socioeconomic deprivation indices (NZDep).
-
+\begin{itemize}
+\item In this section we review the confidential report to the Western Pacific Regional Verification Commission for Measles Elimination risk assessment provided by the Ministry, titled \emph {Progress Towards Measles Elimination in New Zealand - Final}.
+\end{itemize}
+
+Overall, the review was very thorough.
+The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
+Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).
+
+Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand.
+Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
+
+Specific epidemiological details were provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}).
+Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}).
+Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
+
+The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
+
+We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe.
+Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
+Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
+However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
+This is particularly in light of our preliminary modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).
+
+The analyses we believe could further inform the understanding of risk from measles infection are:
+\begin{itemize}
+\item Multivariate modelling to account for confounding within the univariate analyses.
+\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
+\item Analyses of changing risk factors through time during outbreaks.
+\item Modelling transmission chains in the population to understand effective reproduction number.
+\item Development of risk maps to understand measles importation.
+\end{itemize}
+
+\subsection{Additional risk analyses}
+\label{sub:risk_analyses}
+
+In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
+These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=1\textwidth]{incidence_1997_2014.pdf}
+     \caption{Measles incidence from 1997 to 2014}
+     \label{fig:incidence1997}
+\end{figure}
+
+In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
+
+\subsection{Additional risk analyses methods}
+
+For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Note this uses all immigration of foreign nationals, coming for whatever purpose.
+We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
+
+%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Total travel numbers (2012)}
+%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$1799655$\tabularnewline
+United Kingdom&$ 401737$\tabularnewline
+China&$ 322076$\tabularnewline
+United States&$ 316058$\tabularnewline
+Fiji&$ 151443$\tabularnewline
+India&$ 107618$\tabularnewline
+Japan&$ 106716$\tabularnewline
+Germany&$  96308$\tabularnewline
+Korea, Republic of&$  87419$\tabularnewline
+France&$  85948$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:immigration12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
+%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
+\hline
+Equatorial Guinea&$34$\tabularnewline
+Somalia&$49$\tabularnewline
+Lesotho&$60$\tabularnewline
+Central African Republic&$65$\tabularnewline
+Papua New Guinea&$67$\tabularnewline
+Chad&$69$\tabularnewline
+Haiti&$69$\tabularnewline
+South Sudan&$70$\tabularnewline
+Gabon&$71$\tabularnewline
+Yemen&$71$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cover12}
+\end{table}
+
+\begin{table}
+\caption{Measles incidence per million (2012)}
+%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
+\hline
+Equatorial Guinea&$1617$\tabularnewline
+Nauru&$1100$\tabularnewline
+Democratic Republic of the Congo&$1096$\tabularnewline
+Somalia&$ 979$\tabularnewline
+Djibouti&$ 824$\tabularnewline
+Sudan&$ 786$\tabularnewline
+Burkina Faso&$ 447$\tabularnewline
+Romania&$ 342$\tabularnewline
+Ukraine&$ 280$\tabularnewline
+Sudan&$ 229$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:incidence12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012}
+%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+Australia&$15537152$\tabularnewline
+United Kingdom&$13386328$\tabularnewline
+Thailand&$ 4774688$\tabularnewline
+Malaysia&$ 4495338$\tabularnewline
+Indonesia&$ 2216871$\tabularnewline
+India&$ 1620172$\tabularnewline
+China&$ 1432659$\tabularnewline
+Ukraine&$  765096$\tabularnewline
+Ireland&$  734128$\tabularnewline
+Romania&$  655107$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:risk12}
+\end{table}
+
+\begin{table}
+\caption{Immigration numbers to New Zealand (2012)}
+%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$809775$\tabularnewline
+United Kingdom&$306177$\tabularnewline
+China&$256036$\tabularnewline
+United States&$194438$\tabularnewline
+Japan&$ 86676$\tabularnewline
+Germany&$ 83608$\tabularnewline
+Korea, Republic of&$ 73459$\tabularnewline
+France&$ 71448$\tabularnewline
+India&$ 69038$\tabularnewline
+Canada&$ 54981$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:immigration12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
+%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
+\hline
+Equatorial Guinea&$34$\tabularnewline
+Somalia&$49$\tabularnewline
+Lesotho&$60$\tabularnewline
+Central African Republic&$65$\tabularnewline
+Papua New Guinea&$67$\tabularnewline
+Chad&$69$\tabularnewline
+Haiti&$69$\tabularnewline
+South Sudan&$70$\tabularnewline
+Gabon&$71$\tabularnewline
+Yemen&$71$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cover12}
+\end{table}
+
+\begin{table}
+\caption{Measles incidence per million (2012)}
+%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
+\hline
+Equatorial Guinea&$1617$\tabularnewline
+Nauru&$1100$\tabularnewline
+Democratic Republic of the Congo&$1096$\tabularnewline
+Somalia&$ 979$\tabularnewline
+Djibouti&$ 824$\tabularnewline
+Sudan&$ 786$\tabularnewline
+Burkina Faso&$ 447$\tabularnewline
+Romania&$ 342$\tabularnewline
+Ukraine&$ 280$\tabularnewline
+Sudan&$ 229$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:incidence12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012}
+%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+United Kingdom&$10202161$\tabularnewline
+Australia&$ 6991116$\tabularnewline
+Malaysia&$ 3226580$\tabularnewline
+Thailand&$ 1576414$\tabularnewline
+China&$ 1138900$\tabularnewline
+India&$ 1039356$\tabularnewline
+Indonesia&$  984022$\tabularnewline
+Ukraine&$  664315$\tabularnewline
+Ireland&$  585413$\tabularnewline
+Romania&$  490731$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:risk12}
+\end{table}
+%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{New Zealander travel numbers (2012)}
+%latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$989880$\tabularnewline
+United States&$121620$\tabularnewline
+Fiji&$104720$\tabularnewline
+United Kingdom&$ 95560$\tabularnewline
+Cook Islands&$ 71960$\tabularnewline
+China&$ 66040$\tabularnewline
+Samoa&$ 46020$\tabularnewline
+Thailand&$ 41100$\tabularnewline
+India&$ 38580$\tabularnewline
+Canada&$ 20400$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:nztravel12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
+%latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+Australia&$8546036$\tabularnewline
+Thailand&$3198274$\tabularnewline
+United Kingdom&$3184166$\tabularnewline
+Malaysia&$1268758$\tabularnewline
+Indonesia&$1232849$\tabularnewline
+India&$ 580816$\tabularnewline
+China&$ 293759$\tabularnewline
+Samoa&$ 243492$\tabularnewline
+Philippines&$ 241740$\tabularnewline
+Nepal&$ 188450$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:nzrisk12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%
 \subsection{Risk analyses methods}
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-## get data
-# set wd
-# setwd("~/Massey 2014/DHayman_20140627")
-# read data
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-                 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-@
-
 \begin{figure}[h!]
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mfrow=c(1,1))
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="",cex.lab=2)
-hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1.8)
-@
+\includegraphics{draftfinalreport-015}
 \end{center}
@@
 \end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## plot cases / age class
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-                   data = test , FUN=sum)
-#caseyr
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-colnames(naive)<-c("AgeInYears","Naive","Cases")
-
-dose1=as.factor(round(vac$Dose1Mths/12,0));
-dose1<-summary(dose1)
-dose1<-as.data.frame(dose1)
-dose1$AgeInYears<-rownames(dose1)
-dose2=as.factor(round(na.omit(vac$Dose2Mths/12,0)));
-dose2<-summary(dose2)
-dose2<-as.data.frame(dose2)
-dose2$AgeInYears<-rownames(dose2)
-
-## need to merge vaccination data with data
-## to get years the cases were from
-datav<-merge(vac,data,by="CaseCode",all=T)
-##
-datav$Dose1Mths[which(is.na(datav$Dose1Mths))] <- 9999
-datav$Dose2Mths[which(is.na(datav$Dose2Mths))] <- 9999
-#########################
-datav$RptYear<-as.factor(datav$RptYear)
-datav$SurvWeek<-as.factor(datav$SurvWeek)
-datav$NZDep01<-as.factor(datav$NZDep01)
-datav$NZDep06<-as.factor(datav$NZDep06)
-datav$NZDep13<-as.factor(datav$NZDep13)
-datav$Dose1Mths<-as.factor(datav$Dose1Mths)
-datav$Dose2Mths<-as.factor(datav$Dose2Mths)
-
-#data$AgeInYears<-as.factor(data$AgeInYears)
-datav$EthnicityPrioritised<-as.factor(datav$EthnicityPrioritised)
-
-timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
-                 data = datav , FUN=sum,na.rm=F,na.action=na.pass)
-
-testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-library(plyr)
-testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
-testv$Dose2Mths<-revalue(testv$Dose2Mths, c("9999"=NA));
-testv$Dose1Mths<-as.numeric(testv$Dose1Mths)
-testv$Dose2Mths<-as.numeric(testv$Dose2Mths)
-testv$D2vac<-ifelse(testv$Dose1Mths > -1 & testv$Dose2Mths >= testv$Dose1Mths,testv$Dose2Mths,NA)
-testv$D1vac<-ifelse(testv$Dose1Mths > -1 & is.na(testv$D2vac) == T,testv$Dose1Mths,NA)
-testv$Unvac<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,testv$AgeInYears,NA)
-
-testv$VC<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,0,
-                 ifelse(testv$Dose1Mths >= 0 & is.na(testv$Dose2Mths) == T,1,2))
-AgeVac<-table(testv$VC,testv$AgeInYears)
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-AgeInYears<-(as.numeric(rownames(AgeVac)))
-AgeVac<-cbind(AgeVac,AgeInYears)       
-AgeInYears<-(as.numeric(rownames(pop)))
-pop<-cbind(pop,AgeInYears)
-colnames(pop)<-c("Population","AgeInYears")
-AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
-# head(AgeV)
-#AgeVac<-t(AgeVac)
-@
 
@@
 For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).
-<<echo=FALSE,include=FALSE,results=hide>>=
-## regression analyses - measles
-# setwd("~/Massey 2014/DHayman_20140627")
-# read data
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-                 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
-
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-                   data = test , FUN=sum)
-caseyr
-test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
-test$AgeInYears<-as.factor(test$AgeInYears)
-tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
-test$AgeInYears<-tage
-#summary(test$EthnicityPrioritised)
-teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
-                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
-
-test$EthnicityPrioritised<-teth
-head(test)
-test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
-test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
-
-# want to set NZ Deprivation to the appropriate year
-
-testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
-                              ifelse (test$RptYear == "2008",test$NZDep06,
-                              ifelse (test$RptYear == "2009",test$NZDep06,
-                              ifelse (test$RptYear == "2010",test$NZDep06,
-                              ifelse (test$RptYear == "2011",test$NZDep06,
-                              ifelse (test$RptYear == "2012",test$NZDep06,
-                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-head(testt)
-tail(testt)
-str(testt)
-testt$NZDep<-as.factor(testt$NZDep-1)
-
-testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
-                      ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
-                 data = testt , FUN=sum)
-dim(testtable)
-colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
-head(testtable)
-
-tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-                                 "9"="6-10","10"="6-10"));
-
-testtable$NZDep<-tnzd
-# head(testtable)
-tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
-                 data = testtable , FUN=sum)
-
-ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
-                 data = testtable , FUN=sum)
-@
 
@@
 \caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(ttyr, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:case}
+%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllr}
+\hline\hline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
+\hline
+1-5&0-2&Asian&$ 11$\tabularnewline
+6-10&0-2&Asian&$  8$\tabularnewline
+1-5&3-5&Asian&$  1$\tabularnewline
+1-5&6-17&Asian&$ 11$\tabularnewline
+6-10&6-17&Asian&$  5$\tabularnewline
+1-5&18-24&Asian&$  3$\tabularnewline
+6-10&18-24&Asian&$  5$\tabularnewline
+1-5&25+&Asian&$ 10$\tabularnewline
+6-10&25+&Asian&$ 13$\tabularnewline
+1-5&0-2&European&$ 83$\tabularnewline
+6-10&0-2&European&$ 64$\tabularnewline
+1-5&3-5&European&$ 42$\tabularnewline
+6-10&3-5&European&$ 17$\tabularnewline
+1-5&6-17&European&$219$\tabularnewline
+6-10&6-17&European&$ 80$\tabularnewline
+1-5&18-24&European&$ 34$\tabularnewline
+6-10&18-24&European&$ 36$\tabularnewline
+1-5&25+&European&$ 78$\tabularnewline
+6-10&25+&European&$ 51$\tabularnewline
+1-5&0-2&Maori&$ 18$\tabularnewline
+6-10&0-2&Maori&$ 48$\tabularnewline
+1-5&3-5&Maori&$  7$\tabularnewline
+6-10&3-5&Maori&$ 11$\tabularnewline
+1-5&6-17&Maori&$ 19$\tabularnewline
+6-10&6-17&Maori&$ 92$\tabularnewline
+1-5&18-24&Maori&$  5$\tabularnewline
+6-10&18-24&Maori&$  8$\tabularnewline
+1-5&25+&Maori&$  2$\tabularnewline
+6-10&25+&Maori&$  6$\tabularnewline
+6-10&0-2&MLA&$  3$\tabularnewline
+6-10&3-5&MLA&$  1$\tabularnewline
+6-10&6-17&MLA&$  1$\tabularnewline
+6-10&18-24&MLA&$  2$\tabularnewline
+6-10&25+&MLA&$  6$\tabularnewline
+1-5&0-2&Pacific&$  5$\tabularnewline
+6-10&0-2&Pacific&$ 58$\tabularnewline
+6-10&3-5&Pacific&$  3$\tabularnewline
+1-5&6-17&Pacific&$  5$\tabularnewline
+6-10&6-17&Pacific&$ 22$\tabularnewline
+1-5&18-24&Pacific&$  1$\tabularnewline
+6-10&18-24&Pacific&$  8$\tabularnewline
+1-5&25+&Pacific&$  2$\tabularnewline
+6-10&25+&Pacific&$ 11$\tabularnewline
+1-5&0-2&None&$  3$\tabularnewline
+1-5&3-5&None&$  1$\tabularnewline
+1-5&6-17&None&$  3$\tabularnewline
+6-10&6-17&None&$  4$\tabularnewline
+1-5&18-24&None&$  2$\tabularnewline
+6-10&18-24&None&$  1$\tabularnewline
+1-5&25+&None&$  5$\tabularnewline
+6-10&25+&None&$  3$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:case}
 \end{table}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-### denominator data
-# setwd("~/Massey_2014/measles/data")
-denom<-read.csv("NZDep2006Denominators.csv",header=T)
-head(denom)
-denom<-denom[,-c(9)]
-summary(denom)
-dim(denom)
-denom<-denom[,1:18]
-row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
-sum(row.has.na)
-denomt <- denom[!row.has.na,]
-head(denomt)
-denomt<-denomt[denomt$Sex_code==99,]
-denomt$Eth_Level<-as.factor(denomt$Eth_Level)
-denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
-dm<-cbind(denomt[,c(3,6,9:18)])
-head(dm)
-dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
-dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
-dm$Age_Label<-dage
-colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-head(dm)
-datadm<-dm[dm$Age=="<6",]
-dataA<-(rbind(datadm,datadm)/2)
-dataA<-round(dataA[,3:12])
-dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
-colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-ddm<-dm[!dm$Age=="<6",]
-dm<-rbind(dataA,ddm)
-dm$Ethnicity <- factor(dm$Ethnicity)
-deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
-                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
-dm$Ethnicity<-deth
-head(dm)
-
-popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
-colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
-popn
-pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
-                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
-                                 "Dep9"="6-10","Dep10"="6-10"));
-
-popn$NZDep<-pnzd
-head(popn)
-tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
-               data = popn , FUN=sum)
-
-tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
-ttyr$merge <- paste(ttyr$NZDep, ttyr$Age, ttyr$Ethnicity)
-ttyr <- ttyr[ttyr$Ethnicity!="None",]
-
-tp$cases <- 0
-cases <- matrix(0, length(tp$merge),1)
-rownames(cases) <- tp$merge
-cases[ttyr$merge,] <- ttyr$Cases
-tp$cases <- cases
-PerCapita<-(tp$cases/tp$Popn)
-PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita,4))
-colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
-@
 
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="Category",cex.lab=2)
-legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
-       col=c(1:4,6),pch=16,bty="n",cex=1.8)
-@
+\includegraphics{draftfinalreport-023}
 \end{center}
@@
 \caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:percap}
+%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllrrr}
+\hline\hline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
+\hline
+1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
+6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
+1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
+6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
+1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
+6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
+1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
+6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
+1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
+6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
+1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
+6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
+1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
+6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
+1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
+6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
+1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
+6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
+1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
+6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
+1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
+6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
+1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
+6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
+1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
+6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
+1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
+6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
+1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
+6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
+1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
+6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
+1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
+6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
+1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
+6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
+1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
+6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
+1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
+6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
+1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
+6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
+1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
+6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
+1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
+6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
+1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
+6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
+1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
+6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:percap}
 \end{table}
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-## plot this figure?
-#hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-tp$Ethnicity<- relevel(tp$Ethnicity, "European")
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
-
-model2<-update(model1,~.-Age:NZDep)
-summary(model2)
-anova(model2,test="F")
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-tpsub<-tp[!(tp$Ethnicity=="MLA"),]
-
-#hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
-
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F")
-
-model2<-update(model1,~.-Ethnicity:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-model3<-update(model2,~.-Age:NZDep)
-summary(model3)
-anova(model3,test="F")
-@
+\subsection{Additional risk analyses results}
+
+Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
+However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}).
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
+These data are shown in Table \ref{table:risk12}.
+The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{np1.pdf}
+     \caption{Non-New Zealander travel}
+     \label{fig:immigration12}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{nznp1.pdf}
+     \caption{New Zealand travellers 2012}
+     \label{fig:nztravellers}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{totnp1.pdf}
+     \caption{Total travellers 2012}
+     \label{fig:tottravellers}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-028}
+\end{center}
+\caption{Measles incidence per million 2012}
+\label{fig:incidence12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-029}
+\end{center}
+\caption{Measles vaccination cover (\%) 2012 }
+\label{fig:cover12}
+\end{figure}
+
+\begin{figure}[h!]
+ \centering
+ \includegraphics[width=0.8\textwidth]{np4.pdf}
+\caption{Risk map from measles incidence and immigration rates}
+\label{fig:risk12}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{nznp4.pdf}
+     \caption{Risk map from measles incidence and New Zealander travel rates}
+     \label{fig:nzrisk12}
+\end{figure}
+
+\begin{figure}
+     \centering
+     \includegraphics[width=0.8\textwidth]{totnp4.pdf}
+     \caption{Risk map from measles incidence and total travel rates}
+     \label{fig:totrisk12}
+\end{figure}
+
+%%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-030}
+\end{center}
+\caption{Trend in global per capita measles incidence}
+\label{fig:trendincidence}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=1\textwidth]{nzers.pdf}
+     \end{center}
+     \caption{Trends in travel}
+     \label{fig:travel}
+\end{figure}
+
+Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
+This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
+Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).
+This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-031}
+\end{center}
+\caption{Trend in total immigrations into New Zealand}
+\label{fig:trendimmigration}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{naive_map.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per District Health Board}
+     \label{fig:naive_map}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+     \label{fig:Northland}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+     \label{fig:Waitemata}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+     \label{fig:Auckland}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+     \label{fig:Counties_Manukau}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+     \label{fig:Waikato}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+     \label{fig:Lakes}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+     \label{fig:BayofPlenty}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+     \label{fig:Tairawhiti}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb9.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+     \label{fig:Taranaki}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+     \label{fig:HawkesBay}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+     \label{fig:Whanganui}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+     \label{fig:Midcentral}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+     \label{fig:Hutt}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+     \label{fig:CapitalandCoast}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+     \label{fig:Wairarapa}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+     \label{fig:NelsonMarlborough}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+     \label{fig:WestCoast}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
+     \label{fig:Canterbury}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
+     \label{fig:SouthCanterbury}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Southern District Health Board}
+     \label{fig:Southern}
+\end{figure}
 
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
-@
+\includegraphics{draftfinalreport-032}
 \end{center}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width =11, height=11>>=
-res<-predict(model3)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-@
+\includegraphics{draftfinalreport-033}
 \end{center}
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
-
 \begin{figure}[h!]
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-@
+\includegraphics{draftfinalreport-035}
 \end{center}
@@
 
-<<label=regmodelanova,include=FALSE>>=
-anova(model3,test="F")
-@
+\begin{Schunk}
+\begin{Sinput}
+> anova(model3,test="F")
+\end{Sinput}
+\begin{Soutput}
+Analysis of Deviance Table
+
+Model: quasipoisson, link: log
+
+Response: cases
+
+Terms added sequentially (first to last)
+
+              Df Deviance Resid. Df Resid. Dev        F   Pr(>F)    
+NULL                             39    1453.62                      
+Age            4  1304.20        35     149.43 183.5795 6.73e-15 ***
+Ethnicity      3    20.00        32     129.43   3.7529 0.028500 *  
+NZDep          1    10.70        31     118.74   6.0236 0.023932 *  
+Age:Ethnicity 12    81.91        19      36.83   3.8430 0.004459 ** 
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<label=regmodel,include=FALSE>>=
-summary(model3)
-@
+\begin{Schunk}
+\begin{Sinput}
+> summary(model3)
+\end{Sinput}
+\begin{Soutput}
+Call:
+glm(formula = cases ~ Age + Ethnicity + NZDep + Age:Ethnicity + 
+    offset(log(Popn)), family = "quasipoisson", data = tpsub)
+
+Deviance Residuals: 
+     Min        1Q    Median        3Q       Max  
+-2.76316  -0.52409   0.04736   0.49883   1.92918
+
+Coefficients:
+                          Estimate Std. Error t value Pr(>|t|)    
+(Intercept)               -6.46738    0.11483 -56.324  < 2e-16 ***
+Age3-5                    -0.91290    0.20539  -4.445 0.000278 ***
+Age6-17                   -0.76175    0.13427  -5.673 1.81e-05 ***
+Age18-24                  -1.50501    0.19366  -7.772 2.57e-07 ***
+Age25+                    -2.95073    0.16079 -18.352 1.51e-13 ***
+EthnicityAsian             0.05304    0.32498   0.163 0.872072    
+EthnicityMaori             0.21245    0.19935   1.066 0.299919    
+EthnicityPacific           1.16032    0.20429   5.680 1.78e-05 ***
+NZDep6-10                 -0.21192    0.08548  -2.479 0.022713 *  
+Age3-5:EthnicityAsian     -2.03154    1.38265  -1.469 0.158112    
+Age6-17:EthnicityAsian    -1.00742    0.47171  -2.136 0.045940 *  
+Age18-24:EthnicityAsian   -0.83115    0.59409  -1.399 0.177917    
+Age25+:EthnicityAsian      0.42108    0.44334   0.950 0.354143    
+Age3-5:EthnicityMaori     -0.38639    0.40959  -0.943 0.357344    
+Age6-17:EthnicityMaori    -0.08552    0.24685  -0.346 0.732812    
+Age18-24:EthnicityMaori   -0.58278    0.44835  -1.300 0.209209    
+Age25+:EthnicityMaori     -1.04821    0.52418  -2.000 0.060036 .  
+Age3-5:EthnicityPacific   -2.13163    0.81388  -2.619 0.016882 *  
+Age6-17:EthnicityPacific  -1.45411    0.33466  -4.345 0.000349 ***
+Age18-24:EthnicityPacific -0.98266    0.51289  -1.916 0.070546 .  
+Age25+:EthnicityPacific   -0.61270    0.43665  -1.403 0.176692    
+---
+Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+
+(Dispersion parameter for quasipoisson family taken to be 1.776063)
+
+    Null deviance: 1453.62  on 39  degrees of freedom
+Residual deviance:   36.83  on 19  degrees of freedom
+AIC: NA
+
+Number of Fisher Scoring iterations: 5
+\end{Soutput}
+\end{Schunk}
 
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-@
-
 \begin{figure}[ht!]
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
+\includegraphics{draftfinalreport-039}
 \end{center}
@@
 
-<<echo=FALSE>>=
-#setwd("C:\Users\dtshayma\Documents\GitHub\measles")
-costtable1<-read.csv("costtable1.csv",header=T)
-@
-
 \begin{table}
 \caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable1, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:direct}
+%latex.default(costtable1, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllll}
+\hline\hline
+\multicolumn{1}{c}{Category}&\multicolumn{1}{c}{January}&\multicolumn{1}{c}{February}&\multicolumn{1}{c}{March}&\multicolumn{1}{c}{Total}\tabularnewline
+\hline
+PHN&55,296&71,175&24,087&150,558\tabularnewline
+PHA&0&0&2,656&2,656\tabularnewline
+Data support&0&7,752&4,552&12,304\tabularnewline
+Supervisors&10,656&10,464&3,232&24,352\tabularnewline
+IMT&32,918&28,624&7,156&68,698\tabularnewline
+SSO&0&2,746&1,186&3,932\tabularnewline
+Measles operations&1,800&10,326&6,678&18,804\tabularnewline
+Operations partner&2,187&14,580&7,290&24,057\tabularnewline
+IMT controller partner&2,916&14,580&7,290&24,786\tabularnewline
+Total&105,773&160,247&64,127&330,147\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:direct}
 \end{table}
@@
 
-<<echo=FALSE>>=
-costtable2<-read.csv("costtable2v2.csv",header=T)
-@
-
 \begin{table}
 \caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable2, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:hosp}
+%latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrlll}
+\hline\hline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Days}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Per.case}&\multicolumn{1}{c}{Per.day}\tabularnewline
+\hline
+2000&$  6$&$ 13$&8,850&1,475&681\tabularnewline
+2001&$ 13$&$ 18$&11,267&867&626\tabularnewline
+2002&$  5$&$  2$&3,869&774&1,934\tabularnewline
+2003&$  9$&$ 12$&10,241&1,138&853\tabularnewline
+2004&$  4$&$  5$&4,765&1,191&953\tabularnewline
+2005&$  3$&$ 11$&5,111&1,704&465\tabularnewline
+2006&$  1$&$  0$&602&602&NC\tabularnewline
+2007&$  5$&$ 25$&82,977&16,595&3,319\tabularnewline
+2008&$  3$&$  1$&3,038&1,013&3,038\tabularnewline
+2009&$ 29$&$ 38$&40,782&1,406&1,073\tabularnewline
+2010&$  5$&$  5$&6,701&1,340&1,340\tabularnewline
+2011&$132$&$189$&205,303&1,555&1,086\tabularnewline
+2012&$ 19$&$ 12$&28,540&1,502&2,378\tabularnewline
+2013&$  4$&$  6$&5,330&1,333&888\tabularnewline
+2014&$ 55$&$133$&132,648&2,412&997\tabularnewline
+TOTAL&$293$&$470$&550,024&1,877&1,170\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:hosp}
  \centering
@@
 
-<<echo=FALSE>>=
-costtable3<-read.csv("costtable3.csv",header=T)
-@
-
 \begin{table}
 \caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable3, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:freq}
+%latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrr}
+\hline\hline
+\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Admitted}&\multicolumn{1}{c}{Proportion}\tabularnewline
+\hline
+\textless  15 months&$ 21$&$10$&$0.47$\tabularnewline
+15 months – 3 years&$  7$&$ 1$&$0.14$\tabularnewline
+4 – 9 years&$  8$&$ 0$&$0.00$\tabularnewline
+10 – 19 years&$132$&$12$&$0.09$\tabularnewline
+\textgreater  19 years&$ 33$&$11$&$0.33$\tabularnewline
+Total&$201$&$34$&$0.17$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:freq}
 \end{table}
@@
 
-<<echo=FALSE>>=
-hosp<-read.csv("hospital.csv",header=T)
-library(ggplot2)
-p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")+theme(axis.title.x = element_text(size=25),axis.title.y = element_text(size=25),axis.text.x  = element_text(size=16),axis.text.y  = element_text(size=16))
-@
-
 \begin{figure}[h!]
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(p)
-@
+\includegraphics{draftfinalreport-047}
 \end{center}
@@
 
-<<echo=FALSE>>=
-costtable4<-read.csv("costtable4v2.csv",header=T)
-@
-
 \begin{table}\small
 \caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable4, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cases}
+%latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllrrl}
+\hline\hline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Gender}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Length.of.stay}&\multicolumn{1}{c}{Cost.per.case}\tabularnewline
+\hline
+2000&F&4,296&$  2$&$  4$&2,148\tabularnewline
+&M&4,554&$  4$&$  9$&1,139\tabularnewline
+&Total&8,850&$  6$&$ 13$&1,475\tabularnewline
+2001&F&3,740&$  5$&$  5$&748\tabularnewline
+&M&7,527&$  8$&$ 13$&941\tabularnewline
+&Total&11,267&$ 13$&$ 18$&867\tabularnewline
+2002&F&924&$  2$&$  0$&462\tabularnewline
+&M&2,945&$  3$&$  2$&982\tabularnewline
+&Total&3,869&$  5$&$  2$&774\tabularnewline
+2003&F&9,766&$  8$&$ 12$&1,221\tabularnewline
+&M&475&$  1$&$  0$&475\tabularnewline
+&Total&10,241&$  9$&$ 12$&1,138\tabularnewline
+2004&F&1,437&$  1$&$  2$&1,437\tabularnewline
+&M&3,328&$  3$&$  3$&1,109\tabularnewline
+&Total&4,765&$  4$&$  5$&1,191\tabularnewline
+2005&F&0&$  0$&$  0$&0\tabularnewline
+&M&5,111&$  3$&$ 11$&1,704\tabularnewline
+&Total&5,111&$  3$&$ 11$&1,704\tabularnewline
+2006&F&0&$  0$&$  0$&0\tabularnewline
+&M&602&$  1$&$  0$&602\tabularnewline
+&Total&602&$  1$&$  0$&602\tabularnewline
+2007&F&1,930&$  1$&$  3$&1,930\tabularnewline
+&M&81,046&$  4$&$ 22$&20,262\tabularnewline
+&Total&82,977&$  5$&$ 25$&16,595\tabularnewline
+2008&F&714&$  1$&$  0$&714\tabularnewline
+&M&2,324&$  2$&$  1$&1,162\tabularnewline
+                    &Total&3,038&$  3$&$  1$&1,013\tabularnewline
+2009&F&11,953&$  7$&$ 15$&1,708\tabularnewline
+&M&28,830&$ 22$&$ 23$&1,310\tabularnewline
+                    &Total&40,782&$ 29$&$ 38$&1,406\tabularnewline
+2010&F&5,884&$  4$&$  5$&1,471\tabularnewline
+&M&817&$  1$&$  0$&817\tabularnewline
+                    &Total&6,701&$  5$&$  5$&1,340\tabularnewline
+2011&F&103,460&$ 66$&$ 86$&1,568\tabularnewline
+&M&101,842&$ 66$&$103$&1,543\tabularnewline
+                    &Total&205,303&$132$&$189$&1,555\tabularnewline
+2012&F&13,054&$  8$&$  6$&1,632\tabularnewline
+&M&15,486&$ 11$&$  6$&1,408\tabularnewline
+                    &Total&28,540&$ 19$&$ 12$&1,502\tabularnewline
+2013&F&1,800&$  1$&$  2$&1,800\tabularnewline
+&M&3,530&$  3$&$  4$&1,177\tabularnewline
+                    &Total&5,330&$  4$&$  6$&1,333\tabularnewline
+2014&F&55,633&$ 21$&$ 46$&2,649\tabularnewline
+&M&77,014&$ 34$&$ 87$&2,265\tabularnewline
+&Total&132,647&$ 55$&$133$&2,412\tabularnewline
+2000-2014&      F&335,431&$166$&$284$&2,021\tabularnewline
+&     M&214,591&$127$&$186$&1,690\tabularnewline
+&   TOTAL&550,022&$293$&$470$&1,877\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cases}
  \centering
@@
 
-<<echo=FALSE>>=
-costtable5<-read.csv("costtable5v2.csv",header=T)
-@
-
 \begin{table}
 \caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable5, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:regression}
+%latex.default(costtable5, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrl}
+\hline\hline
+\multicolumn{1}{c}{Variable}&\multicolumn{1}{c}{Coefficient}&\multicolumn{1}{c}{P.value}\tabularnewline
+\hline
+Intercept&$ 581.39$&\textless  0.001\tabularnewline
+Length of stay (nights)&$ 406.07$&\textless  0.001\tabularnewline
+Gender (0 = F, 1 = M)&$-255.98$&0.006\tabularnewline
+Case age (years)&$  -8.23$&0.007\tabularnewline
+Year of case (vs. 2000)&$  64.35$&\textless  0.001\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:regression}
 \end{table}
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-#naive[is.na(naive)] <- 0
-colnames(naive)<-c("AgeInYears","Naive","Cases")
-@
-
 \begin{figure}
      \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-##
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(round(0.53*naive$Naive) ,col="grey",bg="red",pch=21,ylab="Number to vaccinate",xlab="Age in Years",
-     main="",cex.lab=2)
-points(naive$Naive[0:21],col="grey",bg="orange",pch=22) # 
-points(naive$Naive[3:30],col="grey",bg="yellow",pch=23) 
-points(jitter(round(0.16*naive$Naive)),col="grey",bg="lightblue",pch=21) 
-points(jitter(naive$Naive[5:13]),col="grey",pch=22,bg="skyblue") # 
-points(naive$Naive[1:5],col="grey",pch=23,bg="navyblue") #
-legend("topright",c("53% naive, all ages","53% naive, all 0-21 yr olds","53% naive, all 3-30 yr olds",
-                    "16% naive, all ages","16% naive, all 5-13 yr olds","16% naive, all 1-5 yr olds"),
-       bty="n",pch=c(rep(21:23,2)),col=rep("grey",6),pt.bg=c("red","orange","yellow","lightblue","skyblue","navyblue"),cex=2)
-@
+\includegraphics{draftfinalreport-053}
      \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave}}
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-## simulate measles outbreaks/epidemics using GT and R0 from data.
-# http://www.sciencedirect.com/science/article/pii/S0022519311003146
-mu    <- 12
-sigma <- 3.5
-sigma_logn <- sqrt(log(1 + (sigma/mu)^2))
-mu_logn    <- log(mu) - log(1 + (sigma/mu)^2) / 2
-
-# then exp(rnorm(n, mu_logn, sigma_logn)) simulates from lognormal with the given mean and sd.
-
-# the R0 library 
-require(R0)
-#source("estR0.R")
-
-# generation time 
-genTime <- generation.time(type="lognormal", val=c(12, 3.5))
-
-jm.epid <- function (epid.nb, GT, R0, epid.length, family, negbin.size = NULL, 
-          peak.value = 300000, popn = 300000) 
-{
-  if (class(GT) != "R0.GT") {
-    stop("GT object must be of class R0.GT.")
-  }
-  if (family == "negbin" & is.null(negbin.size)) {
-    negbin.size <- R0/4
-  }
-  GT <- GT$GT
-  epidemics <- matrix(data = 0, nrow = epid.length, ncol = epid.nb)
-  for (n in 1:epid.nb) {
-    sim.epid = c(1, rep(0, epid.length - 1))
-    for (t in 1:epid.length) {
-      if (family == "poisson") {
-        susc <- max(0,popn - sum(sim.epid, na.rm=T))
-    #    peak.value[t]<-susc
-        new <- rbinom(sim.epid[t], susc, min(1, R0/susc))
-        if (sum(new) > susc) # arggrggg
-        {
-    #      cat("doing the new thing", susc, sim.epid[t], sum(sim.epid[1:t]),"\n")
-          new <- rep(0, sim.epid[t])
-          for (i in 1:sim.epid[t]) {
-   #         cat("taking a random number, susc=", susc, "\n")
-            new[i] <- rbinom(1, susc, min(1, R0/susc))
-            susc <- susc - new[i]
-            if (susc == 0)
-            {
- #             cat("got 'em all", sum(new), "\n")
-              break;
-            }
-          }
-        }
-#        cat("generating", new[1], "individuals from first, susc=", susc, "\n")
-      }
-      else if (family == "negbin") {
-        new <- rnbinom(sim.epid[t], size = negbin.size, 
-                       mu = R0)
-      }
-if (is.na(sum(new)))
-  break
-
-      newd <- rmultinom(1, sum(new), GT)[, 1]
-      sim.epid[t:(t + length(GT) - 1)] <- sim.epid[t:(t + 
-                                                        length(GT) - 1)] + newd
-      if (sim.epid[t + 1] > peak.value & t < (epid.length - 
-                                                1)) {
-        sim.epid[(t + 2):epid.length] <- 0
-        break
-      }
-    }
-    sim.epid <- sim.epid[!is.na(sim.epid)]
-    epidemics[, n] <- sim.epid
-  }
-  return(epidemics)
-}
-
-## if R 0 == 2
-set.seed(1) # all infected
-res<-jm.epid(epid.nb=1000,GT=genTime,R0=1,epid.length=365*5,popn=4.2*1e6 * 0.11
-             ,family="poisson",peak.value=4.2*1e6 * 0.11)
-sizes<-colSums(res)
-sizes
-legMd<-as.factor(paste("median =",c(median(sizes))))
-legMn<-as.factor(paste("mean =",c(round(mean(sizes)))))
-@
-
 \begin{figure}
      \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(sizes,xlim=c(0,200),col="grey",breaks=3000,xlab="Outbreak size",main="",cex.lab=2)
-#summary(sizes)
-abline(v=median(sizes),col="orange",lty=3,lwd=10)
-abline(v=mean(sizes),col="red",lty=3,lwd=10)
-#length(which(sizes>5))/1000
-legend('top',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
-@
+\includegraphics{draftfinalreport-055}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
@@
      \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(sizes,breaks=1000,xlab="Outbreak size",main="",cex.lab=2)
-abline(v=median(sizes),col="orange",lty=3,lwd=10)
-abline(v=mean(sizes),col="red",lty=3,lwd=10)
-legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
-@
+\includegraphics{draftfinalreport-056}
      \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
@@
 We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
+
+\subsection{Cost analysis discussion}
+The results presented here are based on available data.
+While some of the data are complete and detailed, this is not true of all the data.
+In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
+For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
+In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
+In addition, similar data would be needed for non-publicly funded hospitals, e.g.
+private clinics.
+Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
+
+Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
+Similar data are needed for the period preceding 2014.
+If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
+However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
+As other studies have demonstrated, direct costs required to manage measles are not linear.
+
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
+In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
+They estimated direct costs per case to be less than US\$500.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
+The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
 
@@
 
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 20 dollars per vaccine}
+%latex.default(costtable20, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrrrrrrrrr}
+\hline\hline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
+\hline
+Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$140.19$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$ 78.93$\tabularnewline
+Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$146.01$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$103.46$\tabularnewline
+Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$166.36$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$ 80.20$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$ 65.11$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$ 61.38$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$ 75.06$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$ 40.68$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$ 60.49$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$ 21.24$\tabularnewline
+Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$ 90.23$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$ 42.18$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$ 58.04$\tabularnewline
+Waikato&$11331$&$226620$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$110.30$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$ 20.85$\tabularnewline
+Waitemata&$17291$&$345820$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$150.07$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$ 22.89$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$ 36.86$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cost20}
+\end{sidewaystable}
+
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 50 dollars per vaccine}
+%latex.default(costtable50, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrrrrrrrrr}
+\hline\hline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
+\hline
+Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$72.02$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$52.49$\tabularnewline
+Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$74.87$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$61.20$\tabularnewline
+Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$78.44$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$52.91$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$45.59$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$43.82$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$50.42$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$32.41$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$43.72$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$18.75$\tabularnewline
+Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$56.86$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$33.03$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$42.20$\tabularnewline
+Waikato&$11331$&$566550$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$63.97$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$18.43$\tabularnewline
+Waitemata&$17291$&$864550$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$75.46$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$19.99$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$29.84$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cost50}
+\end{sidewaystable}
+
 The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}.
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-AgeVac<-table(testv$VC,testv$AgeInYears)
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-@
+\includegraphics{draftfinalreport-061}
 \end{center}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-@
+\includegraphics{draftfinalreport-062}
 \end{center}
@@
 Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
+
+\subsection{Cost analysis summary}
+
+\begin{itemize}
+\item Our initial estimates suggest the ongoing 2013--2014 measles outbreak has cost New Zealand over \$750,000.
+\end{itemize}
 

### 2014-10-28 draftfinalreport.tex
revised South Sudan v Sudan in risk tabs

@@
 Djibouti&$ 824$\tabularnewline
-Sudan&$ 786$\tabularnewline
+South Sudan&$ 786$\tabularnewline
 Burkina Faso&$ 447$\tabularnewline

### 2014-10-29 draftfinalreport.Rnw
revisions

--- 2014-10-28 draftfinalreport.Rnw
@@
 \usepackage{graphics}
+\usepackage{subfig}
 \usepackage{amsmath}
@@
 \usepackage{caption2}
+\usepackage{caption}
 \usepackage{fancyhdr}
+\usepackage{placeins}
+\usepackage{framed}
 \newcommand{\vecA}{\mathbf{A}}
@@
 
-New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, but significant outbreaks of measles after measles introductions from abroad.
-In this draft final report we reviewed the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
-We identified additional analyses that may help understand risk of infection in New Zealand, and present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007, provide updated cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on several alternative situations.
+New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks of measles after measles introductions from abroad.
+In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
+We identified additional analyses that may help understand risk of infection in New Zealand, and present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007.
+We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level.
 We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
@@
 \item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
+\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
-\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$750,000.
-\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, is well above one.
+\item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
+However, New Zealander and immigrant or non--New Zealander travelling is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, as analysed from data from April, is well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
-\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
-\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist range from 0-53\%, leading to an additional 0--250,000 additional vaccinations.
-\item The benefit--cost ratio is dependent on expectations regarding measles transmission: if each measles case in New Zealand continues to infect the same number of secondary cases as we estimate from the current 2013--2014 outbreak, additional vaccination is extremely beneficial, but if outbreak dynamics remain the same as those of the average outbreaks since 1997 and future measles cases are few, there is no benefit to additional vaccination campaigns.
+\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
+\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
+\item The benefit--cost ratio analyses suggest additional vaccination is extremely beneficial financially.
 \end{itemize}
@@
 
-The analyses we believe could further inform the understanding of risk from measles infection are:
-\begin{itemize}
-\item Multivariate modelling to account for confounding within the univariate analyses.
-\item Inclusion of multivariate model results with local level immunity data to develop a risk map for New Zealand.
-\item Analyses of changing risk factors through time during outbreaks.
-\item Modelling transmission chains in the population to understand effective reproduction number.
-\item Development of risk maps to understand measles importation.
-\end{itemize}
-
-\subsection{Additional risk analyses}
+\section{Additional risk analyses}
 \label{sub:risk_analyses}
 
-In this section we provide a brief description of the work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
+In this section we provide work that we believe will help inform the Ministry of Health regarding the understanding of risk from measles.
 These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
+
+We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand, and descriptive analyses of risk of infection due to previous vaccination history, current rates of immunity within the population.
+In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel.
 
@@
 
-In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
-
-\subsection{Additional risk analyses methods}
-
-For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
-Note this uses all immigration of foreign nationals, coming for whatever purpose.
-We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
-
-<<echo=FALSE>>=
-library(xtable)
-dpdf<-read.csv("mapdata.csv",header=T)
-# output top countries
-colnames(dpdf)[1]<-c("country")
-topimmigration <- dpdf[order(-dpdf$immigration),] 
-immigration<-topimmigration[1:10,c(1,4)]
-topincidence <- dpdf[order(-dpdf$incidence),] 
-incidence<-topincidence[1:10,c(1,2)]
-incidence[,2]<-round(incidence[,2],0)
-topvaccine <- dpdf[order(dpdf$cover),] 
-vaccinecover<-topvaccine[1:10,c(1,3)]
-toprisk <- dpdf[order(-dpdf$risk),] 
-risk<-(toprisk[1:10,c(1,5)])
-risk[,2]<-round(risk[,2],0)
-@
-
-\begin{table}
-\caption{Immigration numbers to New Zealand (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(immigration, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:immigration12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT NZ Travelers here
 %%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total Travelers here
-%%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{Lowest national measles vaccine cover (\%, 2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cover12}
-\end{table}
-
-\begin{table}
-\caption{Measles incidence per million (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(incidence, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:incidence12}
-\end{table}
-
-\begin{table}
-\caption{Risk of measles importation to New Zealand in 2012}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(risk, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:risk12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT NZ risk here
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total risk here
-%%%%%%%%%%%%%%%%%%
-\subsection{Risk analyses methods}
+\subsection{Risk of measles infection in New Zealand analyses methods}
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
@@
 
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15>>=
 par(mfrow=c(1,1))
@@
 
-\begin{table}\tiny
-\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\begin{table}[hbtp]
+\footnotesize
 <<results=tex,echo=FALSE>>=
@@
 library(Hmisc)
-latex(ttyr, file="", table.env=FALSE,rowname=NULL)
-@
+latex(ttyr, file="",table.env=FALSE,rowname=NULL)
+@
+\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
 \label{table:case}
@@
 
-\begin{table}\tiny
+\begin{table}[hbtp]
+\footnotesize
 \caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
@@
 @
+
+\subsection{Measles importation risk}
+
+For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) to measure human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
+We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
+We used simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
+We use the data from 2012 because this year had the most complete WHO measles data and yet was most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+
+%%%%%%%%%%%%%%%%%%
+<<echo=FALSE>>=
+library(xtable)
+totdpdf<-read.csv("mapdata_tot.csv",header=T)
+# output top countries
+colnames(totdpdf)[1]<-c("country")
+toptot <- totdpdf[order(-totdpdf$immigration),] 
+immigration<-toptot[1:30,c(1,4)]
+topincidence <- totdpdf[order(-totdpdf$incidence),] 
+incidence<-topincidence[1:30,c(1,2)]
+incidence[,2]<-round(incidence[,2],0)
+topvaccine <- totdpdf[order(totdpdf$cover),] 
+vaccinecover<-topvaccine[1:30,c(1,3)]
+toprisk <- totdpdf[order(-totdpdf$risk),] 
+risk<-(toprisk[1:30,c(1,5)])
+risk[,2]<-round(risk[,2],0)
+@
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cover12}
+\end{table}
+
+\begin{table}
+\caption{Measles incidence per million (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(incidence, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:incidence12}
+\end{table}
+
+\begin{table}
+\caption{Total New Zealand traveller numbers by country (New Zealand nationals and immigrants, 2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(immigration, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:immigration12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%
+<<echo=FALSE>>=
+library(xtable)
+nzdpdf<-read.csv("mapdata_nzers.csv",header=T)
+# output top countries
+colnames(nzdpdf)[1]<-c("country")
+topnz <- nzdpdf[order(-nzdpdf$immigration),] 
+nz<-topnz[1:30,c(1,4)]
+topnzrisk <- nzdpdf[order(-nzdpdf$risk),] 
+nzrisk<-(topnzrisk[1:30,c(1,5)])
+nzrisk[,2]<-round(nzrisk[,2],0)
+@
+
+\begin{table}
+\caption{New Zealander travel numbers (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(nz, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:nztravel12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+<<echo=FALSE>>=
+library(xtable)
+imdpdf<-read.csv("mapdata_immigration.csv",header=T)
+# output top countries
+colnames(imdpdf)[1]<-c("country")
+topim <- imdpdf[order(-imdpdf$immigration),] 
+im<-topim[1:30,c(1,4)]
+topimrisk <- imdpdf[order(-imdpdf$risk),] 
+imrisk<-(topimrisk[1:30,c(1,5)])
+imrisk[,2]<-round(imrisk[,2],0)
+@
+
+\begin{table}
+\caption{Non-New Zealander travel and immigration numbers (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(im, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:imtravel12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(nzrisk, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:nzrisk12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(imrisk, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:imrisk12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(risk, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:risk12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%
 
@@
 id<-as.data.frame(id)
-dpdf<-read.csv("mapdata.csv",header=T)
+dpdf<-read.csv("mapdata_tot.csv",header=T)
 # output top countries
@@
 #np4
-@
+# output top countries
+dpdf<-read.csv("mapdata_immigration.csv",header=T)
+imnp1<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+imnp4<-ggplot(dpdf, aes(map_id = id)) +
+  geom_map(aes(fill = risk), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
+#np4
+dpdf<-read.csv("mapdata_nzers.csv",header=T)
+nznp1<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+nznp4<-ggplot(dpdf, aes(map_id = id)) +
+  geom_map(aes(fill = risk), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
+#np4
+
+@
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np2)
+@
+\end{center}
+\caption{Measles incidence per million 2012}
+\label{fig:incidence12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(np3)
+@
+\end{center}
+\caption{Measles vaccination cover (\%) 2012 }
+\label{fig:cover12}
+\end{figure}
 
@@
 \end{center}
-\caption{Immigration to New Zealand 2012}
+\caption{Total international travel, 2012 }
+\label{fig:travel12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(nznp1)
+@
+\end{center}
+\caption{New Zealander international travel, 2012 }
+\label{fig:nztravel12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(imnp1)
+@
+\end{center}
+\caption{Non-New Zealander international travel and immigration, 2012 }
 \label{fig:immigration12}
-\end{figure}
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT NZ Travellers here
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total Travelers here
-%%%%%%%%%%%%%%%%%%
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np2)
-@
-\end{center}
-\caption{Measles incidence per million 2012}
-\label{fig:incidence12}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np3)
-@
-\end{center}
-\caption{Measles vaccination cover (\%) 2012 }
-\label{fig:cover12}
 \end{figure}
@@
 \end{center}
-\caption{Risk map from measles incidence and immigration rates}
+\caption{Risk map for measles importation, 2012 }
 \label{fig:risk12}
@@
 
-%%%%%%%%%%%%%%%%%%%
-% INSERT NZ risk here
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%
-% INSERT Total risk here
-%%%%%%%%%%%%%%%%%%
-
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(nznp4)
+@
+\end{center}
+\caption{Risk map for measles importation from New Zealander international travel, 2012 }
+\label{fig:nzrisk12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(imnp4)
+@
+\end{center}
+\caption{Risk map for measles importation from non-New Zealander international travel and immigration, 2012 }
+\label{fig:imrisk12}
+\end{figure}
 %%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
@@
 
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=1\textwidth]{nzers.pdf}
+     \end{center}
+     \caption{Trends in international travel}
+     \label{fig:travel}
+\end{figure}
+
 Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
@@
 
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE>>=
-ttimes<-read.csv("times.csv",header=T)
-lmres<-lm(res~time,data=ttimes)
-#summary(lmres)
-plot(ttimes$res,xaxt="n",type="l",xlab="Date",ylab="")
-axis(1, at=1:71, labels=ttimes$mydate) 
-abline(lmres,col="red",
-       lty=2)
-acf(ttimes$res,lag.max = length(ttimes$res),plot=T,main="")
-@
-\end{center}
-\caption{Trend in total immigrations into New Zealand}
-\label{fig:trendimmigration}
-\end{figure}
-
-%%%%%%%%% ADD NAIVE ## BY DHB HERE
-
-%%%%%%%%%%%%%%%%%%%
-% ADD REPORT # 2 RISK ANAYSIS HERE
-%%%%%%%%%%%%%%%%%%
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per District Health Board}
+     \label{fig:naive_map}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+require(reshape2)  # this is the library that lets one flatten out data
+require(ggplot2)   # plotting library
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+@
+
+\begin{figure}[ht!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+plot(pop,xlab="Age",ylab="Population",cex.lab=2)
+points(naive,pch=16)
+legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
+@
+\end{center}
+\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\label{fig:naive}
+\end{figure}
+
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+%     \label{fig:Northland}
+%\end{figure}
+%
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+%     \label{fig:Waitemata}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+%     \label{fig:Auckland}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+%     \label{fig:Counties_Manukau}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+%     \label{fig:Waikato}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+%     \label{fig:Lakes}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+%     \label{fig:BayofPlenty}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+%     \label{fig:Tairawhiti}
+%\end{figure}
+
+%\begin{figure}
+%    \begin{center}
+%    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+%     \label{fig:Taranaki}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+%     \label{fig:HawkesBay}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+%     \label{fig:Whanganui}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+%     \label{fig:Midcentral}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+%     \label{fig:Hutt}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+%     \label{fig:CapitalandCoast}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+%     \label{fig:Wairarapa}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+%     \label{fig:NelsonMarlborough}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+%     \label{fig:WestCoast}
+%\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
+     \label{fig:Canterbury}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
+     \label{fig:SouthCanterbury}
+\end{figure}
+
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Southern District Health Board}
+     \label{fig:Southern}
+\end{figure}
+
 \subsection{Regression analyses results}
@@
 \end{figure}
+
 The significance of the different predictor variables can be seen in the ANOVA results:
-
-<<label=regmodelanova,include=FALSE>>=
-anova(model3,test="F")
-@
-
+\par
+\vspace{5mm} %5mm vertical space
+<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
+library(xtable)
+resan<-xtable(anova(model3,test="F"))
+print(resan,floating = FALSE)
+@
+\vspace{5mm} %5mm vertical space
 A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
-
-<<label=regmodel,include=FALSE>>=
-summary(model3)
-@
-
-Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
+\vspace{5mm} %5mm vertical space
+<<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
+library(xtable)
+xtb<-xtable(summary(model3))
+print(xtb,floating = FALSE)
+@
+\vspace{5mm} %5mm vertical space
+\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
 Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
@@
 Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+AgeVac<-table(testv$VC,testv$AgeInYears)
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
+legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
+@
+\end{center}
+\caption{Age and vaccination status of cases}
+\label{fig:ageandvac}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
+abline(v=12,col="blue")
+hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+abline(v=60,col="red")
+legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
+legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
+@
+\end{center}
+\caption{Age of vaccination of vaccinated measles cases}
+\label{fig:vaccstat}
+\end{figure}
 
@@
 The results of these regression analyses are also discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit})
-
-\subsection{Risk analysis summary}
-\begin{itemize}
-\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
-\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
-\item Risk of measles infection decreases significantly with age
-\item Pacific people are statistically more at risk of measles infection
-\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
-\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk than European or Maori children
-\end{itemize}
 
@@
 
+\subsection{Risk analysis summary}
+\begin{itemize}
+\item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
+\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
+\item Risk of measles infection decreases significantly with age
+\item Pacific people are statistically more at risk of measles infection
+\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
+\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk than European or Maori children
+\end{itemize}
+
 \section{Modelling measles epidemics}
@@
 Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-@
-
-\begin{figure}[ht!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
-\end{center}
-\caption{New Zealand population by age and estimated numbers of naive people in each age class}
-\label{fig:naive}
-\end{figure}
 
@@
 These formulae were applied at a District Health Board (DHB) level, assuming no mixing between DHBs.
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{cases_by_dhb_2009_2014.pdf}
-     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
-     \label{fig:dhbcases}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{outbreaks_for_R0.pdf}
-     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
-     \label{fig:outbreaks}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ ($R_0$) for the outbreaks each year, as classified in Figure~\ref{fig:outbreaks} (page~\pageref{fig:outbreaks})}
-     \label{fig:r0}
-\end{figure}
 
@@
 This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008--2014 was estimated to be \$2,562 (\$852 in forgone wages and \$1,710 in hospital costs).
-
-%%%%%%%% REVISE
-
-This final figure brings an approximate estimate of \$786,479 for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
+Overall, the cost per case from 2008--2014 was estimated to be \4,327 (\$852 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
+
+This final figure brings an approximate estimate of \$809,149 for 187 cases for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
 
@@
 
-\begin{figure}
+\begin{figure}[width=1\textwidth]
      \centering
@@
 par(cex.axis=2)
-plot(round(0.53*naive$Naive) ,col="grey",bg="red",pch=21,ylab="Number to vaccinate",xlab="Age in Years",
-     main="",cex.lab=2)
-points(naive$Naive[0:21],col="grey",bg="orange",pch=22) # 
-points(naive$Naive[3:30],col="grey",bg="yellow",pch=23) 
-points(jitter(round(0.16*naive$Naive)),col="grey",bg="lightblue",pch=21) 
-points(jitter(naive$Naive[5:13]),col="grey",pch=22,bg="skyblue") # 
-points(naive$Naive[1:5],col="grey",pch=23,bg="navyblue") #
-legend("topright",c("53% naive, all ages","53% naive, all 0-21 yr olds","53% naive, all 3-30 yr olds",
-                    "16% naive, all ages","16% naive, all 5-13 yr olds","16% naive, all 1-5 yr olds"),
-       bty="n",pch=c(rep(21:23,2)),col=rep("grey",6),pt.bg=c("red","orange","yellow","lightblue","skyblue","navyblue"),cex=2)
-@
-     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave}}
+plot(naive$Naive ,col="black",bg="black",pch=1,ylab="Number to vaccinate",xlab="Age in Years",
+     main="",cex.lab=2,cex=3)
+points(round(0.28*naive$Naive),col="grey",bg="red",pch=21,cex=2)
+points(x=3:18,y=naive$Naive[3:18],col="grey",bg="orange",pch=23,cex=2) # 
+legend("topright",c("Naive","28% naive, all ages","28% naive, all 2-17 yr olds"),
+       bty="n",pch=c(1,21,23),col=rep("grey",3),pt.bg=c("black","red","orange"),cex=2)
+@
+     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave})}
      \label{fig:numvac}
@@
 set.seed(1) # all infected
-res<-jm.epid(epid.nb=1000,GT=genTime,R0=1,epid.length=365*5,popn=4.2*1e6 * 0.11
-             ,family="poisson",peak.value=4.2*1e6 * 0.11)
+res<-jm.epid(epid.nb=1000,GT=genTime,R0=0.99,epid.length=365*5,popn=331385
+             ,family="poisson",peak.value=331385)
 sizes<-colSums(res)
@@
 par(cex.axis=2)
-hist(sizes,xlim=c(0,200),col="grey",breaks=3000,xlab="Outbreak size",main="",cex.lab=2)
+hist(sizes,xlim=c(0,200),col="grey",breaks=1000,xlab="Outbreak size",main="",cex.lab=2)
 #summary(sizes)
@@
 
-\begin{sidewaystable}[htbp]  
-  \centering  
-  \begin{raggedright}
-  \caption{Cost benefit analyses using simulated epidemic sizes}  
-  \label{tab:cost_sim}
-   \begin{threeparttable}[b] \tiny
-    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
-    \toprule
-
-Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \dagger&	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
-\midrule
-2009-2014 &	0.92-1.19	& 0.92	& 0	    & na	  & 0	      & 130	    & 22  	& 148551	  & 130519	  & 13	& 130	  & 22	& 148551	& 0	      & 0	        & 0	        & na \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	& 20	  & 1600000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 126359360	& 79.97 \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	& 50	  & 4000000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 123959360	& 31.99 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 336060587	& 81.72 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 329815997	& 32.69 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 377382119	& 75.71 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 369805289	& 30.28 \\
-
-\end{raggedright}
-\bottomrule  
-    \end{tabular}%  
-    \begin{tablenotes}\footnotesize  
-        \item $\dagger$ {1000 simulations}
-        \item \sym{*} Proportion of cases hospitalised 0.17
-        \item \sym{**} Wage losses per case = \$852 and cost per hospitalised case = \$1710
-        \item \sym{***} Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
-    \end{threeparttable}  
+<<echo=FALSE>>=
+costtable20<-read.csv("cost_benefit_20.csv",header=T)
+@
+
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 20 dollars per vaccine}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable20, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cost20}
 \end{sidewaystable}
 
-\normalsize
-
-\begin{sidewaystable}[htbp]  
-  \centering  
-  \begin{raggedright}
-  \caption{Cost benefit analyses using historic epidemic sizes since 1997}  
-  \label{tab:cost_ave}
-   \begin{threeparttable}[b] \tiny
-    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
-    \toprule  
-Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \ddagger &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
-\midrule
-
-2009-2014	& 0.92-1.19	& 0.92	& 0	      & na	& 0	      & 130	  & 22	& 148551	  & 130519	& 13	& 130	  & 22	& 148551	& 0	  & 0	      & 0	        & na \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 20	& 1600000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -847009	  & 0.47 \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 50	& 4000000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -3247009	& 0.19 \\ 
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -3118912	& 0.25 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -9363502	& 0.10 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -4007072	& 0.21 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -11583902	& 0.08 \\
-\end{raggedright}
-\bottomrule  
-    \end{tabular}%  
-    \begin{tablenotes}\footnotesize  
-        \item $\ddagger$ Based on average number of cases per year in New Zealand from 1997 onward
-        \item $\dagger$ {1000 simulations}
-        \item \sym{*} Proportion of cases hospitalised 0.17
-        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
-        \item \sym{***} Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
-    \end{threeparttable}  
+<<echo=FALSE>>=
+costtable50<-read.csv("cost_benefit_50.csv",header=T)
+@
+
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 50 dollars per vaccine}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable50, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cost50}
 \end{sidewaystable}
-
-\normalsize
 
@@
 It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-AgeVac<-table(testv$VC,testv$AgeInYears)
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-@
-\end{center}
-\caption{Age and vaccination status of cases}
-\label{fig:ageandvac}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-@
-\end{center}
-\caption{Age of vaccination of vaccinated measles cases}
-\label{fig:vaccstat}
-\end{figure}
 

### 2014-10-29 draftfinalreport.tex
revisions

--- 2014-10-28 draftfinalreport.tex
@@
 
-In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to use quantitative methods to evaluate the risk of measles importation.
-
-\subsection{Additional risk analyses methods}
-
-For our initial analyses, we use arrivals data from New Zealand immigration (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) as a proxy for human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
-Note this uses all immigration of foreign nationals, coming for whatever purpose.
-We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
-
 %%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{Total travel numbers (2012)}
-%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
-\hline
-Australia&$1799655$\tabularnewline
-United Kingdom&$ 401737$\tabularnewline
-China&$ 322076$\tabularnewline
-United States&$ 316058$\tabularnewline
-Fiji&$ 151443$\tabularnewline
-India&$ 107618$\tabularnewline
-Japan&$ 106716$\tabularnewline
-Germany&$  96308$\tabularnewline
-Korea, Republic of&$  87419$\tabularnewline
-France&$  85948$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:immigration12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{Lowest national measles vaccine cover (\%, 2012)}
-%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
-\hline
-Equatorial Guinea&$34$\tabularnewline
-Somalia&$49$\tabularnewline
-Lesotho&$60$\tabularnewline
-Central African Republic&$65$\tabularnewline
-Papua New Guinea&$67$\tabularnewline
-Chad&$69$\tabularnewline
-Haiti&$69$\tabularnewline
-South Sudan&$70$\tabularnewline
-Gabon&$71$\tabularnewline
-Yemen&$71$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:cover12}
-\end{table}
-
-\begin{table}
-\caption{Measles incidence per million (2012)}
-%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
-\hline
-Equatorial Guinea&$1617$\tabularnewline
-Nauru&$1100$\tabularnewline
-Democratic Republic of the Congo&$1096$\tabularnewline
-Somalia&$ 979$\tabularnewline
-Djibouti&$ 824$\tabularnewline
-Sudan&$ 786$\tabularnewline
-Burkina Faso&$ 447$\tabularnewline
-Romania&$ 342$\tabularnewline
-Ukraine&$ 280$\tabularnewline
-Sudan&$ 229$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:incidence12}
-\end{table}
-
-\begin{table}
-\caption{Risk of measles importation to New Zealand in 2012}
-%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
-\hline
-Australia&$15537152$\tabularnewline
-United Kingdom&$13386328$\tabularnewline
-Thailand&$ 4774688$\tabularnewline
-Malaysia&$ 4495338$\tabularnewline
-Indonesia&$ 2216871$\tabularnewline
-India&$ 1620172$\tabularnewline
-China&$ 1432659$\tabularnewline
-Ukraine&$  765096$\tabularnewline
-Ireland&$  734128$\tabularnewline
-Romania&$  655107$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:risk12}
-\end{table}
-
-\begin{table}
-\caption{Immigration numbers to New Zealand (2012)}
-%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
-\hline
-Australia&$809775$\tabularnewline
-United Kingdom&$306177$\tabularnewline
-China&$256036$\tabularnewline
-United States&$194438$\tabularnewline
-Japan&$ 86676$\tabularnewline
-Germany&$ 83608$\tabularnewline
-Korea, Republic of&$ 73459$\tabularnewline
-France&$ 71448$\tabularnewline
-India&$ 69038$\tabularnewline
-Canada&$ 54981$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:immigration12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{Lowest national measles vaccine cover (\%, 2012)}
-%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
-\hline
-Equatorial Guinea&$34$\tabularnewline
-Somalia&$49$\tabularnewline
-Lesotho&$60$\tabularnewline
-Central African Republic&$65$\tabularnewline
-Papua New Guinea&$67$\tabularnewline
-Chad&$69$\tabularnewline
-Haiti&$69$\tabularnewline
-South Sudan&$70$\tabularnewline
-Gabon&$71$\tabularnewline
-Yemen&$71$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:cover12}
-\end{table}
-
-\begin{table}
-\caption{Measles incidence per million (2012)}
-%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
-\hline
-Equatorial Guinea&$1617$\tabularnewline
-Nauru&$1100$\tabularnewline
-Democratic Republic of the Congo&$1096$\tabularnewline
-Somalia&$ 979$\tabularnewline
-Djibouti&$ 824$\tabularnewline
-Sudan&$ 786$\tabularnewline
-Burkina Faso&$ 447$\tabularnewline
-Romania&$ 342$\tabularnewline
-Ukraine&$ 280$\tabularnewline
-Sudan&$ 229$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:incidence12}
-\end{table}
-
-\begin{table}
-\caption{Risk of measles importation to New Zealand in 2012}
-%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
-\hline
-United Kingdom&$10202161$\tabularnewline
-Australia&$ 6991116$\tabularnewline
-Malaysia&$ 3226580$\tabularnewline
-Thailand&$ 1576414$\tabularnewline
-China&$ 1138900$\tabularnewline
-India&$ 1039356$\tabularnewline
-Indonesia&$  984022$\tabularnewline
-Ukraine&$  664315$\tabularnewline
-Ireland&$  585413$\tabularnewline
-Romania&$  490731$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:risk12}
-\end{table}
-%%%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{New Zealander travel numbers (2012)}
-%latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
-\hline
-Australia&$989880$\tabularnewline
-United States&$121620$\tabularnewline
-Fiji&$104720$\tabularnewline
-United Kingdom&$ 95560$\tabularnewline
-Cook Islands&$ 71960$\tabularnewline
-China&$ 66040$\tabularnewline
-Samoa&$ 46020$\tabularnewline
-Thailand&$ 41100$\tabularnewline
-India&$ 38580$\tabularnewline
-Canada&$ 20400$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:nztravel12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
-%latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
-\hline
-Australia&$8546036$\tabularnewline
-Thailand&$3198274$\tabularnewline
-United Kingdom&$3184166$\tabularnewline
-Malaysia&$1268758$\tabularnewline
-Indonesia&$1232849$\tabularnewline
-India&$ 580816$\tabularnewline
-China&$ 293759$\tabularnewline
-Samoa&$ 243492$\tabularnewline
-Philippines&$ 241740$\tabularnewline
-Nepal&$ 188450$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:nzrisk12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%
-\subsection{Risk analyses methods}
+\subsection{Risk of measles infection in New Zealand analyses methods}
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-015}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-002}
 \end{center}
@@
 
-\begin{table}\tiny
-\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\begin{table}[hbtp]
+\footnotesize
 %latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:case}
+\end{tabular}\end{center}\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\label{table:case}
 \end{table}
@@
 \begin{center}
-\includegraphics{draftfinalreport-023}
+\includegraphics{draftfinalreport-010}
 \end{center}
@@
 
+\subsection{Measles importation risk}
+
+For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) to measure human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
+We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
+We used simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
+We use the data from 2012 because this year had the most complete WHO measles data and yet was most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+
+%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (\%, 2012)}
+%latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{cover}\tabularnewline
+\hline
+Equatorial Guinea&$34$\tabularnewline
+Somalia&$49$\tabularnewline
+Lesotho&$60$\tabularnewline
+Central African Republic&$65$\tabularnewline
+Papua New Guinea&$67$\tabularnewline
+Chad&$69$\tabularnewline
+Haiti&$69$\tabularnewline
+South Sudan&$70$\tabularnewline
+Gabon&$71$\tabularnewline
+Yemen&$71$\tabularnewline
+Benin&$72$\tabularnewline
+Lao People's Democratic Republic&$72$\tabularnewline
+Togo&$72$\tabularnewline
+Suriname&$73$\tabularnewline
+Timor-Leste&$73$\tabularnewline
+Paraguay&$74$\tabularnewline
+Mauritania&$75$\tabularnewline
+Namibia&$76$\tabularnewline
+Eritrea&$77$\tabularnewline
+Marshall Islands&$78$\tabularnewline
+Nigeria&$78$\tabularnewline
+Syrian Arab Republic&$78$\tabularnewline
+Ukraine&$79$\tabularnewline
+Congo&$80$\tabularnewline
+Ethiopia&$80$\tabularnewline
+Liberia&$80$\tabularnewline
+Afghanistan&$81$\tabularnewline
+Cameroon&$82$\tabularnewline
+Mozambique&$82$\tabularnewline
+Senegal&$82$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cover12}
+\end{table}
+
+\begin{table}
+\caption{Measles incidence per million (2012)}
+%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
+\hline
+Equatorial Guinea&$1617$\tabularnewline
+Nauru&$1100$\tabularnewline
+Democratic Republic of the Congo&$1096$\tabularnewline
+Somalia&$ 979$\tabularnewline
+Djibouti&$ 824$\tabularnewline
+Sudan&$ 786$\tabularnewline
+Burkina Faso&$ 447$\tabularnewline
+Romania&$ 342$\tabularnewline
+Ukraine&$ 280$\tabularnewline
+Sudan&$ 229$\tabularnewline
+Angola&$ 214$\tabularnewline
+Monaco&$ 132$\tabularnewline
+Nepal&$ 122$\tabularnewline
+Sierra Leone&$ 113$\tabularnewline
+Afghanistan&$  93$\tabularnewline
+Yemen&$  91$\tabularnewline
+Lesotho&$  87$\tabularnewline
+Qatar&$  78$\tabularnewline
+Thailand&$  78$\tabularnewline
+Malaysia&$  64$\tabularnewline
+Zambia&$  64$\tabularnewline
+Ghana&$  64$\tabularnewline
+Indonesia&$  63$\tabularnewline
+Congo&$  60$\tabularnewline
+Uganda&$  56$\tabularnewline
+Libyan Arab Jamahiriya&$  52$\tabularnewline
+Ethiopia&$  47$\tabularnewline
+Pakistan&$  45$\tabularnewline
+Myanmar&$  41$\tabularnewline
+Nigeria&$  38$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:incidence12}
+\end{table}
+
+\begin{table}
+\caption{Total New Zealand traveller numbers by country (New Zealand nationals and immigrants, 2012)}
+%latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$1799655$\tabularnewline
+United Kingdom&$ 401737$\tabularnewline
+China&$ 322076$\tabularnewline
+United States&$ 316058$\tabularnewline
+Fiji&$ 151443$\tabularnewline
+India&$ 107618$\tabularnewline
+Japan&$ 106716$\tabularnewline
+Germany&$  96308$\tabularnewline
+Korea, Republic of&$  87419$\tabularnewline
+France&$  85948$\tabularnewline
+Canada&$  75381$\tabularnewline
+Samoa&$  70567$\tabularnewline
+Malaysia&$  70366$\tabularnewline
+Thailand&$  61358$\tabularnewline
+South Africa&$  45980$\tabularnewline
+Tonga&$  44477$\tabularnewline
+Singapore&$  42580$\tabularnewline
+Philippines&$  39747$\tabularnewline
+Netherlands&$  39151$\tabularnewline
+Hong Kong&$  37323$\tabularnewline
+Indonesia&$  35352$\tabularnewline
+Taiwan&$  34275$\tabularnewline
+Ireland&$  31396$\tabularnewline
+Italy&$  25208$\tabularnewline
+Viet Nam&$  18574$\tabularnewline
+Switzerland&$  18431$\tabularnewline
+Brazil&$  17878$\tabularnewline
+Vanuatu&$  16261$\tabularnewline
+Spain&$  15604$\tabularnewline
+Sweden&$  15479$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:immigration12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{New Zealander travel numbers (2012)}
+%latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$989880$\tabularnewline
+United States&$121620$\tabularnewline
+Fiji&$104720$\tabularnewline
+United Kingdom&$ 95560$\tabularnewline
+Cook Islands&$ 71960$\tabularnewline
+China&$ 66040$\tabularnewline
+Samoa&$ 46020$\tabularnewline
+Thailand&$ 41100$\tabularnewline
+India&$ 38580$\tabularnewline
+Canada&$ 20400$\tabularnewline
+Japan&$ 20040$\tabularnewline
+Malaysia&$ 19860$\tabularnewline
+Indonesia&$ 19660$\tabularnewline
+Hong Kong&$ 18220$\tabularnewline
+Tonga&$ 17760$\tabularnewline
+Singapore&$ 17120$\tabularnewline
+South Africa&$ 15380$\tabularnewline
+Philippines&$ 15220$\tabularnewline
+France&$ 14500$\tabularnewline
+Korea, Republic of&$ 13960$\tabularnewline
+Viet Nam&$ 12920$\tabularnewline
+Germany&$ 12700$\tabularnewline
+Vanuatu&$ 12520$\tabularnewline
+Italy&$ 11820$\tabularnewline
+Taiwan&$ 10460$\tabularnewline
+New Caledonia&$  7340$\tabularnewline
+Ireland&$  6360$\tabularnewline
+French Polynesia&$  6360$\tabularnewline
+Papua New Guinea&$  6140$\tabularnewline
+Netherlands&$  5720$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:nztravel12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Non-New Zealander travel and immigration numbers (2012)}
+%latex.default(im, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$809775$\tabularnewline
+United Kingdom&$306177$\tabularnewline
+China&$256036$\tabularnewline
+United States&$194438$\tabularnewline
+Japan&$ 86676$\tabularnewline
+Germany&$ 83608$\tabularnewline
+Korea, Republic of&$ 73459$\tabularnewline
+France&$ 71448$\tabularnewline
+India&$ 69038$\tabularnewline
+Canada&$ 54981$\tabularnewline
+Malaysia&$ 50506$\tabularnewline
+Fiji&$ 46723$\tabularnewline
+Netherlands&$ 33431$\tabularnewline
+South Africa&$ 30600$\tabularnewline
+Tonga&$ 26717$\tabularnewline
+Singapore&$ 25460$\tabularnewline
+Ireland&$ 25036$\tabularnewline
+Samoa&$ 24547$\tabularnewline
+Philippines&$ 24527$\tabularnewline
+Taiwan&$ 23815$\tabularnewline
+Thailand&$ 20258$\tabularnewline
+Hong Kong&$ 19103$\tabularnewline
+Indonesia&$ 15692$\tabularnewline
+Switzerland&$ 14851$\tabularnewline
+Brazil&$ 14778$\tabularnewline
+Sweden&$ 13719$\tabularnewline
+Italy&$ 13388$\tabularnewline
+Spain&$ 10104$\tabularnewline
+Denmark&$ 10056$\tabularnewline
+Russia&$  8103$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:imtravel12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
+%latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+Australia&$8546036$\tabularnewline
+Thailand&$3198274$\tabularnewline
+United Kingdom&$3184166$\tabularnewline
+Malaysia&$1268758$\tabularnewline
+Indonesia&$1232849$\tabularnewline
+India&$ 580816$\tabularnewline
+China&$ 293759$\tabularnewline
+Samoa&$ 243492$\tabularnewline
+Philippines&$ 241740$\tabularnewline
+Nepal&$ 188450$\tabularnewline
+Romania&$ 164376$\tabularnewline
+Ireland&$ 148715$\tabularnewline
+Spain&$ 141632$\tabularnewline
+Singapore&$ 135591$\tabularnewline
+Ukraine&$ 100781$\tabularnewline
+Viet Nam&$  82248$\tabularnewline
+Sudan&$  78640$\tabularnewline
+Afghanistan&$  74756$\tabularnewline
+Italy&$  72995$\tabularnewline
+United Arab Emirates&$  60508$\tabularnewline
+Pakistan&$  58435$\tabularnewline
+Nauru&$  44000$\tabularnewline
+Democratic Republic of the Congo&$  43850$\tabularnewline
+Qatar&$  43686$\tabularnewline
+Israel&$  40301$\tabularnewline
+Japan&$  35977$\tabularnewline
+Russia&$  35334$\tabularnewline
+Angola&$  34258$\tabularnewline
+Switzerland&$  27308$\tabularnewline
+Saudi Arabia&$  25775$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:nzrisk12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012}
+%latex.default(imrisk, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+United Kingdom&$10202161$\tabularnewline
+Australia&$ 6991116$\tabularnewline
+Malaysia&$ 3226580$\tabularnewline
+Thailand&$ 1576414$\tabularnewline
+China&$ 1138900$\tabularnewline
+India&$ 1039356$\tabularnewline
+Indonesia&$  984022$\tabularnewline
+Ukraine&$  664315$\tabularnewline
+Ireland&$  585413$\tabularnewline
+Romania&$  490731$\tabularnewline
+Philippines&$  389563$\tabularnewline
+Spain&$  260191$\tabularnewline
+Singapore&$  201644$\tabularnewline
+Germany&$  167620$\tabularnewline
+Japan&$  155607$\tabularnewline
+Nepal&$  136076$\tabularnewline
+Israel&$  133876$\tabularnewline
+Samoa&$  129878$\tabularnewline
+Russia&$  120298$\tabularnewline
+Switzerland&$  113281$\tabularnewline
+Pakistan&$   90079$\tabularnewline
+Somalia&$   85191$\tabularnewline
+Italy&$   82679$\tabularnewline
+Nauru&$   82500$\tabularnewline
+Sudan&$   62912$\tabularnewline
+Saudi Arabia&$   56809$\tabularnewline
+Belgium&$   53337$\tabularnewline
+Sweden&$   43273$\tabularnewline
+Afghanistan&$   37098$\tabularnewline
+Viet Nam&$   35993$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:imrisk12}
+\end{table}
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers}
+%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+Australia&$15537152$\tabularnewline
+United Kingdom&$13386328$\tabularnewline
+Thailand&$ 4774688$\tabularnewline
+Malaysia&$ 4495338$\tabularnewline
+Indonesia&$ 2216871$\tabularnewline
+India&$ 1620172$\tabularnewline
+China&$ 1432659$\tabularnewline
+Ukraine&$  765096$\tabularnewline
+Ireland&$  734128$\tabularnewline
+Romania&$  655107$\tabularnewline
+Philippines&$  631303$\tabularnewline
+Spain&$  401823$\tabularnewline
+Samoa&$  373370$\tabularnewline
+Singapore&$  337236$\tabularnewline
+Nepal&$  324526$\tabularnewline
+Germany&$  193081$\tabularnewline
+Japan&$  191585$\tabularnewline
+Israel&$  174177$\tabularnewline
+Italy&$  155674$\tabularnewline
+Russia&$  155632$\tabularnewline
+Pakistan&$  148514$\tabularnewline
+Sudan&$  141552$\tabularnewline
+Switzerland&$  140589$\tabularnewline
+Nauru&$  126500$\tabularnewline
+Viet Nam&$  118241$\tabularnewline
+Afghanistan&$  111854$\tabularnewline
+Somalia&$  104775$\tabularnewline
+Saudi Arabia&$   82584$\tabularnewline
+United Arab Emirates&$   67606$\tabularnewline
+Belgium&$   63784$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:risk12}
+\end{table}
+
+%%%%%%%%%%%%%%%%%%
+
 \subsection{Additional risk analyses results}
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{np1.pdf}
-     \caption{Non-New Zealander travel}
-     \label{fig:immigration12}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{nznp1.pdf}
-     \caption{New Zealand travellers 2012}
-     \label{fig:nztravellers}
-\end{figure}
-
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{totnp1.pdf}
-     \caption{Total travellers 2012}
-     \label{fig:tottravellers}
-\end{figure}
-
 \begin{figure}[h!]
 \begin{center}
-\includegraphics{draftfinalreport-028}
+\includegraphics{draftfinalreport-026}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-029}
+\includegraphics{draftfinalreport-027}
 \end{center}
@@
 \begin{figure}[h!]
- \centering
- \includegraphics[width=0.8\textwidth]{np4.pdf}
-\caption{Risk map from measles incidence and immigration rates}
+\begin{center}
+\includegraphics{draftfinalreport-028}
+\end{center}
+\caption{Total international travel, 2012 }
+\label{fig:travel12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-029}
+\end{center}
+\caption{New Zealander international travel, 2012 }
+\label{fig:nztravel12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-030}
+\end{center}
+\caption{Non-New Zealander international travel and immigration, 2012 }
+\label{fig:immigration12}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-031}
+\end{center}
+\caption{Risk map for measles importation, 2012 }
 \label{fig:risk12}
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{nznp4.pdf}
-     \caption{Risk map from measles incidence and New Zealander travel rates}
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-032}
+\end{center}
+\caption{Risk map for measles importation from New Zealander international travel, 2012 }
 \label{fig:nzrisk12}
@@
 
-\begin{figure}
-     \centering
-     \includegraphics[width=0.8\textwidth]{totnp4.pdf}
-     \caption{Risk map from measles incidence and total travel rates}
-     \label{fig:totrisk12}
-\end{figure}
-
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-033}
+\end{center}
+\caption{Risk map for measles importation from non-New Zealander international travel and immigration, 2012 }
+\label{fig:imrisk12}
+\end{figure}
 %%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
@@
 \begin{center}
-\includegraphics{draftfinalreport-030}
+\includegraphics{draftfinalreport-034}
 \end{center}
@@
      \end{center}
-     \caption{Trends in travel}
+     \caption{Trends in international travel}
      \label{fig:travel}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-031}
-\end{center}
-\caption{Trend in total immigrations into New Zealand}
-\label{fig:trendimmigration}
-\end{figure}
-
 \begin{figure}
      \begin{center}
-     \includegraphics[width=0.6\textwidth]{naive_map.pdf}
+     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
      \end{center}
@@
 
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Northland District Health Board}
-     \label{fig:Northland}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
-     \label{fig:Waitemata}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
-     \label{fig:Auckland}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
-     \label{fig:Counties_Manukau}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
-     \label{fig:Waikato}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
-     \label{fig:Lakes}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
-     \label{fig:BayofPlenty}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
-     \label{fig:Tairawhiti}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb9.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
-     \label{fig:Taranaki}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
-     \label{fig:HawkesBay}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
-     \label{fig:Whanganui}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
-     \label{fig:Midcentral}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
-     \label{fig:Hutt}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
-     \label{fig:CapitalandCoast}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
-     \label{fig:Wairarapa}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
-     \label{fig:NelsonMarlborough}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
-     \label{fig:WestCoast}
-\end{figure}
+\begin{figure}[ht!]
+\begin{center}
+\includegraphics{draftfinalreport-036}
+\end{center}
+\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\label{fig:naive}
+\end{figure}
+
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+%     \label{fig:Northland}
+%\end{figure}
+%
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+%     \label{fig:Waitemata}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+%     \label{fig:Auckland}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+%     \label{fig:Counties_Manukau}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+%     \label{fig:Waikato}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+%     \label{fig:Lakes}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+%     \label{fig:BayofPlenty}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+%     \label{fig:Tairawhiti}
+%\end{figure}
+
+%\begin{figure}
+%    \begin{center}
+%    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+%     \label{fig:Taranaki}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+%     \label{fig:HawkesBay}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+%     \label{fig:Whanganui}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+%     \label{fig:Midcentral}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+%     \label{fig:Hutt}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+%     \label{fig:CapitalandCoast}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+%     \label{fig:Wairarapa}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+%     \label{fig:NelsonMarlborough}
+%\end{figure}
+%
+%\begin{figure}
+%     \begin{center}
+%     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
+%     \end{center}
+%     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+%     \label{fig:WestCoast}
+%\end{figure}
 
@@
 \begin{center}
-\includegraphics{draftfinalreport-032}
+\includegraphics{draftfinalreport-037}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-033}
+\includegraphics{draftfinalreport-038}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-035}
+\includegraphics{draftfinalreport-040}
 \end{center}
@@
 \end{figure}
+
 The significance of the different predictor variables can be seen in the ANOVA results:
-
-\begin{Schunk}
-\begin{Sinput}
-> anova(model3,test="F")
-\end{Sinput}
-\begin{Soutput}
-Analysis of Deviance Table
-
-Model: quasipoisson, link: log
-
-Response: cases
-
-Terms added sequentially (first to last)
-
-              Df Deviance Resid. Df Resid. Dev        F   Pr(>F)    
-NULL                             39    1453.62                      
-Age            4  1304.20        35     149.43 183.5795 6.73e-15 ***
-Ethnicity      3    20.00        32     129.43   3.7529 0.028500 *  
-NZDep          1    10.70        31     118.74   6.0236 0.023932 *  
-Age:Ethnicity 12    81.91        19      36.83   3.8430 0.004459 ** 
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-\end{Soutput}
-\end{Schunk}
-
+\par
+\vspace{5mm} %5mm vertical space
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Wed Oct 29 19:17:39 2014
+\begin{tabular}{lrrrrrr}
+  \hline
+ & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
+  \hline
+NULL &  &  & 39 & 1453.62 &  &  \\ 
+  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
+  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
+  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
+  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
+   \hline
+\end{tabular}\vspace{5mm} %5mm vertical space
 A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
-
-\begin{Schunk}
-\begin{Sinput}
-> summary(model3)
-\end{Sinput}
-\begin{Soutput}
-Call:
-glm(formula = cases ~ Age + Ethnicity + NZDep + Age:Ethnicity + 
-    offset(log(Popn)), family = "quasipoisson", data = tpsub)
-
-Deviance Residuals: 
-     Min        1Q    Median        3Q       Max  
--2.76316  -0.52409   0.04736   0.49883   1.92918
-
-Coefficients:
-                          Estimate Std. Error t value Pr(>|t|)    
-(Intercept)               -6.46738    0.11483 -56.324  < 2e-16 ***
-Age3-5                    -0.91290    0.20539  -4.445 0.000278 ***
-Age6-17                   -0.76175    0.13427  -5.673 1.81e-05 ***
-Age18-24                  -1.50501    0.19366  -7.772 2.57e-07 ***
-Age25+                    -2.95073    0.16079 -18.352 1.51e-13 ***
-EthnicityAsian             0.05304    0.32498   0.163 0.872072    
-EthnicityMaori             0.21245    0.19935   1.066 0.299919    
-EthnicityPacific           1.16032    0.20429   5.680 1.78e-05 ***
-NZDep6-10                 -0.21192    0.08548  -2.479 0.022713 *  
-Age3-5:EthnicityAsian     -2.03154    1.38265  -1.469 0.158112    
-Age6-17:EthnicityAsian    -1.00742    0.47171  -2.136 0.045940 *  
-Age18-24:EthnicityAsian   -0.83115    0.59409  -1.399 0.177917    
-Age25+:EthnicityAsian      0.42108    0.44334   0.950 0.354143    
-Age3-5:EthnicityMaori     -0.38639    0.40959  -0.943 0.357344    
-Age6-17:EthnicityMaori    -0.08552    0.24685  -0.346 0.732812    
-Age18-24:EthnicityMaori   -0.58278    0.44835  -1.300 0.209209    
-Age25+:EthnicityMaori     -1.04821    0.52418  -2.000 0.060036 .  
-Age3-5:EthnicityPacific   -2.13163    0.81388  -2.619 0.016882 *  
-Age6-17:EthnicityPacific  -1.45411    0.33466  -4.345 0.000349 ***
-Age18-24:EthnicityPacific -0.98266    0.51289  -1.916 0.070546 .  
-Age25+:EthnicityPacific   -0.61270    0.43665  -1.403 0.176692    
----
-Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-
-(Dispersion parameter for quasipoisson family taken to be 1.776063)
-
-    Null deviance: 1453.62  on 39  degrees of freedom
-Residual deviance:   36.83  on 19  degrees of freedom
-AIC: NA
-
-Number of Fisher Scoring iterations: 5
-\end{Soutput}
-\end{Schunk}
-
-Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
+\vspace{5mm} %5mm vertical space
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Wed Oct 29 19:17:39 2014
+\begin{tabular}{rrrrr}
+  \hline
+ & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
+  \hline
+(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
+  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
+  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
+  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
+  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
+  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
+  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
+  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
+  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
+  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
+  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
+  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
+  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
+  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
+  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
+  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
+  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
+  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
+  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
+  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
+  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
+   \hline
+\end{tabular}\vspace{5mm} %5mm vertical space
+\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
 Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
@@
 Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-043}
+\end{center}
+\caption{Age and vaccination status of cases}
+\label{fig:ageandvac}
+\end{figure}
+
+\begin{figure}[h!]
+\begin{center}
+\includegraphics{draftfinalreport-044}
+\end{center}
+\caption{Age of vaccination of vaccinated measles cases}
+\label{fig:vaccstat}
+\end{figure}
 
@@
 Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
-
-\begin{figure}[ht!]
-\begin{center}
-\includegraphics{draftfinalreport-039}
-\end{center}
-\caption{New Zealand population by age and estimated numbers of naive people in each age class}
-\label{fig:naive}
-\end{figure}
 
@@
 \begin{center}
-\includegraphics{draftfinalreport-047}
+\includegraphics{draftfinalreport-052}
 \end{center}
@@
 
-\begin{figure}
+\begin{figure}[width=1\textwidth]
      \centering
-\includegraphics{draftfinalreport-053}
-     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave}}
+\includegraphics{draftfinalreport-058}
+     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave})}
      \label{fig:numvac}
@@
      \centering
-\includegraphics{draftfinalreport-055}
+\includegraphics{draftfinalreport-060}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
@@
      \centering
-\includegraphics{draftfinalreport-056}
+\includegraphics{draftfinalreport-061}
      \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
@@
 \subsection{Benefit--cost analyses results}
-
-\begin{sidewaystable}[htbp]  
-  \centering  
-  \begin{raggedright}
-  \caption{Cost benefit analyses using simulated epidemic sizes}  
-  \label{tab:cost_sim}
-   \begin{threeparttable}[b] \tiny
-    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
-    \toprule
-
-Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \dagger&	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
-\midrule
-2009-2014 &	0.92-1.19	& 0.92	& 0	    & na	  & 0	      & 130	    & 22  	& 148551	  & 130519	  & 13	& 130	  & 22	& 148551	& 0	      & 0	        & 0	        & na \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	& 20	  & 1600000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 126359360	& 79.97 \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	& 50	  & 4000000	& 128901	& 21913	& 147295173	& 129415143	& 145	& 1450	& 247	& 1656915	& 127451	& 127959360	& 123959360	& 31.99 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 336060587	& 81.72 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 340032	& 57805	& 388554566	& 341388274	& 116	& 1160	& 197	& 1325532	& 338872	& 340223647	& 329815997	& 32.69 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 377382119	& 75.71 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 382074	& 64953	& 436595960	& 383597966	& 116	& 1160	& 197	& 1325532	& 380914	& 382433339	& 369805289	& 30.28 \\
-
-\end{raggedright}
-\bottomrule  
-    \end{tabular}%  
-    \begin{tablenotes}\footnotesize  
-        \item $\dagger$ {1000 simulations}
-        \item \sym{*} Proportion of cases hospitalised 0.17
-        \item \sym{**} Wage losses per case = \$852 and cost per hospitalised case = \$1710
-        \item \sym{***} Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
-    \end{threeparttable}  
-\end{sidewaystable}
-
-\normalsize
-
-\begin{sidewaystable}[htbp]  
-  \centering  
-  \begin{raggedright}
-  \caption{Cost benefit analyses using historic epidemic sizes since 1997}  
-  \label{tab:cost_ave}
-   \begin{threeparttable}[b] \tiny
-    \begin{tabular}{p{0.8cm}p{0.8cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.8cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}p{0.9cm}}  
-    \toprule  
-Years $R_v$ estimated from & $R_v$ range &  $R_v$  for simulations &	Additional vaccines required to reduce $R_v$ to 1 &	Costs per vaccine (\$) &	Present value of discounted vaccination costs (\$) &	Mean total cases in over 10 years without additional vaccination \ddagger &	Total hospitalised cases\sym{*}	& Total undiscounted case costs without additional vaccination (\$)\sym{**} &	Present value of discounted case costs of baseline program (\$)(Equation~\ref{eq:pc}) &	Mean epidemic size in immune pop over 10 years with $R_v$ = 1 \dagger &	Total cases over 10 years following additional vaccination\sym{***} &	Total hospitalised cases following additional vaccination &	Total costs for cases assuming supplemental vaccination (\$) &	Total cases reduce by vaccination alternative vs baseline &	Present value of discounted benefits from reducing cases with supplemental vaccination program	& Discounted net benefit of supplemental vaccination program &	Benefit--cost ratio\\
-\midrule
-
-2009-2014	& 0.92-1.19	& 0.92	& 0	      & na	& 0	      & 130	  & 22	& 148551	  & 130519	& 13	& 130	  & 22	& 148551	& 0	  & 0	      & 0	        & na \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 20	& 1600000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -847009	  & 0.47 \\
-2009-2014	& 0.92-1.19	& 1.19	& 80000	  & 50	& 4000000	& 2200	& 374	& 2513940	  & 2208775	& 145	& 1450	& 247	& 1656915	& 750	& 752991	& -3247009	& 0.19 \\ 
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 20	& 4163060	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -3118912	& 0.25 \\
-2013-2014	& 1.82-2.13	& 1.82	& 208153	& 50	& 10407650& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -9363502	& 0.10 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 20	& 5051220	& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -4007072	& 0.21 \\
-2013-2014	& 1.82-2.13	& 2.13	& 252561	& 50	& 12628050& 2200	& 374	& 2513940	  & 2208775	& 116	& 1160	& 197	& 1325532	& 1040& 1044148	& -11583902	& 0.08 \\
-\end{raggedright}
-\bottomrule  
-    \end{tabular}%  
-    \begin{tablenotes}\footnotesize  
-        \item $\ddagger$ Based on average number of cases per year in New Zealand from 1997 onward
-        \item $\dagger$ {1000 simulations}
-        \item \sym{*} Proportion of cases hospitalised 0.17
-        \item \sym{**} Wage losses per case \$852 and cost per hospitalised case \$1710
-        \item \sym{***} Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
-    \end{threeparttable}  
-\end{sidewaystable}
-
-\normalsize
 
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-061}
-\end{center}
-\caption{Age and vaccination status of cases}
-\label{fig:ageandvac}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-062}
-\end{center}
-\caption{Age of vaccination of vaccinated measles cases}
-\label{fig:vaccstat}
-\end{figure}
-
 \subsection{Benefit--cost analysis discussion}

### 2014-10-29 rworldmaps_measles_travel.R

@@ -1,0 +1,751 @@
+rm(list=ls())
+
+library(adehabitat)
+library(ggplot2)
+library(ggmap)
+library(grid)
+library(gridExtra)
+library(maps)       # Provides functions that let us plot the maps
+library(mapdata)    # Contains the hi-resolution points that mark out the countries.
+library(mapproj)
+library(maptools)
+library(plyr)
+library(rworldmap)
+library(rgdal)
+require(rgeos)
+library(raster)
+library(sp)
+#library(SDMTools)
+gpclibPermit()
+library(utils)
+
+# Data from http://thematicmapping.org/downloads/world_borders.php.
+# Direct link: http://thematicmapping.org/downloads/TM_WORLD_BORDERS_SIMPL-0.3.zip
+# Unpack and put the files in a dir 'data'
+
+getwd()
+#setwd("~/data")
+#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
+# world.map <- readOGR(dsn="C:/Users/David Hayman/Documents/data", layer="TM_WORLD_BORDERS_SIMPL-0.3")
+
+#setwd("C:/Users/David Hayman/Documents/GitHub/measles/maps")
+
+incidence<-read.csv("incidence_series.csv",header=T)
+head(incidence)
+
+cover<-read.csv("coverage_series.csv",header=T)
+head(cover)
+
+immigration<-read.csv("immigration.csv",header=T)
+head(immigration)
+
+population<-read.csv("population.csv", header=T)
+head(population)
+
+annualtravel<-read.csv("annualtravel.csv",header=T)
+head(annualtravel)
+
+monthtravel<-read.csv("monthtravel.csv",header=T)
+head(monthtravel)
+
+#######
+
+# world.ggmap <- fortify(world.map, region = "NAME")
+# save(world.ggmap,file="world.ggmap.Rda")
+load("world.ggmap.Rda")
+n <- length(unique(world.ggmap$id))
+
+id = unique(world.ggmap$id)
+id<-as.data.frame(id)
+colnames(cover)[3]<-"id"
+colnames(incidence)[3]<-"id"
+
+cover$id
+
+summary(immigration$Nationality,maxsum=Inf)
+## match immigration names to IDs from maps
+levels(immigration$Nationality)[match(c("Antigua & Barbuda","Bosnia & Herzegovina","Myanmar","Ivory Coast",
+                                        "Faeroe Islands","Guinea - Bissau","Vatican City","Iran","North Korea",
[…]

### 2014-10-30 draftfinalreport.Rnw
major text revisions
minor editing still required. Bulk done

@@
 \usepackage{framed}
+\usepackage{float}
 \newcommand{\vecA}{\mathbf{A}}
@@
 In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
-We identified additional analyses that may help understand risk of infection in New Zealand, and present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007.
+We identified additional analyses that may help understand risk of infection in New Zealand.
+Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007.
 We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level.
@@
 \item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, as analysed from data from April, is well above one.
-This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
+This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
 \item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
 \item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
-\item The benefit--cost ratio analyses suggest additional vaccination is extremely beneficial financially.
+\item The mean wage losses per measles case is estimated to be \$852
+\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
+\item The mean management costs per case were \$1,765
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>>$1).
+\item Large outbreaks, with a mean size of approximately around cases per year, median of 2 cases, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
 \end{itemize}
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our preliminary modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).
+This is particularly in light of our modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).
 
@@
 
-We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand, and descriptive analyses of risk of infection due to previous vaccination history, current rates of immunity within the population.
+We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:risk_analyses}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:risk_analyses}), and current rates of immunity within the population (section~\ref{sub:risk_analyses}).
 In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\ref{sub:risk_analyses}).
 
@@
 \subsection{Risk of measles infection in New Zealand analyses methods}
+\label{sub:name}
+
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 15>>=
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
 par(mfrow=c(1,1))
@@
 
-\begin{figure}[h!]
+\begin{figure}
 \begin{center}
@@
 
+\subsection{Vaccination history and measles infection}
+
+In this section we describe the vaccination history of the measles cases from 2007--2014.
+
+\subsection{Population immunity}
+
+In this section we describe the population immunity levels at both a national level and that for each DHB in New Zealand.
+To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work.
+Specifically, we use the data in Table \ref{table:sero}, that uses the serosurvey results and the National Immunisation Register for under 6 year olds.
+
+\begin{table}[htdp]
+\begin{center}
+\begin{tabular}{lrrrr}
+\hline
+Age &  Proportion immune\\
+\hline
+0  & 0\\
+1  & 0.89\\
+2	& 0.92\\
+3	& 0.93\\
+4	& 0.93\\
+5	& 0.92\\
+6-13	& 0.8\\
+14-18	& 0.83\\
+19-23	& 0.77\\
+24-32	& 0.85\\
+33-52	& 0.92\\
+>52	& 0.99\\
+\hline
+\end{tabular}
+\end{center}
+\caption{Vaccination coverage and serosurvey estimates of immunity among different age classes}
+\label{table:sero}
+\end{table}%
+
 \subsection{Measles importation risk}
 
-For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}) to measure human movement to and from New Zealand.
+For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel data by country and year to measure human movement to and from New Zealand  (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}).
 We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
-Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
+Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
 We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
@@
 We use the data from 2012 because this year had the most complete WHO measles data and yet was most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+
+\subsection{Regression analyses results}
+\label{sub:regression_results}
+
+The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
+@
+\end{center}
+\caption{Distribution of cases per category used in the final regression model}
+\label{fig:histcasecat}
+\end{figure}
+
+The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width =11, height=11>>=
+res<-predict(model3)
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
+@
+\end{center}
+\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
+\label{fig:predict}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+@
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+@
+\end{center}
+\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
+\label{fig:resid}
+\end{figure}
+
+The significance of the different predictor variables can be seen in the ANOVA results:
+\par
+\vspace{5mm} %5mm vertical space
+<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
+library(xtable)
+resan<-xtable(anova(model3,test="F"))
+print(resan,floating = FALSE)
+@
+\vspace{5mm} %5mm vertical space
+A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
+\vspace{5mm} %5mm vertical space
+<<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
+library(xtable)
+xtb<-xtable(summary(model3))
+print(xtb,floating = FALSE)
+@
+\vspace{5mm} %5mm vertical space
+
+\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
+
+People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
+Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+
+\subsection{Vaccination history and measles infection results}
+
+The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (Figure~\ref{fig:ageandvac}).
+However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second.
+A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (Figure~\ref{fig:vaccstat}).
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+AgeVac<-table(testv$VC,testv$AgeInYears)
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
+legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
+@
+\end{center}
+\caption{Age and vaccination status of cases}
+\label{fig:ageandvac}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
+abline(v=12,col="blue")
+hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+abline(v=60,col="red")
+legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
+legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
+@
+\end{center}
+\caption{Age of vaccination of vaccinated measles cases}
+\label{fig:vaccstat}
+\end{figure}
+
+\subsection{Population immunity results}
+
+The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
+However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
+Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
+The distribution of the numbers of naive and the total naive populations per DHB are given in \ref{fig:Northland}\ref{fig:Waitemata}\ref{fig:Auckland}\ref{fig:Counties_Manukau}\ref{fig:Waikato}\ref{fig:Lakes}\ref{fig:BayofPlenty}\ref{fig:Tairawhiti}\ref{fig:Taranaki}\ref{fig:HawkesBay}\ref{fig:Whanganui}\ref{fig:Midcentral}\ref{fig:Hutt}\ref{fig:CapitalandCoast}\ref{fig:Wairarapa}\ref{fig:NelsonMarlborough}\ref{fig:WestCoast}\ref{fig:Canterbury}\ref{fig:SouthCanterbury}\ref{fig:Southern}.
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+require(reshape2)  # this is the library that lets one flatten out data
+require(ggplot2)   # plotting library
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+@
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+plot(pop,xlab="Age",ylab="Population",cex.lab=2)
+points(naive,pch=16)
+legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
+@
+\end{center}
+\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\label{fig:naive}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per District Health Board}
+     \label{fig:naive_map}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+     \label{fig:Northland}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+     \label{fig:Waitemata}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+     \label{fig:Auckland}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+     \label{fig:Counties_Manukau}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+     \label{fig:Waikato}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+     \label{fig:Lakes}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+     \label{fig:BayofPlenty}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+     \label{fig:Tairawhiti}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+     \label{fig:Taranaki}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+     \label{fig:HawkesBay}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+     \label{fig:Whanganui}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+     \label{fig:Midcentral}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+     \label{fig:Hutt}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+     \label{fig:CapitalandCoast}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+     \label{fig:Wairarapa}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+     \label{fig:NelsonMarlborough}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+     \label{fig:WestCoast}
+\end{figure}
+
+\begin{figure}[H]
+   \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
+     \label{fig:Canterbury}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
+     \label{fig:SouthCanterbury}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Southern District Health Board}
+     \label{fig:Southern}
+\end{figure}
+
+\subsection{Measles importation risk results}
 
@@
 
-\subsection{Additional risk analyses results}
-
-Preliminary analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:immigration12}.
-However, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}).
+Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}.
+However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}.
+Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}.
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
@@
 The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
+The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12}\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12}\ref{fig:imrisk12}
 
@@
 %%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
-\begin{figure}[h!]
+\begin{figure}
 \begin{center}
@@
 This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per District Health Board}
-     \label{fig:naive_map}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-@
-
-\begin{figure}[ht!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
-\end{center}
-\caption{New Zealand population by age and estimated numbers of naive people in each age class}
-\label{fig:naive}
-\end{figure}
-
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Northland District Health Board}
-%     \label{fig:Northland}
-%\end{figure}
-%
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
-%     \label{fig:Waitemata}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
-%     \label{fig:Auckland}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
-%     \label{fig:Counties_Manukau}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
-%     \label{fig:Waikato}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
-%     \label{fig:Lakes}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
-%     \label{fig:BayofPlenty}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
-%     \label{fig:Tairawhiti}
-%\end{figure}
-
-%\begin{figure}
-%    \begin{center}
-%    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
-%     \label{fig:Taranaki}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
-%     \label{fig:HawkesBay}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
-%     \label{fig:Whanganui}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
-%     \label{fig:Midcentral}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
-%     \label{fig:Hutt}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
-%     \label{fig:CapitalandCoast}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
-%     \label{fig:Wairarapa}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
-%     \label{fig:NelsonMarlborough}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
-%     \label{fig:WestCoast}
-%\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
-     \label{fig:Canterbury}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
-     \label{fig:SouthCanterbury}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Southern District Health Board}
-     \label{fig:Southern}
-\end{figure}
-
-\subsection{Regression analyses results}
-\label{sub:regression_results}
-
-The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
-@
-\end{center}
-\caption{Distribution of cases per category used in the final regression model}
-\label{fig:histcasecat}
-\end{figure}
-
-The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width =11, height=11>>=
-res<-predict(model3)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-@
-\end{center}
-\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
-\label{fig:predict}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-@
-\end{center}
-\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
-\label{fig:resid}
-\end{figure}
-
-The significance of the different predictor variables can be seen in the ANOVA results:
-\par
-\vspace{5mm} %5mm vertical space
-<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
-library(xtable)
-resan<-xtable(anova(model3,test="F"))
-print(resan,floating = FALSE)
-@
-\vspace{5mm} %5mm vertical space
-A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
-\vspace{5mm} %5mm vertical space
-<<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
-library(xtable)
-xtb<-xtable(summary(model3))
-print(xtb,floating = FALSE)
-@
-\vspace{5mm} %5mm vertical space
-\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
-
-People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
-Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-AgeVac<-table(testv$VC,testv$AgeInYears)
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-@
-\end{center}
-\caption{Age and vaccination status of cases}
-\label{fig:ageandvac}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-@
-\end{center}
-\caption{Age of vaccination of vaccinated measles cases}
-\label{fig:vaccstat}
-\end{figure}
+Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling.
+The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (Figure~\ref{fig:trendimmigration}).
 
@@
 
-\subsection{Future risk analyses}
-We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those data (not shown) suggest that we require denominator data to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically for the multivariate analyses we wish to perform that detect interactions we require $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the report reviewed above.
-These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data only became available to us on 3 July 2014, provided by the University of Otago.
-
-The following analyses are in progress, for inclusion in later reports:
-\begin{itemize}
-\item Multivariate regression analyses to assess interactions between risk factors that may confound the univariate analyses.
-\item Update of the importation risk analyses using a broader range of years, including modelling the trend in importations.
-\end{itemize}
-Additional data we believe would enable us and the Ministry to better understand measles risk is fine scale (lower than District Health Board (DHB)) immunisation coverage data.
-We understand the National Immunisation Register (NIR) allows tracking of the vaccination status of children and this is very useful, but inclusion of these data at lower (e.g.
-meshblock, census area unit) level would allow better understanding of risk of measles infection and resource allocation because they may allow targeted immunisation programmes.
-Thus the data gap that we have that will hinder us providing fine scale risk maps is:
-\begin{itemize}
-\item Meshblock (or census area unit) level immunisation coverage data to allow targeted immunisation and understanding of risk at a fine scale level.
-\end {itemize}
-An additional data set that would enable us to develop the understanding of measles importation risk is:
-\begin {itemize}
-\item The number of New Zealanders arriving from abroad each year, the countries to which they travelled, and length of travel.
-\end{itemize}
+The distribution of the measles cases (\ref{fig:ageandvac}) appears to reflect the distribution of naive in the population (\ref{fig:naive}).
+The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number ($>16\%$) of cases had been vaccinated at least once.
+However, the majority of these received only one vaccination, and when they were young (\ref{fig:vaccstat}).
+The majority of naive among the New Zealand population is clearly focused in the urban areas.
+Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than District Health Board (DHB)) immunisation coverage data ew not able to provide reliable results.
+We obtained the National Immunisation Register (NIR) data, but frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census.
+Thus the data gap that we have hinder us providing fine scale risk maps.
+
+The distribution of naive among the DHB also varies (Figures~\ref{fig:Northland}~\ref{fig:Waitemata}~\ref{fig:Auckland}~\ref{fig:Counties_Manukau}~\ref{fig:Waikato}~\ref{fig:Lakes}~\ref{fig:BayofPlenty}~\ref{fig:Tairawhiti}~\ref{fig:Taranaki}~\ref{fig:HawkesBay}~\ref{fig:Whanganui}~\ref{fig:Midcentral}~\ref{fig:Hutt}~\ref{fig:CapitalandCoast}~\ref{fig:Wairarapa}~\ref{fig:NelsonMarlborough}~\ref{fig:WestCoast}~\ref{fig:Canterbury}~\ref{fig:SouthCanterbury}~\ref{fig:Southern}).
+We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
+
+Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries most travel to New Zealand is coming from.
+However, the strong seasonality in travel around Christmas suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
 
@@
 \begin{itemize}
+\item Risk of measles infection decreases significantly with age.
+\item Pacific people are statistically more at risk of measles infection.
+\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
+\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk than European or Maori children.
+\item The majority of cases are unvaccinated.
+\item The majority of vaccines failures occur in those people which revieved single vaccinations around 1 year old.
+\item Distribution of numbers of naive among New Zealand is uneven, with the majority predictably in urban areas.
 \item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
-\item There may be seasonal changes in risk of measles importation, though further analyses are needed.
-\item Risk of measles infection decreases significantly with age
-\item Pacific people are statistically more at risk of measles infection
-\item There is some statistical support for those living in better socio-economic situations being at greater risk of measles
-\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk than European or Maori children
+\item There may be seasonal changes in risk of measles importation, with travel numbers peaking in December.
 \end{itemize}
@@
 In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
-All agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
+All modelling studies agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
 \citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
@@
 
-To understand the level of immunity in the population, the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
+To understand the transmission dynamics of measles in the partially immune population and how likely an outbreak was of becoming endemic, we estimated $R_v$ from all the outbreaks in New Zealand since 2009.
 To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
@@
 
-To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we make several simplifying assumptions regarding the relationship between the proportion requiring vaccination, $p_c$, and $R_0$.
-Specifically, we use the simple relationship:
-
-\begin{equation} \label{eq:pc}
-p_c = 1-(1/R_0)
+To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we use the well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population is \citep{diekmann13}
+\begin{equation} \label{eq:eqn1}
+\log\left(1-\Pe\right)+\Ro\Pe=0
 \end{equation}
-To account for the different outbreak sizes we weighted our $R_v$ estimates by outbreak size and used 1.96 standard deviations of the mean of our $R_v$ estimates, to generate 95\% confidence limits from 2009--2014 outbreaks or the 2013--2014 outbreak for our modelling scenarios below.
-
-We then assume that the measles epidemics are occurring in a naive population that is the size of the naive and currently susceptible population remaining in New Zealand and thus $R_0$ is equivalent to $R_v$.
-We estimated the susceptible population in New Zealand from the national level vaccination coverage data provided by the Ministry of Health for this work (approximately 11\%, Figure~\ref{fig:naive}).
-Thus, the $p_c$ applies to the naive population in New Zealand and these are the additional vaccinations required to reduce $R_v$ to one in the overall New Zealand population.
-
-We then used the proportions of the population that were estimated to require vaccination ($p_c$, Equation~\ref{eq:pc}), and the proportions of the naive population per age case from Figure~\ref{fig:naive} to provide different scenarios to achieve those goals.
-These are explored and discussed in relation to the risk analyses (Section~\ref{sub:regression_results}) in the benefit--cost analyses section below (Section~\ref{sub:cost_benefit}).
-
-Finally, we used the same model structure and generation times to estimate the epidemic sizes once vaccination catch up protocols had been implemented assuming $R_v$ < 1 and the population of susceptible people in New Zealand was now reduced by the numbers vaccinated.
-Because the model is stochastic, we ran 1000 simulations and report the mean values, though the distribution of these is shown in the benefit--cost section (Section~\ref{sub:cost_benefit}).
-
-\section{Mick Introduction}
-
-The well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population is \citep{diekmann13}
-$$\log\left(1-\Pe\right)+\Ro\Pe=0$$
+
 where $\Ro$ is the basic reproduction number and $\Pe$ is the proportion of the population infected over the course of the outbreak.
+
 If a proportion $x_0$ of the population is susceptible following vaccination, then the  reproduction number under vaccination is $\Rr_V=x_0\Ro$, and the final size equation becomes
-$$\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0$$
+
+\begin{equation} \label{eq:eqn2}
+\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0
+\end{equation}
+
 Hence the relationship between the proportion initially susceptible and the proportion infected in an epidemic is
-$$x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}$$
+
+\begin{equation} \label{eq:eqn3}
+x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}
+\end{equation}
+
 In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks is $x_0-1/\Ro$.
@@
 The probability density of the $R_v$ estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak, which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
+Of particular note is the ongoing outbreak (as of 12 June 2014), which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
 An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
@@
 
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim report.
+The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim reports.
 The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one.
@@
 
-To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:pc}.
-Depending on $R_v$ the proportion of the national population requiring additional vaccination to make $R_v$ < 1 ranged from 0 ($R_v$ = 0.92, lower 95\% confidence limit) to an upper 95\% confidence limit of 53\% for the 2013--2014 outbreak ($R_v$ = 2.13, Figure~\ref{fig:r0}).
+To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:eqn3}.
+The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\%, with a national average of approximately 28\%, Table~\ref{table:attack}).
 These additional vaccination numbers can be made up in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
@@
 The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
-However, the naive population (Figure~\ref{fig:naive}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary.
-Researchers found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached.
+However, the naive population (Figures~\ref{fig:naive}\ref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary (and see Table~\ref{table:attack}.
+The results of these analyses support other studies.
+Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% naive.
 Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+Future analyses at a smaller, more local level would be useful, but the lack of appropriate data matching between the NIR data and the census unit area data prevented these.
 
@@
 Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$.			}
-\label{Table1}
+\label{table:attack}
 \end{table}%
@@
 \item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number, $R_v$, for measles in the current outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
-\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
- \item The reproduction number, $R_v$, for measles in the current 2013--2014 outbreak is well over one, suggesting that this outbreak has the potential to persist for prolonged periods within the population is it circulating.
- \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
-\end{itemize}
-
-\subsection{Future modelling}
-Future modelling we aim to perform are:
-\begin{itemize}
-\item An update of previous ODE models of measles in the overall population according the differing vaccine coverage scenarios \citep{roberts4}.
-\item Model measles outbreaks with differing scenarios of measles importation into various population groups based on current introduction rates.
+ \item The reproduction number, $R_v$, for measles in the current outbreak is over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
+ \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 17\% to 34\% among DHBs, and 28\% at the national level (approximately 131,500 vaccinations).
 \end{itemize}
@@
 This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008--2014 was estimated to be \4,327 (\$852 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
-
-This final figure brings an approximate estimate of \$809,149 for 187 cases for the current 2014 outbreak, which is comprised of earnings lost, case management and hospitalisation costs.
+Overall, the cost per case from 2008--2014 was estimated to be \$4,327 (\$852 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
+
+This final figure brings an approximate estimate of \$809,149 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs.
 
@@
 To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ distribution from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}).
-We assumed recovery from infection led to immunity and thus constantly reduced the population size by the number recovered.
-We used these values of numbers of predicted cases and the cost figures above to estimate what the cost of not vaccinating additional populations would be.
-We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this.
-We can therefore estimate the number of cases prevented, and the savings made from the additional vaccinations.
-We then use the costs of the catch up vaccination schemes, the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity, and the savings from reduced measles cases to work out the benefit--cost ratio ((B/C), where a B/C ratio > 1 means that the program benefits exceed their costs.
+Primarily, we used Equation~\ref{equation:eqn3} to estimate the proportion of the naive populations in each DHB (Figure~\ref{fig:naive_map}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{sec:epidemic_modelling}).
+We assumed that the cases prevented by this was proportional to the outbreak size.
+However, there is a continued risk of introduction and despite $R_v$ being <1, outbreaks may occur.
+Thus, we simulated expected outbreak sizes with $R_v$ < 1.
+We used these values of numbers of predicted cases prevented and numbers expected despite additional vaccination to calculate the savings.
+The cost figures above were used to estimate what the savings for vaccinating additional populations was, using estimated per case costs.
+The costs of the catch up vaccination schemes were estimated to be two differnt values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
+Values of \$20 and \$50 were used.
+Thus the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity and the vaccination schemes were used to estimate the costs of additional vaccine programs, and the savings from reduced measles cases to work out the benefit, and the B/C ratio estimated.
+A B/C ratio > 1 means that the program benefits exceed their costs.
 A B/C value less than one suggests the costs are higher than the economic benefits.
-Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}:
-
-\begin{equation} \label{eq:disc}
-\textit{present value} = \sum \textit{future value}_y_r * (1/(1+\textit{discount rate})^{yr}
+Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}.
+Thus:
+
+\begin{equation} \label{eq:bc}
+ \textit{B/C} = \frac{\sum\limits_{n=0}^{10} \frac{B_n}{(1+r)^n}}{\sum\limits_{n=0}^{10} \frac{C_n}{(1+r)^n}}
  \end{equation}
-Where $_y_r$ is the year 0 (current) through to 9 (ten years into the future).
-
-Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the susceptible New Zealand population would become infected during the outbreak, or the epidemic it would stutter after a few infects and fail to cause a large epidemic due to stochasticity.
-We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
-However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
-However, to address this, we not only use $R_v$ values of 1 in simulation models to estimate predicted outbreak sizes, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year since 1997.
-This was 220 cases per year.
+Where B is the benefit in saved funds from cases prevented and C is the cost of vaccination, discounted over time, where $n$ is the year and $r$ the discount rate of 0.03.
+
+\subsection{Benefit--cost analyses results}
+
+The estimated vaccinaton rates, with percentages are shown in Table~\ref{table:vacp}.
+<<echo=FALSE>>=
+vacp<-read.csv("vacc_predictions.csv",header=T)
+@
+
+\begin{sidewaystable}\small
+\caption{DHB and National level catch up vaccination rates and estimated outbreak sizes post vaccination}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(vacp, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:vacp}
+\end{sidewaystable}
 
@@
      \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
+<<fig=TRUE,echo=FALSE,width = 11,height=11>>=
 ##
@@
 
+<<echo=FALSE>>=
+costtable20<-read.csv("cost_benefit_20.csv",header=T)
+@
+
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 20 dollars per vaccine}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable20, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cost20}
+\end{sidewaystable}
+
+<<echo=FALSE>>=
+costtable50<-read.csv("cost_benefit_50.csv",header=T)
+@
+
+\begin{sidewaystable}\small
+\caption{Cost benefit analyses with 50 dollars per vaccine}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(costtable50, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cost50}
+\end{sidewaystable}
+
+The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\ref{table:vacp}.
+The benefit--cost results are in Table~\ref{tab:cost20} and Table~\ref{tab:cost50}, for two different vaccine prices.
+The results in the two tables show two alternative trends.
+In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
+However, if the previous recent history of cases since 1997 is the \emph{status quo} and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one is not a cost effective exercise (Table~\ref{tab:cost_ave}).
+
+It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
+
 For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
@@
 
-\subsection{Cost analysis discussion}
+\subsection{Benefit--cost analysis discussion}
+
+Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the susceptible New Zealand population would become infected during the outbreak, or the epidemic it would stutter after a few infects and fail to cause a large epidemic due to stochasticity.
+We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
+However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
+However, to address this, we not only use $R_v$ values of 1 in simulation models to estimate predicted outbreak sizes, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year since 1997.
+This was 220 cases per year.
+
 The results presented here are based on available data.
@@
 
-\subsection{Benefit--cost analyses results}
-
-<<echo=FALSE>>=
-costtable20<-read.csv("cost_benefit_20.csv",header=T)
-@
-
-\begin{sidewaystable}\small
-\caption{Cost benefit analyses with 20 dollars per vaccine}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable20, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cost20}
-\end{sidewaystable}
-
-<<echo=FALSE>>=
-costtable50<-read.csv("cost_benefit_50.csv",header=T)
-@
-
-\begin{sidewaystable}\small
-\caption{Cost benefit analyses with 50 dollars per vaccine}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(costtable50, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cost50}
-\end{sidewaystable}
-
-The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}.
-The results in the two tables show two alternative trends.
-In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
-However, if the previous recent history of cases since 1997 is the \emph{status quo} and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one is not a cost effective exercise (Table~\ref{tab:cost_ave}).
-
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
-
-\subsection{Benefit--cost analysis discussion}
 The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
@@
 
-\subsection{Cost analysis summary}
-
-\begin{itemize}
-\item Our initial estimates suggest the ongoing 2013--2014 measles outbreak has cost New Zealand over \$750,000.
-\end{itemize}
-
 \subsection{Benefit--cost analysis summary}
@@
 \item The mean wage losses per measles case is estimated to be \$852
-\item The mean cost per measles case attending hospital is estimated to be \$1710
+\item The mean cost per measles case attending hospital is estimated to be \$1,710
 \item Approximately 17\% of measles cases attend hospital
-\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
-\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
-\item Large outbreaks, with a mean size of approximately 151 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
-\end{itemize}
-
-\subsection{Future cost-benefit analyses}
-Using the results above we aim to:
-\begin{itemize}
-\item Estimate the costs and benefits for targeted vaccination, based either on the univariate analyses presented to date in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report or adjusted if any additional risk groups are identified in the multivariate analyses (section~\ref{sub:risk_analyses}) or modelling (section~\ref{sec:epidemic_modelling}).
-\end{itemize}
-
-We also require additional clarifications of the data, regarding:
-\begin {itemize}
-\item What hospital costs refer to, such as if a hospital day were 0 does that mean the case stayed in the hospital but not overnight? Or, does it mean the case stayed for less than 24 hours?
-\end {itemize}
-Once more complete data are available, comparisons of these results will be made to other published studies, discussed above.
-
-\section {Summary of key findings}
-\begin{itemize}
-\item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the world.
-\item The cost of the current measles outbreak is estimated to be at least \$750,000.
-\item Analyses of outbreak data suggest that measles $R_v$ values often include 1 and in this year, 2014, are well above one.
-This analysis suggests improved vaccination is a requisite to prevent measles becoming endemic again.
-\item Risk of measles infection decreases significantly with age
-\item Pacific people are statistically more at risk on a per capita basis, as are those living in better socio-economic situations
-\item Pacific and Asian children in the 6--17 year age categories have been at lower risk of measles than European or Maori children of the same age
-\item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 0\% ($R_v$ < 1) to 53\% ($R_v$ 2.13, upper 95\% confidencelimit 2013--2014 outbreak, approximately 250,000 vaccinations), depending on the appropriate reproduction number, $R_v$, for measles in New Zealand.
-\item The mean wage losses per measles case is estimated to be \$852
-\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
-\item For $R_v$ values estimated for the 2013--2014 outbreak and the upper bounds estimated for all outbreaks since 2009, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
-\item For $R_v$ values at the lower bounds of the estimates for all outbreaks since 2009, the benefits of catch up vaccination strategies are not clear and may not be cost effective (<1 B/C ratio).
-\item Large outbreaks, with a mean size of approximately 151 cases per year, median of 2, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
+\item The mean management cost per measles case for the current outbreak, estimated from three months of data is estimated to be \$1,765
+\item Using $R_0$ values estimated for measles in New Zealand prior to vaccination and the current naive population sizes, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
+\item Outbreaks with a mean size of approximately 61 cases per year, but median of 2 cases per year may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
+There is, however, a small chance for some of these outbreaks to reach several thousand in number.
 \end{itemize}

### 2014-10-30 draftfinalreport.tex
major text revisions
minor editing still required. Bulk done

@@
 We use the data from 2012 because this year had the most complete WHO measles data and yet was most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+
+\subsection{Regression analyses results}
+\label{sub:regression_results}
+
+The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-014}
+\end{center}
+\caption{Distribution of cases per category used in the final regression model}
+\label{fig:histcasecat}
+\end{figure}
+
+The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-015}
+\end{center}
+\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
+\label{fig:predict}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-017}
+\end{center}
+\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
+\label{fig:resid}
+\end{figure}
+
+The significance of the different predictor variables can be seen in the ANOVA results:
+\par
+\vspace{5mm} %5mm vertical space
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Thu Oct 30 20:32:30 2014
+\begin{tabular}{lrrrrrr}
+  \hline
+ & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
+  \hline
+NULL &  &  & 39 & 1453.62 &  &  \\ 
+  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
+  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
+  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
+  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
+   \hline
+\end{tabular}\vspace{5mm} %5mm vertical space
+A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
+\vspace{5mm} %5mm vertical space
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Thu Oct 30 20:32:30 2014
+\begin{tabular}{rrrrr}
+  \hline
+ & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
+  \hline
+(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
+  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
+  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
+  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
+  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
+  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
+  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
+  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
+  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
+  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
+  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
+  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
+  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
+  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
+  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
+  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
+  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
+  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
+  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
+  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
+  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
+   \hline
+\end{tabular}\vspace{5mm} %5mm vertical space
+
+\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
+
+People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
+Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+
+\subsection{Vaccination history and measles infection results}
+
+The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (Figure~\ref{fig:ageandvac}).
+However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second.
+A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (Figure~\ref{fig:vaccstat}).
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-020}
+\end{center}
+\caption{Age and vaccination status of cases}
+\label{fig:ageandvac}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-021}
+\end{center}
+\caption{Age of vaccination of vaccinated measles cases}
+\label{fig:vaccstat}
+\end{figure}
+
+\subsection{Population immunity results}
+
+The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
+However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
+Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
+The distribution of the numbers of naive and the total naive populations per DHB are given in \ref{fig:Northland}\ref{fig:Waitemata}\ref{fig:Auckland}\ref{fig:Counties_Manukau}\ref{fig:Waikato}\ref{fig:Lakes}\ref{fig:BayofPlenty}\ref{fig:Tairawhiti}\ref{fig:Taranaki}\ref{fig:HawkesBay}\ref{fig:Whanganui}\ref{fig:Midcentral}\ref{fig:Hutt}\ref{fig:CapitalandCoast}\ref{fig:Wairarapa}\ref{fig:NelsonMarlborough}\ref{fig:WestCoast}\ref{fig:Canterbury}\ref{fig:SouthCanterbury}\ref{fig:Southern}.
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-023}
+\end{center}
+\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\label{fig:naive}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per District Health Board}
+     \label{fig:naive_map}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+     \label{fig:Northland}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+     \label{fig:Waitemata}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+     \label{fig:Auckland}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+     \label{fig:Counties_Manukau}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+     \label{fig:Waikato}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+     \label{fig:Lakes}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+     \label{fig:BayofPlenty}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+     \label{fig:Tairawhiti}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+     \label{fig:Taranaki}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+     \label{fig:HawkesBay}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+     \label{fig:Whanganui}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+     \label{fig:Midcentral}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+     \label{fig:Hutt}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+     \label{fig:CapitalandCoast}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+     \label{fig:Wairarapa}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+     \label{fig:NelsonMarlborough}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+     \label{fig:WestCoast}
+\end{figure}
+
+\begin{figure}[H]
+   \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
+     \label{fig:Canterbury}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
+     \label{fig:SouthCanterbury}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
+     \end{center}
+     \caption{Numbers of naive individuals per age class, Southern District Health Board}
+     \label{fig:Southern}
+\end{figure}
+
+\subsection{Measles importation risk results}
 
@@
 The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-026}
+The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12}\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12}\ref{fig:imrisk12}
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-036}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-027}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-037}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-028}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-038}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-029}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-039}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-030}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-040}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-031}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-041}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-032}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-042}
 \end{center}
@@
 
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-033}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-043}
 \end{center}
@@
 %%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-034}
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-044}
 \end{center}
@@
 This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per District Health Board}
-     \label{fig:naive_map}
-\end{figure}
-
-\begin{figure}[ht!]
-\begin{center}
-\includegraphics{draftfinalreport-036}
-\end{center}
-\caption{New Zealand population by age and estimated numbers of naive people in each age class}
-\label{fig:naive}
-\end{figure}
-
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Northland District Health Board}
-%     \label{fig:Northland}
-%\end{figure}
-%
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
-%     \label{fig:Waitemata}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
-%     \label{fig:Auckland}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
-%     \label{fig:Counties_Manukau}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
-%     \label{fig:Waikato}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
-%     \label{fig:Lakes}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
-%     \label{fig:BayofPlenty}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
-%     \label{fig:Tairawhiti}
-%\end{figure}
-
-%\begin{figure}
-%    \begin{center}
-%    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
-%     \label{fig:Taranaki}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
-%     \label{fig:HawkesBay}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
-%     \label{fig:Whanganui}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
-%     \label{fig:Midcentral}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
-%     \label{fig:Hutt}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
-%     \label{fig:CapitalandCoast}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
-%     \label{fig:Wairarapa}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
-%     \label{fig:NelsonMarlborough}
-%\end{figure}
-%
-%\begin{figure}
-%     \begin{center}
-%     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
-%     \end{center}
-%     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
-%     \label{fig:WestCoast}
-%\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
-     \label{fig:Canterbury}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
-     \label{fig:SouthCanterbury}
-\end{figure}
-
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
-     \end{center}
-     \caption{Numbers of naive individuals per age class, Southern District Health Board}
-     \label{fig:Southern}
-\end{figure}
-
-\subsection{Regression analyses results}
-\label{sub:regression_results}
-
-The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-037}
-\end{center}
-\caption{Distribution of cases per category used in the final regression model}
-\label{fig:histcasecat}
-\end{figure}
-
-The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-038}
-\end{center}
-\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
-\label{fig:predict}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-040}
-\end{center}
-\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
-\label{fig:resid}
-\end{figure}
-
-The significance of the different predictor variables can be seen in the ANOVA results:
-\par
-\vspace{5mm} %5mm vertical space
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Wed Oct 29 19:17:39 2014
-\begin{tabular}{lrrrrrr}
-  \hline
- & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
-  \hline
-NULL &  &  & 39 & 1453.62 &  &  \\ 
-  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
-  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
-  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
-  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
-   \hline
-\end{tabular}\vspace{5mm} %5mm vertical space
-A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
-\vspace{5mm} %5mm vertical space
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Wed Oct 29 19:17:39 2014
-\begin{tabular}{rrrrr}
-  \hline
- & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-  \hline
-(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
-  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
-  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
-  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
-  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
-  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
-  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
-  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
-  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
-  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
-  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
-  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
-  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
-  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
-  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
-  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
-  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
-  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
-  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
-  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
-  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
-   \hline
-\end{tabular}\vspace{5mm} %5mm vertical space
-\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
-
-People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
-Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-043}
-\end{center}
-\caption{Age and vaccination status of cases}
-\label{fig:ageandvac}
-\end{figure}
-
-\begin{figure}[h!]
-\begin{center}
-\includegraphics{draftfinalreport-044}
-\end{center}
-\caption{Age of vaccination of vaccinated measles cases}
-\label{fig:vaccstat}
-\end{figure}
+Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling.
+The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (Figure~\ref{fig:trendimmigration}).
 
@@
 To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.
-Primarily, we simulated 1000 measles outbreaks using the estimated $R_v$ distribution from (Section~\ref{sec:epidemic_modelling}) in the estimated susceptible population of naive New Zealanders (Figure~\ref{fig:naive}).
-We assumed recovery from infection led to immunity and thus constantly reduced the population size by the number recovered.
-We used these values of numbers of predicted cases and the cost figures above to estimate what the cost of not vaccinating additional populations would be.
-We also simulated what we might expect measles outbreaks to look like following introductions in the population, given that $R_v$ would be one, though where our estimates for $R_v$ are already < 1, we use this.
-We can therefore estimate the number of cases prevented, and the savings made from the additional vaccinations.
-We then use the costs of the catch up vaccination schemes, the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity, and the savings from reduced measles cases to work out the benefit--cost ratio ((B/C), where a B/C ratio > 1 means that the program benefits exceed their costs.
+Primarily, we used Equation~\ref{equation:eqn3} to estimate the proportion of the naive populations in each DHB (Figure~\ref{fig:naive_map}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{sec:epidemic_modelling}).
+We assumed that the cases prevented by this was proportional to the outbreak size.
+However, there is a continued risk of introduction and despite $R_v$ being <1, outbreaks may occur.
+Thus, we simulated expected outbreak sizes with $R_v$ < 1.
+We used these values of numbers of predicted cases prevented and numbers expected despite additional vaccination to calculate the savings.
+The cost figures above were used to estimate what the savings for vaccinating additional populations was, using estimated per case costs.
+The costs of the catch up vaccination schemes were estimated to be two differnt values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
+Values of \$20 and \$50 were used.
+Thus the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity and the vaccination schemes were used to estimate the costs of additional vaccine programs, and the savings from reduced measles cases to work out the benefit, and the B/C ratio estimated.
+A B/C ratio > 1 means that the program benefits exceed their costs.
 A B/C value less than one suggests the costs are higher than the economic benefits.
-Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}:
-
-\begin{equation} \label{eq:disc}
-\textit{present value} = \sum \textit{future value}_y_r * (1/(1+\textit{discount rate})^{yr}
+Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}.
+Thus:
+
+\begin{equation} \label{eq:bc}
+ \textit{B/C} = \frac{\sum\limits_{n=0}^{10} \frac{B_n}{(1+r)^n}}{\sum\limits_{n=0}^{10} \frac{C_n}{(1+r)^n}}
  \end{equation}
-Where $_y_r$ is the year 0 (current) through to 9 (ten years into the future).
-
-Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the susceptible New Zealand population would become infected during the outbreak, or the epidemic it would stutter after a few infects and fail to cause a large epidemic due to stochasticity.
-We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
-However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
-However, to address this, we not only use $R_v$ values of 1 in simulation models to estimate predicted outbreak sizes, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year since 1997.
-This was 220 cases per year.
+Where B is the benefit in saved funds from cases prevented and C is the cost of vaccination, discounted over time, where $n$ is the year and $r$ the discount rate of 0.03.
+
+\subsection{Benefit--cost analyses results}
+
+The estimated vaccinaton rates, with percentages are shown in Table~\ref{table:vacp}.
+
+\begin{sidewaystable}\small
+\caption{DHB and National level catch up vaccination rates and estimated outbreak sizes post vaccination}
+%latex.default(vacp, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrrrrrrr}
+\hline\hline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Size}&\multicolumn{1}{c}{Naïve}&\multicolumn{1}{c}{Attack}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Proportion}&\multicolumn{1}{c}{Naïve.post.vaccination}&\multicolumn{1}{c}{Median.outbreak}&\multicolumn{1}{c}{Mean.outbreak}\tabularnewline
+\hline
+Auckland&$ 436350$&$ 52010$&$ 31159$&$ 17920$&$0.34$&$ 34090$&$2$&$ 82$\tabularnewline
+Bay of Plenty&$ 206000$&$ 20679$&$  8437$&$  4585$&$0.22$&$ 16094$&$2$&$ 71$\tabularnewline
+Canterbury&$ 482180$&$ 51357$&$ 24695$&$ 13687$&$0.27$&$ 37670$&$2$&$ 62$\tabularnewline
+Capital and Coast&$ 283700$&$ 32625$&$ 18403$&$ 10461$&$0.32$&$ 22164$&$3$&$ 96$\tabularnewline
+Counties Manukau&$ 469300$&$ 55544$&$ 32903$&$ 18880$&$0.34$&$ 36664$&$3$&$ 50$\tabularnewline
+Hawke's Bay&$ 151700$&$ 15602$&$  6846$&$  3751$&$0.24$&$ 11851$&$2$&$ 56$\tabularnewline
+Hutt Valley&$ 138380$&$ 15198$&$  7836$&$  4388$&$0.29$&$ 10810$&$2$&$ 86$\tabularnewline
+Lakes&$  98196$&$ 10558$&$  5192$&$  2886$&$0.27$&$  7672$&$2$&$ 62$\tabularnewline
+MidCentral&$ 162560$&$ 17328$&$  8348$&$  4628$&$0.27$&$ 12700$&$2$&$ 75$\tabularnewline
+Nelson Marlborough&$ 137000$&$ 13059$&$  4411$&$  2356$&$0.18$&$ 10703$&$3$&$ 90$\tabularnewline
+Northland&$ 151690$&$ 14921$&$  5688$&$  3071$&$0.21$&$ 11850$&$3$&$ 70$\tabularnewline
+South Canterbury&$  55620$&$  5238$&$  1678$&$   893$&$0.17$&$  4345$&$3$&$ 72$\tabularnewline
+Southern&$ 297420$&$ 31607$&$ 15115$&$  8371$&$0.26$&$ 23236$&$2$&$102$\tabularnewline
+Tairawhiti&$  43650$&$  4769$&$  2431$&$  1359$&$0.28$&$  3410$&$2$&$ 47$\tabularnewline
+Taranaki&$ 109750$&$ 11473$&$  5262$&$  2899$&$0.25$&$  8574$&$3$&$ 68$\tabularnewline
+Waikato&$ 359310$&$ 39402$&$ 20248$&$ 11331$&$0.29$&$ 28071$&$2$&$ 95$\tabularnewline
+Wairarapa&$  41112$&$  3932$&$  1346$&$   720$&$0.18$&$  3212$&$3$&$ 59$\tabularnewline
+Waitemata&$ 525550$&$ 58350$&$ 30774$&$ 17291$&$0.30$&$ 41059$&$2$&$ 70$\tabularnewline
+West Coast&$  32151$&$  3197$&$  1265$&$   685$&$0.21$&$  2512$&$2$&$ 50$\tabularnewline
+Whanganui&$  60120$&$  6075$&$  2530$&$  1378$&$0.23$&$  4697$&$2$&$ 58$\tabularnewline
+Total&$4241739$&$462924$&$234567$&$131540$&$0.28$&$331384$&$2$&$106$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:vacp}
+\end{sidewaystable}
 
@@
      \centering
-\includegraphics{draftfinalreport-058}
+\includegraphics{draftfinalreport-060}
      \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave})}
@@
      \centering
-\includegraphics{draftfinalreport-060}
+\includegraphics{draftfinalreport-062}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
@@
      \centering
-\includegraphics{draftfinalreport-061}
+\includegraphics{draftfinalreport-063}
      \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
@@
 \end{figure}
-
-For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
-We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
-We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
-
-\subsection{Cost analysis discussion}
-The results presented here are based on available data.
-While some of the data are complete and detailed, this is not true of all the data.
-In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
-For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
-In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
-In addition, similar data would be needed for non-publicly funded hospitals, e.g.
-private clinics.
-Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
-
-Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
-Similar data are needed for the period preceding 2014.
-If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
-However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
-As other studies have demonstrated, direct costs required to manage measles are not linear.
-
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
-This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
-The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
-In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
-They estimated direct costs per case to be less than US\$500.
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
-In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
-In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
-The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
-
-\subsection{Benefit--cost analyses results}
 
@@
 
-The benefit--cost results are in Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}.
+The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\ref{table:vacp}.
+The benefit--cost results are in Table~\ref{tab:cost20} and Table~\ref{tab:cost50}, for two different vaccine prices.
 The results in the two tables show two alternative trends.
@@
 
+For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
+Specifically, we used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
+We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
+We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
+
 \subsection{Benefit--cost analysis discussion}
+
+Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the susceptible New Zealand population would become infected during the outbreak, or the epidemic it would stutter after a few infects and fail to cause a large epidemic due to stochasticity.
+We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
+However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
+However, to address this, we not only use $R_v$ values of 1 in simulation models to estimate predicted outbreak sizes, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year since 1997.
+This was 220 cases per year.
+
+The results presented here are based on available data.
+While some of the data are complete and detailed, this is not true of all the data.
+In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
+For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
+In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
+In addition, similar data would be needed for non-publicly funded hospitals, e.g.
+private clinics.
+Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
+
+Detailed measles outbreak management costs were provided for the period of January 1 -- March 9, 2014.
+Similar data are needed for the period preceding 2014.
+If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
+However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
+As other studies have demonstrated, direct costs required to manage measles are not linear.
+
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
+This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
+In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
+They estimated direct costs per case to be less than US\$500.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
+The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
+
 The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.

### 2014-11-02 draftfinalreport.Rnw
Revised text
Beginning of tidying drafts.

@@
 \item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
+\item Estimates of the proportion of the currently naive New Zealand population (11\%) requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
 \item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
-However, New Zealander and immigrant or non--New Zealander travelling is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
-\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, as analysed from data from April, is well above one.
+However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this 2013--2014 outbreak, as analysed from data until June, is well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
-\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
 \item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
@@
 \item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
-\item The mean management costs per case were \$1,765
+\item The mean management cost per case was \$1,765
 \item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>>$1).
-\item Large outbreaks, with a mean size of approximately around cases per year, median of 2 cases, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
+\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
 \end{itemize}
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sec:}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}).
-Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there is spatial clustering of cases.
+Three figures, \emph{Figures 12, 13,} and \emph{28}, show that there was spatial clustering of cases.
 
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including important analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
 
@@
 Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
+Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
 These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
@@
 This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of individual NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 and NZDep 6--10.
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 and NZDep 6--10.
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
@@
 \end{center}
-\caption{Per capita rates of measles infections broken down by ethnicity, see Table~\ref{table:percap} for details}
+\caption{Per capita rates of measles infections for 2007--2014 broken down by ethnicity (Table~\ref{table:percap} for details)}
 \label{fig:percap}
@@
 
-\subsection{Vaccination history and measles infection}
-
-In this section we describe the vaccination history of the measles cases from 2007--2014.
-
-\subsection{Population immunity}
-
-In this section we describe the population immunity levels at both a national level and that for each DHB in New Zealand.
+\subsection{Vaccination history of measles infection and population immunity estimation methods}
+
+In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR.
+We describe the population immunity levels at both a national level and that for each DHB in New Zealand.
 To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work.
@@
 \end{center}
-\caption{Vaccination coverage and serosurvey estimates of immunity among different age classes}
+\caption{Vaccination coverage (0--5 years) and serosurvey estimates ($>5$ years) of immunity among different age classes used in these analyses}
 \label{table:sero}
@@
 
-\subsection{Measles importation risk}
+\subsection{Measles importation risk methods}
 
@@
 
-The distribution of the cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
+The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
 \begin{figure}
@@
 \end{center}
-\caption{Distribution of cases per category used in the final regression model}
+\caption{Distribution of measles cases per category used in the final regression model}
 \label{fig:histcasecat}
@@
 
-The significance of the different predictor variables can be seen in the ANOVA results:
+The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
 \par
 \vspace{5mm} %5mm vertical space
+\begin{table}
 <<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
@@
 @
+\caption{Significance of different predictor variables for measles risk factors from 2007-2014}
+\label{table:anova}
+\end{table}
 \vspace{5mm} %5mm vertical space
-A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
+A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.
+
 \vspace{5mm} %5mm vertical space
@@
 @
+\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+\label{table:regression}
+\end{table}
 \vspace{5mm} %5mm vertical space
@@
 \par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
-
-People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (Figure~\ref{fig:ageinyears}).
+
+People of Pacific origin are also over-represented as measles cases ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:}).
 Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
@@
 \end{center}
-\caption{Age and vaccination status of cases}
+\caption{Age and vaccination status of measles cases, 2007--2014}
 \label{fig:ageandvac}
@@
 \end{center}
-\caption{Age of vaccination of vaccinated measles cases}
+\caption{Age of vaccination of vaccinated measles cases, 2007--2014}
 \label{fig:vaccstat}
@@
 Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
-The distribution of the numbers of naive and the total naive populations per DHB are given in \ref{fig:Northland}\ref{fig:Waitemata}\ref{fig:Auckland}\ref{fig:Counties_Manukau}\ref{fig:Waikato}\ref{fig:Lakes}\ref{fig:BayofPlenty}\ref{fig:Tairawhiti}\ref{fig:Taranaki}\ref{fig:HawkesBay}\ref{fig:Whanganui}\ref{fig:Midcentral}\ref{fig:Hutt}\ref{fig:CapitalandCoast}\ref{fig:Wairarapa}\ref{fig:NelsonMarlborough}\ref{fig:WestCoast}\ref{fig:Canterbury}\ref{fig:SouthCanterbury}\ref{fig:Southern}.
+The distribution of the numbers of naive and the total naive populations per DHB, assuming national immunisation and immunity rates are representative, are given in Figures~\ref{fig:Northland},\ref{fig:Waitemata},\ref{fig:Auckland},\ref{fig:Counties_Manukau},\ref{fig:Waikato},\ref{fig:Lakes},\ref{fig:BayofPlenty},\ref{fig:Tairawhiti},\ref{fig:Taranaki},\ref{fig:HawkesBay},\ref{fig:Whanganui},\ref{fig:Midcentral},\ref{fig:Hutt},\ref{fig:CapitalandCoast},\ref{fig:Wairarapa},\ref{fig:NelsonMarlborough},\ref{fig:WestCoast},\ref{fig:Canterbury},\ref{fig:SouthCanterbury}, and \ref{fig:Southern}.
 
@@
 \end{center}
-\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\caption{New Zealand population by age and estimated numbers of naive people in each age class using national immunity data (Table~\ref{table:sero})}
 \label{fig:naive}
@@
      \end{center}
-     \caption{Numbers of naive individuals per District Health Board}
+     \caption{Numbers of naive individuals per District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:naive_map}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+     \caption{Numbers of naive individuals per age class, Northland District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Northland}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+     \caption{Numbers of naive individuals per age class, Waitemata District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Waitemata}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+     \caption{Numbers of naive individuals per age class, Auckland District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Auckland}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Counties_Manukau}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+     \caption{Numbers of naive individuals per age class, Waikato District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Waikato}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+     \caption{Numbers of naive individuals per age class, Lakes District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Lakes}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:BayofPlenty}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Tairawhiti}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+     \caption{Numbers of naive individuals per age class, Taranaki District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Taranaki}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:HawkesBay}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+     \caption{Numbers of naive individuals per age class, Whanganui District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Whanganui}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+     \caption{Numbers of naive individuals per age class, Midcentral District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Midcentral}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+     \caption{Numbers of naive individuals per age class, Hutt District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Hutt}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:CapitalandCoast}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Wairarapa}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:NelsonMarlborough}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+     \caption{Numbers of naive individuals per age class, West Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:WestCoast}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
+     \caption{Numbers of naive individuals per age class, Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Canterbury}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
+     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:SouthCanterbury}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Southern District Health Board}
+     \caption{Numbers of naive individuals per age class, Southern District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Southern}
@@
 \begin{table}
-\caption{Measles incidence per million (2012)}
+\caption{Highest measles incidence per million (2012)}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{New Zealander travel numbers (2012)}
+\caption{New Zealander travel numbers by destination (2012)}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by travellers numbers$*$measles incidence}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by travellers numbers$*$measles incidence}
 <<results=tex,echo=FALSE>>=
@@
      \includegraphics[width=0.9\textwidth]{cases_by_dhb_2009_2014.pdf}
-     \caption{Measles cases by district health board (DHB) from 2009 to 2014}
+     \caption{Measles cases by District Health Board (DHB) from 2009 to 2014}
      \label{fig:dhbcases}
@@
      \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
-     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation}
+     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation from 2009 to 2014}
      \label{fig:outbreaks}
@@
      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ (Average $R_0$) for the outbreaks each year, as classified by outbreaks in Figure~\ref{fig:outbreaks}}
+     \caption{Estimates of $R_v$ (Average $R_0$) for the outbreaks each year, as classified by outbreaks in Figure~\ref{fig:outbreaks}. Sporadic cases are excluded.}
      \label{fig:r0}
@@
 \begin{table}
-\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
+\caption{Reported direct costs, with the number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
 <<results=tex,echo=FALSE>>=

### 2014-11-04 draftfinalreport.Rnw
Draft revisions

@@
 \usepackage[utf8]{inputenc}
+\usepackage{enumitem}
 \usepackage{graphicx}
@@
 
-\title{Draft final report\\ Measles risk assessment, modelling and benefit--cost analysis\\ \vspace{2 mm} {\large David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French}}
-\author{mEpiLab and EpiCentre,\\ Infectious Diseases Research Centre,\\
-Massey University,\\
-Palmerston North 4442,\\
-New Zealand\\
-\href{mailto: D.T.S.Hayman@massey.ac.nz}{D.T.S.Hayman@massey.ac.nz}}  %\texttt formats the text to a typewriter style font
+\title{Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand}
+\author{{David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French.} \\\\
+{\small mEpiLab and EpiCentre,}\\
+{\small   Infectious Diseases Research Centre,}\\
+{\small   Massey University, Palmerston North 4442,New Zealand}\\
+{\small and}\\
+{\small Institute of Natural \& Mathematical Sciences,}\\
+{\small   New Zealand Institute for Advanced Study and the Infectious Disease Research Centre,}\\
+{\small   Massey University, Private Bag 102 904, North Shore Mail Centre, Auckland, New Zealand. } \\ 
+Email: \texttt{D.T.S.Hayman@massey.ac.nz}}
+\date{}
 \maketitle
@@
 Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007.
-We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level.
+We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level.
 We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
@@
 \item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
-\item Estimates of the proportion of the currently naive New Zealand population (11\%) requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
+\item Estimates of the proportion of the currently na\"{\i}ve New Zealand population (11\%) requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
@@
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
-\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
-\item The mean wage losses per measles case is estimated to be \$852
+\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$805,000.
+\item The mean wage losses per measles case is estimated to be \$839
 \item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
@@
 
-The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
-However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
-Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
-The distribution of the numbers of naive and the total naive populations per DHB, assuming national immunisation and immunity rates are representative, are given in Figures~\ref{fig:Northland},\ref{fig:Waitemata},\ref{fig:Auckland},\ref{fig:Counties_Manukau},\ref{fig:Waikato},\ref{fig:Lakes},\ref{fig:BayofPlenty},\ref{fig:Tairawhiti},\ref{fig:Taranaki},\ref{fig:HawkesBay},\ref{fig:Whanganui},\ref{fig:Midcentral},\ref{fig:Hutt},\ref{fig:CapitalandCoast},\ref{fig:Wairarapa},\ref{fig:NelsonMarlborough},\ref{fig:WestCoast},\ref{fig:Canterbury},\ref{fig:SouthCanterbury}, and \ref{fig:Southern}.
+The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
+However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
+Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
+The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in:
+\begin{itemize}[noitemsep,nolistsep]
+\item Northland: Figure~\ref{fig:Northland}
+\item Waitemata: Figure~\ref{fig:Waitemata}
+\item Auckland: Figure~\ref{fig:Auckland}
+\item Counties Manukau: Figure~\ref{fig:Counties_Manukau}
+\item Waikato: Figure~\ref{fig:Waikato}
+\item Lakes: Figure~\ref{fig:Lakes}
+\item Bay of Plenty: Figure~\ref{fig:BayofPlenty}
+\item Tairawhiti: Figure~\ref{fig:Tairawhiti}
+\item Taranaki: Figure~\ref{fig:Taranaki}
+\item Hawke's Bay: Figure~\ref{fig:HawkesBay}
+\item Whanganui: Figure~\ref{fig:Whanganui}
+\item Mid-Central: Figure~\ref{fig:Midcentral}
+\item Hutt: Figure~\ref{fig:Hutt}
+\item Capital and Coast: Figure~\ref{fig:CapitalandCoast}
+\item Wairarapa: Figure~\ref{fig:Wairarapa}
+\item Nelson Marlborough: Figure~\ref{fig:NelsonMarlborough}
+\item West Coast: Figure~\ref{fig:WestCoast}
+\item Canterbury: Figure~\ref{fig:Canterbury}
+\item South Canterbury: Figure~\ref{fig:SouthCanterbury}
+\item Southern: Figure~\ref{fig:Southern}
+\end{itemize}
 
@@
 \end{center}
-\caption{New Zealand population by age and estimated numbers of naive people in each age class using national immunity data (Table~\ref{table:sero})}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (Table~\ref{table:sero})}
 \label{fig:naive}
@@
      \end{center}
-     \caption{Numbers of naive individuals per District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:naive_map}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Northland District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Northland}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Waitemata District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Waitemata}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Auckland District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Auckland}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Counties_Manukau}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Waikato District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Waikato}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Lakes District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Lakes}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:BayofPlenty}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Tairawhiti}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Taranaki District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Taranaki}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:HawkesBay}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Whanganui District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Whanganui}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Midcentral District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Midcentral}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Hutt District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Hutt}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:CapitalandCoast}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Wairarapa}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:NelsonMarlborough}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, West Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:WestCoast}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Canterbury}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:SouthCanterbury}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Southern District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Southern}
@@
 
-Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}.
-However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}.
-Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}.
+Globally, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12} and \ref{fig:incidence12}.
+Analyses of 2012 data, the most complete and recent year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}.
+However, travel by New Zealanders is dominated by that to Australia (Table~\ref{table:nztravel12} and Figure~\ref{fig:nztravel12}).
+Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
@@
 The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
-The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12}\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12}\ref{fig:imrisk12}
+The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and \ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and \ref{fig:imrisk12}
 
@@
 
-Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:trendimmigration}).
+Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:travel}).
 This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
-Of note, however, is the clear seasonality in immigration (Figure~\ref{fig:trendimmigration}).
+Of note, however, is the clear seasonality in immigration and travel (Figure~\ref{fig:travel}).
 This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
 Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling.
-The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (Figure~\ref{fig:trendimmigration}).
+The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (Figure~\ref{fig:travel}).
 
@@
 Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
-A key issue with incorporating spatial immunisation data has been the denominator data, and how to deal with people of greater age than those recorded in the NIR.
-These analyses are possible future directions for this work.
-The results of these regression analyses are also discussed in the benefit--cost section below (Section~\ref{sub:cost_benefit})
-
-The distribution of the measles cases (\ref{fig:ageandvac}) appears to reflect the distribution of naive in the population (\ref{fig:naive}).
-The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number ($>16\%$) of cases had been vaccinated at least once.
-However, the majority of these received only one vaccination, and when they were young (\ref{fig:vaccstat}).
-The majority of naive among the New Zealand population is clearly focused in the urban areas.
-Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than District Health Board (DHB)) immunisation coverage data ew not able to provide reliable results.
-We obtained the National Immunisation Register (NIR) data, but frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census.
+A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data.
+The data suggest that census data and NIR data re recording children living in different DHB to that which they are vaccinated in, or subsequent to the census or NIR data moving.
+The leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated.
+Another issue was how to deal with people of greater age than those recorded in the NIR.
+Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.
+
+The distribution of the measles cases (Figure~\ref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve in the population (Figure~\ref{fig:naive}).
+The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number of cases ($>16\%$) had been vaccinated at least once (Figure~\ref{fig:ageandvac}).
+However, the majority of these vaccinated cases received only one vaccination, and were vaccinated when they were young (Figure~\ref{fig:vaccstat}).
+
+The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (Figure~\ref{fig:naive_map}).
+Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than District Health Board (DHB)) immunisation coverage data were not able to provide reliable results, as discussed above.
+The National Immunisation Register (NIR) data, but frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census.
 Thus the data gap that we have hinder us providing fine scale risk maps.
 
-The distribution of naive among the DHB also varies (Figures~\ref{fig:Northland}~\ref{fig:Waitemata}~\ref{fig:Auckland}~\ref{fig:Counties_Manukau}~\ref{fig:Waikato}~\ref{fig:Lakes}~\ref{fig:BayofPlenty}~\ref{fig:Tairawhiti}~\ref{fig:Taranaki}~\ref{fig:HawkesBay}~\ref{fig:Whanganui}~\ref{fig:Midcentral}~\ref{fig:Hutt}~\ref{fig:CapitalandCoast}~\ref{fig:Wairarapa}~\ref{fig:NelsonMarlborough}~\ref{fig:WestCoast}~\ref{fig:Canterbury}~\ref{fig:SouthCanterbury}~\ref{fig:Southern}).
+The distribution of na\"{\i}ve among the DHB also varies.
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
+However, it is worth bearing in mind that the data here assume that the national NIR vaccination coverage and serosurvey results are appropriate for all DHB.
 
@@
 However, the strong seasonality in travel around Christmas suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
+The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}).
+However, recently Australia was declare free of endemic measles and so the 2012 data must reflect imported measles there.
+However, other areas of high risk include the United Kingdom, where measles persists, South, Southeast and East Asia, where measles is endemic (Table~\ref{table:risk12} and Figure~\ref{fig:risk12}).
 
@@
 \item The majority of vaccines failures occur in those people which revieved single vaccinations around 1 year old.
-\item Distribution of numbers of naive among New Zealand is uneven, with the majority predictably in urban areas.
+\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHB with larger urban areas.
 \item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
@@
 All modelling studies agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one.
+\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one and eliminate endemic measles.
 They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
@@
 
-In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks is $x_0-1/\Ro$.
+In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks is
+
+\begin{equation} \label{eq:prop}
+$x_0-1/\Ro$.
+\end{equation}
 
@@
 The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current outbreak 1.82--2.13.
 The probability density of the $R_v$ estimates for each outbreak all include one.
@@
 
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0} and in the previous interim reports.
-The probability density of the $R_v$ estimates for each outbreak all include one, except for the ongoing 2013--2014 outbreak, which has an $R_v$ well above one.
-As discussed in the previous interim report, the important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
-However, the results of these differences are presented and discussed in the benefit--cost analysis section (Section~\ref{sub:cost_benefit}).
-The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current outbreak 1.82--2.13.
-
-The regularity of these outbreaks also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we explore the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
-
-The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
-However, sub-clinical and underreporting may lower the estimate.
-The relative contributions of both to our estimates are currently unknown.
-
-To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:eqn3}.
-The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\%, with a national average of approximately 28\%, Table~\ref{table:attack}).
+If fewer cases occur during each outbreak, then this suggests that $R_v$ is lower, but that measles is introduced to New Zealand more frequently.
+The regularity of these outbreaks, however, imply that the regular (approximately yearly) importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we include the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+
+To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:prop}.
+The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (Table~\ref{table:attack}).
 These additional vaccination numbers can be made up in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+However, they require differing numbers of vaccinations per DHB.
+Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (Table~\ref{table:attack}).
+The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (Table~\ref{table:attack}).
 
 The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
-However, the naive population (Figures~\ref{fig:naive}\ref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary (and see Table~\ref{table:attack}.
-The results of these analyses support other studies.
-Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% naive.
-Measles vaccination in various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+However, the na\"{\i}ve population (Figures~\ref{fig:naive} and \ref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary (Table~\ref{table:attack}).
+The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics.
+Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% na\"{\i}ve.
+Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
 Future analyses at a smaller, more local level would be useful, but the lack of appropriate data matching between the NIR data and the census unit area data prevented these.
@@
 \item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
-\item The reproduction number for measles in a partially immune population is often close to one, suggesting increased population level immunity is required to prevent this measles persisting.
+\item The reproduction number for measles in the partially immune New Zealand population is often close to one, suggesting increased population level immunity is required to ensure prevention of measles persistence following importation.
  \item The reproduction number, $R_v$, for measles in the current outbreak is over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
- \item Additional vaccination levels to push $R_v$ below one among the currently naive population in New Zealand range from 17\% to 34\% among DHBs, and 28\% at the national level (approximately 131,500 vaccinations).
+ \item Additional vaccination levels to push $R_v$ below one among the currently na\"{\i}ve population in New Zealand range from 17\% to 34\% among DHBs, and 28\% at the national level (approximately 131,500 vaccinations).
 \end{itemize}
@@
 
-In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current measles outbreak.
+In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current 2013--2014 measles outbreak.
 
@@
 Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$298 per contact.
+
 It was estimated that for the one week that the Iowa Department of Public Health (DPH) investigated a case in 2004, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination \citep{dayan5}.
@@
 
-Mean wages for New Zealand workers, by age and gender were obtained for the period, 2008--2013 from the New Zealand Income Survey (Statistics New Zealand, 2013).
-Measles cases were assumed to not work for a period of 5 days.
-Similarly, a care taker was assumed to not work for 5 days if the case were less than 20 years of age.
-In order to calculate the wage loss associated with the care taker, it was assumed that the person was a female between the ages of 35-39.
-Age and gender information for the 192 publicly funded hospital discharges with a measles primary diagnosis from 2008--2013 were matched to the New Zealand wage file to calculate lost wages due to measles.
+Wages lost due to measles were calculated for the period January 2008 - August 2014.
+Calculations were based on the assumptions that 5 days of work were lost for each case; however, individuals between 0-14 years of age were not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15-19 years of age was 41.9\%  (Statistics New Zealand, 2013).
+If the case were < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
+Total wage lost for the 247 cases and care givers was estimated to be \$207,155.
+This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases <15 years of age.
+Overall, the cost per case from 2008 - 2014 was estimated to be \$2,513 (\$839 in forgone wages and \$1,710 in hospital costs).
 
@@
 This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008--2014 was estimated to be \$4,327 (\$852 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
+Overall, the cost per case from 2008--2014 was estimated to be \$4,327 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
 
@@
 To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.
-Primarily, we used Equation~\ref{equation:eqn3} to estimate the proportion of the naive populations in each DHB (Figure~\ref{fig:naive_map}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{sec:epidemic_modelling}).
+Primarily, we used Equation~\ref{equation:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{sec:epidemic_modelling}).
 We assumed that the cases prevented by this was proportional to the outbreak size.
-However, there is a continued risk of introduction and despite $R_v$ being <1, outbreaks may occur.
+However, there is a continued risk of introduction and despite $R_v$ being <1, smaller outbreaks may occur.
 Thus, we simulated expected outbreak sizes with $R_v$ < 1.
 We used these values of numbers of predicted cases prevented and numbers expected despite additional vaccination to calculate the savings.
-The cost figures above were used to estimate what the savings for vaccinating additional populations was, using estimated per case costs.
+The cost figures above were used to estimate what the savings for vaccinating additional populations was, using estimated per case costs saved.
 The costs of the catch up vaccination schemes were estimated to be two differnt values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
-Values of \$20 and \$50 were used.
-Thus the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity and the vaccination schemes were used to estimate the costs of additional vaccine programs, and the savings from reduced measles cases to work out the benefit, and the B/C ratio estimated.
+Values of \$20 and \$50 per vaccine were used.
+Thus the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity and the vaccination schemes were used to estimate the costs of additional vaccine programs and costs from continued measles introductions.
+The financial savings from reduced measles cases and these costs were used to work out the estimated B/C ratio.
 A B/C ratio > 1 means that the program benefits exceed their costs.
@@
 \begin{equation} \label{eq:bc}
- \textit{B/C} = \frac{\sum\limits_{n=0}^{10} \frac{B_n}{(1+r)^n}}{\sum\limits_{n=0}^{10} \frac{C_n}{(1+r)^n}}
+ \textit{B/C} = \frac{\sum\limits_{n=0}^{9} \frac{B_n}{(1+r)^n}}{\sum\limits_{n=0}^{9} \frac{C_n}{(1+r)^n}}
  \end{equation}
-Where B is the benefit in saved funds from cases prevented and C is the cost of vaccination, discounted over time, where $n$ is the year and $r$ the discount rate of 0.03.
+Where B is the benefit in saved funds from cases prevented and C is the cost of vaccination, discounted over time, where $n$ is the year and $r$ the annual discount rate of 0.03 (3\%).
 
@@
 \begin{sidewaystable}\small
-\caption{DHB and National level catch up vaccination rates and estimated outbreak sizes post vaccination}
+\caption{DHB and National level catch up vaccination rates and estimated outbreak sizes post vaccination. Size: DHB Population, Statistics NZ 2013;  	
+Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
+Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
+Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}
+
 <<results=tex,echo=FALSE>>=
@@
 
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population and using the upper 95\% confidence limits of our $R_v$ estimates (Section~\ref{sec:epidemic_modelling}) are shown in Figure~\ref{fig:numvac}.
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using the equation~\ref{eq:prop} are shown in Figure~\ref{fig:numvac}.
 
@@
 legend("topright",c("Naive","28% naive, all ages","28% naive, all 2-17 yr olds"),
-       bty="n",pch=c(1,21,23),col=rep("grey",3),pt.bg=c("black","red","orange"),cex=2)
-@
-     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave})}
+       bty="n",pch=c(1,21,23),col=c("black","grey","grey"),pt.bg=c("black","red","orange"),cex=2)
+@
+     \caption{The estimated proportion of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost20} and ~\ref{tab:cost50})}
      \label{fig:numvac}
@@
 
-The expected number of cases in New Zealand, assuming homogenous mixing in a naive population of 11\% of the population (the current status) and assuming measles $R_v$ were 1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
-These simulations show that even in scenarios when $R_v$ is one, and thus stochastically should fail to persist (i.e.
-become endemic), large outbreaks can occur due to stochastic processes.
-Though the median value from 1000 simulations is low (2 cases), and thus most measles introductions will be single cases or lead to minor outbreaks, in this modelling exercise the mean value was 151 cases and the maximum nearly 25,000 cases.
+The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ were <1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
+These simulations show that even in scenarios when $R_v$ is lower than one, and thus stochastically should fail to persist (i.e.
+become endemic), outbreaks can occur due to stochastic processes following measles importation.
+The most likely scenario is that very few cases occur, with the median value from 1000 simulations low (2 cases).
+Thus most measles introductions will be single cases or lead to minor outbreaks.
+In this modelling exercise the mean value was 61 cases, suggesting this is likely to be an outbreak size to be expected.
+The maximum predicted number of cases was nearly 8,000 cases.
+Note, however, that larger values were very rarely predicted and this does not take into account spatial variation or heterogeneous contact rates.
 
@@
 #length(which(sizes>5))/1000
-legend('top',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
-@
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
+legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
+@
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
 The full distribution of results can be seen in Figure~\ref{fig:sim1}}
@@
 @
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
 A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
@@
 \begin{sidewaystable}\small
-\caption{Cost benefit analyses with 20 dollars per vaccine}
+\caption{Benefit--cost analyses with 20 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 \label{table:cost20}
+\begin{tablenotes}\footnotesize  
+        \item For numbers to vaccinate see Table~\ref{table:attack}
+        \item Proportion of cases hospitalised 0.17
+        \item Wage losses per case \$839 and cost per hospitalised case \$1710
+        \item Based on 10 introductions of measles, one per year
+      \end{tablenotes}  
 \end{sidewaystable}
+
+Vacc  Vacc.costs	Wages.saved	Manage.saved	Hospitalised	Hosp.saved	Costs.save	Outbreak	OB.costs	Benefit/cost
 
@@
 \begin{sidewaystable}\small
-\caption{Cost benefit analyses with 50 dollars per vaccine}
+\caption{Benefit--costs analyses with 50 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 \label{table:cost50}
+\begin{tablenotes}\footnotesize  
+        \item For numbers to vaccinate see Table~\ref{table:attack}
+        \item Proportion of cases hospitalised 0.17
+        \item Wage losses per case \$839 and cost per hospitalised case \$1710
+        \item Based on 10 introductions of measles, one per year
+      \end{tablenotes}  
 \end{sidewaystable}
 
-The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\ref{table:vacp}.
-The benefit--cost results are in Table~\ref{tab:cost20} and Table~\ref{tab:cost50}, for two different vaccine prices.
-The results in the two tables show two alternative trends.
-In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
-However, if the previous recent history of cases since 1997 is the \emph{status quo} and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one is not a cost effective exercise (Table~\ref{tab:cost_ave}).
-
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
-
 For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
+Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
 We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
@@
 
-\subsection{Benefit--cost analysis discussion}
-
-Our estimates for the numbers of measles cases are highly variable: if $R_v$ were greater than one, either the measles outbreak would take off and all the susceptible New Zealand population would become infected during the outbreak, or the epidemic it would stutter after a few infects and fail to cause a large epidemic due to stochasticity.
-We used the average of 1000 simulations to estimate the mean of these stochastic simulations for our benefit--cost analysis.
-However, these simulations assume a homogeneously mixed population and thus give large epidemic sizes once the epidemic gets beyond the initial stuttering infection chains and this assumption is discussed below.
-However, to address this, we not only use $R_v$ values of 1 in simulation models to estimate predicted outbreak sizes, but we also compared our results to those using the mean number of measles cases seen in New Zealand per year since 1997.
-This was 220 cases per year.
+The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\ref{table:vacp}.
+The benefit--cost results are in Table~\ref{table:cost20} and Table~\ref{table:cost50}, for two different vaccine prices.
+The results in the two tables show the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Tables~\ref{table:cost20} and \ref{table:cost50}).
+
+It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
+Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children.
+
+\subsection{Benefit--cost analyses discussion}
+
+Our estimates of the costs of the measles outbreaks in New Zealand suggest that measles management per case in New Zealand is high.
+We used the mean values per case in the absence of alternative data and used as much data as possible.
+The exact costs will vary, however, our benefit--cost analyses suggest that in all cases catch up vaccination schemes will be financially beneficial for New Zealand (see Tables~\ref{table:cost20} and \ref{table:cost50}).
 
@@
 While some of the data are complete and detailed, this is not true of all the data.
-In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data are needed.
+In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data would be better.
 For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
@@
 private clinics.
-Other factors that we aim to investigate are whether or not a linear term for case age is appropriate, or what if any interaction there might be between age and length of stay in hospital.
+Other factors that would be useful to investigate in the future include the relationship between the costs over time.
+For these analyses we use the mean cost per case.
+However, as stated previously, it is uncertain as to whether that is a valid measure or not and whether there may be other relationships.
+Furthermore, a linear term for case age may not be appropriate, and we have not accounted for alternative interactions there might be between age and length of stay in hospital.
 
@@
 Similar data are needed for the period preceding 2014.
-If detailed data, such as that provided for early 2014, are not available, aggregated data would be acceptable.
-However, it is unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
-As other studies have demonstrated, direct costs required to manage measles are not linear.
-
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively (Carabin, et al., 2002).
-This and other findings will be compared and contrasted with New Zealand costs, once more complete New Zealand data are made available.
+In the abscence of these data we have used mean case data from aggregated data available.
+It may be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014, but we have done so for these purposes.
+However, other studies have demonstrated direct costs required to manage measles are not linear.
+
+In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively  \citep{carabin2}.
+These values are substantially lower than our estimates, however, it is noticable that these estimates are also from more than a decade ago.
 The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
@@
 They estimated direct costs per case to be less than US\$500.
+Thus, our estimates of the additional costs, including indirect costs and wages lost may be appropriate.
+
 The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
@@
 The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
-
-The results presented here are based on available data, and only a 10 week period for the 2013--2014 outbreak.
-The majority of the cost analyses were discussed in our previous interim report and though some details have changed, we will focus here on the benefit--cost analyses.
-However, for all analyses more detailed costs of all aspects of hospitalisation would be beneficial.
-So too would information of the cost of measles vaccination in New Zealand.
-
-We have used standard discounting from healthcare of 3\% per annum \citep{honeycutt6}, but again there may be official New Zealand discounting rates that we are unaware of.
-For our benefit--cost analyses we have not included measles outbreak management costs, though these were provided for the period of January 1 -- March 9, 2014 (Table ~\ref{table:direct}).
-Including these could be done, but we would like some further data of these before including them in our benefit--cost analyses.
-It is likely to be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014.
-
-In other nations' outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively \citep{carabin2}.
-In our analyses if we include loss of earnings and hospitalisation costs, our estimates are higher, around US\$959 per case.
-This figure is from NZ\$ 852 for loss of wages and 0.17 * NZ \$ 1710 for costs related to attending hospital.
-However, costs can vary widely, with the containment of a single case and two secondary cases of measles in 2004 in Iowa, USA, estimated to cost US\$142,542 alone \citep{dayan5}.
-
-Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies, if we presume the $R_v$ is greater than 1 and measles will continue to circulate among the presently naive and susceptible New Zealand population until everyone has been infected.
+They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
+The work by \citep{zhou4} also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see Table~\ref{table:zhou}).
+However, the cost of measles in their analyses was substantially higher than mumps and rubella (Table~\ref{table:zhou}).
+
+\begin{table}[htdp]\small
+\begin{center}
+\begin{tabular}{lrrrr}
+\hline
+Parameter, disease & Direct costs & Indirect costs & Total costs\\
+\hline
+Cost of disease without vaccination program \\
+\hline
+Measles & \$2,645,779,861 & \$3,228,846,601 & \$5,874,626,462 \\
+Mumps & \$936,032,273 & \$522,974,252 & \$1,459,006,525 \\
+Rubella & \$88,352,366 & \$292,537,083 & \$380,889,449 \\
+Congenital rubella syndrome & \$114,726,378 & \$57,975,659 & \$172,702,037 \\
+Subtotal & \$3,784,890,878 & \$4,102,333,595 & \$7,887,224,473 \\
+\hline
+Cost of disease with vaccination program \\
+\hline
+Measles & \$647,488 & \$586,595 & \$1,234,083 \\
+Mumps & \$1,960,182  & \$1,124,249  & \$3,084,431 \\
+Rubella  & \$260,982  & \$258,240  & \$519,222 \\
+Congenital rubella syndrome  & \$2,662,760  & \$1,345,595  & \$4,008,355 \\
+Subtotal  & \$5,531,412  & \$3,314,679  & \$8,846,091 \\
+Costs averted by MMR vaccination program & \$3,779,359,466 & \$4,099,018,916 & \$7,878,378,382 \\
+\hline
+MMR vaccination program costs \\
+\hline
+Vaccine & \$147,802,803 &  & \$147,802,803 \\
+Administration  & \$94,756,438 & & \$94,756,438 \\
+Transportation  & \$7,164,227 &  & \$7,164,227 \\
+Parental productivity &   & \$34,787,438 & \$34,787,438 \\
+Adverse events  & \$16,413,145  & \$2,154,826  & \$18,567,971 \\
+Subtotal  & \$266,136,613  & \$36,942,264  & \$303,078,877 \\
+Net present value (net saving)  & \$3,513,222,853  & \$4,062,076,652  & \$7,575,299,505 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Summary of all measles, mumps, and rubella diseases and measles-mumps-rubella vaccination costs, USA, from \citep{zhou4}}
+\label{table:zhou}
+\end{table}%
+
+Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies.
+If we presume the $R_0$ estimated for measles in New Zealand is representative and following 28\% vaccination of the na\"{\i}ve measles will be unable to persist ($R_v$ < 1), our B/C ratio is substantially greater than 1.
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
-However, if we presume that in a totally naive population $R_0$ for measles is around 12, and sometimes estimated to be more than 18 (i.e.
-1 case infects 18 others on average, \citep{anderson91}) then the majority of naive people becoming infected is not unrealistic given some of our $R_v$ estimates (Figure~\ref{fig:r0}).
-The benefits of catch up vaccination are clear once $R_v$ is greater than one (Table~\ref{tab:cost_sim}).
-If measles continues to cause smaller outbreaks, and/or $R_v$ is less than one (Figure~\ref{fig:r0}), then there is little financial benefit in additional vaccination (Table~\ref{tab:cost_sim} and Table~\ref{tab:cost_ave}), though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
-
-Finally, our model of measles introductions is a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local naive population.
+However,  \citep{zhou4} estimate the B/C ratio to be much greater than one for MMR vaccination (see Table~\ref{table:zhou2}).
+As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis.
+
+Our results may be conservative.
+We presume that in a totally na\"{\i}ve population $R_0$ for measles was around 12.
+However, $R_0$ for measles has been estimated to be more than 18 (i.e.
+1 case infects 18 others on average, \citep{anderson91}).
+The benefits of catch up vaccination are clear once $R_v$ is greater than one (Tables~\ref{tab:cost20} and ~\ref{tab:cost20}, and previous interim report).
+However, as noted in our previous report, if measles $R_v$ in the currently na\"{\i}ve population is less than one, the benefits will not be so clear, though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
+
+\begin{table}[htdp]
+\tiny
+\begin{center}
+\begin{tabular}{lrrrr}
+\hline
+Parameter & B/C direct costs & B/C societal
+perspective & Costs per disc. yrs life saved\\
+\hline
+Base case & 14.2 & 26.0 & \$4195 \\
+Discount rate 0\% & 21.7 & 54.2 & \$1628 \\
+Discount rate 5\% & 12.3 & 19.6 & \$6569 \\
+Discount rate 8\% & 10.8 & 15.5 & \$10,615 \\
+\hline
+\end{tabular}
+\end{center}
+\caption{Benefit--cost ratios, USA, from \citep{zhou4}}
+\label{table:zhou2}
+\end{table}%
+
+Finally, our model of measles introductions remains a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local na\"{\i}ve population.
 The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{sec:epidemic_modelling}).
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis (see previous report) may produce some large and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis (see previous report) may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
+Given that, the greater the vaccination coverage, the smaller the outbreaks will be.
 
@@
 \begin{itemize}
-\item The mean wage losses per measles case is estimated to be \$852
+\item The mean wage losses per measles case is estimated to be \$839
 \item The mean cost per measles case attending hospital is estimated to be \$1,710
@@
 \item The mean management cost per measles case for the current outbreak, estimated from three months of data is estimated to be \$1,765
-\item Using $R_0$ values estimated for measles in New Zealand prior to vaccination and the current naive population sizes, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
+\item Using $R_0$ values estimated for measles in New Zealand prior to vaccination and the current na\"{\i}ve population sizes, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
 \item Outbreaks with a mean size of approximately 61 cases per year, but median of 2 cases per year may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.

### 2014-11-04 draftfinalreport.tex
Draft revisions

@@
 \item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
+\item Estimates of the proportion of the currently na\"{\i}ve New Zealand population (11\%) requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
 \item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
-However, New Zealander and immigrant or non--New Zealander travelling is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
-\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, as analysed from data from April, is well above one.
+However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this 2013--2014 outbreak, as analysed from data until June, is well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
-\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
-\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
-\item The mean wage losses per measles case is estimated to be \$852
+\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$805,000.
+\item The mean wage losses per measles case is estimated to be \$839
 \item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
-\item The mean management costs per case were \$1,765
+\item The mean management cost per case was \$1,765
 \item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>>$1).
-\item Large outbreaks, with a mean size of approximately around cases per year, median of 2 cases, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
+\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
 \end{itemize}
@@
 
-The significance of the different predictor variables can be seen in the ANOVA results:
+The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
 \par
 \vspace{5mm} %5mm vertical space
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Thu Oct 30 20:32:30 2014
+\begin{table}
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Mon Nov 03 20:46:33 2014
 \begin{tabular}{lrrrrrr}
@@
    \hline
-\end{tabular}\vspace{5mm} %5mm vertical space
-A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
+\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014}
+\label{table:anova}
+\end{table}
 \vspace{5mm} %5mm vertical space
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Thu Oct 30 20:32:30 2014
+A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.
+
+\vspace{5mm} %5mm vertical space
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Mon Nov 03 20:46:33 2014
 \begin{tabular}{rrrrr}
@@
    \hline
-\end{tabular}\vspace{5mm} %5mm vertical space
+\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+\label{table:regression}
+\end{table}
+\vspace{5mm} %5mm vertical space
 
 \par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
-
-People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (Figure~\ref{fig:ageinyears}).
+
+People of Pacific origin are also over-represented as measles cases ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:}).
 Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
@@
 
-The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
-However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
-Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
-The distribution of the numbers of naive and the total naive populations per DHB are given in \ref{fig:Northland}\ref{fig:Waitemata}\ref{fig:Auckland}\ref{fig:Counties_Manukau}\ref{fig:Waikato}\ref{fig:Lakes}\ref{fig:BayofPlenty}\ref{fig:Tairawhiti}\ref{fig:Taranaki}\ref{fig:HawkesBay}\ref{fig:Whanganui}\ref{fig:Midcentral}\ref{fig:Hutt}\ref{fig:CapitalandCoast}\ref{fig:Wairarapa}\ref{fig:NelsonMarlborough}\ref{fig:WestCoast}\ref{fig:Canterbury}\ref{fig:SouthCanterbury}\ref{fig:Southern}.
+The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
+However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
+Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
+The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in:
+\begin{itemize}[noitemsep,nolistsep]
+\item Northland: Figure~\ref{fig:Northland}
+\item Waitemata: Figure~\ref{fig:Waitemata}
+\item Auckland: Figure~\ref{fig:Auckland}
+\item Counties Manukau: Figure~\ref{fig:Counties_Manukau}
+\item Waikato: Figure~\ref{fig:Waikato}
+\item Lakes: Figure~\ref{fig:Lakes}
+\item Bay of Plenty: Figure~\ref{fig:BayofPlenty}
+\item Tairawhiti: Figure~\ref{fig:Tairawhiti}
+\item Taranaki: Figure~\ref{fig:Taranaki}
+\item Hawke's Bay: Figure~\ref{fig:HawkesBay}
+\item Whanganui: Figure~\ref{fig:Whanganui}
+\item Mid-Central: Figure~\ref{fig:Midcentral}
+\item Hutt: Figure~\ref{fig:Hutt}
+\item Capital and Coast: Figure~\ref{fig:CapitalandCoast}
+\item Wairarapa: Figure~\ref{fig:Wairarapa}
+\item Nelson Marlborough: Figure~\ref{fig:NelsonMarlborough}
+\item West Coast: Figure~\ref{fig:WestCoast}
+\item Canterbury: Figure~\ref{fig:Canterbury}
+\item South Canterbury: Figure~\ref{fig:SouthCanterbury}
+\item Southern: Figure~\ref{fig:Southern}
+\end{itemize}
 
@@
 \end{center}
-\caption{New Zealand population by age and estimated numbers of naive people in each age class}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (Table~\ref{table:sero})}
 \label{fig:naive}
@@
      \end{center}
-     \caption{Numbers of naive individuals per District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:naive_map}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Northland District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Northland}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Waitemata District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Waitemata}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Auckland District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Auckland}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Counties Manukau District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Counties_Manukau}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Waikato District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Waikato}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Lakes District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Lakes}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Bay of Plenty District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:BayofPlenty}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Tairawhiti District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Tairawhiti}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Taranaki District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Taranaki}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Hawke's Bay District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:HawkesBay}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Whanganui District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Whanganui}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Midcentral District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Midcentral}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Hutt District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Hutt}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Capital and Coast District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:CapitalandCoast}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Wairarapa District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Wairarapa}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Nelson Marlborough District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:NelsonMarlborough}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, West Coast District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:WestCoast}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Canterbury District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Canterbury}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, South Canterbury District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:SouthCanterbury}
@@
      \end{center}
-     \caption{Numbers of naive individuals per age class, Southern District Health Board}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (Table~\ref{table:sero})}
      \label{fig:Southern}
@@
 \begin{table}
-\caption{Measles incidence per million (2012)}
+\caption{Highest measles incidence per million (2012)}
 %latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{New Zealander travel numbers (2012)}
+\caption{New Zealander travel numbers by destination (2012)}
 %latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealand travellers in 2012}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by travellers numbers$*$measles incidence}
 %latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by travellers numbers$*$measles incidence}
 %latex.default(imrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
+\caption{Reported direct costs, with the number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
 %latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{sidewaystable}\small
-\caption{DHB and National level catch up vaccination rates and estimated outbreak sizes post vaccination}
+\caption{DHB and National level catch up vaccination rates and estimated outbreak sizes post vaccination. Size: DHB Population, Statistics NZ 2013;  	
+Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
+Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
+Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}
+
 %latex.default(vacp, file = "", table.env = FALSE, rowname = NULL)%
@@
 \includegraphics{draftfinalreport-060}
-     \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave})}
+     \caption{The estimated proportion of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost20} and ~\ref{tab:cost50})}
      \label{fig:numvac}
@@
 \includegraphics{draftfinalreport-062}
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
 The full distribution of results can be seen in Figure~\ref{fig:sim1}}
@@
 \includegraphics{draftfinalreport-063}
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
 A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
@@
 \begin{sidewaystable}\small
-\caption{Cost benefit analyses with 20 dollars per vaccine}
+\caption{Benefit--cost analyses with 20 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
 %latex.default(costtable20, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
-\hline
-Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$140.19$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$ 78.93$\tabularnewline
-Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$146.01$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$103.46$\tabularnewline
-Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$166.36$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$ 80.20$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$ 65.11$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$ 61.38$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$ 75.06$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$ 40.68$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$ 60.49$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$ 21.24$\tabularnewline
-Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$ 90.23$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$ 42.18$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$ 58.04$\tabularnewline
-Waikato&$11331$&$226620$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$110.30$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$ 20.85$\tabularnewline
-Waitemata&$17291$&$345820$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$150.07$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$ 22.89$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$ 36.86$\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc.costs}&\multicolumn{1}{c}{Wages.saved}&\multicolumn{1}{c}{Manage.saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp.saved}&\multicolumn{1}{c}{Costs.save}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
+\hline
+Auckland&$17920$&$358400$&$26142401$&$55010965$&$5297$&$9057921$&$79260619$&$ 82$&$209524$&$139.56$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7078643$&$14895456$&$1434$&$2452636$&$21461595$&$ 71$&$181417$&$ 78.58$\tabularnewline
+Canterbury&$13687$&$273740$&$20719105$&$43598824$&$4198$&$7178837$&$62817837$&$ 62$&$158420$&$145.36$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15440117$&$32490349$&$3129$&$5349752$&$46812580$&$ 96$&$245296$&$102.99$\tabularnewline
+Counties Manukau&$18880$&$377600$&$27605617$&$58089983$&$5594$&$9564902$&$83696914$&$ 50$&$127758$&$165.62$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5743794$&$12086558$&$1164$&$1990132$&$17414493$&$ 56$&$143089$&$ 79.84$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6574404$&$13834395$&$1332$&$2277925$&$19932803$&$ 86$&$219745$&$ 64.82$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4356088$&$ 9166434$&$ 883$&$1509314$&$13207136$&$ 62$&$158420$&$ 61.10$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7003972$&$14738327$&$1419$&$2426764$&$21235202$&$ 75$&$191638$&$ 74.72$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3700829$&$ 7787585$&$ 750$&$1282278$&$11220469$&$ 90$&$229965$&$ 40.49$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4772232$&$10042118$&$ 967$&$1653502$&$14468834$&$ 70$&$178862$&$ 60.22$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1407842$&$ 2962496$&$ 285$&$ 487795$&$ 4268408$&$ 72$&$183972$&$ 21.15$\tabularnewline
+Southern&$ 8371$&$167420$&$12681485$&$26685411$&$2570$&$4393931$&$38448739$&$102$&$260627$&$ 89.82$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2039609$&$ 4291911$&$ 413$&$ 706692$&$ 6183849$&$ 47$&$120093$&$ 41.99$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4414818$&$ 9290019$&$ 895$&$1529663$&$13385198$&$ 68$&$173752$&$ 57.76$\tabularnewline
+Waikato&$11331$&$226620$&$16988072$&$35747682$&$3442$&$5886094$&$51505793$&$ 95$&$242741$&$109.74$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1129294$&$ 2376352$&$ 229$&$ 391282$&$ 3423884$&$ 59$&$150755$&$ 20.73$\tabularnewline
+Waitemata&$17291$&$345820$&$25819386$&$54331250$&$5232$&$8946002$&$78281277$&$ 70$&$178862$&$149.20$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1061335$&$ 2233347$&$ 215$&$ 367736$&$ 3217840$&$ 50$&$127758$&$ 22.75$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2122670$&$ 4466695$&$ 430$&$ 735471$&$ 6435680$&$ 58$&$148200$&$ 36.62$\tabularnewline
 \hline
 \end{tabular}\end{center}\label{table:cost20}
+\begin{tablenotes}\footnotesize  
+        \item For numbers to vaccinate see Table~\ref{table:attack}
+        \item Proportion of cases hospitalised 0.17
+        \item Wage losses per case \$839 and cost per hospitalised case \$1710
+        \item Based on 10 introductions of measles, one per year
+      \end{tablenotes}  
 \end{sidewaystable}
 
+Vacc  Vacc.costs	Wages.saved	Manage.saved	Hospitalised	Hosp.saved	Costs.save	Outbreak	OB.costs	Benefit/cost
+
 \begin{sidewaystable}\small
-\caption{Cost benefit analyses with 50 dollars per vaccine}
+\caption{Benefit--costs analyses with 50 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
 %latex.default(costtable50, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
-\hline
-Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$72.02$\tabularnewline
-Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$52.49$\tabularnewline
-Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$74.87$\tabularnewline
-Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$61.20$\tabularnewline
-Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$78.44$\tabularnewline
-Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$52.91$\tabularnewline
-Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$45.59$\tabularnewline
-Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$43.82$\tabularnewline
-MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$50.42$\tabularnewline
-Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$32.41$\tabularnewline
-Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$43.72$\tabularnewline
-South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$18.75$\tabularnewline
-Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$56.86$\tabularnewline
-Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$33.03$\tabularnewline
-Taranaki&$ 2899$&$144950$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$42.20$\tabularnewline
-Waikato&$11331$&$566550$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$63.97$\tabularnewline
-Wairarapa&$  720$&$ 36000$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$18.43$\tabularnewline
-Waitemata&$17291$&$864550$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$75.46$\tabularnewline
-West Coast&$  685$&$ 34250$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$19.99$\tabularnewline
-Whanganui&$ 1378$&$ 68900$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$29.84$\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc.costs}&\multicolumn{1}{c}{Wages.saved}&\multicolumn{1}{c}{Manage.saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp.saved}&\multicolumn{1}{c}{Costs.save}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
+\hline
+Auckland&$17920$&$896000$&$26142401$&$55010965$&$5297$&$9057921$&$79260619$&$ 82$&$209524$&$71.70$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7078643$&$14895456$&$1434$&$2452636$&$21461595$&$ 71$&$181417$&$52.26$\tabularnewline
+Canterbury&$13687$&$684350$&$20719105$&$43598824$&$4198$&$7178837$&$62817837$&$ 62$&$158420$&$74.54$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15440117$&$32490349$&$3129$&$5349752$&$46812580$&$ 96$&$245296$&$60.93$\tabularnewline
+Counties Manukau&$18880$&$944000$&$27605617$&$58089983$&$5594$&$9564902$&$83696914$&$ 50$&$127758$&$78.09$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5743794$&$12086558$&$1164$&$1990132$&$17414493$&$ 56$&$143089$&$52.67$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6574404$&$13834395$&$1332$&$2277925$&$19932803$&$ 86$&$219745$&$45.39$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4356088$&$ 9166434$&$ 883$&$1509314$&$13207136$&$ 62$&$158420$&$43.63$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7003972$&$14738327$&$1419$&$2426764$&$21235202$&$ 75$&$191638$&$50.20$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3700829$&$ 7787585$&$ 750$&$1282278$&$11220469$&$ 90$&$229965$&$32.26$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4772232$&$10042118$&$ 967$&$1653502$&$14468834$&$ 70$&$178862$&$43.53$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1407842$&$ 2962496$&$ 285$&$ 487795$&$ 4268408$&$ 72$&$183972$&$18.67$\tabularnewline
+Southern&$ 8371$&$418550$&$12681485$&$26685411$&$2570$&$4393931$&$38448739$&$102$&$260627$&$56.61$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2039609$&$ 4291911$&$ 413$&$ 706692$&$ 6183849$&$ 47$&$120093$&$32.89$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4414818$&$ 9290019$&$ 895$&$1529663$&$13385198$&$ 68$&$173752$&$42.00$\tabularnewline
+Waikato&$11331$&$566550$&$16988072$&$35747682$&$3442$&$5886094$&$51505793$&$ 95$&$242741$&$63.64$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1129294$&$ 2376352$&$ 229$&$ 391282$&$ 3423884$&$ 59$&$150755$&$18.33$\tabularnewline
+Waitemata&$17291$&$864550$&$25819386$&$54331250$&$5232$&$8946002$&$78281277$&$ 70$&$178862$&$75.02$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1061335$&$ 2233347$&$ 215$&$ 367736$&$ 3217840$&$ 50$&$127758$&$19.86$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2122670$&$ 4466695$&$ 430$&$ 735471$&$ 6435680$&$ 58$&$148200$&$29.64$\tabularnewline
 \hline
 \end{tabular}\end{center}\label{table:cost50}
+\begin{tablenotes}\footnotesize  
+        \item For numbers to vaccinate see Table~\ref{table:attack}
+        \item Proportion of cases hospitalised 0.17
+        \item Wage losses per case \$839 and cost per hospitalised case \$1710
+        \item Based on 10 introductions of measles, one per year
+      \end{tablenotes}  
 \end{sidewaystable}
 
-The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\ref{table:vacp}.
-The benefit--cost results are in Table~\ref{tab:cost20} and Table~\ref{tab:cost50}, for two different vaccine prices.
-The results in the two tables show two alternative trends.
-In scenarios where we simulate the expected outbreaks in a homogeneously mixed population and $R_v$ is above 1 the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Table~\ref{tab:cost_sim}).
-However, if the previous recent history of cases since 1997 is the \emph{status quo} and what we may expect in the future then the \emph{additional} effort to vaccinate the currently naive population of New Zealand based on $R_v$ estimates that were greater than one is not a cost effective exercise (Table~\ref{tab:cost_ave}).
-
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
-
 For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$852 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
+Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
 We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).

### 2014-11-07 draftfinalreport.Rnw
revised text NF comments

--- 2014-11-02 draftfinalreport.Rnw
@@
 
-New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks of measles after measles introductions from abroad.
+New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks of measles following measles introductions from abroad.
 In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
@@
 Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007.
-We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level.
+We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, considering pre- and post-vaccination scenarios, based on the numbers of naive people at the District Health Board (DHB) and national level.
 We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
@@
 \item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
-\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
-\item Estimates of the proportion of the currently naive New Zealand population (11\%) requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
+\item The regression analyses suggest that age is a particularly strong risk factor for measles.
+However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis.
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
+People of European ethnicity, however, make up the majority of cases.
+\item Of the 11\% of the population of New Zealand that are currently considered to be immunologically naive, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
 \item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
-However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
-\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this 2013--2014 outbreak, as analysed from data until June, is well above one.
+However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
@@
 \item The mean management cost per case was \$1,765
-\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>>$1).
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1).
 \item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sec:}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 
-\section{Additional risk analyses}
+Thus in this report we include regression analyses that help understand risk factors for being measles cases in New zealand (Section~\ref{sub:risk_analyses}, descriptions of the New Zealand population in terms of vaccination coverage and immunity (Section~\ref{sub:immunestat}), and use those data to model the likely measles outbreak sizes in each Distric Health Board and the country (Section~\ref{sec:epidemic_modelling}.
+We estimate the costs of previous and current measles outbreaks (Section~\ref{sub:cost}) and uses these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch up vaccination programmes (Section~\ref{sub:cost_benefit}).
+This report includes further developments of the analyses provided in previous interim reports.
+
+\section{Risk analyses}
 \label{sub:risk_analyses}
@@
 \subsection{Risk of measles infection in New Zealand analyses methods}
-\label{sub:name}
 
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
+Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to adjust for confounding due to a lack of independence among risk factors.
+Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to adjust for confounding and test whether interactions among covariates provide additional information on risk in demographically--defined subgroups over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
 These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
-We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.
+We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of measles infection risk.
 
@@
 The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}).
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
                  data = testtable , FUN=sum)
+ttyr <- ttyr[ttyr$Ethnicity=="None",]
 @
@@
 @
-\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
-\label{table:case}
+\caption{Absolute number of measles cases in specific age and socio-economic deprivation categories from 2007-2014 with no ethnicity recorded}
+\label{table:none}
 \end{table}
@@
 # setwd("~/Massey_2014/measles/data")
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+                 data = data , FUN=sum)
+library(reshape)
+library(reshape2)
+library(plyr)
+library(pscl)
+
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+head(test)
+dim(test)
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+                   data = test , FUN=sum)
+caseyr
+test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
+test$AgeInYears<-as.factor(test$AgeInYears)
+tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
+test$AgeInYears<-tage
+#summary(test$EthnicityPrioritised)
+teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
+                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
+
+test$EthnicityPrioritised<-teth
+head(test)
+test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
+test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
+
+# want to set NZ Deprivation to the appropriate year
+
+testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
+                              ifelse (test$RptYear == "2008",test$NZDep06,
+                              ifelse (test$RptYear == "2009",test$NZDep06,
+                              ifelse (test$RptYear == "2010",test$NZDep06,
+                              ifelse (test$RptYear == "2011",test$NZDep06,
+                              ifelse (test$RptYear == "2012",test$NZDep06,
+                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
+head(testt)
+tail(testt)
+str(testt)
+testt$NZDep<-as.factor(testt$NZDep-1)
+
+testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
+                      ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
+                 data = testt , FUN=sum)
+dim(testtable)
+colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
+head(testtable)
+
+tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
+                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
+                                 "9"="6-10","10"="6-10"));
+
+testtable$NZDep<-tnzd
+# head(testtable)
+tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
+                 data = testtable , FUN=sum)
+
+ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
+                 data = testtable , FUN=sum)
 denom<-read.csv("NZDep2006Denominators.csv",header=T)
@@
 PerCapita<-(tp$cases/tp$Popn)
-PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita,4))
+PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita*10000,4))
 colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
@@
 Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
+However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category.
+Along with small population sizes (Table~\ref{table:percap}), there are several issues with estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11,height=9>>=
-par(mar=c(5,6,4,2)+0.1)
+<<fig=TRUE,echo=FALSE,width = 15,height=11>>=
+par(mar=c(12,6,4,2)+0.1)
 par(cex.axis=2)
-plot(tp$cases/tp$Popn,main="",ylab="Per Capita Rate",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="Category",cex.lab=2)
+plot(tp$cases/tp$Popn*10000,main="",ylab="Per Capita Rate Per 10,000",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
+     xlab="",cex.lab=2,xaxt="n")
+axis(side = 1, at = 1:50, labels=rep(c("0-2yr, 1-5","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),las=2,cex.axis=1.5)
+mtext(side=1,line=10,text="Age and NZDep",cex=2)
 legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
        col=c(1:4,6),pch=16,bty="n",cex=1.8)
-@
-\end{center}
-\caption{Per capita rates of measles infections for 2007--2014 broken down by ethnicity (Table~\ref{table:percap} for details)}
+par(mar = c(5,4,4,2) + 0.1)
+@
+\end{center}
+\caption{Per capita rates of measles infections for specific age and New Zealand deprivation indices (NZDep) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
 \label{fig:percap}
@@
 \footnotesize
-\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\caption{Numbers of measles cases, population sizes and per capita rates per 10,000 of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
 <<results=tex,echo=FALSE>>=
@@
 \subsection{Vaccination history of measles infection and population immunity estimation methods}
+\label{sub:immunestat}
 
@@
 par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Cases",main='',col='grey',breaks=20,ylab="",cex.lab=2)
+hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
 @
@@
 @
-\caption{Significance of different predictor variables for measles risk factors from 2007-2014}
+\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
 \label{table:anova}
@@
 
-People of Pacific origin are also over-represented as measles cases ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
+Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:}).
-Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+The very young (<2 years old) are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}).
+Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
 
@@
 The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
-However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
+However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the 3--5 year old age group to greater levels than in older young people  (Figure~\ref{fig:naive}).
 Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
-The distribution of the numbers of naive and the total naive populations per DHB, assuming national immunisation and immunity rates are representative, are given in Figures~\ref{fig:Northland},\ref{fig:Waitemata},\ref{fig:Auckland},\ref{fig:Counties_Manukau},\ref{fig:Waikato},\ref{fig:Lakes},\ref{fig:BayofPlenty},\ref{fig:Tairawhiti},\ref{fig:Taranaki},\ref{fig:HawkesBay},\ref{fig:Whanganui},\ref{fig:Midcentral},\ref{fig:Hutt},\ref{fig:CapitalandCoast},\ref{fig:Wairarapa},\ref{fig:NelsonMarlborough},\ref{fig:WestCoast},\ref{fig:Canterbury},\ref{fig:SouthCanterbury}, and \ref{fig:Southern}.
+The distribution of the numbers of naive and the total naive populations per DHB, assuming national immunisation and immunity rates are representative, are given in Figures~\ref{fig:Northland}, \ref{fig:Waitemata}, \ref{fig:Auckland}, \ref{fig:Counties_Manukau}, \ref{fig:Waikato}, \ref{fig:Lakes}, \ref{fig:BayofPlenty}, \ref{fig:Tairawhiti}, \ref{fig:Taranaki}, \ref{fig:HawkesBay}, \ref{fig:Whanganui}, \ref{fig:Midcentral}, \ref{fig:Hutt}, \ref{fig:CapitalandCoast}, \ref{fig:Wairarapa}, \ref{fig:NelsonMarlborough}, \ref{fig:WestCoast}, \ref{fig:Canterbury}, \ref{fig:SouthCanterbury}, and \ref{fig:Southern}.
 
@@
 risk[,2]<-round(risk[,2],0)
+toprisk2<-(toprisk[1:10,c(1,2,3,4)])
+toprisk2[,2]<-round(toprisk2[,2],0)
 @
@@
 
-Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}.
+\begin{table}
+\caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation.
+incidence: measles incidence per million; cover: percent vaccination cover; immigration: the number of travellers entering New Zealand from the country}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(toprisk2, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:toprisk}
+\end{table}
+
+Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}.
 However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}.
-Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}.
+Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
+The
+
+Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}.
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
@@
 \end{center}
-\caption{Measles incidence per million 2012}
+\caption{Measles incidence per million, 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
 \label{fig:incidence12}
@@
 \end{center}
-\caption{Measles vaccination cover (\%) 2012 }
+\caption{Measles vaccination cover (\%), 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
 \label{fig:cover12}
@@
 
-The distribution of the measles cases (\ref{fig:ageandvac}) appears to reflect the distribution of naive in the population (\ref{fig:naive}).
+The distribution of the measles cases (\ref{fig:ageandvac}) appears to reflect the distribution of naive individuals in the population (\ref{fig:naive}).
 The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number ($>16\%$) of cases had been vaccinated at least once.
@@
 
-The distribution of naive among the DHB also varies (Figures~\ref{fig:Northland}~\ref{fig:Waitemata}~\ref{fig:Auckland}~\ref{fig:Counties_Manukau}~\ref{fig:Waikato}~\ref{fig:Lakes}~\ref{fig:BayofPlenty}~\ref{fig:Tairawhiti}~\ref{fig:Taranaki}~\ref{fig:HawkesBay}~\ref{fig:Whanganui}~\ref{fig:Midcentral}~\ref{fig:Hutt}~\ref{fig:CapitalandCoast}~\ref{fig:Wairarapa}~\ref{fig:NelsonMarlborough}~\ref{fig:WestCoast}~\ref{fig:Canterbury}~\ref{fig:SouthCanterbury}~\ref{fig:Southern}).
+The distribution of naive among the DHBs also varies (Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 \subsection{Cost analyses methods}
+\label{sub:cost}
+
 Costs were evaluated as either direct or indirect.
@@
 \subsection{Cost analyses results}
-\label{sub:cost_benefit}
 
@@
 \subsection{Benefit--cost analyses methods}
+\label{sub:cost_benefit}
+
 To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.

### 2014-11-07 draftfinalreport.tex
revised text NF comments

--- 2014-10-30 draftfinalreport.tex
@@
 \item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
-\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
+\item Age is the best predictor of measles infection in multivariate regression analyses, with the highest rates overall in children aged 0-2.
+Other high risk sub groups were identified, including:  people of Pacific ethnicity; the less socially deprived; and European and Maori school age children.
+Individuals in these groups were more likely to have been measles cases in outbreaks since 2007.
+\item Of the 11\% of the population of New Zealand that are currently considered to be immunologically naive, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
 \item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
-However, New Zealander and immigrant or non--New Zealander travelling is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
-\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this year, 2014, as analysed from data from April, is well above one.
+However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
+\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
-\item Estimates of the proportion of the currently naive New Zealand population requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
 \item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$809,000.
@@
 \item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
-\item The mean management costs per case were \$1,765
-\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>>$1).
-\item Large outbreaks, with a mean size of approximately around cases per year, median of 2 cases, but peak size of up to many thousands, may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
+\item The mean management cost per case was \$1,765
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1).
+\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
 \end{itemize}
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles, thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 \subsection{Risk of measles infection in New Zealand analyses methods}
-\label{sub:name}
 
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically we required $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
+Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to adjust for confounding due to a lack of independence among risk factors.
+Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to adjust for confounding and test whether interactions among covariates provide additional information on risk in demographically--defined subgroups over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
 These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
-We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.
+We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of measles infection risk.
 
@@
 The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:case}).
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}).
 
@@
 \hline
-1-5&0-2&Asian&$ 11$\tabularnewline
-6-10&0-2&Asian&$  8$\tabularnewline
-1-5&3-5&Asian&$  1$\tabularnewline
-1-5&6-17&Asian&$ 11$\tabularnewline
-6-10&6-17&Asian&$  5$\tabularnewline
-1-5&18-24&Asian&$  3$\tabularnewline
-6-10&18-24&Asian&$  5$\tabularnewline
-1-5&25+&Asian&$ 10$\tabularnewline
-6-10&25+&Asian&$ 13$\tabularnewline
-1-5&0-2&European&$ 83$\tabularnewline
-6-10&0-2&European&$ 64$\tabularnewline
-1-5&3-5&European&$ 42$\tabularnewline
-6-10&3-5&European&$ 17$\tabularnewline
-1-5&6-17&European&$219$\tabularnewline
-6-10&6-17&European&$ 80$\tabularnewline
-1-5&18-24&European&$ 34$\tabularnewline
-6-10&18-24&European&$ 36$\tabularnewline
-1-5&25+&European&$ 78$\tabularnewline
-6-10&25+&European&$ 51$\tabularnewline
-1-5&0-2&Maori&$ 18$\tabularnewline
-6-10&0-2&Maori&$ 48$\tabularnewline
-1-5&3-5&Maori&$  7$\tabularnewline
-6-10&3-5&Maori&$ 11$\tabularnewline
-1-5&6-17&Maori&$ 19$\tabularnewline
-6-10&6-17&Maori&$ 92$\tabularnewline
-1-5&18-24&Maori&$  5$\tabularnewline
-6-10&18-24&Maori&$  8$\tabularnewline
-1-5&25+&Maori&$  2$\tabularnewline
-6-10&25+&Maori&$  6$\tabularnewline
-6-10&0-2&MLA&$  3$\tabularnewline
-6-10&3-5&MLA&$  1$\tabularnewline
-6-10&6-17&MLA&$  1$\tabularnewline
-6-10&18-24&MLA&$  2$\tabularnewline
-6-10&25+&MLA&$  6$\tabularnewline
-1-5&0-2&Pacific&$  5$\tabularnewline
-6-10&0-2&Pacific&$ 58$\tabularnewline
-6-10&3-5&Pacific&$  3$\tabularnewline
-1-5&6-17&Pacific&$  5$\tabularnewline
-6-10&6-17&Pacific&$ 22$\tabularnewline
-1-5&18-24&Pacific&$  1$\tabularnewline
-6-10&18-24&Pacific&$  8$\tabularnewline
-1-5&25+&Pacific&$  2$\tabularnewline
-6-10&25+&Pacific&$ 11$\tabularnewline
-1-5&0-2&None&$  3$\tabularnewline
-1-5&3-5&None&$  1$\tabularnewline
-1-5&6-17&None&$  3$\tabularnewline
-6-10&6-17&None&$  4$\tabularnewline
-1-5&18-24&None&$  2$\tabularnewline
-6-10&18-24&None&$  1$\tabularnewline
-1-5&25+&None&$  5$\tabularnewline
-6-10&25+&None&$  3$\tabularnewline
-\hline
-\end{tabular}\end{center}\caption{Absolute number of measles cases in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
-\label{table:case}
+1-5&0-2&None&$3$\tabularnewline
+1-5&3-5&None&$1$\tabularnewline
+1-5&6-17&None&$3$\tabularnewline
+6-10&6-17&None&$4$\tabularnewline
+1-5&18-24&None&$2$\tabularnewline
+6-10&18-24&None&$1$\tabularnewline
+1-5&25+&None&$5$\tabularnewline
+6-10&25+&None&$3$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Absolute number of measles cases in specific age and socio-economic deprivation categories from 2007-2014 with no ethnicity recorded}
+\label{table:none}
 \end{table}
@@
 \end{center}
-\caption{Per capita rates of measles infections broken down by ethnicity, see Table~\ref{table:percap} for details}
+\caption{Per capita rates of measles infections for specific age and New Zealand deprivation indices (NZDep) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
 \label{fig:percap}
@@
 \footnotesize
-\caption{Numbers of measles cases, population sizes and per capita rates of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\caption{Numbers of measles cases, population sizes and per capita rates per 10,000 of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
 %latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-1-5&0-2&Asian&$   6094$&$ 11$&$0.0018$\tabularnewline
-6-10&0-2&Asian&$   6806$&$  8$&$0.0012$\tabularnewline
-1-5&3-5&Asian&$   6094$&$  1$&$0.0002$\tabularnewline
-6-10&3-5&Asian&$   6806$&$  0$&$0.0000$\tabularnewline
-1-5&6-17&Asian&$  33918$&$ 11$&$0.0003$\tabularnewline
-6-10&6-17&Asian&$  28905$&$  5$&$0.0002$\tabularnewline
-1-5&18-24&Asian&$  22917$&$  3$&$0.0001$\tabularnewline
-6-10&18-24&Asian&$  34107$&$  5$&$0.0001$\tabularnewline
-1-5&25+&Asian&$  96357$&$ 10$&$0.0001$\tabularnewline
-6-10&25+&Asian&$  98715$&$ 13$&$0.0001$\tabularnewline
-1-5&0-2&European&$  57872$&$ 83$&$0.0014$\tabularnewline
-6-10&0-2&European&$  45445$&$ 64$&$0.0014$\tabularnewline
-1-5&3-5&European&$  57872$&$ 42$&$0.0007$\tabularnewline
-6-10&3-5&European&$  45445$&$ 17$&$0.0004$\tabularnewline
-1-5&6-17&European&$ 264330$&$219$&$0.0008$\tabularnewline
-6-10&6-17&European&$ 182937$&$ 80$&$0.0004$\tabularnewline
-1-5&18-24&European&$ 107649$&$ 34$&$0.0003$\tabularnewline
-6-10&18-24&European&$ 117840$&$ 36$&$0.0003$\tabularnewline
-1-5&25+&European&$1001916$&$ 78$&$0.0001$\tabularnewline
-6-10&25+&European&$ 724317$&$ 51$&$0.0001$\tabularnewline
-1-5&0-2&Maori&$  10003$&$ 18$&$0.0018$\tabularnewline
-6-10&0-2&Maori&$  30104$&$ 48$&$0.0016$\tabularnewline
-1-5&3-5&Maori&$  10003$&$  7$&$0.0007$\tabularnewline
-6-10&3-5&Maori&$  30104$&$ 11$&$0.0004$\tabularnewline
-1-5&6-17&Maori&$  40461$&$ 19$&$0.0005$\tabularnewline
-6-10&6-17&Maori&$ 116640$&$ 92$&$0.0008$\tabularnewline
-1-5&18-24&Maori&$  15360$&$  5$&$0.0003$\tabularnewline
-6-10&18-24&Maori&$  48495$&$  8$&$0.0002$\tabularnewline
-1-5&25+&Maori&$  71217$&$  2$&$0.0000$\tabularnewline
-6-10&25+&Maori&$ 192729$&$  6$&$0.0000$\tabularnewline
-1-5&0-2&MLA&$    728$&$  0$&$0.0000$\tabularnewline
-6-10&0-2&MLA&$   1290$&$  3$&$0.0023$\tabularnewline
-1-5&3-5&MLA&$    728$&$  0$&$0.0000$\tabularnewline
-6-10&3-5&MLA&$   1290$&$  1$&$0.0008$\tabularnewline
-1-5&6-17&MLA&$   2991$&$  0$&$0.0000$\tabularnewline
-6-10&6-17&MLA&$   4539$&$  1$&$0.0002$\tabularnewline
-1-5&18-24&MLA&$   1710$&$  0$&$0.0000$\tabularnewline
-6-10&18-24&MLA&$   3078$&$  2$&$0.0006$\tabularnewline
-1-5&25+&MLA&$   8028$&$  0$&$0.0000$\tabularnewline
-6-10&25+&MLA&$  10335$&$  6$&$0.0006$\tabularnewline
-1-5&0-2&Pacific&$   2093$&$  5$&$0.0024$\tabularnewline
-6-10&0-2&Pacific&$  13124$&$ 58$&$0.0044$\tabularnewline
-1-5&3-5&Pacific&$   2093$&$  0$&$0.0000$\tabularnewline
-6-10&3-5&Pacific&$  13124$&$  3$&$0.0002$\tabularnewline
-1-5&6-17&Pacific&$   8541$&$  5$&$0.0006$\tabularnewline
-6-10&6-17&Pacific&$  51183$&$ 22$&$0.0004$\tabularnewline
-1-5&18-24&Pacific&$   3972$&$  1$&$0.0003$\tabularnewline
-6-10&18-24&Pacific&$  22098$&$  8$&$0.0004$\tabularnewline
-1-5&25+&Pacific&$  18492$&$  2$&$0.0001$\tabularnewline
-6-10&25+&Pacific&$  91533$&$ 11$&$0.0001$\tabularnewline
+1-5&0-2&Asian&$   6094$&$ 11$&$18.0505$\tabularnewline
+6-10&0-2&Asian&$   6806$&$  8$&$11.7543$\tabularnewline
+1-5&3-5&Asian&$   6094$&$  1$&$ 1.6410$\tabularnewline
+6-10&3-5&Asian&$   6806$&$  0$&$ 0.0000$\tabularnewline
+1-5&6-17&Asian&$  33918$&$ 11$&$ 3.2431$\tabularnewline
+6-10&6-17&Asian&$  28905$&$  5$&$ 1.7298$\tabularnewline
+1-5&18-24&Asian&$  22917$&$  3$&$ 1.3091$\tabularnewline
+6-10&18-24&Asian&$  34107$&$  5$&$ 1.4660$\tabularnewline
+1-5&25+&Asian&$  96357$&$ 10$&$ 1.0378$\tabularnewline
+6-10&25+&Asian&$  98715$&$ 13$&$ 1.3169$\tabularnewline
+1-5&0-2&European&$  57872$&$ 83$&$14.3420$\tabularnewline
+6-10&0-2&European&$  45445$&$ 64$&$14.0830$\tabularnewline
+1-5&3-5&European&$  57872$&$ 42$&$ 7.2574$\tabularnewline
+6-10&3-5&European&$  45445$&$ 17$&$ 3.7408$\tabularnewline
+1-5&6-17&European&$ 264330$&$219$&$ 8.2851$\tabularnewline
+6-10&6-17&European&$ 182937$&$ 80$&$ 4.3731$\tabularnewline
+1-5&18-24&European&$ 107649$&$ 34$&$ 3.1584$\tabularnewline
+6-10&18-24&European&$ 117840$&$ 36$&$ 3.0550$\tabularnewline
+1-5&25+&European&$1001916$&$ 78$&$ 0.7785$\tabularnewline
+6-10&25+&European&$ 724317$&$ 51$&$ 0.7041$\tabularnewline
+1-5&0-2&Maori&$  10003$&$ 18$&$17.9946$\tabularnewline
+6-10&0-2&Maori&$  30104$&$ 48$&$15.9447$\tabularnewline
+1-5&3-5&Maori&$  10003$&$  7$&$ 6.9979$\tabularnewline
+6-10&3-5&Maori&$  30104$&$ 11$&$ 3.6540$\tabularnewline
+1-5&6-17&Maori&$  40461$&$ 19$&$ 4.6959$\tabularnewline
+6-10&6-17&Maori&$ 116640$&$ 92$&$ 7.8875$\tabularnewline
+1-5&18-24&Maori&$  15360$&$  5$&$ 3.2552$\tabularnewline
+6-10&18-24&Maori&$  48495$&$  8$&$ 1.6497$\tabularnewline
+1-5&25+&Maori&$  71217$&$  2$&$ 0.2808$\tabularnewline
+6-10&25+&Maori&$ 192729$&$  6$&$ 0.3113$\tabularnewline
+1-5&0-2&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
+6-10&0-2&MLA&$   1290$&$  3$&$23.2558$\tabularnewline
+1-5&3-5&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
+6-10&3-5&MLA&$   1290$&$  1$&$ 7.7519$\tabularnewline
+1-5&6-17&MLA&$   2991$&$  0$&$ 0.0000$\tabularnewline
+6-10&6-17&MLA&$   4539$&$  1$&$ 2.2031$\tabularnewline
+1-5&18-24&MLA&$   1710$&$  0$&$ 0.0000$\tabularnewline
+6-10&18-24&MLA&$   3078$&$  2$&$ 6.4977$\tabularnewline
+1-5&25+&MLA&$   8028$&$  0$&$ 0.0000$\tabularnewline
+6-10&25+&MLA&$  10335$&$  6$&$ 5.8055$\tabularnewline
+1-5&0-2&Pacific&$   2093$&$  5$&$23.8892$\tabularnewline
+6-10&0-2&Pacific&$  13124$&$ 58$&$44.1938$\tabularnewline
+1-5&3-5&Pacific&$   2093$&$  0$&$ 0.0000$\tabularnewline
+6-10&3-5&Pacific&$  13124$&$  3$&$ 2.2859$\tabularnewline
+1-5&6-17&Pacific&$   8541$&$  5$&$ 5.8541$\tabularnewline
+6-10&6-17&Pacific&$  51183$&$ 22$&$ 4.2983$\tabularnewline
+1-5&18-24&Pacific&$   3972$&$  1$&$ 2.5176$\tabularnewline
+6-10&18-24&Pacific&$  22098$&$  8$&$ 3.6202$\tabularnewline
+1-5&25+&Pacific&$  18492$&$  2$&$ 1.0815$\tabularnewline
+6-10&25+&Pacific&$  91533$&$ 11$&$ 1.2018$\tabularnewline
 \hline
@@
 
-\subsection{Vaccination history and measles infection}
-
-In this section we describe the vaccination history of the measles cases from 2007--2014.
-
-\subsection{Population immunity}
-
-In this section we describe the population immunity levels at both a national level and that for each DHB in New Zealand.
+\subsection{Vaccination history of measles infection and population immunity estimation methods}
+\label{sub:immunestat}
+
+In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR.
+We describe the population immunity levels at both a national level and that for each DHB in New Zealand.
 To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work.
@@
 
-The significance of the different predictor variables can be seen in the ANOVA results:
+The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
 \par
 \vspace{5mm} %5mm vertical space
+\begin{table}
 % latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Thu Oct 30 20:32:30 2014
+% Fri Nov 07 16:14:03 2014
 \begin{tabular}{lrrrrrr}
@@
    \hline
-\end{tabular}\vspace{5mm} %5mm vertical space
-A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen here:
+\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
+\label{table:anova}
+\end{table}
+\vspace{5mm} %5mm vertical space
+A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.
+
 \vspace{5mm} %5mm vertical space
 % latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Thu Oct 30 20:32:30 2014
+% Fri Nov 07 16:14:03 2014
 \begin{tabular}{rrrrr}
@@
    \hline
-\end{tabular}\vspace{5mm} %5mm vertical space
+\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+\label{table:regression}
+\end{table}
+\vspace{5mm} %5mm vertical space
 
 \par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood decreases with age (Figure~\ref{fig:ageinyears}).
-
-People of Pacific origin are also over-represented ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-In later outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figure~\ref{fig:ageinyears}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories.
-Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (Figure~\ref{fig:ageinyears}).
+
+The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
+Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young (<2 years old) are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}).
+Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
 
@@
 The majority of the naive among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
-However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
+However, the distribution of naive at a national level shows that the recent MMR vaccination schemes are reducing the proportions of naive in the 3--5 year old age group to greater levels than in older young people  (Figure~\ref{fig:naive}).
 Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of naive people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
-The distribution of the numbers of naive and the total naive populations per DHB are given in \ref{fig:Northland}\ref{fig:Waitemata}\ref{fig:Auckland}\ref{fig:Counties_Manukau}\ref{fig:Waikato}\ref{fig:Lakes}\ref{fig:BayofPlenty}\ref{fig:Tairawhiti}\ref{fig:Taranaki}\ref{fig:HawkesBay}\ref{fig:Whanganui}\ref{fig:Midcentral}\ref{fig:Hutt}\ref{fig:CapitalandCoast}\ref{fig:Wairarapa}\ref{fig:NelsonMarlborough}\ref{fig:WestCoast}\ref{fig:Canterbury}\ref{fig:SouthCanterbury}\ref{fig:Southern}.
+The distribution of the numbers of naive and the total naive populations per DHB, assuming national immunisation and immunity rates are representative, are given in Figures~\ref{fig:Northland}, \ref{fig:Waitemata}, \ref{fig:Auckland}, \ref{fig:Counties_Manukau}, \ref{fig:Waikato}, \ref{fig:Lakes}, \ref{fig:BayofPlenty}, \ref{fig:Tairawhiti}, \ref{fig:Taranaki}, \ref{fig:HawkesBay}, \ref{fig:Whanganui}, \ref{fig:Midcentral}, \ref{fig:Hutt}, \ref{fig:CapitalandCoast}, \ref{fig:Wairarapa}, \ref{fig:NelsonMarlborough}, \ref{fig:WestCoast}, \ref{fig:Canterbury}, \ref{fig:SouthCanterbury}, and \ref{fig:Southern}.
 
@@
 
-Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}.
+\begin{table}
+\caption{Measles and travel data for the top 10 countries identified as high risk for measles importation.
+incidence: measles incidence per million; cover: percent vaccination cover; immigration: the number of travellers entering New Zealand from the country}
+%latex.default(toprisk2, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}&\multicolumn{1}{c}{cover}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$  9$&$94$&$1799655$\tabularnewline
+United Kingdom&$ 33$&$93$&$ 401737$\tabularnewline
+Thailand&$ 78$&$98$&$  61358$\tabularnewline
+Malaysia&$ 64$&$95$&$  70366$\tabularnewline
+Indonesia&$ 63$&$84$&$  35352$\tabularnewline
+India&$ 15$&$85$&$ 107618$\tabularnewline
+China&$  4$&$99$&$ 322076$\tabularnewline
+Ukraine&$280$&$79$&$   2733$\tabularnewline
+Ireland&$ 23$&$92$&$  31396$\tabularnewline
+Romania&$342$&$94$&$   1913$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:toprisk}
+\end{table}
+
+Analyses of 2012 data, the most complete and current year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}.
 However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}.
-Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}.
+Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
+The
+
+Vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12}\ref{fig:incidence12}.
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
@@
 \begin{center}
-\includegraphics{draftfinalreport-036}
-\end{center}
-\caption{Measles incidence per million 2012}
+\includegraphics{draftfinalreport-037}
+\end{center}
+\caption{Measles incidence per million, 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
 \label{fig:incidence12}
@@
 \begin{center}
-\includegraphics{draftfinalreport-037}
-\end{center}
-\caption{Measles vaccination cover (\%) 2012 }
+\includegraphics{draftfinalreport-038}
+\end{center}
+\caption{Measles vaccination cover (\%), 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
 \label{fig:cover12}
@@
 \begin{center}
-\includegraphics{draftfinalreport-038}
+\includegraphics{draftfinalreport-039}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-039}
+\includegraphics{draftfinalreport-040}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-040}
+\includegraphics{draftfinalreport-041}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-041}
+\includegraphics{draftfinalreport-042}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-042}
+\includegraphics{draftfinalreport-043}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-043}
+\includegraphics{draftfinalreport-044}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-044}
+\includegraphics{draftfinalreport-045}
 \end{center}
@@
 
-The distribution of naive among the DHB also varies (Figures~\ref{fig:Northland}~\ref{fig:Waitemata}~\ref{fig:Auckland}~\ref{fig:Counties_Manukau}~\ref{fig:Waikato}~\ref{fig:Lakes}~\ref{fig:BayofPlenty}~\ref{fig:Tairawhiti}~\ref{fig:Taranaki}~\ref{fig:HawkesBay}~\ref{fig:Whanganui}~\ref{fig:Midcentral}~\ref{fig:Hutt}~\ref{fig:CapitalandCoast}~\ref{fig:Wairarapa}~\ref{fig:NelsonMarlborough}~\ref{fig:WestCoast}~\ref{fig:Canterbury}~\ref{fig:SouthCanterbury}~\ref{fig:Southern}).
+The distribution of naive among the DHB also varies (Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 \begin{center}
-\includegraphics{draftfinalreport-052}
+\includegraphics{draftfinalreport-053}
 \end{center}
@@
      \centering
-\includegraphics{draftfinalreport-060}
+\includegraphics{draftfinalreport-061}
      \caption{The estimated proportion of the currently naive New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost_sim} and ~\ref{tab:cost_ave})}
@@
      \centering
-\includegraphics{draftfinalreport-062}
+\includegraphics{draftfinalreport-063}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1.
@@
      \centering
-\includegraphics{draftfinalreport-063}
+\includegraphics{draftfinalreport-064}
      \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 11\% susceptible to measles infection using an $R_v$ =1, showing the rare but possible large epidemic sizes possible.
@@
 \hline
-Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$140.19$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$ 78.93$\tabularnewline
-Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$146.01$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$103.46$\tabularnewline
-Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$166.36$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$ 80.20$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$ 65.11$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$ 61.38$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$ 75.06$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$ 40.68$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$ 60.49$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$ 21.24$\tabularnewline
-Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$ 90.23$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$ 42.18$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$ 58.04$\tabularnewline
-Waikato&$11331$&$226620$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$110.30$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$ 20.85$\tabularnewline
-Waitemata&$17291$&$345820$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$150.07$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$ 22.89$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$ 36.86$\tabularnewline
+Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$32.45$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$11.31$\tabularnewline
+Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$33.96$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$17.66$\tabularnewline
+Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$50.79$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$11.62$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$ 8.76$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$ 8.08$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$10.62$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$ 4.80$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$ 7.86$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$ 2.31$\tabularnewline
+Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$13.92$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$ 5.06$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$ 7.49$\tabularnewline
+Waikato&$11331$&$226620$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$19.51$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$ 2.26$\tabularnewline
+Waitemata&$17291$&$345820$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$36.89$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$ 2.51$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$ 4.29$\tabularnewline
 \hline
@@
 \hline
-Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$72.02$\tabularnewline
-Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$52.49$\tabularnewline
-Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$74.87$\tabularnewline
-Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$61.20$\tabularnewline
-Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$78.44$\tabularnewline
-Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$52.91$\tabularnewline
-Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$45.59$\tabularnewline
-Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$43.82$\tabularnewline
-MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$50.42$\tabularnewline
-Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$32.41$\tabularnewline
-Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$43.72$\tabularnewline
-South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$18.75$\tabularnewline
-Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$56.86$\tabularnewline
-Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$33.03$\tabularnewline
-Taranaki&$ 2899$&$144950$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$42.20$\tabularnewline
-Waikato&$11331$&$566550$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$63.97$\tabularnewline
-Wairarapa&$  720$&$ 36000$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$18.43$\tabularnewline
-Waitemata&$17291$&$864550$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$75.46$\tabularnewline
-West Coast&$  685$&$ 34250$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$19.99$\tabularnewline
-Whanganui&$ 1378$&$ 68900$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$29.84$\tabularnewline
+Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$2095239$&$26.62$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$1814170$&$10.55$\tabularnewline
+Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$1584205$&$27.82$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$2452962$&$15.80$\tabularnewline
+Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$1277585$&$37.84$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$1430895$&$10.81$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$2197445$&$ 8.28$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$1584205$&$ 7.68$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$1916377$&$ 9.93$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$2299652$&$ 4.66$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$1788618$&$ 7.48$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$1839722$&$ 2.28$\tabularnewline
+Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$2606273$&$12.77$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$1200930$&$ 4.90$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$1737515$&$ 7.14$\tabularnewline
+Waikato&$11331$&$566550$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$2427411$&$17.29$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$1507550$&$ 2.23$\tabularnewline
+Waitemata&$17291$&$864550$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$1788618$&$29.68$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$1277585$&$ 2.47$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$1481998$&$ 4.18$\tabularnewline
 \hline

### 2014-11-07 dhbcensus.R

@@ -30,6 +30,7 @@ dhbpop<-read.csv("dhbcensus.csv",header=T)
 df.expanded <- dhbpop[rep(row.names(dhbpop),5), 1:21]
 df<-df.expanded[with(df.expanded, order(Age)), ]
 df<-df[,2:21]/5
+popdhb<-colSums(df)
 naivedhb<-df*(1-impop[1:90])
 head(naivedhb)
 matplot(naivedhb,type="l")
@@ -42,7 +43,7 @@ for (i in 1:20){
   pdf(paste("dhb", i, ".pdf", sep = ""))
   plot(naivedhb[,i],ylim=c(0,8000),main=colnames(naivedhb)[i],xlab="Age",ylab="Numbers",
        pch=16,cex=1.1)
-  legend("topright",legend=c("Total naive", round(sum(naivedhb[,i]))),bty="n")
+  legend("topright",legend=c(c("Total naive", round(sum(naivedhb[,i]))),c("Total population", signif(popdhb[i],5))),bty="n")
   dev.off()
 }  
 

### 2014-11-07 draftfinalreport.Rnw
edited final draft

--- 2014-11-04 draftfinalreport.Rnw
@@
 \usepackage{graphicx}
-\usepackage{caption2}
+%\usepackage{caption2}
 \usepackage{caption}
@@
 \usepackage{float}
+\usepackage{authblk}
 \newcommand{\vecA}{\mathbf{A}}
@@
 \newcommand{\Rc}{\mathcal{R}_c}
+\newcommand{\Pv}{\mathcal{P}_v}
 \newcommand{\Rr}{\mathcal{R}}
@@
 \newcommand{\Inf}{\mbox{Inf}\,}
-
+%\newcommand{\Ro}{R_0}
 \DeclareMathOperator{\var}{var}
@@
 \title{Draft Final Report: Measles risk assessment, modelling, and benefit--cost analyses for New Zealand}
-\author{{David T S Hayman, Tim Carpenter,\\ Jonathan C Marshall, Mick Roberts, Nigel P French.} \\\\
-{\small mEpiLab and EpiCentre,}\\
-{\small   Infectious Diseases Research Centre,}\\
-{\small   Massey University, Palmerston North 4442,New Zealand}\\
-{\small and}\\
-{\small Institute of Natural \& Mathematical Sciences,}\\
-{\small   New Zealand Institute for Advanced Study and the Infectious Disease Research Centre,}\\
-{\small   Massey University, Private Bag 102 904, North Shore Mail Centre, Auckland, New Zealand. } \\ 
-Email: \texttt{D.T.S.Hayman@massey.ac.nz}}
+\author[1]{David T S Hayman\thanks{D.T.S.Hayman@massey.ac.nz}}
+\author[1] {Jonathan C Marshall}
+\author[1] {Nigel P French}
+\author[2] {Tim Carpenter}
+\author[3] {Mick G Roberts}
+\affil[1] {$^{m}$EpiLab,Infectious Diseases Research Centre, Massey University, Palmerston North 4442,New Zealand}
+\affil[2] {EpiCentre,Infectious Diseases Research Centre, Massey University, Palmerston North 4442,New Zealand}
+\affil[3] {Institute of Natural \& Mathematical Sciences,New Zealand Institute for Advanced Study and the Infectious Disease Research Centre, Massey University, Private Bag 102 904, North Shore Mail Centre, Auckland, New Zealand}
 \date{}
@@
 
-New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks of measles after measles introductions from abroad.
+New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks after measles introductions from abroad.
 In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
-We identified additional analyses that may help understand risk of infection in New Zealand.
+We identify additional analyses that may help understand risk of infection in New Zealand.
 Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007.
@@
 We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
-Our key findings were:
+Our key findings are:
 
@@
 \item The \emph {Progress Towards Measles Elimination in New Zealand - Final} report was of high quality and contained substantial information and useful analyses.
-\item Age is the best predictor of risk of measles infection in multivariate regression analyses, though some groups, such as people of Pacific ethnicity, the less socially deprived, and European and Maori school age children have been more likely to be measles cases in outbreaks since 2007.
-\item Estimates of the proportion of the currently na\"{\i}ve New Zealand population (11\%) requiring additional vaccination to ensure measles does not persist is approximately 28\%, leading to an additional 131,500 vaccinations.
+\item The regression analyses suggest that age is a particularly strong risk factor for measles.
+However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis.
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
+People of European ethnicity, however, make up the majority of cases.
+\item Of the 11\% of the population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
 \item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
-However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during the winter, and non--New Zealanders summer.
-\item Analyses of outbreak data suggest that measles basic reproductive number ($R_v$, the number of secondary infections) values often include 1 and this 2013--2014 outbreak, as analysed from data until June, is well above one.
+However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
+\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
 \item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$805,000.
-\item The mean wage losses per measles case is estimated to be \$839
+\item The mean wage loss per measles case is estimated to be \$839
 \item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
-\item The mean management cost per case was \$1,765
-\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>>$1).
+\item The mean management cost per case is \$1,765
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1).
 \item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
@@
 As a member of the World Health Organization (WHO) Western Pacific Region, New Zealand is committed to work towards measles elimination, defined as the interruption of endemic (domestic) measles virus transmission, as achieved in the Americas in 2002.
-The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced that in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.
+The Western Pacific Region is expected to be the second WHO region to  achieve measles elimination and it was announced in March 2014 that Australia, Macao, Mongolia and the Republic of Korea have achieved measles elimination.
 
@@
 The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}).
-In 2013, prior to the 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
+In 2013, but prior to the current 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
 
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current overall population immunity estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sec:}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-Since 2009, all the outbreaks in New Zealand were linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
-Under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
+Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
+Those under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
 
@@
 
-Overall, the review was very thorough.
-The report included substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
+Overall, the review is very thorough.
+The report includes substantial background information on measles immunisation in New Zealand (\emph{Section 1.3}), the epidemiology of measles in New Zealand (\emph{Section 2}), the quality of epidemiological surveillance and laboratory testing for measles (\emph{Section 3}), and the levels of population immunity against the virus (\emph{Section 4}).
 Additional details are included for many aspects of measles epidemiology and control, not least regarding the recent MMR coverage rates by birth cohort in New Zealand (\emph{Section 4.2}) and the sustainability of the national immunisation programme (\emph{Section 5}).
@@
 Within the report there are many tables and figures which give considerable detail on the measles situation in New Zealand.
-Overall these were of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
-
-Specific epidemiological details were provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}).
+Overall these are of high quality, reporting both absolute measles case numbers and rates per 100,000 population in New Zealand.
+
+Specific epidemiological details are provided for the 2011--2012 outbreak including \emph{Figure 4}, the number and classification of measles notifications in New Zealand by month and year (2011 and 2012), with additional breakdown by age group in both years (\emph{Figure 5}) and per 100,000 population (\emph{Figures 6--8}).
 Similar presentation of the case data are provided for ethnicity (\emph{Figures 9--10}) and New Zealand Index of Deprivation (NZDep) (\emph{Figures 11--13}).
@@
 
-We agree with the report conclusions that testing for measles is performed appropriately within the required timeframe.
+We agree with the \emph {Progress Towards Measles Elimination in New Zealand - Final} report conclusions that testing for measles is performed appropriately within the required timeframe.
 Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
@@
 
-We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:risk_analyses}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:risk_analyses}), and current rates of immunity within the population (section~\ref{sub:risk_analyses}).
+We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:immunestat}), and current rates of immunity within the population (section~\ref{sub:immunestat}).
 In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
@@
 \subsection{Risk of measles infection in New Zealand analyses methods}
-\label{sub:name}
-
+\label{sub:regression}
 We received the raw EpiSurv measles case data from The Institute of Environmental Science and Research Ltd (ESR) on 27 June 2014.
-Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to avoid confounding results due to a lack of independence among risk factors.
-Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to test whether interactions among case covariates provide additional information on risk over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
+Initial analyses of those ESR data (not shown) suggested that denominator data were required to perform multivariate analyses to adjust for confounding due to a lack of independence among risk factors.
+Specifically $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data for New Zealand were required to adjust for confounding and test whether interactions among covariates provide additional information on risk in demographically--defined subgroups over the univariate analyses performed in the \emph{Progress Towards Measles Elimination in New Zealand - Final} report.
 These $\textsf{Age} \times \textsf{Prioritised Ethnicity} \times \textsf{NZDep}$ data were provided to us on 3 July 2014 by the University of Otago.
-We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of risk of measles infection.
+We used these denominator data to determine if there were interactions among specific age categories, Prioritized Ethnicities, and NZDep that might exist among cases allowing better understanding of measles infection risk.
 
@@
 par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="",cex.lab=2)
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
 hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1.8)
+legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
 @
@@
 
-This large number of categories, some with small population sizes, lead to both overdispersion and zeroinflation, as there were many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 and NZDep 6--10.
+This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases in, particularly in the adult age classes.
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 (least deprived) and NZDep 6--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}).
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
@@
 Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for Middle Eastern, Latin American and African ethnicities category, and along with small population sizes (Table~\ref{table:percap}), and there several issues with estimating the denominator data for this group (University of Otago, personal communication).
+However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
@@
 \begin{equation} \label{eq:reg}
- log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a * x_e)+ log(population)  + \epsilon
+ \log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a \times x_e)+ \log(population)  + \epsilon
   \end{equation}
-Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.
+Where $\alpha$ is the intercept, $y$ signifies cases, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity, $_n$ signifies NZDep, and $\epsilon$ is the error term.
 
@@
 \subsection{Vaccination history of measles infection and population immunity estimation methods}
+\label{sub:immunestat}
 
 In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR.
-We describe the population immunity levels at both a national level and that for each DHB in New Zealand.
+We describe the population immunity levels at both a national level and for each DHB in New Zealand.
 To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work.
-Specifically, we use the data in Table \ref{table:sero}, that uses the serosurvey results and the National Immunisation Register for under 6 year olds.
+Specifically, we use the data in Table \ref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.
 
@@
 
-For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel data by country and year to measure human movement to and from New Zealand  (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}).
+For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year, to measure human movement to and from New Zealand  (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}).
 We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
 Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
-We used the WHO data to determine per capita measles cases for each year and used these data and the number of immigrants to New Zealand to begin to understand where measles was likely to be imported from.
-We used simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
-We use the data from 2012 because this year had the most complete WHO measles data and yet was most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from.
+We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
+We use the data from 2012 because this year had the most complete WHO measles data and yet is most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
 
@@
 \end{center}
-\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the cases (Table~\ref{table:percap})}
+\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the case data (Table~\ref{table:percap})}
 \label{fig:predict}
@@
 
-People of Pacific origin are also over-represented as measles cases ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001), NZDep levels 6-10 under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
+Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young are still most likely to be infected, but of school aged children older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:}).
-Interestingly the regression results suggest risk of measles cases in the 6--17 year age category was greater for Europeans and Maori.
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}).
+Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
 
@@
 par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",cex.lab=2)
+barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
 legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
@@
 par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="")
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
 abline(v=12,col="blue")
@@
 The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
-However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve in the very young to greater levels than in older young people  (Figure~\ref{fig:naive}).
+However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (Figure~\ref{fig:naive}).
 Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
-The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in:
+The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures:
 \begin{itemize}[noitemsep,nolistsep]
@@
 \begin{table}
-\caption{Lowest national measles vaccine cover (\%, 2012)}
+\caption{Lowest national measles vaccine cover (\%), 2012}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by travellers numbers$*$measles incidence}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:imrisk12} and ~\ref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by travellers numbers$*$measles incidence}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
 <<results=tex,echo=FALSE>>=
@@
 
-Globally, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12} and \ref{fig:incidence12}.
-Analyses of 2012 data, the most complete and recent year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}.
-However, travel by New Zealanders is dominated by that to Australia (Table~\ref{table:nztravel12} and Figure~\ref{fig:nztravel12}).
-Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
-Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
-These data are shown in Table \ref{table:risk12}.
-The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
-The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and \ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and \ref{fig:imrisk12}
+\begin{table}
+\caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation.
+incidence: measles incidence per million; cover: percent vaccination cover (\%); immigration: the number of travellers entering New Zealand from the country}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(toprisk2, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:toprisk}
+\end{table}
+
+Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (Tables~\ref{table:incidence12} and ~\ref{table:cover12}, and Figures~\ref{fig:incidence12} and ~\ref{fig:cover12}).
+Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}.
+However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}.
+Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation.
+Though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
+These data are shown in Table~\ref{table:risk12}.
+The data for all the variables for each nation state and territories for 2012 are plotted in Figures~\ref{fig:immigration12}, ~\ref{fig:incidence12}, and ~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
+The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.
+
+The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in Table~\ref{table:toprisk}.
 
@@
 \begin{figure}
-     \begin{center}
-     \includegraphics[width=1\textwidth]{nzers.pdf}
-     \end{center}
-     \caption{Trends in international travel}
+     \centering
+     \includegraphics[width=0.9\textwidth]{nzers.pdf}
+     \caption{Trends in international travel, determined by immigration arrival data and New Zealander travel departure destinations}
      \label{fig:travel}
@@
 
-Though global incidence of measles in declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:travel}).
+Though the global incidence of measles is declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:travel}).
 This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
@@
 However, our analyses also highlight other groups that are at greater risk of measles.
-In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
+In particular, Pacific people are at greater risk, as are the less deprived (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
 Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses.
-Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the National Immunisation Register (NIR).
+Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the NIR.
 A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data.
-The data suggest that census data and NIR data re recording children living in different DHB to that which they are vaccinated in, or subsequent to the census or NIR data moving.
-The leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated.
+The data suggest that census data and NIR data are recording children living in different DHBs to that in which they are vaccinated, or subsequent to the census or NIR data moving.
+This leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated.
 Another issue was how to deal with people of greater age than those recorded in the NIR.
@@
 
-The distribution of the measles cases (Figure~\ref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve in the population (Figure~\ref{fig:naive}).
-The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number of cases ($>16\%$) had been vaccinated at least once (Figure~\ref{fig:ageandvac}).
-However, the majority of these vaccinated cases received only one vaccination, and were vaccinated when they were young (Figure~\ref{fig:vaccstat}).
-
+The distribution of the measles cases (Figure~\ref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (Figure~\ref{fig:naive}).
+The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number ($>16\%$) of cases had been vaccinated at least once.
+However, the majority of these received only one vaccination, and were vaccinated when they were young (\ref{fig:vaccstat}).
 The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (Figure~\ref{fig:naive_map}).
-Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than District Health Board (DHB)) immunisation coverage data were not able to provide reliable results, as discussed above.
-The National Immunisation Register (NIR) data, but frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census.
-Thus the data gap that we have hinder us providing fine scale risk maps.
-
-The distribution of na\"{\i}ve among the DHB also varies.
+
+Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than DHB) immunisation coverage data are not able to provide reliable results, as discussed above.
+Frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census.
+Given these data gaps we are unable to provide finer scale risk maps.
+
+The distribution of na\"{\i}ve among the DHBs also varies (Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 
-Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries most travel to New Zealand is coming from.
+Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand.
 However, the strong seasonality in travel around Christmas suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
 The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}).
-However, recently Australia was declare free of endemic measles and so the 2012 data must reflect imported measles there.
-However, other areas of high risk include the United Kingdom, where measles persists, South, Southeast and East Asia, where measles is endemic (Table~\ref{table:risk12} and Figure~\ref{fig:risk12}).
+However, recently Australia was declared free of endemic measles and so the 2012 data must reflect imported measles there.
+However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (Table~\ref{table:risk12} and Figure~\ref{fig:risk12}).
 
@@
 \item The majority of cases are unvaccinated.
-\item The majority of vaccines failures occur in those people which revieved single vaccinations around 1 year old.
-\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHB with larger urban areas.
+\item The majority of vaccine failures occur in those people which recieved single vaccinations around 1 year old.
+\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas.
 \item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
@@
 The original mathematical model for the dynamics of measles in New Zealand prepared in 1996 \citep{tobias98} successfully predicted the 1997 epidemic, which was curtailed by a mass vaccination campaign \citep{mansoor98,roberts0}.
-Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics.
-The model developed by \citep{roberts0} supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
-The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a} and later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintaining high coverage rates in order to prevent future epidemics \citep{roberts4}.
-
-These results were comparable to others, for example: \citep{babad95} suggested two-dose schedule for England and Wales, with the second vaccination given at age four; and \citep{gay98} recommended a second vaccination at either 18 months or five years, to complement the first vaccination at 12 months in Canada.
-In addition, \citep{agur93} found that vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals would prevent epidemics in Israel.
+Subsequent extension of this work in 1998 showed that the then current schedule of MMR1 at 15 months and MMR2 at 11 years was insufficient to prevent further epidemics \citep{roberts0}.
+The model supported the change in the immunisation schedule that took effect in January 2001, at which time MMR2 was changed from delivery at 11 years to delivery before the age of five.
+The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a}.
+Later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintain high coverage rates in order to prevent future epidemics \citep{roberts4}.
+
+These results were comparable to others, for example: a two-dose schedule for England and Wales, with the second vaccination given at age four \citep{babad95}; and a second vaccination at either 18 months or five years were recommened \citep{gay98} to complement the first vaccination at 12 months in Canada.
+In addition, vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was found to predict the prevention of measles epidemics in Israel \citep{agur93}.
 All modelling studies agree that two vaccinations at no less than five years apart are necessary to prevent measles epidemics.
-\citep{wallinga1} took existing policies in eight European countries and estimated the coverage rates required to reduce $R_v$ below one and eliminate endemic measles.
+Existing policies in eight European countries were analysed and researchers estimated the coverage rates required to reduce $R_v$ below one and eliminate endemic measles \citep{wallinga1}.
 They found that results depended on the age at delivery, but no strategy succeeded if coverage rates were below approximately 87\%.
 
-Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear differential equation (ODE) models have reached similar conclusions.
+Numerous models for measles vaccination strategies for various regions \citep{agur93, babad95, edmunds0, gay98, wallinga1} based on sets of nonlinear ordinary differential equation (ODE) models have reached similar conclusions.
 The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, \citep{glass4} in the Netherlands showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, in the Netherlands analyses showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate \citep{glass4}.
 
 The quantity that determines whether an epidemic will occur is the basic reproduction number of the infection, $R_0$.
-This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann0}.
+This is defined as the expected number of secondary infections that would arise from a single primary infection introduced into a fully susceptible population \citep{anderson91, diekmann13}.
 If $R_0 > 1$ an epidemic will occur following an introduction of infection.
@@
 To do this we estimated $R_t$, following an adaptation of the methods in \citep{obidia12,wallinga4}.
-We were required to compute the generation time for measles to do so.
+We are required to compute the generation time for measles to do so.
 The generation time is the average time an index case infects others after becoming infected.
@@
      \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
-     \caption{Measles data classified as outbreaks for reproductive number of the infection ($R_v$) estimation from 2009 to 2014}
+     \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014}
      \label{fig:outbreaks}
@@
      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ (Average $R_0$) for the outbreaks each year, as classified by outbreaks in Figure~\ref{fig:outbreaks}. Sporadic cases are excluded.}
+     \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in Figure~\ref{fig:outbreaks}. Sporadic cases are excluded.}
      \label{fig:r0}
@@
 
-To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we use the well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population is \citep{diekmann13}
+To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we use the well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population \citep{diekmann13}:
+
 \begin{equation} \label{eq:eqn1}
@@
 
-In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks is
+In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks, $\Pv$ is
 
 \begin{equation} \label{eq:prop}
-$x_0-1/\Ro$.
+\Pv = x_0-1/\Ro
 \end{equation}
 
-These formulae were applied at a District Health Board (DHB) level, assuming no mixing between DHBs.
+These formulae are applied at a District Health Board (DHB) level, assuming no mixing between DHBs.
 
@@
 Of particular note is the ongoing outbreak (as of 12 June 2014), which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
-An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_0$ is necessarily over one, and so the comparison with others must be cautious.
+An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
 
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain the level of immunisation to high enough levels that measles does not become endemic.
+Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
 The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given it's $R_v$ is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproductive numbers may be lower than estimated.
+This implies that the current outbreak may persist within the population for a substantial period, given its $R_v$ is approximately twice that of the 2011-2012 outbreak.
+A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproduction numbers may be lower than estimated.
 However, sub-clinical and underreporting may lower the estimate.
@@
 If fewer cases occur during each outbreak, then this suggests that $R_v$ is lower, but that measles is introduced to New Zealand more frequently.
-The regularity of these outbreaks, however, imply that the regular (approximately yearly) importation of measles is an ongoing process.
+The regularity of these outbreaks, however, implies that the approximately yearly importation of measles is an ongoing process.
 Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we include the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
@@
 The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (Table~\ref{table:attack}).
-These additional vaccination numbers can be made up in a number of different ways, and these are discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
 However, they require differing numbers of vaccinations per DHB.
@@
 Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% na\"{\i}ve.
-Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate.
-Future analyses at a smaller, more local level would be useful, but the lack of appropriate data matching between the NIR data and the census unit area data prevented these.
+Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult.
+Future analyses at a smaller, more local level would be useful, but the lack of appropriate data matching between the NIR data and the census unit area data currently prevent these.
 
@@
 
-Approximately 50 years ago, approximately 135 million cases and 7--8 million deaths were believed to occur in the world due to measles \citep{clements4}.
+About 50 years ago, approximately 135 million cases of measles causing 7--8 million deaths were believed to occur in the world \citep{clements4}.
 Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities.
 \citep{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
-Similarly, the number of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
+Similarly, the incidence of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
 They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination.
@@
 
-The annual cost of treating and controlling measles in 11 industrialised countries was estimated to cost more than US\$150 million \citep{carabin3}.
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to be more than US\$150 million \citep{carabin3}.
 The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}.
@@
 In 2001, it reported losses related to measles of approximately US\$50 million.
-The economic impact of a large measles outbreak in Italy, 2002--03 examined the costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis.
+The economic impact of a large measles outbreak in Italy, 2002--03, has been evaluated.
+The costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis was calculated.
 The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
@@
 Germany implemented a two-dose measles vaccination program in 1991 and has seen the benefits in recent years.
-In 2001 more than 6,000 cases were reported in Germany but by 2004 this number fell to 122 \citep{wichmann9}.
+In 2001 more than 6,000 cases were reported, but by 2004 this number fell to 122 \citep{wichmann9}.
 However, in 2005 more than 500 cases were reported by the middle of the year in two German states, with the vast majority (>95\%) in non-vaccinated children \citep{siedler6}.
@@
 Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case.
-In addition, to determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
-
-Although measles was declared eliminated from the United States in 2000, it remains a concern due to the endemic nature of it around the world \citep{parker6}.
-Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported measles.
+To determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
+
+Although measles was declared eliminated from the United States in 2000, it remains a concern due to its endemic nature around the world \citep{parker6}.
+Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported cases.
 \citep{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
@@
 Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak.
-Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overhead.
+Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overheads.
 This amounted to a direct cost of \$4,932 per measles case.
@@
 The cost of hospitalisation for the 3-day stay by the index case was US\$931.
-Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents' exposed to measles (US\$240) for a total of US\$23,816.
+Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents exposed to measles (US\$240) for a total of US\$23,816.
 
@@
 \subsection{Cost analyses methods}
-Costs were evaluated as either direct or indirect.
+\label{sub:cost}
+
+Costs are evaluated as either direct or indirect.
 Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
@@
 The HVA measures the potential future earnings of an individual and discounts it into a present value.
-Typically this is 3\% but 5\% has also been used in a sensitivity analysis, which is more compared to non-human life calculations and will tend to reduce the present value of the future earnings (saved by avoiding a case).
-
-Data for the current measles outbreak were obtained from the New Zealand Ministry of Health, from 2008 through June 2014.
-Data included information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, case weight and associated cost.
+Typically this is 3\% but 5\% has also been used in a sensitivity analysis, which will tend to reduce the present value of the future earnings (saved by avoiding a case).
+
+Data to estimate the costs relating to the current measles outbreak were obtained from the New Zealand Ministry of Health, for the period 2008 through June 2014.
+Data included information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, and associated cost.
 
@@
 Data, for the period January 1 - March 9, 2014, reported salaries for people involved with the measles outbreak management medical team.
-The costs were reported as direct, additional (above normal budgeting) costs required to enable the management of measles.
+The costs are reported as direct, additional (above normal budgeting) costs required to enable the management of measles.
 It includes a breakdown by individual performing the work and whether it was during the normal work schedule (Monday to Friday, M-F) or weekends.
-Normal work was calculated as $1.2 \times \textsf{full time equivalent (FTE)} \times \textsf{number of days worked}$.
-Overtime was calculated as $1.6\times\textsf{FTE}$ (M-F) and $2.0 \times \textsf{FTE}$ (weekend).
-A full day was considered as 8 hours worked.
-Salary (hourly) rates were calculated for the following: public health nurse (PHN, \$36), public health assistant (PHA, \$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), incident management team (IMT), which had the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and safety/security officer (SSO) (\$26).
-In addition, measles operations personnel were calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
-
-Wages lost due to measles were calculated for the period January 2008 - August 2014.
-Calculations were based on the assumptions that 5 days of work were lost for each case; however, individuals between 0-14 years of age were not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15-19 years of age was 41.9\%  (Statistics New Zealand, 2013).
-If the case were < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
+Normal work is calculated as $1.2 \times \textsf{full time equivalent (FTE)} \times \textsf{number of days worked}$.
+Overtime is calculated as $1.6\times\textsf{FTE}$ (M-F) and $2.0 \times \textsf{FTE}$ (weekend).
+A full day is considered as 8 hours worked.
+Salary (hourly) rates are calculated for the following: public health nurse (PHN, \$36), public health assistant (PHA, \$22), data support (\$26), data support (temporary) (\$33), management and programme supervisors (\$40), incident management team (IMT), which have the following work titles: incident controller (\$96), administrator (\$24), planning and intel (\$40), logistics (\$36), communications (\$45), informatics (\$40), operations (\$40), and safety/security officer (SSO) (\$26).
+In addition, measles operations personnel are calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
+
+Wages lost due to measles are calculated for the period January 2008 - August 2014.
+Calculations are based on the assumptions that 5 days of work are lost for each case; however, individuals between 0-14 years of age are not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15-19 years of age was 41.9\%  (Statistics New Zealand, 2013).
+If the case is < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
 Total wage lost for the 247 cases and care givers was estimated to be \$207,155.
@@
 \begin{table}
-\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014.
+PHN: public health nurse; PHA: public health assistant; IMT: incident management team; SSO: safety/security officer (SSO)}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Reported direct costs, with the number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
+\caption{Reported direct costs in NZ\$, with the number of cases, length of hospital day, total cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, 2000--2014}
 <<results=tex,echo=FALSE>>=
@@
 library(Hmisc)
+colnames(costtable2) <- c("Year", "Cases", "Days", "Cost","Cost per case","Cost per day")
 latex(costtable2, file="", table.env=FALSE,rowname=NULL)
@@
 \begin{table}\small
-\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
+\caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\ref{table:regression})}
 <<results=tex,echo=FALSE>>=
@@
 library(Hmisc)
+colnames(costtable4) <- c("Year", "Gender", "Cost","Cases","Length of stay","Cost per case")
 latex(costtable4, file="", table.env=FALSE,rowname=NULL)
@@
 
-Regression analyses showed statistically significant associations between cost of hospitalisation and three variables, length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
+Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
 Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.
-The cost was \$256 less if the case were a male.
+The cost was \$256 less if the case was a male.
 It increased of approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014.
@@
 
-Wages lost due to measles were calculated for the period January 2008 -- August 2014.
-Calculations were based on the assumption that 5 days of work were lost for each case; however, individuals under 15 years of age were not assumed to be employed and therefore did not suffer an income loss.
-If the case were less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
+Wages lost due to measles are calculated for the period January 2008 -- August 2014.
+Calculations are based on the assumption that 5 days of work are lost for each case; however, individuals under 15 years of age are not assumed to be employed and therefore did not suffer an income loss.
+If the case are less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
 Total wage lost for the 247 cases and care givers was estimated to be \$210,436.
@@
 \subsection{Benefit--cost analyses methods}
-To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we did several things.
+\label{sub:cost_benefit}
+
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we carried out a number of analyses.
 Primarily, we used Equation~\ref{equation:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{sec:epidemic_modelling}).
-We assumed that the cases prevented by this was proportional to the outbreak size.
+
+We assume that the number of cases prevented by this is proportional to the outbreak size.
 However, there is a continued risk of introduction and despite $R_v$ being <1, smaller outbreaks may occur.
 Thus, we simulated expected outbreak sizes with $R_v$ < 1.
-We used these values of numbers of predicted cases prevented and numbers expected despite additional vaccination to calculate the savings.
-The cost figures above were used to estimate what the savings for vaccinating additional populations was, using estimated per case costs saved.
-The costs of the catch up vaccination schemes were estimated to be two differnt values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
-Values of \$20 and \$50 per vaccine were used.
-Thus the costs of the expected measles-related costs due to constant introduction of measles despite increased population immunity and the vaccination schemes were used to estimate the costs of additional vaccine programs and costs from continued measles introductions.
-The financial savings from reduced measles cases and these costs were used to work out the estimated B/C ratio.
-A B/C ratio > 1 means that the program benefits exceed their costs.
-A B/C value less than one suggests the costs are higher than the economic benefits.
-Lastly, benefits were assessed over a 10--year time period, using a discounting rate of 3\% discount per year for the costs, as is common for healthcare discounting \citep{honeycutt6}.
-Thus:
+We use these values of numbers of predicted cases prevented and numbers expected despite additional vaccination to calculate the savings.
+
+The cost figures above are used to estimate what the savings for vaccinating additional populations are, using estimated per case costs saved.
+The costs of the catch up vaccination schemes are estimated to be two different values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
+Values of \$20 and \$50 per vaccine are used.
+Thus the expected measles-related costs due to constant introduction of measles despite increased population immunity and the costs of the vaccination schemes are used to estimate the costs of additional vaccine programs for the B/C analysis.
+The financial savings from reduced measles cases are used to work out the benefits for the additional vaccination programs.
+The costs for the catch up vaccination programs are presumed to be over a single year.
+The benefits for the cases saved are presumed to be over a ten year period.
+Because benefits are assessed over a 10--year time period, a discounting rate of 3\% discount per year for the costs saved was used, as is common for healthcare discounting \citep{honeycutt6}.
+Measles introductions are expected to be annually occurring events, though the costs per year for each event was also discounted.
+Thus, the B/C ration was:
 
@@
 Where B is the benefit in saved funds from cases prevented and C is the cost of vaccination, discounted over time, where $n$ is the year and $r$ the annual discount rate of 0.03 (3\%).
+A B/C ratio > 1 means that the program benefits exceed their costs. A B/C value less than one suggests the costs are higher than the economic benefits.
 
@@
 
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using the equation~\ref{eq:prop} are shown in Figure~\ref{fig:numvac}.
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using equation~\ref{eq:prop} are shown in Figure~\ref{fig:numvac}.
 
@@
 @
-     \caption{The estimated proportion of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost20} and ~\ref{tab:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
      \label{fig:numvac}
@@
 
-The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ were <1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
-These simulations show that even in scenarios when $R_v$ is lower than one, and thus stochastically should fail to persist (i.e.
+The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ is <1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
+These simulations show that even in scenarios when $R_v$ is lower than one, and thus deterministically should fail to persist (i.e.
 become endemic), outbreaks can occur due to stochastic processes following measles importation.
@@
 The maximum predicted number of cases was nearly 8,000 cases.
-Note, however, that larger values were very rarely predicted and this does not take into account spatial variation or heterogeneous contact rates.
+Note, however, that larger values are very rarely predicted and this does not take into account spatial variation or heterogeneous contact rates.
 
@@
 
-Vacc  Vacc.costs	Wages.saved	Manage.saved	Hospitalised	Hosp.saved	Costs.save	Outbreak	OB.costs	Benefit/cost
-
 <<echo=FALSE>>=
@@
 library(Hmisc)
+
 latex(costtable50, file="", table.env=FALSE,rowname=NULL)
@@
 
-Our estimates of the costs of the measles outbreaks in New Zealand suggest that measles management per case in New Zealand is high.
+Our estimates of the costs of the measles outbreaks in New Zealand suggest that measles management costs per case in New Zealand is high.
 We used the mean values per case in the absence of alternative data and used as much data as possible.
@@
 The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
-In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak (Dayan et al., 2005).
+In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak \citep{dayan5}.
 They estimated direct costs per case to be less than US\$500.
@@
 
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively \citep{prouty1}.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively in Minnesota \citep{prouty1}.
 In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
@@
 
-\begin{table}[htdp]\small
+\begin{sidewaystable}[htdp]\small
 \begin{center}
@@
 \label{table:zhou}
-\end{table}%
+\end{sidewaystable}%
 
 Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies.
-If we presume the $R_0$ estimated for measles in New Zealand is representative and following 28\% vaccination of the na\"{\i}ve measles will be unable to persist ($R_v$ < 1), our B/C ratio is substantially greater than 1.
+If we presume the $R_0$ estimated for measles in New Zealand is representative and following 28\% vaccination of the na\"{\i}ve population, measles will be unable to persist ($R_v$ < 1), our B/C ratio is substantially greater than 1.
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
@@
 However, $R_0$ for measles has been estimated to be more than 18 (i.e.
-1 case infects 18 others on average, \citep{anderson91}).
-The benefits of catch up vaccination are clear once $R_v$ is greater than one (Tables~\ref{tab:cost20} and ~\ref{tab:cost20}, and previous interim report).
+1 case infects 18 others on average, \citep{anderson91}) in other countries.
+It is unlikely $R_0$ was so high as 18 in New Zealand, as measles would still be endemic given current vaccination rates.
+However, the benefits of catch up vaccination are clear once $R_v$ is greater than one (Tables~\ref{table:cost20} and ~\ref{table:cost20}, and previous interim report).
 However, as noted in our previous report, if measles $R_v$ in the currently na\"{\i}ve population is less than one, the benefits will not be so clear, though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
@@
 
-\bibitem[Diekmann et al.(2000)]{diekmann0}
-Diekmann, O. and  J.~A.~P. Heesterbeek (2000).
-\newblock \emph{Mathematical epidemiology of infectious diseases: model building, analysis and interpretation}.
-Chichester: Wiley.
-
 \bibitem[Diekmann et al.(2013)]{diekmann13}

### 2014-11-07 draftfinalreport.tex
edited final draft

--- 2014-11-04 draftfinalreport.tex
@@
 \end{center}
-\caption{Per capita rates of measles infections for 2007--2014 broken down by ethnicity (Table~\ref{table:percap} for details)}
+\caption{Per capita rates of measles infections for specific age and New Zealand deprivation indices (NZDep) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
 \label{fig:percap}
@@
 \begin{equation} \label{eq:reg}
- log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a * x_e)+ log(population)  + \epsilon
+ \log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a \times x_e)+ \log(population)  + \epsilon
   \end{equation}
-Where $\alpha$ is the intercept, $y$ cases, $_a$ age, $_e$ Prioritized Ethnicity, $_n$ NZDep, and $\epsilon$ the error term.
+Where $\alpha$ is the intercept, $y$ signifies cases, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity, $_n$ signifies NZDep, and $\epsilon$ is the error term.
 
 \subsection{Vaccination history of measles infection and population immunity estimation methods}
+\label{sub:immunestat}
 
 In this section we describe the vaccination history of the measles cases from 2007--2014 outbreaks using the data provided by ESR.
-We describe the population immunity levels at both a national level and that for each DHB in New Zealand.
+We describe the population immunity levels at both a national level and for each DHB in New Zealand.
 To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work.
-Specifically, we use the data in Table \ref{table:sero}, that uses the serosurvey results and the National Immunisation Register for under 6 year olds.
+Specifically, we use the data in Table \ref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.
 
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Mon Nov 03 20:46:33 2014
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Fri Nov 07 22:00:40 2014
 \begin{tabular}{lrrrrrr}
@@
    \hline
-\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014}
+\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
 \label{table:anova}
@@
 \vspace{5mm} %5mm vertical space
-% latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Mon Nov 03 20:46:33 2014
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Fri Nov 07 22:00:40 2014
 \begin{tabular}{rrrrr}
@@
 \begin{table}
-\caption{Lowest national measles vaccine cover (\%, 2012)}
+\caption{Lowest national measles vaccine cover (\%), 2012}
 %latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by travellers numbers$*$measles incidence}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:imrisk12} and ~\ref{table:risk12}}
 %latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by travellers numbers$*$measles incidence}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:risk12}}
 %latex.default(imrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
 %latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-Globally, vaccination coverage is lowest and measles incidence highest in less developed nations (Table \ref{table:cover12} and Table \ref{table:incidence12}, and Figures~\ref{fig:cover12} and \ref{fig:incidence12}.
-Analyses of 2012 data, the most complete and recent year of data, suggest immigration (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:imtravel12}.
-However, travel by New Zealanders is dominated by that to Australia (Table~\ref{table:nztravel12} and Figure~\ref{fig:nztravel12}).
-Together, these mean that the greatest travel location for New Zealanders and immigration travel is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
-Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
-These data are shown in Table \ref{table:risk12}.
-The data for all the variables for each nation state and territories for 2012 are plotted in Figure~\ref{fig:immigration12}, Figure~\ref{fig:incidence12}, and Figure~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
-The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and \ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and \ref{fig:imrisk12}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport-036}
-\end{center}
-\caption{Measles incidence per million 2012}
+\begin{table}
+\caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation.
+incidence: measles incidence per million; cover: percent vaccination cover (\%); immigration: the number of travellers entering New Zealand from the country}
+%latex.default(toprisk2, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lrrr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}&\multicolumn{1}{c}{cover}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$  9$&$94$&$1799655$\tabularnewline
+United Kingdom&$ 33$&$93$&$ 401737$\tabularnewline
+Thailand&$ 78$&$98$&$  61358$\tabularnewline
+Malaysia&$ 64$&$95$&$  70366$\tabularnewline
+Indonesia&$ 63$&$84$&$  35352$\tabularnewline
+India&$ 15$&$85$&$ 107618$\tabularnewline
+China&$  4$&$99$&$ 322076$\tabularnewline
+Ukraine&$280$&$79$&$   2733$\tabularnewline
+Ireland&$ 23$&$92$&$  31396$\tabularnewline
+Romania&$342$&$94$&$   1913$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:toprisk}
+\end{table}
+
+Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (Tables~\ref{table:incidence12} and ~\ref{table:cover12}, and Figures~\ref{fig:incidence12} and ~\ref{fig:cover12}).
+Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}.
+However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}.
+Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation.
+Though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
+These data are shown in Table~\ref{table:risk12}.
+The data for all the variables for each nation state and territories for 2012 are plotted in Figures~\ref{fig:immigration12}, ~\ref{fig:incidence12}, and ~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
+The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.
+
+The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in Table~\ref{table:toprisk}.
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-037}
+\end{center}
+\caption{Measles incidence per million, 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
 \label{fig:incidence12}
@@
 \begin{table}
-\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014 (see text for abbreviations)}
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014.
+PHN: public health nurse; PHA: public health assistant; IMT: incident management team; SSO: safety/security officer (SSO)}
 %latex.default(costtable1, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Reported direct costs, with the number of cases, length of hospital day, cost, cost per case and cost per day for patients with measles as the primary diagnosis, 2000--2014}
+\caption{Reported direct costs in NZ\$, with the number of cases, length of hospital day, total cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, 2000--2014}
 %latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Days}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Per.case}&\multicolumn{1}{c}{Per.day}\tabularnewline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Days}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cost per case}&\multicolumn{1}{c}{Cost per day}\tabularnewline
 \hline
@@
 \begin{table}\small
-\caption{Number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender, 2000--2014}
+\caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\ref{table:regression})}
 %latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Gender}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Length.of.stay}&\multicolumn{1}{c}{Cost.per.case}\tabularnewline
+\multicolumn{1}{c}{Year}&\multicolumn{1}{c}{Gender}&\multicolumn{1}{c}{Cost}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{Length of stay}&\multicolumn{1}{c}{Cost per case}\tabularnewline
 \hline
@@
      \centering
-\includegraphics{draftfinalreport-060}
-     \caption{The estimated proportion of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{tab:cost20} and ~\ref{tab:cost50})}
+\includegraphics{draftfinalreport-061}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
      \label{fig:numvac}
@@
      \centering
-\includegraphics{draftfinalreport-062}
+\includegraphics{draftfinalreport-063}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
@@
      \centering
-\includegraphics{draftfinalreport-063}
+\includegraphics{draftfinalreport-064}
      \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc.costs}&\multicolumn{1}{c}{Wages.saved}&\multicolumn{1}{c}{Manage.saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp.saved}&\multicolumn{1}{c}{Costs.save}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
-\hline
-Auckland&$17920$&$358400$&$26142401$&$55010965$&$5297$&$9057921$&$79260619$&$ 82$&$209524$&$139.56$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7078643$&$14895456$&$1434$&$2452636$&$21461595$&$ 71$&$181417$&$ 78.58$\tabularnewline
-Canterbury&$13687$&$273740$&$20719105$&$43598824$&$4198$&$7178837$&$62817837$&$ 62$&$158420$&$145.36$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15440117$&$32490349$&$3129$&$5349752$&$46812580$&$ 96$&$245296$&$102.99$\tabularnewline
-Counties Manukau&$18880$&$377600$&$27605617$&$58089983$&$5594$&$9564902$&$83696914$&$ 50$&$127758$&$165.62$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5743794$&$12086558$&$1164$&$1990132$&$17414493$&$ 56$&$143089$&$ 79.84$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6574404$&$13834395$&$1332$&$2277925$&$19932803$&$ 86$&$219745$&$ 64.82$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4356088$&$ 9166434$&$ 883$&$1509314$&$13207136$&$ 62$&$158420$&$ 61.10$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7003972$&$14738327$&$1419$&$2426764$&$21235202$&$ 75$&$191638$&$ 74.72$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3700829$&$ 7787585$&$ 750$&$1282278$&$11220469$&$ 90$&$229965$&$ 40.49$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4772232$&$10042118$&$ 967$&$1653502$&$14468834$&$ 70$&$178862$&$ 60.22$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1407842$&$ 2962496$&$ 285$&$ 487795$&$ 4268408$&$ 72$&$183972$&$ 21.15$\tabularnewline
-Southern&$ 8371$&$167420$&$12681485$&$26685411$&$2570$&$4393931$&$38448739$&$102$&$260627$&$ 89.82$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2039609$&$ 4291911$&$ 413$&$ 706692$&$ 6183849$&$ 47$&$120093$&$ 41.99$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4414818$&$ 9290019$&$ 895$&$1529663$&$13385198$&$ 68$&$173752$&$ 57.76$\tabularnewline
-Waikato&$11331$&$226620$&$16988072$&$35747682$&$3442$&$5886094$&$51505793$&$ 95$&$242741$&$109.74$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1129294$&$ 2376352$&$ 229$&$ 391282$&$ 3423884$&$ 59$&$150755$&$ 20.73$\tabularnewline
-Waitemata&$17291$&$345820$&$25819386$&$54331250$&$5232$&$8946002$&$78281277$&$ 70$&$178862$&$149.20$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1061335$&$ 2233347$&$ 215$&$ 367736$&$ 3217840$&$ 50$&$127758$&$ 22.75$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2122670$&$ 4466695$&$ 430$&$ 735471$&$ 6435680$&$ 58$&$148200$&$ 36.62$\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
+\hline
+Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$32.45$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$11.31$\tabularnewline
+Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$33.96$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$17.66$\tabularnewline
+Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$50.79$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$11.62$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$ 8.76$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$ 8.08$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$10.62$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$ 4.80$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$ 7.86$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$ 2.31$\tabularnewline
+Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$13.92$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$ 5.06$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$ 7.49$\tabularnewline
+Waikato&$11331$&$226620$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$19.51$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$ 2.26$\tabularnewline
+Waitemata&$17291$&$345820$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$36.89$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$ 2.51$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$ 4.29$\tabularnewline
 \hline
@@
 
-Vacc  Vacc.costs	Wages.saved	Manage.saved	Hospitalised	Hosp.saved	Costs.save	Outbreak	OB.costs	Benefit/cost
-
 \begin{sidewaystable}\small
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc.costs}&\multicolumn{1}{c}{Wages.saved}&\multicolumn{1}{c}{Manage.saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp.saved}&\multicolumn{1}{c}{Costs.save}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
-\hline
-Auckland&$17920$&$896000$&$26142401$&$55010965$&$5297$&$9057921$&$79260619$&$ 82$&$209524$&$71.70$\tabularnewline
-Bay of Plenty&$ 4585$&$229250$&$ 7078643$&$14895456$&$1434$&$2452636$&$21461595$&$ 71$&$181417$&$52.26$\tabularnewline
-Canterbury&$13687$&$684350$&$20719105$&$43598824$&$4198$&$7178837$&$62817837$&$ 62$&$158420$&$74.54$\tabularnewline
-Capital and Coast&$10461$&$523050$&$15440117$&$32490349$&$3129$&$5349752$&$46812580$&$ 96$&$245296$&$60.93$\tabularnewline
-Counties Manukau&$18880$&$944000$&$27605617$&$58089983$&$5594$&$9564902$&$83696914$&$ 50$&$127758$&$78.09$\tabularnewline
-Hawke's Bay&$ 3751$&$187550$&$ 5743794$&$12086558$&$1164$&$1990132$&$17414493$&$ 56$&$143089$&$52.67$\tabularnewline
-Hutt Valley&$ 4388$&$219400$&$ 6574404$&$13834395$&$1332$&$2277925$&$19932803$&$ 86$&$219745$&$45.39$\tabularnewline
-Lakes&$ 2886$&$144300$&$ 4356088$&$ 9166434$&$ 883$&$1509314$&$13207136$&$ 62$&$158420$&$43.63$\tabularnewline
-MidCentral&$ 4628$&$231400$&$ 7003972$&$14738327$&$1419$&$2426764$&$21235202$&$ 75$&$191638$&$50.20$\tabularnewline
-Nelson Marlborough&$ 2356$&$117800$&$ 3700829$&$ 7787585$&$ 750$&$1282278$&$11220469$&$ 90$&$229965$&$32.26$\tabularnewline
-Northland&$ 3071$&$153550$&$ 4772232$&$10042118$&$ 967$&$1653502$&$14468834$&$ 70$&$178862$&$43.53$\tabularnewline
-South Canterbury&$  893$&$ 44650$&$ 1407842$&$ 2962496$&$ 285$&$ 487795$&$ 4268408$&$ 72$&$183972$&$18.67$\tabularnewline
-Southern&$ 8371$&$418550$&$12681485$&$26685411$&$2570$&$4393931$&$38448739$&$102$&$260627$&$56.61$\tabularnewline
-Tairawhiti&$ 1359$&$ 67950$&$ 2039609$&$ 4291911$&$ 413$&$ 706692$&$ 6183849$&$ 47$&$120093$&$32.89$\tabularnewline
-Taranaki&$ 2899$&$144950$&$ 4414818$&$ 9290019$&$ 895$&$1529663$&$13385198$&$ 68$&$173752$&$42.00$\tabularnewline
-Waikato&$11331$&$566550$&$16988072$&$35747682$&$3442$&$5886094$&$51505793$&$ 95$&$242741$&$63.64$\tabularnewline
-Wairarapa&$  720$&$ 36000$&$ 1129294$&$ 2376352$&$ 229$&$ 391282$&$ 3423884$&$ 59$&$150755$&$18.33$\tabularnewline
-Waitemata&$17291$&$864550$&$25819386$&$54331250$&$5232$&$8946002$&$78281277$&$ 70$&$178862$&$75.02$\tabularnewline
-West Coast&$  685$&$ 34250$&$ 1061335$&$ 2233347$&$ 215$&$ 367736$&$ 3217840$&$ 50$&$127758$&$19.86$\tabularnewline
-Whanganui&$ 1378$&$ 68900$&$ 2122670$&$ 4466695$&$ 430$&$ 735471$&$ 6435680$&$ 58$&$148200$&$29.64$\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
+\hline
+Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$2095239$&$26.62$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$1814170$&$10.55$\tabularnewline
+Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$1584205$&$27.82$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$2452962$&$15.80$\tabularnewline
+Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$1277585$&$37.84$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$1430895$&$10.81$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$2197445$&$ 8.28$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$1584205$&$ 7.68$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$1916377$&$ 9.93$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$2299652$&$ 4.66$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$1788618$&$ 7.48$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$1839722$&$ 2.28$\tabularnewline
+Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$2606273$&$12.77$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$1200930$&$ 4.90$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$1737515$&$ 7.14$\tabularnewline
+Waikato&$11331$&$566550$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$2427411$&$17.29$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$1507550$&$ 2.23$\tabularnewline
+Waitemata&$17291$&$864550$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$1788618$&$29.68$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$1277585$&$ 2.47$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$1481998$&$ 4.18$\tabularnewline
 \hline

### 2014-11-07 draftfinalreport.Rnw
Final edits but not compiling correctly

@@
 \usepackage{natbib}
+\usepackage{graphicx}
 \usepackage{graphics}
@@
 \author[3] {Mick G Roberts}
-\affil[1] {$^{m}$EpiLab,Infectious Diseases Research Centre, Massey University, Palmerston North 4442,New Zealand}
-\affil[2] {EpiCentre,Infectious Diseases Research Centre, Massey University, Palmerston North 4442,New Zealand}
-\affil[3] {Institute of Natural \& Mathematical Sciences,New Zealand Institute for Advanced Study and the Infectious Disease Research Centre, Massey University, Private Bag 102 904, North Shore Mail Centre, Auckland, New Zealand}
+\affil[1] {$^{m}$EpiLab, Infectious Diseases Research Centre, Massey University, Palmerston North 4442, New Zealand}
+\affil[2] {EpiCentre, Infectious Diseases Research Centre, Massey University, Palmerston North 4442, New Zealand}
+\affil[3] {Institute of Natural \& Mathematical Sciences, New Zealand Institute for Advanced Study and the Infectious Disease Research Centre, Massey University, Private Bag 102 904, North Shore Mail Centre, Auckland, New Zealand}
 \date{}
@@
 \item The regression analyses suggest that age is a particularly strong risk factor for measles.
+People of European ethnicity, however, make up the majority of cases.
 However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis.
 In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
-People of European ethnicity, however, make up the majority of cases.
 \item Of the 11\% of the population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
@@
 However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
-\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one.
+\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
@@
 \item The mean wage loss per measles case is estimated to be \$839
-\item The mean cost per measles case attending hospital is estimated to be \$1710, and approximately 17\% of measles cases attend hospital
+\item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
 \item The mean management cost per case is \$1,765
 \item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1).
-\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out with additional vaccinations.
+\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but possibly a peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out following additional vaccinations.
 \end{itemize}
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section \ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section~\ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{sec:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see Section~\ref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
 
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{sec:epidemic_modelling}).
-
-Thus in this report we include regression analyses that help understand risk factors for being measles cases in New zealand (Section~\ref{sub:risk_analyses}, descriptions of the New Zealand population in terms of vaccination coverage and immunity (Section~\ref{sub:immunestat}), and use those data to model the likely measles outbreak sizes in each Distric Health Board and the country (Section~\ref{sec:epidemic_modelling}.
-We estimate the costs of previous and current measles outbreaks (Section~\ref{sub:cost}) and uses these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch up vaccination programmes (Section~\ref{sub:cost_benefit}).
+This is particularly in light of our modelling and risk analyses results (Section~\ref{sub:risk_analyses} and Section~\ref{section:epidemic_modelling}).
+
+Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (Section~\ref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (Section~\ref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (Section~\ref{section:epidemic_modelling}).
+We estimate the costs of previous and the current measles outbreaks (Section~\ref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (Section~\ref{sub:cost_benefit}).
 This report includes further developments of the analyses provided in previous interim reports.
@@
 
-We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:immunestat}), and current rates of immunity within the population (section~\ref{sub:immunestat}).
+We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (Section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (Section~\ref{sub:immunestat}), and current rates of immunity within the population (Section~\ref{sub:immunestat}).
 In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\ref{sub:risk_analyses}).
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (Section~\ref{{sub:imp_risk}}).
 
@@
      \includegraphics[width=1\textwidth]{incidence_1997_2014.pdf}
-     \caption{Measles incidence from 1997 to 2014}
+     \caption{Measles incidence in New Zealand from 1997 to 2014}
      \label{fig:incidence1997}
@@
 \end{center}
-\caption{Age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
 \label{fig:ageinyears}
@@
 @
-\caption{Absolute number of measles cases in specific age and socio-economic deprivation categories from 2007-2014 with no ethnicity recorded}
+\caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
 \label{table:none}
@@
 The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the Middle Eastern/Latin American/African category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
 However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
@@
 \end{center}
-\caption{Per capita rates of measles infections for specific age and New Zealand deprivation indices (NZDep) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
+\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
 \label{fig:percap}
@@
 \footnotesize
-\caption{Numbers of measles cases, population sizes and per capita rates per 10,000 of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\caption{Numbers of measles cases (Cases), population sizes (Ppulation) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
 <<results=tex,echo=FALSE>>=
@@
 \end{center}
-\caption{Vaccination coverage (0--5 years) and serosurvey estimates ($>5$ years) of immunity among different age classes used in these analyses}
+\caption{Vaccination coverage (0--5 years) and serosurvey estimates of immunity ($>5$ years) among different age classes used in these analyses}
 \label{table:sero}
@@
 \subsection{Measles importation risk methods}
-
-For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year, to measure human movement to and from New Zealand  (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}).
+\label{sub:imp_risk}
+For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand.
 We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
@@
 We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
-We use the data from 2012 because this year had the most complete WHO measles data and yet is most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+We present the data from 2012 because this year had the most complete WHO measles data and yet is most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
 
@@
 \end{center}
-\caption{Distribution of measles cases per category used in the final regression model}
+\caption{Distribution of measles cases per category (Table~\ref{table:percap}) used in the final regression model (Equation~\ref{eq:reg})}
 \label{fig:histcasecat}
@@
 \vspace{5mm} %5mm vertical space
+
 A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.
@@
 \vspace{5mm} %5mm vertical space
+\begin{table}
 <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
@@
 \par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (Figure~\ref{fig:ageinyears}).
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also Figure~\ref{fig:ageinyears}).
 
@@
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}).
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}, Table~\ref{table:sero}, and Section~\ref{sub:popim}).
 This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}).
@@
 \subsection{Population immunity results}
+\label{sub:popim}
 
@@
 
-\begin{table}
-\caption{Lowest national measles vaccine cover (\%), 2012}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:cover12}
-\end{table}
-
-\begin{table}
-\caption{Highest measles incidence per million (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(incidence, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:incidence12}
-\end{table}
-
-\begin{table}
-\caption{Total New Zealand traveller numbers by country (New Zealand nationals and immigrants, 2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(immigration, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:immigration12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%
 <<echo=FALSE>>=
@@
 @
-
-\begin{table}
-\caption{New Zealander travel numbers by destination (2012)}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(nz, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:nztravel12}
-\end{table}
 
@@
 \begin{table}
+\caption{Highest measles incidence per million (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(incidence, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:incidence12}
+\end{table}
+
+\begin{table}
+\caption{Lowest national measles vaccine cover (\%), 2012}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:cover12}
+\end{table}
+
+\begin{table}
 \caption{Non-New Zealander travel and immigration numbers (2012)}
@@
 
+\begin{table}
+\caption{New Zealander travel numbers by destination (2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(nz, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:nztravel12}
+\end{table}
+
+\begin{table}
+\caption{Total New Zealand traveller numbers by country (New Zealand nationals and immigrants, 2012)}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(immigration, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:immigration12}
+\end{table}
+
 %%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(risk, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:risk12}
+\end{table}
 
@@
 \label{table:imrisk12}
-\end{table}
-
-\begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(risk, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:risk12}
 \end{table}
@@
 Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}.
-However, travel by New Zealanders is dominated by that to Australia \ref{table:nztravel12} and \ref{fig:nztravel12}.
-Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\ref{table:immigration12} and \ref{fig:immigration12}.
+However, travel by New Zealanders is dominated by that to Australia (Tables~\ref{table:nztravel12} and \ref{fig:nztravel12}).
+Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}.
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation.
-Though immigration is lower from some Asian countries, travel from (and thus we presume to) some Asian countries also poses a high risk of measles importation to New Zealand.
+Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand.
 These data are shown in Table~\ref{table:risk12}.
-The data for all the variables for each nation state and territories for 2012 are plotted in Figures~\ref{fig:immigration12}, ~\ref{fig:incidence12}, and ~\ref{fig:cover12} and the risk map for measles incidence and immigration in Figure~\ref{fig:risk12}.
-The breakdown by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.
+The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.
 
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
+print(imnp1)
+@
+\end{center}
+\caption{Non-New Zealander international travel and immigration, 2012 }
+\label{fig:immigration12}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+print(nznp1)
+@
+\end{center}
+\caption{New Zealander international travel, 2012 }
+\label{fig:nztravel12}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
 print(np1)
@@
 \label{fig:travel12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(nznp1)
-@
-\end{center}
-\caption{New Zealander international travel, 2012 }
-\label{fig:nztravel12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(imnp1)
-@
-\end{center}
-\caption{Non-New Zealander international travel and immigration, 2012 }
-\label{fig:immigration12}
 \end{figure}
@@
 
-\begin{figure}
-     \centering
+\begin{figure}[H]
+     \begin{center}
      \includegraphics[width=0.9\textwidth]{nzers.pdf}
@@
 However, our analyses also highlight other groups that are at greater risk of measles.
-In particular, Pacific people are at greater risk, as are the less deprived (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
+In particular, Pacific people are at greater risk per capita, as are the less deprived (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
 Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses.
@@
 Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand.
-However, the strong seasonality in travel around Christmas suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
+However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
 The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}).
@@
 \item Risk of measles infection decreases significantly with age.
-\item Pacific people are statistically more at risk of measles infection.
+\item European ethnicity comprise the largest number of measles cases.
+\item Pacific people are statistically more at risk per capita of measles infection.
 \item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
-\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk than European or Maori children.
+\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk per capita than European or Maori children.
 \item The majority of cases are unvaccinated.
@@
 \section{Modelling measles epidemics}
-\label{sec:epidemic_modelling}
+\label{section:epidemic_modelling}
 
@@
 
-These results were comparable to others, for example: a two-dose schedule for England and Wales, with the second vaccination given at age four \citep{babad95}; and a second vaccination at either 18 months or five years were recommened \citep{gay98} to complement the first vaccination at 12 months in Canada.
+These results were comparable to others, for example: a two-dose schedule for England and Wales, with the second vaccination given at age four \citep{babad95}; and a second vaccination at either 18 months or five years were recommended to complement the first vaccination at 12 months in Canada \citep{gay98}.
 In addition, vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was found to predict the prevention of measles epidemics in Israel \citep{agur93}.
@@
 The differences in the models have been in the details of the representation of the infectious period, and in the ways in which the age and contact structures of the population have been specified.
-While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, in the Netherlands analyses showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation is difficult to evaluate \citep{glass4}.
+While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, in the Netherlands analyses showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation difficult to evaluate \citep{glass4}.
 
@@
 
-\begin{figure}
+\begin{figure}[H]
       \centering
@@
 
-\begin{figure}
+\begin{figure}[H]
      \centering
@@
 The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
-The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current outbreak 1.82--2.13.
+The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current 2013--2014 outbreak 1.82--2.13.
 The probability density of the $R_v$ estimates for each outbreak all include one.
@@
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
+Given the risk of importation of measles as highlighted in Section~\ref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
 The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
@@
 The regularity of these outbreaks, however, implies that the approximately yearly importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:risk_analyses} and our previous report is likely to continue, we include the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+Given the risk of importation of measles as highlighted in Section~\ref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
 
@@
 The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics.
-Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% na\"{\i}ve.
-Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, but \citep{glass4} showed that in the Netherlands high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult.
+Vaccinating 85\% of susceptible children aged one to seven years at five-yearly intervals was suggested to be sufficient to prevent epidemics in Israel \citep{agur93}, but nearly all other studies in Europe suggest no strategies succeeded if coverage rates were below approximately 87\%, which the population level immunity in New Zealand has only just reached, with approximately 11\% currently na\"{\i}ve.
+Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, whereas \citep{glass4} showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult.
 Future analyses at a smaller, more local level would be useful, but the lack of appropriate data matching between the NIR data and the census unit area data currently prevent these.
@@
 Thirty years later, it was estimated there were still approximately 45 million cases of measles occurring annually, including 6 million measles-related fatalities.
-\citep{wolfson7} estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005.
-Similarly, the incidence of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
-They estimated approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination.
+It was estimated that in 1999 measles was responsible for more than 30 million disability adjusted life years (DALYs) lost and 12 million in 2005 \citep{wolfson7}.
+The incidence of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
+Approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination \citep{wolfson7}.
 The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \citep{who13}.
@@
 Several studies have been conducted in the United States to assess the economic impact of recent measles outbreaks due to imported cases.
-\citep{ortegasanchez14} estimated the economic impact to public health departments in the US as the result of 16 outbreaks in 2011.
+The economic impact to public health departments in the US as the result of 16 outbreaks in 2011 has been estimated \citep{ortegasanchez14}.
 The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$2.7 and 5.3 million.
@@
 In the Republic of Korea, the economics of alternative measles vaccination programs were compared.
-All of the alternatives were found to be economically efficient (benefit/cost ratio (B/C) > 1.0), with the alternative using two doses of the MMR program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27).
+All of the alternatives were found to be economically efficient (benefit/cost ratio (B/C) > 1.0), with the alternative using two doses of the MMR program, with a catch-up campaign for measles and rubella being the most favourable (B/C = 1.27), and B/C values 10.8 -- 54.2 have been estimated for MMR vaccination in the USA \citep{zhou4}.
 
@@
 Data to estimate the costs relating to the current measles outbreak were obtained from the New Zealand Ministry of Health, for the period 2008 through June 2014.
-Data included information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, and associated cost.
+Data include information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, and associated cost.
 
@@
 The costs are reported as direct, additional (above normal budgeting) costs required to enable the management of measles.
-It includes a breakdown by individual performing the work and whether it was during the normal work schedule (Monday to Friday, M-F) or weekends.
+The data includes a breakdown by individual performing the work and whether it was during the normal work schedule (Monday to Friday, M-F) or weekends.
 Normal work is calculated as $1.2 \times \textsf{full time equivalent (FTE)} \times \textsf{number of days worked}$.
@@
 
-To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{sec:epidemic_modelling}), we carried out a number of analyses.
-Primarily, we used Equation~\ref{equation:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{sec:epidemic_modelling}).
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{section:epidemic_modelling}), we carried out a number of analyses.
+Primarily, we used Equation~\ref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{section:epidemic_modelling}).
 
@@
 
-The cost figures above are used to estimate what the savings for vaccinating additional populations are, using estimated per case costs saved.
+The cost figures above are used to estimate the savings for vaccinating additional populations, using estimated per case costs saved.
 The costs of the catch up vaccination schemes are estimated to be two different values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
@@
 \begin{sidewaystable}\small
-\caption{DHB and National level catch up vaccination rates and estimated outbreak sizes post vaccination. Size: DHB Population, Statistics NZ 2013;  	
+\caption{DHB and national level catch up vaccination rates and estimated outbreak sizes post catch-up vaccination. Size: DHB Population, Statistics NZ 2013;  	
 Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
@@
 @
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (Section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
      \label{fig:numvac}
@@
 @
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
 The full distribution of results can be seen in Figure~\ref{fig:sim1}}
@@
 @
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
 A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 library(Hmisc)
+colnames(costtable20)<-c("Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
 latex(costtable20, file="", table.env=FALSE,rowname=NULL)
 @
-\label{table:cost20}
-\begin{tablenotes}\footnotesize  
-        \item For numbers to vaccinate see Table~\ref{table:attack}
-        \item Proportion of cases hospitalised 0.17
-        \item Wage losses per case \$839 and cost per hospitalised case \$1710
-        \item Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
 \end{sidewaystable}
@@
 \begin{sidewaystable}\small
-\caption{Benefit--costs analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
+\caption{Benefit--cost analyses with 50 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 library(Hmisc)
-
+colnames(costtable50)<-c("Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
 latex(costtable50, file="", table.env=FALSE,rowname=NULL)
@@
 \label{table:cost50}
-\begin{tablenotes}\footnotesize  
-        \item For numbers to vaccinate see Table~\ref{table:attack}
-        \item Proportion of cases hospitalised 0.17
-        \item Wage losses per case \$839 and cost per hospitalised case \$1710
-        \item Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
 \end{sidewaystable}
@@
 For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1710 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
+Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
 We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
@@
 For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
-In addition to this information, similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals.
-In addition, similar data would be needed for non-publicly funded hospitals, e.g.
-private clinics.
+Similar data would be needed for cases occurring outside the period 2011--2013 at publicly funded hospitals, non-publicly funded hospitals, and private clinics.
 Other factors that would be useful to investigate in the future include the relationship between the costs over time.
@@
 It may be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014, but we have done so for these purposes.
-However, other studies have demonstrated direct costs required to manage measles are not linear.
 
@@
 They estimated direct costs per case to be less than US\$500.
+However, the combined direct costs of \euro 520 per case in Germany \citep{siedler6} suggests costs may be higher.
 Thus, our estimates of the additional costs, including indirect costs and wages lost may be appropriate.
@@
 They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
-The work by \citep{zhou4} also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see Table~\ref{table:zhou}).
-However, the cost of measles in their analyses was substantially higher than mumps and rubella (Table~\ref{table:zhou}).
+Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see Table~\ref{table:zhou}) \citep{zhou4}.
+However, the cost of measles in their analyses was substantially higher than mumps and rubella (Table~\ref{table:zhou}) \citep{zhou4}.
 
@@
 
-Our estimates for the benefit--cost ratio of catch up vaccination are also much higher than those in some studies.
+Our estimates for the benefit--cost ratio of catch up vaccination are also higher than those in some studies.
 If we presume the $R_0$ estimated for measles in New Zealand is representative and following 28\% vaccination of the na\"{\i}ve population, measles will be unable to persist ($R_v$ < 1), our B/C ratio is substantially greater than 1.
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
-However,  \citep{zhou4} estimate the B/C ratio to be much greater than one for MMR vaccination (see Table~\ref{table:zhou2}).
+However,  the estimated B/C ratio for MMR has been estimated to be between 10.2 and 54.2, much greater than one (see Table~\ref{table:zhou2}) \citep{zhou4}.
 As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis.
 
-Our results may be conservative.
+Our results for the benefits of MMR vaccination may be conservative.
 We presume that in a totally na\"{\i}ve population $R_0$ for measles was around 12.
@@
 It is unlikely $R_0$ was so high as 18 in New Zealand, as measles would still be endemic given current vaccination rates.
-However, the benefits of catch up vaccination are clear once $R_v$ is greater than one (Tables~\ref{table:cost20} and ~\ref{table:cost20}, and previous interim report).
-However, as noted in our previous report, if measles $R_v$ in the currently na\"{\i}ve population is less than one, the benefits will not be so clear, though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
-
-\begin{table}[htdp]
+However, the benefits of catch up vaccination are clear if $R_v$ is greater than one (Tables~\ref{table:cost20} and ~\ref{table:cost20}) and our analyses do not include the additional benefits of mumps and rubella immunity.
+As noted in our previous report, if measles $R_v$ in the currently na\"{\i}ve population is less than one, the benefits will not be so clear, though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
+
+\begin{sidewaystable}[htdp]
 \tiny
@@
 \end{center}
-\caption{Benefit--cost ratios, USA, from \citep{zhou4}}
+\caption{Benefit--cost ratios for Measles-Mumps-Rubella (MMR) vaccination, USA, from \citep{zhou4}}
 \label{table:zhou2}
-\end{table}%
+\end{sidewaystable}%
 
 Finally, our model of measles introductions remains a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local na\"{\i}ve population.
-The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{sec:epidemic_modelling}).
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis (see previous report) may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
+The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{section:epidemic_modelling}).
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
 Given that, the greater the vaccination coverage, the smaller the outbreaks will be.
@@
 \item The mean wage losses per measles case is estimated to be \$839
-\item The mean cost per measles case attending hospital is estimated to be \$1,710
+\item The mean cost per measles case attending hospital is estimated to be \$1,877
 \item Approximately 17\% of measles cases attend hospital

### 2014-11-07 draftfinalreport.tex
Final edits but not compiling correctly

@@
 \hline
-\end{tabular}\end{center}\caption{Absolute number of measles cases in specific age and socio-economic deprivation categories from 2007-2014 with no ethnicity recorded}
+\end{tabular}\end{center}\caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
 \label{table:none}
@@
 \footnotesize
-\caption{Numbers of measles cases, population sizes and per capita rates per 10,000 of measles in specific age, ethnicity and socio-economic deprivation categories from 2007-2014}
+\caption{Numbers of measles cases (Cases), population sizes (Ppulation) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
 %latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
@@
 % latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Fri Nov 07 22:00:40 2014
+% Sat Nov 08 00:20:12 2014
 \begin{tabular}{lrrrrrr}
@@
 \vspace{5mm} %5mm vertical space
+\begin{table}
 % latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Fri Nov 07 22:00:40 2014
+% Sat Nov 08 00:20:12 2014
 \begin{tabular}{rrrrr}
@@
 %%%%%%%%%%%%%%%%%%%%%
+
+%%%%%%%%%%%%%%%%%%%%%
+
+\begin{table}
+\caption{Highest measles incidence per million (2012)}
+%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
+\hline
+Equatorial Guinea&$1617$\tabularnewline
+Nauru&$1100$\tabularnewline
+Democratic Republic of the Congo&$1096$\tabularnewline
+Somalia&$ 979$\tabularnewline
+Djibouti&$ 824$\tabularnewline
+Sudan&$ 786$\tabularnewline
+Burkina Faso&$ 447$\tabularnewline
+Romania&$ 342$\tabularnewline
+Ukraine&$ 280$\tabularnewline
+Sudan&$ 229$\tabularnewline
+Angola&$ 214$\tabularnewline
+Monaco&$ 132$\tabularnewline
+Nepal&$ 122$\tabularnewline
+Sierra Leone&$ 113$\tabularnewline
+Afghanistan&$  93$\tabularnewline
+Yemen&$  91$\tabularnewline
+Lesotho&$  87$\tabularnewline
+Qatar&$  78$\tabularnewline
+Thailand&$  78$\tabularnewline
+Malaysia&$  64$\tabularnewline
+Zambia&$  64$\tabularnewline
+Ghana&$  64$\tabularnewline
+Indonesia&$  63$\tabularnewline
+Congo&$  60$\tabularnewline
+Uganda&$  56$\tabularnewline
+Libyan Arab Jamahiriya&$  52$\tabularnewline
+Ethiopia&$  47$\tabularnewline
+Pakistan&$  45$\tabularnewline
+Myanmar&$  41$\tabularnewline
+Nigeria&$  38$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:incidence12}
+\end{table}
 
@@
 \begin{table}
-\caption{Highest measles incidence per million (2012)}
-%latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
+\caption{Non-New Zealander travel and immigration numbers (2012)}
+%latex.default(im, file = "", table.env = FALSE, rowname = NULL)%
 \begin{center}
@@
 \hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{incidence}\tabularnewline
-\hline
-Equatorial Guinea&$1617$\tabularnewline
-Nauru&$1100$\tabularnewline
-Democratic Republic of the Congo&$1096$\tabularnewline
-Somalia&$ 979$\tabularnewline
-Djibouti&$ 824$\tabularnewline
-Sudan&$ 786$\tabularnewline
-Burkina Faso&$ 447$\tabularnewline
-Romania&$ 342$\tabularnewline
-Ukraine&$ 280$\tabularnewline
-Sudan&$ 229$\tabularnewline
-Angola&$ 214$\tabularnewline
-Monaco&$ 132$\tabularnewline
-Nepal&$ 122$\tabularnewline
-Sierra Leone&$ 113$\tabularnewline
-Afghanistan&$  93$\tabularnewline
-Yemen&$  91$\tabularnewline
-Lesotho&$  87$\tabularnewline
-Qatar&$  78$\tabularnewline
-Thailand&$  78$\tabularnewline
-Malaysia&$  64$\tabularnewline
-Zambia&$  64$\tabularnewline
-Ghana&$  64$\tabularnewline
-Indonesia&$  63$\tabularnewline
-Congo&$  60$\tabularnewline
-Uganda&$  56$\tabularnewline
-Libyan Arab Jamahiriya&$  52$\tabularnewline
-Ethiopia&$  47$\tabularnewline
-Pakistan&$  45$\tabularnewline
-Myanmar&$  41$\tabularnewline
-Nigeria&$  38$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:incidence12}
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$809775$\tabularnewline
+United Kingdom&$306177$\tabularnewline
+China&$256036$\tabularnewline
+United States&$194438$\tabularnewline
+Japan&$ 86676$\tabularnewline
+Germany&$ 83608$\tabularnewline
+Korea, Republic of&$ 73459$\tabularnewline
+France&$ 71448$\tabularnewline
+India&$ 69038$\tabularnewline
+Canada&$ 54981$\tabularnewline
+Malaysia&$ 50506$\tabularnewline
+Fiji&$ 46723$\tabularnewline
+Netherlands&$ 33431$\tabularnewline
+South Africa&$ 30600$\tabularnewline
+Tonga&$ 26717$\tabularnewline
+Singapore&$ 25460$\tabularnewline
+Ireland&$ 25036$\tabularnewline
+Samoa&$ 24547$\tabularnewline
+Philippines&$ 24527$\tabularnewline
+Taiwan&$ 23815$\tabularnewline
+Thailand&$ 20258$\tabularnewline
+Hong Kong&$ 19103$\tabularnewline
+Indonesia&$ 15692$\tabularnewline
+Switzerland&$ 14851$\tabularnewline
+Brazil&$ 14778$\tabularnewline
+Sweden&$ 13719$\tabularnewline
+Italy&$ 13388$\tabularnewline
+Spain&$ 10104$\tabularnewline
+Denmark&$ 10056$\tabularnewline
+Russia&$  8103$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:imtravel12}
+\end{table}
+
+\begin{table}
+\caption{New Zealander travel numbers by destination (2012)}
+%latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
+\hline
+Australia&$989880$\tabularnewline
+United States&$121620$\tabularnewline
+Fiji&$104720$\tabularnewline
+United Kingdom&$ 95560$\tabularnewline
+Cook Islands&$ 71960$\tabularnewline
+China&$ 66040$\tabularnewline
+Samoa&$ 46020$\tabularnewline
+Thailand&$ 41100$\tabularnewline
+India&$ 38580$\tabularnewline
+Canada&$ 20400$\tabularnewline
+Japan&$ 20040$\tabularnewline
+Malaysia&$ 19860$\tabularnewline
+Indonesia&$ 19660$\tabularnewline
+Hong Kong&$ 18220$\tabularnewline
+Tonga&$ 17760$\tabularnewline
+Singapore&$ 17120$\tabularnewline
+South Africa&$ 15380$\tabularnewline
+Philippines&$ 15220$\tabularnewline
+France&$ 14500$\tabularnewline
+Korea, Republic of&$ 13960$\tabularnewline
+Viet Nam&$ 12920$\tabularnewline
+Germany&$ 12700$\tabularnewline
+Vanuatu&$ 12520$\tabularnewline
+Italy&$ 11820$\tabularnewline
+Taiwan&$ 10460$\tabularnewline
+New Caledonia&$  7340$\tabularnewline
+Ireland&$  6360$\tabularnewline
+French Polynesia&$  6360$\tabularnewline
+Papua New Guinea&$  6140$\tabularnewline
+Netherlands&$  5720$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:nztravel12}
 \end{table}
@@
 
-%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%
 
 \begin{table}
-\caption{New Zealander travel numbers by destination (2012)}
-%latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
+%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
 \begin{center}
@@
 \hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
-\hline
-Australia&$989880$\tabularnewline
-United States&$121620$\tabularnewline
-Fiji&$104720$\tabularnewline
-United Kingdom&$ 95560$\tabularnewline
-Cook Islands&$ 71960$\tabularnewline
-China&$ 66040$\tabularnewline
-Samoa&$ 46020$\tabularnewline
-Thailand&$ 41100$\tabularnewline
-India&$ 38580$\tabularnewline
-Canada&$ 20400$\tabularnewline
-Japan&$ 20040$\tabularnewline
-Malaysia&$ 19860$\tabularnewline
-Indonesia&$ 19660$\tabularnewline
-Hong Kong&$ 18220$\tabularnewline
-Tonga&$ 17760$\tabularnewline
-Singapore&$ 17120$\tabularnewline
-South Africa&$ 15380$\tabularnewline
-Philippines&$ 15220$\tabularnewline
-France&$ 14500$\tabularnewline
-Korea, Republic of&$ 13960$\tabularnewline
-Viet Nam&$ 12920$\tabularnewline
-Germany&$ 12700$\tabularnewline
-Vanuatu&$ 12520$\tabularnewline
-Italy&$ 11820$\tabularnewline
-Taiwan&$ 10460$\tabularnewline
-New Caledonia&$  7340$\tabularnewline
-Ireland&$  6360$\tabularnewline
-French Polynesia&$  6360$\tabularnewline
-Papua New Guinea&$  6140$\tabularnewline
-Netherlands&$  5720$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:nztravel12}
+\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
+\hline
+Australia&$15537152$\tabularnewline
+United Kingdom&$13386328$\tabularnewline
+Thailand&$ 4774688$\tabularnewline
+Malaysia&$ 4495338$\tabularnewline
+Indonesia&$ 2216871$\tabularnewline
+India&$ 1620172$\tabularnewline
+China&$ 1432659$\tabularnewline
+Ukraine&$  765096$\tabularnewline
+Ireland&$  734128$\tabularnewline
+Romania&$  655107$\tabularnewline
+Philippines&$  631303$\tabularnewline
+Spain&$  401823$\tabularnewline
+Samoa&$  373370$\tabularnewline
+Singapore&$  337236$\tabularnewline
+Nepal&$  324526$\tabularnewline
+Germany&$  193081$\tabularnewline
+Japan&$  191585$\tabularnewline
+Israel&$  174177$\tabularnewline
+Italy&$  155674$\tabularnewline
+Russia&$  155632$\tabularnewline
+Pakistan&$  148514$\tabularnewline
+Sudan&$  141552$\tabularnewline
+Switzerland&$  140589$\tabularnewline
+Nauru&$  126500$\tabularnewline
+Viet Nam&$  118241$\tabularnewline
+Afghanistan&$  111854$\tabularnewline
+Somalia&$  104775$\tabularnewline
+Saudi Arabia&$   82584$\tabularnewline
+United Arab Emirates&$   67606$\tabularnewline
+Belgium&$   63784$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:risk12}
 \end{table}
-
-%%%%%%%%%%%%%%%%%%%%%
-
-\begin{table}
-\caption{Non-New Zealander travel and immigration numbers (2012)}
-%latex.default(im, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{immigration}\tabularnewline
-\hline
-Australia&$809775$\tabularnewline
-United Kingdom&$306177$\tabularnewline
-China&$256036$\tabularnewline
-United States&$194438$\tabularnewline
-Japan&$ 86676$\tabularnewline
-Germany&$ 83608$\tabularnewline
-Korea, Republic of&$ 73459$\tabularnewline
-France&$ 71448$\tabularnewline
-India&$ 69038$\tabularnewline
-Canada&$ 54981$\tabularnewline
-Malaysia&$ 50506$\tabularnewline
-Fiji&$ 46723$\tabularnewline
-Netherlands&$ 33431$\tabularnewline
-South Africa&$ 30600$\tabularnewline
-Tonga&$ 26717$\tabularnewline
-Singapore&$ 25460$\tabularnewline
-Ireland&$ 25036$\tabularnewline
-Samoa&$ 24547$\tabularnewline
-Philippines&$ 24527$\tabularnewline
-Taiwan&$ 23815$\tabularnewline
-Thailand&$ 20258$\tabularnewline
-Hong Kong&$ 19103$\tabularnewline
-Indonesia&$ 15692$\tabularnewline
-Switzerland&$ 14851$\tabularnewline
-Brazil&$ 14778$\tabularnewline
-Sweden&$ 13719$\tabularnewline
-Italy&$ 13388$\tabularnewline
-Spain&$ 10104$\tabularnewline
-Denmark&$ 10056$\tabularnewline
-Russia&$  8103$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:imtravel12}
-\end{table}
-
-%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
-%latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{c}{country}&\multicolumn{1}{c}{risk}\tabularnewline
-\hline
-Australia&$15537152$\tabularnewline
-United Kingdom&$13386328$\tabularnewline
-Thailand&$ 4774688$\tabularnewline
-Malaysia&$ 4495338$\tabularnewline
-Indonesia&$ 2216871$\tabularnewline
-India&$ 1620172$\tabularnewline
-China&$ 1432659$\tabularnewline
-Ukraine&$  765096$\tabularnewline
-Ireland&$  734128$\tabularnewline
-Romania&$  655107$\tabularnewline
-Philippines&$  631303$\tabularnewline
-Spain&$  401823$\tabularnewline
-Samoa&$  373370$\tabularnewline
-Singapore&$  337236$\tabularnewline
-Nepal&$  324526$\tabularnewline
-Germany&$  193081$\tabularnewline
-Japan&$  191585$\tabularnewline
-Israel&$  174177$\tabularnewline
-Italy&$  155674$\tabularnewline
-Russia&$  155632$\tabularnewline
-Pakistan&$  148514$\tabularnewline
-Sudan&$  141552$\tabularnewline
-Switzerland&$  140589$\tabularnewline
-Nauru&$  126500$\tabularnewline
-Viet Nam&$  118241$\tabularnewline
-Afghanistan&$  111854$\tabularnewline
-Somalia&$  104775$\tabularnewline
-Saudi Arabia&$   82584$\tabularnewline
-United Arab Emirates&$   67606$\tabularnewline
-Belgium&$   63784$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:risk12}
-\end{table}
-
 %%%%%%%%%%%%%%%%%%
@@
 \end{center}
+\caption{Non-New Zealander international travel and immigration, 2012 }
+\label{fig:immigration12}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-040}
+\end{center}
+\caption{New Zealander international travel, 2012 }
+\label{fig:nztravel12}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport-041}
+\end{center}
 \caption{Total international travel, 2012 }
@@
 \begin{center}
-\includegraphics{draftfinalreport-040}
-\end{center}
-\caption{New Zealander international travel, 2012 }
-\label{fig:nztravel12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport-041}
-\end{center}
-\caption{Non-New Zealander international travel and immigration, 2012 }
-\label{fig:immigration12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
 \includegraphics{draftfinalreport-042}
@@
 \includegraphics{draftfinalreport-061}
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination given alternative $R_v$ values to reduce the $R_v$ to one (Section~\ref{sec:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (Section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
      \label{fig:numvac}
@@
 \includegraphics{draftfinalreport-063}
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
 The full distribution of results can be seen in Figure~\ref{fig:sim1}}
@@
 \includegraphics{draftfinalreport-064}
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{sec:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
 A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable20, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
-\hline
-Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$209524$&$32.45$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$181417$&$11.31$\tabularnewline
-Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$158420$&$33.96$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$245296$&$17.66$\tabularnewline
-Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$127758$&$50.79$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$143089$&$11.62$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$219745$&$ 8.76$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$158420$&$ 8.08$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$191638$&$10.62$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$229965$&$ 4.80$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$178862$&$ 7.86$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$183972$&$ 2.31$\tabularnewline
-Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$260627$&$13.92$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$120093$&$ 5.06$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$173752$&$ 7.49$\tabularnewline
-Waikato&$11331$&$226620$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$242741$&$19.51$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$150755$&$ 2.26$\tabularnewline
-Waitemata&$17291$&$345820$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$178862$&$36.89$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$127758$&$ 2.51$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$148200$&$ 4.29$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:cost20}
-\begin{tablenotes}\footnotesize  
-        \item For numbers to vaccinate see Table~\ref{table:attack}
-        \item Proportion of cases hospitalised 0.17
-        \item Wage losses per case \$839 and cost per hospitalised case \$1710
-        \item Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
-\end{sidewaystable}
+\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}&\multicolumn{1}{c}{NA}\tabularnewline
+\hline
+Auckland&$17920$&$358400$&$26547468$&$55010964.560$&$5297.03$&$ 9942525.31$&$80393738.230$&$ 82$&$2115692.588$&$32.494231870$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895455.820$&$1434.29$&$ 2692162.33$&$21768412.640$&$ 71$&$1831880.168$&$11.316613160$\tabularnewline
+Canterbury&$13687$&$273740$&$21040140$&$43598824.410$&$4198.15$&$ 7879927.55$&$63715888.370$&$ 62$&$1599670.006$&$34.010648060$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15679356$&$32490348.880$&$3128.51$&$ 5872213.27$&$47481817.920$&$ 96$&$2476908.396$&$17.676674730$\tabularnewline
+Counties Manukau&$18880$&$377600$&$28033356$&$58089982.570$&$5593.51$&$10499018.27$&$84893455.150$&$ 50$&$1290056.456$&$50.905841450$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558.090$&$1163.82$&$ 2184490.14$&$17663453.000$&$ 56$&$1444863.231$&$11.621585550$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395.140$&$1332.12$&$ 2500389.24$&$20217764.780$&$ 86$&$2218897.104$&$ 8.764963262$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434.353$&$ 882.64$&$ 1656715.28$&$13395946.240$&$ 62$&$1599670.006$&$ 8.082555219$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327.040$&$1419.16$&$ 2663763.32$&$21538782.590$&$ 75$&$1935084.684$&$10.622562600$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585.118$&$ 749.87$&$ 1407505.99$&$11380878.060$&$ 90$&$2322101.621$&$ 4.803635909$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118.370$&$ 966.96$&$ 1814983.92$&$14675682.240$&$ 70$&$1806079.039$&$ 7.858468434$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962495.540$&$ 285.26$&$  535433.02$&$ 4329429.467$&$ 72$&$1857681.297$&$ 2.308362644$\tabularnewline
+Southern&$ 8371$&$167420$&$12877980$&$26685411.260$&$2569.55$&$ 4823045.35$&$38998406.670$&$102$&$2631715.170$&$13.932305620$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911.000$&$ 413.27$&$  775707.79$&$ 6272254.490$&$ 47$&$1212653.069$&$ 5.058950796$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4483224$&$ 9290018.791$&$ 894.54$&$ 1679051.58$&$13576554.140$&$ 68$&$1754476.780$&$ 7.490691249$\tabularnewline
+Waikato&$11331$&$226620$&$17251296$&$35747681.580$&$3442.16$&$ 6460934.32$&$52242126.250$&$ 95$&$2451107.267$&$19.509875740$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1146792$&$ 2376352.203$&$ 228.82$&$  429495.14$&$ 3472831.980$&$ 59$&$1522266.618$&$ 2.259977498$\tabularnewline
+Waitemata&$17291$&$345820$&$26219448$&$54331250.150$&$5231.58$&$ 9819675.66$&$79400394.760$&$ 70$&$1806079.039$&$36.897825290$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1077780$&$ 2233347.353$&$ 215.05$&$  403648.85$&$ 3263842.834$&$ 50$&$1290056.456$&$ 2.503414513$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2155560$&$ 4466694.706$&$ 430.10$&$  807297.70$&$ 6527685.668$&$ 58$&$1496465.489$&$ 4.283186675$\tabularnewline
+\hline
+\end{tabular}\end{center}\end{sidewaystable}
 
 \begin{sidewaystable}\small
-\caption{Benefit--costs analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate; \textit{Vacc.costs} is cost for the catch up vaccination programme; \textit{Wages.saved} is wages of care givers and cases saved; \textit{Manage.saved} is management costs saved; \textit{Hospitalised} is number of hospitalisations saved; \textit{Hosp.saved} is the hospitalisation costs saved; \textit{Costs.save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB.costs}	is costs expected due to continued measles importations, based on the same discounted costs as before; \textit{Benefit/cost} is the benefit--cost ratio.}
+\caption{Benefit--cost analyses with 50 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable50, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vaccines}&\multicolumn{1}{c}{Vac.costs}&\multicolumn{1}{c}{Wage.loss}&\multicolumn{1}{c}{Management}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hospitalisation}&\multicolumn{1}{c}{Costs}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB.costs}&\multicolumn{1}{c}{Benefit.cost}\tabularnewline
-\hline
-Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$9057921$&$79616516$&$ 82$&$2095239$&$26.62$\tabularnewline
-Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$2452636$&$21557962$&$ 71$&$1814170$&$10.55$\tabularnewline
-Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$7178837$&$63099902$&$ 62$&$1584205$&$27.82$\tabularnewline
-Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$5349752$&$47022778$&$ 96$&$2452962$&$15.80$\tabularnewline
-Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$9564902$&$84072731$&$ 50$&$1277585$&$37.84$\tabularnewline
-Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$1990132$&$17492688$&$ 56$&$1430895$&$10.81$\tabularnewline
-Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$2277925$&$20022305$&$ 86$&$2197445$&$ 8.28$\tabularnewline
-Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$1509314$&$13266438$&$ 62$&$1584205$&$ 7.68$\tabularnewline
-MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$2426764$&$21330552$&$ 75$&$1916377$&$ 9.93$\tabularnewline
-Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$1282278$&$11270851$&$ 90$&$2299652$&$ 4.66$\tabularnewline
-Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$1653502$&$14533802$&$ 70$&$1788618$&$ 7.48$\tabularnewline
-South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$ 487795$&$ 4287574$&$ 72$&$1839722$&$ 2.28$\tabularnewline
-Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$4393931$&$38621382$&$102$&$2606273$&$12.77$\tabularnewline
-Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$ 706692$&$ 6211616$&$ 47$&$1200930$&$ 4.90$\tabularnewline
-Taranaki&$ 2899$&$144950$&$ 4488486$&$ 9290019$&$ 895$&$1529663$&$13449923$&$ 68$&$1737515$&$ 7.14$\tabularnewline
-Waikato&$11331$&$566550$&$17291792$&$35747682$&$3442$&$5886094$&$51772645$&$ 95$&$2427411$&$17.29$\tabularnewline
-Wairarapa&$  720$&$ 36000$&$ 1150830$&$ 2376352$&$ 229$&$ 391282$&$ 3442806$&$ 59$&$1507550$&$ 2.23$\tabularnewline
-Waitemata&$17291$&$864550$&$26342544$&$54331250$&$5232$&$8946002$&$78740929$&$ 70$&$1788618$&$29.68$\tabularnewline
-West Coast&$  685$&$ 34250$&$ 1084105$&$ 2233347$&$ 215$&$ 367736$&$ 3237846$&$ 50$&$1277585$&$ 2.47$\tabularnewline
-Whanganui&$ 1378$&$ 68900$&$ 2170740$&$ 4466695$&$ 430$&$ 735471$&$ 6477915$&$ 58$&$1481998$&$ 4.18$\tabularnewline
+\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}&\multicolumn{1}{c}{NA}\tabularnewline
+\hline
+Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$ 9942525$&$80393738$&$ 82$&$2115693$&$26.69$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$ 2692162$&$21768413$&$ 71$&$1831880$&$10.56$\tabularnewline
+Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$ 7879928$&$63715888$&$ 62$&$1599670$&$27.90$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$ 5872213$&$47481818$&$ 96$&$2476908$&$15.83$\tabularnewline
+Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$10499018$&$84893455$&$ 50$&$1290056$&$38.00$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$ 2184490$&$17663453$&$ 56$&$1444863$&$10.82$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$ 2500389$&$20217765$&$ 86$&$2218897$&$ 8.29$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$ 1656715$&$13395946$&$ 62$&$1599670$&$ 7.68$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$ 2663763$&$21538783$&$ 75$&$1935085$&$ 9.94$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$ 1407506$&$11380878$&$ 90$&$2322102$&$ 4.66$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$ 1814984$&$14675682$&$ 70$&$1806079$&$ 7.49$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$  535433$&$ 4329429$&$ 72$&$1857681$&$ 2.28$\tabularnewline
+Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$ 4823045$&$38998407$&$102$&$2631715$&$12.79$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$  775708$&$ 6272254$&$ 47$&$1212653$&$ 4.90$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4483224$&$ 9290019$&$ 895$&$ 1679052$&$13576554$&$ 68$&$1754477$&$ 7.15$\tabularnewline
+Waikato&$11331$&$566550$&$17251296$&$35747682$&$3442$&$ 6460934$&$52242126$&$ 95$&$2451107$&$17.31$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1146792$&$ 2376352$&$ 229$&$  429495$&$ 3472832$&$ 59$&$1522267$&$ 2.23$\tabularnewline
+Waitemata&$17291$&$864550$&$26219448$&$54331250$&$5232$&$ 9819676$&$79400395$&$ 70$&$1806079$&$29.73$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1077780$&$ 2233347$&$ 215$&$  403649$&$ 3263843$&$ 50$&$1290056$&$ 2.46$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2155560$&$ 4466695$&$ 430$&$  807298$&$ 6527686$&$ 58$&$1496465$&$ 4.17$\tabularnewline
 \hline
 \end{tabular}\end{center}\label{table:cost50}
-\begin{tablenotes}\footnotesize  
-        \item For numbers to vaccinate see Table~\ref{table:attack}
-        \item Proportion of cases hospitalised 0.17
-        \item Wage losses per case \$839 and cost per hospitalised case \$1710
-        \item Based on 10 introductions of measles, one per year
-      \end{tablenotes}  
 \end{sidewaystable}

### 2014-11-07 draftfinalreport.Rnw
Not compiling, but text and figures all complete
Will not compile - seems to not be able to define cross references

@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see Section~\ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see section~\ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see Section~\ref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
 
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modelling and risk analyses results (Section~\ref{sub:risk_analyses} and Section~\ref{section:epidemic_modelling}).
-
-Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (Section~\ref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (Section~\ref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (Section~\ref{section:epidemic_modelling}).
-We estimate the costs of previous and the current measles outbreaks (Section~\ref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (Section~\ref{sub:cost_benefit}).
+This is particularly in light of our modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{section:epidemic_modelling}).
+
+Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (section~\ref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (section~\ref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (section~\ref{section:epidemic_modelling}).
+We estimate the costs of previous and the current measles outbreaks (section~\ref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (section~\ref{sub:cost_benefit}).
 This report includes further developments of the analyses provided in previous interim reports.
@@
 
-We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (Section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (Section~\ref{sub:immunestat}), and current rates of immunity within the population (Section~\ref{sub:immunestat}).
+We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:immunestat}), and current rates of immunity within the population (section~\ref{sub:immunestat}).
 In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (Section~\ref{{sub:imp_risk}}).
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\ref{{sub:imp_risk}}).
 
@@
      \centering
-     \includegraphics[width=1\textwidth]{incidence_1997_2014.pdf}
+     \includegraphics[width=0.9\textwidth]{incidence_1997_2014.pdf}
      \caption{Measles incidence in New Zealand from 1997 to 2014}
@@
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}, Table~\ref{table:sero}, and Section~\ref{sub:popim}).
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}, Table~\ref{table:sero}, and section~\ref{sub:popim}).
 This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}).
@@
 
-\subsection{Risk analysis discussion}
+\subsection*{Risk analysis discussion}
 
@@
 
-The distribution of na\"{\i}ve among the DHBs also varies (Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
+The distribution of na\"{\i}ve among the DHBs also varies (section~\ref{sub:popim}, Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in Section~\ref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
+Given the risk of importation of measles as highlighted in section~\ref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
 The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
@@
 The regularity of these outbreaks, however, implies that the approximately yearly importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in Section~\ref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+Given the risk of importation of measles as highlighted in section~\ref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (section~\ref{sub:cost_benefit}).
 
@@
 The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (Table~\ref{table:attack}).
-These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (Section~\ref{sub:cost_benefit}).
+These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (section~\ref{sub:cost_benefit}).
 However, they require differing numbers of vaccinations per DHB.
@@
 
-To estimate the benefits from additional vaccinations, as estimated from the above modelling section (Section~\ref{section:epidemic_modelling}), we carried out a number of analyses.
-Primarily, we used Equation~\ref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (Section~\ref{section:epidemic_modelling}).
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section (section~\ref{section:epidemic_modelling}), we carried out a number of analyses.
+Primarily, we used Equation~\ref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (section~\ref{section:epidemic_modelling}).
 
@@
 @
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (Section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
      \label{fig:numvac}
@@
 @
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
 The full distribution of results can be seen in Figure~\ref{fig:sim1}}
@@
 @
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
 A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
@@
 
-For the cost analyses we used the values from the above cost section (Section~\ref{sub:cost_benefit}).
+For the cost analyses we used the values from the above cost section (section~\ref{sub:cost_benefit}).
 Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
-We estimated there would be approximately one introduction of measles per year (Section~\ref{sub:risk_analyses}).
+We estimated there would be approximately one introduction of measles per year (section~\ref{sub:risk_analyses}).
 We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
@@
 Finally, our model of measles introductions remains a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local na\"{\i}ve population.
-The spatial effects of measles transmission may have affected both our multivariate regression analyses (Section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (Section~\ref{section:epidemic_modelling}).
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure ~\ref{fig:sim1}).
+The spatial effects of measles transmission may have affected both our multivariate regression analyses (section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (section~\ref{section:epidemic_modelling}).
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}).
 Given that, the greater the vaccination coverage, the smaller the outbreaks will be.

### 2014-11-07 draftfinalreport.tex
Not compiling, but text and figures all complete
Will not compile - seems to not be able to define cross references

@@
 % latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Sat Nov 08 00:20:12 2014
+% Sat Nov 08 01:33:38 2014
 \begin{tabular}{lrrrrrr}
@@
 % latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Sat Nov 08 00:20:12 2014
+% Sat Nov 08 01:33:38 2014
 \begin{tabular}{rrrrr}
@@
 \includegraphics{draftfinalreport-061}
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (Section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
      \label{fig:numvac}
@@
 \includegraphics{draftfinalreport-063}
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
 The full distribution of results can be seen in Figure~\ref{fig:sim1}}
@@
 \includegraphics{draftfinalreport-064}
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (Section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
 A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}

### 2014-11-07 draftfinalreport.Rnw
hmmm

@@
 \usepackage{natbib}
-\usepackage{graphicx}
 \usepackage{graphics}
-\usepackage{subfig}
 \usepackage{amsmath}
@@
 \usepackage[utf8]{inputenc}
-\usepackage{enumitem}
 \usepackage{graphicx}
@@
 \usepackage{todonotes}
+\usepackage{subfig}
+\usepackage{enumitem}
 \usepackage{pdflscape}
@@
 \usepackage{amsmath}
-\usepackage{graphicx}
 %\usepackage{caption2}
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see section~\ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see section~\autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\autoref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
 
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{section:epidemic_modelling}).
-
-Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (section~\ref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (section~\ref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (section~\ref{section:epidemic_modelling}).
-We estimate the costs of previous and the current measles outbreaks (section~\ref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (section~\ref{sub:cost_benefit}).
+This is particularly in light of our modelling and risk analyses results (section~\autoref{sub:risk_analyses} and section~\autoref{section:epidemic_modelling}).
+
+Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (section~\autoref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (section~\autoref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (section~\autoref{section:epidemic_modelling}).
+We estimate the costs of previous and the current measles outbreaks (section~\autoref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (section~\autoref{sub:cost_benefit}).
 This report includes further developments of the analyses provided in previous interim reports.
@@
 
-We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:immunestat}), and current rates of immunity within the population (section~\ref{sub:immunestat}).
-In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\ref{{sub:imp_risk}}).
+We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\autoref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\autoref{sub:immunestat}), and current rates of immunity within the population (section~\autoref{sub:immunestat}).
+In light of the apparent increasing trend in measles incidence in the last few years (Figure~\autoref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\autoref{{sub:imp_risk}}).
 
@@
 With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
+Because for measles cases the very young appear to be disproportionately affected (Figure~\autoref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
 
@@
 The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}).
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\autoref{table:none}).
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
 
-The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
+The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\autoref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\autoref{fig:percap}, Table~\autoref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\autoref{table:percap}).
 
@@
 \end{center}
-\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
+\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (Table~\autoref{table:percap} for details)}
 \label{fig:percap}
@@
 To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work.
-Specifically, we use the data in Table \ref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.
+Specifically, we use the data in Table \autoref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.
 
@@
 
-The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
+The distribution of the meases cases per category used in the regression analyses are in Figure~\autoref{fig:histcasecat}.
 \begin{figure}
@@
 \end{center}
-\caption{Distribution of measles cases per category (Table~\ref{table:percap}) used in the final regression model (Equation~\ref{eq:reg})}
+\caption{Distribution of measles cases per category (Table~\autoref{table:percap}) used in the final regression model (Equation~\autoref{eq:reg})}
 \label{fig:histcasecat}
@@
 
-The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
+The predicted values from the regression model plotted against the reported cases are shown in Figure~\autoref{fig:predict}, and the residuals are shown in Figure~\autoref{fig:resid}.
 \begin{figure}
@@
 \end{center}
-\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the case data (Table~\ref{table:percap})}
+\caption{Regression model (Equation~\autoref{eq:reg}) predictions plotted against the case data (Table~\autoref{table:percap})}
 \label{fig:predict}
@@
 \end{center}
-\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
+\caption{Histogram of residuals from the regression model (Equation~\autoref{eq:reg})}
 \label{fig:resid}
@@
 
-The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
+The significance of the different predictor variables can be seen in the ANOVA results (Table~\autoref{table:anova}).
 \par
@@
 @
-\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
+\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\autoref{eq:reg} and a quasipoisson error structure}
 \label{table:anova}
@@
 
-A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.
+A summary of the regression model (Equation~\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\autoref{table:regression}.
 
@@
 \par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also Figure~\ref{fig:ageinyears}).
-
-The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also Figure~\autoref{fig:ageinyears}).
+
+The majority of measles cases are among people of European ancestry (Table~\autoref{table:percap} and Figure~\autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
 Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}, Table~\ref{table:sero}, and section~\ref{sub:popim}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}).
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\autoref{fig:ageinyears} and ~\autoref{fig:ageandvac}, Table~\autoref{table:sero}, and section~\autoref{sub:popim}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\autoref{table:percap}).
 Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
@@
 
-The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (Figure~\ref{fig:ageandvac}).
+The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (Figure~\autoref{fig:ageandvac}).
 However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second.
-A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (Figure~\ref{fig:vaccstat}).
+A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (Figure~\autoref{fig:vaccstat}).
 
@@
 
-The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
-However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (Figure~\ref{fig:naive}).
-Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
+The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (Figure~\autoref{fig:naive}).
+However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (Figure~\autoref{fig:naive}).
+Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (Figure~\autoref{fig:naive_map}).
 The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures:
 \begin{itemize}[noitemsep,nolistsep]
-\item Northland: Figure~\ref{fig:Northland}
-\item Waitemata: Figure~\ref{fig:Waitemata}
-\item Auckland: Figure~\ref{fig:Auckland}
-\item Counties Manukau: Figure~\ref{fig:Counties_Manukau}
-\item Waikato: Figure~\ref{fig:Waikato}
-\item Lakes: Figure~\ref{fig:Lakes}
-\item Bay of Plenty: Figure~\ref{fig:BayofPlenty}
-\item Tairawhiti: Figure~\ref{fig:Tairawhiti}
-\item Taranaki: Figure~\ref{fig:Taranaki}
-\item Hawke's Bay: Figure~\ref{fig:HawkesBay}
-\item Whanganui: Figure~\ref{fig:Whanganui}
-\item Mid-Central: Figure~\ref{fig:Midcentral}
-\item Hutt: Figure~\ref{fig:Hutt}
-\item Capital and Coast: Figure~\ref{fig:CapitalandCoast}
-\item Wairarapa: Figure~\ref{fig:Wairarapa}
-\item Nelson Marlborough: Figure~\ref{fig:NelsonMarlborough}
-\item West Coast: Figure~\ref{fig:WestCoast}
-\item Canterbury: Figure~\ref{fig:Canterbury}
-\item South Canterbury: Figure~\ref{fig:SouthCanterbury}
-\item Southern: Figure~\ref{fig:Southern}
+\item Northland: Figure~\autoref{fig:Northland}
+\item Waitemata: Figure~\autoref{fig:Waitemata}
+\item Auckland: Figure~\autoref{fig:Auckland}
+\item Counties Manukau: Figure~\autoref{fig:Counties_Manukau}
+\item Waikato: Figure~\autoref{fig:Waikato}
+\item Lakes: Figure~\autoref{fig:Lakes}
+\item Bay of Plenty: Figure~\autoref{fig:BayofPlenty}
+\item Tairawhiti: Figure~\autoref{fig:Tairawhiti}
+\item Taranaki: Figure~\autoref{fig:Taranaki}
+\item Hawke's Bay: Figure~\autoref{fig:HawkesBay}
+\item Whanganui: Figure~\autoref{fig:Whanganui}
+\item Mid-Central: Figure~\autoref{fig:Midcentral}
+\item Hutt: Figure~\autoref{fig:Hutt}
+\item Capital and Coast: Figure~\autoref{fig:CapitalandCoast}
+\item Wairarapa: Figure~\autoref{fig:Wairarapa}
+\item Nelson Marlborough: Figure~\autoref{fig:NelsonMarlborough}
+\item West Coast: Figure~\autoref{fig:WestCoast}
+\item Canterbury: Figure~\autoref{fig:Canterbury}
+\item South Canterbury: Figure~\autoref{fig:SouthCanterbury}
+\item Southern: Figure~\autoref{fig:Southern}
 \end{itemize}
@@
 \end{center}
-\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (Table~\ref{table:sero})}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (Table~\autoref{table:sero})}
 \label{fig:naive}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:naive_map}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Northland}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Waitemata}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Auckland}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Counties_Manukau}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Waikato}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Lakes}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:BayofPlenty}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Tairawhiti}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Taranaki}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:HawkesBay}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Whanganui}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Midcentral}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Hutt}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:CapitalandCoast}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Wairarapa}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:NelsonMarlborough}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:WestCoast}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Canterbury}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:SouthCanterbury}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (Table~\autoref{table:sero})}
      \label{fig:Southern}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\autoref{table:nzrisk12} and ~\autoref{table:imrisk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:imrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\autoref{table:imrisk12} and ~\autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\autoref{table:nzrisk12} and ~\autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 
-Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (Tables~\ref{table:incidence12} and ~\ref{table:cover12}, and Figures~\ref{fig:incidence12} and ~\ref{fig:cover12}).
-Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}.
-However, travel by New Zealanders is dominated by that to Australia (Tables~\ref{table:nztravel12} and \ref{fig:nztravel12}).
-Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}.
+Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (Tables~\autoref{table:incidence12} and ~\autoref{table:cover12}, and Figures~\autoref{fig:incidence12} and ~\autoref{fig:cover12}).
+Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \autoref{table:imtravel12} and Figure~\autoref{fig:immigration12}.
+However, travel by New Zealanders is dominated by that to Australia (Tables~\autoref{table:nztravel12} and \autoref{fig:nztravel12}).
+Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (Table~\autoref{table:immigration12} and Figure~\autoref{fig:immigration12}.
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation.
 Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand.
-These data are shown in Table~\ref{table:risk12}.
-The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.
-
-The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in Table~\ref{table:toprisk}.
+These data are shown in Table~\autoref{table:risk12}.
+The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in Tables~\autoref{table:nzrisk12} and ~\autoref{table:imrisk12} and Figures~\autoref{fig:nzrisk12} and ~\autoref{fig:imrisk12}.
+
+The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in Table~\autoref{table:toprisk}.
 
@@
 
-Though the global incidence of measles is declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:travel}).
+Though the global incidence of measles is declining, in recent years that decline has slowed (Figure~\autoref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\autoref{fig:travel}).
 This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
-Of note, however, is the clear seasonality in immigration and travel (Figure~\ref{fig:travel}).
+Of note, however, is the clear seasonality in immigration and travel (Figure~\autoref{fig:travel}).
 This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
 Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling.
-The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (Figure~\ref{fig:travel}).
+The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (Figure~\autoref{fig:travel}).
 
@@
 
-The distribution of the measles cases (Figure~\ref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (Figure~\ref{fig:naive}).
+The distribution of the measles cases (Figure~\autoref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (Figure~\autoref{fig:naive}).
 The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number ($>16\%$) of cases had been vaccinated at least once.
-However, the majority of these received only one vaccination, and were vaccinated when they were young (\ref{fig:vaccstat}).
-The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (Figure~\ref{fig:naive_map}).
+However, the majority of these received only one vaccination, and were vaccinated when they were young (\autoref{fig:vaccstat}).
+The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (Figure~\autoref{fig:naive_map}).
 
@@
 
-The distribution of na\"{\i}ve among the DHBs also varies (section~\ref{sub:popim}, Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
+The distribution of na\"{\i}ve among the DHBs also varies (section~\autoref{sub:popim}, Figures~\autoref{fig:Northland}, ~\autoref{fig:Waitemata}, ~\autoref{fig:Auckland}, ~\autoref{fig:Counties_Manukau}, ~\autoref{fig:Waikato}, ~\autoref{fig:Lakes}, ~\autoref{fig:BayofPlenty}, ~\autoref{fig:Tairawhiti}, ~\autoref{fig:Taranaki}, ~\autoref{fig:HawkesBay}, ~\autoref{fig:Whanganui}, ~\autoref{fig:Midcentral}, ~\autoref{fig:Hutt}, ~\autoref{fig:CapitalandCoast}, ~\autoref{fig:Wairarapa}, ~\autoref{fig:NelsonMarlborough}, ~\autoref{fig:WestCoast}, ~\autoref{fig:Canterbury}, ~\autoref{fig:SouthCanterbury}, ~\autoref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
-The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}).
+The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\autoref{table:immigration12} and Figure~\autoref{fig:immigration12}).
 However, recently Australia was declared free of endemic measles and so the 2012 data must reflect imported measles there.
-However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (Table~\ref{table:risk12} and Figure~\ref{fig:risk12}).
+However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (Table~\autoref{table:risk12} and Figure~\autoref{fig:risk12}).
 
@@
 We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
-The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\autoref{fig:dhbcases}).
+The outbreaks we used in our analyses are shown in Figure~\autoref{fig:outbreaks}.
 
@@
      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in Figure~\ref{fig:outbreaks}. Sporadic cases are excluded.}
+     \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in Figure~\autoref{fig:outbreaks}. Sporadic cases are excluded.}
      \label{fig:r0}
@@
 
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The estimated $R_v$ for each outbreak is shown in Figure~\autoref{fig:r0}.
 The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current 2013--2014 outbreak 1.82--2.13.
@@
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
+Given the risk of importation of measles as highlighted in section~\autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
 The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
@@
 The regularity of these outbreaks, however, implies that the approximately yearly importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (section~\ref{sub:cost_benefit}).
-
-To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:prop}.
-The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (Table~\ref{table:attack}).
-These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (section~\ref{sub:cost_benefit}).
+Given the risk of importation of measles as highlighted in section~\autoref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (section~\autoref{sub:cost_benefit}).
+
+To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\autoref{eq:prop}.
+The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (Table~\autoref{table:attack}).
+These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (section~\autoref{sub:cost_benefit}).
 However, they require differing numbers of vaccinations per DHB.
-Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (Table~\ref{table:attack}).
-The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (Table~\ref{table:attack}).
-
-The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
-However, the na\"{\i}ve population (Figures~\ref{fig:naive} and \ref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary (Table~\ref{table:attack}).
+Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (Table~\autoref{table:attack}).
+The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (Table~\autoref{table:attack}).
+
+The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\autoref{fig:r0}).
+However, the na\"{\i}ve population (Figures~\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\autoref{fig:r0}) suggests that catch up vaccination may be necessary (Table~\autoref{table:attack}).
 The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics.
@@
 \begin{itemize}
-\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\autoref{sub:risk_analyses}).
 \item The reproduction number for measles in the partially immune New Zealand population is often close to one, suggesting increased population level immunity is required to ensure prevention of measles persistence following importation.
@@
 
-Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
-The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\ref{table:hosp}.
+Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\autoref{table:direct}.
+The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\autoref{table:hosp}.
 
@@
 
-The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (Table~\ref{table:hosp}).
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (Table~\autoref{table:hosp}).
 The mean cost per case was \$1,877.
@@
 From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before 1 January 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
-The number of cases by age group is shown in Table~\ref{table:freq}.
+The number of cases by age group is shown in Table~\autoref{table:freq}.
 Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.
@@
 
-The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (Figure~\ref{fig:hosp}).
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (Figure~\autoref{fig:hosp}).
 
@@
 Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days.
-The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\autoref{table:cases}.
 
@@
 \caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\ref{table:regression})}
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\autoref{table:regression})}
 <<results=tex,echo=FALSE>>=
@@
 
-Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
+Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\autoref{table:regression}).
 Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.
@@
 
-To estimate the benefits from additional vaccinations, as estimated from the above modelling section (section~\ref{section:epidemic_modelling}), we carried out a number of analyses.
-Primarily, we used Equation~\ref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (section~\ref{section:epidemic_modelling}).
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section (section~\autoref{section:epidemic_modelling}), we carried out a number of analyses.
+Primarily, we used Equation~\autoref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\autoref{fig:naive_map} and Table~\autoref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (section~\autoref{section:epidemic_modelling}).
 
@@
 
-The estimated vaccinaton rates, with percentages are shown in Table~\ref{table:vacp}.
+The estimated vaccinaton rates, with percentages are shown in Table~\autoref{table:vacp}.
 <<echo=FALSE>>=
@@
 
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using equation~\ref{eq:prop} are shown in Figure~\ref{fig:numvac}.
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using equation~\autoref{eq:prop} are shown in Figure~\autoref{fig:numvac}.
 
@@
 @
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\autoref{section:epidemic_modelling}, Tables~\autoref{table:cost20} and ~\autoref{table:cost50})}
      \label{fig:numvac}
@@
 
-The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ is <1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
+The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ is <1 is shown in Figure~\autoref{fig:sim} and Figure~\autoref{fig:sim1}.
 These simulations show that even in scenarios when $R_v$ is lower than one, and thus deterministically should fail to persist (i.e.
@@
 @
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
-The full distribution of results can be seen in Figure~\ref{fig:sim1}}
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+The full distribution of results can be seen in Figure~\autoref{fig:sim1}}
      \label{fig:sim}
@@
 @
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
-A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+A subset of the distribution of results can be seen in Figure~\autoref{fig:sim}}
      \label{fig:sim1}
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see Table~\autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 \caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see Table~\autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 
-For the cost analyses we used the values from the above cost section (section~\ref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
-We estimated there would be approximately one introduction of measles per year (section~\ref{sub:risk_analyses}).
+For the cost analyses we used the values from the above cost section (section~\autoref{sub:cost_benefit}).
+Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\autoref{table:hosp}).
+We estimated there would be approximately one introduction of measles per year (section~\autoref{sub:risk_analyses}).
 We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
 
-The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\ref{table:vacp}.
-The benefit--cost results are in Table~\ref{table:cost20} and Table~\ref{table:cost50}, for two different vaccine prices.
-The results in the two tables show the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Tables~\ref{table:cost20} and \ref{table:cost50}).
-
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
+The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\autoref{table:vacp}.
+The benefit--cost results are in Table~\autoref{table:cost20} and Table~\autoref{table:cost50}, for two different vaccine prices.
+The results in the two tables show the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Tables~\autoref{table:cost20} and \autoref{table:cost50}).
+
+It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\autoref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\autoref{fig:vaccstat}).
 Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children.
@@
 We used the mean values per case in the absence of alternative data and used as much data as possible.
-The exact costs will vary, however, our benefit--cost analyses suggest that in all cases catch up vaccination schemes will be financially beneficial for New Zealand (see Tables~\ref{table:cost20} and \ref{table:cost50}).
+The exact costs will vary, however, our benefit--cost analyses suggest that in all cases catch up vaccination schemes will be financially beneficial for New Zealand (see Tables~\autoref{table:cost20} and \autoref{table:cost50}).
 
@@
 They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
-Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see Table~\ref{table:zhou}) \citep{zhou4}.
-However, the cost of measles in their analyses was substantially higher than mumps and rubella (Table~\ref{table:zhou}) \citep{zhou4}.
+Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see Table~\autoref{table:zhou}) \citep{zhou4}.
+However, the cost of measles in their analyses was substantially higher than mumps and rubella (Table~\autoref{table:zhou}) \citep{zhou4}.
 
@@
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
-However,  the estimated B/C ratio for MMR has been estimated to be between 10.2 and 54.2, much greater than one (see Table~\ref{table:zhou2}) \citep{zhou4}.
+However,  the estimated B/C ratio for MMR has been estimated to be between 10.2 and 54.2, much greater than one (see Table~\autoref{table:zhou2}) \citep{zhou4}.
 As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis.
@@
 It is unlikely $R_0$ was so high as 18 in New Zealand, as measles would still be endemic given current vaccination rates.
-However, the benefits of catch up vaccination are clear if $R_v$ is greater than one (Tables~\ref{table:cost20} and ~\ref{table:cost20}) and our analyses do not include the additional benefits of mumps and rubella immunity.
+However, the benefits of catch up vaccination are clear if $R_v$ is greater than one (Tables~\autoref{table:cost20} and ~\autoref{table:cost20}) and our analyses do not include the additional benefits of mumps and rubella immunity.
 As noted in our previous report, if measles $R_v$ in the currently na\"{\i}ve population is less than one, the benefits will not be so clear, though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
@@
 Finally, our model of measles introductions remains a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local na\"{\i}ve population.
-The spatial effects of measles transmission may have affected both our multivariate regression analyses (section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (section~\ref{section:epidemic_modelling}).
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}).
+The spatial effects of measles transmission may have affected both our multivariate regression analyses (section~\autoref{sub:risk_analyses}) and will affect the predictions from modelling exercises (section~\autoref{section:epidemic_modelling}).
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\autoref{fig:sim} and Figure~\autoref{fig:sim1}).
 Given that, the greater the vaccination coverage, the smaller the outbreaks will be.

### 2014-11-07 draftfinalreport.tex
hmmm

@@
 The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}).
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\autoref{table:none}).
 
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Sat Nov 08 01:33:38 2014
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Sat Nov 08 03:43:34 2014
 \begin{tabular}{lrrrrrr}
@@
    \hline
-\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
+\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\autoref{eq:reg} and a quasipoisson error structure}
 \label{table:anova}
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Sat Nov 08 01:33:38 2014
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Sat Nov 08 03:43:34 2014
 \begin{tabular}{rrrrr}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\autoref{table:nzrisk12} and ~\autoref{table:imrisk12}}
 %latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:imrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\autoref{table:imrisk12} and ~\autoref{table:risk12}}
 %latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\autoref{table:nzrisk12} and ~\autoref{table:risk12}}
 %latex.default(imrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\ref{table:regression})}
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\autoref{table:regression})}
 %latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-The estimated vaccinaton rates, with percentages are shown in Table~\ref{table:vacp}.
+The estimated vaccinaton rates, with percentages are shown in Table~\autoref{table:vacp}.
 
@@
 \includegraphics{draftfinalreport-061}
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\autoref{section:epidemic_modelling}, Tables~\autoref{table:cost20} and ~\autoref{table:cost50})}
      \label{fig:numvac}
@@
 \includegraphics{draftfinalreport-063}
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
-The full distribution of results can be seen in Figure~\ref{fig:sim1}}
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+The full distribution of results can be seen in Figure~\autoref{fig:sim1}}
      \label{fig:sim}
@@
 \includegraphics{draftfinalreport-064}
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
-A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+A subset of the distribution of results can be seen in Figure~\autoref{fig:sim}}
      \label{fig:sim1}
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see Table~\autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable20, file = "", table.env = FALSE, rowname = NULL)%
@@
 \caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see Table~\autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable50, file = "", table.env = FALSE, rowname = NULL)%

### 2014-11-09 draftfinalreport.Rnw
fix duplicate references
fix missing label ref
fix missing \end{center}
\autoref doesn't need prefixes - it's magic

--- 2014-11-07 draftfinalreport.Rnw
@@
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
-Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see section~\ref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
+Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see \autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
 Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see section~\ref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see \autoref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
 
@@
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modelling and risk analyses results (section~\ref{sub:risk_analyses} and section~\ref{section:epidemic_modelling}).
-
-Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (section~\ref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (section~\ref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (section~\ref{section:epidemic_modelling}).
-We estimate the costs of previous and the current measles outbreaks (section~\ref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (section~\ref{sub:cost_benefit}).
+This is particularly in light of our modelling and risk analyses results (\autoref{sub:risk_analyses} and \autoref{section:epidemic_modelling}).
+
+Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (\autoref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (\autoref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (\autoref{section:epidemic_modelling}).
+We estimate the costs of previous and the current measles outbreaks (\autoref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (\autoref{sub:cost_benefit}).
 This report includes further developments of the analyses provided in previous interim reports.
@@
 
-We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (section~\ref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (section~\ref{sub:immunestat}), and current rates of immunity within the population (section~\ref{sub:immunestat}).
-In light of the apparent increasing trend in measles incidence in the last few years (Figure~\ref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
-To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (section~\ref{{sub:imp_risk}}).
+We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (\autoref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (\autoref{sub:immunestat}), and current rates of immunity within the population (\autoref{sub:immunestat}).
+In light of the apparent increasing trend in measles incidence in the last few years (\autoref{fig:incidence1997}), we reviewed the information on measles importation and the origins of the introductions of measles into New Zealand.
+To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (\autoref{sub:imp_risk}).
 
@@
 With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (Figure~\ref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
+Because for measles cases the very young appear to be disproportionately affected (\autoref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
 
@@
 The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}).
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, \autoref{table:none}).
 <<echo=FALSE,include=FALSE,results=hide>>=
@@
 
-The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in Table~\ref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (Figure~\ref{fig:percap}, Table~\ref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see Table~\ref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
+The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap}, \autoref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (Table~\ref{table:percap}).
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
 
@@
 \end{center}
-\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (Table~\ref{table:percap} for details)}
+\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (\autoref{table:percap} for details)}
 \label{fig:percap}
@@
 
-The distribution of the meases cases per category used in the regression analyses are in Figure~\ref{fig:histcasecat}.
+The distribution of the meases cases per category used in the regression analyses are in \autoref{fig:histcasecat}.
 \begin{figure}
@@
 \end{center}
-\caption{Distribution of measles cases per category (Table~\ref{table:percap}) used in the final regression model (Equation~\ref{eq:reg})}
+\caption{Distribution of measles cases per category (\autoref{table:percap}) used in the final regression model (\autoref{eq:reg})}
 \label{fig:histcasecat}
@@
 
-The predicted values from the regression model plotted against the reported cases are shown in Figure~\ref{fig:predict}, and the residuals are shown in Figure~\ref{fig:resid}.
+The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}.
 \begin{figure}
@@
 \end{center}
-\caption{Regression model (Equation~\ref{eq:reg}) predictions plotted against the case data (Table~\ref{table:percap})}
+\caption{Regression model (\autoref{eq:reg}) predictions plotted against the case data (\autoref{table:percap})}
 \label{fig:predict}
@@
 \end{center}
-\caption{Histogram of residuals from the regression model (Equation~\ref{eq:reg})}
+\caption{Histogram of residuals from the regression model (\autoref{eq:reg})}
 \label{fig:resid}
@@
 
-The significance of the different predictor variables can be seen in the ANOVA results (Table~\ref{table:anova}).
+The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
 \par
@@
 @
-\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
+\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using \autoref{eq:reg} and a quasipoisson error structure}
 \label{table:anova}
@@
 
-A summary of the regression model (Equation~\ref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in Table~\ref{table:regression}.
+A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.
 
@@
 \caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
-\label{table:regression}
+\label{table:regression_riskfactors}
 \end{table}
@@
 \par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also Figure~\ref{fig:ageinyears}).
-
-The majority of measles cases are among people of European ancestry (Table~\ref{table:percap} and Figure~\ref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also \autoref{fig:ageinyears}).
+
+The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
 Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (Figures~\ref{fig:ageinyears} and ~\ref{fig:ageandvac}, Table~\ref{table:sero}, and section~\ref{sub:popim}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (Table~\ref{table:percap}).
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}, \autoref{table:sero}, and \autoref{sub:popim}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}).
 Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
@@
 
-The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (Figure~\ref{fig:ageandvac}).
+The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (\autoref{fig:ageandvac}).
 However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second.
-A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (Figure~\ref{fig:vaccstat}).
+A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
 
@@
 
-The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (Figure~\ref{fig:naive}).
-However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (Figure~\ref{fig:naive}).
-Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (Figure~\ref{fig:naive_map}).
+The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (\autoref{fig:naive}).
+However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (\autoref{fig:naive}).
+Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (\autoref{fig:naive_map}).
 The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures:
 \begin{itemize}[noitemsep,nolistsep]
-\item Northland: Figure~\ref{fig:Northland}
-\item Waitemata: Figure~\ref{fig:Waitemata}
-\item Auckland: Figure~\ref{fig:Auckland}
-\item Counties Manukau: Figure~\ref{fig:Counties_Manukau}
-\item Waikato: Figure~\ref{fig:Waikato}
-\item Lakes: Figure~\ref{fig:Lakes}
-\item Bay of Plenty: Figure~\ref{fig:BayofPlenty}
-\item Tairawhiti: Figure~\ref{fig:Tairawhiti}
-\item Taranaki: Figure~\ref{fig:Taranaki}
-\item Hawke's Bay: Figure~\ref{fig:HawkesBay}
-\item Whanganui: Figure~\ref{fig:Whanganui}
-\item Mid-Central: Figure~\ref{fig:Midcentral}
-\item Hutt: Figure~\ref{fig:Hutt}
-\item Capital and Coast: Figure~\ref{fig:CapitalandCoast}
-\item Wairarapa: Figure~\ref{fig:Wairarapa}
-\item Nelson Marlborough: Figure~\ref{fig:NelsonMarlborough}
-\item West Coast: Figure~\ref{fig:WestCoast}
-\item Canterbury: Figure~\ref{fig:Canterbury}
-\item South Canterbury: Figure~\ref{fig:SouthCanterbury}
-\item Southern: Figure~\ref{fig:Southern}
+\item Northland: \autoref{fig:Northland}
+\item Waitemata: \autoref{fig:Waitemata}
+\item Auckland: \autoref{fig:Auckland}
+\item Counties Manukau: \autoref{fig:Counties_Manukau}
+\item Waikato: \autoref{fig:Waikato}
+\item Lakes: \autoref{fig:Lakes}
+\item Bay of Plenty: \autoref{fig:BayofPlenty}
+\item Tairawhiti: \autoref{fig:Tairawhiti}
+\item Taranaki: \autoref{fig:Taranaki}
+\item Hawke's Bay: \autoref{fig:HawkesBay}
+\item Whanganui: \autoref{fig:Whanganui}
+\item Mid-Central: \autoref{fig:Midcentral}
+\item Hutt: \autoref{fig:Hutt}
+\item Capital and Coast: \autoref{fig:CapitalandCoast}
+\item Wairarapa: \autoref{fig:Wairarapa}
+\item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
+\item West Coast: \autoref{fig:WestCoast}
+\item Canterbury: \autoref{fig:Canterbury}
+\item South Canterbury: \autoref{fig:SouthCanterbury}
+\item Southern: \autoref{fig:Southern}
 \end{itemize}
@@
 \end{center}
-\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (Table~\ref{table:sero})}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (\autoref{table:sero})}
 \label{fig:naive}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:naive_map}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Northland}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Waitemata}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Auckland}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Counties_Manukau}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Waikato}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Lakes}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:BayofPlenty}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Tairawhiti}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Taranaki}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:HawkesBay}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Whanganui}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Midcentral}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Hutt}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:CapitalandCoast}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Wairarapa}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:NelsonMarlborough}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:WestCoast}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Canterbury}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:SouthCanterbury}
@@
      \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (Table~\ref{table:sero})}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (\autoref{table:sero})}
      \label{fig:Southern}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:imrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 
-Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (Tables~\ref{table:incidence12} and ~\ref{table:cover12}, and Figures~\ref{fig:incidence12} and ~\ref{fig:cover12}).
-Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \ref{table:imtravel12} and Figure~\ref{fig:immigration12}.
-However, travel by New Zealanders is dominated by that to Australia (Tables~\ref{table:nztravel12} and \ref{fig:nztravel12}).
-Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}.
+Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}).
+Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \autoref{table:imtravel12} and \autoref{fig:immigration12}.
+However, travel by New Zealanders is dominated by that to Australia (\autoref{table:nztravel12} and \autoref{fig:nztravel12}).
+Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\autoref{table:immigration12} and \autoref{fig:immigration12}.
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation.
 Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand.
-These data are shown in Table~\ref{table:risk12}.
-The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12} and Figures~\ref{fig:nzrisk12} and ~\ref{fig:imrisk12}.
-
-The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in Table~\ref{table:toprisk}.
+These data are shown in \autoref{table:risk12}.
+The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in \autoref{table:nzrisk12} and \autoref{table:imrisk12} and \autoref{fig:nzrisk12} and \autoref{fig:imrisk12}.
+
+The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in \autoref{table:toprisk}.
 
@@
      \includegraphics[width=0.9\textwidth]{nzers.pdf}
+     \end{center}
      \caption{Trends in international travel, determined by immigration arrival data and New Zealander travel departure destinations}
@@
 
-Though the global incidence of measles is declining, in recent years that decline has slowed (Figure~\ref{fig:trendincidence}) and immigration rates to New Zealand have risen (Figure~\ref{fig:travel}).
+Though the global incidence of measles is declining, in recent years that decline has slowed (\autoref{fig:trendincidence}) and immigration rates to New Zealand have risen (\autoref{fig:travel}).
 This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
-Of note, however, is the clear seasonality in immigration and travel (Figure~\ref{fig:travel}).
+Of note, however, is the clear seasonality in immigration and travel (\autoref{fig:travel}).
 This seasonality suggests that there may be period of increased risk of measles importation, though again the interactions with seasonal measles transmission from the nations of origin will be an important factor in determining the risk of measles importation.
 Of interest is the asynchrony, or lack of phase, between peak non-New Zealander immigration and New Zealanders travelling.
-The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (Figure~\ref{fig:travel}).
+The Christmas summer period being an exception to this, when there is substantial New Zealander travel and combined these lead to a very sharp peak in travel rates in December (\autoref{fig:travel}).
 
@@
 
-The distribution of the measles cases (Figure~\ref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (Figure~\ref{fig:naive}).
+The distribution of the measles cases (\autoref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (\autoref{fig:naive}).
 The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number ($>16\%$) of cases had been vaccinated at least once.
-However, the majority of these received only one vaccination, and were vaccinated when they were young (\ref{fig:vaccstat}).
-The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (Figure~\ref{fig:naive_map}).
+However, the majority of these received only one vaccination, and were vaccinated when they were young (\autoref{fig:vaccstat}).
+The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (\autoref{fig:naive_map}).
 
@@
 
-The distribution of na\"{\i}ve among the DHBs also varies (section~\ref{sub:popim}, Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
+The distribution of na\"{\i}ve among the DHBs also varies (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, ~autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
-The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (Table~\ref{table:immigration12} and Figure~\ref{fig:immigration12}).
+The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}).
 However, recently Australia was declared free of endemic measles and so the 2012 data must reflect imported measles there.
-However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (Table~\ref{table:risk12} and Figure~\ref{fig:risk12}).
+However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}).
 
@@
 We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (Figure~\ref{fig:dhbcases}).
-The outbreaks we used in our analyses are shown in Figure~\ref{fig:outbreaks}.
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (\autoref{fig:dhbcases}).
+The outbreaks we used in our analyses are shown in \autoref{fig:outbreaks}.
 
@@
      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-     \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in Figure~\ref{fig:outbreaks}. Sporadic cases are excluded.}
+     \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in \autoref{fig:outbreaks}. Sporadic cases are excluded.}
      \label{fig:r0}
@@
 
-The estimated $R_v$ for each outbreak is shown in Figure~\ref{fig:r0}.
+The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}.
 The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current 2013--2014 outbreak 1.82--2.13.
@@
 These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
+Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
 The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
@@
 The regularity of these outbreaks, however, implies that the approximately yearly importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in section~\ref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (section~\ref{sub:cost_benefit}).
-
-To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use Equation~\ref{eq:prop}.
-The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (Table~\ref{table:attack}).
-These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (section~\ref{sub:cost_benefit}).
+Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (\autoref{sub:cost_benefit}).
+
+To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use \autoref{eq:prop}.
+The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (\autoref{table:attack}).
+These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (\autoref{sub:cost_benefit}).
 However, they require differing numbers of vaccinations per DHB.
-Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (Table~\ref{table:attack}).
-The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (Table~\ref{table:attack}).
-
-The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (Figure~\ref{fig:r0}).
-However, the na\"{\i}ve population (Figures~\ref{fig:naive} and \ref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (Figure~\ref{fig:r0}) suggests that catch up vaccination may be necessary (Table~\ref{table:attack}).
+Estimates for West Coast, Wairarapa, and South Canterbury, for example, are fewer than 1000 (\autoref{table:attack}).
+The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (\autoref{table:attack}).
+
+The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (\autoref{fig:r0}).
+However, the na\"{\i}ve population (\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (\autoref{fig:r0}) suggests that catch up vaccination may be necessary (\autoref{table:attack}).
 The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics.
@@
 \begin{itemize}
-\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see section~\ref{sub:risk_analyses}).
+\item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see \autoref{sub:risk_analyses}).
 \item The reproduction number for measles in the partially immune New Zealand population is often close to one, suggesting increased population level immunity is required to ensure prevention of measles persistence following importation.
@@
 
-Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  Table~\ref{table:direct}.
-The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in Table~\ref{table:hosp}.
+Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  \autoref{table:direct}.
+The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in \autoref{table:hosp}.
 
@@
 
-The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (Table~\ref{table:hosp}).
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (\autoref{table:hosp}).
 The mean cost per case was \$1,877.
@@
 From 16 December, 2013 through 19 June, 2014 there were 201 confirmed measles cases in New Zealand (note 14 of these occurred before 1 January 2014, so 187 occurred from Jan 2013 -- 19 June 2014).
-The number of cases by age group is shown in Table~\ref{table:freq}.
+The number of cases by age group is shown in \autoref{table:freq}.
 Of these 201 cases, 34 (17\%) were admitted to hospital with the highest proportion occurring in the youngest (< 15 months) and oldest (> 19 years) age groups, 47\% and 33\%, respectively.
@@
 
-The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (Figure~\ref{fig:hosp}).
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (\autoref{fig:hosp}).
 
@@
 Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days.
-The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in \autoref{table:cases}.
 
@@
 \caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\ref{table:regression})}
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
 <<results=tex,echo=FALSE>>=
@@
 
-Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation, case age and year of case, and a less strong association with case gender (Table~\ref{table:regression}).
+Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation, case age and year of case, and a less strong association with case gender (\autoref{table:regression_costs}).
 Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.
@@
 @
-\label{table:regression}
+\label{table:regression_costs}
 \end{table}
@@
 
-To estimate the benefits from additional vaccinations, as estimated from the above modelling section (section~\ref{section:epidemic_modelling}), we carried out a number of analyses.
-Primarily, we used Equation~\ref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (Figure~\ref{fig:naive_map} and Table~\ref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (section~\ref{section:epidemic_modelling}).
+To estimate the benefits from additional vaccinations, as estimated from the above modelling section (\autoref{section:epidemic_modelling}), we carried out a number of analyses.
+Primarily, we used \autoref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (\autoref{fig:naive_map} and \autoref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (\autoref{section:epidemic_modelling}).
 
@@
 
-The estimated vaccinaton rates, with percentages are shown in Table~\ref{table:vacp}.
+The estimated vaccinaton rates, with percentages are shown in \autoref{table:vacp}.
 <<echo=FALSE>>=
@@
 
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using equation~\ref{eq:prop} are shown in Figure~\ref{fig:numvac}.
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using \autoref{eq:prop} are shown in \autoref{fig:numvac}.
 
@@
 @
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (\autoref{section:epidemic_modelling}, \autoref{table:cost20} and \autoref{table:cost50})}
      \label{fig:numvac}
@@
 
-The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ is <1 is shown in Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}.
+The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ is <1 is shown in \autoref{fig:sim} and \autoref{fig:sim1}.
 These simulations show that even in scenarios when $R_v$ is lower than one, and thus deterministically should fail to persist (i.e.
@@
 @
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
-The full distribution of results can be seen in Figure~\ref{fig:sim1}}
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+The full distribution of results can be seen in \autoref{fig:sim1}}
      \label{fig:sim}
@@
 @
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
-A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+A subset of the distribution of results can be seen in \autoref{fig:sim}}
      \label{fig:sim1}
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 @
+\label{table:cost20}
 \end{sidewaystable}
@@
 \caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 
-For the cost analyses we used the values from the above cost section (section~\ref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (Table~\ref{table:hosp}).
-We estimated there would be approximately one introduction of measles per year (section~\ref{sub:risk_analyses}).
+For the cost analyses we used the values from the above cost section (\autoref{sub:cost_benefit}).
+Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (\autoref{table:hosp}).
+We estimated there would be approximately one introduction of measles per year (\autoref{sub:risk_analyses}).
 We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
 
-The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in Table~\ref{table:vacp}.
-The benefit--cost results are in Table~\ref{table:cost20} and Table~\ref{table:cost50}, for two different vaccine prices.
-The results in the two tables show the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (Tables~\ref{table:cost20} and \ref{table:cost50}).
-
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (Figure~\ref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (Figure~\ref{fig:vaccstat}).
+The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in \autoref{table:vacp}.
+The benefit--cost results are in \autoref{table:cost20} and \autoref{table:cost50}, for two different vaccine prices.
+The results in the two tables show the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (\autoref{table:cost20} and \autoref{table:cost50}).
+
+It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (\autoref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (\autoref{fig:vaccstat}).
 Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children.
@@
 We used the mean values per case in the absence of alternative data and used as much data as possible.
-The exact costs will vary, however, our benefit--cost analyses suggest that in all cases catch up vaccination schemes will be financially beneficial for New Zealand (see Tables~\ref{table:cost20} and \ref{table:cost50}).
+The exact costs will vary, however, our benefit--cost analyses suggest that in all cases catch up vaccination schemes will be financially beneficial for New Zealand (see \autoref{table:cost20} and \autoref{table:cost50}).
 
@@
 They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
-Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see Table~\ref{table:zhou}) \citep{zhou4}.
-However, the cost of measles in their analyses was substantially higher than mumps and rubella (Table~\ref{table:zhou}) \citep{zhou4}.
+Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see \autoref{table:zhou}) \citep{zhou4}.
+However, the cost of measles in their analyses was substantially higher than mumps and rubella (\autoref{table:zhou}) \citep{zhou4}.
 
@@
 Estimates in Korea suggest catch up vaccination schemes have a benefit--cost ratio of just over one \citep{bae13}.
-However,  the estimated B/C ratio for MMR has been estimated to be between 10.2 and 54.2, much greater than one (see Table~\ref{table:zhou2}) \citep{zhou4}.
+However,  the estimated B/C ratio for MMR has been estimated to be between 10.2 and 54.2, much greater than one (see \autoref{table:zhou2}) \citep{zhou4}.
 As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis.
@@
 It is unlikely $R_0$ was so high as 18 in New Zealand, as measles would still be endemic given current vaccination rates.
-However, the benefits of catch up vaccination are clear if $R_v$ is greater than one (Tables~\ref{table:cost20} and ~\ref{table:cost20}) and our analyses do not include the additional benefits of mumps and rubella immunity.
+However, the benefits of catch up vaccination are clear if $R_v$ is greater than one (\autoref{table:cost20} and ref{table:cost20}) and our analyses do not include the additional benefits of mumps and rubella immunity.
 As noted in our previous report, if measles $R_v$ in the currently na\"{\i}ve population is less than one, the benefits will not be so clear, though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
@@
 Finally, our model of measles introductions remains a simple one, and more complex models may predict smaller outbreaks depending on contact structure and other scenarios, such as the size of the local na\"{\i}ve population.
-The spatial effects of measles transmission may have affected both our multivariate regression analyses (section~\ref{sub:risk_analyses}) and will affect the predictions from modelling exercises (section~\ref{section:epidemic_modelling}).
-Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (Figure~\ref{fig:sim} and Figure~\ref{fig:sim1}).
+The spatial effects of measles transmission may have affected both our multivariate regression analyses (\autoref{sub:risk_analyses}) and will affect the predictions from modelling exercises (\autoref{section:epidemic_modelling}).
+Whatever happens, however, it is clear that there will be ongoing costs to maintain New Zealand free of endemic measles and introductions occurring on an annual basis may produce some larger and costly outbreaks, even if vaccination cover is high and $R_v$ is less than one (\autoref{fig:sim} and \autoref{fig:sim1}).
 Given that, the greater the vaccination coverage, the smaller the outbreaks will be.

### 2014-11-09 draftfinalreport.Rnw
last table need not be \tiny
fix column names on cost-benefit tables

@@
 library(Hmisc)
-colnames(costtable20)<-c("Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
+colnames(costtable20)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
 latex(costtable20, file="", table.env=FALSE,rowname=NULL)
@@
 library(Hmisc)
-colnames(costtable50)<-c("Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
+colnames(costtable50)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
 latex(costtable50, file="", table.env=FALSE,rowname=NULL)
@@
 \begin{sidewaystable}[htdp]
-\tiny
 \begin{center}
@@
 \hline
-Parameter & B/C direct costs & B/C societal
-perspective & Costs per disc. yrs life saved\\
+Parameter & B/C direct costs & B/C societal perspective & Costs per disc. yrs life saved\\
 \hline

### 2014-11-09 draftfinalreport.Rnw
reduce size of table 19 a bit
autoref typo corrected

@@
 
-The distribution of na\"{\i}ve among the DHBs also varies (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, ~autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).
+The distribution of na\"{\i}ve among the DHBs also varies (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 
-\begin{table}\small
+\begin{table}\footnotesize
 \caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.

### 2014-11-09 draftfinalreport.tex
autoref typo corrected

--- 2014-11-07 draftfinalreport.tex
@@
 The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, Table~\ref{table:none}).
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, \autoref{table:none}).
 
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Sat Nov 08 01:33:38 2014
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Mon Nov 10 12:38:31 2014
 \begin{tabular}{lrrrrrr}
@@
    \hline
-\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using Equation~\ref{eq:reg} and a quasipoisson error structure}
+\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using \autoref{eq:reg} and a quasipoisson error structure}
 \label{table:anova}
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Sat Nov 08 01:33:38 2014
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Mon Nov 10 12:38:31 2014
 \begin{tabular}{rrrrr}
@@
 \end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
-\label{table:regression}
+\label{table:regression_riskfactors}
 \end{table}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:imrisk12}}
+\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
 %latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:imrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
 %latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also Tables~\ref{table:nzrisk12} and ~\ref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
 %latex.default(imrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 
-The distribution of na\"{\i}ve among the DHBs also varies (section~\ref{sub:popim}, Figures~\ref{fig:Northland}, ~\ref{fig:Waitemata}, ~\ref{fig:Auckland}, ~\ref{fig:Counties_Manukau}, ~\ref{fig:Waikato}, ~\ref{fig:Lakes}, ~\ref{fig:BayofPlenty}, ~\ref{fig:Tairawhiti}, ~\ref{fig:Taranaki}, ~\ref{fig:HawkesBay}, ~\ref{fig:Whanganui}, ~\ref{fig:Midcentral}, ~\ref{fig:Hutt}, ~\ref{fig:CapitalandCoast}, ~\ref{fig:Wairarapa}, ~\ref{fig:NelsonMarlborough}, ~\ref{fig:WestCoast}, ~\ref{fig:Canterbury}, ~\ref{fig:SouthCanterbury}, ~\ref{fig:Southern}).
+The distribution of na\"{\i}ve among the DHBs also varies (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 Due to the small number of cases spending a week or more in the hospital, the regression analysis to determine the association between cost of hospitalisation was limited to the 285 cases hospitalised for seven or fewer days.
-The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in Table~\ref{table:cases}.
-
-\begin{table}\small
+The number of cases, length of hospital stay, cost, cost per case and cost per day for patients with measles as the primary diagnosis, by year and gender for 2000--2014 appear in \autoref{table:cases}.
+
+\begin{table}\footnotesize
 \caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see Table~\ref{table:regression})}
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
 %latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
@@
 &     M&214,591&$127$&$186$&1,690\tabularnewline
-&   TOTAL&550,022&$293$&$470$&1,877\tabularnewline
+&TOTAL&550,022&$293$&$470$&1,877\tabularnewline
 \hline
@@
 \hline
-\end{tabular}\end{center}\label{table:regression}
+\end{tabular}\end{center}\label{table:regression_costs}
 \end{table}
@@
 
-The estimated vaccinaton rates, with percentages are shown in Table~\ref{table:vacp}.
+The estimated vaccinaton rates, with percentages are shown in \autoref{table:vacp}.
 
@@
 \includegraphics{draftfinalreport-061}
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (section~\ref{section:epidemic_modelling}, Tables~\ref{table:cost20} and ~\ref{table:cost50})}
+     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (\autoref{section:epidemic_modelling}, \autoref{table:cost20} and \autoref{table:cost50})}
      \label{fig:numvac}
@@
 \includegraphics{draftfinalreport-063}
-     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
-The full distribution of results can be seen in Figure~\ref{fig:sim1}}
+     \caption{A subset of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
+The full distribution of results can be seen in \autoref{fig:sim1}}
      \label{fig:sim}
@@
 \includegraphics{draftfinalreport-064}
-     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (section~\ref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
-A subset of the distribution of results can be seen in Figure~\ref{fig:sim}}
+     \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
+A subset of the distribution of results can be seen in \autoref{fig:sim}}
      \label{fig:sim1}
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable20, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}&\multicolumn{1}{c}{NA}\tabularnewline
-\hline
-Auckland&$17920$&$358400$&$26547468$&$55010964.560$&$5297.03$&$ 9942525.31$&$80393738.230$&$ 82$&$2115692.588$&$32.494231870$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895455.820$&$1434.29$&$ 2692162.33$&$21768412.640$&$ 71$&$1831880.168$&$11.316613160$\tabularnewline
-Canterbury&$13687$&$273740$&$21040140$&$43598824.410$&$4198.15$&$ 7879927.55$&$63715888.370$&$ 62$&$1599670.006$&$34.010648060$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15679356$&$32490348.880$&$3128.51$&$ 5872213.27$&$47481817.920$&$ 96$&$2476908.396$&$17.676674730$\tabularnewline
-Counties Manukau&$18880$&$377600$&$28033356$&$58089982.570$&$5593.51$&$10499018.27$&$84893455.150$&$ 50$&$1290056.456$&$50.905841450$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558.090$&$1163.82$&$ 2184490.14$&$17663453.000$&$ 56$&$1444863.231$&$11.621585550$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395.140$&$1332.12$&$ 2500389.24$&$20217764.780$&$ 86$&$2218897.104$&$ 8.764963262$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434.353$&$ 882.64$&$ 1656715.28$&$13395946.240$&$ 62$&$1599670.006$&$ 8.082555219$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327.040$&$1419.16$&$ 2663763.32$&$21538782.590$&$ 75$&$1935084.684$&$10.622562600$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585.118$&$ 749.87$&$ 1407505.99$&$11380878.060$&$ 90$&$2322101.621$&$ 4.803635909$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118.370$&$ 966.96$&$ 1814983.92$&$14675682.240$&$ 70$&$1806079.039$&$ 7.858468434$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962495.540$&$ 285.26$&$  535433.02$&$ 4329429.467$&$ 72$&$1857681.297$&$ 2.308362644$\tabularnewline
-Southern&$ 8371$&$167420$&$12877980$&$26685411.260$&$2569.55$&$ 4823045.35$&$38998406.670$&$102$&$2631715.170$&$13.932305620$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911.000$&$ 413.27$&$  775707.79$&$ 6272254.490$&$ 47$&$1212653.069$&$ 5.058950796$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4483224$&$ 9290018.791$&$ 894.54$&$ 1679051.58$&$13576554.140$&$ 68$&$1754476.780$&$ 7.490691249$\tabularnewline
-Waikato&$11331$&$226620$&$17251296$&$35747681.580$&$3442.16$&$ 6460934.32$&$52242126.250$&$ 95$&$2451107.267$&$19.509875740$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1146792$&$ 2376352.203$&$ 228.82$&$  429495.14$&$ 3472831.980$&$ 59$&$1522266.618$&$ 2.259977498$\tabularnewline
-Waitemata&$17291$&$345820$&$26219448$&$54331250.150$&$5231.58$&$ 9819675.66$&$79400394.760$&$ 70$&$1806079.039$&$36.897825290$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1077780$&$ 2233347.353$&$ 215.05$&$  403648.85$&$ 3263842.834$&$ 50$&$1290056.456$&$ 2.503414513$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2155560$&$ 4466694.706$&$ 430.10$&$  807297.70$&$ 6527685.668$&$ 58$&$1496465.489$&$ 4.283186675$\tabularnewline
-\hline
-\end{tabular}\end{center}\end{sidewaystable}
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}\tabularnewline
+\hline
+Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$ 9942525$&$80393738$&$ 82$&$2115693$&$32.5$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$ 2692162$&$21768413$&$ 71$&$1831880$&$11.3$\tabularnewline
+Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$ 7879928$&$63715888$&$ 62$&$1599670$&$34.0$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$ 5872213$&$47481818$&$ 96$&$2476908$&$17.7$\tabularnewline
+Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$10499018$&$84893455$&$ 50$&$1290056$&$50.9$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$ 2184490$&$17663453$&$ 56$&$1444863$&$11.6$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$ 2500389$&$20217765$&$ 86$&$2218897$&$ 8.8$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$ 1656715$&$13395946$&$ 62$&$1599670$&$ 8.1$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$ 2663763$&$21538783$&$ 75$&$1935085$&$10.6$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$ 1407506$&$11380878$&$ 90$&$2322102$&$ 4.8$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$ 1814984$&$14675682$&$ 70$&$1806079$&$ 7.9$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$  535433$&$ 4329429$&$ 72$&$1857681$&$ 2.3$\tabularnewline
+Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$ 4823045$&$38998407$&$102$&$2631715$&$13.9$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$  775708$&$ 6272254$&$ 47$&$1212653$&$ 5.1$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4483224$&$ 9290019$&$ 895$&$ 1679052$&$13576554$&$ 68$&$1754477$&$ 7.5$\tabularnewline
+Waikato&$11331$&$226620$&$17251296$&$35747682$&$3442$&$ 6460934$&$52242126$&$ 95$&$2451107$&$19.5$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1146792$&$ 2376352$&$ 229$&$  429495$&$ 3472832$&$ 59$&$1522267$&$ 2.3$\tabularnewline
+Waitemata&$17291$&$345820$&$26219448$&$54331250$&$5232$&$ 9819676$&$79400395$&$ 70$&$1806079$&$36.9$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1077780$&$ 2233347$&$ 215$&$  403649$&$ 3263843$&$ 50$&$1290056$&$ 2.5$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2155560$&$ 4466695$&$ 430$&$  807298$&$ 6527686$&$ 58$&$1496465$&$ 4.3$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:cost20}
+\end{sidewaystable}
 
@@
 \caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see Table~\ref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable50, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}&\multicolumn{1}{c}{NA}\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}\tabularnewline
 \hline

### 2015-03-18 draftfinalreport_v2.Rnw
revised draft

--- 2014-11-09 draftfinalreport.Rnw
@@
 Previous measles analyses, including two in New Zealand by Prof.
-Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
+Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles %\citep{roberts0,roberts4}.
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
@@
 \begin{figure}
-     \centering
+     %\centering
      \includegraphics[width=0.9\textwidth]{incidence_1997_2014.pdf}
@@
 
-%%%%%%%%%%%%%%%%%%
 \subsection{Risk of measles infection in New Zealand analyses methods}
@@
 
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mfrow=c(1,1))
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-@
-\end{center}
-\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
-\label{fig:ageinyears}
-\end{figure}
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mfrow=c(1,1))
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+% @
+% \end{center}
+% \caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+% \label{fig:ageinyears}
+% \end{figure}
 
@@
 
+<<echo=FALSE,include=FALSE,results=hide>>=
+
+## get data
+# set wd
+# setwd("~/Massey 2014/DHayman_20140627")
+# read data
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+data = data , FUN=sum)
+library(reshape)
+library(reshape2)
+library(plyr)
+library(pscl)
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+head(test)
+dim(test)
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mfrow=c(1,1))
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+## plot cases / age class
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+data = test , FUN=sum)
+
+require(reshape2)  # this is the library that lets one flatten out data
+require(ggplot2)   # plotting library
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+
+## match cases per age
+AgeInYears<-0:100
+naive<-cbind(AgeInYears,naive)
+colnames(naive)<-c("AgeInYears","Naive")
+naive<-merge(naive,caseyr,by="AgeInYears",all=T)
+colnames(naive)<-c("AgeInYears","Naive","Cases")
+
+dose1=as.factor(round(vac$Dose1Mths/12,0));
+dose1<-summary(dose1)
+dose1<-as.data.frame(dose1)
+dose1$AgeInYears<-rownames(dose1)
+dose2=as.factor(round(na.omit(vac$Dose2Mths/12,0)));
+dose2<-summary(dose2)
+dose2<-as.data.frame(dose2)
+dose2$AgeInYears<-rownames(dose2)
+
+## need to merge vaccination data with data
+## to get years the cases were from
+datav<-merge(vac,data,by="CaseCode",all=T)
+##
+datav$Dose1Mths[which(is.na(datav$Dose1Mths))] <- 9999
+datav$Dose2Mths[which(is.na(datav$Dose2Mths))] <- 9999
+#########################
+datav$RptYear<-as.factor(datav$RptYear)
+datav$SurvWeek<-as.factor(datav$SurvWeek)
+datav$NZDep01<-as.factor(datav$NZDep01)
+datav$NZDep06<-as.factor(datav$NZDep06)
+datav$NZDep13<-as.factor(datav$NZDep13)
+datav$Dose1Mths<-as.factor(datav$Dose1Mths)
+datav$Dose2Mths<-as.factor(datav$Dose2Mths)
+
+#data$AgeInYears<-as.factor(data$AgeInYears)
+datav$EthnicityPrioritised<-as.factor(datav$EthnicityPrioritised)
+
+timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
+data = datav , FUN=sum,na.rm=F,na.action=na.pass)
+
+testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+library(plyr)
+testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
+testv$Dose2Mths<-revalue(testv$Dose2Mths, c("9999"=NA));
+testv$Dose1Mths<-as.numeric(testv$Dose1Mths)
+testv$Dose2Mths<-as.numeric(testv$Dose2Mths)
+testv$D2vac<-ifelse(testv$Dose1Mths > -1 & testv$Dose2Mths >= testv$Dose1Mths,testv$Dose2Mths,NA)
+testv$D1vac<-ifelse(testv$Dose1Mths > -1 & is.na(testv$D2vac) == T,testv$Dose1Mths,NA)
+testv$Unvac<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,testv$AgeInYears,NA)
+
+testv$VC<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,0,
+ifelse(testv$Dose1Mths >= 0 & is.na(testv$Dose2Mths) == T,1,2))
+AgeVac<-table(testv$VC,testv$AgeInYears)
+
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+AgeInYears<-(as.numeric(rownames(AgeVac)))
+AgeVac<-cbind(AgeVac,AgeInYears)       
+AgeInYears<-(as.numeric(rownames(pop)))
+pop<-cbind(pop,AgeInYears)
+colnames(pop)<-c("Population","AgeInYears")
+AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
+
+## regression analyses - measles
+# setwd("~/Massey 2014/DHayman_20140627")
+# read data
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+data = data , FUN=sum)
+library(reshape)
+library(reshape2)
+library(plyr)
+library(pscl)
+
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+head(test)
+dim(test)
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+data = test , FUN=sum)
+caseyr
+# test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
+test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
+test$AgeInYears<-as.factor(test$AgeInYears)
+#tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
+tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
+test$AgeInYears<-tage
+#summary(test$EthnicityPrioritised)
+teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
+"Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
+
+test$EthnicityPrioritised<-teth
+head(test)
+test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
+test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
+
+# want to set NZ Deprivation to the appropriate year
+
+testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
+ifelse (test$RptYear == "2008",test$NZDep06,
+ifelse (test$RptYear == "2009",test$NZDep06,
+ifelse (test$RptYear == "2010",test$NZDep06,
+ifelse (test$RptYear == "2011",test$NZDep06,
+ifelse (test$RptYear == "2012",test$NZDep06,
+ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
+head(testt)
+tail(testt)
+str(testt)
+testt$NZDep<-as.factor(testt$NZDep-1)
+
+testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
+~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
+data = testt , FUN=sum)
+dim(testtable)
+colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
+head(testtable)
+
+#tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
+#"5"="1-5","6"="6-10","7"="6-10","8"="6-10",
+#"9"="6-10","10"="6-10"));
+
+tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
+                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
+                                 "9"="8-10","10"="8-10"));
+
+testtable$NZDep<-tnzd
+# head(testtable)
+tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
+data = testtable , FUN=sum)
+
+ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
+data = testtable , FUN=sum)
+ttyr <- ttyr[ttyr$Ethnicity=="None",]
+
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(ttyr, file="",table.env=FALSE,rowname=NULL)
+
+### denominator data
+# setwd("~/Massey_2014/measles/data")
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+data = data , FUN=sum)
+library(reshape)
+library(reshape2)
+library(plyr)
+library(pscl)
+
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+head(test)
+dim(test)
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+data = test , FUN=sum)
+caseyr
+## for MoH desired ages
+test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
+test$AgeInYears<-as.factor(test$AgeInYears)
+tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
+test$AgeInYears<-tage
+#summary(test$EthnicityPrioritised)
+teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
+"Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
+
+test$EthnicityPrioritised<-teth
+head(test)
+test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
+test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
+
+# want to set NZ Deprivation to the appropriate year
+
+testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
+ifelse (test$RptYear == "2008",test$NZDep06,
+ifelse (test$RptYear == "2009",test$NZDep06,
+ifelse (test$RptYear == "2010",test$NZDep06,
+ifelse (test$RptYear == "2011",test$NZDep06,
+ifelse (test$RptYear == "2012",test$NZDep06,
+ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
+head(testt)
+tail(testt)
+str(testt)
+testt$NZDep<-as.factor(testt$NZDep-1)
+
+testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
+~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
+data = testt , FUN=sum)
+dim(testtable)
+colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
+head(testtable)
+
+#tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
+#"5"="1-5","6"="6-10","7"="6-10","8"="6-10",
+#"9"="6-10","10"="6-10"));
+
+tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
+                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
+                                 "9"="8-10","10"="8-10"));
+
+testtable$NZDep<-tnzd
+# head(testtable)
+tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
+data = testtable , FUN=sum)
+
+ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
+data = testtable , FUN=sum)
+denom<-read.csv("NZDep2006Denominators.csv",header=T)
+head(denom)
+denom<-denom[,-c(9)]
+summary(denom)
+dim(denom)
+denom<-denom[,1:18]
+row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
+sum(row.has.na)
+denomt <- denom[!row.has.na,]
+head(denomt)
+denomt<-denomt[denomt$Sex_code==99,]
+denomt$Eth_Level<-as.factor(denomt$Eth_Level)
+denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
+dm<-cbind(denomt[,c(3,6,9:18)])
+head(dm)
+dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
+dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
+dm$Age_Label<-dage
+colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
+head(dm)
+datadm<-dm[dm$Age=="<6",]
+dataA<-(rbind(datadm,datadm)/2)
+dataA<-round(dataA[,3:12])
+dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
+colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
+ddm<-dm[!dm$Age=="<6",]
+dm<-rbind(dataA,ddm)
+dm$Ethnicity <- factor(dm$Ethnicity)
+deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
+"Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
+dm$Ethnicity<-deth
+head(dm)
+
+## here's the additional fudge the MoH asked for.
+## add 5 year olds to next class
+dm[11:15,3:12]<-dm[11:15,3:12]+round(dm[6:10,3:12]/3,0)
+dm[6:10,3:12]<-round(dm[6:10,3:12]*2/3,0)
+## add 2 year olds to next class
+dm[6:10,3:12]<-dm[6:10,3:12]+round(dm[1:5,3:12]/3,0)
+dm[1:5,3:12]<-round(dm[1:5,3:12]*2/3,0)
+## 
+# rename
+dm$Age<-revalue(dm$Age, c("0-2"="0-1", "3-5"="2-4","6-17"="5-17"));
+
+popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
+colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
+popn
+#pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
+#"Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
+#"Dep9"="6-10","Dep10"="6-10"));
+
+pnzd<-revalue(popn$NZDep, c("Dep1"="1-3","Dep2"="1-3","Dep3"="1-3","Dep4"="4-7",
+                            "Dep5"="4-7","Dep6"="4-7","Dep7"="4-7","Dep8"="8-10",
+                            "Dep9"="8-10","Dep10"="8-10"));
+
+popn$NZDep<-pnzd
+head(popn)
+tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
+data = popn , FUN=sum)
+
+tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
+ttyr$merge <- paste(ttyr$NZDep, ttyr$Age, ttyr$Ethnicity)
+ttyr <- ttyr[ttyr$Ethnicity!="None",]
+
+tp$cases <- 0
+cases <- matrix(0, length(tp$merge),1)
+rownames(cases) <- tp$merge
+cases[ttyr$merge,] <- ttyr$Cases
+tp$cases <- cases
+PerCapita<-(tp$cases/tp$Popn)
+PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita*10000,4))
+colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
+
+#par(mar=c(12,6,4,2)+0.1)
+#par(cex.axis=2)
+#plot(tp$cases/tp$Popn*10000,main="",
+#        ylab="Per Capita Rate Per 10,000",
+#        pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
+#xlab="",cex.lab=2,xaxt="n")
+#axis(side = 1, at = 1:75, labels=rep(c("0-2yr, 1-3","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),
+#     las=2,cex.axis=1.5)
+#mtext(side=1,line=10,text="Age and NZDep",cex=2)
+#legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
+#col=c(1:4,6),pch=16,bty="n",cex=1.8)
+#par(mar = c(5,4,4,2) + 0.1)
+#
+
+tp$perCap<-with(tp,tp$cases/tp$Popn*10000)
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, cases, fill=Age)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%  # theme(text=element_text(size=45))+
+%   ylab("cases")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, cases, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+% #  theme(text=element_text(size=45))+
+%   ylab("cases")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(Age, cases, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+% #  theme(text=element_text(size=45))+
+%   ylab("cases")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, perCap, fill=Age)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%  # theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(Age, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+tpsub<-tp[!(tp$Ethnicity=="MLA"),]
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tpsub, aes(NZDep, perCap, fill=Age)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tpsub, aes(NZDep, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tpsub, aes(Age, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(PerCap, file="", table.env=FALSE,rowname=NULL)
+
+## plot this figure?
+#hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+tp$Ethnicity<- relevel(tp$Ethnicity, "European")
+model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
+summary(model)
+
+model1<-update(model,~.-Age:Ethnicity:NZDep)
+summary(model1)
+anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
+
+model2<-update(model1,~.-Age:NZDep)
+summary(model2)
+anova(model2,test="F")
+
+## use tpsub
+#hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
+
+model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
+summary(model)
+
+model1<-update(model,~.-Age:Ethnicity:NZDep)
+summary(model1)
+anova(model1,test="F")
+
+#model2<-update(model1,~.-Ethnicity:NZDep)
+#summary(model2)
+#anova(model2,test="F")
+model2<-update(model1,~.-Age:NZDep)
+summary(model2)
+anova(model2,test="F")
+
+model3<-update(model2,~.-Ethnicity:NZDep)
+summary(model3)
+anova(model3,test="F")
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(cex.axis=2)
+% par(mar=c(5,6,4,2)+0.1)
+% hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+res<-predict(model3)
+#res<-predict(model2)
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% #hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+% hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+library(xtable)
+resan<-xtable(anova(model3,test="F"))
+#resan<-xtable(anova(model2,test="F"))
+print(resan,floating = FALSE)
+
+library(xtable)
+#xtb<-xtable(summary(model3))
+xtb<-xtable(summary(model2))
+print(xtb,floating = FALSE)
+
+AgeVac<-table(testv$VC,testv$AgeInYears)
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
+% legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
+% abline(v=12,col="blue")
+% hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+% abline(v=60,col="red")
+% legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
+% legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+DOB<-as.numeric(levels(testv$RptYear))[testv$RptYear]-testv$AgeInYears
+#hist(DOB,breaks=100,col="grey",main="2007- 2014 measles cases",xlab="Date of Birth")
+
+DOBVac<-table(testv$VC,DOB)
+row.names(DOBVac)<-c("Unvaccinated","Dose1","Dose2")
+DOBVac<-t(DOBVac)
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1)
+% barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
+% legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1)
+% par(mfrow=c(2,1))
+% barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+% legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1)
+% barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+% legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1.5)
+% par(mfrow=c(1,1))
+% barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
+% legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+
+require(reshape2)  # this is the library that lets one flatten out data
+require(ggplot2)   # plotting library
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+@
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(cex.axis=2)
+% par(mar=c(5,6,4,2)+0.1)
+% plot(pop,xlab="Age",ylab="Population",cex.lab=2)
+% points(naive,pch=16)
+% legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+## match cases per age
+AgeInYears<-0:100
+naive<-cbind(AgeInYears,naive)
+colnames(naive)<-c("AgeInYears","Naive")
+naive<-merge(naive,caseyr,by="AgeInYears",all=T)
+#naive[is.na(naive)] <- 0
+colnames(naive)<-c("AgeInYears","Naive","Cases")
+##
+
+# proportions of the naive population the wuld be covered if vaccinated by age groups now
+propvCohort<-c(
+round((sum(naive$Naive[3:6])/sum(naive$Naive)),2),
+round((sum(naive$Naive[6:12])/sum(naive$Naive)),2),
+round((sum(naive$Naive[6:17])/sum(naive$Naive)),2),
+round((sum(naive$Naive[6:19])/sum(naive$Naive)),2),
+round((sum(naive$Naive[12:17])/sum(naive$Naive)),2),
+round((sum(naive$Naive[12:19])/sum(naive$Naive)),2),
+round((sum(naive$Naive[1:30])/sum(naive$Naive)),2),
+round((sum(naive$Naive[1:45])/sum(naive$Naive)),2),
+round((sum(naive$Naive[8:100])/sum(naive$Naive)),2))
+names(propvCohort)<-c("2-5","5-11","5-16","5-18","11-16","11-18","0-29","0-44",">8")
+write.table(propvCohort,"vaccine_proportions_Cohort.txt",sep=",",row.names=T,col.names=F)
+
+propsv = NULL
+for (i in 1:20 ) {
+  propsv[i] = round((sum(naive$Naive[i:13])/sum(naive$Naive)),2)
+}
+propsvAge = NULL
+for (i in 1:20 ) {
+  propsvAge[i]=paste(c(min(i:20)-1, max(i:20)-1),collapse="-")
+}
+
+table120<-as.data.frame(propsv,propsvAge)
+table120<-cbind(Age=rownames(table120),table120)
+rownames(table120)<-1:20
+colnames(table120)<-c("Age","Proportion")
+
+propsvrev = NULL
+for (i in 20:1 ) {
+  propsvrev[i] = round((sum(naive$Naive[i:1])/sum(naive$Naive)),2)
+}
+propsvrevAge = NULL
+for (i in 20:1 ) {
+  propsvrevAge[i]=paste(c(min(i:1)-1, max(i:1)-1),collapse="-")
+}
+
+table201<-as.data.frame(propsvrev,propsvrevAge)
+table201<-cbind(Age=rownames(table201),table201)
+rownames(table201)<-1:20
+colnames(table201)<-c("Age","Proportion")
+
+propsv11 = NULL
+for (i in 1:11 ) {
+  propsv11[i] = round((sum(naive$Naive[i:11])/sum(naive$Naive)),2)
+}
+propsv11Age = NULL
+for (i in 1:11 ) {
+  propsv11Age[i]=paste(c(min(i:11)-1, max(i:11)-1),collapse="-")
+}
+
+table11<-as.data.frame(propsv11,propsv11Age)
+table11<-cbind(Age=rownames(table11),table11)
+rownames(table11)<-1:11
+colnames(table11)<-c("Age","Proportion")
+
+write.table(table11,"vaccine_proportions_11yrOlds.txt",sep=",",row.names=F)
+write.table(table120,"vaccine_proportions_1_20yrOlds.txt",sep=",",row.names=F)
+write.table(table201,"vaccine_proportions_20_1yrOlds.txt",sep=",",row.names=F)
+
+###
+@
+
 The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
@@
 
-\vspace{5mm} %5mm vertical space
-\begin{table}
-<<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
-library(xtable)
-xtb<-xtable(summary(model3))
-print(xtb,floating = FALSE)
-@
-\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
-\label{table:regression_riskfactors}
-\end{table}
-\vspace{5mm} %5mm vertical space
+% \vspace{5mm} %5mm vertical space
+% \begin{table}
+% % <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
+% % library(xtable)
+% % xtb<-xtable(summary(model3))
+% % print(xtb,floating = FALSE)
+% % @
+% \caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+% \label{table:regression_riskfactors}
+% \end{table}
+% \vspace{5mm} %5mm vertical space
 
@@
 
+%%%%
+<<echo=FALSE,include=FALSE,results=hide>>=
+library(maptools)
+
+# read in data from NIR
+data <- read.table("NIR_MMR_data_20060101-20140731/NIR_MMR_data_20060101-20140731.txt", header=T, sep="|")
+
+# read in domicile <-> AU map
+data_dom <- read.csv("NIR_MMR_data_20060101-20140731/domicile_code.csv")
+
+# merge NIR with domicile/AU so we have AU for each NIR
+data_merge <- merge(data, data_dom, by.x = "DOMICILE_CODE", by.y="dom", all=T)
+
+# TODO: figure out how many kids are immunised in each AU
+# find number of kids by BIRTH_CALENDAR_YEAR, AU, VACCINE_DOSE
+
+num_kids <- table(data_merge$BIRTH_CALENDAR_YEAR, data_merge$area.unit, data_merge$VACCINE_DOSE)
+# drop the 0's from vaccine_dose
+dose1 <- t(num_kids[,,2])
+dose2 <- t(num_kids[,,3])
+dose1 <- as.data.frame(dose1[,-ncol(dose1)])
+dose2 <- as.data.frame(dose2[,-ncol(dose2)])
+dose1$cau <- rownames(dose1)
+dose2$cau <- rownames(dose2)
+
+# read in shapefile for AU's
+au <- readShapeSpatial("AU2013_GV_Full")
+
+au_data <- slot(au, "data")
+
+# read in denominator data for AU's
+data_denom<-read.csv("2013-mb-dataset-Total-New-Zealand-individual-part-1_AU.csv",header=T)
+denom<-data_denom[,1]
+for (i in 1:5)
+  denom <- cbind(denom, data_denom[,3]/5)
+for (i in 1:5)
+  denom <- cbind(denom, data_denom[,4]/5)
+colnames(denom)[2:11] <- 2013:(2013-9)
+denom <- denom[,c(1,9:2)]
+
+dose2_merge <- merge(dose2, denom, by.x="cau", by.y="denom", all=T)
+diff <- dose2_merge[,2:8+7] - dose2_merge[,2:8]
+image(as.matrix(diff > 0))
+
+dose2_num <- data.frame(cau=dose2_merge$cau, vacc = rowSums(dose2_merge[,2:5]), total = rowSums(dose2_merge[,10:13]))
+diff <- dose2_num$total - dose2_num$vacc
+
+dose1_merge <- merge(dose1, denom, by.x="cau", by.y="denom", all=T)
+diff <- dose1_merge[,2:8] - dose1_merge[,2:8+7]
+image(as.matrix(diff > 0))
+
+au_new <- merge(au_data, dose2_num, by.x="AU2013", by.y="cau", all.x=T)
+au_new$prop <- au_new$vacc / au_new$total
+library(RColorBrewer)
+library(classInt)
+
+fixedBreaks <- c(0,0.5,0.7, 0.9, 1.1, 1.3, 1.5, Inf)
+au_new$prop[!is.finite(au_new$prop)] <- NA
+breaks <- classIntervals(au_new$prop, n=7, style="fixed", fixedBreaks=fixedBreaks)
+pal <- brewer.pal("Spectral", n=7)
+
+cols <- findColours(breaks, pal)
+plot(au, col=cols)
+
+## plot to show Ministry the data issues...
+fixedBreaks <- c(0,0.6,0.8,1, 1.2,1.4,Inf)
+au_new$prop[!is.finite(au_new$prop)] <- NA
+breaks <- classIntervals(au_new$prop, n=6, style="fixed", fixedBreaks=fixedBreaks)
+pal <- brewer.pal("RdGy", n=6)
+pal<-c("red","orange","grey","blue","green","yellow")
+cols <- findColours(breaks, pal)
+@
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mfrow=c(1,1))
+pdf(paste("nir_census.pdf"), width=7, height=5)
+plot(au, col=cols)
+legend("topright",c("0-60%","61-80%","81-100%","101-120%","121-140%",">141%","NA")
+       ,fill=c("red","orange","grey","blue","green","yellow","white"),
+       bty="n",inset=0.1,title="Percent vaccinated")
+## par("usr")
+##
+par(fig = c(0.05, 0.4, 0.5,0.9), mar=c(5,5,2,2), new=TRUE)
+hist(au_new$prop[au_new$prop < 2], breaks=50,col="grey",
+     xlab="Proportion vaccinated",main="")
+abline(v=1,col="red",lwd=3)
+#text(x=0.2,y=150,"Area Unit")
+#dev.off()
+par(mfrow=c(1,1))
+
+hist(au_new$prop, breaks=100,col="grey",xlab="Proportion vaccination",main="All domiciles")
+abline(v=1,col="red")
+@
+\end{center}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator}
+\label{fig:naive_au}
+\end{figure}
+%%%%
+
 The majority of the na\"{\i}ve among the general New Zealand population are in their first years of life (\autoref{fig:naive}).
@@
 The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures:
-\begin{itemize}[noitemsep,nolistsep]
-\item Northland: \autoref{fig:Northland}
-\item Waitemata: \autoref{fig:Waitemata}
-\item Auckland: \autoref{fig:Auckland}
-\item Counties Manukau: \autoref{fig:Counties_Manukau}
-\item Waikato: \autoref{fig:Waikato}
-\item Lakes: \autoref{fig:Lakes}
-\item Bay of Plenty: \autoref{fig:BayofPlenty}
-\item Tairawhiti: \autoref{fig:Tairawhiti}
-\item Taranaki: \autoref{fig:Taranaki}
-\item Hawke's Bay: \autoref{fig:HawkesBay}
-\item Whanganui: \autoref{fig:Whanganui}
-\item Mid-Central: \autoref{fig:Midcentral}
-\item Hutt: \autoref{fig:Hutt}
-\item Capital and Coast: \autoref{fig:CapitalandCoast}
-\item Wairarapa: \autoref{fig:Wairarapa}
-\item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
-\item West Coast: \autoref{fig:WestCoast}
-\item Canterbury: \autoref{fig:Canterbury}
-\item South Canterbury: \autoref{fig:SouthCanterbury}
-\item Southern: \autoref{fig:Southern}
-\end{itemize}
+% \begin{itemize}[noitemsep,nolistsep]
+% \item Northland: \autoref{fig:Northland}
+% \item Waitemata: \autoref{fig:Waitemata}
+% \item Auckland: \autoref{fig:Auckland}
+% \item Counties Manukau: \autoref{fig:Counties_Manukau}
+% \item Waikato: \autoref{fig:Waikato}
+% \item Lakes: \autoref{fig:Lakes}
+% \item Bay of Plenty: \autoref{fig:BayofPlenty}
+% \item Tairawhiti: \autoref{fig:Tairawhiti}
+% \item Taranaki: \autoref{fig:Taranaki}
+% \item Hawke's Bay: \autoref{fig:HawkesBay}
+% \item Whanganui: \autoref{fig:Whanganui}
+% \item Mid-Central: \autoref{fig:Midcentral}
+% \item Hutt: \autoref{fig:Hutt}
+% \item Capital and Coast: \autoref{fig:CapitalandCoast}
+% \item Wairarapa: \autoref{fig:Wairarapa}
+% \item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
+% \item West Coast: \autoref{fig:WestCoast}
+% \item Canterbury: \autoref{fig:Canterbury}
+% \item South Canterbury: \autoref{fig:SouthCanterbury}
+% \item Southern: \autoref{fig:Southern}
+% \end{itemize}
 

### 2015-03-18 draftfinalreport_v2.tex
revised draft

--- 2014-11-09 draftfinalreport.tex
@@
 \begin{document}
-\input{draftfinalreport-concordance}
+\input{draftfinalreport_v2-concordance}
 
@@
 
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport-002}
-\end{center}
-\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
-\label{fig:ageinyears}
-\end{figure}
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mfrow=c(1,1))
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+% @
+% \end{center}
+% \caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+% \label{fig:ageinyears}
+% \end{figure}
 
@@
 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mfrow=c(1,1))
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, cases, fill=Age)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%  # theme(text=element_text(size=45))+
+%   ylab("cases")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, cases, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+% #  theme(text=element_text(size=45))+
+%   ylab("cases")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(Age, cases, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+% #  theme(text=element_text(size=45))+
+%   ylab("cases")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, perCap, fill=Age)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%  # theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(NZDep, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tp, aes(Age, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tpsub, aes(NZDep, perCap, fill=Age)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tpsub, aes(NZDep, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% ggplot(tpsub, aes(Age, perCap, fill=Ethnicity)) + 
+%   geom_bar(stat="identity", position="dodge")+
+%   #theme(text=element_text(size=45))+
+%   ylab("Per capita per 10000")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(cex.axis=2)
+% par(mar=c(5,6,4,2)+0.1)
+% hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% #hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+% hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
+% legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=2)
+% hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
+% abline(v=12,col="blue")
+% hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+% abline(v=60,col="red")
+% legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
+% legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1)
+% barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
+% legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1)
+% par(mfrow=c(2,1))
+% barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+% legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1)
+% barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+% legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+% 
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(mar=c(5,6,4,2)+0.1)
+% par(cex.axis=1.5)
+% par(mfrow=c(1,1))
+% barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
+% legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+% par(cex.axis=2)
+% par(mar=c(5,6,4,2)+0.1)
+% plot(pop,xlab="Age",ylab="Population",cex.lab=2)
+% points(naive,pch=16)
+% legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
+% @
+% \end{center}
+% \caption{test}
+% %\label{fig:test}
+% \end{figure}
+
 The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
@@
 \begin{center}
-\includegraphics{draftfinalreport-010}
+\includegraphics{draftfinalreport_v2-019}
 \end{center}
@@
 \hline
-1-5&0-2&Asian&$   6094$&$ 11$&$18.0505$\tabularnewline
-6-10&0-2&Asian&$   6806$&$  8$&$11.7543$\tabularnewline
-1-5&3-5&Asian&$   6094$&$  1$&$ 1.6410$\tabularnewline
-6-10&3-5&Asian&$   6806$&$  0$&$ 0.0000$\tabularnewline
-1-5&6-17&Asian&$  33918$&$ 11$&$ 3.2431$\tabularnewline
-6-10&6-17&Asian&$  28905$&$  5$&$ 1.7298$\tabularnewline
-1-5&18-24&Asian&$  22917$&$  3$&$ 1.3091$\tabularnewline
-6-10&18-24&Asian&$  34107$&$  5$&$ 1.4660$\tabularnewline
-1-5&25+&Asian&$  96357$&$ 10$&$ 1.0378$\tabularnewline
-6-10&25+&Asian&$  98715$&$ 13$&$ 1.3169$\tabularnewline
-1-5&0-2&European&$  57872$&$ 83$&$14.3420$\tabularnewline
-6-10&0-2&European&$  45445$&$ 64$&$14.0830$\tabularnewline
-1-5&3-5&European&$  57872$&$ 42$&$ 7.2574$\tabularnewline
-6-10&3-5&European&$  45445$&$ 17$&$ 3.7408$\tabularnewline
-1-5&6-17&European&$ 264330$&$219$&$ 8.2851$\tabularnewline
-6-10&6-17&European&$ 182937$&$ 80$&$ 4.3731$\tabularnewline
-1-5&18-24&European&$ 107649$&$ 34$&$ 3.1584$\tabularnewline
-6-10&18-24&European&$ 117840$&$ 36$&$ 3.0550$\tabularnewline
-1-5&25+&European&$1001916$&$ 78$&$ 0.7785$\tabularnewline
-6-10&25+&European&$ 724317$&$ 51$&$ 0.7041$\tabularnewline
-1-5&0-2&Maori&$  10003$&$ 18$&$17.9946$\tabularnewline
-6-10&0-2&Maori&$  30104$&$ 48$&$15.9447$\tabularnewline
-1-5&3-5&Maori&$  10003$&$  7$&$ 6.9979$\tabularnewline
-6-10&3-5&Maori&$  30104$&$ 11$&$ 3.6540$\tabularnewline
-1-5&6-17&Maori&$  40461$&$ 19$&$ 4.6959$\tabularnewline
-6-10&6-17&Maori&$ 116640$&$ 92$&$ 7.8875$\tabularnewline
-1-5&18-24&Maori&$  15360$&$  5$&$ 3.2552$\tabularnewline
-6-10&18-24&Maori&$  48495$&$  8$&$ 1.6497$\tabularnewline
-1-5&25+&Maori&$  71217$&$  2$&$ 0.2808$\tabularnewline
-6-10&25+&Maori&$ 192729$&$  6$&$ 0.3113$\tabularnewline
-1-5&0-2&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
-6-10&0-2&MLA&$   1290$&$  3$&$23.2558$\tabularnewline
-1-5&3-5&MLA&$    728$&$  0$&$ 0.0000$\tabularnewline
-6-10&3-5&MLA&$   1290$&$  1$&$ 7.7519$\tabularnewline
-1-5&6-17&MLA&$   2991$&$  0$&$ 0.0000$\tabularnewline
-6-10&6-17&MLA&$   4539$&$  1$&$ 2.2031$\tabularnewline
-1-5&18-24&MLA&$   1710$&$  0$&$ 0.0000$\tabularnewline
-6-10&18-24&MLA&$   3078$&$  2$&$ 6.4977$\tabularnewline
-1-5&25+&MLA&$   8028$&$  0$&$ 0.0000$\tabularnewline
-6-10&25+&MLA&$  10335$&$  6$&$ 5.8055$\tabularnewline
-1-5&0-2&Pacific&$   2093$&$  5$&$23.8892$\tabularnewline
-6-10&0-2&Pacific&$  13124$&$ 58$&$44.1938$\tabularnewline
-1-5&3-5&Pacific&$   2093$&$  0$&$ 0.0000$\tabularnewline
-6-10&3-5&Pacific&$  13124$&$  3$&$ 2.2859$\tabularnewline
-1-5&6-17&Pacific&$   8541$&$  5$&$ 5.8541$\tabularnewline
-6-10&6-17&Pacific&$  51183$&$ 22$&$ 4.2983$\tabularnewline
-1-5&18-24&Pacific&$   3972$&$  1$&$ 2.5176$\tabularnewline
-6-10&18-24&Pacific&$  22098$&$  8$&$ 3.6202$\tabularnewline
-1-5&25+&Pacific&$  18492$&$  2$&$ 1.0815$\tabularnewline
-6-10&25+&Pacific&$  91533$&$ 11$&$ 1.2018$\tabularnewline
+1-3&0-1&Asian&$  2269$&$  6$&$26.4434$\tabularnewline
+4-7&0-1&Asian&$  3748$&$ 10$&$26.6809$\tabularnewline
+8-10&0-1&Asian&$  2583$&$  3$&$11.6144$\tabularnewline
+1-3&2-4&Asian&$  3403$&$  1$&$ 2.9386$\tabularnewline
+4-7&2-4&Asian&$  5623$&$  0$&$ 0.0000$\tabularnewline
+8-10&2-4&Asian&$  3874$&$  0$&$ 0.0000$\tabularnewline
+1-3&5-17&Asian&$ 20625$&$  6$&$ 2.9091$\tabularnewline
+4-7&5-17&Asian&$ 29721$&$  8$&$ 2.6917$\tabularnewline
+8-10&5-17&Asian&$ 16777$&$  2$&$ 1.1921$\tabularnewline
+1-3&18-24&Asian&$ 12057$&$  2$&$ 1.6588$\tabularnewline
+4-7&18-24&Asian&$ 23904$&$  3$&$ 1.2550$\tabularnewline
+8-10&18-24&Asian&$ 21063$&$  3$&$ 1.4243$\tabularnewline
+1-3&25+&Asian&$ 54258$&$  4$&$ 0.7372$\tabularnewline
+4-7&25+&Asian&$ 86055$&$ 11$&$ 1.2783$\tabularnewline
+8-10&25+&Asian&$ 54759$&$  8$&$ 1.4609$\tabularnewline
+1-3&0-1&European&$ 23961$&$ 40$&$16.6938$\tabularnewline
+4-7&0-1&European&$ 28051$&$ 58$&$20.6766$\tabularnewline
+8-10&0-1&European&$ 16866$&$ 24$&$14.2298$\tabularnewline
+1-3&2-4&European&$ 35942$&$ 37$&$10.2944$\tabularnewline
+4-7&2-4&European&$ 42076$&$ 22$&$ 5.2286$\tabularnewline
+8-10&2-4&European&$ 25299$&$  6$&$ 2.3716$\tabularnewline
+1-3&5-17&European&$180875$&$160$&$ 8.8459$\tabularnewline
+4-7&5-17&European&$194100$&$118$&$ 6.0793$\tabularnewline
+8-10&5-17&European&$106731$&$ 40$&$ 3.7477$\tabularnewline
+1-3&18-24&European&$ 63051$&$ 24$&$ 3.8064$\tabularnewline
+4-7&18-24&European&$ 94422$&$ 26$&$ 2.7536$\tabularnewline
+8-10&18-24&European&$ 68016$&$ 20$&$ 2.9405$\tabularnewline
+1-3&25+&European&$619698$&$ 54$&$ 0.8714$\tabularnewline
+4-7&25+&European&$734550$&$ 46$&$ 0.6262$\tabularnewline
+8-10&25+&European&$371985$&$ 29$&$ 0.7796$\tabularnewline
+1-3&0-1&Maori&$  3265$&$  9$&$27.5651$\tabularnewline
+4-7&0-1&Maori&$  8378$&$ 13$&$15.5168$\tabularnewline
+8-10&0-1&Maori&$ 15094$&$ 32$&$21.2005$\tabularnewline
+1-3&2-4&Maori&$  4897$&$  4$&$ 8.1683$\tabularnewline
+4-7&2-4&Maori&$ 12568$&$ 12$&$ 9.5481$\tabularnewline
+8-10&2-4&Maori&$ 22642$&$ 10$&$ 4.4166$\tabularnewline
+1-3&5-17&Maori&$ 21750$&$  8$&$ 3.6782$\tabularnewline
+4-7&5-17&Maori&$ 53963$&$ 33$&$ 6.1153$\tabularnewline
+8-10&5-17&Maori&$ 94758$&$ 74$&$ 7.8094$\tabularnewline
+1-3&18-24&Maori&$  7329$&$  3$&$ 4.0933$\tabularnewline
+4-7&18-24&Maori&$ 20862$&$  2$&$ 0.9587$\tabularnewline
+8-10&18-24&Maori&$ 35664$&$  8$&$ 2.2432$\tabularnewline
+1-3&25+&Maori&$ 35934$&$  1$&$ 0.2783$\tabularnewline
+4-7&25+&Maori&$ 87192$&$  3$&$ 0.3441$\tabularnewline
+8-10&25+&Maori&$140820$&$  4$&$ 0.2841$\tabularnewline
+1-3&0-1&MLA&$   273$&$  0$&$ 0.0000$\tabularnewline
+4-7&0-1&MLA&$   519$&$  2$&$38.5356$\tabularnewline
+8-10&0-1&MLA&$   553$&$  1$&$18.0832$\tabularnewline
+1-3&2-4&MLA&$   410$&$  0$&$ 0.0000$\tabularnewline
+4-7&2-4&MLA&$   778$&$  0$&$ 0.0000$\tabularnewline
+8-10&2-4&MLA&$   830$&$  1$&$12.0482$\tabularnewline
+1-3&5-17&MLA&$  1832$&$  0$&$ 0.0000$\tabularnewline
+4-7&5-17&MLA&$  3121$&$  0$&$ 0.0000$\tabularnewline
+8-10&5-17&MLA&$  3250$&$  1$&$ 3.0769$\tabularnewline
+1-3&18-24&MLA&$   906$&$  0$&$ 0.0000$\tabularnewline
+4-7&18-24&MLA&$  1806$&$  0$&$ 0.0000$\tabularnewline
+8-10&18-24&MLA&$  2076$&$  2$&$ 9.6339$\tabularnewline
+1-3&25+&MLA&$  4569$&$  0$&$ 0.0000$\tabularnewline
+4-7&25+&MLA&$  7452$&$  5$&$ 6.7096$\tabularnewline
+8-10&25+&MLA&$  6342$&$  1$&$ 1.5768$\tabularnewline
+1-3&0-1&Pacific&$   643$&$  1$&$15.5521$\tabularnewline
+4-7&0-1&Pacific&$  2113$&$  6$&$28.3956$\tabularnewline
+8-10&0-1&Pacific&$  7389$&$ 48$&$64.9614$\tabularnewline
+1-3&2-4&Pacific&$   964$&$  0$&$ 0.0000$\tabularnewline
+4-7&2-4&Pacific&$  3169$&$  1$&$ 3.1556$\tabularnewline
+8-10&2-4&Pacific&$ 11084$&$ 10$&$ 9.0220$\tabularnewline
+1-3&5-17&Pacific&$  4212$&$  2$&$ 4.7483$\tabularnewline
+4-7&5-17&Pacific&$ 13419$&$  4$&$ 2.9808$\tabularnewline
+8-10&5-17&Pacific&$ 47165$&$ 21$&$ 4.4525$\tabularnewline
+1-3&18-24&Pacific&$  1707$&$  0$&$ 0.0000$\tabularnewline
+4-7&18-24&Pacific&$  6174$&$  4$&$ 6.4788$\tabularnewline
+8-10&18-24&Pacific&$ 18189$&$  5$&$ 2.7489$\tabularnewline
+1-3&25+&Pacific&$  8580$&$  1$&$ 1.1655$\tabularnewline
+4-7&25+&Pacific&$ 26688$&$  3$&$ 1.1241$\tabularnewline
+8-10&25+&Pacific&$ 74757$&$  9$&$ 1.2039$\tabularnewline
 \hline
@@
 \begin{center}
-\includegraphics{draftfinalreport-014}
+\includegraphics{draftfinalreport_v2-023}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-015}
+\includegraphics{draftfinalreport_v2-024}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-017}
+\includegraphics{draftfinalreport_v2-026}
 \end{center}
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Mon Nov 10 12:38:31 2014
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Wed Mar 18 16:47:24 2015
 \begin{tabular}{lrrrrrr}
@@
   \hline
-NULL &  &  & 39 & 1453.62 &  &  \\ 
-  Age & 4 & 1304.20 & 35 & 149.43 & 183.58 & 0.0000 \\ 
-  Ethnicity & 3 & 20.00 & 32 & 129.43 & 3.75 & 0.0285 \\ 
-  NZDep & 1 & 10.70 & 31 & 118.74 & 6.02 & 0.0239 \\ 
-  Age:Ethnicity & 12 & 81.91 & 19 & 36.83 & 3.84 & 0.0045 \\ 
+NULL &  &  & 59 & 1532.80 &  &  \\ 
+  Age & 4 & 1345.17 & 55 & 187.63 & 204.01 & 0.0000 \\ 
+  Ethnicity & 3 & 19.88 & 52 & 167.75 & 4.02 & 0.0141 \\ 
+  NZDep & 2 & 17.10 & 50 & 150.65 & 5.19 & 0.0102 \\ 
+  Age:Ethnicity & 12 & 80.80 & 38 & 69.85 & 4.09 & 0.0004 \\ 
    \hline
@@
 
-\vspace{5mm} %5mm vertical space
-\begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Mon Nov 10 12:38:31 2014
-\begin{tabular}{rrrrr}
-  \hline
- & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
-  \hline
-(Intercept) & -6.4674 & 0.1148 & -56.32 & 0.0000 \\ 
-  Age3-5 & -0.9129 & 0.2054 & -4.44 & 0.0003 \\ 
-  Age6-17 & -0.7618 & 0.1343 & -5.67 & 0.0000 \\ 
-  Age18-24 & -1.5050 & 0.1937 & -7.77 & 0.0000 \\ 
-  Age25+ & -2.9507 & 0.1608 & -18.35 & 0.0000 \\ 
-  EthnicityAsian & 0.0530 & 0.3250 & 0.16 & 0.8721 \\ 
-  EthnicityMaori & 0.2124 & 0.1994 & 1.07 & 0.2999 \\ 
-  EthnicityPacific & 1.1603 & 0.2043 & 5.68 & 0.0000 \\ 
-  NZDep6-10 & -0.2119 & 0.0855 & -2.48 & 0.0227 \\ 
-  Age3-5:EthnicityAsian & -2.0315 & 1.3827 & -1.47 & 0.1581 \\ 
-  Age6-17:EthnicityAsian & -1.0074 & 0.4717 & -2.14 & 0.0459 \\ 
-  Age18-24:EthnicityAsian & -0.8311 & 0.5941 & -1.40 & 0.1779 \\ 
-  Age25+:EthnicityAsian & 0.4211 & 0.4433 & 0.95 & 0.3541 \\ 
-  Age3-5:EthnicityMaori & -0.3864 & 0.4096 & -0.94 & 0.3573 \\ 
-  Age6-17:EthnicityMaori & -0.0855 & 0.2469 & -0.35 & 0.7328 \\ 
-  Age18-24:EthnicityMaori & -0.5828 & 0.4484 & -1.30 & 0.2092 \\ 
-  Age25+:EthnicityMaori & -1.0482 & 0.5242 & -2.00 & 0.0600 \\ 
-  Age3-5:EthnicityPacific & -2.1316 & 0.8139 & -2.62 & 0.0169 \\ 
-  Age6-17:EthnicityPacific & -1.4541 & 0.3347 & -4.35 & 0.0003 \\ 
-  Age18-24:EthnicityPacific & -0.9827 & 0.5129 & -1.92 & 0.0705 \\ 
-  Age25+:EthnicityPacific & -0.6127 & 0.4367 & -1.40 & 0.1767 \\ 
-   \hline
-\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
-\label{table:regression_riskfactors}
-\end{table}
-\vspace{5mm} %5mm vertical space
+% \vspace{5mm} %5mm vertical space
+% \begin{table}
+% % <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
+% % library(xtable)
+% % xtb<-xtable(summary(model3))
+% % print(xtb,floating = FALSE)
+% % @
+% \caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+% \label{table:regression_riskfactors}
+% \end{table}
+% \vspace{5mm} %5mm vertical space
 
@@
 \begin{center}
-\includegraphics{draftfinalreport-020}
+\includegraphics{draftfinalreport_v2-028}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-021}
+\includegraphics{draftfinalreport_v2-029}
 \end{center}
@@
 \label{sub:popim}
+
+%%%%
+
+\begin{figure}
+\begin{center}
+\includegraphics{draftfinalreport_v2-031}
+\end{center}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator}
+\label{fig:naive_au}
+\end{figure}
+%%%%
 
@@
 \begin{center}
-\includegraphics{draftfinalreport-023}
+\includegraphics{draftfinalreport_v2-033}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-037}
+\includegraphics{draftfinalreport_v2-047}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-038}
+\includegraphics{draftfinalreport_v2-048}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-039}
+\includegraphics{draftfinalreport_v2-049}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-040}
+\includegraphics{draftfinalreport_v2-050}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-041}
+\includegraphics{draftfinalreport_v2-051}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-042}
+\includegraphics{draftfinalreport_v2-052}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-043}
+\includegraphics{draftfinalreport_v2-053}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-044}
+\includegraphics{draftfinalreport_v2-054}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-045}
+\includegraphics{draftfinalreport_v2-055}
 \end{center}
@@
 \begin{center}
-\includegraphics{draftfinalreport-053}
+\includegraphics{draftfinalreport_v2-063}
 \end{center}
@@
      \centering
-\includegraphics{draftfinalreport-061}
+\includegraphics{draftfinalreport_v2-071}
      \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (\autoref{section:epidemic_modelling}, \autoref{table:cost20} and \autoref{table:cost50})}
@@
      \centering
-\includegraphics{draftfinalreport-063}
+\includegraphics{draftfinalreport_v2-073}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
@@
      \centering
-\includegraphics{draftfinalreport-064}
+\includegraphics{draftfinalreport_v2-074}
      \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.

### 2015-03-18 revised_draft.R

@@ -444,7 +444,7 @@ dev.off()
 # This chunk uses "results=tex"
 library(Hmisc)
 latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-
+write.csv(PerCap,"percap.csv")
 ## plot this figure?
 #hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
 tp$Ethnicity<- relevel(tp$Ethnicity, "European")
@@ -479,7 +479,8 @@ anova(model2,test="F")
 model3<-update(model2,~.-Ethnicity:NZDep)
 summary(model3)
 anova(model3,test="F")
-
+anovap<-anova(model3,test="F")
+write.csv(anovap,"anovap.csv")
 
 pdf(paste("Cases_regmodel.pdf"), width=7, height=5)
 par(cex.axis=2)

### 2015-03-18 draftfinalreport_v2.Rnw
clean up

@@
 We identify additional analyses that may help understand risk of infection in New Zealand.
-Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007.
+Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007 until mid-2014.
 We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level.
@@
 However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis.
-In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
-\item Of the 11\% of the population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--3).
+However, this was driven by young Pacific children, as 2--24 year old Pacific children are less at risk than European, as are 5--17 year old Asian children.
+\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
-\item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
+\item Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December.
 However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
-\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one.
+\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed 1 and the 2013--2014 outbreaks, as analysed from data until June, is well above 1 (mean estimate 3.66, range 3.4--3.92 for the last outbreak).
+The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
-\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$805,000.
-\item The mean wage loss per measles case is estimated to be \$839
+\item The cost of the 2013--2014 measles outbreaks is estimated to be approximately \$1,332,539 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
+\item The mean wage loss per measles case is estimated to be approximately \$852.
 \item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
-\item The mean management cost per case is \$1,765
-\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1).
-\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but possibly a peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out following additional vaccinations.
+\item The mean public health service cost per case is \$1,765 excluding hospitalisation costs.
+\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170.4 per day.
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed \$8,214 per vaccine before the costs exceed the benefits.
+\item Measles introductions to New Zealand with a mean size of 61 cases per year, median of 2 cases, but rarely a peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out following additional vaccinations.
 \end{itemize}
@@
 The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
-The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}).
-In 2013, but prior to the current 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and other significant outbreaks occurred in the Auckland and Waikato regions in 2013--2014.
+The outbreak in 2011--2012 lasted for more than 12 months.
+In 2013--2014 outbreaks started at the end of December 2013, and the last finished in August 2014.
+In 2013, but prior to the 2013--2014 outbreaks, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
 
@@
 Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see \autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
-Those under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
+Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children (especially young adults) and children under two years of age.
+Those under two year olds are thought to be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mfrow=c(1,1))
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-% @
-% \end{center}
-% \caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
-% \label{fig:ageinyears}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mfrow=c(1,1))
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+@
+\end{center}
+\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+\label{fig:ageinyears}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mfrow=c(1,1))
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mfrow=c(1,1))
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, cases, fill=Age)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%  # theme(text=element_text(size=45))+
-%   ylab("cases")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, cases, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-% #  theme(text=element_text(size=45))+
-%   ylab("cases")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(Age, cases, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-% #  theme(text=element_text(size=45))+
-%   ylab("cases")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, perCap, fill=Age)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%  # theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(Age, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(NZDep, cases, fill=Age)) + 
+  geom_bar(stat="identity", position="dodge")+
+ # theme(text=element_text(size=45))+
+  ylab("cases")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(NZDep, cases, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+#  theme(text=element_text(size=45))+
+  ylab("cases")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(Age, cases, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+#  theme(text=element_text(size=45))+
+  ylab("cases")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(NZDep, perCap, fill=Age)) + 
+  geom_bar(stat="identity", position="dodge")+
+ # theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(NZDep, perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(Age, perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tpsub, aes(NZDep, perCap, fill=Age)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tpsub, aes(NZDep, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tpsub, aes(Age, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(NZDep, perCap, fill=Age)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(NZDep, perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(Age, perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(cex.axis=2)
-% par(mar=c(5,6,4,2)+0.1)
-% hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% #hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-% hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+#hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
-% legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
-% abline(v=12,col="blue")
-% hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-% abline(v=60,col="red")
-% legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-% legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
+legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
+abline(v=12,col="blue")
+hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+abline(v=60,col="red")
+legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
+legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(cex.axis=2)
-% par(mar=c(5,6,4,2)+0.1)
-% plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-% points(naive,pch=16)
-% legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+plot(pop,xlab="Age",ylab="Population",cex.lab=2)
+points(naive,pch=16)
+legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
 
@@
 write.table(propvCohort,"vaccine_proportions_Cohort.txt",sep=",",row.names=T,col.names=F)
+@
+
+\begin{table}[hbtp]
+\footnotesize
+\caption{proportions of the naive population the wuld be covered if vaccinated by age groups}
+<<results=tex,echo=FALSE>>=
+# This chunk uses "results=tex"
+library(Hmisc)
+latex(propvCohort, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:vaccine_cohort}
+\end{table}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
 

### 2015-03-19 draftfinalreport_v2.Rnw
added additional plots

@@
 
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(merge, cases, fill=Ethnicity)) + 
++     geom_bar(stat="identity", position="dodge")+
++     # theme(text=element_text(size=45))+
++     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+tpsub$order <- factor(tpsub$merge, as.character(tpsub$merge))
+ggplot(tpsub, aes(x=order, cases, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("cases")+theme(axis.text.x = element_text(angle = 90, vjust=0,hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(x=order, perCap, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("per capita")+theme(axis.text.x = element_text(angle = 90, vjust=0, hjust = 1))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+tpsub1<-tpsub[ order(tpsub$Age), ]
+tpsub1$order <- factor(tpsub1$merge, as.character(tpsub1$merge))
+ggplot(tpsub1, aes(order, cases, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub1, aes(order, perCap, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("per captita")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
 <<echo=FALSE,include=FALSE,results=hide>>=

### 2015-03-19 draftfinalreport_v2.Rnw
revisions of maps and text for min

@@
 In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--3).
-However, this was driven by young Pacific children, as 2--24 year old Pacific children are less at risk than European, as are 5--17 year old Asian children.
+However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
 \item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
@@
 A measles risk assessment has been undertaken by the Ministry of Health to better assess current and future population immunity and high risk groups.
-Given the current measles outbreak, measles control is a priority for the Ministry and resources are available to control this outbreak and decrease the risk of future outbreaks.
+Given the size of the 2013--2014 outbreaks, prevention of further measles outbreaks is a priority for the Ministry.
 \begin{itemize}
@@
 The report concludes that New Zealand's surveillance system has been performing well and that the Ministry is confident that measles has not been circulating since June 2012 and has not become endemic in NZ.
-We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see \autoref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the current outbreak.
+We agree with the statement that measles did not become endemic and provide some preliminary analyses on the outbreaks since endemic measles elimination (see \autoref{section:epidemic_modelling}) that give information regarding the likelihood of measles persisting within the population and becoming endemic, including analysis of the 2013--2014 outbreaks.
 
@@
 Thus in this report we include regression analyses that help understand risk factors for being measles cases in New Zealand (\autoref{sub:risk_analyses}), descriptions of the New Zealand population in terms of vaccination coverage and immunity (\autoref{sub:immunestat}), and we use those data to model the likely measles outbreak sizes in each District Health Board (DHB) and the country (\autoref{section:epidemic_modelling}).
-We estimate the costs of previous and the current measles outbreaks (\autoref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (\autoref{sub:cost_benefit}).
+We estimate the costs of previous and the 2013--2014 measles outbreaks (\autoref{sub:cost}) and use these estimates, the modelling results and the population data to estimate the benefit--cost ratio for catch-up vaccination programmes (\autoref{sub:cost_benefit}).
 This report includes further developments of the analyses provided in previous interim reports.
@@
 These analyses are intended to build on the analyses already included in the \emph {Progress Towards Measles Elimination in New Zealand - Final} report reviewed above.
-
 We include multivariate modelling to account for confounding within the univariate analyses for measles cases in New Zealand (\autoref{sub:regression}), descriptive analyses of risk of infection due to previous vaccination history (\autoref{sub:immunestat}), and current rates of immunity within the population (\autoref{sub:immunestat}).
@@
 To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (\autoref{sub:imp_risk}).
+The details of the methods and approaches are given in the text below for each section.
 
@@
 DOB<-as.numeric(levels(testv$RptYear))[testv$RptYear]-testv$AgeInYears
-#hist(DOB,breaks=100,col="grey",main="2007- 2014 measles cases",xlab="Date of Birth")
-
+ hist(DOB,breaks=100,col="grey",main="2007- 2014 measles cases",xlab="Date of Birth")
  DOBVac<-table(testv$VC,DOB)
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1)
-% barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
-% legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1)
-% par(mfrow=c(2,1))
-% barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
-% legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1)
-% barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
-% legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1.5)
-% par(mfrow=c(1,1))
-% barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
-% legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
+ \begin{figure}
+ \begin{center}
+ <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1)
+ barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
+ legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
+@
+ \end{center}
+ \caption{test}
+%\label{fig:test}
+ \end{figure}
+
+ \begin{figure}
+ \begin{center}
+ <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1)
+ par(mfrow=c(2,1))
+ barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+ legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1)
+ barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
+ @
+ \end{center}
+ \caption{test}
+\label{fig:test}
+ \end{figure}
+
+ \begin{figure}
+ \begin{center}
+ <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1.5)
+ par(mfrow=c(1,1))
+ barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
+ legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
+ @
+ \end{center}
+ \caption{test}
+\label{fig:test}
+ \end{figure}
 
@@
 
+\subsection{Regression analyses results}
+\label{sub:regression_results}
+
+The distribution of the measles cases per category used in the regression analyses are in \autoref{fig:histcasecat}.
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
+@
+\end{center}
+\caption{Distribution of measles cases per category (\autoref{table:percap}) used in the final regression model (\autoref{eq:reg})}
+\label{fig:histcasecat}
+\end{figure}
+
+The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}.
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width =11, height=11>>=
+res<-predict(model3)
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
+@
+\end{center}
+\caption{Regression model (\autoref{eq:reg}) predictions plotted against the case data (\autoref{table:percap})}
+\label{fig:predict}
+\end{figure}
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+@
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+@
+\end{center}
+\caption{Histogram of residuals from the regression model (\autoref{eq:reg})}
+\label{fig:resid}
+\end{figure}
+
+The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
+\par
+\vspace{5mm} %5mm vertical space
+\begin{table}
+<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
+library(xtable)
+resan<-xtable(anova(model3,test="F"))
+print(resan,floating = FALSE)
+@
+\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using \autoref{eq:reg} and a quasipoisson error structure}
+\label{table:anova}
+\end{table}
+\vspace{5mm} %5mm vertical space
+
+A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.
+
+\vspace{5mm} %5mm vertical space
+\begin{table}
+  <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
+  library(xtable)
+  xtb<-xtable(summary(model3))
+  print(xtb,floating = FALSE)
+  @
+\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+\label{table:regression_riskfactors}
+\end{table}
+\vspace{5mm} %5mm vertical space
+
+\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also \autoref{fig:ageinyears}).
+
+The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
+Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}, \autoref{table:sero}, and \autoref{sub:popim}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}).
+Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
+
 \subsection{Vaccination history of measles infection and population immunity estimation methods}
@@
 To do this we use the census data from NZ statistics (Statistics New Zealand, 2013) and the vaccination and serosurvey data provided by the Ministry of Health at the commencement of this work.
-Specifically, we use the data in Table \autoref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.
+Specifically, we use the data in \autoref{table:sero}, derived from the serosurvey results and the National Immunisation Register (NIR) for under 6 year olds.
 
@@
 
-\subsection{Measles importation risk methods}
-\label{sub:imp_risk}
-For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
-Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
-We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from.
-We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
-We present the data from 2012 because this year had the most complete WHO measles data and yet is most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
-
-\subsection{Regression analyses results}
-\label{sub:regression_results}
-
-The distribution of the meases cases per category used in the regression analyses are in \autoref{fig:histcasecat}.
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
-@
-\end{center}
-\caption{Distribution of measles cases per category (\autoref{table:percap}) used in the final regression model (\autoref{eq:reg})}
-\label{fig:histcasecat}
-\end{figure}
-
-The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}.
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width =11, height=11>>=
-res<-predict(model3)
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-@
-\end{center}
-\caption{Regression model (\autoref{eq:reg}) predictions plotted against the case data (\autoref{table:percap})}
-\label{fig:predict}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-@
-\end{center}
-\caption{Histogram of residuals from the regression model (\autoref{eq:reg})}
-\label{fig:resid}
-\end{figure}
-
-The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
-\par
-\vspace{5mm} %5mm vertical space
-\begin{table}
-<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
-library(xtable)
-resan<-xtable(anova(model3,test="F"))
-print(resan,floating = FALSE)
-@
-\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using \autoref{eq:reg} and a quasipoisson error structure}
-\label{table:anova}
-\end{table}
-\vspace{5mm} %5mm vertical space
-
-A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.
-
-% \vspace{5mm} %5mm vertical space
-% \begin{table}
-% % <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
-% % library(xtable)
-% % xtb<-xtable(summary(model3))
-% % print(xtb,floating = FALSE)
-% % @
-% \caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
-% \label{table:regression_riskfactors}
-% \end{table}
-% \vspace{5mm} %5mm vertical space
-
-\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also \autoref{fig:ageinyears}).
-
-The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
-Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}, \autoref{table:sero}, and \autoref{sub:popim}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}).
-Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
-
 \subsection{Vaccination history and measles infection results}
@@
 A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
+Plotting the cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).
 
@@
 \end{figure}
+
+% ## insert figures here
 
@@
 \end{figure}
+
+\subsection{Measles importation risk methods}
+\label{sub:imp_risk}
+For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
+We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from.
+We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
+We present the data from 2012 and 2013 because 2012 had the most complete WHO measles data, whereas 2013 is most recent, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
 
@@
 \end{figure}
-%%%%
 
@@
 The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures:
-% \begin{itemize}[noitemsep,nolistsep]
-% \item Northland: \autoref{fig:Northland}
-% \item Waitemata: \autoref{fig:Waitemata}
-% \item Auckland: \autoref{fig:Auckland}
-% \item Counties Manukau: \autoref{fig:Counties_Manukau}
-% \item Waikato: \autoref{fig:Waikato}
-% \item Lakes: \autoref{fig:Lakes}
-% \item Bay of Plenty: \autoref{fig:BayofPlenty}
-% \item Tairawhiti: \autoref{fig:Tairawhiti}
-% \item Taranaki: \autoref{fig:Taranaki}
-% \item Hawke's Bay: \autoref{fig:HawkesBay}
-% \item Whanganui: \autoref{fig:Whanganui}
-% \item Mid-Central: \autoref{fig:Midcentral}
-% \item Hutt: \autoref{fig:Hutt}
-% \item Capital and Coast: \autoref{fig:CapitalandCoast}
-% \item Wairarapa: \autoref{fig:Wairarapa}
-% \item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
-% \item West Coast: \autoref{fig:WestCoast}
-% \item Canterbury: \autoref{fig:Canterbury}
-% \item South Canterbury: \autoref{fig:SouthCanterbury}
-% \item Southern: \autoref{fig:Southern}
-% \end{itemize}
+
+\begin{itemize}[noitemsep,nolistsep]
+\item Northland: \autoref{fig:Northland}
+\item Waitemata: \autoref{fig:Waitemata}
+\item Auckland: \autoref{fig:Auckland}
+\item Counties Manukau: \autoref{fig:Counties_Manukau}
+\item Waikato: \autoref{fig:Waikato}
+\item Lakes: \autoref{fig:Lakes}
+\item Bay of Plenty: \autoref{fig:BayofPlenty}
+\item Tairawhiti: \autoref{fig:Tairawhiti}
+\item Taranaki: \autoref{fig:Taranaki}
+\item Hawke's Bay: \autoref{fig:HawkesBay}
+\item Whanganui: \autoref{fig:Whanganui}
+\item Mid-Central: \autoref{fig:Midcentral}
+\item Hutt: \autoref{fig:Hutt}
+\item Capital and Coast: \autoref{fig:CapitalandCoast}
+\item Wairarapa: \autoref{fig:Wairarapa}
+\item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
+\item West Coast: \autoref{fig:WestCoast}
+\item Canterbury: \autoref{fig:Canterbury}
+\item South Canterbury: \autoref{fig:SouthCanterbury}
+\item Southern: \autoref{fig:Southern}
+\end{itemize}
+
+Plotting the lower level data available from the NIR database highlights the issue of surveillance data requiring appropriate denominator data, as these analyses show that there is a lack of correspondence between the national census data and the NIR data (\autoref{fig:nir})
 
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
+\caption{Risk of measles importation to New Zealand in 2013, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 \caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation.
-incidence: measles incidence per million; cover: percent vaccination cover (\%); immigration: the number of travellers entering New Zealand from the country}
+incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of travellers entering New Zealand from the country}
 <<results=tex,echo=FALSE>>=
@@
 
-Globally, analyses of the most complete and recent year of data available, 2012, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}).
+Globally, analyses of the most complete and recent year of data available, 2013, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}).
 Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \autoref{table:imtravel12} and \autoref{fig:immigration12}.
@@
 Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\autoref{table:immigration12} and \autoref{fig:immigration12}.
-Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2012 and immigration numbers in 2012, suggest that those countries with greatest international travel, Australia and New Zealand, pose the greatest risk to New Zealand for measles importation.
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2013 and immigration numbers in 2013, suggest that those countries with greatest international travel, Australia, UK, China and South East Asian countries, pose the greatest risk to New Zealand for measles importation.
 Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand.
@@
 The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in \autoref{table:toprisk}.
+WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday.
 
@@
 \end{center}
-\caption{Measles vaccination cover (\%), 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
+\caption{Measles vaccination cover (\% coverage of children that receive one dose of measles vaccine by their second birthday.), 2012.
+Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Grey indicates not reported.}
 \label{fig:cover12}
@@
 Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses.
-Additional data we have been provided by the Ministry of Health but are yet to incorporate in our risk analyses are finer scale (domicile level) immunisation coverage data from the NIR.
 A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data.
@@
 \item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
-\item There is some statistical support for Pacific and Asian children in the 6--17 year age categories being at lower risk per capita than European or Maori children.
+\item There is some statistical support for 2--24 year old Pacific people being less at risk than European and Maori of the same age, as are 5--17 year old Asian children.
 \item The majority of cases are unvaccinated.
@@
 The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}.
-The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current 2013--2014 outbreak 1.82--2.13.
+The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the 2013--2014 outbreak 1.82--2.13.
 The probability density of the $R_v$ estimates for each outbreak all include one.

### 2015-03-19 maps/rworldmaps_measles_travel.R

@@ -383,6 +383,7 @@ dim(m3)
 head(immigration)
 class(immigration$Month.of.Arrival)
 summary(immigration$Month.of.Arrival)
+immigration<-subset(immigration, Age.Range=="0-19 Years")
 newdata<-immigration[which(immigration$Month.of.Arrival%in%c("2013-01",
                                                            "2013-02",
                                                            "2013-03",

### 2015-03-20 draftfinalreport_v2.Rnw
change to draft

@@
 Previous measles analyses, including two in New Zealand by Prof.
-Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles %\citep{roberts0,roberts4}.
+Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
@@
 <<echo=FALSE,include=FALSE,results=hide>>=
+library(adehabitat)
+library(classInt)
+library(ggplot2)   # plotting library
+library(grid)
+library(gridExtra)
+library(ggmap)
+library(Hmisc)
+library(maps)       # Provides functions that let us plot the maps
+library(mapdata)    # Contains the hi-resolution points that mark out the countries.
+library(maptools)
+library(mapproj)
+library(plyr)
+library(pscl)
+library(RColorBrewer)
+library(rworldmap)
+library(reshape)
+library(reshape2)
+library(rgeos)
+library(rgdal)
+library(raster)
+library(sp)
+library(utils)
+library(xtable)
+#library(SDMTools)
+#gpclibPermit()
+
+# read data
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+# names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+                 data = data , FUN=sum)
+time_yr<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+                   data = time_yr , FUN=sum)
+
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+naive<-round(pop-(pop*impop))
+
+## match cases per age
+AgeInYears<-0:100
+naive<-cbind(AgeInYears,naive)
+colnames(naive)<-c("AgeInYears","Naive")
+naive<-merge(naive,caseyr,by="AgeInYears",all=T)
+colnames(naive)<-c("AgeInYears","Naive","Cases")
+
+dose1=as.factor(round(vac$Dose1Mths/12,0));
+dose1<-summary(dose1)
+dose1<-as.data.frame(dose1)
+dose1$AgeInYears<-rownames(dose1)
+dose2=as.factor(round(na.omit(vac$Dose2Mths/12,0)));
+dose2<-summary(dose2)
+dose2<-as.data.frame(dose2)
+dose2$AgeInYears<-rownames(dose2)
+
+## need to merge vaccination data with data
+## to get years the cases were from
+datav<-merge(vac,data,by="CaseCode",all=T)
+##
+datav$Dose1Mths[which(is.na(datav$Dose1Mths))] <- 9999
+datav$Dose2Mths[which(is.na(datav$Dose2Mths))] <- 9999
+#########################
+datav$RptYear<-as.factor(datav$RptYear)
+datav$SurvWeek<-as.factor(datav$SurvWeek)
+datav$NZDep01<-as.factor(datav$NZDep01)
+datav$NZDep06<-as.factor(datav$NZDep06)
+datav$NZDep13<-as.factor(datav$NZDep13)
+datav$Dose1Mths<-as.factor(datav$Dose1Mths)
+datav$Dose2Mths<-as.factor(datav$Dose2Mths)
+
+#data$AgeInYears<-as.factor(data$AgeInYears)
+datav$EthnicityPrioritised<-as.factor(datav$EthnicityPrioritised)
+
+timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
+                 data = datav , FUN=sum,na.rm=F,na.action=na.pass)
+
+testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+
+testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
+testv$Dose2Mths<-revalue(testv$Dose2Mths, c("9999"=NA));
+testv$Dose1Mths<-as.numeric(testv$Dose1Mths)
+testv$Dose2Mths<-as.numeric(testv$Dose2Mths)
+testv$D2vac<-ifelse(testv$Dose1Mths > -1 & testv$Dose2Mths >= testv$Dose1Mths,testv$Dose2Mths,NA)
+testv$D1vac<-ifelse(testv$Dose1Mths > -1 & is.na(testv$D2vac) == T,testv$Dose1Mths,NA)
+testv$Unvac<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,testv$AgeInYears,NA)
+
+testv$VC<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,0,
+                 ifelse(testv$Dose1Mths >= 0 & is.na(testv$Dose2Mths) == T,1,2))
+AgeVac<-table(testv$VC,testv$AgeInYears)
+
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+AgeInYears<-(as.numeric(rownames(AgeVac)))
+AgeVac<-cbind(AgeVac,AgeInYears)       
+AgeInYears<-(as.numeric(rownames(pop)))
+pop<-cbind(pop,AgeInYears)
+colnames(pop)<-c("Population","AgeInYears")
+AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
+
+### denominator data
+# setwd("~/Massey_2014/measles/data")
+data<-read.csv("DHayman_20140627.csv",header=T)
+vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
+names(data)
+data$RptYear<-as.factor(data$RptYear)
+data$SurvWeek<-as.factor(data$SurvWeek)
+data$NZDep01<-as.factor(data$NZDep01)
+data$NZDep06<-as.factor(data$NZDep06)
+data$NZDep13<-as.factor(data$NZDep13)
+#data$AgeInYears<-as.factor(data$AgeInYears)
+data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
+
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
+                 data = data , FUN=sum)
+
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+caseyr<-aggregate( DiseaseName ~ AgeInYears, 
+                   data = test , FUN=sum)
+test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
+test$AgeInYears<-as.factor(test$AgeInYears)
+tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
+test$AgeInYears<-tage
+#summary(test$EthnicityPrioritised)
+teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
+                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
+
+test$EthnicityPrioritised<-teth
+
+test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
+test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
+
+# want to set NZ Deprivation to the appropriate year
+
+testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
+                              ifelse (test$RptYear == "2008",test$NZDep06,
+                              ifelse (test$RptYear == "2009",test$NZDep06,
+                              ifelse (test$RptYear == "2010",test$NZDep06,
+                              ifelse (test$RptYear == "2011",test$NZDep06,
+                              ifelse (test$RptYear == "2012",test$NZDep06,
+                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
+testt$NZDep<-as.factor(testt$NZDep-1)
+
+testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
+                      ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
+                 data = testt , FUN=sum)
+
+colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
+
+tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
+                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
+                                 "9"="6-10","10"="6-10"));
+
+testtable$NZDep<-tnzd
+
+tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
+                 data = testtable , FUN=sum)
+
+ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
+                 data = testtable , FUN=sum)
+denom<-read.csv("NZDep2006Denominators.csv",header=T)
+denom<-denom[,-c(9)]
+
+denom<-denom[,1:18]
+row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
+
+denomt <- denom[!row.has.na,]
+
+denomt<-denomt[denomt$Sex_code==99,]
+denomt$Eth_Level<-as.factor(denomt$Eth_Level)
+denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
+dm<-cbind(denomt[,c(3,6,9:18)])
+
+dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
+dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
+dm$Age_Label<-dage
+colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
+
+datadm<-dm[dm$Age=="<6",]
+dataA<-(rbind(datadm,datadm)/2)
+dataA<-round(dataA[,3:12])
+dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
+colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
+ddm<-dm[!dm$Age=="<6",]
+dm<-rbind(dataA,ddm)
+dm$Ethnicity <- factor(dm$Ethnicity)
+deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
+                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
+dm$Ethnicity<-deth
+head(dm)
+
+popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
+colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
+pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
+                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
+                                 "Dep9"="6-10","Dep10"="6-10"));
+
+popn$NZDep<-pnzd
+tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
+               data = popn , FUN=sum)
+
+tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
+ttyr$merge <- paste(ttyr$NZDep, ttyr$Age, ttyr$Ethnicity)
+ttyr <- ttyr[ttyr$Ethnicity!="None",]
+
+tp$cases <- 0
+cases <- matrix(0, length(tp$merge),1)
+rownames(cases) <- tp$merge
+cases[ttyr$merge,] <- ttyr$Cases
+tp$cases <- cases
+PerCapita<-(tp$cases/tp$Popn)
+PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita*10000,4))
+colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
+
 ## get data
@@
 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
+
 test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mfrow=c(1,1))
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-@
-\end{center}
-\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
-\label{fig:ageinyears}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## plot cases / age class
+
 caseyr<-aggregate( DiseaseName ~ AgeInYears, 
 data = test , FUN=sum)
-#caseyr
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
+
 popimmune<-read.csv("PopnImmunityAll.csv",header=T)
@@
 rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
+
 naive<-round(pop-(pop*impop))
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
+
 ## match cases per age
@@
 testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-library(plyr)
+
 testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
@@
 AgeVac<-table(testv$VC,testv$AgeInYears)
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
+
 row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
@@
 AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
-# head(AgeV)
-#AgeVac<-t(AgeVac)
-@
-
-This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 (least deprived) and NZDep 6--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}).
-We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
-By doing so, we reduce the zeroinflation present in the data.
-
-The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, \autoref{table:none}).
-<<echo=FALSE,include=FALSE,results=hide>>=
+
 ## regression analyses - measles
@@
 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
 
 test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
+
 caseyr<-aggregate( DiseaseName ~ AgeInYears, 
 data = test , FUN=sum)
-caseyr
-test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
+
+# test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
+test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
 test$AgeInYears<-as.factor(test$AgeInYears)
-tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
+#tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
+tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
 test$AgeInYears<-tage
@@
 test$EthnicityPrioritised<-teth
-head(test)
+
 test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
@@
 ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-head(testt)
-tail(testt)
-str(testt)
+
 testt$NZDep<-as.factor(testt$NZDep-1)
@@
 
-tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-                                 "9"="6-10","10"="6-10"));
+tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
+                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
+                                 "9"="8-10","10"="8-10"));
 
 testtable$NZDep<-tnzd
-# head(testtable)
+
 tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
@@
 ttyr <- ttyr[ttyr$Ethnicity=="None",]
-@
-
-\begin{table}[hbtp]
-\footnotesize
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(ttyr, file="",table.env=FALSE,rowname=NULL)
-@
-\caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
-\label{table:none}
-\end{table}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
 ### denominator data
-# setwd("~/Massey_2014/measles/data")
+
 data<-read.csv("DHayman_20140627.csv",header=T)
 vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
+
 data$RptYear<-as.factor(data$RptYear)
@@
 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
 
 test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
+
 caseyr<-aggregate( DiseaseName ~ AgeInYears, 
@@
 caseyr
-test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
+## for MoH desired ages
+test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
 test$AgeInYears<-as.factor(test$AgeInYears)
-tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
+tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
 test$AgeInYears<-tage
@@
 
-tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-                                 "9"="6-10","10"="6-10"));
+tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
+                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
+                                 "9"="8-10","10"="8-10"));
 
@@
 
-popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
-colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
-popn
-pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
-                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
-                                 "Dep9"="6-10","Dep10"="6-10"));
-
-popn$NZDep<-pnzd
-head(popn)
-tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
-               data = popn , FUN=sum)
-
-tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
-ttyr$merge <- paste(ttyr$NZDep, ttyr$Age, ttyr$Ethnicity)
-ttyr <- ttyr[ttyr$Ethnicity!="None",]
-
-tp$cases <- 0
-cases <- matrix(0, length(tp$merge),1)
-rownames(cases) <- tp$merge
-cases[ttyr$merge,] <- ttyr$Cases
-tp$cases <- cases
-PerCapita<-(tp$cases/tp$Popn)
-PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita*10000,4))
-colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-
-## get data
-# set wd
-# setwd("~/Massey 2014/DHayman_20140627")
-# read data
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mfrow=c(1,1))
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-## plot cases / age class
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-data = test , FUN=sum)
-
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-colnames(naive)<-c("AgeInYears","Naive","Cases")
-
-dose1=as.factor(round(vac$Dose1Mths/12,0));
-dose1<-summary(dose1)
-dose1<-as.data.frame(dose1)
-dose1$AgeInYears<-rownames(dose1)
-dose2=as.factor(round(na.omit(vac$Dose2Mths/12,0)));
-dose2<-summary(dose2)
-dose2<-as.data.frame(dose2)
-dose2$AgeInYears<-rownames(dose2)
-
-## need to merge vaccination data with data
-## to get years the cases were from
-datav<-merge(vac,data,by="CaseCode",all=T)
-##
-datav$Dose1Mths[which(is.na(datav$Dose1Mths))] <- 9999
-datav$Dose2Mths[which(is.na(datav$Dose2Mths))] <- 9999
-#########################
-datav$RptYear<-as.factor(datav$RptYear)
-datav$SurvWeek<-as.factor(datav$SurvWeek)
-datav$NZDep01<-as.factor(datav$NZDep01)
-datav$NZDep06<-as.factor(datav$NZDep06)
-datav$NZDep13<-as.factor(datav$NZDep13)
-datav$Dose1Mths<-as.factor(datav$Dose1Mths)
-datav$Dose2Mths<-as.factor(datav$Dose2Mths)
-
-#data$AgeInYears<-as.factor(data$AgeInYears)
-datav$EthnicityPrioritised<-as.factor(datav$EthnicityPrioritised)
-
-timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
-data = datav , FUN=sum,na.rm=F,na.action=na.pass)
-
-testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-library(plyr)
-testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
-testv$Dose2Mths<-revalue(testv$Dose2Mths, c("9999"=NA));
-testv$Dose1Mths<-as.numeric(testv$Dose1Mths)
-testv$Dose2Mths<-as.numeric(testv$Dose2Mths)
-testv$D2vac<-ifelse(testv$Dose1Mths > -1 & testv$Dose2Mths >= testv$Dose1Mths,testv$Dose2Mths,NA)
-testv$D1vac<-ifelse(testv$Dose1Mths > -1 & is.na(testv$D2vac) == T,testv$Dose1Mths,NA)
-testv$Unvac<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,testv$AgeInYears,NA)
-
-testv$VC<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,0,
-ifelse(testv$Dose1Mths >= 0 & is.na(testv$Dose2Mths) == T,1,2))
-AgeVac<-table(testv$VC,testv$AgeInYears)
-
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-AgeInYears<-(as.numeric(rownames(AgeVac)))
-AgeVac<-cbind(AgeVac,AgeInYears)       
-AgeInYears<-(as.numeric(rownames(pop)))
-pop<-cbind(pop,AgeInYears)
-colnames(pop)<-c("Population","AgeInYears")
-AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
-
-## regression analyses - measles
-# setwd("~/Massey 2014/DHayman_20140627")
-# read data
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
-
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-data = test , FUN=sum)
-caseyr
-# test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
-test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
-test$AgeInYears<-as.factor(test$AgeInYears)
-#tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
-tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
-test$AgeInYears<-tage
-#summary(test$EthnicityPrioritised)
-teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
-"Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
-
-test$EthnicityPrioritised<-teth
-head(test)
-test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
-test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
-
-# want to set NZ Deprivation to the appropriate year
-
-testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
-ifelse (test$RptYear == "2008",test$NZDep06,
-ifelse (test$RptYear == "2009",test$NZDep06,
-ifelse (test$RptYear == "2010",test$NZDep06,
-ifelse (test$RptYear == "2011",test$NZDep06,
-ifelse (test$RptYear == "2012",test$NZDep06,
-ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-head(testt)
-tail(testt)
-str(testt)
-testt$NZDep<-as.factor(testt$NZDep-1)
-
-testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
-~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
-data = testt , FUN=sum)
-dim(testtable)
-colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
-head(testtable)
-
-#tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-#"5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-#"9"="6-10","10"="6-10"));
-
-tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
-                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
-                                 "9"="8-10","10"="8-10"));
-
-testtable$NZDep<-tnzd
-# head(testtable)
-tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
-data = testtable , FUN=sum)
-
-ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
-data = testtable , FUN=sum)
-ttyr <- ttyr[ttyr$Ethnicity=="None",]
-
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(ttyr, file="",table.env=FALSE,rowname=NULL)
-
-### denominator data
-# setwd("~/Massey_2014/measles/data")
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
-
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-head(test)
-dim(test)
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-data = test , FUN=sum)
-caseyr
-## for MoH desired ages
-test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
-test$AgeInYears<-as.factor(test$AgeInYears)
-tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
-test$AgeInYears<-tage
-#summary(test$EthnicityPrioritised)
-teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
-"Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
-
-test$EthnicityPrioritised<-teth
-head(test)
-test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
-test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
-
-# want to set NZ Deprivation to the appropriate year
-
-testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
-ifelse (test$RptYear == "2008",test$NZDep06,
-ifelse (test$RptYear == "2009",test$NZDep06,
-ifelse (test$RptYear == "2010",test$NZDep06,
-ifelse (test$RptYear == "2011",test$NZDep06,
-ifelse (test$RptYear == "2012",test$NZDep06,
-ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-head(testt)
-tail(testt)
-str(testt)
-testt$NZDep<-as.factor(testt$NZDep-1)
-
-testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
-~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
-data = testt , FUN=sum)
-dim(testtable)
-colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
-head(testtable)
-
-#tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-#"5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-#"9"="6-10","10"="6-10"));
-
-tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
-                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
-                                 "9"="8-10","10"="8-10"));
-
-testtable$NZDep<-tnzd
-# head(testtable)
-tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
-data = testtable , FUN=sum)
-
-ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
-data = testtable , FUN=sum)
-denom<-read.csv("NZDep2006Denominators.csv",header=T)
-head(denom)
-denom<-denom[,-c(9)]
-summary(denom)
-dim(denom)
-denom<-denom[,1:18]
-row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
-sum(row.has.na)
-denomt <- denom[!row.has.na,]
-head(denomt)
-denomt<-denomt[denomt$Sex_code==99,]
-denomt$Eth_Level<-as.factor(denomt$Eth_Level)
-denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
-dm<-cbind(denomt[,c(3,6,9:18)])
-head(dm)
-dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
-dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
-dm$Age_Label<-dage
-colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-head(dm)
-datadm<-dm[dm$Age=="<6",]
-dataA<-(rbind(datadm,datadm)/2)
-dataA<-round(dataA[,3:12])
-dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
-colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-ddm<-dm[!dm$Age=="<6",]
-dm<-rbind(dataA,ddm)
-dm$Ethnicity <- factor(dm$Ethnicity)
-deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
-"Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
-dm$Ethnicity<-deth
-head(dm)
-
 ## here's the additional fudge the MoH asked for.
@@
 popn
-#pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
-#"Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
-#"Dep9"="6-10","Dep10"="6-10"));
 
@@
 
-#par(mar=c(12,6,4,2)+0.1)
-#par(cex.axis=2)
-#plot(tp$cases/tp$Popn*10000,main="",
-#        ylab="Per Capita Rate Per 10,000",
-#        pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-#xlab="",cex.lab=2,xaxt="n")
-#axis(side = 1, at = 1:75, labels=rep(c("0-2yr, 1-3","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),
-#     las=2,cex.axis=1.5)
-#mtext(side=1,line=10,text="Age and NZDep",cex=2)
-#legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
-#col=c(1:4,6),pch=16,bty="n",cex=1.8)
-#par(mar = c(5,4,4,2) + 0.1)
-#
-
 tp$perCap<-with(tp,tp$cases/tp$Popn*10000)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(NZDep, cases, fill=Age)) + 
-  geom_bar(stat="identity", position="dodge")+
- # theme(text=element_text(size=45))+
-  ylab("cases")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(NZDep, cases, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-#  theme(text=element_text(size=45))+
-  ylab("cases")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(Age, cases, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-#  theme(text=element_text(size=45))+
-  ylab("cases")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(NZDep, perCap, fill=Age)) + 
-  geom_bar(stat="identity", position="dodge")+
- # theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(NZDep, perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(Age, perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
+
 tpsub<-tp[!(tp$Ethnicity=="MLA"),]
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(NZDep, perCap, fill=Age)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(NZDep, perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(Age, perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(merge, cases, fill=Ethnicity)) + 
-+     geom_bar(stat="identity", position="dodge")+
-+     # theme(text=element_text(size=45))+
-+     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-tpsub$order <- factor(tpsub$merge, as.character(tpsub$merge))
-ggplot(tpsub, aes(x=order, cases, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("cases")+theme(axis.text.x = element_text(angle = 90, vjust=0,hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(x=order, perCap, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("per capita")+theme(axis.text.x = element_text(angle = 90, vjust=0, hjust = 1))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+tpsub$ordert <- factor(tpsub$merge, as.character(tpsub$merge))
+
 tpsub1<-tpsub[ order(tpsub$Age), ]
-tpsub1$order <- factor(tpsub1$merge, as.character(tpsub1$merge))
-ggplot(tpsub1, aes(order, cases, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub1, aes(order, perCap, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("per captita")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-
-## plot this figure?
-#hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
+tpsub1$ordert <- factor(tpsub1$merge, as.character(tpsub1$merge))
+
 tp$Ethnicity<- relevel(tp$Ethnicity, "European")
@@
 ## use tpsub
-#hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
 
@@
 anova(model3,test="F")
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-res<-predict(model3)
-#res<-predict(model2)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-#hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-library(xtable)
-resan<-xtable(anova(model3,test="F"))
-#resan<-xtable(anova(model2,test="F"))
-print(resan,floating = FALSE)
-
-library(xtable)
-#xtb<-xtable(summary(model3))
-xtb<-xtable(summary(model2))
-print(xtb,floating = FALSE)
 
@@
 AgeVac<-t(AgeVac)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
+
 DOB<-as.numeric(levels(testv$RptYear))[testv$RptYear]-testv$AgeInYears
@@
  DOBVac<-t(DOBVac)
- @
-
- \begin{figure}
- \begin{center}
- <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1)
- barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
- legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
-@
- \end{center}
- \caption{test}
-%\label{fig:test}
- \end{figure}
-
- \begin{figure}
- \begin{center}
- <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1)
- par(mfrow=c(2,1))
- barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
- legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1)
- barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
-legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
- @
- \end{center}
- \caption{test}
-\label{fig:test}
- \end{figure}
-
- \begin{figure}
- \begin{center}
- <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1.5)
- par(mfrow=c(1,1))
- barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
- legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
- @
- \end{center}
- \caption{test}
-\label{fig:test}
- \end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
+
 popimmune<-read.csv("PopnImmunityAll.csv",header=T)
@@
 naive<-round(pop-(pop*impop))
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
 
@@
 write.table(propvCohort,"vaccine_proportions_Cohort.txt",sep=",",row.names=T,col.names=F)
-@
-
-\begin{table}[hbtp]
-\footnotesize
-\caption{proportions of the naive population the wuld be covered if vaccinated by age groups}
-<<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
-latex(propvCohort, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:vaccine_cohort}
-\end{table}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
 
@@
 
-write.table(table11,"vaccine_proportions_11yrOlds.txt",sep=",",row.names=F)
-write.table(table120,"vaccine_proportions_1_20yrOlds.txt",sep=",",row.names=F)
-write.table(table201,"vaccine_proportions_20_1yrOlds.txt",sep=",",row.names=F)
-
 ###
-@
+
+tp$Ethnicity<- relevel(tp$Ethnicity, "European")
+model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
+summary(model)
+
+model1<-update(model,~.-Age:Ethnicity:NZDep)
+summary(model1)
+anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
+
+model2<-update(model1,~.-Age:NZDep)
+summary(model2)
+anova(model2,test="F")
+
+tpsub<-tp[!(tp$Ethnicity=="MLA"),]
+
+model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
+summary(model)
+
+model1<-update(model,~.-Age:Ethnicity:NZDep)
+summary(model1)
+anova(model1,test="F")
+
+model2<-update(model1,~.-Ethnicity:NZDep)
+summary(model2)
+anova(model2,test="F")
+
+model3<-update(model2,~.-Age:NZDep)
+summary(model3)
+anova(model3,test="F")
+
+res<-predict(model3)
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+
+AgeVac<-table(testv$VC,testv$AgeInYears)
+row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
+AgeVac<-t(AgeVac)
+
+# read in data from NIR
+data_d <- read.table("NIR_MMR_data_20060101-20140731/NIR_MMR_data_20060101-20140731.txt", header=T, sep="|")
+
+# read in domicile <-> AU map
+data_dom <- read.csv("NIR_MMR_data_20060101-20140731/domicile_code.csv")
+
+# merge NIR with domicile/AU so we have AU for each NIR
+data_merge <- merge(data_d, data_dom, by.x = "DOMICILE_CODE", by.y="dom", all=T)
+
+# TODO: figure out how many kids are immunised in each AU
+# find number of kids by BIRTH_CALENDAR_YEAR, AU, VACCINE_DOSE
+
+num_kids <- table(data_merge$BIRTH_CALENDAR_YEAR, data_merge$area.unit, data_merge$VACCINE_DOSE)
+# drop the 0's from vaccine_dose
+dose1 <- t(num_kids[,,2])
+dose2 <- t(num_kids[,,3])
+dose1 <- as.data.frame(dose1[,-ncol(dose1)])
+dose2 <- as.data.frame(dose2[,-ncol(dose2)])
+dose1$cau <- rownames(dose1)
+dose2$cau <- rownames(dose2)
+
+# read in shapefile for AU's
+au <- readShapeSpatial("AU2013_GV_Full")
+
+au_data <- slot(au, "data")
+
+# read in denominator data for AU's
+data_denom<-read.csv("2013-mb-dataset-Total-New-Zealand-individual-part-1_AU.csv",header=T)
+denom<-data_denom[,1]
+for (i in 1:5)
+  denom <- cbind(denom, data_denom[,3]/5)
+for (i in 1:5)
+  denom <- cbind(denom, data_denom[,4]/5)
+colnames(denom)[2:11] <- 2013:(2013-9)
+denom <- denom[,c(1,9:2)]
+
+dose2_merge <- merge(dose2, denom, by.x="cau", by.y="denom", all=T)
+diff <- dose2_merge[,2:8+7] - dose2_merge[,2:8]
+image(as.matrix(diff > 0))
+
+dose2_num <- data.frame(cau=dose2_merge$cau, vacc = rowSums(dose2_merge[,2:5]), total = rowSums(dose2_merge[,10:13]))
+diff <- dose2_num$total - dose2_num$vacc
+
+dose1_merge <- merge(dose1, denom, by.x="cau", by.y="denom", all=T)
+diff <- dose1_merge[,2:8] - dose1_merge[,2:8+7]
+image(as.matrix(diff > 0))
+
+au_new <- merge(au_data, dose2_num, by.x="AU2013", by.y="cau", all.x=T)
+au_new$prop <- au_new$vacc / au_new$total
+
+fixedBreaks <- c(0,0.5,0.7, 0.9, 1.1, 1.3, 1.5, Inf)
+au_new$prop[!is.finite(au_new$prop)] <- NA
+breaks <- classIntervals(au_new$prop, n=7, style="fixed", fixedBreaks=fixedBreaks)
+pal <- brewer.pal("Spectral", n=7)
+
+cols <- findColours(breaks, pal)
+plot(au, col=cols)
+
+## plot to show Ministry the data issues...
+fixedBreaks <- c(0,0.6,0.8,1, 1.2,1.4,Inf)
+au_new$prop[!is.finite(au_new$prop)] <- NA
+breaks <- classIntervals(au_new$prop, n=6, style="fixed", fixedBreaks=fixedBreaks)
+pal <- brewer.pal("RdGy", n=6)
+pal<-c("red","orange","grey","blue","green","yellow")
+cols <- findColours(breaks, pal)
+
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+
+totdpdf<-read.csv("mapdata_tot.csv",header=T)
+# output top countries
+colnames(totdpdf)[1]<-c("country")
+toptot <- totdpdf[order(-totdpdf$immigration),] 
+immigration<-toptot[1:30,c(1,4)]
+topincidence <- totdpdf[order(-totdpdf$incidence),] 
+incidence<-topincidence[1:30,c(1,2)]
+incidence[,2]<-round(incidence[,2],0)
+topvaccine <- totdpdf[order(totdpdf$cover),] 
+vaccinecover<-topvaccine[1:30,c(1,3)]
+toprisk <- totdpdf[order(-totdpdf$risk),] 
+risk<-(toprisk[1:30,c(1,5)])
+risk[,2]<-round(risk[,2],0)
+toprisk2<-(toprisk[1:10,c(1,2,3,4)])
+toprisk2[,2]<-round(toprisk2[,2],0)
+
+nzdpdf<-read.csv("mapdata_nzers.csv",header=T)
+# output top countries
+colnames(nzdpdf)[1]<-c("country")
+topnz <- nzdpdf[order(-nzdpdf$immigration),] 
+nz<-topnz[1:30,c(1,4)]
+topnzrisk <- nzdpdf[order(-nzdpdf$risk),] 
+nzrisk<-(topnzrisk[1:30,c(1,5)])
+nzrisk[,2]<-round(nzrisk[,2],0)
+
+imdpdf<-read.csv("mapdata_immigration.csv",header=T)
+# output top countries
+colnames(imdpdf)[1]<-c("country")
+topim <- imdpdf[order(-imdpdf$immigration),] 
+im<-topim[1:30,c(1,4)]
+topimrisk <- imdpdf[order(-imdpdf$risk),] 
+imrisk<-(topimrisk[1:30,c(1,5)])
+imrisk[,2]<-round(imrisk[,2],0)
+# haven't had time to sort out which packages I need
+
+#getwd()
+#setwd("~/data")
+#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
+load("world.ggmap.Rda")
+n <- length(unique(world.ggmap$id))
+id = unique(world.ggmap$id)
+id<-as.data.frame(id)
+dpdf<-read.csv("mapdata_tot.csv",header=T)
+# output top countries
+
+np1<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+np2<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = incidence), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+np3<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = cover), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+  xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
+
+#grid.arrange(np1, np2, np3, ncol=3, main="")
+
+np4<-ggplot(dpdf, aes(map_id = id)) +
+  geom_map(aes(fill = risk), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
+#np4
+# output top countries
+dpdf<-read.csv("mapdata_immigration.csv",header=T)
+imnp1<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+imnp4<-ggplot(dpdf, aes(map_id = id)) +
+  geom_map(aes(fill = risk), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
+#np4
+dpdf<-read.csv("mapdata_nzers.csv",header=T)
+nznp1<-ggplot(dpdf, aes(map_id = id))  +
+  geom_map(aes(fill = immigration), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+   theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
+
+nznp4<-ggplot(dpdf, aes(map_id = id)) +
+  geom_map(aes(fill = risk), map =world.ggmap) +
+  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
+   xlab("longitude") +
+  ylab("latitude") +
+  theme(text = element_text(size=30))+
+  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
+#np4
+
+change<-read.csv("change.csv",header=T)
+costtable1<-read.csv("costtable1.csv",header=T)
+costtable2<-read.csv("costtable2v2.csv",header=T)
+costtable3<-read.csv("costtable3.csv",header=T)
+
+hosp<-read.csv("hospital.csv",header=T)
+
+p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")+theme(axis.title.x = element_text(size=25),axis.title.y = element_text(size=25),axis.text.x  = element_text(size=16),axis.text.y  = element_text(size=16))
+
+costtable4<-read.csv("costtable4v2.csv",header=T)
+costtable5<-read.csv("costtable5v2.csv",header=T)
+vacp<-read.csv("vacc_predictions.csv",header=T)
+
+costtable20<-read.csv("cost_benefit_20.csv",header=T)
+costtable50<-read.csv("cost_benefit_50.csv",header=T)
+
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+pop<-read.csv("popnsize.csv",header=T)
+colnames(pop)<-0:100
+pop<-t(pop)
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
+         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+length(pop)
+naive<-round(pop-(pop*impop))
+
+## match cases per age
+AgeInYears<-0:100
+naive<-cbind(AgeInYears,naive)
+colnames(naive)<-c("AgeInYears","Naive")
+naive<-merge(naive,caseyr,by="AgeInYears",all=T)
+#naive[is.na(naive)] <- 0
+colnames(naive)<-c("AgeInYears","Naive","Cases")
+@
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mfrow=c(1,1))
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+hist(time_yr$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
+legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+@
+\end{center}
+\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+\label{fig:ageinyears}
+\end{figure}
+
+This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases in, particularly in the adult age classes.
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 (least deprived) and NZDep 6--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}).
+We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
+By doing so, we reduce the zeroinflation present in the data.
+
+The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, \autoref{table:none}).
+
+\begin{table}[hbtp]
+\footnotesize
+<<results=tex,echo=FALSE>>=# This chunk uses "results=tex"
+latex(ttyr, file="",table.env=FALSE,rowname=NULL)
+@
+\caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
+\label{table:none}
+\end{table}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(x=NZDep, y=cases, fill=Age)) + 
+  geom_bar(stat="identity", position="dodge")+
+ # theme(text=element_text(size=45))+
+  ylab("cases")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(x=NZDep, y=cases, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+#  theme(text=element_text(size=45))+
+  ylab("cases")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(x=Age, y=cases, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+#  theme(text=element_text(size=45))+
+  ylab("cases")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(x=NZDep,y= perCap, fill=Age)) + 
+  geom_bar(stat="identity", position="dodge")+
+ # theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(x=NZDep, y=perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(x=Age,y= perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(x=NZDep,y= perCap, fill=Age)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(x=NZDep,y= perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(x=Age, y=perCap, fill=Ethnicity)) + 
+  geom_bar(stat="identity", position="dodge")+
+  #theme(text=element_text(size=45))+
+  ylab("Per capita per 10000")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tp, aes(x=merge, y=cases, fill=Ethnicity)) + 
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(x=ordert, y=cases, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("cases")+theme(axis.text.x = element_text(angle = 90, vjust=0,hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub, aes(x=ordert, y=perCap, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("per capita")+theme(axis.text.x = element_text(angle = 90, vjust=0, hjust = 1))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+
+ggplot(tpsub1, aes(x=ordert,y= cases, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ggplot(tpsub1, aes(x=ordert,y= perCap, fill=Ethnicity)) +
+     geom_bar(stat="identity", position="dodge")+
+     # theme(text=element_text(size=45))+
+     ylab("per captita")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+<<echo=FALSE,include=FALSE>>=
+#cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+@
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+#hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
+legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(mar=c(5,6,4,2)+0.1)
+par(cex.axis=2)
+hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
+abline(v=12,col="blue")
+hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
+abline(v=60,col="red")
+legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
+legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+ \begin{figure}
+ \begin{center}
+ <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1)
+ barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
+ legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
+@
+ \end{center}
+ \caption{test}
+%\label{fig:test}
+ \end{figure}
+
+ \begin{figure}
+ \begin{center}
+ <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1)
+ par(mfrow=c(2,1))
+ barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+ legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1)
+ barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
+ @
+ \end{center}
+ \caption{test}
+\label{fig:test}
+ \end{figure}
+
+ \begin{figure}
+ \begin{center}
+ <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+ par(mar=c(5,6,4,2)+0.1)
+ par(cex.axis=1.5)
+ par(mfrow=c(1,1))
+ barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
+ legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
+ @
+ \end{center}
+ \caption{test}
+\label{fig:test}
+ \end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
+par(cex.axis=2)
+par(mar=c(5,6,4,2)+0.1)
+plot(pop,xlab="Age",ylab="Population",cex.lab=2)
+points(naive,pch=16)
+legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
+@
+\end{center}
+\caption{test}
+%\label{fig:test}
+\end{figure}
+
+\begin{table}[hbtp]
+\footnotesize
+\caption{Per capita measles rates}
+<<results=tex,echo=FALSE>>=
+latex(PerCap, file="", table.env=FALSE,rowname=NULL)
+@
+\label{table:percap_rates}
+\end{table}
+
+\begin{table}[hbtp]
+\footnotesize
+\caption{proportions of the naive population the would be covered if vaccinated by age groups}
+<<results=tex,echo=FALSE>>=
+latex(propvCohort, file="", table.env=FALSE)
+@
+\label{table:vaccine_cohort}
+\end{table}
 
@@
 
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=11>>=
-par(mar=c(12,6,4,2)+0.1)
-par(cex.axis=2)
-plot(tp$cases/tp$Popn*10000,main="",ylab="Per Capita Rate Per 10,000",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-     xlab="",cex.lab=2,xaxt="n")
-axis(side = 1, at = 1:50, labels=rep(c("0-2yr, 1-5","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),las=2,cex.axis=1.5)
-mtext(side=1,line=10,text="Age and NZDep",cex=2)
-legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
-       col=c(1:4,6),pch=16,bty="n",cex=1.8)
-par(mar = c(5,4,4,2) + 0.1)
-@
-\end{center}
-\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (\autoref{table:percap} for details)}
-\label{fig:percap}
-\end{figure}
+% \begin{figure}
+% \begin{center}
+% <<fig=TRUE,echo=FALSE,width = 15,height=11>>=
+% par(mar=c(12,6,4,2)+0.1)
+% par(cex.axis=2)
+% plot(tp$cases/tp$Popn*10000,main="",ylab="Per Capita Rate Per 10,000",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
+%      xlab="",cex.lab=2,xaxt="n")
+% axis(side = 1, at = 1:50, labels=rep(c("0-2yr, 1-5","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),las=2,cex.axis=1.5)
+% mtext(side=1,line=10,text="Age and NZDep",cex=2)
+% legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
+%        col=c(1:4,6),pch=16,bty="n",cex=1.8)
+% par(mar = c(5,4,4,2) + 0.1)
+% @
+% \end{center}
+% \caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (\autoref{table:percap} for details)}
+% \label{fig:percap}
+% \end{figure}
 
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-## plot this figure?
-#hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
-tp$Ethnicity<- relevel(tp$Ethnicity, "European")
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
-
-model2<-update(model1,~.-Age:NZDep)
-summary(model2)
-anova(model2,test="F")
-@
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-tpsub<-tp[!(tp$Ethnicity=="MLA"),]
-
-#hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
-
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F")
-
-model2<-update(model1,~.-Ethnicity:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-model3<-update(model2,~.-Age:NZDep)
-summary(model3)
-anova(model3,test="F")
-@
-
 \subsection{Regression analyses results}
@@
 <<fig=TRUE,echo=FALSE,width =11, height=11>>=
-res<-predict(model3)
 par(mar=c(5,6,4,2)+0.1)
@@
 \end{figure}
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
 
@@
 <<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
-library(xtable)
 resan<-xtable(anova(model3,test="F"))
@@
   <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
-  library(xtable)
   xtb<-xtable(summary(model3))
@@
 <<fig=TRUE,echo=FALSE,width = 11>>=
-AgeVac<-table(testv$VC,testv$AgeInYears)
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
 par(mar=c(5,6,4,2)+0.1)
@@
 
-%%%%
-<<echo=FALSE,include=FALSE,results=hide>>=
-library(maptools)
-
-# read in data from NIR
-data <- read.table("NIR_MMR_data_20060101-20140731/NIR_MMR_data_20060101-20140731.txt", header=T, sep="|")
-
-# read in domicile <-> AU map
-data_dom <- read.csv("NIR_MMR_data_20060101-20140731/domicile_code.csv")
-
-# merge NIR with domicile/AU so we have AU for each NIR
-data_merge <- merge(data, data_dom, by.x = "DOMICILE_CODE", by.y="dom", all=T)
-
-# TODO: figure out how many kids are immunised in each AU
-# find number of kids by BIRTH_CALENDAR_YEAR, AU, VACCINE_DOSE
-
-num_kids <- table(data_merge$BIRTH_CALENDAR_YEAR, data_merge$area.unit, data_merge$VACCINE_DOSE)
-# drop the 0's from vaccine_dose
-dose1 <- t(num_kids[,,2])
-dose2 <- t(num_kids[,,3])
-dose1 <- as.data.frame(dose1[,-ncol(dose1)])
-dose2 <- as.data.frame(dose2[,-ncol(dose2)])
-dose1$cau <- rownames(dose1)
-dose2$cau <- rownames(dose2)
-
-# read in shapefile for AU's
-au <- readShapeSpatial("AU2013_GV_Full")
-
-au_data <- slot(au, "data")
-
-# read in denominator data for AU's
-data_denom<-read.csv("2013-mb-dataset-Total-New-Zealand-individual-part-1_AU.csv",header=T)
-denom<-data_denom[,1]
-for (i in 1:5)
-  denom <- cbind(denom, data_denom[,3]/5)
-for (i in 1:5)
-  denom <- cbind(denom, data_denom[,4]/5)
-colnames(denom)[2:11] <- 2013:(2013-9)
-denom <- denom[,c(1,9:2)]
-
-dose2_merge <- merge(dose2, denom, by.x="cau", by.y="denom", all=T)
-diff <- dose2_merge[,2:8+7] - dose2_merge[,2:8]
-image(as.matrix(diff > 0))
-
-dose2_num <- data.frame(cau=dose2_merge$cau, vacc = rowSums(dose2_merge[,2:5]), total = rowSums(dose2_merge[,10:13]))
-diff <- dose2_num$total - dose2_num$vacc
-
-dose1_merge <- merge(dose1, denom, by.x="cau", by.y="denom", all=T)
-diff <- dose1_merge[,2:8] - dose1_merge[,2:8+7]
-image(as.matrix(diff > 0))
-
-au_new <- merge(au_data, dose2_num, by.x="AU2013", by.y="cau", all.x=T)
-au_new$prop <- au_new$vacc / au_new$total
-library(RColorBrewer)
-library(classInt)
-
-fixedBreaks <- c(0,0.5,0.7, 0.9, 1.1, 1.3, 1.5, Inf)
-au_new$prop[!is.finite(au_new$prop)] <- NA
-breaks <- classIntervals(au_new$prop, n=7, style="fixed", fixedBreaks=fixedBreaks)
-pal <- brewer.pal("Spectral", n=7)
-
-cols <- findColours(breaks, pal)
-plot(au, col=cols)
-
-## plot to show Ministry the data issues...
-fixedBreaks <- c(0,0.6,0.8,1, 1.2,1.4,Inf)
-au_new$prop[!is.finite(au_new$prop)] <- NA
-breaks <- classIntervals(au_new$prop, n=6, style="fixed", fixedBreaks=fixedBreaks)
-pal <- brewer.pal("RdGy", n=6)
-pal<-c("red","orange","grey","blue","green","yellow")
-cols <- findColours(breaks, pal)
-@
-
 \begin{figure}
@@
 par(mfrow=c(1,1))
-pdf(paste("nir_census.pdf"), width=7, height=5)
+# pdf(paste("nir_census.pdf"), width=7, height=5)
 plot(au, col=cols)
@@
 #dev.off()
-par(mfrow=c(1,1))
-
-hist(au_new$prop, breaks=100,col="grey",xlab="Proportion vaccination",main="All domiciles")
-abline(v=1,col="red")
 @
@@
 \label{fig:naive_au}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE,width = 11>>=
+par(mfrow=c(1,1))
+par(cex.axis=1)
+hist(au_new$prop, breaks=100,col="grey",xlab="Proportion vaccination",main="")
+abline(v=1,col="red")
+@
+\end{center}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator}
+\label{fig:naive_au_prop}
 \end{figure}
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-@
-
 \begin{figure}
@@
 
-%%%%%%%%%%%%%%%%%%
-<<echo=FALSE>>=
-library(xtable)
-totdpdf<-read.csv("mapdata_tot.csv",header=T)
-# output top countries
-colnames(totdpdf)[1]<-c("country")
-toptot <- totdpdf[order(-totdpdf$immigration),] 
-immigration<-toptot[1:30,c(1,4)]
-topincidence <- totdpdf[order(-totdpdf$incidence),] 
-incidence<-topincidence[1:30,c(1,2)]
-incidence[,2]<-round(incidence[,2],0)
-topvaccine <- totdpdf[order(totdpdf$cover),] 
-vaccinecover<-topvaccine[1:30,c(1,3)]
-toprisk <- totdpdf[order(-totdpdf$risk),] 
-risk<-(toprisk[1:30,c(1,5)])
-risk[,2]<-round(risk[,2],0)
-toprisk2<-(toprisk[1:10,c(1,2,3,4)])
-toprisk2[,2]<-round(toprisk2[,2],0)
-@
-
-%%%%%%%%%%%%%%%%%%%%%
-
-<<echo=FALSE>>=
-library(xtable)
-nzdpdf<-read.csv("mapdata_nzers.csv",header=T)
-# output top countries
-colnames(nzdpdf)[1]<-c("country")
-topnz <- nzdpdf[order(-nzdpdf$immigration),] 
-nz<-topnz[1:30,c(1,4)]
-topnzrisk <- nzdpdf[order(-nzdpdf$risk),] 
-nzrisk<-(topnzrisk[1:30,c(1,5)])
-nzrisk[,2]<-round(nzrisk[,2],0)
-@
-
-%%%%%%%%%%%%%%%%%%%%%
-<<echo=FALSE>>=
-library(xtable)
-imdpdf<-read.csv("mapdata_immigration.csv",header=T)
-# output top countries
-colnames(imdpdf)[1]<-c("country")
-topim <- imdpdf[order(-imdpdf$immigration),] 
-im<-topim[1:30,c(1,4)]
-topimrisk <- imdpdf[order(-imdpdf$risk),] 
-imrisk<-(topimrisk[1:30,c(1,5)])
-imrisk[,2]<-round(imrisk[,2],0)
-@
-
 \begin{table}
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(incidence, file="", table.env=FALSE,rowname=NULL)
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(vaccinecover, file="", table.env=FALSE,rowname=NULL)
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(im, file="", table.env=FALSE,rowname=NULL)
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(nz, file="", table.env=FALSE,rowname=NULL)
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(immigration, file="", table.env=FALSE,rowname=NULL)
@@
 \end{table}
-
-%%%%%%%%%%%%%%%%%%%%%
 
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(risk, file="", table.env=FALSE,rowname=NULL)
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(nzrisk, file="", table.env=FALSE,rowname=NULL)
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(imrisk, file="", table.env=FALSE,rowname=NULL)
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(toprisk2, file="", table.env=FALSE,rowname=NULL)
@@
 
-<<echo=FALSE>>=
-# haven't had time to sort out which packages I need
-library(maps)       # Provides functions that let us plot the maps
-library(mapdata)    # Contains the hi-resolution points that mark out the countries.
-library(rworldmap)
-library(adehabitat)
-library(rgdal)
-library(raster)
-library(maps)
-library(maptools)
-library(sp)
-#library(SDMTools)
-library(ggplot2)
-require(rgeos)
-library(ggmap)
-library(mapdata)
-#gpclibPermit()
-library(mapproj)
-library(plyr)
-library(grid)
-library(gridExtra)
-#gpclibPermit()
-library(utils)
-library(rgdal)
-#getwd()
-#setwd("~/data")
-#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
-load("world.ggmap.Rda")
-n <- length(unique(world.ggmap$id))
-id = unique(world.ggmap$id)
-id<-as.data.frame(id)
-dpdf<-read.csv("mapdata_tot.csv",header=T)
-# output top countries
-
-np1<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np2<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = incidence), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np3<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = cover), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
-
-#grid.arrange(np1, np2, np3, ncol=3, main="")
-
-np4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-# output top countries
-dpdf<-read.csv("mapdata_immigration.csv",header=T)
-imnp1<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-imnp4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-dpdf<-read.csv("mapdata_nzers.csv",header=T)
-nznp1<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-nznp4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-
-@
-
 \begin{figure}
@@
 \end{figure}
-%%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
+
 \begin{figure}
@@
 <<fig=TRUE,echo=FALSE>>=
-change<-read.csv("change.csv",header=T)
 plot(change[18:25,1],change[18:25,4],type="l",xlab="Year",ylab="",ylim=c(0,max(change[,4])),
@@
 Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than DHB) immunisation coverage data are not able to provide reliable results, as discussed above.
-Frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census.
+Frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census (\autoref{fig:}).
 Given these data gaps we are unable to provide finer scale risk maps.
@@
 The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a}.
-Later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintain high coverage rates in order to prevent future epidemics \citep{roberts4}.
+Later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintain high coverage rates in order to prevent future epidemics and poor schedules were not considered \citep{roberts4}.
 
@@
 While analyses suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics, in the Netherlands analyses showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection and effective immunisation difficult to evaluate \citep{glass4}.
+No such models exist for New Zealand currently.
 
@@
 
+% \begin{figure}[H]
+%      \centering
+%      \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
+%      \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014}
+%      \label{fig:outbreaks}
+% \end{figure}
+% 
+% \begin{figure}[H]
+%      \centering
+%      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
+%      \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in \autoref{fig:outbreaks}. Sporadic cases are excluded.}
+%      \label{fig:r0}
+% \end{figure}
+
 \begin{figure}[H]
      \centering
-     \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
+     \includegraphics[width=0.9\textwidth]{outbreaks_2009_2014.pdf}
      \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014}
@@
      \centering
-     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
+     \includegraphics[width=0.9\textwidth]{averageR0_all_ob.pdf}
      \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in \autoref{fig:outbreaks}. Sporadic cases are excluded.}
@@
 The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}.
-The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the 2013--2014 outbreak 1.82--2.13.
-The probability density of the $R_v$ estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak (as of 12 June 2014), which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
-An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
-
-These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
-The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given its $R_v$ is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproduction numbers may be lower than estimated.
-However, sub-clinical and underreporting may lower the estimate.
+our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014 but for which data were available until 12 June, a mean estimate of 3.66 (range 3.4--3.92).
+The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one.
+Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect this outbreak to persist if conditions remain the same.
+An important caveat to this outbreak analysis is that because this 2013--2014 outbreak that ended in August was ongoing when the data were collected, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
+However, this is not the only case for which the $R_v$ is greater than one.
+
+These analyses also imply that the regular importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to increase immunisation to high enough levels that measles does not become endemic.
+The measles outbreaks in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+This implies that outbreaks may persist within the population for a substantial period, even if $R_v$ values are not much greater than one.
+A caveat to this and other $R_v$ estimates is that the outbreaks may include some sporadic cases misclassified, but also exclude some sporadic cases as yet not associated with another outbreak.
+Thus the true basic reproduction numbers may differ to those estimated.
+However, undetected cases may lower the estimate.
 The relative contributions of both to our estimates are currently unknown.
-
-If fewer cases occur during each outbreak, then this suggests that $R_v$ is lower, but that measles is introduced to New Zealand more frequently.
-The regularity of these outbreaks, however, implies that the approximately yearly importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (\autoref{sub:cost_benefit}).
 
@@
 The estimated numbers to vaccinate in the Auckland area, however, are higher for each DHB, with those estimated for Waitemata, Auckland, and Counties Manukau DHBs all over 17,000  (\autoref{table:attack}).
+Note that there is also evidence for vaccine failure (\autoref{fig:vaccstat}) and pockets of susceptible populations \citep{glass4} may require greater vaccination effort than these estimates and this is considered further in the \autoref{sec:cb}.
 
 The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (\autoref{fig:r0}).
-However, the na\"{\i}ve population (\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for the 2013--2014 outbreak (\autoref{fig:r0}) suggests that catch up vaccination may be necessary (\autoref{table:attack}).
+However, the na\"{\i}ve population (\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for some of the 2009--2014 outbreaks (\autoref{fig:r0}) suggests that catch up vaccination is necessary (\autoref{table:attack}).
 The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics.
@@
 \hline
-DHB  & Size  & Na\"{\i}ve    &   Attack	& Vacc		\\
+DHB  & Size  & Na\"{\i}ve    &   Attack	& Vacc	& Vacc (\%)	\\
 \hline
-Auckland	&	436350	&	52010	&	31159	&	17920	\\
-Bay of Plenty	&	206000	&	20679	&	8437	&	4585	\\
-Canterbury	&	482180	&	51357	&	24695	&	13687	\\
-Capital and Coast	&	283700	&	32625	&	18403	&	10461	\\
-Counties Manukau	&	469300	&	55544	&	32903	&	18880	\\
-Hawke's Bay	&	151700	&	15602	&	6846	&	3751	\\
-Hutt Valley	&	138380	&	15198	&	7836	&	4388	\\
-Lakes	&	98196	&	10558	&	5192	&	2886	\\
-MidCentral	&	162560	&	17328	&	8348	&	4628	\\
-Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	\\
-Northland	&	151690	&	14921	&	5688	&	3071	\\
-South Canterbury	&	55620	&	5238	&	1678	&	893	\\
-Southern	&	297420	&	31607	&	15115	&	8371	\\
-Tairawhiti &	43650	&	4769	&	2431	&	1359	\\
-Taranaki	&	109750	&	11473	&	5262	&	2899	\\
-Waikato	&	359310	&	39402	&	20248	&	11331	\\
-Wairarapa	&	41112	&	3932	&	1346	&	720	\\
-Waitemata	&	525550	&	58350	&	30774	&	17291	\\
-West Coast	&	32151	&	3197	&	1265	&	685	\\
-Whanganui	&	60120	&	6075	&	2530	&	1378	\\
+Auckland	&	436350	&	52010	&	31159	&	17920	& 34\\
+Bay of Plenty	&	206000	&	20679	&	8437	&	4585	& 22\\
+Canterbury	&	482180	&	51357	&	24695	&	13687	& 27\\
+Capital and Coast	&	283700	&	32625	&	18403	&	10461 & 32	\\
+Counties Manukau	&	469300	&	55544	&	32903	&	18880	& 33\\
+Hawke's Bay	&	151700	&	15602	&	6846	&	3751 &	& 24\\
+Hutt Valley	&	138380	&	15198	&	7836	&	4388	& 29\\
+Lakes	&	98196	&	10558	&	5192	&	2886	& 27\\
+MidCentral	&	162560	&	17328	&	8348	&	4628	& 27\\
+Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	& 18\\
+Northland	&	151690	&	14921	&	5688	&	3071	& 21\\
+South Canterbury	&	55620	&	5238	&	1678	&	893	& 17\\
+Southern	&	297420	&	31607	&	15115	&	8371	& 26\\
+Tairawhiti &	43650	&	4769	&	2431	&	1359	& 28\\
+Taranaki	&	109750	&	11473	&	5262	&	2899	& 25\\
+Waikato	&	359310	&	39402	&	20248	&	11331	& 29\\
+Wairarapa	&	41112	&	3932	&	1346	&	720	& 18\\
+Waitemata	&	525550	&	58350	&	30774	&	17291	& 30\\
+West Coast	&	32151	&	3197	&	1265	&	685	& 21\\
+Whanganui	&	60120	&	6075	&	2530	&	1378	& 23\\
 \hline			
-TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 \\
+TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 & 28\\
 \hline
@@
 Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
-Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$.			}
+Attack:		Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;	
+Vacc:		Number of the currently na\"{\i}ve to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Vacc (\%): Proportion of the currently na\"{\i}ve population to vaccinate.
+}
 \label{table:attack}
@@
 \section{Cost analyses}
+\label{sec:cb}
 
 In this section we provide a review of the costs of measles from other locations and an analysis of the costs involved with the current 2013--2014 measles outbreak.
+All exchange rates are from the Reserve Bank of New Zealand.\footnote{http://www.rbnz.govt.nz/statistics/tables/b1/}
 
@@
 An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
-In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case.
+In that study, they estimated the health-care provider costs to be approximately \euro 229,000,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$426046} or \euro 373\footnote{Jan 2005 \$ to \euro rate 0.5375: \$694} per case.
 Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised.
-The mean costs of the hospitalised patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623.
-In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing.
-Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case.
+The mean costs of the hospitalised patients was  \euro 1,877,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$3492} including one patient with encephalitis at a cost of \euro 35,623.\footnote{Jan 2005 \$ to \euro rate 0.5375: \$66275} In addition to the health-care provider costs, additional costs of \euro 89,400\footnote{Jan 2005 \$ to \euro rate 0.5375: \$166325} were incurred by the district public health office, the majority (\euro 85,000,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$158140} 95.1\%) for personnel, \euro 2,300\footnote{Jan 2005 \$ to \euro rate 0.5375: \$4279} (2.6\%) for vaccination, and \euro 2,100\footnote{Jan 2005 \$ to \euro rate 0.5375: \$3907} (2.3\%) for serologic testing.
+Therefore the combined direct costs of these 612 cases amounted to \euro 318,400,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$592372} or \euro 520\footnote{Jan 2005 \$ to \euro rate 0.5375: \$967} per case.
 To determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
@@
 The economic impact to public health departments in the US as the result of 16 outbreaks in 2011 has been estimated \citep{ortegasanchez14}.
-The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$2.7 and 5.3 million.
-Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$298 per contact.
+The outbreaks lasted an average of 22 days and resulted in 107 confirmed cases; however, from these 107 cases, they estimated between approximately 8,900 and 17,500 contacts with confirmed cases, requiring between 42,600 and 83,100 personnel hours at a cost of between US\$2.7\footnote{Jan 2011 \$ to US\$ rate 0.7653: \$3.53million} and 5.3\footnote{Jan 2011 \$ to US\$ rate 0.7653: \$6.93million} million.
+Overall, it was estimated that each contact required 4.7 personnel hours at a cost of US\$298\footnote{Jan 2011 \$ to US\$ rate 0.7653: \$389} per contact.
 
 It was estimated that for the one week that the Iowa Department of Public Health (DPH) investigated a case in 2004, 2,525 hours were used to identify contacts, set up vaccination clinics, and institute and enforce quarantine orders for those who refused vaccination \citep{dayan5}.
-In total, it was estimated the direct costs associated with three cases of measles was US\$142,452, or nearly US\$50,000 per case.
+In total, it was estimated the direct costs associated with three cases of measles was US\$142,452,\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$211856} or nearly US\$50,000\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$74360} per case.
 
@@
 Direct cost information was obtained from approximately 100 public health officers and infection-control officials needed to control the outbreak.
-Direct cost for those completing a survey showed the outbreak cost at least \$167,685, 83\% of which (\$139,023) was for wages, salaries and overheads.
-This amounted to a direct cost of \$4,932 per measles case.
+Direct cost for those completing a survey showed the outbreak cost at least US\$167,685,\footnote{Jan 2005 \$ to US\$ rate 0.704: \$238189} 83\% of which (US\$139,023\footnote{Jan 2005 \$ to US\$ rate 0.704: \$197476}) was for wages, salaries and overheads.
+This amounted to a direct cost of US\$4,932\footnote{Jan 2005 \$ to US\$ rate 0.704: \$7006} per measles case.
 These costs did not include either patient care or indirect costs, which would have made the total and per case cost higher.
@@
 In addition, medical costs were incurred due to vaccination, immunoglobulin, testing for measles immunity, hospitalisation, transportation and diagnosis.
-In total, 387 hours were associated with this single case, resulting in a cost of US\$11,881.
-In addition, per-contact costs amounted to US\$264.
-The cost of hospitalisation for the 3-day stay by the index case was US\$931.
-Additional costs were associated with physician visits (US\$294), vaccine and immunoglobulin (US\$1,765), mileage (US\$205) and immunologic screening tests for the parents exposed to measles (US\$240) for a total of US\$23,816.
+In total, 387 hours were associated with this single case, resulting in a cost of US\$11,881.\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$16327} In addition, per-contact costs amounted to US\$264.\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$363} The cost of hospitalisation for the 3-day stay by the index case was US\$931.\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$1279} Additional costs were associated with physician visits (US\$294\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$404}), vaccine and immunoglobulin (US\$1,765\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$2425}), mileage (US\$205\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$282}) and immunologic screening tests for the parents exposed to measles (US\$240\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$330}) for a total of US\$23,816.\footnote{Jan 2010 \$ to US\$ rate 0.7277: \$32728}
 
@@
 
-The purpose of the current study is to estimate the cost of the current measles outbreak in New Zealand.
+The purpose of the current study is to estimate the cost of the 2013--2014 measles outbreaks in New Zealand.
 Using this information, we will then evaluate the economics of alternative measles control strategies in order to provide additional information to public health officials and decision makers.
@@
 Costs are evaluated as either direct or indirect.
-Direct costs included physician consultations, hospitalisations, drugs, vaccination, long-term care for chronic sequelae, special education costs.
+Direct costs included physician/general practicioner (GP) consultations, hospitalisations, laboratory testing, drugs, vaccination, long-term care for chronic sequelae, special education costs.
 Direct costs can be divided into medical and non-medical \citep{saha13}.
 Direct medical costs include costs for diagnosis, treatment, continuing care, rehabilitation and terminal care.
-Personnel time (investigation and emergency response), materials (phone calls, vaccine), personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
+Personnel time (investigation and emergency response, including primary care), materials (phone calls, vaccine), personnel (cost, wages and fringe benefits), overhead costs, public information, and mileage are estimated when calculating direct medical costs.
 Direct non-medical costs include transportation to and from health care providers.
@@
 Indirect costs are productivity losses for the case and/or health care provider, e.g.
-parent of a school child.
+parent of a school child, as well as those relating to case-contacts.
 Indirect costs included work loss for cases and caregivers.
@@
 
-Cost of the Auckland Regional Public Health Service (ARPHS) for measles response were obtained from the Ministry of Health.
+To determine the cost per case for the direct medical costs, cost of the Auckland Regional Public Health Service (ARPHS) for measles response were obtained from the Ministry of Health.
 Data, for the period January 1 - March 9, 2014, reported salaries for people involved with the measles outbreak management medical team.
@@
 In addition, measles operations personnel are calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
+Senior medical officer (SMO) costs for GP visits were estimated as approximately \$20 per consultation,\footnote{personal communication} though we allow flexibility in this value for the benefit--cost analyses later.
+These costs are applied on a per case basis below.
 
 Wages lost due to measles are calculated for the period January 2008 - August 2014.
+We assume that the losses to the economy are the same, whether the the employee is able to obtain sick leave pay, as this is the general loss to the economy as work days are lost.
 Calculations are based on the assumptions that 5 days of work are lost for each case; however, individuals between 0-14 years of age are not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15-19 years of age was 41.9\%  (Statistics New Zealand, 2013).
 If the case is < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
-Total wage lost for the 247 cases and care givers was estimated to be \$207,155.
-This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases <15 years of age.
-Overall, the cost per case from 2008 - 2014 was estimated to be \$2,513 (\$839 in forgone wages and \$1,710 in hospital costs).
 
@@
 Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  \autoref{table:direct}.
-The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in \autoref{table:hosp}.
-
-<<echo=FALSE>>=
-#setwd("C:\Users\dtshayma\Documents\GitHub\measles")
-costtable1<-read.csv("costtable1.csv",header=T)
-@
+The reported direct medical costs do not include hospital medical costs, which are reported separately in \autoref{table:hosp}.
 
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(costtable1, file="", table.env=FALSE,rowname=NULL)
@@
 
-The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (\autoref{table:hosp}).
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital from 2001--2014 (part year) was \$550,024 (\autoref{table:hosp}).
 The mean cost per case was \$1,877.
 The mean cost per day of stay in the hospital was \$1,170.
-
-<<echo=FALSE>>=
-costtable2<-read.csv("costtable2v2.csv",header=T)
-@
 
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 colnames(costtable2) <- c("Year", "Cases", "Days", "Cost","Cost per case","Cost per day")
@@
 
-<<echo=FALSE>>=
-costtable3<-read.csv("costtable3.csv",header=T)
-@
-
 \begin{table}
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(costtable3, file="", table.env=FALSE,rowname=NULL)
@@
 The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (\autoref{fig:hosp}).
-
-<<echo=FALSE>>=
-hosp<-read.csv("hospital.csv",header=T)
-library(ggplot2)
-p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")+theme(axis.title.x = element_text(size=25),axis.title.y = element_text(size=25),axis.text.x  = element_text(size=16),axis.text.y  = element_text(size=16))
-@
 
@@
 
-<<echo=FALSE>>=
-costtable4<-read.csv("costtable4v2.csv",header=T)
-@
-
 \begin{table}\footnotesize
@@
 Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
+
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 colnames(costtable4) <- c("Year", "Gender", "Cost","Cases","Length of stay","Cost per case")
@@
 
-<<echo=FALSE>>=
-costtable5<-read.csv("costtable5v2.csv",header=T)
-@
-
 \begin{table}
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(costtable5, file="", table.env=FALSE,rowname=NULL)
@@
 
-Wages lost due to measles are calculated for the period January 2008 -- August 2014.
-Calculations are based on the assumption that 5 days of work are lost for each case; however, individuals under 15 years of age are not assumed to be employed and therefore did not suffer an income loss.
-If the case are less than 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
-Total wage lost for the 247 cases and care givers was estimated to be \$210,436.
+Wages lost for the period January 2008 -- August 2014 for the 247 cases and care givers was estimated to be \$210,436.
 This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
 Overall, the cost per case from 2008--2014 was estimated to be \$4,327 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
+\\
+OR
+\\
+Total wage lost for the 247 cases and care givers was estimated to be \$207,155.
+This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases <15 years of age.
+Overall, the cost per case from 2008 - 2014 was estimated to be \$2,513 (\$839 in forgone wages and \$1,710 in hospital costs).
 
@@
 The costs of the catch up vaccination schemes are estimated to be two different values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
-Values of \$20 and \$50 per vaccine are used.
+Values of \$20 and \$50 per vaccine are used, based on US literature, however, this assumes all costs are included with the vaccine delivery.
+Therefore, we also estimate approximate cost a vaccine (including the delivery costs) would need to be before the vaccination costs exceed the benefits.
 Thus the expected measles-related costs due to constant introduction of measles despite increased population immunity and the costs of the vaccination schemes are used to estimate the costs of additional vaccine programs for the B/C analysis.
@@
 Measles introductions are expected to be annually occurring events, though the costs per year for each event was also discounted.
-Thus, the B/C ration was:
+Thus, the B/C ratio was:
 
@@
 The estimated vaccinaton rates, with percentages are shown in \autoref{table:vacp}.
-<<echo=FALSE>>=
-vacp<-read.csv("vacc_predictions.csv",header=T)
-@
 
@@
 Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
-Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}
+Attack:	  Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;
+Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed the small numbers of large outbreaks.
+The distribution of these outbreaks sizes can be seen in \autoref{fig:sim1} and \autoref{fig:sim}}
 
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 latex(vacp, file="", table.env=FALSE,rowname=NULL)
@@
 The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using \autoref{eq:prop} are shown in \autoref{fig:numvac}.
-
-<<echo=FALSE,include=FALSE,results=hide>>=
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-#naive[is.na(naive)] <- 0
-colnames(naive)<-c("AgeInYears","Naive","Cases")
-@
 
@@
 ##
+par(mfrow=c(1,1))
 par(mar=c(5,6,4,2)+0.1)
@@
 sizes<-colSums(res)
-sizes
+#sizes
 legMd<-as.factor(paste("median =",c(median(sizes))))
@@
 #summary(sizes)
-abline(v=median(sizes),col="orange",lty=3,lwd=10)
-abline(v=mean(sizes),col="red",lty=3,lwd=10)
+abline(v=median(sizes),col="orange",lty=3,lwd=8)
+abline(v=mean(sizes),col="red",lty=3,lwd=8)
 #length(which(sizes>5))/1000
-legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
+legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=8,cex=2)
 @
@@
 par(cex.axis=2)
-hist(sizes,breaks=1000,xlab="Outbreak size",main="",cex.lab=2)
-abline(v=median(sizes),col="orange",lty=3,lwd=10)
-abline(v=mean(sizes),col="red",lty=3,lwd=10)
-legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=10,cex=2)
+hist(sizes,breaks=1000,xlab="Outbreak size",main="",cex.lab=2,ylim=c(0,20))
+abline(v=median(sizes),col="orange",lty=3,lwd=8)
+abline(v=mean(sizes),col="red",lty=3,lwd=8)
+legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=8,cex=2)
 @
@@
 \end{figure}
-
-<<echo=FALSE>>=
-costtable20<-read.csv("cost_benefit_20.csv",header=T)
-@
 
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 colnames(costtable20)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
@@
 \end{sidewaystable}
-
-<<echo=FALSE>>=
-costtable50<-read.csv("cost_benefit_50.csv",header=T)
-@
 
@@
 <<results=tex,echo=FALSE>>=
-# This chunk uses "results=tex"
-library(Hmisc)
 colnames(costtable50)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
@@
 For the cost analyses we used the values from the above cost section (\autoref{sub:cost_benefit}).
-Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (\autoref{table:hosp}).
+Specifically, we used the average cost of a case for the analyses to be \$839 lost in wages, \$1765.49 in case management costs, and \$1877 in hospitalisation costs for those attending hospital, with 17\% of cases predicted to be hospitalised (\autoref{table:hosp}).
 We estimated there would be approximately one introduction of measles per year (\autoref{sub:risk_analyses}).
@@
 Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children.
+However, there are a number of different vaccination schemes that will lead to sufficient coverage and these are given in \autoref{table:cohort_vac}.
+The appropriate cover could be reached by catch up vaccination of all school aged children.
+
+\begin{table}[htdp]
+\begin{center}
+\begin{tabular}{lrrrr}
+\hline
+Age range in years for vaccination & Proportion of the currently na\"{\i}ve population it would cover \\
+\hline
+2--5 & 0.04\\
+5--11 & 0.15\\
+5--16 & 0.27\\
+5--18 & 0.31\\
+11--16 & 0.14\\
+11--18 & 0.18\\
+0--29 & 0.72\\
+0--44 & 0.89\\
+>8 & 0.8 \\ 
+\hline
+\end{tabular}
+\end{center}
+\label{table:cohort_vac}
+\end{table}%
 
@@
 
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively  \citep{carabin2}.
+In other outbreaks, the average cost per measles case was estimated to be US\$254\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$571}, US\$276\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$621}, and US\$307\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$691} for Canada, the Netherlands, and the UK, respectively  \citep{carabin2}.
 These values are substantially lower than our estimates, however, it is noticable that these estimates are also from more than a decade ago.
-The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
-In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak \citep{dayan5}.
-They estimated direct costs per case to be less than US\$500.
-However, the combined direct costs of \euro 520 per case in Germany \citep{siedler6} suggests costs may be higher.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$211990} In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak \citep{dayan5}.
+They estimated direct costs per case to be less than US\$500.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$774} However, the combined direct costs of \euro 520\footnote{Jan 2005 \$ to \euro rate 0.5375: \$967} per case in Germany \citep{siedler6} suggests costs may be higher.
 Thus, our estimates of the additional costs, including indirect costs and wages lost may be appropriate.
 
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively in Minnesota \citep{prouty1}.
-In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
-In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
-The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$38790} and US\$78,448,\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$97974} respectively in Minnesota \citep{prouty1}.
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$221612} per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$221612} The cost of a case of measles was estimated to range from \$71\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$138} (no complications and no hospitalisation) to \$29,556\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$57524} (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$2401874} (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
 Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see \autoref{table:zhou}) \citep{zhou4}.

### 2015-03-20 dhbcensus.R

@@ -43,7 +43,10 @@ for (i in 1:20){
   pdf(paste("dhb", i, ".pdf", sep = ""))
   plot(naivedhb[,i],ylim=c(0,8000),main=colnames(naivedhb)[i],xlab="Age",ylab="Numbers",
        pch=16,cex=1.1)
-  legend("topright",legend=c(c("Total naive", round(sum(naivedhb[,i]))),c("Total population", signif(popdhb[i],5))),bty="n")
+  legend("topright",legend=c(c("Total naive", round(sum(naivedhb[,i]))),
+                             c("Total population", signif(popdhb[i],5)),
+                             c("Percent naive",round(sum(naivedhb[,i])/popdhb[i],3)*100)),
+                             bty="n")
   dev.off()
 }  
 

### 2015-03-20 revised_draft.R

@@ -1,5 +1,27 @@
 rm(list=ls())
-
+library(adehabitat)
+library(classInt)
+library(ggplot2)   # plotting library
+library(grid)
+library(gridExtra)
+library(ggmap)
+library(Hmisc)
+library(maps)       # Provides functions that let us plot the maps
+library(mapdata)    # Contains the hi-resolution points that mark out the countries.
+library(maptools)
+library(mapproj)
+library(plyr)
+library(pscl)
+library(RColorBrewer)
+library(rworldmap)
+library(reshape)
+library(reshape2)
+library(rgeos)
+library(rgdal)
+library(raster)
+library(sp)
+library(utils)
+library(xtable)
 ## get data
 # set wd
 # setwd("~/Massey 2014/DHayman_20140627")
@@ -18,10 +40,7 @@ data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
 
 time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
 data = data , FUN=sum)
-library(reshape)
-library(reshape2)
-library(plyr)
-library(pscl)
+
 test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
 head(test)
 dim(test)
@@ -38,8 +57,6 @@ dev.off()
 caseyr<-aggregate( DiseaseName ~ AgeInYears, 
 data = test , FUN=sum)
 
-require(reshape2)  # this is the library that lets one flatten out data
-require(ggplot2)   # plotting library
 popimmune<-read.csv("PopnImmunityAll.csv",header=T)
 popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
 pop<-read.csv("popnsize.csv",header=T)
@@ -88,7 +105,6 @@ timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+Ethn
 data = datav , FUN=sum,na.rm=F,na.action=na.pass)
 
[…]

### 2015-03-20 draftfinalreport_v2.Rnw
Major revisions to text

@@
 \documentclass{article}
-
+\usepackage{longtable}
 \usepackage{times}
@@
 
-\begin{figure}
+\begin{sidewaysfigure}
      %\centering
@@
      \label{fig:incidence1997}
-\end{figure}
+\end{sidewaysfigure}
 
@@
 <<echo=FALSE,include=FALSE,results=hide>>=
+rm(list=ls())
 library(adehabitat)
@@
 library(xtable)
-#library(SDMTools)
-#gpclibPermit()
-
+## get data
+# set wd
+# setwd("~/Massey 2014/DHayman_20140627")
 # read data
@@
 vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-# names(data)
+names(data)
 data$RptYear<-as.factor(data$RptYear)
@@
 
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-                 data = data , FUN=sum)
-time_yr<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,data = data , FUN=sum)
+
+test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
+head(test)
+dim(test)
 
 caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-                   data = time_yr , FUN=sum)
+data = test , FUN=sum)
 
@@
 pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5), rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
+
 naive<-round(pop-(pop*impop))
@@
 
-timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
-                 data = datav , FUN=sum,na.rm=F,na.action=na.pass)
+timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, data = datav , FUN=sum,na.rm=F,na.action=na.pass)
 
 testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-
 testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
@@
 
-### denominator data
-# setwd("~/Massey_2014/measles/data")
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-names(data)
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-                 data = data , FUN=sum)
-
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-                   data = test , FUN=sum)
-test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
-test$AgeInYears<-as.factor(test$AgeInYears)
-tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
-test$AgeInYears<-tage
-#summary(test$EthnicityPrioritised)
-teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
-                                     "Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
-
-test$EthnicityPrioritised<-teth
-
-test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
-test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
-
-# want to set NZ Deprivation to the appropriate year
-
-testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
-                              ifelse (test$RptYear == "2008",test$NZDep06,
-                              ifelse (test$RptYear == "2009",test$NZDep06,
-                              ifelse (test$RptYear == "2010",test$NZDep06,
-                              ifelse (test$RptYear == "2011",test$NZDep06,
-                              ifelse (test$RptYear == "2012",test$NZDep06,
-                              ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-testt$NZDep<-as.factor(testt$NZDep-1)
-
-testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
-                      ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
-                 data = testt , FUN=sum)
-
-colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
-
-tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
-                                 "5"="1-5","6"="6-10","7"="6-10","8"="6-10",
-                                 "9"="6-10","10"="6-10"));
-
-testtable$NZDep<-tnzd
-
-tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
-                 data = testtable , FUN=sum)
-
-ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
-                 data = testtable , FUN=sum)
-denom<-read.csv("NZDep2006Denominators.csv",header=T)
-denom<-denom[,-c(9)]
-
-denom<-denom[,1:18]
-row.has.na <- apply(denom, 1, function(x){any(is.na(x))})
-
-denomt <- denom[!row.has.na,]
-
-denomt<-denomt[denomt$Sex_code==99,]
-denomt$Eth_Level<-as.factor(denomt$Eth_Level)
-denomt<-denomt[denomt$Eth_Level %in% c("24","21","66",'22','77'),]
-dm<-cbind(denomt[,c(3,6,9:18)])
-
-dm<-dm[dm$Age_Label %in% c("A: 0- 5 yrs", "A: 6-17 yrs","A:18-24 yrs","A:25-64 yrs","A:65+ yrs"),]
-dage<-revalue(dm$Age_Label, c("A: 0- 5 yrs"="<6", "A: 6-17 yrs"="6-17","A:18-24 yrs"="18-24","A:25-64 yrs"="25+","A:65+ yrs"="25+"));
-dm$Age_Label<-dage
-colnames(dm)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-
-datadm<-dm[dm$Age=="<6",]
-dataA<-(rbind(datadm,datadm)/2)
-dataA<-round(dataA[,3:12])
-dataA<-cbind(c(rep("0-2",5),rep("3-5",5)),rep(datadm$Ethnicity[1:5],2),dataA)
-colnames(dataA)<-c('Age','Ethnicity',"Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10")
-ddm<-dm[!dm$Age=="<6",]
-dm<-rbind(dataA,ddm)
-dm$Ethnicity <- factor(dm$Ethnicity)
-deth<-revalue(dm$Ethnicity, c("Asian (Prioritised)"="Asian","European (NZ European and Other European)"="European",
-                              "Maori (Prioritised)"="Maori","MELAA"="MLA","Pacific People (Prioritised)"="Pacific"));
-dm$Ethnicity<-deth
-head(dm)
-
-popn<-melt(dm,id.vars=c("Age","Ethnicity"),measure.vars=c("Dep1","Dep2","Dep3","Dep4",'Dep5','Dep6','Dep7','Dep8','Dep9',"Dep10"))
-colnames(popn)<-c("Age","Ethnicity","NZDep","Popn")
-pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
-                                 "Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
-                                 "Dep9"="6-10","Dep10"="6-10"));
-
-popn$NZDep<-pnzd
-tp<-aggregate( cbind(Popn) ~ NZDep+Age+Ethnicity, 
-               data = popn , FUN=sum)
-
-tp$merge <- paste(tp$NZDep, tp$Age, tp$Ethnicity)
-ttyr$merge <- paste(ttyr$NZDep, ttyr$Age, ttyr$Ethnicity)
-ttyr <- ttyr[ttyr$Ethnicity!="None",]
-
-tp$cases <- 0
-cases <- matrix(0, length(tp$merge),1)
-rownames(cases) <- tp$merge
-cases[ttyr$merge,] <- ttyr$Cases
-tp$cases <- cases
-PerCapita<-(tp$cases/tp$Popn)
-PerCap<-cbind(tp[,c(1:4,6)],round(PerCapita*10000,4))
-colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
-
-## get data
-# set wd
-# setwd("~/Massey 2014/DHayman_20140627")
+## regression analyses - measles
 # read data
@@
 
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-data = data , FUN=sum)
+time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear, data = data , FUN=sum)
 
@@
 
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-
-naive<-round(pop-(pop*impop))
-
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-colnames(naive)<-c("AgeInYears","Naive","Cases")
-
-dose1=as.factor(round(vac$Dose1Mths/12,0));
-dose1<-summary(dose1)
-dose1<-as.data.frame(dose1)
-dose1$AgeInYears<-rownames(dose1)
-dose2=as.factor(round(na.omit(vac$Dose2Mths/12,0)));
-dose2<-summary(dose2)
-dose2<-as.data.frame(dose2)
-dose2$AgeInYears<-rownames(dose2)
-
-## need to merge vaccination data with data
-## to get years the cases were from
-datav<-merge(vac,data,by="CaseCode",all=T)
-##
-datav$Dose1Mths[which(is.na(datav$Dose1Mths))] <- 9999
-datav$Dose2Mths[which(is.na(datav$Dose2Mths))] <- 9999
-#########################
-datav$RptYear<-as.factor(datav$RptYear)
-datav$SurvWeek<-as.factor(datav$SurvWeek)
-datav$NZDep01<-as.factor(datav$NZDep01)
-datav$NZDep06<-as.factor(datav$NZDep06)
-datav$NZDep13<-as.factor(datav$NZDep13)
-datav$Dose1Mths<-as.factor(datav$Dose1Mths)
-datav$Dose2Mths<-as.factor(datav$Dose2Mths)
-
-#data$AgeInYears<-as.factor(data$AgeInYears)
-datav$EthnicityPrioritised<-as.factor(datav$EthnicityPrioritised)
-
-timev<-aggregate( cbind( DiseaseName) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear+ Dose1Mths +Dose2Mths, 
-data = datav , FUN=sum,na.rm=F,na.action=na.pass)
-
-testv<-subset(timev, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-
-testv$Dose1Mths<-revalue(testv$Dose1Mths, c("9999"=NA));
-testv$Dose2Mths<-revalue(testv$Dose2Mths, c("9999"=NA));
-testv$Dose1Mths<-as.numeric(testv$Dose1Mths)
-testv$Dose2Mths<-as.numeric(testv$Dose2Mths)
-testv$D2vac<-ifelse(testv$Dose1Mths > -1 & testv$Dose2Mths >= testv$Dose1Mths,testv$Dose2Mths,NA)
-testv$D1vac<-ifelse(testv$Dose1Mths > -1 & is.na(testv$D2vac) == T,testv$Dose1Mths,NA)
-testv$Unvac<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,testv$AgeInYears,NA)
-
-testv$VC<-ifelse(is.na(testv$Dose1Mths) == T & is.na(testv$Dose2Mths) == T,0,
-ifelse(testv$Dose1Mths >= 0 & is.na(testv$Dose2Mths) == T,1,2))
-AgeVac<-table(testv$VC,testv$AgeInYears)
-
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-AgeInYears<-(as.numeric(rownames(AgeVac)))
-AgeVac<-cbind(AgeVac,AgeInYears)       
-AgeInYears<-(as.numeric(rownames(pop)))
-pop<-cbind(pop,AgeInYears)
-colnames(pop)<-c("Population","AgeInYears")
-AgeV<-merge(pop,AgeVac,by="AgeInYears",all=T)
-
-## regression analyses - measles
-# setwd("~/Massey 2014/DHayman_20140627")
-# read data
+test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
+test$AgeInYears<-as.factor(test$AgeInYears)
+tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
+test$AgeInYears<-tage
+teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA","Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
+
+test$EthnicityPrioritised<-teth
+test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
+test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
+
+# want to set NZ Deprivation to the appropriate year
+
+testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
+ifelse (test$RptYear == "2008",test$NZDep06,
+ifelse (test$RptYear == "2009",test$NZDep06,
+ifelse (test$RptYear == "2010",test$NZDep06,
+ifelse (test$RptYear == "2011",test$NZDep06,
+ifelse (test$RptYear == "2012",test$NZDep06,
+ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
+
+testt$NZDep<-as.factor(testt$NZDep-1)
+
+testtable<-aggregate( cbind( DiseaseName) ~ NZDep+AgeInYears+EthnicityPrioritised + RptYear,  data = testt , FUN=sum)
+dim(testtable)
+colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
+head(testtable)
+
+tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
+                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
+                                 "9"="8-10","10"="8-10"));
+
+testtable$NZDep<-tnzd
+# head(testtable)
+tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
+data = testtable , FUN=sum)
+
+ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
+data = testtable , FUN=sum)
+ttyr <- ttyr[ttyr$Ethnicity=="None",]
+
+# latex(ttyr, file="",table.env=FALSE,rowname=NULL)
+
+### denominator data
 data<-read.csv("DHayman_20140627.csv",header=T)
@@
 test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-
-caseyr<-aggregate( DiseaseName ~ AgeInYears, 
-data = test , FUN=sum)
-
-# test$AgeInYears<-findInterval(test$AgeInYears,c(3,6,18,25))
-test$AgeInYears<-findInterval(test$AgeInYears,c(2,5,18,25))
-test$AgeInYears<-as.factor(test$AgeInYears)
-#tage<-revalue(test$AgeInYears, c("0"="0-2", "1"="3-5","2"="6-17","3"="18-24","4"="25+"));
-tage<-revalue(test$AgeInYears, c("0"="0-1", "1"="2-4","2"="5-17","3"="18-24","4"="25+"));
-test$AgeInYears<-tage
-#summary(test$EthnicityPrioritised)
-teth<-revalue(test$EthnicityPrioritised, c("European or Other"="European", "Middle Eastern/Latin American/African"="MLA",
-"Pacific Peoples"="Pacific","Response cannot be classified"="None","Unknown"="None"));
-
-test$EthnicityPrioritised<-teth
-
-test <- within(test, NZDep06[NZDep06 == "0"]<-NA)
-test <- within(test, NZDep13[NZDep13 == "0"]<-NA)
-
-# want to set NZ Deprivation to the appropriate year
-
-testt <- within(test, NZDep<- ifelse (test$RptYear == "2007",test$NZDep06,
-ifelse (test$RptYear == "2008",test$NZDep06,
-ifelse (test$RptYear == "2009",test$NZDep06,
-ifelse (test$RptYear == "2010",test$NZDep06,
-ifelse (test$RptYear == "2011",test$NZDep06,
-ifelse (test$RptYear == "2012",test$NZDep06,
-ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-
-testt$NZDep<-as.factor(testt$NZDep-1)
-
-testtable<-aggregate( cbind( DiseaseName) #+ as.numeric(SurvWeek))
-~ NZDep+AgeInYears+EthnicityPrioritised + RptYear, 
-data = testt , FUN=sum)
-dim(testtable)
-colnames(testtable)<-c("NZDep","Age","Ethnicity","Year","Cases")
-head(testtable)
-
-tnzd<-revalue(testtable$NZDep, c("1"="1-3","2"="1-3","3"="1-3","4"="4-7",
-                                 "5"="4-7","6"="4-7","7"="4-7","8"="8-10",
-                                 "9"="8-10","10"="8-10"));
-
-testtable$NZDep<-tnzd
-
-tt<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity + Year, 
-data = testtable , FUN=sum)
-
-ttyr<-aggregate( cbind(Cases) ~ NZDep+Age+Ethnicity, 
-data = testtable , FUN=sum)
-ttyr <- ttyr[ttyr$Ethnicity=="None",]
-### denominator data
-
-data<-read.csv("DHayman_20140627.csv",header=T)
-vac<-read.csv("DHayman_20140715_Vacc.csv",header=T)
-
-data$RptYear<-as.factor(data$RptYear)
-data$SurvWeek<-as.factor(data$SurvWeek)
-data$NZDep01<-as.factor(data$NZDep01)
-data$NZDep06<-as.factor(data$NZDep06)
-data$NZDep13<-as.factor(data$NZDep13)
-#data$AgeInYears<-as.factor(data$AgeInYears)
-data$EthnicityPrioritised<-as.factor(data$EthnicityPrioritised)
-
-time<-aggregate( cbind( DiseaseName ) ~ NZDep01 +NZDep06+NZDep13+AgeInYears+EthnicityPrioritised + SurvWeek + RptYear,# + Dose1Mths + Dose2Mths, 
-data = data , FUN=sum)
-
-test<-subset(time, (RptYear %in% c("2007","2008","2009","2010","2011","2012","2013","2014")))
-
+head(test)
+dim(test)
 caseyr<-aggregate( DiseaseName ~ AgeInYears, 
@@
 ifelse (test$RptYear == "2013",test$NZDep13,NZDep13))))))))
-
+head(testt)
+tail(testt)
+str(testt)
 testt$NZDep<-as.factor(testt$NZDep-1)
@@
 head(testtable)
+
+#tnzd<-revalue(testtable$NZDep, c("1"="1-5","2"="1-5","3"="1-5","4"="1-5",
+#"5"="1-5","6"="6-10","7"="6-10","8"="6-10",
+#"9"="6-10","10"="6-10"));
 
@@
 popn
+#pnzd<-revalue(popn$NZDep, c("Dep1"="1-5","Dep2"="1-5","Dep3"="1-5","Dep4"="1-5",
+#"Dep5"="1-5","Dep6"="6-10","Dep7"="6-10","Dep8"="6-10",
+#"Dep9"="6-10","Dep10"="6-10"));
 
@@
 tp$perCap<-with(tp,tp$cases/tp$Popn*10000)
-
 tpsub<-tp[!(tp$Ethnicity=="MLA"),]
@@
 
+#hist(tp$cases,xlab="Cases",main='Histogram of cases per category',col='grey',breaks=20)
 tp$Ethnicity<- relevel(tp$Ethnicity, "European")
@@
 ## use tpsub
+#hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
 
@@
 anova(model3,test="F")
+anovap<-anova(model3,test="F")
+
+res<-predict(model3)
+#res<-predict(model2)
+
+cor(exp(res),tpsub$cases)
+cor.test(exp(res),tpsub$cases)
+
+resan<-xtable(anova(model3,test="F"))
+#resan<-xtable(anova(model2,test="F"))
+# print(resan,floating = FALSE)
+
+#xtb<-xtable(summary(model3))
+xtb<-xtable(summary(model2))
+# print(xtb,floating = FALSE)
 
@@
 DOB<-as.numeric(levels(testv$RptYear))[testv$RptYear]-testv$AgeInYears
- hist(DOB,breaks=100,col="grey",main="2007- 2014 measles cases",xlab="Date of Birth")
+#hist(DOB,breaks=100,col="grey",main="2007- 2014 measles cases",xlab="Date of Birth")
+
 DOBVac<-table(testv$VC,DOB)
@@
 DOBVac<-t(DOBVac)
-
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
 
@@
 ##
+###
 
@@
 
+# write.table(table11,"vaccine_proportions_11yrOlds.txt",sep=",",row.names=F)
+# write.table(table120,"vaccine_proportions_1_20yrOlds.txt",sep=",",row.names=F)
+# write.table(table201,"vaccine_proportions_20_1yrOlds.txt",sep=",",row.names=F)
+
 ###
-
-tp$Ethnicity<- relevel(tp$Ethnicity, "European")
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tp,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F") # F test, not Chisq, because dispersion estimated by moments
-
-model2<-update(model1,~.-Age:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-tpsub<-tp[!(tp$Ethnicity=="MLA"),]
-
-model<-glm(cases~Age*Ethnicity*NZDep+offset(log(Popn)),data=tpsub,family="quasipoisson")
-summary(model)
-
-model1<-update(model,~.-Age:Ethnicity:NZDep)
-summary(model1)
-anova(model1,test="F")
-
-model2<-update(model1,~.-Ethnicity:NZDep)
-summary(model2)
-anova(model2,test="F")
-
-model3<-update(model2,~.-Age:NZDep)
-summary(model3)
-anova(model3,test="F")
-
-res<-predict(model3)
-cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-
-AgeVac<-table(testv$VC,testv$AgeInYears)
-row.names(AgeVac)<-c("Unvaccinated","Dose1","Dose2")
-AgeVac<-t(AgeVac)
-
-# read in data from NIR
-data_d <- read.table("NIR_MMR_data_20060101-20140731/NIR_MMR_data_20060101-20140731.txt", header=T, sep="|")
-
-# read in domicile <-> AU map
-data_dom <- read.csv("NIR_MMR_data_20060101-20140731/domicile_code.csv")
-
-# merge NIR with domicile/AU so we have AU for each NIR
-data_merge <- merge(data_d, data_dom, by.x = "DOMICILE_CODE", by.y="dom", all=T)
-
-# TODO: figure out how many kids are immunised in each AU
-# find number of kids by BIRTH_CALENDAR_YEAR, AU, VACCINE_DOSE
-
-num_kids <- table(data_merge$BIRTH_CALENDAR_YEAR, data_merge$area.unit, data_merge$VACCINE_DOSE)
-# drop the 0's from vaccine_dose
-dose1 <- t(num_kids[,,2])
-dose2 <- t(num_kids[,,3])
-dose1 <- as.data.frame(dose1[,-ncol(dose1)])
-dose2 <- as.data.frame(dose2[,-ncol(dose2)])
-dose1$cau <- rownames(dose1)
-dose2$cau <- rownames(dose2)
-
-# read in shapefile for AU's
-au <- readShapeSpatial("AU2013_GV_Full")
-
-au_data <- slot(au, "data")
-
-# read in denominator data for AU's
-data_denom<-read.csv("2013-mb-dataset-Total-New-Zealand-individual-part-1_AU.csv",header=T)
-denom<-data_denom[,1]
-for (i in 1:5)
-  denom <- cbind(denom, data_denom[,3]/5)
-for (i in 1:5)
-  denom <- cbind(denom, data_denom[,4]/5)
-colnames(denom)[2:11] <- 2013:(2013-9)
-denom <- denom[,c(1,9:2)]
-
-dose2_merge <- merge(dose2, denom, by.x="cau", by.y="denom", all=T)
-diff <- dose2_merge[,2:8+7] - dose2_merge[,2:8]
-image(as.matrix(diff > 0))
-
-dose2_num <- data.frame(cau=dose2_merge$cau, vacc = rowSums(dose2_merge[,2:5]), total = rowSums(dose2_merge[,10:13]))
-diff <- dose2_num$total - dose2_num$vacc
-
-dose1_merge <- merge(dose1, denom, by.x="cau", by.y="denom", all=T)
-diff <- dose1_merge[,2:8] - dose1_merge[,2:8+7]
-image(as.matrix(diff > 0))
-
-au_new <- merge(au_data, dose2_num, by.x="AU2013", by.y="cau", all.x=T)
-au_new$prop <- au_new$vacc / au_new$total
-
-fixedBreaks <- c(0,0.5,0.7, 0.9, 1.1, 1.3, 1.5, Inf)
-au_new$prop[!is.finite(au_new$prop)] <- NA
-breaks <- classIntervals(au_new$prop, n=7, style="fixed", fixedBreaks=fixedBreaks)
-pal <- brewer.pal("Spectral", n=7)
-
-cols <- findColours(breaks, pal)
-plot(au, col=cols)
-
-## plot to show Ministry the data issues...
-fixedBreaks <- c(0,0.6,0.8,1, 1.2,1.4,Inf)
-au_new$prop[!is.finite(au_new$prop)] <- NA
-breaks <- classIntervals(au_new$prop, n=6, style="fixed", fixedBreaks=fixedBreaks)
-pal <- brewer.pal("RdGy", n=6)
-pal<-c("red","orange","grey","blue","green","yellow")
-cols <- findColours(breaks, pal)
-
 popimmune<-read.csv("PopnImmunityAll.csv",header=T)
@@
 
-#getwd()
-#setwd("~/data")
-#unzip("TM_WORLD_BORDERS_SIMPL-0.3.zip")
-load("world.ggmap.Rda")
-n <- length(unique(world.ggmap$id))
-id = unique(world.ggmap$id)
-id<-as.data.frame(id)
-dpdf<-read.csv("mapdata_tot.csv",header=T)
-# output top countries
-
-np1<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np2<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = incidence), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-np3<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = cover), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-  xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "red", high = "blue", guide = "colorbar")
-
-#grid.arrange(np1, np2, np3, ncol=3, main="")
-
-np4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-# output top countries
-dpdf<-read.csv("mapdata_immigration.csv",header=T)
-imnp1<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-imnp4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-dpdf<-read.csv("mapdata_nzers.csv",header=T)
-nznp1<-ggplot(dpdf, aes(map_id = id))  +
-  geom_map(aes(fill = immigration), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-   theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "blue", high = "red", guide = "colorbar")
-
-nznp4<-ggplot(dpdf, aes(map_id = id)) +
-  geom_map(aes(fill = risk), map =world.ggmap) +
-  expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
-   xlab("longitude") +
-  ylab("latitude") +
-  theme(text = element_text(size=30))+
-  scale_fill_gradient("",low = "yellow", high = "red", guide = "colorbar")
-#np4
-
 change<-read.csv("change.csv",header=T)
@@
 costtable3<-read.csv("costtable3.csv",header=T)
-
-hosp<-read.csv("hospital.csv",header=T)
-
-p<-ggplot(data=hosp, aes(x=Days, y=Number)) + geom_bar(stat="identity")+theme(axis.title.x = element_text(size=25),axis.title.y = element_text(size=25),axis.text.x  = element_text(size=16),axis.text.y  = element_text(size=16))
-
 costtable4<-read.csv("costtable4v2.csv",header=T)
 costtable5<-read.csv("costtable5v2.csv",header=T)
-vacp<-read.csv("vacc_predictions.csv",header=T)
-
 costtable20<-read.csv("cost_benefit_20.csv",header=T)
@@
 
-popimmune<-read.csv("PopnImmunityAll.csv",header=T)
-popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
-pop<-read.csv("popnsize.csv",header=T)
-colnames(pop)<-0:100
-pop<-t(pop)
-impop<-c(popimmune$Immunity[1:6],rep(popimmune$Immunity[7],8),rep(popimmune$Immunity[8],5),rep(popimmune$Immunity[9],5),
-         rep(popimmune$Immunity[10],9),rep(popimmune$Immunity[11],20),rep(popimmune$Immunity[12],48))
-length(pop)
-naive<-round(pop-(pop*impop))
-
-## match cases per age
-AgeInYears<-0:100
-naive<-cbind(AgeInYears,naive)
-colnames(naive)<-c("AgeInYears","Naive")
-naive<-merge(naive,caseyr,by="AgeInYears",all=T)
-#naive[is.na(naive)] <- 0
-colnames(naive)<-c("AgeInYears","Naive","Cases")
+vacp<-read.csv("vacc_predictions.csv",header=T)
+hosp<-read.csv("hospital.csv",header=T)
+resan<-xtable(anova(model3,test="F"))
+xtb<-xtable(summary(model3))
+
 @
@@
      \begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mfrow=c(1,1))
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-hist(time_yr$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-@
+     \includegraphics[width=0.9\textwidth]{case_age_dist.pdf}
      \end{center}
@@
 
-\begin{table}[hbtp]
-\footnotesize
+\begin{table}[hbtp]\scriptsize
 <<results=tex,echo=FALSE>>=# This chunk uses "results=tex"
@@
 
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(x=NZDep, y=cases, fill=Age)) + 
-  geom_bar(stat="identity", position="dodge")+
- # theme(text=element_text(size=45))+
-  ylab("cases")
-@
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{percap_ordert.pdf}
 \end{center}
 \caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(x=NZDep, y=cases, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-#  theme(text=element_text(size=45))+
-  ylab("cases")
-@
+\label{fig:percap_all}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{case_ordert.pdf}
 \end{center}
 \caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(x=Age, y=cases, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-#  theme(text=element_text(size=45))+
-  ylab("cases")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(x=NZDep,y= perCap, fill=Age)) + 
-  geom_bar(stat="identity", position="dodge")+
- # theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(x=NZDep, y=perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(x=Age,y= perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(x=NZDep,y= perCap, fill=Age)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(x=NZDep,y= perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(x=Age, y=perCap, fill=Ethnicity)) + 
-  geom_bar(stat="identity", position="dodge")+
-  #theme(text=element_text(size=45))+
-  ylab("Per capita per 10000")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tp, aes(x=merge, y=cases, fill=Ethnicity)) + 
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(x=ordert, y=cases, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("cases")+theme(axis.text.x = element_text(angle = 90, vjust=0,hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub, aes(x=ordert, y=perCap, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("per capita")+theme(axis.text.x = element_text(angle = 90, vjust=0, hjust = 1))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-
-ggplot(tpsub1, aes(x=ordert,y= cases, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("cases")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-ggplot(tpsub1, aes(x=ordert,y= perCap, fill=Ethnicity)) +
-     geom_bar(stat="identity", position="dodge")+
-     # theme(text=element_text(size=45))+
-     ylab("per captita")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-<<echo=FALSE,include=FALSE>>=
-#cor(exp(res),tpsub$cases)
-cor.test(exp(res),tpsub$cases)
-@
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-#hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
- \begin{figure}
- \begin{center}
- <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1)
- barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
- legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
-@
- \end{center}
- \caption{test}
-%\label{fig:test}
- \end{figure}
-
- \begin{figure}
- \begin{center}
- <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1)
- par(mfrow=c(2,1))
- barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
- legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1)
- barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
-legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
- @
- \end{center}
- \caption{test}
-\label{fig:test}
- \end{figure}
-
- \begin{figure}
- \begin{center}
- <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
- par(mar=c(5,6,4,2)+0.1)
- par(cex.axis=1.5)
- par(mfrow=c(1,1))
- barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
- legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
- @
- \end{center}
- \caption{test}
-\label{fig:test}
- \end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
-\end{center}
-\caption{test}
-%\label{fig:test}
-\end{figure}
-
-\begin{table}[hbtp]
-\footnotesize
+\label{fig:cases_all}
+\end{sidewaysfigure}
+
+Two way interactions are shown in \autoref{sec:app} (\autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).
+
+\begin{table}[hbtp]\scriptsize
 \caption{Per capita measles rates}
@@
 
-\begin{table}[hbtp]
-\footnotesize
-\caption{proportions of the naive population the would be covered if vaccinated by age groups}
+\begin{table}[hbtp]\footnotesize
+\caption{Proportions of the naive population the would be covered if vaccinated by age groups}
 <<results=tex,echo=FALSE>>=
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=11>>=
-% par(mar=c(12,6,4,2)+0.1)
-% par(cex.axis=2)
-% plot(tp$cases/tp$Popn*10000,main="",ylab="Per Capita Rate Per 10,000",pch=16,cex=2,col=c(rep(1,10),rep(2,10),rep(3,10),rep(4,10),rep(6,10)),
-%      xlab="",cex.lab=2,xaxt="n")
-% axis(side = 1, at = 1:50, labels=rep(c("0-2yr, 1-5","0-2yr, 6-10","3-5yr, 1-5","3-5yr, 6-10","6-17yr, 1-5","6-17yr, 6-10","18-24yr, 1-5",'18-24yr, 1-5',"25+yr, 1-5","25+yr, 6-10"),5),las=2,cex.axis=1.5)
-% mtext(side=1,line=10,text="Age and NZDep",cex=2)
-% legend("topleft",c("Asian","European","Maori","MLA","Pacific"),
-%        col=c(1:4,6),pch=16,bty="n",cex=1.8)
-% par(mar = c(5,4,4,2) + 0.1)
-% @
-% \end{center}
-% \caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (\autoref{table:percap} for details)}
-% \label{fig:percap}
-% \end{figure}
-
-\begin{table}[hbtp]
-\footnotesize
-\caption{Numbers of measles cases (Cases), population sizes (Ppulation) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
+\begin{table}[hbtp] \scriptsize
+\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
 <<results=tex,echo=FALSE>>=
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
-@
+     \includegraphics[width=0.9\textwidth]{Cases_regmodel.pdf}
 \end{center}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width =11, height=11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-@
+     \includegraphics[width=0.9\textwidth]{Cases_regmodel_prediction.pdf}
 \end{center}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-@
+     \includegraphics[width=0.9\textwidth]{Cases_regmodel_resid.pdf}
 \end{center}
@@
 The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
-\par
+
 \vspace{5mm} %5mm vertical space
 \begin{table}
-<<label=regmodelanova,include=FALSE,echo=FALSE,results=tex>>=
-resan<-xtable(anova(model3,test="F"))
+<<include=FALSE,echo=FALSE,results=tex>>=
 print(resan,floating = FALSE)
@@
 \end{table}
-\vspace{5mm} %5mm vertical space
 
@@
 \begin{table}
-  <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
-  xtb<-xtable(summary(model3))
+<<include=FALSE,echo=FALSE,results=tex>>=
   print(xtb,floating = FALSE)
@@
 
-\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
+Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
 Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also \autoref{fig:ageinyears}).
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
-legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-@
+     \includegraphics[width=0.9\textwidth]{vacc_status.pdf}
 \end{center}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
-abline(v=12,col="blue")
-hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-abline(v=60,col="red")
-legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-@
+     \includegraphics[width=0.9\textwidth]{vacc_age.pdf}
 \end{center}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mfrow=c(1,1))
-# pdf(paste("nir_census.pdf"), width=7, height=5)
-plot(au, col=cols)
-legend("topright",c("0-60%","61-80%","81-100%","101-120%","121-140%",">141%","NA")
-       ,fill=c("red","orange","grey","blue","green","yellow","white"),
-       bty="n",inset=0.1,title="Percent vaccinated")
-## par("usr")
-##
-par(fig = c(0.05, 0.4, 0.5,0.9), mar=c(5,5,2,2), new=TRUE)
-hist(au_new$prop[au_new$prop < 2], breaks=50,col="grey",
-     xlab="Proportion vaccinated",main="")
-abline(v=1,col="red",lwd=3)
-#text(x=0.2,y=150,"Area Unit")
-#dev.off()
-@
+     \includegraphics[width=0.9\textwidth]{naive_au.pdf}
 \end{center}
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mfrow=c(1,1))
-par(cex.axis=1)
-hist(au_new$prop, breaks=100,col="grey",xlab="Proportion vaccination",main="")
-abline(v=1,col="red")
-@
+     \includegraphics[width=0.9\textwidth]{naive_au_prop.pdf}
 \end{center}
@@
 Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (\autoref{fig:naive_map}).
-The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures:
-
+The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures in the \autoref{sec:app}:
 \begin{itemize}[noitemsep,nolistsep]
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(cex.axis=2)
-par(mar=c(5,6,4,2)+0.1)
-plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-points(naive,pch=16)
-legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-@
+     \includegraphics[width=0.9\textwidth]{naive_allPop.pdf}
 \end{center}
@@
      \begin{center}
-     \includegraphics[width=1.2\textwidth]{naive_map.pdf}
+     \includegraphics[width=0.9\textwidth]{naive_map.pdf}
      \end{center}
@@
      \label{fig:naive_map}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Northland}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Waitemata}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Auckland}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Counties_Manukau}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Waikato}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Lakes}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:BayofPlenty}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Tairawhiti}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Taranaki}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:HawkesBay}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Whanganui}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Midcentral}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Hutt}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:CapitalandCoast}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Wairarapa}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:NelsonMarlborough}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:WestCoast}
-\end{figure}
-
-\begin{figure}[H]
-   \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Canterbury}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:SouthCanterbury}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Southern}
 \end{figure}
@@
 
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np2)
-@
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{incidence12.pdf}
 \end{center}
@@
 \label{fig:incidence12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np3)
-@
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{cover12.pdf}
 \end{center}
@@
 \label{fig:cover12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(imnp1)
-@
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{immigration12.pdf}
 \end{center}
@@
 \label{fig:immigration12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(nznp1)
-@
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{nztravel12.pdf}
 \end{center}
@@
 \label{fig:nztravel12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np1)
-@
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{travel12.pdf}
 \end{center}
@@
 \label{fig:travel12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(np4)
-@
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{risk12.pdf}
 \end{center}
@@
 \label{fig:risk12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(nznp4)
-@
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{nzrisk12.pdf}
 \end{center}
@@
 \label{fig:nzrisk12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(imnp4)
-@
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{imrisk12.pdf}
 \end{center}
@@
 \label{fig:imrisk12}
-\end{figure}
+\end{sidewaysfigure}
 
@@
 
-\begin{figure}[H]
+\begin{sidewaysfigure}[H]
       \centering
@@
      \label{fig:dhbcases}
-\end{figure}
-
-% \begin{figure}[H]
-%      \centering
-%      \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
-%      \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014}
-%      \label{fig:outbreaks}
-% \end{figure}
-% 
-% \begin{figure}[H]
-%      \centering
-%      \includegraphics[width=0.9\textwidth]{averageR0.pdf}
-%      \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in \autoref{fig:outbreaks}. Sporadic cases are excluded.}
-%      \label{fig:r0}
-% \end{figure}
-
-\begin{figure}[H]
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}[H]
      \centering
@@
      \label{fig:outbreaks}
-\end{figure}
-
-\begin{figure}[H]
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}[H]
      \centering
@@
      \label{fig:r0}
-\end{figure}
+\end{sidewaysfigure}
 
@@
 
-\begin{table}[htdp]
-
-\begin{center}
-
-\begin{tabular}{lrrrr}
+\begin{table}[htdp]\small
+\begin{center}
+\begin{tabular}{lrrrrr}
 \hline
@@
 Counties Manukau	&	469300	&	55544	&	32903	&	18880	& 33\\
-Hawke's Bay	&	151700	&	15602	&	6846	&	3751 &	& 24\\
+Hawke's Bay	&	151700	&	15602	&	6846	&	3751 & 24\\
 Hutt Valley	&	138380	&	15198	&	7836	&	4388	& 29\\
@@
 An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
-In that study, they estimated the health-care provider costs to be approximately \euro 229,000,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$426046} or \euro 373\footnote{Jan 2005 \$ to \euro rate 0.5375: \$694} per case.
+In that study, they estimated the health-care provider costs to be approximately \euro 229,000,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$426046} or \euro 373\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$694} per case.
 Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised.
-The mean costs of the hospitalised patients was  \euro 1,877,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$3492} including one patient with encephalitis at a cost of \euro 35,623.\footnote{Jan 2005 \$ to \euro rate 0.5375: \$66275} In addition to the health-care provider costs, additional costs of \euro 89,400\footnote{Jan 2005 \$ to \euro rate 0.5375: \$166325} were incurred by the district public health office, the majority (\euro 85,000,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$158140} 95.1\%) for personnel, \euro 2,300\footnote{Jan 2005 \$ to \euro rate 0.5375: \$4279} (2.6\%) for vaccination, and \euro 2,100\footnote{Jan 2005 \$ to \euro rate 0.5375: \$3907} (2.3\%) for serologic testing.
-Therefore the combined direct costs of these 612 cases amounted to \euro 318,400,\footnote{Jan 2005 \$ to \euro rate 0.5375: \$592372} or \euro 520\footnote{Jan 2005 \$ to \euro rate 0.5375: \$967} per case.
+The mean costs of the hospitalised patients was  \euro 1,877,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$3492} including one patient with encephalitis at a cost of \euro 35,623.\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$66275} In addition to the health-care provider costs, additional costs of \euro 89,400\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$166325} were incurred by the district public health office, the majority (\euro 85,000,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$158140} 95.1\%) for personnel, \euro 2,300\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$4279} (2.6\%) for vaccination, and \euro 2,100\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$3907} (2.3\%) for serologic testing.
+Therefore the combined direct costs of these 612 cases amounted to \euro 318,400,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$592372} or \euro 520\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$967} per case.
 To determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
@@
 \begin{center}
-<<fig=TRUE,echo=FALSE,width = 11>>=
-print(p)
-@
+     \includegraphics[width=0.8\textwidth]{p.pdf}
 \end{center}
@@
 <<results=tex,echo=FALSE>>=
+colnames(vacp) <- c("Age range", "Proportion")
 latex(vacp, file="", table.env=FALSE,rowname=NULL)
@@
 
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using \autoref{eq:prop} are shown in \autoref{fig:numvac}.
-
-\begin{figure}[width=1\textwidth]
-     \centering
-<<fig=TRUE,echo=FALSE,width = 11,height=11>>=
-##
-par(mfrow=c(1,1))
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-plot(naive$Naive ,col="black",bg="black",pch=1,ylab="Number to vaccinate",xlab="Age in Years",
-     main="",cex.lab=2,cex=3)
-points(round(0.28*naive$Naive),col="grey",bg="red",pch=21,cex=2)
-points(x=3:18,y=naive$Naive[3:18],col="grey",bg="orange",pch=23,cex=2) # 
-legend("topright",c("Naive","28% naive, all ages","28% naive, all 2-17 yr olds"),
-       bty="n",pch=c(1,21,23),col=c("black","grey","grey"),pt.bg=c("black","red","orange"),cex=2)
-@
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (\autoref{section:epidemic_modelling}, \autoref{table:cost20} and \autoref{table:cost50})}
-     \label{fig:numvac}
-\end{figure}
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using \autoref{eq:prop} are shown in \autoref{table:vacp}.
 
@@
 
-<<echo=FALSE,include=FALSE,results=hide>>=
-## simulate measles outbreaks/epidemics using GT and R0 from data.
-# http://www.sciencedirect.com/science/article/pii/S0022519311003146
-mu    <- 12
-sigma <- 3.5
-sigma_logn <- sqrt(log(1 + (sigma/mu)^2))
-mu_logn    <- log(mu) - log(1 + (sigma/mu)^2) / 2
-
-# then exp(rnorm(n, mu_logn, sigma_logn)) simulates from lognormal with the given mean and sd.
-
-# the R0 library 
-require(R0)
-#source("estR0.R")
-
-# generation time 
-genTime <- generation.time(type="lognormal", val=c(12, 3.5))
-
-jm.epid <- function (epid.nb, GT, R0, epid.length, family, negbin.size = NULL, 
-          peak.value = 300000, popn = 300000) 
-{
-  if (class(GT) != "R0.GT") {
-    stop("GT object must be of class R0.GT.")
-  }
-  if (family == "negbin" & is.null(negbin.size)) {
-    negbin.size <- R0/4
-  }
-  GT <- GT$GT
-  epidemics <- matrix(data = 0, nrow = epid.length, ncol = epid.nb)
-  for (n in 1:epid.nb) {
-    sim.epid = c(1, rep(0, epid.length - 1))
-    for (t in 1:epid.length) {
-      if (family == "poisson") {
-        susc <- max(0,popn - sum(sim.epid, na.rm=T))
-    #    peak.value[t]<-susc
-        new <- rbinom(sim.epid[t], susc, min(1, R0/susc))
-        if (sum(new) > susc) # arggrggg
-        {
-    #      cat("doing the new thing", susc, sim.epid[t], sum(sim.epid[1:t]),"\n")
-          new <- rep(0, sim.epid[t])
-          for (i in 1:sim.epid[t]) {
-   #         cat("taking a random number, susc=", susc, "\n")
-            new[i] <- rbinom(1, susc, min(1, R0/susc))
-            susc <- susc - new[i]
-            if (susc == 0)
-            {
- #             cat("got 'em all", sum(new), "\n")
-              break;
-            }
-          }
-        }
-#        cat("generating", new[1], "individuals from first, susc=", susc, "\n")
-      }
-      else if (family == "negbin") {
-        new <- rnbinom(sim.epid[t], size = negbin.size, 
-                       mu = R0)
-      }
-if (is.na(sum(new)))
-  break
-
-      newd <- rmultinom(1, sum(new), GT)[, 1]
-      sim.epid[t:(t + length(GT) - 1)] <- sim.epid[t:(t + 
-                                                        length(GT) - 1)] + newd
-      if (sim.epid[t + 1] > peak.value & t < (epid.length - 
-                                                1)) {
-        sim.epid[(t + 2):epid.length] <- 0
-        break
-      }
-    }
-    sim.epid <- sim.epid[!is.na(sim.epid)]
-    epidemics[, n] <- sim.epid
-  }
-  return(epidemics)
-}
-
-## if R 0 == 2
-set.seed(1) # all infected
-res<-jm.epid(epid.nb=1000,GT=genTime,R0=0.99,epid.length=365*5,popn=331385
-             ,family="poisson",peak.value=331385)
-sizes<-colSums(res)
-#sizes
-legMd<-as.factor(paste("median =",c(median(sizes))))
-legMn<-as.factor(paste("mean =",c(round(mean(sizes)))))
-@
-
 \begin{figure}
      \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(sizes,xlim=c(0,200),col="grey",breaks=1000,xlab="Outbreak size",main="",cex.lab=2)
-#summary(sizes)
-abline(v=median(sizes),col="orange",lty=3,lwd=8)
-abline(v=mean(sizes),col="red",lty=3,lwd=8)
-#length(which(sizes>5))/1000
-legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=8,cex=2)
-@
+     \includegraphics[width=0.9\textwidth]{sim.pdf}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
@@
      \centering
-<<fig=TRUE,echo=FALSE,width = 11>>=
-par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(sizes,breaks=1000,xlab="Outbreak size",main="",cex.lab=2,ylim=c(0,20))
-abline(v=median(sizes),col="orange",lty=3,lwd=8)
-abline(v=mean(sizes),col="red",lty=3,lwd=8)
-legend('topright',c(levels(legMd),levels(legMn)),lty=rep(3,2),col=c("orange","red"),bty="n",lwd=8,cex=2)
-@
+     \includegraphics[width=0.9\textwidth]{sim1.pdf}
   \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
-colnames(costtable20)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
+colnames(costtable20)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hosp saved", "Contacts","Costs saved", "Outbreak", "OB costs", "B/C")
 latex(costtable20, file="", table.env=FALSE,rowname=NULL)
@@
 \caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
-colnames(costtable50)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hospitalised", "Hosp saved", "Costs saved", "Outbreak", "OB costs", "B/C")
+colnames(costtable50)<-c("DHB", "Vacc", "Vacc costs", "Wages saved", "Manage saved", "Hosp saved", "Contacts","Costs saved", "Outbreak", "OB costs", "B/C")
 latex(costtable50, file="", table.env=FALSE,rowname=NULL)
@@
 
-\begin{table}[htdp]
+\begin{table}[htdp]\small
 \begin{center}
@@
 \hline
-Age range in years for vaccination & Proportion of the currently na\"{\i}ve population it would cover \\
+Age range (yrs) & Proportion of na\"{\i}ve population\\
 \hline
@@
 \end{tabular}
+\caption{Proportion of the currently na\"{\i}ve New Zealand population that would be vaccinated if all na\"{\i}ve people in each age range were vaccinated.
+The estimated additional proportion of the population requiring vaccination is 28\%}
 \end{center}
@@
 The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$211990} In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak \citep{dayan5}.
-They estimated direct costs per case to be less than US\$500.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$774} However, the combined direct costs of \euro 520\footnote{Jan 2005 \$ to \euro rate 0.5375: \$967} per case in Germany \citep{siedler6} suggests costs may be higher.
+They estimated direct costs per case to be less than US\$500.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$774} However, the combined direct costs of \euro 520\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$967} per case in Germany \citep{siedler6} suggests costs may be higher.
 Thus, our estimates of the additional costs, including indirect costs and wages lost may be appropriate.
@@
 
+\section{Appendix}
+\label{sec:app}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{case_age_nzdep.pdf}
+     \end{center}
+     \caption{Full details are given in (\autoref{fig:cases_all})}
+     \label{fig:CaseAgeNzdep}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{case_eth_nzdep.pdf}
+     \end{center}
+     \caption{Full details are given in (\autoref{fig:cases_all})}
+     \label{fig:CaseNzdepEth}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{case_age_eth.pdf}
+     \end{center}
+     \caption{Full details are given in (\autoref{fig:cases_all})}
+     \label{fig:CaseAgeEth}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{percap_nzdep_age.pdf}
+     \end{center}
+     \caption{Full details are given in (\autoref{fig:percap_all})}
+     \label{fig:PerCapAgeNzdep}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{percap_eth_nzdep.pdf}
+     \end{center}
+     \caption{Full details are given in (\autoref{fig:percap_all})}
+     \label{fig:PerCapNzdepEth}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{percap_eth_age.pdf}
+     \end{center}
+     \caption{Full details are given in (\autoref{fig:percap_all})}
+     \label{fig:PerCapAgeEth}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb1.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Northland}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb2.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Waitemata}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb3.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Auckland}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb4.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Counties_Manukau}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb5.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Waikato}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb6.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Lakes}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb7.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:BayofPlenty}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb8.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Tairawhiti}
+\end{figure}
+
+\begin{figure}[H]
+  \begin{center}
+    \includegraphics[width=0.9\textwidth]{dhb9.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Taranaki}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb10.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:HawkesBay}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb11.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Whanganui}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb12.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Midcentral}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb13.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Hutt}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb14.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:CapitalandCoast}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb15.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Wairarapa}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb16.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:NelsonMarlborough}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb17.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:WestCoast}
+\end{figure}
+
+\begin{figure}[H]
+   \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb18.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Canterbury}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb19.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:SouthCanterbury}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{dhb20.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:Southern}
+\end{figure}
+
 \end{document}

### 2015-03-20 draftfinalreport_v2.tex
Major revisions to text

@@
 However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis.
-In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
-\item Of the 11\% of the population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
+In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--3).
+However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
+\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
-\item Peak overall travel rates, and thus presumably risk from measles importation, is in December.
+\item Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December.
 However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
-\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include 1 and the current 2013--2014 outbreak, as analysed from data until June, is well above one.
+\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed 1 and the 2013--2014 outbreaks, as analysed from data until June, is well above 1 (mean estimate 3.66, range 3.4--3.92 for the last outbreak).
+The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92.
 This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
-\item The cost of the current 2013--2014 measles outbreak is estimated to be at least \$805,000.
-\item The mean wage loss per measles case is estimated to be \$839
+\item The cost of the 2013--2014 measles outbreaks is estimated to be approximately \$1,332,539 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
+\item The mean wage loss per measles case is estimated to be approximately \$852.
 \item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
-\item The mean management cost per case is \$1,765
-\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1).
-\item Outbreaks, with a mean size of 61 cases per year, median of 2 cases, but possibly a peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out following additional vaccinations.
+\item The mean public health service cost per case is \$1,765 excluding hospitalisation costs.
+\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170.4 per day.
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed \$8,214 per vaccine before the costs exceed the benefits.
+\item Measles introductions to New Zealand with a mean size of 61 cases per year, median of 2 cases, but rarely a peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out following additional vaccinations.
 \end{itemize}
@@
 The last widespread measles outbreaks in New Zealand occurred in 1991 and in 1997.
-Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and another significant outbreak is currently ongoing in the Auckland and Waikato regions.
-The outbreak in 2011--2012 lasted for more than 12 months and the current 2013--2014 outbreak started at the end of December 2013 and is ongoing (as of \date{3 July 2014}).
-In 2013, but prior to the current 2013--2014 outbreak, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
+Since then, smaller but significant outbreaks have occurred in 2009 (mainly in Canterbury) and in 2011--2012 (mainly in the Auckland region) and other significant outbreaks occurred in the Auckland and Waikato regions in 2013--2014.
+The outbreak in 2011--2012 lasted for more than 12 months.
+In 2013--2014 outbreaks started at the end of December 2013, and the last finished in August 2014.
+In 2013, but prior to the 2013--2014 outbreaks, New Zealand was advised by the Western Pacific Regional Verification Commission for Measles Elimination (RVC) that it can request verification of non-endemic status three years after the last case of the 2011--12 outbreak in June 2012.
 
 Previous measles analyses, including two in New Zealand by Prof.
-Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles %\citep{roberts0,roberts4}.
+Roberts, estimated the interruption of measles virus transmission can be achieved by herd immunity when approximately 95 percent of the population is homogeneously immune to measles \citep{roberts0,roberts4}.
 Thus, while New Zealand immunisation activities have led to measles outbreaks becoming less frequent, with decreasing numbers of cases, outbreaks still occur (as described above).
 Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see \autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children and children under two years of age.
-Those under two year olds are thought be be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
+Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children (especially young adults) and children under two years of age.
+Those under two year olds are thought to be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
 
@@
 To help understand the risk of measles importation, with a particular goal of enabling the Ministry to better inform travellers and understand high risk periods, we sought to quantitatively evaluate the risk of measles importation from travel (\autoref{sub:imp_risk}).
-
-\begin{figure}
+The details of the methods and approaches are given in the text below for each section.
+
+\begin{sidewaysfigure}
      %\centering
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mfrow=c(1,1))
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-% @
-% \end{center}
-% \caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
-% \label{fig:ageinyears}
-% \end{figure}
+\begin{figure}
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{case_age_dist.pdf}
+     \end{center}
+\caption{Numbers and age of measles cases in years in New Zealand for two periods, 1997--2014 and 2007--2014}
+\label{fig:ageinyears}
+\end{figure}
 
@@
 
-\begin{table}[hbtp]
-\footnotesize
+\begin{table}[hbtp]\scriptsize
 %latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
 \begin{center}
-\begin{tabular}{lllr}
+\begin{tabular}{lllrl}
 \hline\hline
-\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}\tabularnewline
-\hline
-1-5&0-2&None&$3$\tabularnewline
-1-5&3-5&None&$1$\tabularnewline
-1-5&6-17&None&$3$\tabularnewline
-6-10&6-17&None&$4$\tabularnewline
-1-5&18-24&None&$2$\tabularnewline
-6-10&18-24&None&$1$\tabularnewline
-1-5&25+&None&$5$\tabularnewline
-6-10&25+&None&$3$\tabularnewline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{merge}\tabularnewline
+\hline
+1-3&0-1&Asian&$  6$&1-3 0-1 Asian\tabularnewline
+4-7&0-1&Asian&$ 10$&4-7 0-1 Asian\tabularnewline
+8-10&0-1&Asian&$  3$&8-10 0-1 Asian\tabularnewline
+1-3&2-4&Asian&$  1$&1-3 2-4 Asian\tabularnewline
+1-3&5-17&Asian&$  6$&1-3 5-17 Asian\tabularnewline
+4-7&5-17&Asian&$  8$&4-7 5-17 Asian\tabularnewline
+8-10&5-17&Asian&$  2$&8-10 5-17 Asian\tabularnewline
+1-3&18-24&Asian&$  2$&1-3 18-24 Asian\tabularnewline
+4-7&18-24&Asian&$  3$&4-7 18-24 Asian\tabularnewline
+8-10&18-24&Asian&$  3$&8-10 18-24 Asian\tabularnewline
+1-3&25+&Asian&$  4$&1-3 25+ Asian\tabularnewline
+4-7&25+&Asian&$ 11$&4-7 25+ Asian\tabularnewline
+8-10&25+&Asian&$  8$&8-10 25+ Asian\tabularnewline
+1-3&0-1&European&$ 40$&1-3 0-1 European\tabularnewline
+4-7&0-1&European&$ 58$&4-7 0-1 European\tabularnewline
+8-10&0-1&European&$ 24$&8-10 0-1 European\tabularnewline
+1-3&2-4&European&$ 37$&1-3 2-4 European\tabularnewline
+4-7&2-4&European&$ 22$&4-7 2-4 European\tabularnewline
+8-10&2-4&European&$  6$&8-10 2-4 European\tabularnewline
+1-3&5-17&European&$160$&1-3 5-17 European\tabularnewline
+4-7&5-17&European&$118$&4-7 5-17 European\tabularnewline
+8-10&5-17&European&$ 40$&8-10 5-17 European\tabularnewline
+1-3&18-24&European&$ 24$&1-3 18-24 European\tabularnewline
+4-7&18-24&European&$ 26$&4-7 18-24 European\tabularnewline
+8-10&18-24&European&$ 20$&8-10 18-24 European\tabularnewline
+1-3&25+&European&$ 54$&1-3 25+ European\tabularnewline
+4-7&25+&European&$ 46$&4-7 25+ European\tabularnewline
+8-10&25+&European&$ 29$&8-10 25+ European\tabularnewline
+1-3&0-1&Maori&$  9$&1-3 0-1 Maori\tabularnewline
+4-7&0-1&Maori&$ 13$&4-7 0-1 Maori\tabularnewline
+8-10&0-1&Maori&$ 32$&8-10 0-1 Maori\tabularnewline
+1-3&2-4&Maori&$  4$&1-3 2-4 Maori\tabularnewline
+4-7&2-4&Maori&$ 12$&4-7 2-4 Maori\tabularnewline
+8-10&2-4&Maori&$ 10$&8-10 2-4 Maori\tabularnewline
+1-3&5-17&Maori&$  8$&1-3 5-17 Maori\tabularnewline
+4-7&5-17&Maori&$ 33$&4-7 5-17 Maori\tabularnewline
+8-10&5-17&Maori&$ 74$&8-10 5-17 Maori\tabularnewline
+1-3&18-24&Maori&$  3$&1-3 18-24 Maori\tabularnewline
+4-7&18-24&Maori&$  2$&4-7 18-24 Maori\tabularnewline
+8-10&18-24&Maori&$  8$&8-10 18-24 Maori\tabularnewline
+1-3&25+&Maori&$  1$&1-3 25+ Maori\tabularnewline
+4-7&25+&Maori&$  3$&4-7 25+ Maori\tabularnewline
+8-10&25+&Maori&$  4$&8-10 25+ Maori\tabularnewline
+4-7&0-1&MLA&$  2$&4-7 0-1 MLA\tabularnewline
+8-10&0-1&MLA&$  1$&8-10 0-1 MLA\tabularnewline
+8-10&2-4&MLA&$  1$&8-10 2-4 MLA\tabularnewline
+8-10&5-17&MLA&$  1$&8-10 5-17 MLA\tabularnewline
+8-10&18-24&MLA&$  2$&8-10 18-24 MLA\tabularnewline
+4-7&25+&MLA&$  5$&4-7 25+ MLA\tabularnewline
+8-10&25+&MLA&$  1$&8-10 25+ MLA\tabularnewline
+1-3&0-1&Pacific&$  1$&1-3 0-1 Pacific\tabularnewline
+4-7&0-1&Pacific&$  6$&4-7 0-1 Pacific\tabularnewline
+8-10&0-1&Pacific&$ 48$&8-10 0-1 Pacific\tabularnewline
+4-7&2-4&Pacific&$  1$&4-7 2-4 Pacific\tabularnewline
+8-10&2-4&Pacific&$ 10$&8-10 2-4 Pacific\tabularnewline
+1-3&5-17&Pacific&$  2$&1-3 5-17 Pacific\tabularnewline
+4-7&5-17&Pacific&$  4$&4-7 5-17 Pacific\tabularnewline
+8-10&5-17&Pacific&$ 21$&8-10 5-17 Pacific\tabularnewline
+4-7&18-24&Pacific&$  4$&4-7 18-24 Pacific\tabularnewline
+8-10&18-24&Pacific&$  5$&8-10 18-24 Pacific\tabularnewline
+1-3&25+&Pacific&$  1$&1-3 25+ Pacific\tabularnewline
+4-7&25+&Pacific&$  3$&4-7 25+ Pacific\tabularnewline
+8-10&25+&Pacific&$  9$&8-10 25+ Pacific\tabularnewline
 \hline
@@
 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mfrow=c(1,1))
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
-% hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-% legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, cases, fill=Age)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%  # theme(text=element_text(size=45))+
-%   ylab("cases")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, cases, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-% #  theme(text=element_text(size=45))+
-%   ylab("cases")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(Age, cases, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-% #  theme(text=element_text(size=45))+
-%   ylab("cases")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, perCap, fill=Age)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%  # theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(NZDep, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tp, aes(Age, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tpsub, aes(NZDep, perCap, fill=Age)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tpsub, aes(NZDep, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% ggplot(tpsub, aes(Age, perCap, fill=Ethnicity)) + 
-%   geom_bar(stat="identity", position="dodge")+
-%   #theme(text=element_text(size=45))+
-%   ylab("Per capita per 10000")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(cex.axis=2)
-% par(mar=c(5,6,4,2)+0.1)
-% hist(tpsub$cases,xlab="Number of cases",main='',col='grey',breaks=20,ylab="Number of categories",cex.lab=2)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% plot(exp(res),tpsub$cases,xlab="cases",ylab='predictions',main="",pch=16,col="darkgrey",cex.lab=2,cex=2)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% #hist(model3$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-% hist(model2$residuals,main="",xlab="residuals",col="grey",cex.lab=2,ylab="")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% barplot(t(AgeVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=2)
-% legend("topright",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1.8,bty="n")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=2)
-% hist(vac$Dose1Mths,breaks=100,col=rgb(0,0,1,1/4),xlab="Age in months",main="",cex.lab=2,ylab="Frequency")
-% abline(v=12,col="blue")
-% hist(vac$Dose2Mths,breaks=50,add=T,col=rgb(1,0,0,1/4))
-% abline(v=60,col="red")
-% legend("topright",c("Dose 1","Dose 2"),col=c(rgb(0,0,1,1/4),rgb(1,0,0,1/4)),pch=15,bty="n",cex=1.8)
-% legend("top",c("12 months","60 months"),col=c("blue","red"),lty=1,bty="n",cex=1.8)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1)
-% barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
-% legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1)
-% par(mfrow=c(2,1))
-% barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
-% legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1)
-% barplot(DOBVac[,3],col=rgb(1,0,0,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
-% legend("topleft",c("Dose 2"),col=rgb(1,0,0,1/4),pch=15,bty="n",cex=1)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-% 
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(mar=c(5,6,4,2)+0.1)
-% par(cex.axis=1.5)
-% par(mfrow=c(1,1))
-% barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
-% legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-% \begin{figure}
-% \begin{center}
-% <<fig=TRUE,echo=FALSE,width = 15,height=12>>=
-% par(cex.axis=2)
-% par(mar=c(5,6,4,2)+0.1)
-% plot(pop,xlab="Age",ylab="Population",cex.lab=2)
-% points(naive,pch=16)
-% legend("topright",c("Population","Naive"),pch=c(1,16),bty="n",cex=1.8)
-% @
-% \end{center}
-% \caption{test}
-% %\label{fig:test}
-% \end{figure}
-
-The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap}, \autoref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
-Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-019}
-\end{center}
-\caption{Per capita rates of measles infections for specific age categories (0--2, 3--5, 6--17, 18--24, and 24+ years) and New Zealand deprivation indices (NZDep, 1--5 and 6--10) for 2007--2014, broken down by ethnicity (\autoref{table:percap} for details)}
-\label{fig:percap}
-\end{figure}
-
-\begin{table}[hbtp]
-\footnotesize
-\caption{Numbers of measles cases (Cases), population sizes (Ppulation) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{percap_ordert.pdf}
+\end{center}
+\caption{test}
+\label{fig:percap_all}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{case_ordert.pdf}
+\end{center}
+\caption{test}
+\label{fig:cases_all}
+\end{sidewaysfigure}
+
+Two way interactions are shown in \autoref{sec:app} (\autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).
+
+\begin{table}[hbtp]\scriptsize
+\caption{Per capita measles rates}
 %latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
+\end{tabular}\end{center}\label{table:percap_rates}
+\end{table}
+
+\begin{table}[hbtp]\footnotesize
+\caption{Proportions of the naive population the would be covered if vaccinated by age groups}
+%latex.default(propvCohort, file = "", table.env = FALSE)%
+\begin{center}
+\begin{tabular}{lr}
+\hline\hline
+\multicolumn{1}{l}{propvCohort}&\multicolumn{1}{c}{}\tabularnewline
+\hline
+2-5&$0.04$\tabularnewline
+5-11&$0.15$\tabularnewline
+5-16&$0.27$\tabularnewline
+5-18&$0.31$\tabularnewline
+11-16&$0.14$\tabularnewline
+11-18&$0.18$\tabularnewline
+0-29&$0.72$\tabularnewline
+0-44&$0.89$\tabularnewline
+\textgreater 8&$0.80$\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:vaccine_cohort}
+\end{table}
+
+The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap}, \autoref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
+Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
+
+\begin{table}[hbtp] \scriptsize
+\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
+%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{lllrrr}
+\hline\hline
+\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
+\hline
+1-3&0-1&Asian&$  2269$&$  6$&$26.4434$\tabularnewline
+4-7&0-1&Asian&$  3748$&$ 10$&$26.6809$\tabularnewline
+8-10&0-1&Asian&$  2583$&$  3$&$11.6144$\tabularnewline
+1-3&2-4&Asian&$  3403$&$  1$&$ 2.9386$\tabularnewline
+4-7&2-4&Asian&$  5623$&$  0$&$ 0.0000$\tabularnewline
+8-10&2-4&Asian&$  3874$&$  0$&$ 0.0000$\tabularnewline
+1-3&5-17&Asian&$ 20625$&$  6$&$ 2.9091$\tabularnewline
+4-7&5-17&Asian&$ 29721$&$  8$&$ 2.6917$\tabularnewline
+8-10&5-17&Asian&$ 16777$&$  2$&$ 1.1921$\tabularnewline
+1-3&18-24&Asian&$ 12057$&$  2$&$ 1.6588$\tabularnewline
+4-7&18-24&Asian&$ 23904$&$  3$&$ 1.2550$\tabularnewline
+8-10&18-24&Asian&$ 21063$&$  3$&$ 1.4243$\tabularnewline
+1-3&25+&Asian&$ 54258$&$  4$&$ 0.7372$\tabularnewline
+4-7&25+&Asian&$ 86055$&$ 11$&$ 1.2783$\tabularnewline
+8-10&25+&Asian&$ 54759$&$  8$&$ 1.4609$\tabularnewline
+1-3&0-1&European&$ 23961$&$ 40$&$16.6938$\tabularnewline
+4-7&0-1&European&$ 28051$&$ 58$&$20.6766$\tabularnewline
+8-10&0-1&European&$ 16866$&$ 24$&$14.2298$\tabularnewline
+1-3&2-4&European&$ 35942$&$ 37$&$10.2944$\tabularnewline
+4-7&2-4&European&$ 42076$&$ 22$&$ 5.2286$\tabularnewline
+8-10&2-4&European&$ 25299$&$  6$&$ 2.3716$\tabularnewline
+1-3&5-17&European&$180875$&$160$&$ 8.8459$\tabularnewline
+4-7&5-17&European&$194100$&$118$&$ 6.0793$\tabularnewline
+8-10&5-17&European&$106731$&$ 40$&$ 3.7477$\tabularnewline
+1-3&18-24&European&$ 63051$&$ 24$&$ 3.8064$\tabularnewline
+4-7&18-24&European&$ 94422$&$ 26$&$ 2.7536$\tabularnewline
+8-10&18-24&European&$ 68016$&$ 20$&$ 2.9405$\tabularnewline
+1-3&25+&European&$619698$&$ 54$&$ 0.8714$\tabularnewline
+4-7&25+&European&$734550$&$ 46$&$ 0.6262$\tabularnewline
+8-10&25+&European&$371985$&$ 29$&$ 0.7796$\tabularnewline
+1-3&0-1&Maori&$  3265$&$  9$&$27.5651$\tabularnewline
+4-7&0-1&Maori&$  8378$&$ 13$&$15.5168$\tabularnewline
+8-10&0-1&Maori&$ 15094$&$ 32$&$21.2005$\tabularnewline
+1-3&2-4&Maori&$  4897$&$  4$&$ 8.1683$\tabularnewline
+4-7&2-4&Maori&$ 12568$&$ 12$&$ 9.5481$\tabularnewline
+8-10&2-4&Maori&$ 22642$&$ 10$&$ 4.4166$\tabularnewline
+1-3&5-17&Maori&$ 21750$&$  8$&$ 3.6782$\tabularnewline
+4-7&5-17&Maori&$ 53963$&$ 33$&$ 6.1153$\tabularnewline
+8-10&5-17&Maori&$ 94758$&$ 74$&$ 7.8094$\tabularnewline
+1-3&18-24&Maori&$  7329$&$  3$&$ 4.0933$\tabularnewline
+4-7&18-24&Maori&$ 20862$&$  2$&$ 0.9587$\tabularnewline
+8-10&18-24&Maori&$ 35664$&$  8$&$ 2.2432$\tabularnewline
+1-3&25+&Maori&$ 35934$&$  1$&$ 0.2783$\tabularnewline
+4-7&25+&Maori&$ 87192$&$  3$&$ 0.3441$\tabularnewline
+8-10&25+&Maori&$140820$&$  4$&$ 0.2841$\tabularnewline
+1-3&0-1&MLA&$   273$&$  0$&$ 0.0000$\tabularnewline
+4-7&0-1&MLA&$   519$&$  2$&$38.5356$\tabularnewline
+8-10&0-1&MLA&$   553$&$  1$&$18.0832$\tabularnewline
+1-3&2-4&MLA&$   410$&$  0$&$ 0.0000$\tabularnewline
+4-7&2-4&MLA&$   778$&$  0$&$ 0.0000$\tabularnewline
+8-10&2-4&MLA&$   830$&$  1$&$12.0482$\tabularnewline
+1-3&5-17&MLA&$  1832$&$  0$&$ 0.0000$\tabularnewline
+4-7&5-17&MLA&$  3121$&$  0$&$ 0.0000$\tabularnewline
+8-10&5-17&MLA&$  3250$&$  1$&$ 3.0769$\tabularnewline
+1-3&18-24&MLA&$   906$&$  0$&$ 0.0000$\tabularnewline
+4-7&18-24&MLA&$  1806$&$  0$&$ 0.0000$\tabularnewline
+8-10&18-24&MLA&$  2076$&$  2$&$ 9.6339$\tabularnewline
+1-3&25+&MLA&$  4569$&$  0$&$ 0.0000$\tabularnewline
+4-7&25+&MLA&$  7452$&$  5$&$ 6.7096$\tabularnewline
+8-10&25+&MLA&$  6342$&$  1$&$ 1.5768$\tabularnewline
+1-3&0-1&Pacific&$   643$&$  1$&$15.5521$\tabularnewline
+4-7&0-1&Pacific&$  2113$&$  6$&$28.3956$\tabularnewline
+8-10&0-1&Pacific&$  7389$&$ 48$&$64.9614$\tabularnewline
+1-3&2-4&Pacific&$   964$&$  0$&$ 0.0000$\tabularnewline
+4-7&2-4&Pacific&$  3169$&$  1$&$ 3.1556$\tabularnewline
+8-10&2-4&Pacific&$ 11084$&$ 10$&$ 9.0220$\tabularnewline
+1-3&5-17&Pacific&$  4212$&$  2$&$ 4.7483$\tabularnewline
+4-7&5-17&Pacific&$ 13419$&$  4$&$ 2.9808$\tabularnewline
+8-10&5-17&Pacific&$ 47165$&$ 21$&$ 4.4525$\tabularnewline
+1-3&18-24&Pacific&$  1707$&$  0$&$ 0.0000$\tabularnewline
+4-7&18-24&Pacific&$  6174$&$  4$&$ 6.4788$\tabularnewline
+8-10&18-24&Pacific&$ 18189$&$  5$&$ 2.7489$\tabularnewline
+1-3&25+&Pacific&$  8580$&$  1$&$ 1.1655$\tabularnewline
+4-7&25+&Pacific&$ 26688$&$  3$&$ 1.1241$\tabularnewline
+8-10&25+&Pacific&$ 74757$&$  9$&$ 1.2039$\tabularnewline
+\hline
 \end{tabular}\end{center}\label{table:percap}
@@
 
+\subsection{Regression analyses results}
+\label{sub:regression_results}
+
+The distribution of the measles cases per category used in the regression analyses are in \autoref{fig:histcasecat}.
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{Cases_regmodel.pdf}
+\end{center}
+\caption{Distribution of measles cases per category (\autoref{table:percap}) used in the final regression model (\autoref{eq:reg})}
+\label{fig:histcasecat}
+\end{figure}
+
+The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}.
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{Cases_regmodel_prediction.pdf}
+\end{center}
+\caption{Regression model (\autoref{eq:reg}) predictions plotted against the case data (\autoref{table:percap})}
+\label{fig:predict}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{Cases_regmodel_resid.pdf}
+\end{center}
+\caption{Histogram of residuals from the regression model (\autoref{eq:reg})}
+\label{fig:resid}
+\end{figure}
+
+The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
+
+\vspace{5mm} %5mm vertical space
+\begin{table}
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Fri Mar 20 22:48:21 2015
+\begin{tabular}{lrrrrrr}
+  \hline
+ & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
+  \hline
+NULL &  &  & 59 & 1532.80 &  &  \\ 
+  Age & 4 & 1345.17 & 55 & 187.63 & 204.01 & 0.0000 \\ 
+  Ethnicity & 3 & 19.88 & 52 & 167.75 & 4.02 & 0.0141 \\ 
+  NZDep & 2 & 17.10 & 50 & 150.65 & 5.19 & 0.0102 \\ 
+  Age:Ethnicity & 12 & 80.80 & 38 & 69.85 & 4.09 & 0.0004 \\ 
+   \hline
+\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using \autoref{eq:reg} and a quasipoisson error structure}
+\label{table:anova}
+\end{table}
+
+A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.
+
+\vspace{5mm} %5mm vertical space
+\begin{table}
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Fri Mar 20 22:48:21 2015
+\begin{tabular}{rrrrr}
+  \hline
+ & Estimate & Std. Error & t value & Pr($>$$|$t$|$) \\ 
+  \hline
+(Intercept) & -5.9180 & 0.3002 & -19.71 & 0.0000 \\ 
+  Age2-4 & -3.3499 & 1.3172 & -2.54 & 0.0152 \\ 
+  Age5-17 & -2.2407 & 0.4357 & -5.14 & 0.0000 \\ 
+  Age18-24 & -2.7392 & 0.5412 & -5.06 & 0.0000 \\ 
+  Age25+ & -2.9353 & 0.3980 & -7.37 & 0.0000 \\ 
+  EthnicityEuropean & -0.2464 & 0.3168 & -0.78 & 0.4415 \\ 
+  EthnicityMaori & -0.0383 & 0.3433 & -0.11 & 0.9118 \\ 
+  EthnicityPacific & 0.9736 & 0.3439 & 2.83 & 0.0074 \\ 
+  NZDep4-7 & -0.2609 & 0.0948 & -2.75 & 0.0090 \\ 
+  NZDep8-10 & -0.3043 & 0.1040 & -2.93 & 0.0058 \\ 
+  Age2-4:EthnicityEuropean & 2.3148 & 1.3319 & 1.74 & 0.0903 \\ 
+  Age5-17:EthnicityEuropean & 1.2453 & 0.4566 & 2.73 & 0.0096 \\ 
+  Age18-24:EthnicityEuropean & 1.0187 & 0.5743 & 1.77 & 0.0841 \\ 
+  Age25+:EthnicityEuropean & -0.2344 & 0.4298 & -0.55 & 0.5887 \\ 
+  Age2-4:EthnicityMaori & 2.2135 & 1.3524 & 1.64 & 0.1099 \\ 
+  Age5-17:EthnicityMaori & 1.1421 & 0.4844 & 2.36 & 0.0236 \\ 
+  Age18-24:EthnicityMaori & 0.4465 & 0.6710 & 0.67 & 0.5098 \\ 
+  Age25+:EthnicityMaori & -1.2693 & 0.6285 & -2.02 & 0.0505 \\ 
+  Age2-4:EthnicityPacific & 1.3350 & 1.3838 & 0.96 & 0.3408 \\ 
+  Age5-17:EthnicityPacific & -0.3256 & 0.5299 & -0.61 & 0.5426 \\ 
+  Age18-24:EthnicityPacific & -0.0167 & 0.7113 & -0.02 & 0.9814 \\ 
+  Age25+:EthnicityPacific & -0.8973 & 0.5614 & -1.60 & 0.1183 \\ 
+   \hline
+\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+\label{table:regression_riskfactors}
+\end{table}
+\vspace{5mm} %5mm vertical space
+
+Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
+Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also \autoref{fig:ageinyears}).
+
+The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
+Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}, \autoref{table:sero}, and \autoref{sub:popim}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}).
+Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
+
 \subsection{Vaccination history of measles infection and population immunity estimation methods}
@@
 
+\subsection{Vaccination history and measles infection results}
+
+The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (\autoref{fig:ageandvac}).
+However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second.
+A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
+Plotting the cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).
+
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{vacc_status.pdf}
+\end{center}
+\caption{Age and vaccination status of measles cases, 2007--2014}
+\label{fig:ageandvac}
+\end{figure}
+
+% ## insert figures here
+
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{vacc_age.pdf}
+\end{center}
+\caption{Age of vaccination of vaccinated measles cases, 2007--2014}
+\label{fig:vaccstat}
+\end{figure}
+
 \subsection{Measles importation risk methods}
@@
 We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
-We present the data from 2012 because this year had the most complete WHO measles data and yet is most recent, thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
-
-\subsection{Regression analyses results}
-\label{sub:regression_results}
-
-The distribution of the meases cases per category used in the regression analyses are in \autoref{fig:histcasecat}.
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-023}
-\end{center}
-\caption{Distribution of measles cases per category (\autoref{table:percap}) used in the final regression model (\autoref{eq:reg})}
-\label{fig:histcasecat}
-\end{figure}
-
-The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}.
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-024}
-\end{center}
-\caption{Regression model (\autoref{eq:reg}) predictions plotted against the case data (\autoref{table:percap})}
-\label{fig:predict}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-026}
-\end{center}
-\caption{Histogram of residuals from the regression model (\autoref{eq:reg})}
-\label{fig:resid}
-\end{figure}
-
-The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
-\par
-\vspace{5mm} %5mm vertical space
-\begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Wed Mar 18 16:47:24 2015
-\begin{tabular}{lrrrrrr}
-  \hline
- & Df & Deviance & Resid. Df & Resid. Dev & F & Pr($>$F) \\ 
-  \hline
-NULL &  &  & 59 & 1532.80 &  &  \\ 
-  Age & 4 & 1345.17 & 55 & 187.63 & 204.01 & 0.0000 \\ 
-  Ethnicity & 3 & 19.88 & 52 & 167.75 & 4.02 & 0.0141 \\ 
-  NZDep & 2 & 17.10 & 50 & 150.65 & 5.19 & 0.0102 \\ 
-  Age:Ethnicity & 12 & 80.80 & 38 & 69.85 & 4.09 & 0.0004 \\ 
-   \hline
-\end{tabular}\caption{Significance of different predictor variables for measles risk factors from 2007-2014, estimated using \autoref{eq:reg} and a quasipoisson error structure}
-\label{table:anova}
-\end{table}
-\vspace{5mm} %5mm vertical space
-
-A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.
-
-% \vspace{5mm} %5mm vertical space
-% \begin{table}
-% % <<label=regmodel,include=FALSE,results=tex,echo=FALSE>>=
-% % library(xtable)
-% % xtb<-xtable(summary(model3))
-% % print(xtb,floating = FALSE)
-% % @
-% \caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
-% \label{table:regression_riskfactors}
-% \end{table}
-% \vspace{5mm} %5mm vertical space
-
-\par Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also \autoref{fig:ageinyears}).
-
-The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
-Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
-In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}, \autoref{table:sero}, and \autoref{sub:popim}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}).
-Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
-
-\subsection{Vaccination history and measles infection results}
-
-The majority of the measles cases (82.8\%, 955/1154) from 2007--2014 were in unvaccinated people (\autoref{fig:ageandvac}).
-However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second.
-A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-028}
-\end{center}
-\caption{Age and vaccination status of measles cases, 2007--2014}
-\label{fig:ageandvac}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-029}
-\end{center}
-\caption{Age of vaccination of vaccinated measles cases, 2007--2014}
-\label{fig:vaccstat}
-\end{figure}
+We present the data from 2012 and 2013 because 2012 had the most complete WHO measles data, whereas 2013 is most recent, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
 
@@
 
-%%%%
-
 \begin{figure}
 \begin{center}
-\includegraphics{draftfinalreport_v2-031}
+     \includegraphics[width=0.9\textwidth]{naive_au.pdf}
 \end{center}
@@
 \end{figure}
-%%%%
+
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{naive_au_prop.pdf}
+\end{center}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator}
+\label{fig:naive_au_prop}
+\end{figure}
 
@@
 Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (\autoref{fig:naive_map}).
-The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures:
-% \begin{itemize}[noitemsep,nolistsep]
-% \item Northland: \autoref{fig:Northland}
-% \item Waitemata: \autoref{fig:Waitemata}
-% \item Auckland: \autoref{fig:Auckland}
-% \item Counties Manukau: \autoref{fig:Counties_Manukau}
-% \item Waikato: \autoref{fig:Waikato}
-% \item Lakes: \autoref{fig:Lakes}
-% \item Bay of Plenty: \autoref{fig:BayofPlenty}
-% \item Tairawhiti: \autoref{fig:Tairawhiti}
-% \item Taranaki: \autoref{fig:Taranaki}
-% \item Hawke's Bay: \autoref{fig:HawkesBay}
-% \item Whanganui: \autoref{fig:Whanganui}
-% \item Mid-Central: \autoref{fig:Midcentral}
-% \item Hutt: \autoref{fig:Hutt}
-% \item Capital and Coast: \autoref{fig:CapitalandCoast}
-% \item Wairarapa: \autoref{fig:Wairarapa}
-% \item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
-% \item West Coast: \autoref{fig:WestCoast}
-% \item Canterbury: \autoref{fig:Canterbury}
-% \item South Canterbury: \autoref{fig:SouthCanterbury}
-% \item Southern: \autoref{fig:Southern}
-% \end{itemize}
+The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures in the \autoref{sec:app}:
+\begin{itemize}[noitemsep,nolistsep]
+\item Northland: \autoref{fig:Northland}
+\item Waitemata: \autoref{fig:Waitemata}
+\item Auckland: \autoref{fig:Auckland}
+\item Counties Manukau: \autoref{fig:Counties_Manukau}
+\item Waikato: \autoref{fig:Waikato}
+\item Lakes: \autoref{fig:Lakes}
+\item Bay of Plenty: \autoref{fig:BayofPlenty}
+\item Tairawhiti: \autoref{fig:Tairawhiti}
+\item Taranaki: \autoref{fig:Taranaki}
+\item Hawke's Bay: \autoref{fig:HawkesBay}
+\item Whanganui: \autoref{fig:Whanganui}
+\item Mid-Central: \autoref{fig:Midcentral}
+\item Hutt: \autoref{fig:Hutt}
+\item Capital and Coast: \autoref{fig:CapitalandCoast}
+\item Wairarapa: \autoref{fig:Wairarapa}
+\item Nelson Marlborough: \autoref{fig:NelsonMarlborough}
+\item West Coast: \autoref{fig:WestCoast}
+\item Canterbury: \autoref{fig:Canterbury}
+\item South Canterbury: \autoref{fig:SouthCanterbury}
+\item Southern: \autoref{fig:Southern}
+\end{itemize}
+
+Plotting the lower level data available from the NIR database highlights the issue of surveillance data requiring appropriate denominator data, as these analyses show that there is a lack of correspondence between the national census data and the NIR data (\autoref{fig:nir})
 
@@
 \begin{center}
-\includegraphics{draftfinalreport_v2-033}
+     \includegraphics[width=0.9\textwidth]{naive_allPop.pdf}
 \end{center}
@@
 
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb1.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Northland District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Northland}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb2.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waitemata District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Waitemata}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb3.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Auckland District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Auckland}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb4.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Counties Manukau District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Counties_Manukau}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb5.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Waikato District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Waikato}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb6.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Lakes District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Lakes}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb7.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Bay of Plenty District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:BayofPlenty}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb8.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Tairawhiti District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Tairawhiti}
-\end{figure}
-
-\begin{figure}[H]
-  \begin{center}
-    \includegraphics[width=0.6\textwidth]{dhb9.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Taranaki District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Taranaki}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb10.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hawke's Bay District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:HawkesBay}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb11.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Whanganui District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Whanganui}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb12.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Midcentral District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Midcentral}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb13.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Hutt District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Hutt}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb14.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Capital and Coast District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:CapitalandCoast}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb15.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Wairarapa District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Wairarapa}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb16.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Nelson Marlborough District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:NelsonMarlborough}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb17.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, West Coast District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:WestCoast}
-\end{figure}
-
-\begin{figure}[H]
-   \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb18.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Canterbury}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb19.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, South Canterbury District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:SouthCanterbury}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.6\textwidth]{dhb20.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per age class, Southern District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:Southern}
-\end{figure}
-
 \subsection{Measles importation risk results}
-
-%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-%%%%%%%%%%%%%%%%%%%%%
-
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2012, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
+\caption{Risk of measles importation to New Zealand in 2013, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
 %latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
 %latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2012, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
+\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
 %latex.default(imrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation.
-incidence: measles incidence per million; cover: percent vaccination cover (\%); immigration: the number of travellers entering New Zealand from the country}
+incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of travellers entering New Zealand from the country}
 %latex.default(toprisk2, file = "", table.env = FALSE, rowname = NULL)%
@@
 The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in \autoref{table:toprisk}.
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-047}
+WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday.
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{incidence12.pdf}
 \end{center}
@@
 \label{fig:incidence12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-048}
-\end{center}
-\caption{Measles vaccination cover (\%), 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{cover12.pdf}
+\end{center}
+\caption{Measles vaccination cover (\% coverage of children that receive one dose of measles vaccine by their second birthday.), 2012.
+Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+Grey indicates not reported.}
 \label{fig:cover12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-049}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{immigration12.pdf}
 \end{center}
@@
 \label{fig:immigration12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-050}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{nztravel12.pdf}
 \end{center}
@@
 \label{fig:nztravel12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-051}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{travel12.pdf}
 \end{center}
@@
 \label{fig:travel12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-052}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{risk12.pdf}
 \end{center}
@@
 \label{fig:risk12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-053}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{nzrisk12.pdf}
 \end{center}
@@
 \label{fig:nzrisk12}
-\end{figure}
-
-\begin{figure}
-\begin{center}
-\includegraphics{draftfinalreport_v2-054}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{imrisk12.pdf}
 \end{center}
@@
 \label{fig:imrisk12}
-\end{figure}
-%%%%%%%%%%% REVISE FIGURE %%%%%%%%%%%
+\end{sidewaysfigure}
+
 \begin{figure}
 \begin{center}
-\includegraphics{draftfinalreport_v2-055}
+\includegraphics{draftfinalreport_v2-017}
 \end{center}
@@
      \label{fig:dhbcases}
-\end{figure}
-
-\begin{figure}[H]
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}[H]
      \centering
-     \includegraphics[width=0.9\textwidth]{outbreaks_for_R0.pdf}
+     \includegraphics[width=0.9\textwidth]{outbreaks_2009_2014.pdf}
      \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014}
      \label{fig:outbreaks}
-\end{figure}
-
-\begin{figure}[H]
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}[H]
      \centering
-     \includegraphics[width=0.9\textwidth]{averageR0.pdf}
+     \includegraphics[width=0.9\textwidth]{averageR0_all_ob.pdf}
      \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in \autoref{fig:outbreaks}. Sporadic cases are excluded.}
      \label{fig:r0}
-\end{figure}
+\end{sidewaysfigure}
 
@@
 The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}.
-The 95\% confidence intervals for our analyses suggest the $R_v$ for the 2009--2014 outbreaks is 0.92--1.19, and for the current 2013--2014 outbreak 1.82--2.13.
-The probability density of the $R_v$ estimates for each outbreak all include one.
-Of particular note is the ongoing outbreak (as of 12 June 2014), which has an $R_v$ well above one and thus we may expect this outbreak to persist if conditions remain the same.
-An important caveat to this outbreak analysis is that because this 2013--2014 outbreak is an ongoing outbreak, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
-
-These analyses also imply that the regular (approximately yearly) importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to maintain immunisation to high enough levels that measles does not become endemic.
-The measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
-This implies that the current outbreak may persist within the population for a substantial period, given its $R_v$ is approximately twice that of the 2011-2012 outbreak.
-A caveat to this and other $R_v$ estimates is that the 2013--2014 outbreak may include some sporadic cases and thus the true basic reproduction numbers may be lower than estimated.
-However, sub-clinical and underreporting may lower the estimate.
+our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014 but for which data were available until 12 June, a mean estimate of 3.66 (range 3.4--3.92).
+The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one.
+Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect this outbreak to persist if conditions remain the same.
+An important caveat to this outbreak analysis is that because this 2013--2014 outbreak that ended in August was ongoing when the data were collected, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
+However, this is not the only case for which the $R_v$ is greater than one.
+
+These analyses also imply that the regular importation of measles is an ongoing process.
+Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to increase immunisation to high enough levels that measles does not become endemic.
+The measles outbreaks in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+This implies that outbreaks may persist within the population for a substantial period, even if $R_v$ values are not much greater than one.
+A caveat to this and other $R_v$ estimates is that the outbreaks may include some sporadic cases misclassified, but also exclude some sporadic cases as yet not associated with another outbreak.
+Thus the true basic reproduction numbers may differ to those estimated.
+However, undetected cases may lower the estimate.
 The relative contributions of both to our estimates are currently unknown.
-
-If fewer cases occur during each outbreak, then this suggests that $R_v$ is lower, but that measles is introduced to New Zealand more frequently.
-The regularity of these outbreaks, however, implies that the approximately yearly importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, we include the effects of this in the benefit--cost section (\autoref{sub:cost_benefit}).
 
@@
 
-\begin{table}[htdp]
-
-\begin{center}
-
-\begin{tabular}{lrrrr}
-\hline
-DHB  & Size  & Na\"{\i}ve    &   Attack	& Vacc		\\
-\hline
-Auckland	&	436350	&	52010	&	31159	&	17920	\\
-Bay of Plenty	&	206000	&	20679	&	8437	&	4585	\\
-Canterbury	&	482180	&	51357	&	24695	&	13687	\\
-Capital and Coast	&	283700	&	32625	&	18403	&	10461	\\
-Counties Manukau	&	469300	&	55544	&	32903	&	18880	\\
-Hawke's Bay	&	151700	&	15602	&	6846	&	3751	\\
-Hutt Valley	&	138380	&	15198	&	7836	&	4388	\\
-Lakes	&	98196	&	10558	&	5192	&	2886	\\
-MidCentral	&	162560	&	17328	&	8348	&	4628	\\
-Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	\\
-Northland	&	151690	&	14921	&	5688	&	3071	\\
-South Canterbury	&	55620	&	5238	&	1678	&	893	\\
-Southern	&	297420	&	31607	&	15115	&	8371	\\
-Tairawhiti &	43650	&	4769	&	2431	&	1359	\\
-Taranaki	&	109750	&	11473	&	5262	&	2899	\\
-Waikato	&	359310	&	39402	&	20248	&	11331	\\
-Wairarapa	&	41112	&	3932	&	1346	&	720	\\
-Waitemata	&	525550	&	58350	&	30774	&	17291	\\
-West Coast	&	32151	&	3197	&	1265	&	685	\\
-Whanganui	&	60120	&	6075	&	2530	&	1378	\\
-\hline			
-TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 \\
+\begin{table}[htdp]\small
+\begin{center}
+\begin{tabular}{lrrrrr}
+\hline
+DHB  & Size  & Na\"{\i}ve    &   Attack	& Vacc	& Vacc (\%)	\\
+\hline
+Auckland	&	436350	&	52010	&	31159	&	17920	& 34\\
+Bay of Plenty	&	206000	&	20679	&	8437	&	4585	& 22\\
+Canterbury	&	482180	&	51357	&	24695	&	13687	& 27\\
+Capital and Coast	&	283700	&	32625	&	18403	&	10461 & 32	\\
+Counties Manukau	&	469300	&	55544	&	32903	&	18880	& 33\\
+Hawke's Bay	&	151700	&	15602	&	6846	&	3751 & 24\\
+Hutt Valley	&	138380	&	15198	&	7836	&	4388	& 29\\
+Lakes	&	98196	&	10558	&	5192	&	2886	& 27\\
+MidCentral	&	162560	&	17328	&	8348	&	4628	& 27\\
+Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	& 18\\
+Northland	&	151690	&	14921	&	5688	&	3071	& 21\\
+South Canterbury	&	55620	&	5238	&	1678	&	893	& 17\\
+Southern	&	297420	&	31607	&	15115	&	8371	& 26\\
+Tairawhiti &	43650	&	4769	&	2431	&	1359	& 28\\
+Taranaki	&	109750	&	11473	&	5262	&	2899	& 25\\
+Waikato	&	359310	&	39402	&	20248	&	11331	& 29\\
+Wairarapa	&	41112	&	3932	&	1346	&	720	& 18\\
+Waitemata	&	525550	&	58350	&	30774	&	17291	& 30\\
+West Coast	&	32151	&	3197	&	1265	&	685	& 21\\
+Whanganui	&	60120	&	6075	&	2530	&	1378	& 23\\
+\hline			
+TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 & 28\\
 \hline
@@
 An economic analysis was performed of the 614 measles cases reported in an 8-month period in Duisburg in the state of North Rhine-Wesphalia (NRW).
-In that study, they estimated the health-care provider costs to be approximately \euro 229,000, or \euro 373 per case.
+In that study, they estimated the health-care provider costs to be approximately \euro 229,000,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$426046} or \euro 373\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$694} per case.
 Approximately 78\% of these costs were associated with the 95 (15.5\%) of the cases that were hospitalised.
-The mean costs of the hospitalised patients was  \euro 1,877, including one patient with encephalitis at a cost of \euro 35,623.
-In addition to the health-care provider costs, additional costs of \euro 89,400 were incurred by the district public health office, the majority (\euro 85,000, 95.1\%) for personnel, \euro 2,300 (2.6\%) for vaccination, and \euro 2,100 (2.3\%) for serologic testing.
-Therefore the combined direct costs of these 612 cases amounted to \euro 318,400, or \euro 520 per case.
+The mean costs of the hospitalised patients was  \euro 1,877,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$3492} including one patient with encephalitis at a cost of \euro 35,623.\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$66275} In addition to the health-care provider costs, additional costs of \euro 89,400\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$166325} were incurred by the district public health office, the majority (\euro 85,000,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$158140} 95.1\%) for personnel, \euro 2,300\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$4279} (2.6\%) for vaccination, and \euro 2,100\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$3907} (2.3\%) for serologic testing.
+Therefore the combined direct costs of these 612 cases amounted to \euro 318,400,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$592372} or \euro 520\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$967} per case.
 To determine the total impact, it would be necessary to include the indirect losses associated with lost production of cases and care givers.
@@
 Direct costs for measles management in New Zealand for the 10-week period, January 1 -- March 9, 2014 are shown in  \autoref{table:direct}.
-The reported direct medical costs do not appear to include hospital medical costs, which are reported separately in \autoref{table:hosp}.
+The reported direct medical costs do not include hospital medical costs, which are reported separately in \autoref{table:hosp}.
 
@@
 
-The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital was \$550,024 (\autoref{table:hosp}).
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital from 2001--2014 (part year) was \$550,024 (\autoref{table:hosp}).
 The mean cost per case was \$1,877.
@@
 \begin{center}
-\includegraphics{draftfinalreport_v2-063}
+     \includegraphics[width=0.8\textwidth]{p.pdf}
 \end{center}
@@
 Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
+
 %latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
@@
 Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-Attack:		Number infected in DHB in an outbreak of measles $\left(\Pe\right)$;	
-Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}
+Attack:	  Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;
+Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed the small numbers of large outbreaks.
+The distribution of these outbreaks sizes can be seen in \autoref{fig:sim1} and \autoref{fig:sim}}
 
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Size}&\multicolumn{1}{c}{Naïve}&\multicolumn{1}{c}{Attack}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Proportion}&\multicolumn{1}{c}{Naïve.post.vaccination}&\multicolumn{1}{c}{Median.outbreak}&\multicolumn{1}{c}{Mean.outbreak}\tabularnewline
+\multicolumn{1}{c}{Age range}&\multicolumn{1}{c}{Proportion}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}\tabularnewline
 \hline
@@
 
-The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using \autoref{eq:prop} are shown in \autoref{fig:numvac}.
-
-\begin{figure}[width=1\textwidth]
-     \centering
-\includegraphics{draftfinalreport_v2-071}
-     \caption{The estimated numbers of the currently na\"{\i}ve New Zealand population requiring additional vaccination to reduce the $R_v$ to less than one (\autoref{section:epidemic_modelling}, \autoref{table:cost20} and \autoref{table:cost50})}
-     \label{fig:numvac}
-\end{figure}
+The numbers of susceptible people to vaccinate in New Zealand, assuming a homogeneously mixed population to achieve the 28\% currently na\"{i}ve catch up in vaccination numbers and using \autoref{eq:prop} are shown in \autoref{table:vacp}.
 
@@
      \centering
-\includegraphics{draftfinalreport_v2-073}
+     \includegraphics[width=0.9\textwidth]{sim.pdf}
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
@@
      \centering
-\includegraphics{draftfinalreport_v2-074}
+     \includegraphics[width=0.9\textwidth]{sim1.pdf}
   \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable20, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}\tabularnewline
-\hline
-Auckland&$17920$&$358400$&$26547468$&$55010965$&$5297$&$ 9942525$&$80393738$&$ 82$&$2115693$&$32.5$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7188324$&$14895456$&$1434$&$ 2692162$&$21768413$&$ 71$&$1831880$&$11.3$\tabularnewline
-Canterbury&$13687$&$273740$&$21040140$&$43598824$&$4198$&$ 7879928$&$63715888$&$ 62$&$1599670$&$34.0$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15679356$&$32490349$&$3129$&$ 5872213$&$47481818$&$ 96$&$2476908$&$17.7$\tabularnewline
-Counties Manukau&$18880$&$377600$&$28033356$&$58089983$&$5594$&$10499018$&$84893455$&$ 50$&$1290056$&$50.9$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5832792$&$12086558$&$1164$&$ 2184490$&$17663453$&$ 56$&$1444863$&$11.6$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6676272$&$13834395$&$1332$&$ 2500389$&$20217765$&$ 86$&$2218897$&$ 8.8$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4423584$&$ 9166434$&$ 883$&$ 1656715$&$13395946$&$ 62$&$1599670$&$ 8.1$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7112496$&$14738327$&$1419$&$ 2663763$&$21538783$&$ 75$&$1935085$&$10.6$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3758172$&$ 7787585$&$ 750$&$ 1407506$&$11380878$&$ 90$&$2322102$&$ 4.8$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4846176$&$10042118$&$ 967$&$ 1814984$&$14675682$&$ 70$&$1806079$&$ 7.9$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1429656$&$ 2962496$&$ 285$&$  535433$&$ 4329429$&$ 72$&$1857681$&$ 2.3$\tabularnewline
-Southern&$ 8371$&$167420$&$12877980$&$26685411$&$2570$&$ 4823045$&$38998407$&$102$&$2631715$&$13.9$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2071212$&$ 4291911$&$ 413$&$  775708$&$ 6272254$&$ 47$&$1212653$&$ 5.1$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4483224$&$ 9290019$&$ 895$&$ 1679052$&$13576554$&$ 68$&$1754477$&$ 7.5$\tabularnewline
-Waikato&$11331$&$226620$&$17251296$&$35747682$&$3442$&$ 6460934$&$52242126$&$ 95$&$2451107$&$19.5$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1146792$&$ 2376352$&$ 229$&$  429495$&$ 3472832$&$ 59$&$1522267$&$ 2.3$\tabularnewline
-Waitemata&$17291$&$345820$&$26219448$&$54331250$&$5232$&$ 9819676$&$79400395$&$ 70$&$1806079$&$36.9$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1077780$&$ 2233347$&$ 215$&$  403649$&$ 3263843$&$ 50$&$1290056$&$ 2.5$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2155560$&$ 4466695$&$ 430$&$  807298$&$ 6527686$&$ 58$&$1496465$&$ 4.3$\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Contacts}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}\tabularnewline
+\hline
+Auckland&$17920$&$358400$&$26546458.800$&$55634144.560$&$ 9942525.31$&$81779020.980$&$152792322.900$&$ 82$&$4020979.645$&$34.9$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7188050.737$&$15064195.820$&$ 2692162.33$&$22143509.100$&$ 41371957.640$&$ 71$&$3481579.936$&$11.6$\tabularnewline
+Canterbury&$13687$&$273740$&$21039340.160$&$44092724.410$&$ 7879927.55$&$64813791.300$&$121095234.500$&$ 62$&$3040252.902$&$36.5$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15678759.950$&$32858408.880$&$ 5872213.27$&$48299987.910$&$ 90241571.220$&$ 96$&$4707488.364$&$18.4$\tabularnewline
+Counties Manukau&$18880$&$377600$&$28032290.320$&$58748042.570$&$10499018.27$&$86356273.550$&$161344260.100$&$ 50$&$2451816.857$&$57.0$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5832570.267$&$12223478.090$&$ 2184490.14$&$17967815.970$&$ 33570276.400$&$ 56$&$2746034.879$&$11.9$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6676018.202$&$13991115.140$&$ 2500389.24$&$20566141.670$&$ 38424873.780$&$ 86$&$4217124.993$&$ 8.9$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4423415.838$&$ 9270274.353$&$ 1656715.28$&$13626774.830$&$ 25459666.240$&$ 62$&$3040252.902$&$ 8.2$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7112225.619$&$14905287.040$&$ 2663763.32$&$21909922.240$&$ 40935534.240$&$ 75$&$3677725.285$&$10.9$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3758029.134$&$ 7875805.118$&$ 1407505.99$&$11576984.550$&$ 21629928.310$&$ 90$&$4413270.342$&$ 4.8$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4845991.773$&$10155878.370$&$ 1814983.92$&$14928562.260$&$ 27891868.560$&$ 70$&$3432543.599$&$ 8.0$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1429601.652$&$ 2996055.540$&$  535433.02$&$ 4404030.849$&$  8228297.370$&$ 72$&$3530616.273$&$ 2.3$\tabularnewline
+Southern&$ 8371$&$167420$&$12877490.450$&$26987711.260$&$ 4823045.35$&$39670397.070$&$ 74118423.570$&$102$&$5001706.387$&$14.3$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2071133.263$&$ 4340531.000$&$  775707.79$&$ 6380333.130$&$ 11920733.560$&$ 47$&$2304707.845$&$ 5.1$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4483053.571$&$ 9395258.791$&$ 1679051.58$&$13810494.830$&$ 25802920.600$&$ 68$&$3334470.925$&$ 7.6$\tabularnewline
+Waikato&$11331$&$226620$&$17250640.190$&$36152641.580$&$ 6460934.32$&$53142322.180$&$ 99288775.420$&$ 95$&$4658452.027$&$20.3$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1146748.405$&$ 2403272.203$&$  429495.14$&$ 3532673.136$&$  6600290.978$&$ 59$&$2893143.891$&$ 2.3$\tabularnewline
+Waitemata&$17291$&$345820$&$26218451.270$&$54946730.150$&$ 9819675.66$&$80768560.990$&$150904423.900$&$ 70$&$3432543.599$&$39.9$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1077739.028$&$ 2258647.353$&$  403648.85$&$ 3320082.851$&$  6203096.647$&$ 50$&$2451816.857$&$ 2.5$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2155478.057$&$ 4517294.706$&$  807297.70$&$ 6640165.701$&$ 12406193.290$&$ 58$&$2844107.554$&$ 4.3$\tabularnewline
 \hline
@@
 \caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hospitalised} is number of hospitalisations saved, given the proportion of cases hospitalised 0.17; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable50, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline\hline
-\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hospitalised}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}\tabularnewline
-\hline
-Auckland&$17920$&$896000$&$26547468$&$55010965$&$5297$&$ 9942525$&$80393738$&$ 82$&$2115693$&$26.69$\tabularnewline
-Bay of Plenty&$ 4585$&$229250$&$ 7188324$&$14895456$&$1434$&$ 2692162$&$21768413$&$ 71$&$1831880$&$10.56$\tabularnewline
-Canterbury&$13687$&$684350$&$21040140$&$43598824$&$4198$&$ 7879928$&$63715888$&$ 62$&$1599670$&$27.90$\tabularnewline
-Capital and Coast&$10461$&$523050$&$15679356$&$32490349$&$3129$&$ 5872213$&$47481818$&$ 96$&$2476908$&$15.83$\tabularnewline
-Counties Manukau&$18880$&$944000$&$28033356$&$58089983$&$5594$&$10499018$&$84893455$&$ 50$&$1290056$&$38.00$\tabularnewline
-Hawke's Bay&$ 3751$&$187550$&$ 5832792$&$12086558$&$1164$&$ 2184490$&$17663453$&$ 56$&$1444863$&$10.82$\tabularnewline
-Hutt Valley&$ 4388$&$219400$&$ 6676272$&$13834395$&$1332$&$ 2500389$&$20217765$&$ 86$&$2218897$&$ 8.29$\tabularnewline
-Lakes&$ 2886$&$144300$&$ 4423584$&$ 9166434$&$ 883$&$ 1656715$&$13395946$&$ 62$&$1599670$&$ 7.68$\tabularnewline
-MidCentral&$ 4628$&$231400$&$ 7112496$&$14738327$&$1419$&$ 2663763$&$21538783$&$ 75$&$1935085$&$ 9.94$\tabularnewline
-Nelson Marlborough&$ 2356$&$117800$&$ 3758172$&$ 7787585$&$ 750$&$ 1407506$&$11380878$&$ 90$&$2322102$&$ 4.66$\tabularnewline
-Northland&$ 3071$&$153550$&$ 4846176$&$10042118$&$ 967$&$ 1814984$&$14675682$&$ 70$&$1806079$&$ 7.49$\tabularnewline
-South Canterbury&$  893$&$ 44650$&$ 1429656$&$ 2962496$&$ 285$&$  535433$&$ 4329429$&$ 72$&$1857681$&$ 2.28$\tabularnewline
-Southern&$ 8371$&$418550$&$12877980$&$26685411$&$2570$&$ 4823045$&$38998407$&$102$&$2631715$&$12.79$\tabularnewline
-Tairawhiti&$ 1359$&$ 67950$&$ 2071212$&$ 4291911$&$ 413$&$  775708$&$ 6272254$&$ 47$&$1212653$&$ 4.90$\tabularnewline
-Taranaki&$ 2899$&$144950$&$ 4483224$&$ 9290019$&$ 895$&$ 1679052$&$13576554$&$ 68$&$1754477$&$ 7.15$\tabularnewline
-Waikato&$11331$&$566550$&$17251296$&$35747682$&$3442$&$ 6460934$&$52242126$&$ 95$&$2451107$&$17.31$\tabularnewline
-Wairarapa&$  720$&$ 36000$&$ 1146792$&$ 2376352$&$ 229$&$  429495$&$ 3472832$&$ 59$&$1522267$&$ 2.23$\tabularnewline
-Waitemata&$17291$&$864550$&$26219448$&$54331250$&$5232$&$ 9819676$&$79400395$&$ 70$&$1806079$&$29.73$\tabularnewline
-West Coast&$  685$&$ 34250$&$ 1077780$&$ 2233347$&$ 215$&$  403649$&$ 3263843$&$ 50$&$1290056$&$ 2.46$\tabularnewline
-Whanganui&$ 1378$&$ 68900$&$ 2155560$&$ 4466695$&$ 430$&$  807298$&$ 6527686$&$ 58$&$1496465$&$ 4.17$\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Vacc costs}&\multicolumn{1}{c}{Wages saved}&\multicolumn{1}{c}{Manage saved}&\multicolumn{1}{c}{Hosp saved}&\multicolumn{1}{c}{Contacts}&\multicolumn{1}{c}{Costs saved}&\multicolumn{1}{c}{Outbreak}&\multicolumn{1}{c}{OB costs}&\multicolumn{1}{c}{B/C}\tabularnewline
+\hline
+Auckland&$17920$&$896000$&$26546458.800$&$55634144.560$&$81779020.980$&$ 9942525.31$&$152792322.900$&$ 82$&$4020979.645$&$31.1$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7188050.737$&$15064195.820$&$22143509.100$&$ 2692162.33$&$ 41371957.640$&$ 71$&$3481579.936$&$11.1$\tabularnewline
+Canterbury&$13687$&$684350$&$21039340.160$&$44092724.410$&$64813791.300$&$ 7879927.55$&$121095234.500$&$ 62$&$3040252.902$&$32.5$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15678759.950$&$32858408.880$&$48299987.910$&$ 5872213.27$&$ 90241571.220$&$ 96$&$4707488.364$&$17.3$\tabularnewline
+Counties Manukau&$18880$&$944000$&$28032290.320$&$58748042.570$&$86356273.550$&$10499018.27$&$161344260.100$&$ 50$&$2451816.857$&$47.5$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5832570.267$&$12223478.090$&$17967815.970$&$ 2184490.14$&$ 33570276.400$&$ 56$&$2746034.879$&$11.4$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6676018.202$&$13991115.140$&$20566141.670$&$ 2500389.24$&$ 38424873.780$&$ 86$&$4217124.993$&$ 8.7$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4423415.838$&$ 9270274.353$&$13626774.830$&$ 1656715.28$&$ 25459666.240$&$ 62$&$3040252.902$&$ 8.0$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7112225.619$&$14905287.040$&$21909922.240$&$ 2663763.32$&$ 40935534.240$&$ 75$&$3677725.285$&$10.5$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3758029.134$&$ 7875805.118$&$11576984.550$&$ 1407505.99$&$ 21629928.310$&$ 90$&$4413270.342$&$ 4.8$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4845991.773$&$10155878.370$&$14928562.260$&$ 1814983.92$&$ 27891868.560$&$ 70$&$3432543.599$&$ 7.8$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1429601.652$&$ 2996055.540$&$ 4404030.849$&$  535433.02$&$  8228297.370$&$ 72$&$3530616.273$&$ 2.3$\tabularnewline
+Southern&$ 8371$&$418550$&$12877490.450$&$26987711.260$&$39670397.070$&$ 4823045.35$&$ 74118423.570$&$102$&$5001706.387$&$13.7$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2071133.263$&$ 4340531.000$&$ 6380333.130$&$  775707.79$&$ 11920733.560$&$ 47$&$2304707.845$&$ 5.0$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4483053.571$&$ 9395258.791$&$13810494.830$&$ 1679051.58$&$ 25802920.600$&$ 68$&$3334470.925$&$ 7.4$\tabularnewline
+Waikato&$11331$&$566550$&$17250640.190$&$36152641.580$&$53142322.180$&$ 6460934.32$&$ 99288775.420$&$ 95$&$4658452.027$&$19.0$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1146748.405$&$ 2403272.203$&$ 3532673.136$&$  429495.14$&$  6600290.978$&$ 59$&$2893143.891$&$ 2.3$\tabularnewline
+Waitemata&$17291$&$864550$&$26218451.270$&$54946730.150$&$80768560.990$&$ 9819675.66$&$150904423.900$&$ 70$&$3432543.599$&$35.1$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1077739.028$&$ 2258647.353$&$ 3320082.851$&$  403648.85$&$  6203096.647$&$ 50$&$2451816.857$&$ 2.5$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2155478.057$&$ 4517294.706$&$ 6640165.701$&$  807297.70$&$ 12406193.290$&$ 58$&$2844107.554$&$ 4.3$\tabularnewline
 \hline
@@
 Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children.
+However, there are a number of different vaccination schemes that will lead to sufficient coverage and these are given in \autoref{table:cohort_vac}.
+The appropriate cover could be reached by catch up vaccination of all school aged children.
+
+\begin{table}[htdp]\small
+\begin{center}
+\begin{tabular}{lrrrr}
+\hline
+Age range (yrs) & Proportion of na\"{\i}ve population\\
+\hline
+2--5 & 0.04\\
+5--11 & 0.15\\
+5--16 & 0.27\\
+5--18 & 0.31\\
+11--16 & 0.14\\
+11--18 & 0.18\\
+0--29 & 0.72\\
+0--44 & 0.89\\
+>8 & 0.8 \\ 
+\hline
+\end{tabular}
+\caption{Proportion of the currently na\"{\i}ve New Zealand population that would be vaccinated if all na\"{\i}ve people in each age range were vaccinated.
+The estimated additional proportion of the population requiring vaccination is 28\%}
+\end{center}
+\label{table:cohort_vac}
+\end{table}%
 
@@
 
-In other outbreaks, the average cost per measles case was estimated to be US\$254, US\$276, and US\$307 for Canada, the Netherlands, and the UK, respectively  \citep{carabin2}.
+In other outbreaks, the average cost per measles case was estimated to be US\$254\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$571}, US\$276\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$621}, and US\$307\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$691} for Canada, the Netherlands, and the UK, respectively  \citep{carabin2}.
 These values are substantially lower than our estimates, however, it is noticable that these estimates are also from more than a decade ago.
-The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.
-In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak \citep{dayan5}.
-They estimated direct costs per case to be less than US\$500.
-However, the combined direct costs of \euro 520 per case in Germany \citep{siedler6} suggests costs may be higher.
+The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$211990} In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak \citep{dayan5}.
+They estimated direct costs per case to be less than US\$500.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$774} However, the combined direct costs of \euro 520\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$967} per case in Germany \citep{siedler6} suggests costs may be higher.
 Thus, our estimates of the additional costs, including indirect costs and wages lost may be appropriate.
 
-The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059 and US\$78,448, respectively in Minnesota \citep{prouty1}.
-In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864 per person.
-In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.
-The cost of a case of measles was estimated to range from \$71 (no complications and no hospitalisation) to \$29,556 (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083 (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
+The annual cost for long-term care of people with moderate of severe mental retardation over a period of 50 years is estimated at US\$31,059\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$38790} and US\$78,448,\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$97974} respectively in Minnesota \citep{prouty1}.
+In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$221612} per person.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$221612} The cost of a case of measles was estimated to range from \$71\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$138} (no complications and no hospitalisation) to \$29,556\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$57524} (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$2401874} (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
 Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see \autoref{table:zhou}) \citep{zhou4}.

### 2015-03-20 dhbcensus.R

@@ -42,7 +42,7 @@ colnames(naivedhb)<-c("Northland","Waitemata","Auckland","Counties Manukau",
 for (i in 1:20){
   pdf(paste("dhb", i, ".pdf", sep = ""))
   plot(naivedhb[,i],ylim=c(0,8000),main=colnames(naivedhb)[i],xlab="Age",ylab="Numbers",
-       pch=16,cex=1.1)
+       pch=16,cex=1.1,cex.lab=1.1,cex.main=1.1,cex.axis=1.1)
   legend("topright",legend=c(c("Total naive", round(sum(naivedhb[,i]))),
                              c("Total population", signif(popdhb[i],5)),
                              c("Percent naive",round(sum(naivedhb[,i])/popdhb[i],3)*100)),

### 2015-03-20 revised_draft.R

@@ -48,10 +48,10 @@ dim(test)
 pdf(paste("case_age_dist.pdf"), width=7, height=5)
 par(mfrow=c(1,1))
 par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=2)
-hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=2)
+par(cex.axis=1)
+hist(time$AgeInYears,col="grey",xlab="Age in years",main="",breaks=90,include.lowest=TRUE,right=F,ylab="Frequency",cex.lab=1)
 hist(test$AgeInYears,col="black",breaks=90,include.lowest=TRUE,right=F,add=T)
-legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=2)
+legend("topright",c("1997-2014","2007-2014"),col=c("grey","black"),pch=15,bty="n",cex=1)
 dev.off()
 ## plot cases / age class
 caseyr<-aggregate( DiseaseName ~ AgeInYears, 
@@ -445,17 +445,17 @@ ggplot(tpsub1, aes(x=ordert,y= cases, fill=Ethnicity)) +
 dev.off()
 ##
 pdf(paste("percap_merge.pdf"), width=10, height=5)
-ggplot(tp, aes(x=merge, y=cases, fill=Ethnicity)) + 
+ggplot(tp, aes(x=merge, y=perCap, fill=Ethnicity)) + 
   geom_bar(stat="identity", position="dodge")+
   # theme(text=element_text(size=45))+
-  ylab("cases")+xlab("")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
+  ylab("Per capita per 10000")+xlab("")+theme(axis.text.x = element_text(angle = 90, hjust = 0))
 dev.off()
 #
 pdf(paste("percap_ordert.pdf"), width=10, height=5)
-ggplot(tpsub, aes(x=ordert, y=cases, fill=Ethnicity)) +
+ggplot(tpsub, aes(x=ordert, y=perCap, fill=Ethnicity)) +
   geom_bar(stat="identity", position="dodge")+
   # theme(text=element_text(size=45))+
-  ylab("cases")+xlab("")+theme(axis.text.x = element_text(angle = 90, vjust=0,hjust = 0))
[…]

### 2015-03-21 draftfinalreport_v2.Rnw
Tidying and revisions
Eg relevel for the regression

@@
 #hist(tpsub$cases,xlab="Cases",main='Histogram of cases per category',col="darkgrey",breaks=20)
+tpsub$Ethnicity<- relevel(tpsub$Ethnicity, "European")
 
@@
 resan<-xtable(anova(model3,test="F"))
-#resan<-xtable(anova(model2,test="F"))
-# print(resan,floating = FALSE)
-
-#xtb<-xtable(summary(model3))
-xtb<-xtable(summary(model2))
-# print(xtb,floating = FALSE)
+xtb<-xtable(summary(model3))
 
@@
 round((sum(naive$Naive[8:100])/sum(naive$Naive)),2))
-names(propvCohort)<-c("2-5","5-11","5-16","5-18","11-16","11-18","0-29","0-44",">8")
+yr_range<-c("2-5","5-11","5-16","5-18","11-16","11-18","0-29","0-44",">8")
+propvCohort<-cbind(yr_range,propvCohort)
+propvCohort<-as.data.frame(propvCohort)
 write.table(propvCohort,"vaccine_proportions_Cohort.txt",sep=",",row.names=T,col.names=F)
-
 propsv = NULL
@@
 hosp<-read.csv("hospital.csv",header=T)
-resan<-xtable(anova(model3,test="F"))
-xtb<-xtable(summary(model3))
 
@@
 
-\begin{table}[hbtp]\scriptsize
-<<results=tex,echo=FALSE>>=# This chunk uses "results=tex"
-latex(ttyr, file="",table.env=FALSE,rowname=NULL)
-@
-\caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
-\label{table:none}
-\end{table}
+% \begin{table}[hbtp]\scriptsize
+% <<results=tex,echo=FALSE>>=# This chunk uses "results=tex"
+% latex(ttyr, file="",table.env=FALSE,rowname=NULL)
+% @
+% \caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
+% \label{table:none}
+% \end{table}
 
@@
 
-\begin{table}[hbtp]\scriptsize
-\caption{Per capita measles rates}
+% \begin{table}[hbtp]\scriptsize
+% \caption{Per capita measles rates}
+% <<results=tex,echo=FALSE>>=
+% latex(PerCap, file="", table.env=FALSE,rowname=NULL)
+% @
+% \label{table:percap_rates}
+% \end{table}
+
+\begin{table}[hbtp]\footnotesize
+\caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
 <<results=tex,echo=FALSE>>=
-latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:percap_rates}
-\end{table}
-
-\begin{table}[hbtp]\footnotesize
-\caption{Proportions of the naive population the would be covered if vaccinated by age groups}
-<<results=tex,echo=FALSE>>=
-latex(propvCohort, file="", table.env=FALSE)
+colnames(propvCohort) <- c("Age range", "Proportion")
+latex(propvCohort, file="", table.env=FALSE,rowname=NULL)
 @
@@
 <<results=tex,echo=FALSE>>=
+colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
 latex(PerCap, file="", table.env=FALSE,rowname=NULL)
@@
 
-\begin{figure}
+\begin{sidewaysfigure}
 \begin{center}
@@
 \label{fig:naive_au}
-\end{figure}
+\end{sidewaysfigure}
 
@@
 
-\begin{sidewaysfigure}[H]
-      \centering
+\begin{sidewaysfigure}
+\begin{center}
      \includegraphics[width=0.9\textwidth]{cases_by_dhb_2009_2014.pdf}
+          \end{center}
 \caption{Measles cases by District Health Board (DHB) from 2009 to 2014}
@@
 
-\begin{sidewaysfigure}[H]
-     \centering
+\begin{sidewaysfigure}
+\begin{center}
      \includegraphics[width=0.9\textwidth]{outbreaks_2009_2014.pdf}
+     \end{center}
      \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014}
@@
 
-\begin{sidewaysfigure}[H]
-     \centering
+\begin{sidewaysfigure}
+\begin{center}
      \includegraphics[width=0.9\textwidth]{averageR0_all_ob.pdf}
+     \end{center}
      \caption{Estimates of $R_v$ for the outbreaks each year, as classified by outbreaks in \autoref{fig:outbreaks}. Sporadic cases are excluded.}
@@
 <<results=tex,echo=FALSE>>=
-colnames(vacp) <- c("Age range", "Proportion")
 latex(vacp, file="", table.env=FALSE,rowname=NULL)

### 2015-03-21 draftfinalreport_v2.tex
Tidying and revisions
Eg relevel for the regression

@@
 
-\begin{table}[hbtp]\scriptsize
-%latex.default(ttyr, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lllrl}
-\hline\hline
-\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Cases}&\multicolumn{1}{c}{merge}\tabularnewline
-\hline
-1-3&0-1&Asian&$  6$&1-3 0-1 Asian\tabularnewline
-4-7&0-1&Asian&$ 10$&4-7 0-1 Asian\tabularnewline
-8-10&0-1&Asian&$  3$&8-10 0-1 Asian\tabularnewline
-1-3&2-4&Asian&$  1$&1-3 2-4 Asian\tabularnewline
-1-3&5-17&Asian&$  6$&1-3 5-17 Asian\tabularnewline
-4-7&5-17&Asian&$  8$&4-7 5-17 Asian\tabularnewline
-8-10&5-17&Asian&$  2$&8-10 5-17 Asian\tabularnewline
-1-3&18-24&Asian&$  2$&1-3 18-24 Asian\tabularnewline
-4-7&18-24&Asian&$  3$&4-7 18-24 Asian\tabularnewline
-8-10&18-24&Asian&$  3$&8-10 18-24 Asian\tabularnewline
-1-3&25+&Asian&$  4$&1-3 25+ Asian\tabularnewline
-4-7&25+&Asian&$ 11$&4-7 25+ Asian\tabularnewline
-8-10&25+&Asian&$  8$&8-10 25+ Asian\tabularnewline
-1-3&0-1&European&$ 40$&1-3 0-1 European\tabularnewline
-4-7&0-1&European&$ 58$&4-7 0-1 European\tabularnewline
-8-10&0-1&European&$ 24$&8-10 0-1 European\tabularnewline
-1-3&2-4&European&$ 37$&1-3 2-4 European\tabularnewline
-4-7&2-4&European&$ 22$&4-7 2-4 European\tabularnewline
-8-10&2-4&European&$  6$&8-10 2-4 European\tabularnewline
-1-3&5-17&European&$160$&1-3 5-17 European\tabularnewline
-4-7&5-17&European&$118$&4-7 5-17 European\tabularnewline
-8-10&5-17&European&$ 40$&8-10 5-17 European\tabularnewline
-1-3&18-24&European&$ 24$&1-3 18-24 European\tabularnewline
-4-7&18-24&European&$ 26$&4-7 18-24 European\tabularnewline
-8-10&18-24&European&$ 20$&8-10 18-24 European\tabularnewline
-1-3&25+&European&$ 54$&1-3 25+ European\tabularnewline
-4-7&25+&European&$ 46$&4-7 25+ European\tabularnewline
-8-10&25+&European&$ 29$&8-10 25+ European\tabularnewline
-1-3&0-1&Maori&$  9$&1-3 0-1 Maori\tabularnewline
-4-7&0-1&Maori&$ 13$&4-7 0-1 Maori\tabularnewline
-8-10&0-1&Maori&$ 32$&8-10 0-1 Maori\tabularnewline
-1-3&2-4&Maori&$  4$&1-3 2-4 Maori\tabularnewline
-4-7&2-4&Maori&$ 12$&4-7 2-4 Maori\tabularnewline
-8-10&2-4&Maori&$ 10$&8-10 2-4 Maori\tabularnewline
-1-3&5-17&Maori&$  8$&1-3 5-17 Maori\tabularnewline
-4-7&5-17&Maori&$ 33$&4-7 5-17 Maori\tabularnewline
-8-10&5-17&Maori&$ 74$&8-10 5-17 Maori\tabularnewline
-1-3&18-24&Maori&$  3$&1-3 18-24 Maori\tabularnewline
-4-7&18-24&Maori&$  2$&4-7 18-24 Maori\tabularnewline
-8-10&18-24&Maori&$  8$&8-10 18-24 Maori\tabularnewline
-1-3&25+&Maori&$  1$&1-3 25+ Maori\tabularnewline
-4-7&25+&Maori&$  3$&4-7 25+ Maori\tabularnewline
-8-10&25+&Maori&$  4$&8-10 25+ Maori\tabularnewline
-4-7&0-1&MLA&$  2$&4-7 0-1 MLA\tabularnewline
-8-10&0-1&MLA&$  1$&8-10 0-1 MLA\tabularnewline
-8-10&2-4&MLA&$  1$&8-10 2-4 MLA\tabularnewline
-8-10&5-17&MLA&$  1$&8-10 5-17 MLA\tabularnewline
-8-10&18-24&MLA&$  2$&8-10 18-24 MLA\tabularnewline
-4-7&25+&MLA&$  5$&4-7 25+ MLA\tabularnewline
-8-10&25+&MLA&$  1$&8-10 25+ MLA\tabularnewline
-1-3&0-1&Pacific&$  1$&1-3 0-1 Pacific\tabularnewline
-4-7&0-1&Pacific&$  6$&4-7 0-1 Pacific\tabularnewline
-8-10&0-1&Pacific&$ 48$&8-10 0-1 Pacific\tabularnewline
-4-7&2-4&Pacific&$  1$&4-7 2-4 Pacific\tabularnewline
-8-10&2-4&Pacific&$ 10$&8-10 2-4 Pacific\tabularnewline
-1-3&5-17&Pacific&$  2$&1-3 5-17 Pacific\tabularnewline
-4-7&5-17&Pacific&$  4$&4-7 5-17 Pacific\tabularnewline
-8-10&5-17&Pacific&$ 21$&8-10 5-17 Pacific\tabularnewline
-4-7&18-24&Pacific&$  4$&4-7 18-24 Pacific\tabularnewline
-8-10&18-24&Pacific&$  5$&8-10 18-24 Pacific\tabularnewline
-1-3&25+&Pacific&$  1$&1-3 25+ Pacific\tabularnewline
-4-7&25+&Pacific&$  3$&4-7 25+ Pacific\tabularnewline
-8-10&25+&Pacific&$  9$&8-10 25+ Pacific\tabularnewline
-\hline
-\end{tabular}\end{center}\caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
-\label{table:none}
-\end{table}
+% \begin{table}[hbtp]\scriptsize
+% <<results=tex,echo=FALSE>>=# This chunk uses "results=tex"
+% latex(ttyr, file="",table.env=FALSE,rowname=NULL)
+% @
+% \caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
+% \label{table:none}
+% \end{table}
 
@@
 
-\begin{table}[hbtp]\scriptsize
-\caption{Per capita measles rates}
+% \begin{table}[hbtp]\scriptsize
+% \caption{Per capita measles rates}
+% <<results=tex,echo=FALSE>>=
+% latex(PerCap, file="", table.env=FALSE,rowname=NULL)
+% @
+% \label{table:percap_rates}
+% \end{table}
+
+\begin{table}[hbtp]\footnotesize
+\caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
+%latex.default(propvCohort, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{ll}
+\hline\hline
+\multicolumn{1}{c}{Age range}&\multicolumn{1}{c}{Proportion}\tabularnewline
+\hline
+2-5&0.04\tabularnewline
+5-11&0.15\tabularnewline
+5-16&0.27\tabularnewline
+5-18&0.31\tabularnewline
+11-16&0.14\tabularnewline
+11-18&0.18\tabularnewline
+0-29&0.72\tabularnewline
+0-44&0.89\tabularnewline
+\textgreater8&0.8\tabularnewline
+\hline
+\end{tabular}\end{center}\label{table:vaccine_cohort}
+\end{table}
+
+The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap}, \autoref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
+Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
+
+\begin{table}[hbtp] \scriptsize
+\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
 %latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:percap_rates}
-\end{table}
-
-\begin{table}[hbtp]\footnotesize
-\caption{Proportions of the naive population the would be covered if vaccinated by age groups}
-%latex.default(propvCohort, file = "", table.env = FALSE)%
-\begin{center}
-\begin{tabular}{lr}
-\hline\hline
-\multicolumn{1}{l}{propvCohort}&\multicolumn{1}{c}{}\tabularnewline
-\hline
-2-5&$0.04$\tabularnewline
-5-11&$0.15$\tabularnewline
-5-16&$0.27$\tabularnewline
-5-18&$0.31$\tabularnewline
-11-16&$0.14$\tabularnewline
-11-18&$0.18$\tabularnewline
-0-29&$0.72$\tabularnewline
-0-44&$0.89$\tabularnewline
-\textgreater 8&$0.80$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:vaccine_cohort}
-\end{table}
-
-The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap}, \autoref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
-Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
-
-\begin{table}[hbtp] \scriptsize
-\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
-%latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{lllrrr}
-\hline\hline
-\multicolumn{1}{c}{NZDep}&\multicolumn{1}{c}{Age}&\multicolumn{1}{c}{Ethnicity}&\multicolumn{1}{c}{Population}&\multicolumn{1}{c}{Cases 1}&\multicolumn{1}{c}{Per capita}\tabularnewline
-\hline
-1-3&0-1&Asian&$  2269$&$  6$&$26.4434$\tabularnewline
-4-7&0-1&Asian&$  3748$&$ 10$&$26.6809$\tabularnewline
-8-10&0-1&Asian&$  2583$&$  3$&$11.6144$\tabularnewline
-1-3&2-4&Asian&$  3403$&$  1$&$ 2.9386$\tabularnewline
-4-7&2-4&Asian&$  5623$&$  0$&$ 0.0000$\tabularnewline
-8-10&2-4&Asian&$  3874$&$  0$&$ 0.0000$\tabularnewline
-1-3&5-17&Asian&$ 20625$&$  6$&$ 2.9091$\tabularnewline
-4-7&5-17&Asian&$ 29721$&$  8$&$ 2.6917$\tabularnewline
-8-10&5-17&Asian&$ 16777$&$  2$&$ 1.1921$\tabularnewline
-1-3&18-24&Asian&$ 12057$&$  2$&$ 1.6588$\tabularnewline
-4-7&18-24&Asian&$ 23904$&$  3$&$ 1.2550$\tabularnewline
-8-10&18-24&Asian&$ 21063$&$  3$&$ 1.4243$\tabularnewline
-1-3&25+&Asian&$ 54258$&$  4$&$ 0.7372$\tabularnewline
-4-7&25+&Asian&$ 86055$&$ 11$&$ 1.2783$\tabularnewline
-8-10&25+&Asian&$ 54759$&$  8$&$ 1.4609$\tabularnewline
-1-3&0-1&European&$ 23961$&$ 40$&$16.6938$\tabularnewline
-4-7&0-1&European&$ 28051$&$ 58$&$20.6766$\tabularnewline
-8-10&0-1&European&$ 16866$&$ 24$&$14.2298$\tabularnewline
-1-3&2-4&European&$ 35942$&$ 37$&$10.2944$\tabularnewline
-4-7&2-4&European&$ 42076$&$ 22$&$ 5.2286$\tabularnewline
-8-10&2-4&European&$ 25299$&$  6$&$ 2.3716$\tabularnewline
-1-3&5-17&European&$180875$&$160$&$ 8.8459$\tabularnewline
-4-7&5-17&European&$194100$&$118$&$ 6.0793$\tabularnewline
-8-10&5-17&European&$106731$&$ 40$&$ 3.7477$\tabularnewline
-1-3&18-24&European&$ 63051$&$ 24$&$ 3.8064$\tabularnewline
-4-7&18-24&European&$ 94422$&$ 26$&$ 2.7536$\tabularnewline
-8-10&18-24&European&$ 68016$&$ 20$&$ 2.9405$\tabularnewline
-1-3&25+&European&$619698$&$ 54$&$ 0.8714$\tabularnewline
-4-7&25+&European&$734550$&$ 46$&$ 0.6262$\tabularnewline
-8-10&25+&European&$371985$&$ 29$&$ 0.7796$\tabularnewline
-1-3&0-1&Maori&$  3265$&$  9$&$27.5651$\tabularnewline
-4-7&0-1&Maori&$  8378$&$ 13$&$15.5168$\tabularnewline
-8-10&0-1&Maori&$ 15094$&$ 32$&$21.2005$\tabularnewline
-1-3&2-4&Maori&$  4897$&$  4$&$ 8.1683$\tabularnewline
-4-7&2-4&Maori&$ 12568$&$ 12$&$ 9.5481$\tabularnewline
-8-10&2-4&Maori&$ 22642$&$ 10$&$ 4.4166$\tabularnewline
-1-3&5-17&Maori&$ 21750$&$  8$&$ 3.6782$\tabularnewline
-4-7&5-17&Maori&$ 53963$&$ 33$&$ 6.1153$\tabularnewline
-8-10&5-17&Maori&$ 94758$&$ 74$&$ 7.8094$\tabularnewline
-1-3&18-24&Maori&$  7329$&$  3$&$ 4.0933$\tabularnewline
-4-7&18-24&Maori&$ 20862$&$  2$&$ 0.9587$\tabularnewline
-8-10&18-24&Maori&$ 35664$&$  8$&$ 2.2432$\tabularnewline
-1-3&25+&Maori&$ 35934$&$  1$&$ 0.2783$\tabularnewline
-4-7&25+&Maori&$ 87192$&$  3$&$ 0.3441$\tabularnewline
-8-10&25+&Maori&$140820$&$  4$&$ 0.2841$\tabularnewline
-1-3&0-1&MLA&$   273$&$  0$&$ 0.0000$\tabularnewline
-4-7&0-1&MLA&$   519$&$  2$&$38.5356$\tabularnewline
-8-10&0-1&MLA&$   553$&$  1$&$18.0832$\tabularnewline
-1-3&2-4&MLA&$   410$&$  0$&$ 0.0000$\tabularnewline
-4-7&2-4&MLA&$   778$&$  0$&$ 0.0000$\tabularnewline
-8-10&2-4&MLA&$   830$&$  1$&$12.0482$\tabularnewline
-1-3&5-17&MLA&$  1832$&$  0$&$ 0.0000$\tabularnewline
-4-7&5-17&MLA&$  3121$&$  0$&$ 0.0000$\tabularnewline
-8-10&5-17&MLA&$  3250$&$  1$&$ 3.0769$\tabularnewline
-1-3&18-24&MLA&$   906$&$  0$&$ 0.0000$\tabularnewline
-4-7&18-24&MLA&$  1806$&$  0$&$ 0.0000$\tabularnewline
-8-10&18-24&MLA&$  2076$&$  2$&$ 9.6339$\tabularnewline
-1-3&25+&MLA&$  4569$&$  0$&$ 0.0000$\tabularnewline
-4-7&25+&MLA&$  7452$&$  5$&$ 6.7096$\tabularnewline
-8-10&25+&MLA&$  6342$&$  1$&$ 1.5768$\tabularnewline
-1-3&0-1&Pacific&$   643$&$  1$&$15.5521$\tabularnewline
-4-7&0-1&Pacific&$  2113$&$  6$&$28.3956$\tabularnewline
-8-10&0-1&Pacific&$  7389$&$ 48$&$64.9614$\tabularnewline
-1-3&2-4&Pacific&$   964$&$  0$&$ 0.0000$\tabularnewline
-4-7&2-4&Pacific&$  3169$&$  1$&$ 3.1556$\tabularnewline
-8-10&2-4&Pacific&$ 11084$&$ 10$&$ 9.0220$\tabularnewline
-1-3&5-17&Pacific&$  4212$&$  2$&$ 4.7483$\tabularnewline
-4-7&5-17&Pacific&$ 13419$&$  4$&$ 2.9808$\tabularnewline
-8-10&5-17&Pacific&$ 47165$&$ 21$&$ 4.4525$\tabularnewline
-1-3&18-24&Pacific&$  1707$&$  0$&$ 0.0000$\tabularnewline
-4-7&18-24&Pacific&$  6174$&$  4$&$ 6.4788$\tabularnewline
-8-10&18-24&Pacific&$ 18189$&$  5$&$ 2.7489$\tabularnewline
-1-3&25+&Pacific&$  8580$&$  1$&$ 1.1655$\tabularnewline
-4-7&25+&Pacific&$ 26688$&$  3$&$ 1.1241$\tabularnewline
-8-10&25+&Pacific&$ 74757$&$  9$&$ 1.2039$\tabularnewline
-\hline
 \end{tabular}\end{center}\label{table:percap}
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Fri Mar 20 22:48:21 2015
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Sun Mar 22 01:29:41 2015
 \begin{tabular}{lrrrrrr}
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-3 package
-% Fri Mar 20 22:48:21 2015
+% latex table generated in R 3.0.2 by xtable 1.7-1 package
+% Sun Mar 22 01:29:41 2015
 \begin{tabular}{rrrrr}
@@
   \hline
-(Intercept) & -5.9180 & 0.3002 & -19.71 & 0.0000 \\ 
-  Age2-4 & -3.3499 & 1.3172 & -2.54 & 0.0152 \\ 
-  Age5-17 & -2.2407 & 0.4357 & -5.14 & 0.0000 \\ 
-  Age18-24 & -2.7392 & 0.5412 & -5.06 & 0.0000 \\ 
-  Age25+ & -2.9353 & 0.3980 & -7.37 & 0.0000 \\ 
-  EthnicityEuropean & -0.2464 & 0.3168 & -0.78 & 0.4415 \\ 
-  EthnicityMaori & -0.0383 & 0.3433 & -0.11 & 0.9118 \\ 
-  EthnicityPacific & 0.9736 & 0.3439 & 2.83 & 0.0074 \\ 
+(Intercept) & -6.1644 & 0.1267 & -48.65 & 0.0000 \\ 
+  Age2-4 & -1.0351 & 0.1972 & -5.25 & 0.0000 \\ 
+  Age5-17 & -0.9954 & 0.1368 & -7.28 & 0.0000 \\ 
+  Age18-24 & -1.7204 & 0.1926 & -8.93 & 0.0000 \\ 
+  Age25+ & -3.1697 & 0.1622 & -19.55 & 0.0000 \\ 
+  EthnicityAsian & 0.2464 & 0.3168 & 0.78 & 0.4415 \\ 
+  EthnicityMaori & 0.2081 & 0.2121 & 0.98 & 0.3327 \\ 
+  EthnicityPacific & 1.2200 & 0.2134 & 5.72 & 0.0000 \\ 
   NZDep4-7 & -0.2609 & 0.0948 & -2.75 & 0.0090 \\ 
   NZDep8-10 & -0.3043 & 0.1040 & -2.93 & 0.0058 \\ 
-  Age2-4:EthnicityEuropean & 2.3148 & 1.3319 & 1.74 & 0.0903 \\ 
-  Age5-17:EthnicityEuropean & 1.2453 & 0.4566 & 2.73 & 0.0096 \\ 
-  Age18-24:EthnicityEuropean & 1.0187 & 0.5743 & 1.77 & 0.0841 \\ 
-  Age25+:EthnicityEuropean & -0.2344 & 0.4298 & -0.55 & 0.5887 \\ 
-  Age2-4:EthnicityMaori & 2.2135 & 1.3524 & 1.64 & 0.1099 \\ 
-  Age5-17:EthnicityMaori & 1.1421 & 0.4844 & 2.36 & 0.0236 \\ 
-  Age18-24:EthnicityMaori & 0.4465 & 0.6710 & 0.67 & 0.5098 \\ 
-  Age25+:EthnicityMaori & -1.2693 & 0.6285 & -2.02 & 0.0505 \\ 
-  Age2-4:EthnicityPacific & 1.3350 & 1.3838 & 0.96 & 0.3408 \\ 
-  Age5-17:EthnicityPacific & -0.3256 & 0.5299 & -0.61 & 0.5426 \\ 
-  Age18-24:EthnicityPacific & -0.0167 & 0.7113 & -0.02 & 0.9814 \\ 
-  Age25+:EthnicityPacific & -0.8973 & 0.5614 & -1.60 & 0.1183 \\ 
+  Age2-4:EthnicityAsian & -2.3148 & 1.3319 & -1.74 & 0.0903 \\ 
+  Age5-17:EthnicityAsian & -1.2453 & 0.4566 & -2.73 & 0.0096 \\ 
+  Age18-24:EthnicityAsian & -1.0187 & 0.5743 & -1.77 & 0.0841 \\ 
+  Age25+:EthnicityAsian & 0.2344 & 0.4298 & 0.55 & 0.5887 \\ 
+  Age2-4:EthnicityMaori & -0.1013 & 0.3644 & -0.28 & 0.7826 \\ 
+  Age5-17:EthnicityMaori & -0.1032 & 0.2521 & -0.41 & 0.6846 \\ 
+  Age18-24:EthnicityMaori & -0.5723 & 0.4409 & -1.30 & 0.2022 \\ 
+  Age25+:EthnicityMaori & -1.0349 & 0.5127 & -2.02 & 0.0506 \\ 
+  Age2-4:EthnicityPacific & -0.9798 & 0.4676 & -2.10 & 0.0429 \\ 
+  Age5-17:EthnicityPacific & -1.5709 & 0.3312 & -4.74 & 0.0000 \\ 
+  Age18-24:EthnicityPacific & -1.0354 & 0.5003 & -2.07 & 0.0453 \\ 
+  Age25+:EthnicityPacific & -0.6629 & 0.4279 & -1.55 & 0.1296 \\ 
    \hline
@@
 \begin{center}
-\includegraphics{draftfinalreport_v2-017}
+\includegraphics{draftfinalreport_v2-015}
 \end{center}
@@
 \hline\hline
-\multicolumn{1}{c}{Age range}&\multicolumn{1}{c}{Proportion}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}&\multicolumn{1}{c}{NA}\tabularnewline
+\multicolumn{1}{c}{DHB}&\multicolumn{1}{c}{Size}&\multicolumn{1}{c}{Naïve}&\multicolumn{1}{c}{Attack}&\multicolumn{1}{c}{Vacc}&\multicolumn{1}{c}{Proportion}&\multicolumn{1}{c}{Naïve.post.vaccination}&\multicolumn{1}{c}{Median.outbreak}&\multicolumn{1}{c}{Mean.outbreak}\tabularnewline
 \hline
@@
 \hline
-Auckland&$17920$&$358400$&$26546458.800$&$55634144.560$&$ 9942525.31$&$81779020.980$&$152792322.900$&$ 82$&$4020979.645$&$34.9$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7188050.737$&$15064195.820$&$ 2692162.33$&$22143509.100$&$ 41371957.640$&$ 71$&$3481579.936$&$11.6$\tabularnewline
-Canterbury&$13687$&$273740$&$21039340.160$&$44092724.410$&$ 7879927.55$&$64813791.300$&$121095234.500$&$ 62$&$3040252.902$&$36.5$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15678759.950$&$32858408.880$&$ 5872213.27$&$48299987.910$&$ 90241571.220$&$ 96$&$4707488.364$&$18.4$\tabularnewline
-Counties Manukau&$18880$&$377600$&$28032290.320$&$58748042.570$&$10499018.27$&$86356273.550$&$161344260.100$&$ 50$&$2451816.857$&$57.0$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5832570.267$&$12223478.090$&$ 2184490.14$&$17967815.970$&$ 33570276.400$&$ 56$&$2746034.879$&$11.9$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6676018.202$&$13991115.140$&$ 2500389.24$&$20566141.670$&$ 38424873.780$&$ 86$&$4217124.993$&$ 8.9$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4423415.838$&$ 9270274.353$&$ 1656715.28$&$13626774.830$&$ 25459666.240$&$ 62$&$3040252.902$&$ 8.2$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7112225.619$&$14905287.040$&$ 2663763.32$&$21909922.240$&$ 40935534.240$&$ 75$&$3677725.285$&$10.9$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3758029.134$&$ 7875805.118$&$ 1407505.99$&$11576984.550$&$ 21629928.310$&$ 90$&$4413270.342$&$ 4.8$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4845991.773$&$10155878.370$&$ 1814983.92$&$14928562.260$&$ 27891868.560$&$ 70$&$3432543.599$&$ 8.0$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1429601.652$&$ 2996055.540$&$  535433.02$&$ 4404030.849$&$  8228297.370$&$ 72$&$3530616.273$&$ 2.3$\tabularnewline
-Southern&$ 8371$&$167420$&$12877490.450$&$26987711.260$&$ 4823045.35$&$39670397.070$&$ 74118423.570$&$102$&$5001706.387$&$14.3$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2071133.263$&$ 4340531.000$&$  775707.79$&$ 6380333.130$&$ 11920733.560$&$ 47$&$2304707.845$&$ 5.1$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4483053.571$&$ 9395258.791$&$ 1679051.58$&$13810494.830$&$ 25802920.600$&$ 68$&$3334470.925$&$ 7.6$\tabularnewline
-Waikato&$11331$&$226620$&$17250640.190$&$36152641.580$&$ 6460934.32$&$53142322.180$&$ 99288775.420$&$ 95$&$4658452.027$&$20.3$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1146748.405$&$ 2403272.203$&$  429495.14$&$ 3532673.136$&$  6600290.978$&$ 59$&$2893143.891$&$ 2.3$\tabularnewline
-Waitemata&$17291$&$345820$&$26218451.270$&$54946730.150$&$ 9819675.66$&$80768560.990$&$150904423.900$&$ 70$&$3432543.599$&$39.9$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1077739.028$&$ 2258647.353$&$  403648.85$&$ 3320082.851$&$  6203096.647$&$ 50$&$2451816.857$&$ 2.5$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2155478.057$&$ 4517294.706$&$  807297.70$&$ 6640165.701$&$ 12406193.290$&$ 58$&$2844107.554$&$ 4.3$\tabularnewline
+Auckland&$17920$&$358400$&$26546459$&$55634145$&$ 9942525$&$81779021$&$152792323$&$ 82$&$4020980$&$34.9$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7188051$&$15064196$&$ 2692162$&$22143509$&$ 41371958$&$ 71$&$3481580$&$11.6$\tabularnewline
+Canterbury&$13687$&$273740$&$21039340$&$44092724$&$ 7879928$&$64813791$&$121095235$&$ 62$&$3040253$&$36.5$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15678760$&$32858409$&$ 5872213$&$48299988$&$ 90241571$&$ 96$&$4707488$&$18.4$\tabularnewline
+Counties Manukau&$18880$&$377600$&$28032290$&$58748043$&$10499018$&$86356274$&$161344260$&$ 50$&$2451817$&$57.0$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5832570$&$12223478$&$ 2184490$&$17967816$&$ 33570276$&$ 56$&$2746035$&$11.9$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6676018$&$13991115$&$ 2500389$&$20566142$&$ 38424874$&$ 86$&$4217125$&$ 8.9$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4423416$&$ 9270274$&$ 1656715$&$13626775$&$ 25459666$&$ 62$&$3040253$&$ 8.2$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7112226$&$14905287$&$ 2663763$&$21909922$&$ 40935534$&$ 75$&$3677725$&$10.9$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3758029$&$ 7875805$&$ 1407506$&$11576985$&$ 21629928$&$ 90$&$4413270$&$ 4.8$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4845992$&$10155878$&$ 1814984$&$14928562$&$ 27891869$&$ 70$&$3432544$&$ 8.0$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1429602$&$ 2996056$&$  535433$&$ 4404031$&$  8228297$&$ 72$&$3530616$&$ 2.3$\tabularnewline
+Southern&$ 8371$&$167420$&$12877490$&$26987711$&$ 4823045$&$39670397$&$ 74118424$&$102$&$5001706$&$14.3$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2071133$&$ 4340531$&$  775708$&$ 6380333$&$ 11920734$&$ 47$&$2304708$&$ 5.1$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4483054$&$ 9395259$&$ 1679052$&$13810495$&$ 25802921$&$ 68$&$3334471$&$ 7.6$\tabularnewline
+Waikato&$11331$&$226620$&$17250640$&$36152642$&$ 6460934$&$53142322$&$ 99288775$&$ 95$&$4658452$&$20.3$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1146748$&$ 2403272$&$  429495$&$ 3532673$&$  6600291$&$ 59$&$2893144$&$ 2.3$\tabularnewline
+Waitemata&$17291$&$345820$&$26218451$&$54946730$&$ 9819676$&$80768561$&$150904424$&$ 70$&$3432544$&$39.9$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1077739$&$ 2258647$&$  403649$&$ 3320083$&$  6203097$&$ 50$&$2451817$&$ 2.5$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2155478$&$ 4517295$&$  807298$&$ 6640166$&$ 12406193$&$ 58$&$2844108$&$ 4.3$\tabularnewline
 \hline
@@
 \hline
-Auckland&$17920$&$896000$&$26546458.800$&$55634144.560$&$81779020.980$&$ 9942525.31$&$152792322.900$&$ 82$&$4020979.645$&$31.1$\tabularnewline
-Bay of Plenty&$ 4585$&$229250$&$ 7188050.737$&$15064195.820$&$22143509.100$&$ 2692162.33$&$ 41371957.640$&$ 71$&$3481579.936$&$11.1$\tabularnewline
-Canterbury&$13687$&$684350$&$21039340.160$&$44092724.410$&$64813791.300$&$ 7879927.55$&$121095234.500$&$ 62$&$3040252.902$&$32.5$\tabularnewline
-Capital and Coast&$10461$&$523050$&$15678759.950$&$32858408.880$&$48299987.910$&$ 5872213.27$&$ 90241571.220$&$ 96$&$4707488.364$&$17.3$\tabularnewline
-Counties Manukau&$18880$&$944000$&$28032290.320$&$58748042.570$&$86356273.550$&$10499018.27$&$161344260.100$&$ 50$&$2451816.857$&$47.5$\tabularnewline
-Hawke's Bay&$ 3751$&$187550$&$ 5832570.267$&$12223478.090$&$17967815.970$&$ 2184490.14$&$ 33570276.400$&$ 56$&$2746034.879$&$11.4$\tabularnewline
-Hutt Valley&$ 4388$&$219400$&$ 6676018.202$&$13991115.140$&$20566141.670$&$ 2500389.24$&$ 38424873.780$&$ 86$&$4217124.993$&$ 8.7$\tabularnewline
-Lakes&$ 2886$&$144300$&$ 4423415.838$&$ 9270274.353$&$13626774.830$&$ 1656715.28$&$ 25459666.240$&$ 62$&$3040252.902$&$ 8.0$\tabularnewline
-MidCentral&$ 4628$&$231400$&$ 7112225.619$&$14905287.040$&$21909922.240$&$ 2663763.32$&$ 40935534.240$&$ 75$&$3677725.285$&$10.5$\tabularnewline
-Nelson Marlborough&$ 2356$&$117800$&$ 3758029.134$&$ 7875805.118$&$11576984.550$&$ 1407505.99$&$ 21629928.310$&$ 90$&$4413270.342$&$ 4.8$\tabularnewline
-Northland&$ 3071$&$153550$&$ 4845991.773$&$10155878.370$&$14928562.260$&$ 1814983.92$&$ 27891868.560$&$ 70$&$3432543.599$&$ 7.8$\tabularnewline
-South Canterbury&$  893$&$ 44650$&$ 1429601.652$&$ 2996055.540$&$ 4404030.849$&$  535433.02$&$  8228297.370$&$ 72$&$3530616.273$&$ 2.3$\tabularnewline
-Southern&$ 8371$&$418550$&$12877490.450$&$26987711.260$&$39670397.070$&$ 4823045.35$&$ 74118423.570$&$102$&$5001706.387$&$13.7$\tabularnewline
-Tairawhiti&$ 1359$&$ 67950$&$ 2071133.263$&$ 4340531.000$&$ 6380333.130$&$  775707.79$&$ 11920733.560$&$ 47$&$2304707.845$&$ 5.0$\tabularnewline
-Taranaki&$ 2899$&$144950$&$ 4483053.571$&$ 9395258.791$&$13810494.830$&$ 1679051.58$&$ 25802920.600$&$ 68$&$3334470.925$&$ 7.4$\tabularnewline
-Waikato&$11331$&$566550$&$17250640.190$&$36152641.580$&$53142322.180$&$ 6460934.32$&$ 99288775.420$&$ 95$&$4658452.027$&$19.0$\tabularnewline
-Wairarapa&$  720$&$ 36000$&$ 1146748.405$&$ 2403272.203$&$ 3532673.136$&$  429495.14$&$  6600290.978$&$ 59$&$2893143.891$&$ 2.3$\tabularnewline
-Waitemata&$17291$&$864550$&$26218451.270$&$54946730.150$&$80768560.990$&$ 9819675.66$&$150904423.900$&$ 70$&$3432543.599$&$35.1$\tabularnewline
-West Coast&$  685$&$ 34250$&$ 1077739.028$&$ 2258647.353$&$ 3320082.851$&$  403648.85$&$  6203096.647$&$ 50$&$2451816.857$&$ 2.5$\tabularnewline
-Whanganui&$ 1378$&$ 68900$&$ 2155478.057$&$ 4517294.706$&$ 6640165.701$&$  807297.70$&$ 12406193.290$&$ 58$&$2844107.554$&$ 4.3$\tabularnewline
+Auckland&$17920$&$896000$&$26546459$&$55634145$&$81779021$&$ 9942525$&$152792323$&$ 82$&$4020980$&$31.1$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7188051$&$15064196$&$22143509$&$ 2692162$&$ 41371958$&$ 71$&$3481580$&$11.1$\tabularnewline
+Canterbury&$13687$&$684350$&$21039340$&$44092724$&$64813791$&$ 7879928$&$121095235$&$ 62$&$3040253$&$32.5$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15678760$&$32858409$&$48299988$&$ 5872213$&$ 90241571$&$ 96$&$4707488$&$17.3$\tabularnewline
+Counties Manukau&$18880$&$944000$&$28032290$&$58748043$&$86356274$&$10499018$&$161344260$&$ 50$&$2451817$&$47.5$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5832570$&$12223478$&$17967816$&$ 2184490$&$ 33570276$&$ 56$&$2746035$&$11.4$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6676018$&$13991115$&$20566142$&$ 2500389$&$ 38424874$&$ 86$&$4217125$&$ 8.7$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4423416$&$ 9270274$&$13626775$&$ 1656715$&$ 25459666$&$ 62$&$3040253$&$ 8.0$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7112226$&$14905287$&$21909922$&$ 2663763$&$ 40935534$&$ 75$&$3677725$&$10.5$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3758029$&$ 7875805$&$11576985$&$ 1407506$&$ 21629928$&$ 90$&$4413270$&$ 4.8$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4845992$&$10155878$&$14928562$&$ 1814984$&$ 27891869$&$ 70$&$3432544$&$ 7.8$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1429602$&$ 2996056$&$ 4404031$&$  535433$&$  8228297$&$ 72$&$3530616$&$ 2.3$\tabularnewline
+Southern&$ 8371$&$418550$&$12877490$&$26987711$&$39670397$&$ 4823045$&$ 74118424$&$102$&$5001706$&$13.7$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2071133$&$ 4340531$&$ 6380333$&$  775708$&$ 11920734$&$ 47$&$2304708$&$ 5.0$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4483054$&$ 9395259$&$13810495$&$ 1679052$&$ 25802921$&$ 68$&$3334471$&$ 7.4$\tabularnewline
+Waikato&$11331$&$566550$&$17250640$&$36152642$&$53142322$&$ 6460934$&$ 99288775$&$ 95$&$4658452$&$19.0$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1146748$&$ 2403272$&$ 3532673$&$  429495$&$  6600291$&$ 59$&$2893144$&$ 2.3$\tabularnewline
+Waitemata&$17291$&$864550$&$26218451$&$54946730$&$80768561$&$ 9819676$&$150904424$&$ 70$&$3432544$&$35.1$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1077739$&$ 2258647$&$ 3320083$&$  403649$&$  6203097$&$ 50$&$2451817$&$ 2.5$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2155478$&$ 4517295$&$ 6640166$&$  807298$&$ 12406193$&$ 58$&$2844108$&$ 4.3$\tabularnewline
 \hline

### 2015-03-21 dhbcensus.R

@@ -33,7 +33,7 @@ df<-df[,2:21]/5
 popdhb<-colSums(df)
 naivedhb<-df*(1-impop[1:90])
 head(naivedhb)
-matplot(naivedhb,type="l")
+#matplot(naivedhb,type="l")
 colnames(naivedhb)<-c("Northland","Waitemata","Auckland","Counties Manukau",
                       "Waikato","Lakes","Bay of Plenty","Tairawhiti","Taranaki",
                       "Hawke's Bay","Whanganui","Midcentral","Hutt","Capital and Coast",
@@ -41,8 +41,9 @@ colnames(naivedhb)<-c("Northland","Waitemata","Auckland","Counties Manukau",
                       "Southern")
 for (i in 1:20){
   pdf(paste("dhb", i, ".pdf", sep = ""))
-  plot(naivedhb[,i],ylim=c(0,8000),main=colnames(naivedhb)[i],xlab="Age",ylab="Numbers",
-       pch=16,cex=1.1,cex.lab=1.1,cex.main=1.1,cex.axis=1.1)
+  barplot(naivedhb[,i],ylim=c(0,8000),main=colnames(naivedhb)[i],xlab="Age",ylab="Numbers",
+       #type="l",
+       cex=1.1,cex.lab=1.1,cex.main=1.1,cex.axis=1.1)
   legend("topright",legend=c(c("Total naive", round(sum(naivedhb[,i]))),
                              c("Total population", signif(popdhb[i],5)),
                              c("Percent naive",round(sum(naivedhb[,i])/popdhb[i],3)*100)),
@@ -56,7 +57,7 @@ library(maptools)
 
 # read in shapefile for AU's
 dhb <- readShapeSpatial("nz-district-health-boards-2012.shp")
-plot(dhb,xlim=c(170,180))
+#plot(dhb,xlim=c(170,180))
 
 head(naivedhb)
 # library(plyr)

### 2015-03-22 draftfinalreport_v2.Rnw
Hopefully final draft...

@@
 \item The regression analyses suggest that age is a particularly strong risk factor for measles.
-People of European ethnicity, however, make up the majority of cases.
+People of European ethnicity make up the majority of cases.
 However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis.
-In particular, Pacific people are at greater risk, as are the more wealthy (NZDep 1--3).
+Increased wealth (NZDep 1--3) was a risk factor.
+Pacific people are at greater risk per capita.
 However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
@@
 However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
-\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed 1 and the 2013--2014 outbreaks, as analysed from data until June, is well above 1 (mean estimate 3.66, range 3.4--3.92 for the last outbreak).
+\item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed one, including 2013--2014 outbreaks.
+As analysed from data until June, the last outbreak to end in 2014 had an $R_v$ well above 1 (mean estimate 3.66, range 3.4--3.92).
 The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92.
-This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population.
-\item The cost of the 2013--2014 measles outbreaks is estimated to be approximately \$1,332,539 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
-\item The mean wage loss per measles case is estimated to be approximately \$852.
+This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
+\item The cost of 2014 measles cases is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
+\item The mean wage loss per measles case is estimated to be approximately \$839.
 \item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
@@
 \item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170.4 per day.
-\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed \$8,214 per vaccine before the costs exceed the benefits.
-\item Measles introductions to New Zealand with a mean size of 61 cases per year, median of 2 cases, but rarely a peak size of up to many thousands, may occur due to importation, despite $R_v$ being below one and the epidemic predicted to die out following additional vaccinations.
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$8,214 per vaccine before the costs exceed the benefits.
+\item Measles introductions to New Zealand with a median size of 2 cases per year were predicted by simulation following catch up vaccination to ensure $R_v$ is < 1.
+Thus, typically individual cases and secondary cases will be the expectation.
+However, the mean of 61 cases was predicted, because peak outbreak sizes of tens, hundreds, and very rarely up to many thousands are predicted following importation, despite $R_v$ being one and the outbreak predicted to die out following additional vaccinations.
+Thus, increased vaccination beyond the 28\% of the currently 11\% na\'{\i}ve population required may be useful.
 \end{itemize}
@@
 With the 10 NZDep classes, Prioritized Ethnicities, and the age classes above, the numerous combinations of variables led us to have 250 categories.
-Because for measles cases the very young appear to be disproportionately affected (\autoref{fig:ageinyears}), we split the 0--5 age category into two classes, 0--2 and 3--5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
+Because for measles cases the very young appear to be disproportionately affected (\autoref{fig:ageinyears}), we split the 0--5 age category into additional classes.
+Following discussion with the Ministry, these were 0--1, 2--4, and 5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
+The 5 year olds were added to the 6--17 year old category.
+No breakdown of 6--17 year category was performed as constant birth rates for the period could not be assumed.
 
@@
 
-totdpdf<-read.csv("mapdata_tot.csv",header=T)
+totdpdf<-read.csv("2013mapdata_tot.csv",header=T)
 # output top countries
@@
 
-nzdpdf<-read.csv("mapdata_nzers.csv",header=T)
+nzdpdf<-read.csv("2013mapdata_nzers.csv",header=T)
 # output top countries
@@
 
-imdpdf<-read.csv("mapdata_immigration.csv",header=T)
+imdpdf<-read.csv("2013mapdata_immigration.csv",header=T)
 # output top countries
@@
 This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to two: NZDep 1--5 (least deprived) and NZDep 6--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}).
+Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to three: NZDep 1--3 (least deprived), NZDep 4--7, and NZDep 8--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}) following discussion with the Ministry regarding the most appropriate categories.
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
@@
 The Prioritized Ethnicities for cases are:  European; Maori; Pacific Peoples; Asian; Middle Eastern/Latin American/African (MLA); Other Ethnicity; Residual Categories.
-For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity (see \emph{None}, \autoref{table:none}).
-
-% \begin{table}[hbtp]\scriptsize
-% <<results=tex,echo=FALSE>>=# This chunk uses "results=tex"
-% latex(ttyr, file="",table.env=FALSE,rowname=NULL)
-% @
-% \caption{Absolute number of measles cases with no ethnicity recorded for specific age and socio-economic deprivation (NZDep) categories from 2007-2014.}
-% \label{table:none}
-% \end{table}
+For the analyses in this report only the first five are used, as these categories cover the overwhelming number of cases, with only 1.9\% (22/1137) of cases having no Prioritized Ethnicity.
 
@@
 \begin{center}
+     \includegraphics[width=0.9\textwidth]{case_ordert.pdf}
+\end{center}
+\caption{Measles cases from 2007--2014 (\autoref{table:percap}). The x-axis shows NZDep:Age Range:Prioritized ethnicity. Per capita values are in \autoref{fig:percap_all}}
+\label{fig:cases_all}
+\end{sidewaysfigure}
+
+\begin{sidewaysfigure}
+\begin{center}
      \includegraphics[width=0.9\textwidth]{percap_ordert.pdf}
 \end{center}
-\caption{test}
+\caption{Per capita measles values from 2007--2014 (\autoref{table:percap}). The x-axis shows NZDep:Age Range:Prioritized ethnicity. Absolute cases are in \autoref{fig:cases_all}}
 \label{fig:percap_all}
@@
 
-\begin{sidewaysfigure}
-\begin{center}
-     \includegraphics[width=0.9\textwidth]{case_ordert.pdf}
-\end{center}
-\caption{test}
-\label{fig:cases_all}
-\end{sidewaysfigure}
-
-Two way interactions are shown in \autoref{sec:app} (\autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).
-
-% \begin{table}[hbtp]\scriptsize
-% \caption{Per capita measles rates}
-% <<results=tex,echo=FALSE>>=
-% latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-% @
-% \label{table:percap_rates}
-% \end{table}
+The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap_all}, \autoref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), creating issues for estimating the denominator data for this group (University of Otago, personal communication).
+Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
+This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
+The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}.
+Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).
+
+\begin{table}[hbtp] \scriptsize
+<<results=tex,echo=FALSE>>=
+colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
+latex(PerCap, file="", table.env=FALSE,rowname=NULL)
+@
+\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
+\label{table:percap}
+\end{table}
 
 \begin{table}[hbtp]\footnotesize
-\caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
 \label{table:vaccine_cohort}
-\end{table}
-
-The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap}, \autoref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
-Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
-This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
-
-\begin{table}[hbtp] \scriptsize
-\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
-<<results=tex,echo=FALSE>>=
-colnames(PerCap)<-c("NZDep","Age","Ethnicity","Population","Cases","Per capita")
-latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-@
-\label{table:percap}
 \end{table}
@@
 
+The cases and per capita measles values are in \autoref{fig:cases_all} and \autoref{fig:percap_all}.
 The distribution of the measles cases per category used in the regression analyses are in \autoref{fig:histcasecat}.
+The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}.
+The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
+A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.
+
 \begin{figure}
@@
 
-The predicted values from the regression model plotted against the reported cases are shown in \autoref{fig:predict}, and the residuals are shown in \autoref{fig:resid}.
 \begin{figure}
@@
 \end{center}
-\caption{Regression model (\autoref{eq:reg}) predictions plotted against the case data (\autoref{table:percap})}
+\caption{Regression model predictions plotted against the case data (\autoref{table:percap}), estimated using \autoref{eq:reg} and a quasipoisson error structure}
 \label{fig:predict}
@@
 \end{center}
-\caption{Histogram of residuals from the regression model (\autoref{eq:reg})}
+\caption{Histogram of residuals from the regression model (\autoref{eq:reg}) and a quasipoisson error structure}
 \label{fig:resid}
 \end{figure}
-
-The significance of the different predictor variables can be seen in the ANOVA results (\autoref{table:anova}).
 
@@
 
-A summary of the regression model (\autoref{eq:reg}) with the individual effects and the statistical support for the estimated coefficients can be seen in \autoref{table:regression_riskfactors}.
-
 \vspace{5mm} %5mm vertical space
@@
 @
-\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+\caption{Summary of the regression model results, estimated using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-2014}
 \label{table:regression_riskfactors}
@@
 
-Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case.
-Indeed, all age categories are significantly less likely to be measles cases compared to 0-2 year olds, and the likelihood generally decreases with age (see also \autoref{fig:ageinyears}).
-
-The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:ageinyears}), however, people of Pacific origin are also over-represented as measles cases on a per capita basis ($\beta$ = 1.16, standard error (SE) = 0.2, p--value < 0.0001).
-Those people in NZDep levels 6-10 are under-represented ($\beta$ = -0.21, SE = 0.085, p--value = 0.02), and there are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 6--17 age classes.
+Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case (\autoref{table:regression_riskfactors}).
+Indeed, all age categories are significantly less likely to be measles cases compared to 0--1 year olds, and the likelihood generally decreases with age (\autoref{table:regression_riskfactors}, \autoref{fig:cases_all}, \autoref{fig:percap_all}, see also \autoref{fig:ageinyears}).
+
+The majority of measles cases are among people of European ancestry (\autoref{table:percap} and \autoref{fig:cases_all}).
+Indeed, the majority of cases were in more wealthy European 5--17 year olds (\autoref{fig:cases_all}).
+However, per capita rates for this large age class were lower than in the very young (0--1 year olds, \autoref{fig:percap_all}).
+People of Pacific origin are over-represented as measles cases on a per capita basis ($\beta$ = 1.22, standard error (SE) = 0.21, p--value < 0.0001; \autoref{table:regression_riskfactors}), and in particular the 0--1 year old less wealthy (NZDep 8--10) (\autoref{fig:percap_all}.
+Note however that generally those people in NZDep levels 4--10 are less represented on a per capita basis (NZDep 4--7 $\beta$ = -0.26, SE = 0.09, p--value 0.009; NZDep 8--10 $\beta$ = -0.3, SE = 0.1, p--value == 0.006; \autoref{table:regression_riskfactors}), and in European and Maori ethnicities there is a trend in the opposite way than Pacific peoples, with the more wealthy more likely to be cases (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
+
+There are some other age:ethnicity classes that are significantly less represented in the data compared to Europeans in those ages classes, particularly in the 5--17 age classes.
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}, \autoref{table:sero}, and \autoref{sub:popim}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}).
-Interestingly the regression results suggest risk of measles in the 6--17 year age category was greater for Europeans and Maori.
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap},\autoref{table:sero}, and \autoref{sub:popim}).
+Interestingly the regression results suggest risk of measles in the 5--17 year age category was greater for Europeans and Maori (\autoref{table:percap}, \autoref{table:regression_riskfactors}, \autoref{fig:percap_all}).
 
@@
 \end{center}
-\caption{Vaccination coverage (0--5 years) and serosurvey estimates of immunity ($>5$ years) among different age classes used in these analyses}
+\caption{Vaccination coverage (0--5 years) and serosurvey estimates of immunity ($>5$ years) among different age classes used in these analyses, as provided by the Ministry of Health.
+Note these values are from vaccination records and serosurveys.
+Vaccine efficacy for 0--5 year olds was 99\%, 6--13 year olds 96\%, and for \leq 14 equivocal serological results were considered non-immune}
 \label{table:sero}
@@
 
-% ## insert figures here
-
 \begin{figure}
@@
 \end{center}
-\caption{Age of vaccination of vaccinated measles cases, 2007--2014}
+\caption{Age of vaccination of vaccinated measles cases, 2007--2014. Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
 \label{fig:vaccstat}
@@
 
-\subsection{Measles importation risk methods}
-\label{sub:imp_risk}
-For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand.
-We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
-Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
-We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from.
-We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
-We present the data from 2012 and 2013 because 2012 had the most complete WHO measles data, whereas 2013 is most recent, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{dob_vacc_unvac.pdf}
+\end{center}
+\caption{Year of birth and vaccination status of measles cases, 2007--2014}
+\label{fig:yearandvac}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{dob_vacc_dose.pdf}
+\end{center}
+\caption{Year of birth and vaccination status of measles cases, 2007--2014}
+\label{fig:yearandvacf}
+\end{figure}
 
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{naive_au.pdf}
-\end{center}
-\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator}
+     \includegraphics[width=0.9\textwidth]{nir_census.pdf}
+\end{center}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register (NIR) level data and national census data as the denominator.
+Note no values should be greater than 1 in the frequency distribution (upper left) or 100\% on the map, demonstrating a lack of concordance between NIR and census data.
+The full distribution is shown in \autoref{fig:naive_au_prop}}
 \label{fig:naive_au}
@@
 \end{center}
-\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator.
+Note no values should be greater than 1 (100\% vaccine cover), and the maximum discordance in > 1600\%.}
 \label{fig:naive_au_prop}
@@
 Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (\autoref{fig:naive_map}).
-The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures in the \autoref{sec:app}:
+The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures in the appendix (\autoref{sec:app}):
 \begin{itemize}[noitemsep,nolistsep]
@@
 
-Plotting the lower level data available from the NIR database highlights the issue of surveillance data requiring appropriate denominator data, as these analyses show that there is a lack of correspondence between the national census data and the NIR data (\autoref{fig:nir})
+Plotting the lower level data available from the NIR database highlights the issue of surveillance data requiring appropriate denominator data, as these analyses show that there is a lack of correspondence between the national census data and the NIR data (\autoref{fig:naive_au}, \autoref{fig:naive_au_prop})
 
@@
 
+\subsection{Measles importation risk methods}
+\label{sub:imp_risk}
+For our measles importation risk analyses, we use arrivals data from New Zealand immigration and New Zealander travel destination data by country and year (\href{http://www.immigration.govt.nz/}{www.immigration.govt.nz}), to measure human movement to and from New Zealand.
+We collated country population size, measles incidence and measles vaccination cover from the WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
+WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday.
+Note the immigration figures use all immigration of foreign nationals, coming for whatever purpose, and includes non-New Zealanders resident in New Zealand, but not yet holders of New Zealand passports.
+We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from.
+We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
+We use 0--19 year old data for the immigration figures, as this is the only category available and it was determined this was more appropriate than using all age classes as > 19 year olds are believe to be immune to measles more or less globally.
+These data were not available for New Zealand travellers, so we use all travel.
+We present the data from 2013 because this year was the most recent with the most complete WHO measles data, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+Data for 2013 are incomplete, but substantially better than 2014.
+
 \subsection{Measles importation risk results}
@@
 \begin{table}
-\caption{Highest measles incidence per million (2012)}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Highest measles incidence per million (2013)}
 \label{table:incidence12}
@@
 \begin{table}
-\caption{Lowest national measles vaccine cover (\%), 2012}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Lowest national measles vaccine cover (\%), 2013}
 \label{table:cover12}
@@
 \begin{table}
-\caption{Non-New Zealander travel and immigration numbers (2012)}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Non-New Zealander travel and immigration numbers, 0--19 year olds, 2013}
 \label{table:imtravel12}
@@
 \begin{table}
-\caption{New Zealander travel numbers by destination (2012)}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{New Zealander travel numbers by destination, 2013}
 \label{table:nztravel12}
@@
 \begin{table}
-\caption{Total New Zealand traveller numbers by country (New Zealand nationals and immigrants, 2012)}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Total New Zealand traveller numbers by country (New Zealand nationals and 0--19 year old immigrants, 2013)}
 \label{table:immigration12}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2013, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Risk of measles importation to New Zealand in 2013, estimated by 0--19 year old international and all New Zealander traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
 \label{table:risk12}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2013, estimated by all traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
 \label{table:nzrisk12}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Risk of measles importation to New Zealand due to non-New Zealander 0--19 year old travel and immigration in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
 \label{table:imrisk12}
@@
 
-%%%%%%%%%%%%%%%%%%
-
 \begin{table}
-\caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation.
-incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of travellers entering New Zealand from the country}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Measles and travel data from 2013 for the top 10 countries identified as high risk for measles importation.
+incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of 0--19 year old travellers entering New Zealand from the country}
 \label{table:toprisk}
@@
 Globally, analyses of the most complete and recent year of data available, 2013, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}).
-Immigration by non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia, United Kingdom, China, and the United States, as shown in Table \autoref{table:imtravel12} and \autoref{fig:immigration12}.
-However, travel by New Zealanders is dominated by that to Australia (\autoref{table:nztravel12} and \autoref{fig:nztravel12}).
-Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\autoref{table:immigration12} and \autoref{fig:immigration12}.
-Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2013 and immigration numbers in 2013, suggest that those countries with greatest international travel, Australia, UK, China and South East Asian countries, pose the greatest risk to New Zealand for measles importation.
+Immigration by 0--19 year old non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia \autoref{table:imtravel12} and \autoref{fig:immigration12}.
+A number of other countries then follow, including United Kingdom, China, Japan and the United States.
+Travel by New Zealanders is also dominated by that to Australia (\autoref{table:nztravel12} and \autoref{fig:nztravel12}).
+Together, these mean that the greatest travel location for New Zealanders and immigration origin is from Australia (\autoref{table:immigration12} and \autoref{fig:immigration12}).
+However, many South Pacific nations also appear in the most visited country list by New Zealanders.
+Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2013 and immigration numbers of 0--19 year olds in 2013, suggest that those countries with greatest international travel, Australia, UK, China and South East Asian countries, pose the greatest risk to New Zealand for measles importation.
 Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand.
+Note the South Pacific island do not appear to pose a greater risk, despite large numbers of travellers of New Zealand origin, perhaps because of low measles incidence in these nations.
 These data are shown in \autoref{table:risk12}.
 The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in \autoref{table:nzrisk12} and \autoref{table:imrisk12} and \autoref{fig:nzrisk12} and \autoref{fig:imrisk12}.
-
 The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in \autoref{table:toprisk}.
-WHO measles coverage is the percent of children that receive one dose of measles vaccine by their second birthday.
 
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{incidence12.pdf}
-\end{center}
-\caption{Measles incidence per million, 2012. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
+     \includegraphics[width=0.9\textwidth]{2013totnp2.pdf}
+\end{center}
+\caption{Measles incidence per million, 2013. Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}). Grey indicates not reported.}
 \label{fig:incidence12}
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{cover12.pdf}
-\end{center}
-\caption{Measles vaccination cover (\% coverage of children that receive one dose of measles vaccine by their second birthday.), 2012.
+     \includegraphics[width=0.9\textwidth]{2013totnp3.pdf}
+\end{center}
+\caption{Measles vaccination cover (\% coverage of children that receive one dose of measles vaccine by their second birthday.), 2013.
 Data: WHO (\href{http://www.who.int/research/en/}{www.who.int/research/en/}).
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{immigration12.pdf}
-\end{center}
-\caption{Non-New Zealander international travel and immigration, 2012 }
+     \includegraphics[width=0.9\textwidth]{2013np1.pdf}
+\end{center}
+\caption{Non-New Zealander international travel and immigration by 0--19 year olds, 2013)}
 \label{fig:immigration12}
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{nztravel12.pdf}
-\end{center}
-\caption{New Zealander international travel, 2012 }
+     \includegraphics[width=0.9\textwidth]{2013nznp1.pdf}
+\end{center}
+\caption{New Zealander international travel, 2013}
 \label{fig:nztravel12}
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{travel12.pdf}
-\end{center}
-\caption{Total international travel, 2012 }
+     \includegraphics[width=0.9\textwidth]{2013totnp1.pdf}
+\end{center}
+\caption{Total international travel by 0--19 year old non New Zealanders and all ages of New Zealanders, 2013}
 \label{fig:travel12}
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{risk12.pdf}
-\end{center}
-\caption{Risk map for measles importation, 2012 }
+     \includegraphics[width=0.9\textwidth]{2013totnp4.pdf}
+\end{center}
+\caption{Risk map for measles importation, based on measles incidence and non New Zealander 0--19 year old and all ages of New Zealander travel figures, 2013)}
 \label{fig:risk12}
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{nzrisk12.pdf}
-\end{center}
-\caption{Risk map for measles importation from New Zealander international travel, 2012 }
+     \includegraphics[width=0.9\textwidth]{2013nznp4.pdf}
+\end{center}
+\caption{Risk map for measles importation from New Zealander international travel, based on measles incidence and travel figures, 2013}
 \label{fig:nzrisk12}
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{imrisk12.pdf}
-\end{center}
-\caption{Risk map for measles importation from non-New Zealander international travel and immigration, 2012 }
+     \includegraphics[width=0.9\textwidth]{2013np4.pdf}
+\end{center}
+\caption{Risk map for measles importation from 0--19 year old non-New Zealander international travel and immigration and measles incidence, 2013}
 \label{fig:imrisk12}
@@
      \end{center}
-     \caption{Trends in international travel, determined by immigration arrival data and New Zealander travel departure destinations}
+     \caption{Trends in international travel, determined by immigration arrival data and New Zealander travel departure destinations. Note this is for all age categories.}
      \label{fig:travel}
@@
 However, our analyses also highlight other groups that are at greater risk of measles.
-In particular, Pacific people are at greater risk per capita, as are the less deprived (NZDep 1--5), and European and Maori 6--17 year old children compared to Asian and Pacific ethnicity children of the same age.
+In particular, Pacific people are at greater risk per capita, as are the less deprived (NZDep 1--3), and European and Maori 5--17 year old children compared to Asian and Pacific ethnicity children of the same age.
 Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses.
+For example, increased numbers of more wealthy Europeans were measles cases, as was true for Maori.
+However, the alternative situation was clear for Pacific people, especially of the 0--1 year olds (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
+Results such as this may simply due to the spatial effects due the communities the outbreaks were occuring in, rather than an effect of wealth.
 A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data.
@@
 Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than DHB) immunisation coverage data are not able to provide reliable results, as discussed above.
-Frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census (\autoref{fig:}).
-Given these data gaps we are unable to provide finer scale risk maps.
+Frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census (\autoref{fig:naive_au_prop}).
+Given these data gaps we are unable to provide finer scale risk maps (\autoref{fig:naive_au}).
 
@@
 The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}).
-However, recently Australia was declared free of endemic measles and so the 2012 data must reflect imported measles there.
-However, other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}).
+However, recently Australia was declared free of endemic measles and so the 2013 data must reflect imported measles there.
+Other interaction terms beyond the product of travel and infection incidence may be more appropriate but this will require substantial further analyses.
+Other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}).
+The 2012 data were more complete than the 2013 data, which suffers from data deficiencies, but due to the dynamic nature of measles epidemiology and consultation with the Ministry we present the 2013 data.
+Lastly, more refined data on the age of travellers may improve the risk assessment, but these are currently unavailable.
 
@@
 We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their temporal and geographic correlations (\autoref{fig:dhbcases}).
+We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their reported outbreak numbers.
+The geographic locations of cases are shown in \autoref{fig:dhbcases}.
 The outbreaks we used in our analyses are shown in \autoref{fig:outbreaks}.
+Note these analyses do not use those sporadic cases not linked to other outbreaks.
 
@@
      \end{center}
-     \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014}
+     \caption{Measles data classified as outbreaks for reproduction number of the infection ($R_v$) estimation from 2009 to 2014.
+Outbreaks are coloured by reported outbreak number and sporadic cases not given a number are in grey.}
      \label{fig:outbreaks}
@@
 The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}.
-our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014 but for which data were available until 12 June, a mean estimate of 3.66 (range 3.4--3.92).
+Our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014, but for which data were available until 12 June, a mean estimate of 3.66 (range 3.4--3.92).
 The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one.
-Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect this outbreak to persist if conditions remain the same.
-An important caveat to this outbreak analysis is that because this 2013--2014 outbreak that ended in August was ongoing when the data were collected, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
-However, this is not the only case for which the $R_v$ is greater than one.
+Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect outbreaks to occur if conditions remain the same.
+An important caveat to the analysis of the last outbreak is that because this outbreak that ended in August was ongoing when the data were collected, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
+However, this is not the only case for which the $R_v$ is greater than one (\autoref{fig:r0}).
 
 These analyses also imply that the regular importation of measles is an ongoing process.
-Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to increase immunisation to high enough levels that measles does not become endemic.
+Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to increase immunisation to high enough levels that measles does not become endemic and large outbreaks prevented.
 The measles outbreaks in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
@@
 \end{center}
-\caption{Size: DHB Population, Statistics NZ 2013;		
-Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-Attack:		Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;	
-Vacc:		Number of the currently na\"{\i}ve to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Vacc (\%): Proportion of the currently na\"{\i}ve population to vaccinate.
+\caption{Epidemic size calculations. \textit{Size}: DHB Population, Statistics NZ 2013;		
+\textit{Na\"{\i}ve}:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
+\textit{Attack}:		Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;	
+\textit{Vacc}:		Number of the currently na\"{\i}ve to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; \textit{Vacc (\%)}: Proportion of the currently na\"{\i}ve population to vaccinate.
+These simulations assume homogeneous mixing.
 }
@@
 \item Regular introductions of measles pose an ongoing threat to New Zealand's efforts to eliminate measles (also see \autoref{sub:risk_analyses}).
-\item The reproduction number for measles in the partially immune New Zealand population is often close to one, suggesting increased population level immunity is required to ensure prevention of measles persistence following importation.
- \item The reproduction number, $R_v$, for measles in the current outbreak is over one, suggesting that this outbreak has the potential to persist for prolonged periods, with the caveat that this estimate was made during the ongoing outbreak.
+\item The reproduction number for measles in the partially immune New Zealand population is often close to and exceeding one, suggesting increased population level immunity is required to ensure prevention of measles persistence following importation.
  \item Additional vaccination levels to push $R_v$ below one among the currently na\"{\i}ve population in New Zealand range from 17\% to 34\% among DHBs, and 28\% at the national level (approximately 131,500 vaccinations).
@@
 
-The annual cost of treating and controlling measles in 11 industrialised countries was estimated to be more than US\$150 million \citep{carabin3}.
-The estimated cost for a case ranged from US\$189--344 \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755 \citep{carabin2}.
+The annual cost of treating and controlling measles in 11 industrialised countries was estimated to be more than US\$150 million\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$337.5million} \citep{carabin3}.
+The estimated cost for a case ranged from US\$189--344\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$425--774} \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$2175--3948} \citep{carabin2}.
 \citep{stack11} estimated the economic benefits from cases averted due to measles vaccination.
-They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion of costs averted between 2011 and 2020.
+They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion\footnote{Jan 2009 \$ to US\$ rate 0.5526: \$18.1billion} of costs averted between 2011 and 2020.
 Ninety-nine percent of these averted costs were the result of lost productivity due to an estimated 360,000 measles-specific premature mortalities, with the remaining <1\% associated with averted treatment costs and reduced caretaker productivity for the nearly 12 million measles cases avoided.
@@
 Italy has the highest reported annual cost of measles among industrialised countries \citep{carabin3}.
-In 2001, it reported losses related to measles of approximately US\$50 million.
-The economic impact of a large measles outbreak in Italy, 2002--03, has been evaluated.
+In 2001, it reported losses related to measles of approximately US\$50 million.\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$112.5million} The economic impact of a large measles outbreak in Italy, 2002--03, has been evaluated.
 The costs associated with 5,154 hospitalisations where measles was the main discharge diagnosis was calculated.
 The mean length of hospital stay was 5.2 days (median = 4 days and range = 1 to 303 days).
-The total cost of these hospitalisations amounted to \euro 8.83 million (\euro 1 $\approx$ NZ\$2.0 in 2002-03), or approximately \euro 1,700 per case.
-The average cost per non-complicated measles case was  \euro 1,429, while the mean cost of a case with complicated measles was  \euro 2,721.
-The average daily cost of a hospital stay was  \euro 327.
+The total cost of these hospitalisations amounted to \euro 8.83 million\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$18.4million}, or approximately \euro 1,700\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$3539} per case.
+The average cost per non-complicated measles case was  \euro 1,429,\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$2975} while the mean cost of a case with complicated measles was  \euro 2,721.\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$5664} The average daily cost of a hospital stay was  \euro 327.\footnote{Jan 2005 \$ to \euro\ rate 0.5375: \$681}
 
@@
 The mean number of contacts per case was 54 (median = 28, maximum = 206).
-The estimated cost to the public health unit for contact management for the epidemic was in excess of AUS\$48,000, with 90\% of this being associated with staff time.
+The estimated cost to the public health unit for contact management for the epidemic was in excess of AUS\$48,000,{Jan 2005 \$ to AUS\$ rate 0.7692: \$62,400} with 90\% of this being associated with staff time.
 
@@
 To determine the cost per case for the direct medical costs, cost of the Auckland Regional Public Health Service (ARPHS) for measles response were obtained from the Ministry of Health.
+These figures are representative for the most recent outbreaks as most cases were from North Island, and in particular Auckland and Waikato regions.
+However, results benefit--cost analyses are relatively robust to changes in this.
 Data, for the period January 1 - March 9, 2014, reported salaries for people involved with the measles outbreak management medical team.
@@
 In addition, measles operations personnel are calculated at a daily rate of \$600 and operations partners and IMT controller partners at \$729.
-Senior medical officer (SMO) costs for GP visits were estimated as approximately \$20 per consultation,\footnote{personal communication} though we allow flexibility in this value for the benefit--cost analyses later.
+Senior medical officer (SMO) costs for GP visits were estimated as approximately \$20 per consultation,\footnote{personal communication with Ministry of Health} though we allow flexibility in this value for the benefit--cost analyses later.
 These costs are applied on a per case basis below.
@@
 If the case is < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
+Case contact data was used to estimate further wage loss costs.
+These data were provided by the Ministry of Health.
+The data included the number and duration of school exclusions.
 
@@
 \subsection{Cost analyses results}
+\label{sub:cost_res}
 
@@
 \begin{table}
-\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014.
-PHN: public health nurse; PHA: public health assistant; IMT: incident management team; SSO: safety/security officer (SSO)}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014.
+PHN: public health nurse; PHA: public health assistant; IMT: incident management team; SSO: safety/security officer (SSO)}
 \label{table:direct}
@@
 \begin{table}
-\caption{Reported direct costs in NZ\$, with the number of cases, length of hospital day, total cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, 2000--2014}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Reported direct costs in NZ\$, with the number of cases (Cases), length of hospital stay (Days), total cost (Cost, NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, 2000--2014}
 \label{table:hosp}
@@
 \begin{table}
-\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
 \label{table:freq}
@@
 
-The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (\autoref{fig:hosp}).
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days (\autoref{fig:hosp}) and a cost of \$8,213.
 
@@
 \end{center}
-\caption{Number of measles cases attending hospital and stay duration from 2000--2014}
+\caption{Number of measles cases attending hospital and stay duration from 2000--2014 ($n=288$) in New Zealand}
 \label{fig:hosp}
@@
 \begin{table}\footnotesize
-\caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
-
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
 \label{table:cases}
@@
 \begin{table}
-\caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
 \label{table:regression_costs}
@@
 
-Wages lost for the period January 2008 -- August 2014 for the 247 cases and care givers was estimated to be \$210,436.
-This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008--2014 was estimated to be \$4,327 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
-\\
-OR
-\\
-Total wage lost for the 247 cases and care givers was estimated to be \$207,155.
-This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases <15 years of age.
-Overall, the cost per case from 2008 - 2014 was estimated to be \$2,513 (\$839 in forgone wages and \$1,710 in hospital costs).
-
-This final figure brings an approximate estimate of \$809,149 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs.
+Wages lost for the period January 2008 -- August 2014 for the 247 cases and care givers was estimated to be \$207,155.
+This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
+Overall, the cost per case from 2008--2014 was estimated to be \$2923.27 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
+This figure brings an approximate estimate of \$546650.78 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs for 17\%.
+
+The average number of contacts quarantined per case was calculated to be 2.11.
+Average number of days a contact is quarantined was calculated to be 7.3.
+The average wage of a contact per day was estimated to be \$170.39.
+This leads to the estimate of wage losses of \$490794.86 relating to case contacts.
+
+In total, with additional \$20 costs per GP consultation, we estimate that the 2014 measles cases alone cost the New Zealand economy \$1041185.63.
 
@@
 Primarily, we used \autoref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (\autoref{fig:naive_map} and \autoref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (\autoref{section:epidemic_modelling}).
-
 We assume that the number of cases prevented by this is proportional to the outbreak size.
@@
 \begin{sidewaystable}\small
-\caption{DHB and national level catch up vaccination rates and estimated outbreak sizes post catch-up vaccination. Size: DHB Population, Statistics NZ 2013;  	
-Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-Attack:	  Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;
-Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed the small numbers of large outbreaks.
-The distribution of these outbreaks sizes can be seen in \autoref{fig:sim1} and \autoref{fig:sim}}
-
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{DHB and national level catch up vaccination rates and estimated outbreak sizes post catch-up vaccination. \textit{Size}: DHB Population, Statistics NZ 2013;    
+\textit{Na\"{\i}ve}:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
+\textit{Attack}:	  Number infected in DHB in an outbreak of measles in the \textit{absence} of any vaccination $\left(\Pe\right)$;
+\textit{Vacc}:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; \textit{Proportion}: the proportion of the currently na\"{i}ve population requiring vaccination; \textit{Na\"{i}ve.post.vaccination}: the na\"{i}ve population following catch up vaccination; with \textit{Median.outbreak} and \textit{Mean.outbreak}: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed by the small numbers of large outbreaks.
+The distribution of these outbreaks sizes can be seen in \autoref{fig:sim1} and \autoref{fig:sim}}
 \label{table:vacp}
@@
 
-The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (~8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ is <1 is shown in \autoref{fig:sim} and \autoref{fig:sim1}.
+The expected number of cases in New Zealand, assuming homogenous mixing, in a na\"{\i}ve population of subsequently lower (approximately 8\%) numbers of na\"{\i}ve of the population and assuming measles $R_v$ is <1 is shown in \autoref{fig:sim} and \autoref{fig:sim1}.
 These simulations show that even in scenarios when $R_v$ is lower than one, and thus deterministically should fail to persist (i.e.
-become endemic), outbreaks can occur due to stochastic processes following measles importation.
+become endemic or cause an outbreak), outbreaks can occur due to stochastic processes following measles importation.
 The most likely scenario is that very few cases occur, with the median value from 1000 simulations low (2 cases).
-Thus most measles introductions will be single cases or lead to minor outbreaks.
-In this modelling exercise the mean value was 61 cases, suggesting this is likely to be an outbreak size to be expected.
-The maximum predicted number of cases was nearly 8,000 cases.
+Thus most measles introductions will be single cases or lead to minor outbreaks with single secondary cases.
+In this modelling exercise the mean value was 61 cases, suggesting there is likely to be an outbreak of a larger size given enough introductions.
+The maximum predicted number of cases was nearly 8,000 cases (\autoref{fig:sim1}).
 Note, however, that larger values are very rarely predicted and this does not take into account spatial variation or heterogeneous contact rates.
@@
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
-The full distribution of results can be seen in \autoref{fig:sim1}}
+The full distribution of rare results can be seen in \autoref{fig:sim1}}
      \label{fig:sim}
@@
   \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
-A subset of the distribution of results can be seen in \autoref{fig:sim}}
+The full distribution of common results can be seen in \autoref{fig:sim}}
      \label{fig:sim1}
@@
 \begin{sidewaystable}\small
-\caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Benefit--cost analyses with 20 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost20}
@@
 \begin{sidewaystable}\small
-\caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 <<results=tex,echo=FALSE>>=
@@
 @
+\caption{Benefit--cost analyses with 50 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost50}
@@
 We provide two costs for measles vaccinations for our cost analyses, \$20 and \$50, based on US literature.
+We add GP consultation costs of \$20 and the cost of contact exclusions at \$2624.57 (see \autoref{sub:cost_res}).
+Note vaccination and GP costs can be altered by the Ministry of Health in the spreadsheet provided.
 
@@
 While some of the data are complete and detailed, this is not true of all the data.
-In order to perform an accurate analysis of the current measles outbreak in New Zealand, more complete data would be better.
+More complete data would be better.
 For instance, age, gender, ethnicity, year of discharge, length of stay and estimated cost data are available for cases reported by publicly funded hospitals.
@@
 In 2000 expenditures for care in large state mental retardation/developmental disabilities (MR/DD) facilities continued to increase and reached a national annual average of US\$113,864\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$221612} per person.
-In 2000 the average annual expenditures for care in large state MR/DD facilities were \$113,864.\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$221612} The cost of a case of measles was estimated to range from \$71\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$138} (no complications and no hospitalisation) to \$29,556\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$57524} (encephalitis and hospitalisation for 8.7 days).
-They estimated the annual cost of measles vaccination in the US with its vaccination program to be \$1,234,083\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$2401874} (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
+In 2000 the average annual expenditures for care in large state MR/DD facilities were US\$113,864.\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$221612} The cost of a case of measles was estimated to range from US\$71\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$138} (no complications and no hospitalisation) to US\$29,556\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$57524} (encephalitis and hospitalisation for 8.7 days).
+They estimated the annual cost of measles vaccination in the US with its vaccination program to be US\$1,234,083\footnote{Jan 2012 \$ to US\$ rate 0.8007: \$2401874} (52.5\% direct cost and 47.5\% indirect cost) \citep{zhou4}.
 Work by others also finds that measles vaccination is extremely cost effective, however, in their analyses they also include the costs of mumps and rubella (see \autoref{table:zhou}) \citep{zhou4}.
@@
 \end{center}
-\caption{Summary of all measles, mumps, and rubella diseases and measles-mumps-rubella vaccination costs, USA, from \citep{zhou4}}
+\caption{Summary of all measles, mumps, and rubella diseases and measles-mumps-rubella (MMR) vaccination costs, USA, from \citep{zhou4}}
 \label{table:zhou}
@@
 It is unlikely $R_0$ was so high as 18 in New Zealand, as measles would still be endemic given current vaccination rates.
-However, the benefits of catch up vaccination are clear if $R_v$ is greater than one (\autoref{table:cost20} and ref{table:cost20}) and our analyses do not include the additional benefits of mumps and rubella immunity.
+However, the benefits of catch up vaccination are clear if $R_v$ is greater than one (\autoref{table:cost20} and \autoref{table:cost20}) and our analyses do not include the additional benefits of mumps and rubella immunity.
 As noted in our previous report, if measles $R_v$ in the currently na\"{\i}ve population is less than one, the benefits will not be so clear, though there may be medical and other benefits relating to maintaining measles free status that we have not included in our report.
@@
 \item Using $R_0$ values estimated for measles in New Zealand prior to vaccination and the current na\"{\i}ve population sizes, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).
-\item Outbreaks with a mean size of approximately 61 cases per year, but median of 2 cases per year may occur regularly due to importation, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
-There is, however, a small chance for some of these outbreaks to reach several thousand in number.
+\item Introduction of measles are estimated by simulation to lead to median outbreak size of 2 cases, suggesting most infections will be individuals or lead to single secondary cases.
+However, mean sizes of 61 cases we predicted, despite $R_v$ being below one and the epidemic predicted to die out without additional interventions.
+There is, however, a small chance for some of these outbreaks to reach larger sizes.
 \end{itemize}
@@
      \end{center}
-     \caption{Full details are given in (\autoref{fig:cases_all})}
+     \caption{Measles cases by age and NZDep from 2007--2014. Full details are given in (\autoref{fig:cases_all})}
      \label{fig:CaseAgeNzdep}
@@
      \end{center}
-     \caption{Full details are given in (\autoref{fig:cases_all})}
+     \caption{Measles cases by prioritized ethnicity and NZDep from 2007--2014. Full details are given in (\autoref{fig:cases_all})}
      \label{fig:CaseNzdepEth}
@@
      \end{center}
-     \caption{Full details are given in (\autoref{fig:cases_all})}
+     \caption{Measles cases by age and prioritized ethnicity from 2007--2014. Full details are given in (\autoref{fig:cases_all})}
      \label{fig:CaseAgeEth}
@@
      \end{center}
-     \caption{Full details are given in (\autoref{fig:percap_all})}
+     \caption{Per capita measles case rates by age and NZDep from 2007--2014. Full details are given in (\autoref{fig:percap_all})}
      \label{fig:PerCapAgeNzdep}
@@
      \end{center}
-     \caption{Full details are given in (\autoref{fig:percap_all})}
+     \caption{Per capita measles case rates by prioritized ethnicity and NZDep from 2007--2014. Full details are given in (\autoref{fig:percap_all})}
      \label{fig:PerCapNzdepEth}
@@
      \end{center}
-     \caption{Full details are given in (\autoref{fig:percap_all})}
+     \caption{Per capita measles case rates by prioritized ethnicity and age from 2007--2014. Full details are given in (\autoref{fig:percap_all})}
      \label{fig:PerCapAgeEth}

### 2015-03-22 draftfinalreport_v2.tex
Hopefully final draft...

@@
 
-\begin{sidewaysfigure}
-\begin{center}
-     \includegraphics[width=0.9\textwidth]{case_ordert.pdf}
-\end{center}
-\caption{test}
-\label{fig:cases_all}
-\end{sidewaysfigure}
-
-Two way interactions are shown in \autoref{sec:app} (\autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).
-
-% \begin{table}[hbtp]\scriptsize
-% \caption{Per capita measles rates}
-% <<results=tex,echo=FALSE>>=
-% latex(PerCap, file="", table.env=FALSE,rowname=NULL)
-% @
-% \label{table:percap_rates}
-% \end{table}
-
-\begin{table}[hbtp]\footnotesize
-\caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
-%latex.default(propvCohort, file = "", table.env = FALSE, rowname = NULL)%
-\begin{center}
-\begin{tabular}{ll}
-\hline\hline
-\multicolumn{1}{c}{Age range}&\multicolumn{1}{c}{Proportion}\tabularnewline
-\hline
-2-5&0.04\tabularnewline
-5-11&0.15\tabularnewline
-5-16&0.27\tabularnewline
-5-18&0.31\tabularnewline
-11-16&0.14\tabularnewline
-11-18&0.18\tabularnewline
-0-29&0.72\tabularnewline
-0-44&0.89\tabularnewline
-\textgreater8&0.8\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:vaccine_cohort}
-\end{table}
-
 The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap}, \autoref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), thus creating issues for estimating the denominator data for this group (University of Otago, personal communication).
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap_all}, \autoref{table:percap}), leading to very large standard error in regression analyses.
+However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), creating issues for estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
 This left us with 1102/1115 (99\%) of the measles cases with Prioritized Ethnicity recorded from 2007, and 1102/1137 (97\%) of all measles cases recorded since 2007 (\autoref{table:percap}).
+The breakdown of cases and per capita measles values can be see in \autoref{fig:cases_all} and \autoref{fig:percap_all}.
+Two way interactions are shown in the Appendix (\autoref{sec:app}, \autoref{fig:CaseAgeNzdep}, \autoref{fig:CaseNzdepEth}, \autoref{fig:CaseAgeEth}, \autoref{fig:PerCapAgeNzdep}, \autoref{fig:PerCapNzdepEth}, \autoref{fig:PerCapAgeEth}).
 
 \begin{table}[hbtp] \scriptsize
-\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
 %latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:percap}
+\end{tabular}\end{center}\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
+\label{table:percap}
+\end{table}
+
+\begin{table}[hbtp]\footnotesize
+%latex.default(propvCohort, file = "", table.env = FALSE, rowname = NULL)%
+\begin{center}
+\begin{tabular}{ll}
+\hline\hline
+\multicolumn{1}{c}{Age range}&\multicolumn{1}{c}{Proportion}\tabularnewline
+\hline
+2-5&0.04\tabularnewline
+5-11&0.15\tabularnewline
+5-16&0.27\tabularnewline
+5-18&0.31\tabularnewline
+11-16&0.14\tabularnewline
+11-18&0.18\tabularnewline
+0-29&0.72\tabularnewline
+0-44&0.89\tabularnewline
+\textgreater8&0.8\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
+\label{table:vaccine_cohort}
 \end{table}
@@
 % latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Sun Mar 22 01:29:41 2015
+% Sun Mar 22 17:09:42 2015
 \begin{tabular}{lrrrrrr}
@@
 % latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Sun Mar 22 01:29:41 2015
+% Sun Mar 22 17:09:42 2015
 \begin{tabular}{rrrrr}
@@
    \hline
-\end{tabular}\caption{Summary of the regression model results for measles risk factors for cases from 2007-2014}
+\end{tabular}\caption{Summary of the regression model results, estimated using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-2014}
 \label{table:regression_riskfactors}
@@
 \begin{table}
-\caption{Highest measles incidence per million (2012)}
 %latex.default(incidence, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Equatorial Guinea&$1617$\tabularnewline
-Nauru&$1100$\tabularnewline
-Democratic Republic of the Congo&$1096$\tabularnewline
-Somalia&$ 979$\tabularnewline
-Djibouti&$ 824$\tabularnewline
-Sudan&$ 786$\tabularnewline
-Burkina Faso&$ 447$\tabularnewline
-Romania&$ 342$\tabularnewline
-Ukraine&$ 280$\tabularnewline
-Sudan&$ 229$\tabularnewline
-Angola&$ 214$\tabularnewline
-Monaco&$ 132$\tabularnewline
-Nepal&$ 122$\tabularnewline
-Sierra Leone&$ 113$\tabularnewline
-Afghanistan&$  93$\tabularnewline
-Yemen&$  91$\tabularnewline
-Lesotho&$  87$\tabularnewline
-Qatar&$  78$\tabularnewline
-Thailand&$  78$\tabularnewline
-Malaysia&$  64$\tabularnewline
-Zambia&$  64$\tabularnewline
-Ghana&$  64$\tabularnewline
-Indonesia&$  63$\tabularnewline
-Congo&$  60$\tabularnewline
-Uganda&$  56$\tabularnewline
-Libyan Arab Jamahiriya&$  52$\tabularnewline
-Ethiopia&$  47$\tabularnewline
-Pakistan&$  45$\tabularnewline
-Myanmar&$  41$\tabularnewline
-Nigeria&$  38$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:incidence12}
+Georgia&$1806$\tabularnewline
+Namibia&$ 455$\tabularnewline
+Equatorial Guinea&$ 436$\tabularnewline
+Angola&$ 409$\tabularnewline
+Lebanon&$ 379$\tabularnewline
+Nigeria&$ 313$\tabularnewline
+Somalia&$ 311$\tabularnewline
+Sudan&$ 260$\tabularnewline
+Lesotho&$ 251$\tabularnewline
+Uganda&$ 217$\tabularnewline
+Netherlands&$ 159$\tabularnewline
+Central African Republic&$ 132$\tabularnewline
+Turkey&$ 100$\tabularnewline
+Sri Lanka&$ 100$\tabularnewline
+Togo&$  85$\tabularnewline
+Sudan&$  76$\tabularnewline
+Gabon&$  75$\tabularnewline
+Niger&$  71$\tabularnewline
+Nepal&$  68$\tabularnewline
+Benin&$  63$\tabularnewline
+Ethiopia&$  57$\tabularnewline
+Romania&$  51$\tabularnewline
+Pakistan&$  49$\tabularnewline
+Thailand&$  40$\tabularnewline
+Indonesia&$  39$\tabularnewline
+Qatar&$  36$\tabularnewline
+Cameroon&$  35$\tabularnewline
+Syrian Arab Republic&$  34$\tabularnewline
+Djibouti&$  33$\tabularnewline
+United Kingdom&$  31$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Highest measles incidence per million (2013)}
+\label{table:incidence12}
 \end{table}
@@
 \begin{table}
-\caption{Lowest national measles vaccine cover (\%), 2012}
 %latex.default(vaccinecover, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Equatorial Guinea&$34$\tabularnewline
-Somalia&$49$\tabularnewline
-Lesotho&$60$\tabularnewline
-Central African Republic&$65$\tabularnewline
-Papua New Guinea&$67$\tabularnewline
-Chad&$69$\tabularnewline
-Haiti&$69$\tabularnewline
-South Sudan&$70$\tabularnewline
-Gabon&$71$\tabularnewline
-Yemen&$71$\tabularnewline
-Benin&$72$\tabularnewline
-Lao People's Democratic Republic&$72$\tabularnewline
+Central African Republic&$29$\tabularnewline
+Somalia&$29$\tabularnewline
+Equatorial Guinea&$41$\tabularnewline
+South Sudan&$55$\tabularnewline
+Lesotho&$61$\tabularnewline
+Pakistan&$61$\tabularnewline
+Vanuatu&$65$\tabularnewline
+Gabon&$70$\tabularnewline
+Marshall Islands&$70$\tabularnewline
+Papua New Guinea&$70$\tabularnewline
+Timor-Leste&$70$\tabularnewline
+Mali&$72$\tabularnewline
 Togo&$72$\tabularnewline
-Suriname&$73$\tabularnewline
-Timor-Leste&$73$\tabularnewline
-Paraguay&$74$\tabularnewline
-Mauritania&$75$\tabularnewline
-Namibia&$76$\tabularnewline
-Eritrea&$77$\tabularnewline
-Marshall Islands&$78$\tabularnewline
-Nigeria&$78$\tabularnewline
-Syrian Arab Republic&$78$\tabularnewline
-Ukraine&$79$\tabularnewline
+Nigeria&$73$\tabularnewline
+Liberia&$74$\tabularnewline
+San Marino&$74$\tabularnewline
+Eritrea&$75$\tabularnewline
+Paraguay&$75$\tabularnewline
+Solomon Islands&$76$\tabularnewline
+Iraq&$77$\tabularnewline
+Ethiopia&$78$\tabularnewline
+Haiti&$78$\tabularnewline
+Yemen&$78$\tabularnewline
+Chad&$79$\tabularnewline
+Kenya&$79$\tabularnewline
 Congo&$80$\tabularnewline
-Ethiopia&$80$\tabularnewline
-Liberia&$80$\tabularnewline
-Afghanistan&$81$\tabularnewline
-Cameroon&$82$\tabularnewline
-Mozambique&$82$\tabularnewline
-Senegal&$82$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:cover12}
+Djibouti&$80$\tabularnewline
+Mauritania&$80$\tabularnewline
+Zambia&$80$\tabularnewline
+Afghanistan&$82$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Lowest national measles vaccine cover (\%), 2013}
+\label{table:cover12}
 \end{table}
@@
 \begin{table}
-\caption{Non-New Zealander travel and immigration numbers (2012)}
 %latex.default(im, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Australia&$809775$\tabularnewline
-United Kingdom&$306177$\tabularnewline
-China&$256036$\tabularnewline
-United States&$194438$\tabularnewline
-Japan&$ 86676$\tabularnewline
-Germany&$ 83608$\tabularnewline
-Korea, Republic of&$ 73459$\tabularnewline
-France&$ 71448$\tabularnewline
-India&$ 69038$\tabularnewline
-Canada&$ 54981$\tabularnewline
-Malaysia&$ 50506$\tabularnewline
-Fiji&$ 46723$\tabularnewline
-Netherlands&$ 33431$\tabularnewline
-South Africa&$ 30600$\tabularnewline
-Tonga&$ 26717$\tabularnewline
-Singapore&$ 25460$\tabularnewline
-Ireland&$ 25036$\tabularnewline
-Samoa&$ 24547$\tabularnewline
-Philippines&$ 24527$\tabularnewline
-Taiwan&$ 23815$\tabularnewline
-Thailand&$ 20258$\tabularnewline
-Hong Kong&$ 19103$\tabularnewline
-Indonesia&$ 15692$\tabularnewline
-Switzerland&$ 14851$\tabularnewline
-Brazil&$ 14778$\tabularnewline
-Sweden&$ 13719$\tabularnewline
-Italy&$ 13388$\tabularnewline
-Spain&$ 10104$\tabularnewline
-Denmark&$ 10056$\tabularnewline
-Russia&$  8103$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:imtravel12}
+Australia&$130703$\tabularnewline
+China&$ 26380$\tabularnewline
+United Kingdom&$ 25155$\tabularnewline
+Japan&$ 20534$\tabularnewline
+United States&$ 20247$\tabularnewline
+Germany&$ 14793$\tabularnewline
+France&$ 12173$\tabularnewline
+Korea, Republic of&$ 10398$\tabularnewline
+India&$  7013$\tabularnewline
+Fiji&$  6040$\tabularnewline
+Malaysia&$  5981$\tabularnewline
+Thailand&$  5233$\tabularnewline
+Canada&$  5013$\tabularnewline
+Singapore&$  4887$\tabularnewline
+Samoa&$  4047$\tabularnewline
+South Africa&$  3763$\tabularnewline
+Tonga&$  3477$\tabularnewline
+Hong Kong&$  3281$\tabularnewline
+Philippines&$  3147$\tabularnewline
+Taiwan&$  2915$\tabularnewline
+Indonesia&$  2580$\tabularnewline
+Netherlands&$  2509$\tabularnewline
+Brazil&$  2265$\tabularnewline
+Sweden&$  1862$\tabularnewline
+Denmark&$  1605$\tabularnewline
+Switzerland&$  1376$\tabularnewline
+Ireland&$  1331$\tabularnewline
+Chile&$  1157$\tabularnewline
+Saudi Arabia&$  1141$\tabularnewline
+Russia&$  1068$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Non-New Zealander travel and immigration numbers, 0--19 year olds, 2013}
+\label{table:imtravel12}
 \end{table}
@@
 \begin{table}
-\caption{New Zealander travel numbers by destination (2012)}
 %latex.default(nz, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Australia&$989880$\tabularnewline
-United States&$121620$\tabularnewline
-Fiji&$104720$\tabularnewline
-United Kingdom&$ 95560$\tabularnewline
-Cook Islands&$ 71960$\tabularnewline
-China&$ 66040$\tabularnewline
-Samoa&$ 46020$\tabularnewline
-Thailand&$ 41100$\tabularnewline
-India&$ 38580$\tabularnewline
-Canada&$ 20400$\tabularnewline
-Japan&$ 20040$\tabularnewline
-Malaysia&$ 19860$\tabularnewline
-Indonesia&$ 19660$\tabularnewline
-Hong Kong&$ 18220$\tabularnewline
-Tonga&$ 17760$\tabularnewline
-Singapore&$ 17120$\tabularnewline
-South Africa&$ 15380$\tabularnewline
-Philippines&$ 15220$\tabularnewline
-France&$ 14500$\tabularnewline
-Korea, Republic of&$ 13960$\tabularnewline
-Viet Nam&$ 12920$\tabularnewline
-Germany&$ 12700$\tabularnewline
-Vanuatu&$ 12520$\tabularnewline
-Italy&$ 11820$\tabularnewline
-Taiwan&$ 10460$\tabularnewline
-New Caledonia&$  7340$\tabularnewline
-Ireland&$  6360$\tabularnewline
-French Polynesia&$  6360$\tabularnewline
-Papua New Guinea&$  6140$\tabularnewline
-Netherlands&$  5720$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:nztravel12}
+Australia&$1017540$\tabularnewline
+United States&$ 142200$\tabularnewline
+Fiji&$ 112260$\tabularnewline
+United Kingdom&$  96680$\tabularnewline
+China&$  73340$\tabularnewline
+Cook Islands&$  72820$\tabularnewline
+Samoa&$  44060$\tabularnewline
+Thailand&$  42060$\tabularnewline
+India&$  40760$\tabularnewline
+Indonesia&$  25100$\tabularnewline
+Japan&$  21840$\tabularnewline
+Canada&$  21720$\tabularnewline
+Tonga&$  20120$\tabularnewline
+Singapore&$  18300$\tabularnewline
+Hong Kong&$  18000$\tabularnewline
+Malaysia&$  17480$\tabularnewline
+Philippines&$  17260$\tabularnewline
+France&$  15980$\tabularnewline
+Vanuatu&$  15580$\tabularnewline
+South Africa&$  14000$\tabularnewline
+Korea, Republic of&$  13980$\tabularnewline
+Italy&$  13620$\tabularnewline
+Germany&$  12800$\tabularnewline
+Viet Nam&$  12640$\tabularnewline
+Taiwan&$  10300$\tabularnewline
+New Caledonia&$   7620$\tabularnewline
+Spain&$   6560$\tabularnewline
+Netherlands&$   6260$\tabularnewline
+Papua New Guinea&$   6200$\tabularnewline
+Ireland&$   6160$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{New Zealander travel numbers by destination, 2013}
+\label{table:nztravel12}
 \end{table}
@@
 \begin{table}
-\caption{Total New Zealand traveller numbers by country (New Zealand nationals and immigrants, 2012)}
 %latex.default(immigration, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Australia&$1799655$\tabularnewline
-United Kingdom&$ 401737$\tabularnewline
-China&$ 322076$\tabularnewline
-United States&$ 316058$\tabularnewline
-Fiji&$ 151443$\tabularnewline
-India&$ 107618$\tabularnewline
-Japan&$ 106716$\tabularnewline
-Germany&$  96308$\tabularnewline
-Korea, Republic of&$  87419$\tabularnewline
-France&$  85948$\tabularnewline
-Canada&$  75381$\tabularnewline
-Samoa&$  70567$\tabularnewline
-Malaysia&$  70366$\tabularnewline
-Thailand&$  61358$\tabularnewline
-South Africa&$  45980$\tabularnewline
-Tonga&$  44477$\tabularnewline
-Singapore&$  42580$\tabularnewline
-Philippines&$  39747$\tabularnewline
-Netherlands&$  39151$\tabularnewline
-Hong Kong&$  37323$\tabularnewline
-Indonesia&$  35352$\tabularnewline
-Taiwan&$  34275$\tabularnewline
-Ireland&$  31396$\tabularnewline
-Italy&$  25208$\tabularnewline
-Viet Nam&$  18574$\tabularnewline
-Switzerland&$  18431$\tabularnewline
-Brazil&$  17878$\tabularnewline
-Vanuatu&$  16261$\tabularnewline
-Spain&$  15604$\tabularnewline
-Sweden&$  15479$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:immigration12}
+Australia&$1148243$\tabularnewline
+United States&$ 162447$\tabularnewline
+United Kingdom&$ 121835$\tabularnewline
+Fiji&$ 118300$\tabularnewline
+China&$  99720$\tabularnewline
+Samoa&$  48107$\tabularnewline
+India&$  47773$\tabularnewline
+Thailand&$  47293$\tabularnewline
+Japan&$  42374$\tabularnewline
+France&$  28153$\tabularnewline
+Indonesia&$  27680$\tabularnewline
+Germany&$  27593$\tabularnewline
+Canada&$  26733$\tabularnewline
+Korea, Republic of&$  24378$\tabularnewline
+Tonga&$  23597$\tabularnewline
+Malaysia&$  23461$\tabularnewline
+Singapore&$  23187$\tabularnewline
+Hong Kong&$  21281$\tabularnewline
+Philippines&$  20407$\tabularnewline
+South Africa&$  17763$\tabularnewline
+Vanuatu&$  15748$\tabularnewline
+Italy&$  14617$\tabularnewline
+Viet Nam&$  13680$\tabularnewline
+Taiwan&$  13215$\tabularnewline
+Netherlands&$   8769$\tabularnewline
+Ireland&$   7491$\tabularnewline
+Spain&$   7118$\tabularnewline
+Papua New Guinea&$   6457$\tabularnewline
+Sri Lanka&$   5764$\tabularnewline
+Brazil&$   5245$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Total New Zealand traveller numbers by country (New Zealand nationals and 0--19 year old immigrants, 2013)}
+\label{table:immigration12}
 \end{table}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand in 2013, all travellers, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
 %latex.default(risk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Australia&$15537152$\tabularnewline
-United Kingdom&$13386328$\tabularnewline
-Thailand&$ 4774688$\tabularnewline
-Malaysia&$ 4495338$\tabularnewline
-Indonesia&$ 2216871$\tabularnewline
-India&$ 1620172$\tabularnewline
-China&$ 1432659$\tabularnewline
-Ukraine&$  765096$\tabularnewline
-Ireland&$  734128$\tabularnewline
-Romania&$  655107$\tabularnewline
-Philippines&$  631303$\tabularnewline
-Spain&$  401823$\tabularnewline
-Samoa&$  373370$\tabularnewline
-Singapore&$  337236$\tabularnewline
-Nepal&$  324526$\tabularnewline
-Germany&$  193081$\tabularnewline
-Japan&$  191585$\tabularnewline
-Israel&$  174177$\tabularnewline
-Italy&$  155674$\tabularnewline
-Russia&$  155632$\tabularnewline
-Pakistan&$  148514$\tabularnewline
-Sudan&$  141552$\tabularnewline
-Switzerland&$  140589$\tabularnewline
-Nauru&$  126500$\tabularnewline
-Viet Nam&$  118241$\tabularnewline
-Afghanistan&$  111854$\tabularnewline
-Somalia&$  104775$\tabularnewline
-Saudi Arabia&$   82584$\tabularnewline
-United Arab Emirates&$   67606$\tabularnewline
-Belgium&$   63784$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:risk12}
+Australia&$7870820$\tabularnewline
+United Kingdom&$3723960$\tabularnewline
+China&$1928613$\tabularnewline
+Thailand&$1870193$\tabularnewline
+Netherlands&$1391896$\tabularnewline
+Indonesia&$1071788$\tabularnewline
+Philippines&$ 616175$\tabularnewline
+Germany&$ 590184$\tabularnewline
+Sri Lanka&$ 575635$\tabularnewline
+India&$ 532515$\tabularnewline
+Turkey&$ 350451$\tabularnewline
+Viet Nam&$ 169200$\tabularnewline
+Malaysia&$ 156460$\tabularnewline
+Namibia&$ 151083$\tabularnewline
+Nepal&$ 134525$\tabularnewline
+France&$ 119768$\tabularnewline
+Switzerland&$ 109766$\tabularnewline
+Nigeria&$ 101638$\tabularnewline
+Angola&$  98652$\tabularnewline
+Japan&$  76407$\tabularnewline
+Pakistan&$  73853$\tabularnewline
+Uganda&$  70444$\tabularnewline
+Canada&$  62923$\tabularnewline
+Lebanon&$  57601$\tabularnewline
+Russia&$  55743$\tabularnewline
+Romania&$  34971$\tabularnewline
+Ethiopia&$  21876$\tabularnewline
+Sweden&$  21674$\tabularnewline
+Sudan&$  21024$\tabularnewline
+Spain&$  20552$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Risk of measles importation to New Zealand in 2013, estimated by 0--19 year old international and all New Zealander traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:imrisk12}}
+\label{table:risk12}
 \end{table}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
 %latex.default(nzrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Australia&$8546036$\tabularnewline
-Thailand&$3198274$\tabularnewline
-United Kingdom&$3184166$\tabularnewline
-Malaysia&$1268758$\tabularnewline
-Indonesia&$1232849$\tabularnewline
-India&$ 580816$\tabularnewline
-China&$ 293759$\tabularnewline
-Samoa&$ 243492$\tabularnewline
-Philippines&$ 241740$\tabularnewline
-Nepal&$ 188450$\tabularnewline
-Romania&$ 164376$\tabularnewline
-Ireland&$ 148715$\tabularnewline
-Spain&$ 141632$\tabularnewline
-Singapore&$ 135591$\tabularnewline
-Ukraine&$ 100781$\tabularnewline
-Viet Nam&$  82248$\tabularnewline
-Sudan&$  78640$\tabularnewline
-Afghanistan&$  74756$\tabularnewline
-Italy&$  72995$\tabularnewline
-United Arab Emirates&$  60508$\tabularnewline
-Pakistan&$  58435$\tabularnewline
-Nauru&$  44000$\tabularnewline
-Democratic Republic of the Congo&$  43850$\tabularnewline
-Qatar&$  43686$\tabularnewline
-Israel&$  40301$\tabularnewline
-Japan&$  35977$\tabularnewline
-Russia&$  35334$\tabularnewline
-Angola&$  34258$\tabularnewline
-Switzerland&$  27308$\tabularnewline
-Saudi Arabia&$  25775$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:nzrisk12}
+Australia&$6974895$\tabularnewline
+United Kingdom&$2955082$\tabularnewline
+Thailand&$1663255$\tabularnewline
+China&$1418417$\tabularnewline
+Netherlands&$ 993645$\tabularnewline
+Indonesia&$ 971888$\tabularnewline
+Philippines&$ 521154$\tabularnewline
+Sri Lanka&$ 499336$\tabularnewline
+India&$ 454343$\tabularnewline
+Turkey&$ 342245$\tabularnewline
+Germany&$ 273778$\tabularnewline
+Viet Nam&$ 156336$\tabularnewline
+Namibia&$ 145622$\tabularnewline
+Nepal&$ 125990$\tabularnewline
+Malaysia&$ 116573$\tabularnewline
+Angola&$  98243$\tabularnewline
+Nigeria&$  87565$\tabularnewline
+Switzerland&$  79655$\tabularnewline
+Uganda&$  69360$\tabularnewline
+France&$  67982$\tabularnewline
+Pakistan&$  53765$\tabularnewline
+Lebanon&$  53054$\tabularnewline
+Canada&$  51123$\tabularnewline
+Japan&$  39381$\tabularnewline
+Russia&$  38275$\tabularnewline
+Georgia&$  36127$\tabularnewline
+Romania&$  31378$\tabularnewline
+Ethiopia&$  19471$\tabularnewline
+Spain&$  18941$\tabularnewline
+Qatar&$  16373$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Risk of measles importation to New Zealand due to New Zealander travel in 2013, estimated by all traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:imrisk12} and \autoref{table:risk12}}
+\label{table:nzrisk12}
 \end{table}
@@
 \begin{table}
-\caption{Risk of measles importation to New Zealand due to non-New Zealander travel and immigration in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
 %latex.default(imrisk, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-United Kingdom&$10202161$\tabularnewline
-Australia&$ 6991116$\tabularnewline
-Malaysia&$ 3226580$\tabularnewline
-Thailand&$ 1576414$\tabularnewline
-China&$ 1138900$\tabularnewline
-India&$ 1039356$\tabularnewline
-Indonesia&$  984022$\tabularnewline
-Ukraine&$  664315$\tabularnewline
-Ireland&$  585413$\tabularnewline
-Romania&$  490731$\tabularnewline
-Philippines&$  389563$\tabularnewline
-Spain&$  260191$\tabularnewline
-Singapore&$  201644$\tabularnewline
-Germany&$  167620$\tabularnewline
-Japan&$  155607$\tabularnewline
-Nepal&$  136076$\tabularnewline
-Israel&$  133876$\tabularnewline
-Samoa&$  129878$\tabularnewline
-Russia&$  120298$\tabularnewline
-Switzerland&$  113281$\tabularnewline
-Pakistan&$   90079$\tabularnewline
-Somalia&$   85191$\tabularnewline
-Italy&$   82679$\tabularnewline
-Nauru&$   82500$\tabularnewline
-Sudan&$   62912$\tabularnewline
-Saudi Arabia&$   56809$\tabularnewline
-Belgium&$   53337$\tabularnewline
-Sweden&$   43273$\tabularnewline
-Afghanistan&$   37098$\tabularnewline
-Viet Nam&$   35993$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:imrisk12}
+Australia&$895925$\tabularnewline
+United Kingdom&$768878$\tabularnewline
+China&$510197$\tabularnewline
+Netherlands&$398252$\tabularnewline
+Germany&$316406$\tabularnewline
+Thailand&$206938$\tabularnewline
+Indonesia&$ 99899$\tabularnewline
+Philippines&$ 95021$\tabularnewline
+India&$ 78172$\tabularnewline
+Sri Lanka&$ 76299$\tabularnewline
+France&$ 51786$\tabularnewline
+Malaysia&$ 39887$\tabularnewline
+Japan&$ 37026$\tabularnewline
+Switzerland&$ 30111$\tabularnewline
+Pakistan&$ 20088$\tabularnewline
+Russia&$ 17469$\tabularnewline
+Nigeria&$ 14073$\tabularnewline
+Viet Nam&$ 12863$\tabularnewline
+Canada&$ 11799$\tabularnewline
+Sweden&$  9984$\tabularnewline
+Nepal&$  8535$\tabularnewline
+Turkey&$  8206$\tabularnewline
+Namibia&$  5461$\tabularnewline
+Sudan&$  5451$\tabularnewline
+Denmark&$  4874$\tabularnewline
+Lebanon&$  4547$\tabularnewline
+Somalia&$  4046$\tabularnewline
+Romania&$  3593$\tabularnewline
+Afghanistan&$  2437$\tabularnewline
+Ethiopia&$  2405$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Risk of measles importation to New Zealand due to non-New Zealander 0--19 year old travel and immigration in 2013, estimated by traveller numbers multiplied by measles incidence in the source or destination location, see also \autoref{table:nzrisk12} and \autoref{table:risk12}}
+\label{table:imrisk12}
 \end{table}
 
-%%%%%%%%%%%%%%%%%%
-
 \begin{table}
-\caption{Measles and travel data from 2012 for the top 10 countries identified as high risk for measles importation.
-incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of travellers entering New Zealand from the country}
 %latex.default(toprisk2, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Australia&$  9$&$94$&$1799655$\tabularnewline
-United Kingdom&$ 33$&$93$&$ 401737$\tabularnewline
-Thailand&$ 78$&$98$&$  61358$\tabularnewline
-Malaysia&$ 64$&$95$&$  70366$\tabularnewline
-Indonesia&$ 63$&$84$&$  35352$\tabularnewline
-India&$ 15$&$85$&$ 107618$\tabularnewline
-China&$  4$&$99$&$ 322076$\tabularnewline
-Ukraine&$280$&$79$&$   2733$\tabularnewline
-Ireland&$ 23$&$92$&$  31396$\tabularnewline
-Romania&$342$&$94$&$   1913$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:toprisk}
+Australia&$  7$&$94$&$1148243$\tabularnewline
+United Kingdom&$ 31$&$95$&$ 121835$\tabularnewline
+China&$ 19$&$99$&$  99720$\tabularnewline
+Thailand&$ 40$&$99$&$  47293$\tabularnewline
+Netherlands&$159$&$96$&$   8769$\tabularnewline
+Indonesia&$ 39$&$89$&$  27680$\tabularnewline
+Philippines&$ 30$&$90$&$  20407$\tabularnewline
+Germany&$ 21$&$97$&$  27593$\tabularnewline
+Sri Lanka&$100$&$99$&$   5764$\tabularnewline
+India&$ 11$&$88$&$  47773$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Measles and travel data from 2013 for the top 10 countries identified as high risk for measles importation.
+incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of 0--19 year old travellers entering New Zealand from the country}
+\label{table:toprisk}
 \end{table}
@@
 \begin{table}
-\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014.
-PHN: public health nurse; PHA: public health assistant; IMT: incident management team; SSO: safety/security officer (SSO)}
 %latex.default(costtable1, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:direct}
+\end{tabular}\end{center}\caption{Estimated costs (NZ\$) for measles management in New Zealand, January 1 -- March 9, 2014.
+PHN: public health nurse; PHA: public health assistant; IMT: incident management team; SSO: safety/security officer (SSO)}
+\label{table:direct}
 \end{table}
@@
 \begin{table}
-\caption{Reported direct costs in NZ\$, with the number of cases, length of hospital day, total cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, 2000--2014}
 %latex.default(costtable2, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:hosp}
+\end{tabular}\end{center}\caption{Reported direct costs in NZ\$, with the number of cases (Cases), length of hospital stay (Days), total cost (Cost, NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, 2000--2014}
+\label{table:hosp}
  \centering
@@
 \begin{table}
-\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
 %latex.default(costtable3, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:freq}
+\end{tabular}\end{center}\caption{Frequency of measles cases and number and proportion admitted to hospital by age group, 16 December, 2013 -- 19 June, 2014}
+\label{table:freq}
 \end{table}
 
-The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days and a cost of \$8,213 (\autoref{fig:hosp}).
+The length of hospital stay for the 293 cases reported between 2000 and 2014 ranged from 0 to 19 days, with a male patient, who was discharged in 2011 at age 57, after a stay of 19 days (\autoref{fig:hosp}) and a cost of \$8,213.
 
@@
 \begin{table}\footnotesize
-\caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
-
 %latex.default(costtable4, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:cases}
+\end{tabular}\end{center}\caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
+Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
+\label{table:cases}
  \centering
@@
 \begin{table}
-\caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
 %latex.default(costtable5, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:regression_costs}
+\end{tabular}\end{center}\caption{Regression results ($R^{2}_\textsf{adj} = 0.43$, p-value $<0.001$) for measles hospitalisation cost based on length of stay (days), gender, case age and year of case ($n=288$) in New Zealand, 2000 -- 2014}
+\label{table:regression_costs}
 \end{table}
 
-Wages lost for the period January 2008 -- August 2014 for the 247 cases and care givers was estimated to be \$210,436.
-This consisted of \$107,820 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008--2014 was estimated to be \$4,327 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
-\\
-OR
-\\
-Total wage lost for the 247 cases and care givers was estimated to be \$207,155.
-This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases <15 years of age.
-Overall, the cost per case from 2008 - 2014 was estimated to be \$2,513 (\$839 in forgone wages and \$1,710 in hospital costs).
-
-This final figure brings an approximate estimate of \$809,149 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs.
+Wages lost for the period January 2008 -- August 2014 for the 247 cases and care givers was estimated to be \$207,155.
+This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
+Overall, the cost per case from 2008--2014 was estimated to be \$2923.27 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
+This figure brings an approximate estimate of \$546650.78 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs for 17\%.
+
+The average number of contacts quarantined per case was calculated to be 2.11.
+Average number of days a contact is quarantined was calculated to be 7.3.
+The average wage of a contact per day was estimated to be \$170.39.
+This leads to the estimate of wage losses of \$490794.86 relating to case contacts.
+
+In total, with additional \$20 costs per GP consultation, we estimate that the 2014 measles cases alone cost the New Zealand economy \$1041185.63.
 
@@
 \begin{sidewaystable}\small
-\caption{DHB and national level catch up vaccination rates and estimated outbreak sizes post catch-up vaccination. Size: DHB Population, Statistics NZ 2013;  	
-Na\"{\i}ve:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-Attack:	  Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;
-Vacc:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; Proportion: the proportion of the currently na\"{i}ve population requiring vaccination; Na\"{i}ve.post.vaccination: the na\"{i}ve population following catch up vaccination; with Median.outbreak and Mean.outbreak: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed the small numbers of large outbreaks.
-The distribution of these outbreaks sizes can be seen in \autoref{fig:sim1} and \autoref{fig:sim}}
-
 %latex.default(vacp, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\label{table:vacp}
+\end{tabular}\end{center}\caption{DHB and national level catch up vaccination rates and estimated outbreak sizes post catch-up vaccination. \textit{Size}: DHB Population, Statistics NZ 2013;    
+\textit{Na\"{\i}ve}:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
+\textit{Attack}:	  Number infected in DHB in an outbreak of measles in the \textit{absence} of any vaccination $\left(\Pe\right)$;
+\textit{Vacc}:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; \textit{Proportion}: the proportion of the currently na\"{i}ve population requiring vaccination; \textit{Na\"{i}ve.post.vaccination}: the na\"{i}ve population following catch up vaccination; with \textit{Median.outbreak} and \textit{Mean.outbreak}: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed by the small numbers of large outbreaks.
+The distribution of these outbreaks sizes can be seen in \autoref{fig:sim1} and \autoref{fig:sim}}
+\label{table:vacp}
 \end{sidewaystable}
@@
 \begin{sidewaystable}\small
-\caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}	is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable20, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Auckland&$17920$&$358400$&$26546459$&$55634145$&$ 9942525$&$81779021$&$152792323$&$ 82$&$4020980$&$34.9$\tabularnewline
-Bay of Plenty&$ 4585$&$ 91700$&$ 7188051$&$15064196$&$ 2692162$&$22143509$&$ 41371958$&$ 71$&$3481580$&$11.6$\tabularnewline
-Canterbury&$13687$&$273740$&$21039340$&$44092724$&$ 7879928$&$64813791$&$121095235$&$ 62$&$3040253$&$36.5$\tabularnewline
-Capital and Coast&$10461$&$209220$&$15678760$&$32858409$&$ 5872213$&$48299988$&$ 90241571$&$ 96$&$4707488$&$18.4$\tabularnewline
-Counties Manukau&$18880$&$377600$&$28032290$&$58748043$&$10499018$&$86356274$&$161344260$&$ 50$&$2451817$&$57.0$\tabularnewline
-Hawke's Bay&$ 3751$&$ 75020$&$ 5832570$&$12223478$&$ 2184490$&$17967816$&$ 33570276$&$ 56$&$2746035$&$11.9$\tabularnewline
-Hutt Valley&$ 4388$&$ 87760$&$ 6676018$&$13991115$&$ 2500389$&$20566142$&$ 38424874$&$ 86$&$4217125$&$ 8.9$\tabularnewline
-Lakes&$ 2886$&$ 57720$&$ 4423416$&$ 9270274$&$ 1656715$&$13626775$&$ 25459666$&$ 62$&$3040253$&$ 8.2$\tabularnewline
-MidCentral&$ 4628$&$ 92560$&$ 7112226$&$14905287$&$ 2663763$&$21909922$&$ 40935534$&$ 75$&$3677725$&$10.9$\tabularnewline
-Nelson Marlborough&$ 2356$&$ 47120$&$ 3758029$&$ 7875805$&$ 1407506$&$11576985$&$ 21629928$&$ 90$&$4413270$&$ 4.8$\tabularnewline
-Northland&$ 3071$&$ 61420$&$ 4845992$&$10155878$&$ 1814984$&$14928562$&$ 27891869$&$ 70$&$3432544$&$ 8.0$\tabularnewline
-South Canterbury&$  893$&$ 17860$&$ 1429602$&$ 2996056$&$  535433$&$ 4404031$&$  8228297$&$ 72$&$3530616$&$ 2.3$\tabularnewline
-Southern&$ 8371$&$167420$&$12877490$&$26987711$&$ 4823045$&$39670397$&$ 74118424$&$102$&$5001706$&$14.3$\tabularnewline
-Tairawhiti&$ 1359$&$ 27180$&$ 2071133$&$ 4340531$&$  775708$&$ 6380333$&$ 11920734$&$ 47$&$2304708$&$ 5.1$\tabularnewline
-Taranaki&$ 2899$&$ 57980$&$ 4483054$&$ 9395259$&$ 1679052$&$13810495$&$ 25802921$&$ 68$&$3334471$&$ 7.6$\tabularnewline
-Waikato&$11331$&$226620$&$17250640$&$36152642$&$ 6460934$&$53142322$&$ 99288775$&$ 95$&$4658452$&$20.3$\tabularnewline
-Wairarapa&$  720$&$ 14400$&$ 1146748$&$ 2403272$&$  429495$&$ 3532673$&$  6600291$&$ 59$&$2893144$&$ 2.3$\tabularnewline
-Waitemata&$17291$&$345820$&$26218451$&$54946730$&$ 9819676$&$80768561$&$150904424$&$ 70$&$3432544$&$39.9$\tabularnewline
-West Coast&$  685$&$ 13700$&$ 1077739$&$ 2258647$&$  403649$&$ 3320083$&$  6203097$&$ 50$&$2451817$&$ 2.5$\tabularnewline
-Whanganui&$ 1378$&$ 27560$&$ 2155478$&$ 4517295$&$  807298$&$ 6640166$&$ 12406193$&$ 58$&$2844108$&$ 4.3$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:cost20}
+Auckland&$17920$&$358400$&$26132561$&$55634145$&$ 9942525$&$81779021$&$152428668$&$ 82$&$4011409$&$34.88$\tabularnewline
+Bay of Plenty&$ 4585$&$ 91700$&$ 7075979$&$15064196$&$ 2692162$&$22143509$&$ 41273490$&$ 71$&$3473294$&$11.58$\tabularnewline
+Canterbury&$13687$&$273740$&$20711307$&$44092724$&$ 7879928$&$64813791$&$120807021$&$ 62$&$3033017$&$36.53$\tabularnewline
+Capital and Coast&$10461$&$209220$&$15434306$&$32858409$&$ 5872213$&$48299988$&$ 90026791$&$ 96$&$4696284$&$18.35$\tabularnewline
+Counties Manukau&$18880$&$377600$&$27595227$&$58748043$&$10499018$&$86356274$&$160960251$&$ 50$&$2445981$&$57.01$\tabularnewline
+Hawke's Bay&$ 3751$&$ 75020$&$ 5741632$&$12223478$&$ 2184490$&$17967816$&$ 33490377$&$ 56$&$2739499$&$11.90$\tabularnewline
+Hutt Valley&$ 4388$&$ 87760$&$ 6571929$&$13991115$&$ 2500389$&$20566142$&$ 38333420$&$ 86$&$4207088$&$ 8.93$\tabularnewline
+Lakes&$ 2886$&$ 57720$&$ 4354448$&$ 9270274$&$ 1656715$&$13626775$&$ 25399071$&$ 62$&$3033017$&$ 8.22$\tabularnewline
+MidCentral&$ 4628$&$ 92560$&$ 7001336$&$14905287$&$ 2663763$&$21909922$&$ 40838105$&$ 75$&$3668972$&$10.86$\tabularnewline
+Nelson Marlborough&$ 2356$&$ 47120$&$ 3699436$&$ 7875805$&$ 1407506$&$11576985$&$ 21578448$&$ 90$&$4402766$&$ 4.85$\tabularnewline
+Northland&$ 3071$&$ 61420$&$ 4770436$&$10155878$&$ 1814984$&$14928562$&$ 27825484$&$ 70$&$3424374$&$ 7.98$\tabularnewline
+South Canterbury&$  893$&$ 17860$&$ 1407312$&$ 2996056$&$  535433$&$ 4404031$&$  8208714$&$ 72$&$3522213$&$ 2.32$\tabularnewline
+Southern&$ 8371$&$167420$&$12676712$&$26987711$&$ 4823045$&$39670397$&$ 73942017$&$102$&$4989802$&$14.34$\tabularnewline
+Tairawhiti&$ 1359$&$ 27180$&$ 2038841$&$ 4340531$&$  775708$&$ 6380333$&$ 11892361$&$ 47$&$2299223$&$ 5.11$\tabularnewline
+Taranaki&$ 2899$&$ 57980$&$ 4413156$&$ 9395259$&$ 1679052$&$13810495$&$ 25741508$&$ 68$&$3326535$&$ 7.61$\tabularnewline
+Waikato&$11331$&$226620$&$16981678$&$36152642$&$ 6460934$&$53142322$&$ 99052462$&$ 95$&$4647365$&$20.32$\tabularnewline
+Wairarapa&$  720$&$ 14400$&$ 1128869$&$ 2403272$&$  429495$&$ 3532673$&$  6584582$&$ 59$&$2886258$&$ 2.27$\tabularnewline
+Waitemata&$17291$&$345820$&$25809668$&$54946730$&$ 9819676$&$80768561$&$150545262$&$ 70$&$3424374$&$39.93$\tabularnewline
+West Coast&$  685$&$ 13700$&$ 1060936$&$ 2258647$&$  403649$&$ 3320083$&$  6188333$&$ 50$&$2445981$&$ 2.52$\tabularnewline
+Whanganui&$ 1378$&$ 27560$&$ 2121871$&$ 4517295$&$  807298$&$ 6640166$&$ 12376666$&$ 58$&$2837338$&$ 4.32$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Benefit--cost analyses with 20 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\label{table:cost20}
 \end{sidewaystable}
@@
 \begin{sidewaystable}\small
-\caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case); \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 %latex.default(costtable50, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-Auckland&$17920$&$896000$&$26546459$&$55634145$&$81779021$&$ 9942525$&$152792323$&$ 82$&$4020980$&$31.1$\tabularnewline
-Bay of Plenty&$ 4585$&$229250$&$ 7188051$&$15064196$&$22143509$&$ 2692162$&$ 41371958$&$ 71$&$3481580$&$11.1$\tabularnewline
-Canterbury&$13687$&$684350$&$21039340$&$44092724$&$64813791$&$ 7879928$&$121095235$&$ 62$&$3040253$&$32.5$\tabularnewline
-Capital and Coast&$10461$&$523050$&$15678760$&$32858409$&$48299988$&$ 5872213$&$ 90241571$&$ 96$&$4707488$&$17.3$\tabularnewline
-Counties Manukau&$18880$&$944000$&$28032290$&$58748043$&$86356274$&$10499018$&$161344260$&$ 50$&$2451817$&$47.5$\tabularnewline
-Hawke's Bay&$ 3751$&$187550$&$ 5832570$&$12223478$&$17967816$&$ 2184490$&$ 33570276$&$ 56$&$2746035$&$11.4$\tabularnewline
-Hutt Valley&$ 4388$&$219400$&$ 6676018$&$13991115$&$20566142$&$ 2500389$&$ 38424874$&$ 86$&$4217125$&$ 8.7$\tabularnewline
-Lakes&$ 2886$&$144300$&$ 4423416$&$ 9270274$&$13626775$&$ 1656715$&$ 25459666$&$ 62$&$3040253$&$ 8.0$\tabularnewline
-MidCentral&$ 4628$&$231400$&$ 7112226$&$14905287$&$21909922$&$ 2663763$&$ 40935534$&$ 75$&$3677725$&$10.5$\tabularnewline
-Nelson Marlborough&$ 2356$&$117800$&$ 3758029$&$ 7875805$&$11576985$&$ 1407506$&$ 21629928$&$ 90$&$4413270$&$ 4.8$\tabularnewline
-Northland&$ 3071$&$153550$&$ 4845992$&$10155878$&$14928562$&$ 1814984$&$ 27891869$&$ 70$&$3432544$&$ 7.8$\tabularnewline
-South Canterbury&$  893$&$ 44650$&$ 1429602$&$ 2996056$&$ 4404031$&$  535433$&$  8228297$&$ 72$&$3530616$&$ 2.3$\tabularnewline
-Southern&$ 8371$&$418550$&$12877490$&$26987711$&$39670397$&$ 4823045$&$ 74118424$&$102$&$5001706$&$13.7$\tabularnewline
-Tairawhiti&$ 1359$&$ 67950$&$ 2071133$&$ 4340531$&$ 6380333$&$  775708$&$ 11920734$&$ 47$&$2304708$&$ 5.0$\tabularnewline
-Taranaki&$ 2899$&$144950$&$ 4483054$&$ 9395259$&$13810495$&$ 1679052$&$ 25802921$&$ 68$&$3334471$&$ 7.4$\tabularnewline
-Waikato&$11331$&$566550$&$17250640$&$36152642$&$53142322$&$ 6460934$&$ 99288775$&$ 95$&$4658452$&$19.0$\tabularnewline
-Wairarapa&$  720$&$ 36000$&$ 1146748$&$ 2403272$&$ 3532673$&$  429495$&$  6600291$&$ 59$&$2893144$&$ 2.3$\tabularnewline
-Waitemata&$17291$&$864550$&$26218451$&$54946730$&$80768561$&$ 9819676$&$150904424$&$ 70$&$3432544$&$35.1$\tabularnewline
-West Coast&$  685$&$ 34250$&$ 1077739$&$ 2258647$&$ 3320083$&$  403649$&$  6203097$&$ 50$&$2451817$&$ 2.5$\tabularnewline
-Whanganui&$ 1378$&$ 68900$&$ 2155478$&$ 4517295$&$ 6640166$&$  807298$&$ 12406193$&$ 58$&$2844108$&$ 4.3$\tabularnewline
-\hline
-\end{tabular}\end{center}\label{table:cost50}
+Auckland&$17920$&$896000$&$26132561$&$55634145$&$ 9942525$&$81779021$&$152428668$&$ 82$&$4011409$&$31.06$\tabularnewline
+Bay of Plenty&$ 4585$&$229250$&$ 7075979$&$15064196$&$ 2692162$&$22143509$&$ 41273490$&$ 71$&$3473294$&$11.15$\tabularnewline
+Canterbury&$13687$&$684350$&$20711307$&$44092724$&$ 7879928$&$64813791$&$120807021$&$ 62$&$3033017$&$32.50$\tabularnewline
+Capital and Coast&$10461$&$523050$&$15434306$&$32858409$&$ 5872213$&$48299988$&$ 90026791$&$ 96$&$4696284$&$17.25$\tabularnewline
+Counties Manukau&$18880$&$944000$&$27595227$&$58748043$&$10499018$&$86356274$&$160960251$&$ 50$&$2445981$&$47.48$\tabularnewline
+Hawke's Bay&$ 3751$&$187550$&$ 5741632$&$12223478$&$ 2184490$&$17967816$&$ 33490377$&$ 56$&$2739499$&$11.44$\tabularnewline
+Hutt Valley&$ 4388$&$219400$&$ 6571929$&$13991115$&$ 2500389$&$20566142$&$ 38333420$&$ 86$&$4207088$&$ 8.66$\tabularnewline
+Lakes&$ 2886$&$144300$&$ 4354448$&$ 9270274$&$ 1656715$&$13626775$&$ 25399071$&$ 62$&$3033017$&$ 7.99$\tabularnewline
+MidCentral&$ 4628$&$231400$&$ 7001336$&$14905287$&$ 2663763$&$21909922$&$ 40838105$&$ 75$&$3668972$&$10.47$\tabularnewline
+Nelson Marlborough&$ 2356$&$117800$&$ 3699436$&$ 7875805$&$ 1407506$&$11576985$&$ 21578448$&$ 90$&$4402766$&$ 4.77$\tabularnewline
+Northland&$ 3071$&$153550$&$ 4770436$&$10155878$&$ 1814984$&$14928562$&$ 27825484$&$ 70$&$3424374$&$ 7.78$\tabularnewline
+South Canterbury&$  893$&$ 44650$&$ 1407312$&$ 2996056$&$  535433$&$ 4404031$&$  8208714$&$ 72$&$3522213$&$ 2.30$\tabularnewline
+Southern&$ 8371$&$418550$&$12676712$&$26987711$&$ 4823045$&$39670397$&$ 73942017$&$102$&$4989802$&$13.67$\tabularnewline
+Tairawhiti&$ 1359$&$ 67950$&$ 2038841$&$ 4340531$&$  775708$&$ 6380333$&$ 11892361$&$ 47$&$2299223$&$ 5.02$\tabularnewline
+Taranaki&$ 2899$&$144950$&$ 4413156$&$ 9395259$&$ 1679052$&$13810495$&$ 25741508$&$ 68$&$3326535$&$ 7.42$\tabularnewline
+Waikato&$11331$&$566550$&$16981678$&$36152642$&$ 6460934$&$53142322$&$ 99052462$&$ 95$&$4647365$&$19.00$\tabularnewline
+Wairarapa&$  720$&$ 36000$&$ 1128869$&$ 2403272$&$  429495$&$ 3532673$&$  6584582$&$ 59$&$2886258$&$ 2.25$\tabularnewline
+Waitemata&$17291$&$864550$&$25809668$&$54946730$&$ 9819676$&$80768561$&$150545262$&$ 70$&$3424374$&$35.10$\tabularnewline
+West Coast&$  685$&$ 34250$&$ 1060936$&$ 2258647$&$  403649$&$ 3320083$&$  6188333$&$ 50$&$2445981$&$ 2.50$\tabularnewline
+Whanganui&$ 1378$&$ 68900$&$ 2121871$&$ 4517295$&$  807298$&$ 6640166$&$ 12376666$&$ 58$&$2837338$&$ 4.26$\tabularnewline
+\hline
+\end{tabular}\end{center}\caption{Benefit--cost analyses with 50 dollars per vaccine.
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\label{table:cost50}
 \end{sidewaystable}

### 2015-03-22 maps/rworldmaps_measles_travel.R

@@ -434,7 +434,7 @@ dpdf <- data.frame(id = mp$id,
 
 head(dpdf)
 
-np1<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Immigration  (2013)") +
+np1<-ggplot(dpdf, aes(map_id = id)) +#ggtitle("Immigration  (0-19yrs, 2013)") +
   geom_map(aes(fill = immigration), map =world.ggmap) +
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
@@ -445,7 +445,7 @@ pdf(paste("2013np1.pdf"), width=10, height=5)
 np1
 dev.off()
 
-np2<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Incidence (per million) (2013)") +
+np2<-ggplot(dpdf, aes(map_id = id)) +#ggtitle("Incidence (per million) (2013)") +
   geom_map(aes(fill = incidence), map =world.ggmap) +
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
@@ -456,7 +456,7 @@ pdf(paste("2013np2.pdf"), width=10, height=5)
 np2
 dev.off()
 
-np3<-ggplot(dpdf, aes(map_id = id)) +ggtitle("Vaccination cover (%) (2013)") +
+np3<-ggplot(dpdf, aes(map_id = id)) +#ggtitle("Vaccination cover (%) (2013)") +
   geom_map(aes(fill = cover), map =world.ggmap) +
   expand_limits(x = world.ggmap$long, y = world.ggmap$lat) +
   xlab("longitude") +
@@ -472,10 +472,10 @@ grid.arrange(np1, np2, np3, ncol=3, main="Measles (2013)")
 dpdf$risk <- with(dpdf, mp$X2013.x/mp$Numeric.Value*1000* mp$Immigration)
 head(dpdf)
 #####write.csv(dpdf, "mapdata.csv", row.names=FALSE,col.names=T)
-#####write.csv(dpdf, "mapdata_immigration.csv", row.names=FALSE,col.names=T)
+write.csv(dpdf, "2013mapdata_immigration.csv", row.names=FALSE,col.names=T)
 #####dpdf<-read.csv("mapdata.csv",header=T)
 
[…]

### 2015-03-22 revised_draft.R

@@ -563,14 +563,14 @@ DOBVac<-t(DOBVac)
 
 pdf(paste("dob_vacc_status.pdf"), width=7, height=5)
 par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=1.5)
+par(cex.axis=1)
 barplot(t(DOBVac),xlab="Age",col=c("darkgrey","red","orange"),main="",ylab="Frequency",cex.lab=1)
 legend("topleft",c("Unvaccinated","Dose 1","Dose 2"),fill=c("darkgrey","red","orange"),cex=1,bty="n")
 dev.off()
 
 pdf(paste("dob_vacc_dose.pdf"), width=7, height=5)
 par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=1.5)
+par(cex.axis=1)
 par(mfrow=c(2,1))
 barplot(DOBVac[,2],col=rgb(0,0,1,1/4),xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
 legend("topleft",c("Dose 1"),col=rgb(0,0,1,1/4),pch=15,bty="n",cex=1)
@@ -582,10 +582,10 @@ dev.off()
 
 pdf(paste("dob_vacc_unvac.pdf"), width=7, height=5)
 par(mar=c(5,6,4,2)+0.1)
-par(cex.axis=1.5)
+par(cex.axis=1)
 par(mfrow=c(1,1))
-barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1.5,ylab="Frequency")
-legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1.5)
+barplot(DOBVac[,1],col="grey",xlab="Date of Birth",main="",cex.lab=1,ylab="Frequency")
+legend("topleft",c("Unvaccinated"),col="grey",pch=15,bty="n",cex=1)
 dev.off()
 
 popimmune<-read.csv("PopnImmunityAll.csv",header=T)

### 2015-03-22 vaccinationmap.R

@@ -46,7 +46,7 @@ diff <- dose2_num$total - dose2_num$vacc
 
 dose1_merge <- merge(dose1, denom, by.x="cau", by.y="denom", all=T)
 diff <- dose1_merge[,2:8] - dose1_merge[,2:8+7]
-image(as.matrix(diff > 0))
+# image(as.matrix(diff > 0))
 
 au_new <- merge(au_data, dose2_num, by.x="AU2013", by.y="cau", all.x=T)
 au_new$prop <- au_new$vacc / au_new$total
@@ -70,7 +70,7 @@ pal<-c("red","orange","grey","blue","green","yellow")
 cols <- findColours(breaks, pal)
 par(mfrow=c(1,1))
 pdf(paste("nir_census.pdf"), width=7, height=5)
-plot(au, col=cols)
+plot(au, col=cols,lwd=1,lty=3)
 legend("topright",c("0-60%","61-80%","81-100%","101-120%","121-140%",">141%","NA")
        ,fill=c("red","orange","grey","blue","green","yellow","white"),
        bty="n",inset=0.1,title="Percent vaccinated")

### 2015-03-23 draftfinalreport_v2.Rnw
Final edits

@@
 We identify additional analyses that may help understand risk of infection in New Zealand.
-Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007 until mid-2014.
+Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007 until June 2014.
 We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level.
@@
 \item The regression analyses suggest that age is a particularly strong risk factor for measles.
-People of European ethnicity make up the majority of cases.
+Increased wealth (NZDep 1--3) was a risk factor and people of European ethnicity make up the majority of cases.
 However, our analyses also highlight other groups that are at greater risk of measles on a per capita basis.
-Increased wealth (NZDep 1--3) was a risk factor.
 Pacific people are at greater risk per capita.
 However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
-\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations.
+\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations nationally.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
-\item Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December.
+Peak overall overseas travel rates, and thus presumably risk from measles importation, is in December.
 However, New Zealander and immigrant or non--New Zealander travel is otherwise out of phase, with peak travel for New Zealanders during winter, and non--New Zealanders during summer.
@@
 The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92.
-This analysis suggests improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
-\item The cost of 2014 measles cases is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
+These analyses suggest improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
+\item The cost of 187 measles cases in 2014 is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
 \item The mean wage loss per measles case is estimated to be approximately \$839.
@@
 \item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$8,214 per vaccine before the costs exceed the benefits.
-\item Measles introductions to New Zealand with a median size of 2 cases per year were predicted by simulation following catch up vaccination to ensure $R_v$ is < 1.
+\item Following catch up vaccination to ensure $R_v$ is < 1 measles introductions to New Zealand with a median size of 2 cases were predicted by simulation.
 Thus, typically individual cases and secondary cases will be the expectation.
-However, the mean of 61 cases was predicted, because peak outbreak sizes of tens, hundreds, and very rarely up to many thousands are predicted following importation, despite $R_v$ being one and the outbreak predicted to die out following additional vaccinations.
-Thus, increased vaccination beyond the 28\% of the currently 11\% na\'{\i}ve population required may be useful.
+However, the mean outbreak size of 61 cases was predicted, because outbreak sizes of tens, hundreds, and very rarely up to many thousands of cases are predicted among the 1000 simulations following importation, despite $R_v$ being one and the outbreak predicted to die out.
+Thus, increased vaccination beyond the 28\% of the currently 11\% na\'{\i}ve population required may be useful to prevent these rare but costly events.
 \end{itemize}
@@
 Following discussion with the Ministry, these were 0--1, 2--4, and 5 years old for each of the University of Otago denominator data, assuming equal numbers of young were born into each age group over the last five years (which is supported by data from NZ statistics \citep{stats14}).
-The 5 year olds were added to the 6--17 year old category.
-No breakdown of 6--17 year category was performed as constant birth rates for the period could not be assumed.
+The 5 year olds were added to the 6--17 year old category to make a 5--17 year old category.
+No breakdown of 5--17 year category was performed as constant birth rates for the period could not be assumed.
 
@@
 \end{center}
-\caption{Per capita measles values from 2007--2014 (\autoref{table:percap}). The x-axis shows NZDep:Age Range:Prioritized ethnicity. Absolute cases are in \autoref{fig:cases_all}}
+\caption{Per capita measles values from 2007--2014 (\autoref{table:percap}). The x-axis shows NZDep:Age Range:Prioritized ethnicity. Numbers of cases are in \autoref{fig:cases_all}}
 \label{fig:percap_all}
@@
 The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{fig:percap_all}, \autoref{table:percap}), leading to very large standard error in regression analyses.
-However, there are numerous issues with the data for MLA ethnicity category above and beyond the small population sizes (see \autoref{table:percap}), creating issues for estimating the denominator data for this group (University of Otago, personal communication).
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard error in regression analyses.
+However, there are also numerous issues with the data for MLA ethnicity category above and beyond the small population sizes, creating issues for estimating the denominator data for this group (University of Otago, personal communication).
 Thus, we removed this grouping for our subsequent analyses and are left with Asian, European, Maori and Pacific as Prioritized Ethnicities.
@@
 \begin{table}[hbtp] \scriptsize
+\captionsetup{width=1.2\textwidth}
 <<results=tex,echo=FALSE>>=
@@
 @
-\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
+\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014.
+MLA is Middle Eastern/Latin American/African}
 \label{table:percap}
@@
 @
-\caption{Summary of the regression model results, estimated using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-2014}
+\caption{Summary of the regression model results using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-June 2014}
 \label{table:regression_riskfactors}
@@
 However, per capita rates for this large age class were lower than in the very young (0--1 year olds, \autoref{fig:percap_all}).
-People of Pacific origin are over-represented as measles cases on a per capita basis ($\beta$ = 1.22, standard error (SE) = 0.21, p--value < 0.0001; \autoref{table:regression_riskfactors}), and in particular the 0--1 year old less wealthy (NZDep 8--10) (\autoref{fig:percap_all}.
-Note however that generally those people in NZDep levels 4--10 are less represented on a per capita basis (NZDep 4--7 $\beta$ = -0.26, SE = 0.09, p--value 0.009; NZDep 8--10 $\beta$ = -0.3, SE = 0.1, p--value == 0.006; \autoref{table:regression_riskfactors}), and in European and Maori ethnicities there is a trend in the opposite way than Pacific peoples, with the more wealthy more likely to be cases (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
+People of Pacific origin are over-represented as measles cases on a per capita basis ($\beta$ = 1.22, standard error (SE) = 0.21, p--value < 0.0001; \autoref{table:regression_riskfactors}), and in particular the 0--1 year old less wealthy (NZDep 8--10) (\autoref{fig:percap_all}).
+Note however that generally those people in NZDep levels 4--10 are less represented on a per capita basis (NZDep 4--7 $\beta$ = -0.26, SE = 0.09, p--value 0.009; NZDep 8--10 $\beta$ = -0.3, SE = 0.1, p--value == 0.006; \autoref{table:regression_riskfactors}), and in European and Maori ethnicities there is a trend opposite to that in than Pacific people, with the more wealthy more likely to be cases (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
 
@@
 In later measles outbreaks (since 2007) there has been a shift in the distribution of ages infected.
-The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented then the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}).
-This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap},\autoref{table:sero}, and \autoref{sub:popim}).
+The very young (<2 years old) are still most likely to be infected, but school aged children and older teenagers are more likely to be represented than the under tens (\autoref{fig:ageinyears} and \autoref{fig:ageandvac}).
+This pattern suggests that improving vaccination coverage in the young is reducing the burden of measles in those age categories (\autoref{table:percap}, \autoref{table:sero}, and \autoref{sub:popim}).
 Interestingly the regression results suggest risk of measles in the 5--17 year age category was greater for Europeans and Maori (\autoref{table:percap}, \autoref{table:regression_riskfactors}, \autoref{fig:percap_all}).
@@
 Note these values are from vaccination records and serosurveys.
-Vaccine efficacy for 0--5 year olds was 99\%, 6--13 year olds 96\%, and for \leq 14 equivocal serological results were considered non-immune}
+Vaccine efficacy for 0--5 year olds was 99\%, 6--13 year olds 96\%, and for $\leq$ 14 year olds equivocal serological results were considered non-immune}
 \label{table:sero}
@@
 A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
-Plotting the cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).
+Plotting the unvaccinated cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).
 
@@
 \end{center}
-\caption{Age of vaccination of vaccinated measles cases, 2007--2014. Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
+\caption{Age of vaccination of vaccinated measles cases, 2007- June 2014.
+Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
 \label{fig:vaccstat}
@@
 \end{center}
-\caption{Year of birth and vaccination status of measles cases, 2007--2014}
+\caption{Year of birth of unvaccinated of measles cases, 2007- June 2014}
 \label{fig:yearandvac}
@@
 \end{center}
-\caption{Year of birth and vaccination status of measles cases, 2007--2014}
+\caption{Year of birth of unvaccinated of measles cases, 2007- June 2014.
+The year of birth for those vaccinated with measles, mumps, and rubella (MMR) dose 1 is shown in the upper panel, MMR dose 2 the lower panel.}
 \label{fig:yearandvacf}
+\end{figure}
+
+\begin{figure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{naive_allPop.pdf}
+\end{center}
+\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (\autoref{table:sero})}
+\label{fig:naive}
+\end{figure}
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{naive_map.pdf}
+     \end{center}
+     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (\autoref{table:sero})}
+     \label{fig:naive_map}
 \end{figure}
@@
 \caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator.
-Note no values should be greater than 1 (100\% vaccine cover), and the maximum discordance in > 1600\%.}
+Note no values should be greater than 1 (100\% vaccine cover), and the maximum discordance is > 1600\%.}
 \label{fig:naive_au_prop}
@@
 
-\begin{figure}
-\begin{center}
-     \includegraphics[width=0.9\textwidth]{naive_allPop.pdf}
-\end{center}
-\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using national immunity data (\autoref{table:sero})}
-\label{fig:naive}
-\end{figure}
-
-\begin{figure}[H]
-     \begin{center}
-     \includegraphics[width=0.9\textwidth]{naive_map.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:naive_map}
-\end{figure}
-
 \subsection{Measles importation risk methods}
@@
 We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
-We use 0--19 year old data for the immigration figures, as this is the only category available and it was determined this was more appropriate than using all age classes as > 19 year olds are believe to be immune to measles more or less globally.
-These data were not available for New Zealand travellers, so we use all travel.
-We present the data from 2013 because this year was the most recent with the most complete WHO measles data, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
-Data for 2013 are incomplete, but substantially better than 2014.
+We use 0--19 year old data for the immigration figures, as no further details were available and it was determined this was more appropriate than using all age classes because > 19 year olds are believe to be immune to measles more or less globally.
+These age category data were not available for New Zealand travellers, so we use all data on all travellers.
+We present the data from 2013 because this year was the most recent with the relatively complete WHO measles data, and thus accounts for improved measles vaccination coverage following the United Nations Millenium Development Goals' improvements in measles vaccination coverage.
+Data for 2013 are less complete than 2012, but substantially better than 2014.
 
@@
 \caption{Measles and travel data from 2013 for the top 10 countries identified as high risk for measles importation.
-incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of 0--19 year old travellers entering New Zealand from the country}
+incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of New Zealander and 0--19 year old foreign national travellers entering New Zealand from the country}
 \label{table:toprisk}
@@
 In particular, Pacific people are at greater risk per capita, as are the less deprived (NZDep 1--3), and European and Maori 5--17 year old children compared to Asian and Pacific ethnicity children of the same age.
+Furthermore, Pacific origin people of all categories from 2--24 years old were less likely on a per capita basis to be cases than Europeans or Maori.
 Interpretation of these results must still be viewed with some caution, however, as there is very likely a spatial effect that might not be accounted for in these analyses.
@@
 However, the alternative situation was clear for Pacific people, especially of the 0--1 year olds (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
-Results such as this may simply due to the spatial effects due the communities the outbreaks were occuring in, rather than an effect of wealth.
+Results such as this may simply be due to the spatial effects in the communities the outbreaks were occuring in, rather than an effect of wealth.
 A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data.
@@
 
-The distribution of na\"{\i}ve among the DHBs also varies (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).
+The distribution of na\"{\i}ve among the DHBs also varies, though the proportion of na\"{\i}ve overall across DHBs is relatively similar (\autoref{sub:popim}, \autoref{fig:Northland}, \autoref{fig:Waitemata}, \autoref{fig:Auckland}, \autoref{fig:Counties_Manukau}, \autoref{fig:Waikato}, \autoref{fig:Lakes}, \autoref{fig:BayofPlenty}, \autoref{fig:Tairawhiti}, \autoref{fig:Taranaki}, \autoref{fig:HawkesBay}, \autoref{fig:Whanganui}, \autoref{fig:Midcentral}, \autoref{fig:Hutt}, \autoref{fig:CapitalandCoast}, \autoref{fig:Wairarapa}, \autoref{fig:NelsonMarlborough}, \autoref{fig:WestCoast}, \autoref{fig:Canterbury}, \autoref{fig:SouthCanterbury}, \autoref{fig:Southern}).
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
@@
 The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}).
-However, recently Australia was declared free of endemic measles and so the 2013 data must reflect imported measles there.
+Recently Australia was declared free of endemic measles and so the 2013 data must reflect imported measles there.
 Other interaction terms beyond the product of travel and infection incidence may be more appropriate but this will require substantial further analyses.
 Other areas of high risk include the United Kingdom, where measles persists, and South, Southeast and East Asia, where measles is endemic (\autoref{table:risk12} and \autoref{fig:risk12}).
-The 2012 data were more complete than the 2013 data, which suffers from data deficiencies, but due to the dynamic nature of measles epidemiology and consultation with the Ministry we present the 2013 data.
+The WHO 2012 data were more complete than the WHO 2013 data, which suffers from data deficiencies, but due to the dynamic nature of measles epidemiology and consultation with the Ministry we present the WHO 2013 data.
 Lastly, more refined data on the age of travellers may improve the risk assessment, but these are currently unavailable.
@@
 \item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
-\item There is some statistical support for 2--24 year old Pacific people being less at risk than European and Maori of the same age, as are 5--17 year old Asian children.
+\item There is statistical support for 2--24 year old Pacific people being less at risk than European and Maori of the same age, as are 5--17 year old Asian children.
 \item The majority of cases are unvaccinated.
 \item The majority of vaccine failures occur in those people which recieved single vaccinations around 1 year old.
-\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas.
+\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas, though the \%  na\"{\i}ve is similar across DHBs.
 \item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.
@@
 The schedule was changed in 2000 with MMR2 now being administered before 5 years \citep{anon2a}.
-Later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintain high coverage rates in order to prevent future epidemics and poor schedules were not considered \citep{roberts4}.
+Later analyses suggested high levels of vaccination coverage (but less than 95\%) could eliminate measles, but emphasised that it is necessary to maintain high coverage rates in order to prevent future epidemics \citep{roberts4}.
 
@@
 
+\begin{sidewaysfigure}
+\begin{center}
+     \includegraphics[width=0.9\textwidth]{R0_outbreak33.pdf}
+     \end{center}
+     \caption{Estimates of $R_t$ for the 2011--2012 outbreak, as classified by outbreak number in \autoref{fig:outbreaks}, demonstrating how  $R_t$ can vary and how outbreaks with low  $R_v$ can persist.
+95\% confidence intervals are shown}
+     \label{fig:r0_2011}
+\end{sidewaysfigure}
+
 To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we use the well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population \citep{diekmann13}:
@@
 Given the risk of importation of measles as highlighted in \autoref{sub:imp_risk} is likely to continue, these analyses suggest substantial efforts are required to increase immunisation to high enough levels that measles does not become endemic and large outbreaks prevented.
-The measles outbreaks in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months.
+One measles outbreak in 2011--2012 had an $R_v$ of just greater than one, and yet it persisted for over 12 months (\autoref{fig:r0_2011}).
 This implies that outbreaks may persist within the population for a substantial period, even if $R_v$ values are not much greater than one.
 A caveat to this and other $R_v$ estimates is that the outbreaks may include some sporadic cases misclassified, but also exclude some sporadic cases as yet not associated with another outbreak.
+Undetected and unreported cases may lower the estimate.
 Thus the true basic reproduction numbers may differ to those estimated.
-However, undetected cases may lower the estimate.
 The relative contributions of both to our estimates are currently unknown.
@@
 The mean number of contacts per case was 54 (median = 28, maximum = 206).
-The estimated cost to the public health unit for contact management for the epidemic was in excess of AUS\$48,000,{Jan 2005 \$ to AUS\$ rate 0.7692: \$62,400} with 90\% of this being associated with staff time.
+The estimated cost to the public health unit for contact management for the epidemic was in excess of AUS\$48,000,\footnote{Jan 2005 \$ to AUS\$ rate 0.7692: \$62,400} with 90\% of this being associated with staff time.
 
@@
 \caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
+Note \textit{Length of stay} is number of complete days spent in hospital, thus 0 is no nights spent in hospital}
 \label{table:cases}
@@
 
-In total, with additional \$20 costs per GP consultation, we estimate that the 2014 measles cases alone cost the New Zealand economy \$1041185.63.
+In total, with additional \$20 costs per GP consultation, we estimate that the 187 measles cases in 2014 cost the New Zealand economy \$1041185.63.
 
@@
 The cost figures above are used to estimate the savings for vaccinating additional populations, using estimated per case costs saved.
-The costs of the catch up vaccination schemes are estimated to be two different values, to determine how sensitive the benefit--cost ratio ((B/C) was to differing vaccination costs.
+The costs of the catch up vaccination schemes are estimated to be two different values, to determine how sensitive the benefit--cost ratio (B/C) was to differing vaccination costs.
 Values of \$20 and \$50 per vaccine are used, based on US literature, however, this assumes all costs are included with the vaccine delivery.
@@
 become endemic or cause an outbreak), outbreaks can occur due to stochastic processes following measles importation.
-The most likely scenario is that very few cases occur, with the median value from 1000 simulations low (2 cases).
-Thus most measles introductions will be single cases or lead to minor outbreaks with single secondary cases.
+The most likely scenario is that very few cases occur, with the median value from 1000 simulations just 2 cases.
+Thus most measles introductions will be single cases or lead to single secondary cases.
 In this modelling exercise the mean value was 61 cases, suggesting there is likely to be an outbreak of a larger size given enough introductions.
@@
      \caption{A subset of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1.
-The full distribution of rare results can be seen in \autoref{fig:sim1}}
+The full distribution of rare results can be seen in \autoref{fig:sim1}.}
      \label{fig:sim}
@@
   \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
-The full distribution of common results can be seen in \autoref{fig:sim}}
+Note this figure does not show the frequency of common events as the y-axis was limited to 20 to show rare events.
+The full distribution of common results can be seen in \autoref{fig:sim}.}
      \label{fig:sim1}
@@
 \caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
-Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{Note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small.
+The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost20}
@@
 \caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
-Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{Note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small.
+The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost50}
@@
 The model estimates for each DHB, with the vaccination percentages and the expected outbreak size following additional vaccination is shown in \autoref{table:vacp}.
-The benefit--cost results are in \autoref{table:cost20} and \autoref{table:cost50}, for two different vaccine prices.
+The benefit--cost results are in \autoref{table:cost20} and \autoref{table:cost50}, for two different vaccine costs.
 The results in the two tables show the benefits of vaccination are always substantially greater than the costs of the increased supplementary vaccination (\autoref{table:cost20} and \autoref{table:cost50}).
@@
 Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children.
-However, there are a number of different vaccination schemes that will lead to sufficient coverage and these are given in \autoref{table:cohort_vac}.
+However, there are a number of different vaccination schemes that will lead to sufficient coverage and these are given in \autoref{table:coh_vac}.
 The appropriate cover could be reached by catch up vaccination of all school aged children.
@@
 \end{tabular}
+\end{center}
 \caption{Proportion of the currently na\"{\i}ve New Zealand population that would be vaccinated if all na\"{\i}ve people in each age range were vaccinated.
 The estimated additional proportion of the population requiring vaccination is 28\%}
-\end{center}
-\label{table:cohort_vac}
+\label{table:coh_vac}
 \end{table}%
@@
 As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis.
+Our results are relatively insensitive to changes in vaccination cost, requiring very high costs (>\$8,214 per vaccine) before vaccination is predicted to not be beneficial.
+Similarly, longer discounting periods or 5\% discounting did not affect reduce the B/C ration substantially (not shown).
+A spreadsheet is provided for the Ministry to work with.
 

### 2015-03-23 draftfinalreport_v2.tex
Final edits

@@
 \begin{table}[hbtp] \scriptsize
+\captionsetup{width=1.2\textwidth}
 %latex.default(PerCap, file = "", table.env = FALSE, rowname = NULL)%
@@
 \hline
-\end{tabular}\end{center}\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014}
+\end{tabular}\end{center}\caption{Measles cases (Cases), population sizes (Population) and per capita rates per 10,000 (Per capita) of measles in specific age (Age), ethnicity (Ethnicity) and socio-economic deprivation (NZDep) categories from 2007-2014.
+MLA is Middle Eastern/Latin American/African}
 \label{table:percap}
@@
 0-44&0.89\tabularnewline
-\textgreater8&0.8\tabularnewline
+\textgreater 8&0.8\tabularnewline
 \hline
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Sun Mar 22 17:09:42 2015
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Mon Mar 23 13:50:01 2015
 \begin{tabular}{lrrrrrr}
@@
 \begin{table}
-% latex table generated in R 3.0.2 by xtable 1.7-1 package
-% Sun Mar 22 17:09:42 2015
+% latex table generated in R 3.0.2 by xtable 1.7-3 package
+% Mon Mar 23 13:50:01 2015
 \begin{tabular}{rrrrr}
@@
    \hline
-\end{tabular}\caption{Summary of the regression model results, estimated using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-2014}
+\end{tabular}\caption{Summary of the regression model results using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-June 2014}
 \label{table:regression_riskfactors}
@@
 \end{tabular}\end{center}\caption{Measles and travel data from 2013 for the top 10 countries identified as high risk for measles importation.
-incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of 0--19 year old travellers entering New Zealand from the country}
+incidence: measles incidence per million; cover: percent vaccination cover (\% children that receive one dose of measles vaccine by their second birthday.); immigration: the number of New Zealander and 0--19 year old foreign national travellers entering New Zealand from the country}
 \label{table:toprisk}
@@
 \end{tabular}\end{center}\caption{Number of cases, length of hospital stay, cost (NZ\$), cost per case (NZ\$) and cost per day (NZ\$) for patients with measles as the primary diagnosis, by year and gender, 2000--2014.
-Note "Length of stay" is number of complete days spent in hospital, thus 0 is no nights spent in hospital (see \autoref{table:regression_costs})}
+Note \textit{Length of stay} is number of complete days spent in hospital, thus 0 is no nights spent in hospital}
 \label{table:cases}
@@
 \end{tabular}\end{center}\caption{Benefit--cost analyses with 20 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
-Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}	is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{Note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small.
+The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost20}
@@
 \end{tabular}\end{center}\caption{Benefit--cost analyses with 50 dollars per vaccine.
-\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
-Median values are all 2, thus the expected number of cases will be small}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+\textit{Vacc} is numbers to vaccinate (see \autoref{table:attack}); \textit{Vacc costs} is cost for the catch up vaccination programme; \textit{Wages saved} is wages of care givers and cases saved (\$839 per case); \textit{Contacts} is costs saved through contact exclusion; \textit{Manage saved} is management costs saved (\$1,765 per case), plus \$20 GP costs; \textit{Hosp saved} is the hospitalisation costs saved (\$1,877 per case); \textit{Costs save}  is the discounted costs saved; \textit{Outbreak}  is the predicted mean outbreak size despite $R_v < 1$ due to measles importation from 1000 simulations\footnote{Note the mean takes into account rare, but large outbreaks, see \autoref{fig:sim} and \autoref{fig:sim1}.
+Median values are all 2, thus the expected number of cases will be small.
+The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost50}

### 2015-04-21 all_pop_map.R
lots of new maps with age and sero data
city and national levels

@@ -1,0 +1,172 @@
+library(RColorBrewer)
+library(classInt)
+
+data_denom<-read.csv("2013-mb-dataset-Total-New-Zealand-individual-part-1_AU.csv",header=T)
+
+popimmune<-read.csv("PopnImmunityAll.csv",header=T)
+popimmune$Age = factor(popimmune$Age,levels(popimmune$Age)[c(2,3,6,8,10:12,4,5,7,9,1)])
+
+data_denom$AGE_10_14_im <- 4/5 * data_denom$AGE_10_14 * popimmune[7,3] + 1/5  * data_denom$AGE_10_14 * popimmune[8,3]
+data_denom$AGE_15_19_im <- 4/5 * data_denom$AGE_15_19 * popimmune[8,3] + 1/5  * data_denom$AGE_15_19 * popimmune[9,3]
+data_denom$AGE_20_24_im <- 4/5 * data_denom$AGE_20_24 * popimmune[9,3] + 1/5  * data_denom$AGE_20_24 * popimmune[10,3]
+data_denom$AGE_25_29_im <- 4/5 * data_denom$AGE_25_29 * popimmune[10,3] + 1/5  * data_denom$AGE_25_29 * popimmune[11,3]
+data_denom$AGE_30_34_im <-  data_denom$AGE_30_34  * popimmune[11,3]
+data_denom$AGE_35_39_im <-  data_denom$AGE_35_39  * popimmune[11,3]
+data_denom$AGE_40_44_im <-  data_denom$AGE_40_44  * popimmune[11,3]
+data_denom$AGE_45_49_im <-  data_denom$AGE_45_49  * popimmune[11,3]
+data_denom$AGE_50_54_im <-  3/5 * data_denom$AGE_50_54 * popimmune[11,3] + 2/5  * data_denom$AGE_10_14 * popimmune[12,3]
+data_denom$AGE_55_59_im <-  data_denom$AGE_55_59  * popimmune[12,3]
+data_denom$AGE_60_64_im <-  data_denom$AGE_60_64 * popimmune[12,3]
+data_denom$AGE_65_OVER_im <- data_denom$AGE_65_OVER * popimmune[12,3]
+
+# data_denom$AGE_10_14_pc <- data_denom$AGE_10_14_im / data_denom$AGE_10_14
+# data_denom$AGE_15_19_pc <- data_denom$AGE_15_19_im / data_denom$AGE_15_19
[…]

### 2015-04-21 new_denom_data.R

@@ -17,14 +17,14 @@ data_n <- data_n %>% mutate(Year = year(dmy(DATE_OF_BIRTH)))
 grouped_n1 <- data_n %>% filter(VACCINE_DOSE == 1) %>% group_by(DOMICILE_CODE, Year) %>% summarize(vacc1 = length(unique(UNIQUE_PATIENT_IDENTIFIER)))
 grouped_n2 <- data_n %>% filter(VACCINE_DOSE == 2) %>% group_by(DOMICILE_CODE, Year) %>% summarize(vacc2 = length(unique(UNIQUE_PATIENT_IDENTIFIER)))
 
-####
+#############################################################################
 ## from here for each new year
 ###
 
 grouped <- grouped_d %>% left_join(grouped_n1) %>% left_join(grouped_n2)
 
 ## 2006 -- 2014
-grouped <- grouped[grouped$Year == 2014,]
+grouped <- grouped[grouped$Year == 2006,]
 
 ####
 grouped_all_years <- grouped %>% group_by(DOMICILE_CODE) %>% summarize(total=sum(total), vacc1=sum(vacc1, na.rm=T), vacc2=sum(vacc2, na.rm=T))
@@ -42,19 +42,14 @@ data_au <- data_au %>% left_join(grouped_all_years)
 
 in_the_sea <- grepl("^Oceanic", data_au$AU2013_NAM)
 
-pdf(paste("nir_census_MMR1_NIR_2014_cor.pdf"), width=7, height=7)
-plot(data_au$prop1,data_au$prop2,xlab="MMR1",ylab="MMR2")
-txt<-c(round(cor(data_au$prop1,data_au$prop2,use="complete.obs"),2))
-text(x=c(0,0.05),y=c(1.1,1.1),c(expression("R = "),txt))
-dev.off()
 ## plot and save
 
 fixedBreaks <- c(0,0.4,0.5,0.6,0.7,0.8,0.9,0.95,1,Inf)
 
-####
+################
 
-breaks <- cut(data_au$prop1, fixedBreaks)
-# breaks <- cut(data_au$prop2, fixedBreaks)
+# breaks <- cut(data_au$prop1, fixedBreaks)
+ breaks <- cut(data_au$prop2, fixedBreaks)
 
 ####
 
@@ -66,8 +61,8 @@ par(mfrow=c(1,1))
 
 #####
 
-pdf(paste("nir_census_MMR1_NIR_2014.pdf"), width=7, height=5)
[…]

### 2015-04-24 draftfinalreport_v3.Rnw
update introduction

--- 2015-03-23 draftfinalreport_v2.Rnw
@@
 \usepackage{authblk}
+\usepackage{morefloats}
 \newcommand{\vecA}{\mathbf{A}}
@@
 New Zealand has been working towards elimination of endemic (domestic) measles virus transmission, but has suffered from small, yet significant outbreaks after measles introductions from abroad.
-In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region.
-We identify additional analyses that may help understand risk of infection in New Zealand.
-Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks since 2007 until June 2014.
+In this draft final report we review the draft \emph {Progress Towards Measles Elimination in New Zealand - Final} report from the New Zealand Ministry of Health to the World Health Organization (WHO) Western Pacific Region, and identify additional analyses that may help understand risk of infection in New Zealand.
+Here we present the results of statistical analyses of risk factors for measles cases in New Zealand during outbreaks from 2007 to 12~June~2014.
 We provide cost analyses for the measles outbreaks in New Zealand, and include modelling of measles outbreaks, including pre- and post-vaccination scenarios, based on the numbers of na\"{\i}ve people at the District Health Board (DHB) and national level.
-We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve different vaccination coverage levels.
+We provide benefit--cost analyses using the results from those model simulations, along with a number of alternative vaccination strategies to achieve required vaccination coverage levels.
 Our key findings are:
@@
 However, this was driven by young Pacific children, as 2--24 year old Pacific people are less at risk than European and Maori, as are 5--17 year old Asian children.
-\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 131,500 vaccinations nationally.
+\item Of the 11\% of the overall population of New Zealand that are currently considered to be immunologically na\"{\i}ve, it was estimated that approximately 28\% would require additional vaccination to ensure measles does not persist, leading to an additional 128,034 vaccinations nationally.
 \item New Zealand is at risk of frequent measles importation due to travel and endemic measles elsewhere in the globe.
@@
 \item Analyses of outbreak data suggest that measles basic reproduction number ($R_v$, the number of secondary infections in a partially vaccinated population) values often include or exceed one, including 2013--2014 outbreaks.
-As analysed from data until June, the last outbreak to end in 2014 had an $R_v$ well above 1 (mean estimate 3.66, range 3.4--3.92).
+As analysed from data until 12 June 2014, the last outbreak to end in 2014 had an $R_v$ well above 1 (mean estimate 3.66, range 3.4--3.92).
 The range of estimates for $R_v$ between 2009--2014 was 0.18--3.92.
 These analyses suggest improved vaccination is a requisite to prevent measles persisting in the population and that large outbreaks could occur.
-\item The cost of 187 measles cases in 2014 is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
+\item The cost of the first 187 measles cases in 2014 is estimated to be approximately \$1,041,186 due to earnings lost relating to cases and contacts, case management and hospitalisation costs.
 \item The mean wage loss per measles case is estimated to be approximately \$839.
-\item The mean cost per measles case attending hospital is estimated to be \$1,877 per case, and approximately 17\% of measles cases attend hospital
+\item The mean cost of those measles case that attend hospital is estimated to be \$1,877 per attending case. Approximately 17\% of measles cases attend hospital.
 \item The mean public health service cost per case is \$1,765 excluding hospitalisation costs.
-\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170.4 per day.
-\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$8,214 per vaccine before the costs exceed the benefits.
+\item The average number of contacts requiring quarantine was 2.11, requiring 7.3 days of quarantine on average, at a cost of \$170 per day.
+\item The benefit--cost (B/C) ratio analyses suggest additional vaccination is extremely beneficial financially (B/C $>$1), with vaccine costs required to exceed approximately \$5,365 per vaccine before the costs exceed the benefits for all DHBs.
 \item Following catch up vaccination to ensure $R_v$ is < 1 measles introductions to New Zealand with a median size of 2 cases were predicted by simulation.
@@
 However, the mean outbreak size of 61 cases was predicted, because outbreak sizes of tens, hundreds, and very rarely up to many thousands of cases are predicted among the 1000 simulations following importation, despite $R_v$ being one and the outbreak predicted to die out.
-Thus, increased vaccination beyond the 28\% of the currently 11\% na\'{\i}ve population required may be useful to prevent these rare but costly events.
+Thus, increased vaccination beyond the 28\% of the currently 11\% na\"{\i}ve population required may be useful to prevent these rare but costly events.
 \end{itemize}
@@
 Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
-However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onwards, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
+However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onward, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
 This is particularly in light of our modelling and risk analyses results (\autoref{sub:risk_analyses} and \autoref{section:epidemic_modelling}).
@@
 
-# proportions of the naive population the wuld be covered if vaccinated by age groups now
+# proportions of the naive population the would be covered if vaccinated by age groups now
 propvCohort<-c(
@@
 
+t12006<-read.csv("MMR1summary2006.txt",header=T,sep="")
+t22006<-read.csv("MMR2summary2006.txt",header=T,sep="")
+t12007<-read.csv("MMR1summary2007.txt",header=T,sep="")
+t22007<-read.csv("MMR2summary2007.txt",header=T,sep="")
+t12008<-read.csv("MMR1summary2008.txt",header=T,sep="")
+t22008<-read.csv("MMR2summary2008.txt",header=T,sep="")
+t12009<-read.csv("MMR1summary2009.txt",header=T,sep="")
+t22009<-read.csv("MMR2summary2009.txt",header=T,sep="")
+t12010<-read.csv("MMR1summary2010.txt",header=T,sep="")
+t22010<-read.csv("MMR2summary2010.txt",header=T,sep="")
+t12011<-read.csv("MMR1summary2011.txt",header=T,sep="")
+t22011<-read.csv("MMR2summary2011.txt",header=T,sep="")
+t12012<-read.csv("MMR1summary2012.txt",header=T,sep="")
+t22012<-read.csv("MMR2summary2012.txt",header=T,sep="")
+t12013<-read.csv("MMR1summary2013.txt",header=T,sep="")
+t22013<-read.csv("MMR2summary2013.txt",header=T,sep="")
+t12014<-read.csv("MMR1summary2014.txt",header=T,sep="")
+t22014<-read.csv("MMR2summary2014.txt",header=T,sep="")
+
+table_nir_mmr<-rbind(
+t12006,
+t22006,
+t12007,
+t22007,
+t12008,
+t22008,
+t12009,
+t22009,
+t12010,
+t22010,
+t12011,
+t22011,
+t12012,
+t22012,
+t12013,
+t22013,
+t12014,
+t22014)
+colnames(table_nir_mmr)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+yr_nir<-c('MMR1 2006',
+'MMR2 2006',
+'MMR1 2007',
+'MMR2 2007',
+'MMR1 2008',
+'MMR2 2008',
+'MMR1 2009',
+'MMR2 2009',
+'MMR1 2010',
+'MMR2 2010',
+'MMR1 2011',
+'MMR2 2011',
+"MMR1 2012",
+"MMR2 2012",
+"MMR1 2013",
+"MMR2 2013",
+"MMR1 2014",
+"MMR2 2014")
+table_nir_mmr$Year <-yr_nir
+table_nir_mmr<-table_nir_mmr[c(10,1:7)]
+table_nir_mmr[2:7]<-table_nir_mmr[2:7]*100
+latex(table_nir_mmr, file="", table.env=FALSE,rowname=NULL)
+
+R0 = 12.8;
+x0max = 0.3;
+Nx0 = 200;
+P = (1:Nx0)*x0max/(Nx0+1);
+x0 = P/(1-exp(-R0*P));
+Rv = R0*x0;
+Pop = c(436350,206000,482180,283700,469300,151700,138380,98196,162560,137000,151690,
+        55620,297420,43650,109750,359310,41112,525550,32151,60120,4241739);
+Naive = c(51673,20511,50995,32406,55085,15471,15078,10470,17196,12960,14793,5199,
+          31389,4727,11378,39087,3901,57904,3172,6025,459420);
+RvDHB = R0*Naive/Pop;
+
+library(akima)
+PDHB = aspline(x=Rv,y=P,xout=RvDHB);
+
+# The number of cases per DHB
+FSDHB = PDHB$y*Naive;
+# The number of vaccinations short
+VDHB = Naive-Pop/R0;
+
+Z<-data.frame(nrows=21,ncols=5)
+for (i in 1:21){
+Z[i,1] = Pop[i];
+Z[i,2] = Naive[i];
+Z[i,3] = round(FSDHB[i]);
+Z[i,4] = round(VDHB[i]);
+Z[i,5] = round(round(VDHB[i])/Naive[i],2)}
+colnames(Z)<-c("Population","Naive","Outbreak","Vaccination","PC")
+
+write.csv(Z,"dhb_vacc.csv",row.names=F)
+
 @
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{case_ordert.pdf}
+     \includegraphics[width=0.8\textwidth]{case_ordert.pdf}
 \end{center}
@@
 \begin{center}
-     \includegraphics[width=0.9\textwidth]{percap_ordert.pdf}
+     \includegraphics[width=0.8\textwidth]{percap_ordert.pdf}
 \end{center}
@@
 \begin{table}[hbtp] \scriptsize
-\captionsetup{width=1.2\textwidth}
+\captionsetup{width=1\textwidth}
 <<results=tex,echo=FALSE>>=
@@
 
-\begin{table}[hbtp]\footnotesize
-<<results=tex,echo=FALSE>>=
-colnames(propvCohort) <- c("Age range", "Proportion")
-latex(propvCohort, file="", table.env=FALSE,rowname=NULL)
-@
-\caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
-\label{table:vaccine_cohort}
-\end{table}
+% \begin{table}[hbtp]\footnotesize
+% <<results=tex,echo=FALSE>>=
+% colnames(propvCohort) <- c("Age range", "Proportion")
+% latex(propvCohort, file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Proportions of the na\"{\i}ve population the would be covered if vaccinated by age groups}
+% \label{table:vaccine_cohort}
+% \end{table}
 
@@
 @
-\caption{Summary of the regression model results using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-June 2014}
+\caption{Summary of the regression model results using \autoref{eq:reg} and a quasipoisson error structure, for measles risk factors for cases from 2007-12 June 2014}
 \label{table:regression_riskfactors}
 \end{table}
-\vspace{5mm} %5mm vertical space
 
@@
 People of Pacific origin are over-represented as measles cases on a per capita basis ($\beta$ = 1.22, standard error (SE) = 0.21, p--value < 0.0001; \autoref{table:regression_riskfactors}), and in particular the 0--1 year old less wealthy (NZDep 8--10) (\autoref{fig:percap_all}).
-Note however that generally those people in NZDep levels 4--10 are less represented on a per capita basis (NZDep 4--7 $\beta$ = -0.26, SE = 0.09, p--value 0.009; NZDep 8--10 $\beta$ = -0.3, SE = 0.1, p--value == 0.006; \autoref{table:regression_riskfactors}), and in European and Maori ethnicities there is a trend opposite to that in than Pacific people, with the more wealthy more likely to be cases (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
+Note however that generally those people in NZDep levels 4--10 are less represented on a per capita basis (NZDep 4--7 $\beta$ = -0.26, SE = 0.09, p--value 0.009; NZDep 8--10 $\beta$ = -0.3, SE = 0.1, p--value = 0.006; \autoref{table:regression_riskfactors}), and in European and Maori ethnicities there is a trend opposite to that in than Pacific people, with the more wealthy more likely to be cases (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
 
@@
 \hline
-0  & 0\\
-1  & 0.89\\
+0  & 0.28\\
+1  & 0.67\\
 2  & 0.92\\
@@
 \end{center}
-\caption{Vaccination coverage (0--5 years) and serosurvey estimates of immunity ($>5$ years) among different age classes used in these analyses, as provided by the Ministry of Health.
+\caption{Vaccination coverage (0--13 years) and serosurvey estimates of immunity ($>5$ years) among different age classes used in these analyses, as provided by the Ministry of Health.
 Note these values are from vaccination records and serosurveys.
-Vaccine efficacy for 0--5 year olds was 99\%, 6--13 year olds 96\%, and for $\leq$ 14 year olds equivocal serological results were considered non-immune}
+Vaccine efficacy for 0--4 year olds was 96\%, 5--13 year olds 99\%, and for $\geq$ 14 year olds equivocal serological results were considered non-immune.
+Note that 28\% of those $<1$ were considered immune due to passive immunity, following \citep{waaijenborg}}
 \label{table:sero}
@@
 \end{center}
-\caption{Age of vaccination of vaccinated measles cases, 2007- June 2014.
+\caption{Age of vaccination of vaccinated measles cases, 2007- 12 June 2014.
 Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
@@
 \end{center}
-\caption{Year of birth of unvaccinated of measles cases, 2007- June 2014}
+\caption{Year of birth of unvaccinated of measles cases, 2007- 12 June 2014}
 \label{fig:yearandvac}
@@
 \end{center}
-\caption{Year of birth of unvaccinated of measles cases, 2007- June 2014.
+\caption{Year of birth of unvaccinated of measles cases, 2007- 12 June 2014.
 The year of birth for those vaccinated with measles, mumps, and rubella (MMR) dose 1 is shown in the upper panel, MMR dose 2 the lower panel.}
@@
 
-\begin{figure}[H]
+\begin{figure}
      \begin{center}
@@
 
-\begin{sidewaysfigure}
-\begin{center}
-     \includegraphics[width=0.9\textwidth]{nir_census.pdf}
-\end{center}
-\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register (NIR) level data and national census data as the denominator.
-Note no values should be greater than 1 in the frequency distribution (upper left) or 100\% on the map, demonstrating a lack of concordance between NIR and census data.
-The full distribution is shown in \autoref{fig:naive_au_prop}}
-\label{fig:naive_au}
-\end{sidewaysfigure}
-
-\begin{figure}
-\begin{center}
-     \includegraphics[width=0.9\textwidth]{naive_au_prop.pdf}
-\end{center}
-\caption{New Zealand population by age and estimated numbers of na\"{\i}ve people in each age class using National Immunisation Register level data and national census data as the denominator.
-Note no values should be greater than 1 (100\% vaccine cover), and the maximum discordance is > 1600\%.}
-\label{fig:naive_au_prop}
-\end{figure}
+ \vspace{5mm} %5mm vertical space
+\begin{table}
+<<include=FALSE,echo=FALSE,results=tex>>=
+latex(table_nir_mmr, file="", table.env=FALSE,rowname=NULL)
+@
+\caption{Summary of the proportion of 2006--2014 birth cohorts vaccinated against measles-mumps-rubella (MMR) dose 1 and MMR2 across the 2013 area units.
+Area units are aggregations of meshblock data.
+See \autoref{fig:fig12006}, \autoref{fig:fig22006}, \autoref{fig:fig12007}, \autoref{fig:fig22007}, \autoref{fig:fig12008}, \autoref{fig:fig22008}, \autoref{fig:fig12009}, \autoref{fig:fig22009}, \autoref{fig:fig12010}, \autoref{fig:fig22010}, \autoref{fig:fig12011}, \autoref{fig:fig22011}, \autoref{fig:fig12012}, \autoref{fig:fig22012}, \autoref{fig:fig12013}, \autoref{fig:fig22013}, \autoref{fig:fig12014}, and \autoref{fig:fig22014}.}
+\label{table:tab_mmr_nir}
+\end{table}
+
+\begin{sidewaysfigure}
+   \begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2006.pdf}
+    \end{center}
+    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig12006}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22006)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22006[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22006}.}
+% \label{table:tab22006}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2006.pdf}
+\end{center}
+    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig22006}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12007)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12007[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2007 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12007}.}
+% \label{table:tab12007}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+\includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2007.pdf}
+\end{center}
+    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig12007}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22007)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22007[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2007 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22007}.}
+% \label{table:tab22007}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2007.pdf}
+\end{center}
+    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig22007}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12008)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12008[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2008 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12008}.}
+% \label{table:tab12008}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2008.pdf}
+\end{center}
+    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig12008}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22008)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22008[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2008 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22008}.}
+% \label{table:tab22008}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2008.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig22008}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12009)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12009[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2009 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12009}.}
+% \label{table:tab12009}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2009.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig12009}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22009)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22009[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2009 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22009}.}
+% \label{table:tab22009}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2009.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig22009}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12010)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12010[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2010 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12010}.}
+% \label{table:tab12010}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2010.pdf}
+\end{center}
+    \caption{Proportion of the New Zealand population born in 2010 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig12010}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22010)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22010[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2010 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22010}.}
+% \label{table:tab22010}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2010.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2010 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig22010}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12011)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12011[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2011 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12011}.}
+% \label{table:tab12011}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2011.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2011 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig12011}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22011)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22011[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2011 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22011}.}
+% \label{table:tab22011}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2011.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2011 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig22011}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12012)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12012[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2012 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12012}.}
+% \label{table:tab12012}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2012.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2012 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%\end{center}
+\label{fig:fig12012}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22012)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22012[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2012 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22012}.}
+% \label{table:tab22012}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2012.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2012 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig22012}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12013)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12013[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12013}.}
+% \label{table:tab12013}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2013.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2013 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig12013}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22013)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22013[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22013}.}
+% \label{table:tab22013}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2013.pdf}
+\end{center}
+\caption{Proportion of the New Zealand population born in 2013 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+%     \end{center}
+\label{fig:fig22013}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t12014)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t12014[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2014 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12014}.}
+% \label{table:tab12014}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_2014.pdf}
+ \end{center}
+    \caption{Proportion of the New Zealand population born in 2014 vaccinated against measles using MMR1 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig12014}
+\end{sidewaysfigure}
+
+%  \vspace{5mm} %5mm vertical space
+% \begin{table}
+% <<include=FALSE,echo=FALSE,results=tex>>=
+% colnames(t22014)<-c("Min","1st Q","Median","Mean","3rd Q","Max","NA")
+% latex(t22014[1:7], file="", table.env=FALSE,rowname=NULL)
+% @
+% \caption{Summary of the proportion of 2014 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22014}.}
+% \label{table:tab22014}
+% \end{table}
+
+\begin{sidewaysfigure}
+\begin{center}
+    \includegraphics[width=0.9\textwidth]{nir_census_MMR2_NIR_2014.pdf}
+ \end{center}
+\caption{Proportion of the New Zealand population born in 2014 vaccinated against measles using MMR2 in each 2013 area unit.
+Note no adjustment is made for vaccine efficacy.
+The histogram shows the frequency of the proportion vaccinated across the country's area units with 95\% shown in red and the median for the cohort shown in orange.
+See \autoref{table:tab_mmr_nir} and the supplementary data for details.}
+% \end{center}
+\label{fig:fig22014}
+\end{sidewaysfigure}
 
@@
 
-Plotting the lower level data available from the NIR database highlights the issue of surveillance data requiring appropriate denominator data, as these analyses show that there is a lack of correspondence between the national census data and the NIR data (\autoref{fig:naive_au}, \autoref{fig:naive_au_prop})
+Plotting the lower level data from the 2013 area unit data, available from the NIR database, shows that there is heterogeneity at the local level and an obvious temporal lag, as the proportion vaccinated declines with birth cohort (\autoref{fig:mmr_nir}), though recent vaccination levels are consistently high for MMR1.
+For full details, see \autoref{fig:fig12006} (MMR1, 2006), \autoref{fig:fig22006} (MMR2, 2006), \autoref{fig:fig12007} (MMR1, 2007), \autoref{fig:fig22007} (MMR2, 2007), \autoref{fig:fig12008} (MMR1, 2008), \autoref{fig:fig22008} (MMR2, 2008), \autoref{fig:fig12009} (MMR1, 2009), \autoref{fig:fig22009} (MMR2, 2009), \autoref{fig:fig12010} (MMR1, 2010), \autoref{fig:fig22010} (MMR2, 2010), \autoref{fig:fig12011} (MMR1, 2011), \autoref{fig:fig22011} (MMR2, 2011), \autoref{fig:fig12012} (MMR1, 2012), \autoref{fig:fig22012} (MMR2, 2012), \autoref{fig:fig12013} (MMR1, 2013), \autoref{fig:fig22013} (MMR2, 2013), \autoref{fig:fig12014} (MMR1, 2014), \autoref{fig:fig22014} (MMR2, 2014), \autoref{table:tab_mmr_nir} and the supplementary files.
+
+<<echo=FALSE,include=FALSE,results=hide>>=
+MMR1<-c(0.9604,0.9621,0.971,0.9722,0.9667,0.9643,0.9667,0.6333,0)
+MMR2<-c(0.9167,0.9099,0.9048,0.8889,0.7059,0.0345,0.0041,0,0)
+res_bar<-cbind(MMR1,MMR2)
+rownames(res_bar)<-c("2006","2007","2008","2009","2010","2011","2012","2013","2014")
+@
+
+\begin{figure}
+\begin{center}
+<<fig=TRUE,echo=FALSE>>=
+barplot(t(res_bar),beside=T,ylim=c(0,1),xlab="Birth year",ylab="Proportion vaccinated",legend=colnames(res_bar),
+        args.legend = list(x = "topright",bty="n",inset=c(0.1,0.1)))
+abline(h=0.90,col="red",lty=3,lwd=4)
+abline(h=0.95,col="orange",lty=3,lwd=4)
+@
+\end{center}
+\caption{Proportions vaccinated by birth cohort using the NIR database. 95\% and 90\% thresholds are shown by orange and red dotted lines respectively}
+\label{fig:mmr_nir}
+\end{figure}
+
+The correlation between the percentage vaccinated against MMR1 and MMR2 for the 2004 2013 area units is shown in figure \autoref{fig:cor_nir}.
+
+\begin{figure}[H]
+     \begin{center}
+     \includegraphics[width=0.9\textwidth]{nir_census_MMR1_NIR_cor.pdf}
+     \end{center}
+     \caption{The correlation between measles-mumps-rubella (MMR) vaccination doses (MMR1 and MMR2) at the 2013 area unit level from the National Immunisation Register database for 2006--2014.}
+     \label{fig:cor_nir}
+\end{figure}
 
@@
 We use the WHO data to determine per capita measles cases for each year and use these data and the number of immigrants to New Zealand to begin to understand where measles is likely to be imported from.
-We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of mealses importation.
+We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of measles importation.
 We use 0--19 year old data for the immigration figures, as no further details were available and it was determined this was more appropriate than using all age classes because > 19 year olds are believe to be immune to measles more or less globally.
@@
 However, the alternative situation was clear for Pacific people, especially of the 0--1 year olds (\autoref{fig:percap_all} and \autoref{fig:cases_all}).
-Results such as this may simply be due to the spatial effects in the communities the outbreaks were occuring in, rather than an effect of wealth.
-A key issue with incorporating the spatial immunisation data has been the denominator data and the NIR data.
-The data suggest that census data and NIR data are recording children living in different DHBs to that in which they are vaccinated, or subsequent to the census or NIR data moving.
-This leads to some area census units having more than 100\% children vaccinated, with some many times more, and some with very low levels reportedly vaccinated.
-Another issue was how to deal with people of greater age than those recorded in the NIR.
-Thus, these analyses are possible future directions for this work and a focus of future data collection could be to better understand the link between denominator data and NIR data.
+Results such as this may simply be due to the spatial effects in the communities the outbreaks were occurring in, rather than an effect of wealth.
+The risk maps using the spatially explicit NIR vaccination data suggest that there is some heterogeneity in the vaccination cover at the area unit (a level above meshblock) for both MMR1 and MMR2 (see histograms within \autoref{fig:fig12006}, \autoref{fig:fig12007}, \autoref{fig:fig12008}, \autoref{fig:fig12009}, \autoref{fig:fig12010}, \autoref{fig:fig12011}, \autoref{fig:fig12012}, \autoref{fig:fig12013}, \autoref{fig:fig12014}, \autoref{fig:fig22006}, \autoref{fig:fig22007}, \autoref{fig:fig22008}, \autoref{fig:fig22009}, \autoref{fig:fig22010}, \autoref{fig:fig22011}, \autoref{fig:fig22012}, \autoref{fig:fig22013}, and \autoref{fig:fig22014}).
 
 The distribution of the measles cases (\autoref{fig:ageandvac}) appears to reflect the distribution of na\"{\i}ve individuals in the population (\autoref{fig:naive}).
-The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticable that a number ($>16\%$) of cases had been vaccinated at least once.
+The vaccination history of the cases suggests that lack of vaccination cover is the main contributor to the outbreak, though it is noticeable that a number ($>16\%$) of cases had been vaccinated at least once.
 However, the majority of these received only one vaccination, and were vaccinated when they were young (\autoref{fig:vaccstat}).
@@
 
-Unfortunately, additional data we received from the Ministry that we hoped would allow us to provide finer scale results (lower than DHB) immunisation coverage data are not able to provide reliable results, as discussed above.
-Frequently the numbers of vaccinated children in a census area provided in the NIR was greater than the number reported in the census area unit population census (\autoref{fig:naive_au_prop}).
-Given these data gaps we are unable to provide finer scale risk maps (\autoref{fig:naive_au}).
+The additional data we received from the Ministry from the NIR database allows us to look at vaccination coverage as a finer scale.
+There is consistently high vaccination cover by MMR1 (often exceeding 95\%, see \autoref{fig:mmr_nir}), with lower rates for MMR2, but apparently approaching 90\% (\autoref{fig:mmr_nir}).
+The heterogeneity discussed above, however, does not appear to be the same for each area unit, as there less correlation among the percent coverage for the two vaccines (\autoref{fig:cor_nir}) than might be expected, even at a yearly level (not shown).
 
@@
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
-However, it is worth bearing in mind that the data here assume that the national NIR vaccination coverage and serosurvey results are appropriate for all DHB.
+However, it is worth bearing in mind that the data here assume that the serosurvey results are appropriate for all DHB.
 
 Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand.
-However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilence is required for measles vaccination and control efforts.
+However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilance is required for measles vaccination and control efforts.
 The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}).
@@
 \item The majority of cases are unvaccinated.
-\item The majority of vaccine failures occur in those people which recieved single vaccinations around 1 year old.
+\item The majority of vaccine failures occur in those people which received single vaccinations around 1 year old.
 \item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas, though the \%  na\"{\i}ve is similar across DHBs.
@@
 The estimated $R_v$ for each outbreak is shown in \autoref{fig:r0}.
-Our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014, but for which data were available until 12 June, a mean estimate of 3.66 (range 3.4--3.92).
+Our analyses suggest the $R_v$ for the 2009--2014 outbreaks range between 0.18--3.92, and for the last outbreak that ended in August 2014, but for which data were available until 12 June 2014, a mean estimate of 3.66 (range 3.4--3.92).
 The probability density of the $R_v$ estimates for 13 of 20 outbreaks for which we had outbreak numbers include one.
@@
 \hline
-Auckland	&	436350	&	52010	&	31159	&	17920	& 34\\
-Bay of Plenty	&	206000	&	20679	&	8437	&	4585	& 22\\
-Canterbury	&	482180	&	51357	&	24695	&	13687	& 27\\
-Capital and Coast	&	283700	&	32625	&	18403	&	10461 & 32	\\
-Counties Manukau	&	469300	&	55544	&	32903	&	18880	& 33\\
-Hawke's Bay	&	151700	&	15602	&	6846	&	3751 & 24\\
-Hutt Valley	&	138380	&	15198	&	7836	&	4388	& 29\\
-Lakes	&	98196	&	10558	&	5192	&	2886	& 27\\
-MidCentral	&	162560	&	17328	&	8348	&	4628	& 27\\
-Nelson Marlborough	&	137000	&	13059	&	4411	&	2356	& 18\\
-Northland	&	151690	&	14921	&	5688	&	3071	& 21\\
-South Canterbury	&	55620	&	5238	&	1678	&	893	& 17\\
-Southern	&	297420	&	31607	&	15115	&	8371	& 26\\
-Tairawhiti &	43650	&	4769	&	2431	&	1359	& 28\\
-Taranaki	&	109750	&	11473	&	5262	&	2899	& 25\\
-Waikato	&	359310	&	39402	&	20248	&	11331	& 29\\
-Wairarapa	&	41112	&	3932	&	1346	&	720	& 18\\
-Waitemata	&	525550	&	58350	&	30774	&	17291	& 30\\
-West Coast	&	32151	&	3197	&	1265	&	685	& 21\\
-Whanganui	&	60120	&	6075	&	2530	&	1378	& 23\\
+
+Auckland  &	436350 & 51673 & 3628 & 17583 & 34 \\
+Bay of Plenty  &	206000 & 20511 & 811 & 4417 & 22 \\
+Canterbury  &	482180 & 50995 & 2549 & 13325 & 26 \\
+Capital and Coast  &	283700 & 32406 & 2063 & 10242 & 32 \\
+Counties Manukau  &	469300 & 55085 & 3778 & 18421 & 33 \\
+Hawke's Bay  &	151700 & 15471 & 676 & 3619 & 23 \\
+Hutt Valley  &	138380 & 15078 & 833 & 4267 & 28 \\
+Lakes  &	98196 & 10470 & 538 & 2798 & 27 \\
+MidCentral  &	162560 & 17196 & 860 & 4496 & 26 \\
+Nelson Marlborough  &	137000 & 12960 & 401 & 2257 & 17 \\
+Northland  &	151690 & 14793 & 533 & 2942 & 20 \\
+South Canterbury  &	55620 & 5199 & 150 & 854 & 16 \\
+Southern  &	297420 & 31389 & 1557 & 8153 & 26 \\
+Tairawhiti &  43650 & 4727 & 256 & 1317 & 28 \\
+Taranaki  &	109750 & 11378 & 529 & 2804 & 25 \\
+Waikato  &	359310 & 39087 & 2147 & 11016 & 28 \\
+Wairarapa  &	41112 & 3901 & 123 & 689 & 18 \\
+Waitemata  & 525550 & 57904 & 3312 & 16845 & 29 \\
+West Coast  &	32151 & 3172 & 121 & 660 & 21 \\
+Whanganui & 60120 & 6025 & 245 & 1328 & 22 \\
+
 \hline			
-TOTAL   & 		4241739   & 		462924   & 		234567   & 		131539 & 28\\
+TOTAL   & 		4241739   & 		459420   & 		24882   & 		128034 & 28\\
 \hline
@@
 \textit{Na\"{\i}ve}:		DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-\textit{Attack}:		Number infected in DHB in an outbreak of measles in the absence of any vaccination $\left(\Pe\right)$;	
+\textit{Attack}:		Number infected in DHB in an outbreak of measles in the absence of any further vaccination assuming homogeneous mixing across DHB $\left(\Pe\right)$;	
 \textit{Vacc}:		Number of the currently na\"{\i}ve to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; \textit{Vacc (\%)}: Proportion of the currently na\"{\i}ve population to vaccinate.
@@
 \item The reproduction number for measles in the partially immune New Zealand population is often close to and exceeding one, suggesting increased population level immunity is required to ensure prevention of measles persistence following importation.
- \item Additional vaccination levels to push $R_v$ below one among the currently na\"{\i}ve population in New Zealand range from 17\% to 34\% among DHBs, and 28\% at the national level (approximately 131,500 vaccinations).
+ \item Additional vaccination levels to push $R_v$ below one among the currently na\"{\i}ve population in New Zealand range from 17\% to 34\% among DHBs, and 28\% at the national level (approximately 128,034 vaccinations).
 \end{itemize}
@@
 
-Data to estimate the costs relating to the current measles outbreak were obtained from the New Zealand Ministry of Health, for the period 2008 through June 2014.
+Data to estimate the costs relating to the current measles outbreak were obtained from the New Zealand Ministry of Health, for the period 2008 through 12 June 2014.
 Data include information on gender of the case, ethnicity and age of the case at discharge from hospital, days spent in the hospital, year of case, number of events, and associated cost.
@@
 \textit{Na\"{\i}ve}:	DHB na\"{\i}ve population $\left(x_0\times\text{Size}\right)$;		
-\textit{Attack}:	  Number infected in DHB in an outbreak of measles in the \textit{absence} of any vaccination $\left(\Pe\right)$;
-\textit{Vacc}:		Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; \textit{Proportion}: the proportion of the currently na\"{i}ve population requiring vaccination; \textit{Na\"{i}ve.post.vaccination}: the na\"{i}ve population following catch up vaccination; with \textit{Median.outbreak} and \textit{Mean.outbreak}: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed by the small numbers of large outbreaks.
+\textit{Attack}: Number infected in DHB in an outbreak of measles in the \textit{absence} of any catch up vaccination $\left(\Pe\right)$;
+\textit{Vacc}: Number to be vaccinated in DHB to reduce $\Rr_V$ below one  $\left(\left(x_0-1/\Ro\right)\times\text{Size}\right)$; \textit{Proportion}: the proportion of the currently na\"{i}ve population requiring vaccination; \textit{Na\"{i}ve.post.vaccination}: the na\"{i}ve population following catch up vaccination; with \textit{Median.outbreak} and \textit{Mean.outbreak}: the expected median and mean outbreak size post--vaccination catch up from 1000 simulations of a stochastic model}
+\footnote{Note that the median value is what is expected, however, the mean values from 1000 simulations are espeically skewed by the small numbers of large outbreaks.
 The distribution of these outbreaks sizes can be seen in \autoref{fig:sim1} and \autoref{fig:sim}}
@@
   \caption{The distribution of the expected number of measles cases from 1000 simulations of a model (\autoref{section:epidemic_modelling}) in a homogeneously mixed population in New Zealand with 28\% of the currently 11\% susceptible to measles infection vaccinated, using an $R_v$ <1, showing the rare but possible large epidemic sizes possible.
-Note this figure does not show the frequency of common events as the y-axis was limited to 20 to show rare events.
-The full distribution of common results can be seen in \autoref{fig:sim}.}
+  Note this figure does not show the frequency of common events as the y-axis was limited to 20 to show rare events. The full distribution of common results can be seen in \autoref{fig:sim}.}
      \label{fig:sim1}
@@
 Median values are all 2, thus the expected number of cases will be small.
-The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introductions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost20}
@@
 Median values are all 2, thus the expected number of cases will be small.
-The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introdcutions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
+The mean is used as a conservative estimate regarding the benefits of vaccination, because it allows for rare but large outbreaks}; \textit{OB costs}	is costs expected due to continued measles importations based on 10 introductions of measles, one per year, but costs discounted on the same discounted rate; \textit{B/C} is the benefit--cost ratio.}
 \label{table:cost50}
@@
 Similar data are needed for the period preceding 2014.
-In the abscence of these data we have used mean case data from aggregated data available.
+In the absence of these data we have used mean case data from aggregated data available.
 It may be unrealistic to assume that these costs would be linearly related with the number of measles cases, making it difficult to extrapolate these costs outside the reported period for 2014, but we have done so for these purposes.
@@
 In other outbreaks, the average cost per measles case was estimated to be US\$254\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$571}, US\$276\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$621}, and US\$307\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$691} for Canada, the Netherlands, and the UK, respectively  \citep{carabin2}.
-These values are substantially lower than our estimates, however, it is noticable that these estimates are also from more than a decade ago.
+These values are substantially lower than our estimates, however, it is noticeable that these estimates are also from more than a decade ago.
 The containment of a single case (also 2 secondary cases) of measles in 2004 in Iowa, USA was estimated to cost US\$142,542.\footnote{Jan 2004 \$ to US\$ rate 0.6724: \$211990} In this outbreak, more than 2500 hours of personnel time were needed to investigate and respond to the outbreak \citep{dayan5}.
@@
 As stated above, these authors also consider the costs and benefits of mumps and rubella, so making direct comparison difficult without reanalysis.
-Our results are relatively insensitive to changes in vaccination cost, requiring very high costs (>\$8,214 per vaccine) before vaccination is predicted to not be beneficial.
+Our results are relatively insensitive to changes in vaccination cost, requiring very high costs (>\$5,365 per vaccine) before vaccination is predicted to not be beneficial.
 Similarly, longer discounting periods or 5\% discounting did not affect reduce the B/C ration substantially (not shown).
@@
 
+\bibitem[Waaijenborg et al.(2013)]{waaijenborg}
+Waaijenborg, Sandra and Hahn{\'e}, Susan JM and Mollema, Liesbeth and Smits, Gaby P and Berbers, Guy AM and van der Klis, Fiona RM and de Melker, Hester E and Wallinga, Jacco (2013).
+\newblock Waning of maternal antibodies against measles, mumps, rubella, and varicella in communities with contrasting vaccination coverage.
+\newblock \emph{Journal of Infectious Diseases}, 208, 10--16.
+
 \bibitem[Wallinga et al.(2001)]{wallinga1}

### 2015-04-24 draftfinalreport_v3.Rnw
updates background

@@
 Current estimates suggest that approximately 85 to 90 percent of the population is immune to measles (see \autoref{sub:immunestat}), thus the reasons for the ongoing outbreaks are likely due to overall population immunity being less than 95 percent and there being pockets of susceptible, non-immune population remaining.
-Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children (especially young adults) and children under two years of age.
-Those under two year olds are thought to be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
+Since 2009, all the outbreaks in New Zealand have been linked to infections acquired (imported) from overseas, though previous work suggests these outbreaks still largely affect school-aged children, young adults and children under two years of age.
+Those under two years old are thought to be consistently among the most affected age groups because the first of two doses of measles, mumps and rubella vaccine (MMR) is not due until fifteen months.
 

### 2015-04-24 draftfinalreport_v3.Rnw
updtes risk review

@@
 We agree with the \emph {Progress Towards Measles Elimination in New Zealand - Final} report conclusions that testing for measles is performed appropriately within the required timeframe.
-Clearly improving inter-laboratory communication and collaboration and timeliness of the testing and reporting is necessary for rapid responses to measles introductions following measles control.
 Vaccination coverage presented in the report and to ourselves confirms that immunisation levels are approaching 94\% for MMR dose one (birth cohorts 2009 and 2010) and 89\% for MMR dose two (birth cohorts 2006 and 2007).
 However, only Asian and Pacific ethnicities have consistently had MMR dose one coverage approaching or exceeding 95\% for cohorts from 2007 onward, and thus we agree with the report's conclusions that timeliness and coverage of vaccination need improving.
-This is particularly in light of our modelling and risk analyses results (\autoref{sub:risk_analyses} and \autoref{section:epidemic_modelling}).
 

### 2015-04-24 draftfinalreport_v3.Rnw
fix comment on incidence -> notificatoins

@@
      \includegraphics[width=0.9\textwidth]{incidence_1997_2014.pdf}
-     \caption{Measles incidence in New Zealand from 1997 to 2014}
+     \caption{Measles notifications (confirmed and probable cases) in New Zealand from 1997 to 2014}
      \label{fig:incidence1997}
@@
 We used a lognormal distribution with mean 12.0 and standard deviation (s.d.) 3.5 from \citep{klinkenberg11}.
-We then estimated $R_t$ from the incidence data for each outbreak, defining outbreaks in the dataset given their reported outbreak numbers.
+We then estimated $R_t$ from the notification data (confirmed and probable cases) for each outbreak, defining outbreaks in the dataset given their reported outbreak numbers.
 The geographic locations of cases are shown in \autoref{fig:dhbcases}.

### 2015-04-24 draftfinalreport_v3.Rnw
updates 4.1 risk analysis

@@
 
-This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases in, particularly in the adult age classes.
-Furthermore, initial preliminary analyses, including multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions, and therefore we reduced the number of NZDep categories from ten to three: NZDep 1--3 (least deprived), NZDep 4--7, and NZDep 8--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}) following discussion with the Ministry regarding the most appropriate categories.
+This large number of categories, some with small population sizes, leads to both overdispersion and zeroinflation, as there are many categories with zero cases, particularly in the adult age classes.
+Furthermore, initial preliminary multi-- and univariate analyses (not shown) suggested little effect of \textit{individual} NZDep classifications and several higher order interactions.
+Following discussion with the Ministry regarding the most appropriate NZDep categories, we reduced the number of categories from ten to three: NZDep 1--3 (least deprived), NZDep 4--7, and NZDep 8--10 (most deprived) (\href{http://www.otago.ac.nz/wellington/research/hirp/otago020194.html}{www.otago.ac.nz}).
 We also incorporated the 65+ age classes into the 25--64 age category, to make a 25+ age category.
@@
 
-The numbers of cases per category and population sizes for the complete data set from 2007-2014 can be seen in \autoref{table:percap}.
-Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard error in regression analyses.
+The numbers of cases per category and population sizes for the complete data set from 2007--2014 can be seen in \autoref{table:percap}.
+Subsequent regression analyses (not shown) also suggested that the MLA category was over-- or underrepresented in per capita rates given the very small sample sizes for this classification (\autoref{table:percap}), leading to very large standard errors in regression analyses.
 However, there are also numerous issues with the data for MLA ethnicity category above and beyond the small population sizes, creating issues for estimating the denominator data for this group (University of Otago, personal communication).
@@
 
-For all our statistical analyses (including those above not shown) we used a Poisson error structure, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model.
-We also account for differences in population sizes by using an offset term, the log(population size).
+For all our statistical analyses (including those above not shown) we used a Poisson error structure with rate adjusted for population size, but in all cases there was a need to account for overdispersion and thus we used and present the results of a quasipoission regression model.
 We used a model simplification approach, by beginning our analyses with all terms and all interactions, and then simplifying the models through removal of non-significant higher order interaction terms.
@@
 
-\begin{equation} \label{eq:reg}
- \log(y) = \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _n (x_n) + \beta _a{}_e(x_a \times x_e)+ \log(population)  + \epsilon
-  \end{equation}
-Where $\alpha$ is the intercept, $y$ signifies cases, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity, $_n$ signifies NZDep, and $\epsilon$ is the error term.
+\begin{align*} \label{eq:reg}
+ y &\sim \textsf{Poisson}(\lambda n)\\
+ \log(\lambda) &= \alpha + \beta _a (x_a)+ \beta _e(x_e)+ \beta _s (x_s) + \beta _a{}_e(x_a \times x_e)
+  \end{align*}
+Where $y$ is the number of cases, $\lambda$ is cases per capita, $n$ is population size, $\alpha$ is the intercept, $_a$ signifies age, $_e$ signifies Prioritized Ethnicity and $_s$ signifies NZDep.
 

### 2015-04-24 draftfinalreport_v3.Rnw
update results 4.4

@@
 
-The cases and per capita measles values are in \autoref{fig:cases_all} and \autoref{fig:percap_all}.
+The cases and per capita measles rates are in \autoref{fig:cases_all} and \autoref{fig:percap_all}.
 The distribution of the measles cases per category used in the regression analyses are in \autoref{fig:histcasecat}.
@@
 
-Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case (\autoref{table:regression_riskfactors}).
+Apart from over-representation of some MLA categories discussed above and not included here, the results of the regression model suggest that age is a strong predictor of being a measles case (\autoref{table:anova}).
 Indeed, all age categories are significantly less likely to be measles cases compared to 0--1 year olds, and the likelihood generally decreases with age (\autoref{table:regression_riskfactors}, \autoref{fig:cases_all}, \autoref{fig:percap_all}, see also \autoref{fig:ageinyears}).
@@
 However, 12.6\% (154/1154) cases had received their first dose of vaccine and 4.7\% had received their second.
-A further breakdown of those data suggest that majority of those 'vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
+A further breakdown of those data suggest that majority of those `vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
 Plotting the unvaccinated cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).

### 2015-04-24 draftfinalreport_v3.Rnw
update figures in 4.3 for MMR2 date etc.

@@
 \end{center}
-\caption{Age of vaccination of vaccinated measles cases, 2007- 12 June 2014.
-Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (60 months)}
+\caption{Age of vaccination of vaccinated measles cases, 2007--12~June~2014.
+Recommended vaccination age in months is shown for the measles, mumps, rubella (MMR) dose 1 (12 months) and MMR dose 2 (4 years)}
 \label{fig:vaccstat}
@@
 \end{center}
-\caption{Year of birth of unvaccinated of measles cases, 2007- 12 June 2014}
+\caption{Year of birth of unvaccinated of measles cases, 2007--12~June~2014}
 \label{fig:yearandvac}
@@
 \end{center}
-\caption{Year of birth of unvaccinated of measles cases, 2007- 12 June 2014.
-The year of birth for those vaccinated with measles, mumps, and rubella (MMR) dose 1 is shown in the upper panel, MMR dose 2 the lower panel.}
+\caption{The year of birth for those vaccinated with measles, mumps, and rubella (MMR) dose 1 is shown in the upper panel, MMR dose 2 the lower panel.}
 \label{fig:yearandvacf}

### 2015-04-24 draftfinalreport_v3.Rnw
update rest of figures to drop '2013' from area unit in captions etc.

@@
 @
-\caption{Summary of the proportion of 2006--2014 birth cohorts vaccinated against measles-mumps-rubella (MMR) dose 1 and MMR2 across the 2013 area units.
+\caption{Summary of the proportion of 2006--2014 birth cohorts vaccinated against measles-mumps-rubella (MMR) dose 1 and MMR2 across area units.
 Area units are aggregations of meshblock data.
@@
     \end{center}
-    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22006}.}
+% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22006}.}
 % \label{table:tab22006}
@@
 \end{center}
-    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR2 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2006 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2007 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12007}.}
+% \caption{Summary of the proportion of 2007 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12007}.}
 % \label{table:tab12007}
@@
 \end{center}
-    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2007 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22007}.}
+% \caption{Summary of the proportion of 2007 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22007}.}
 % \label{table:tab22007}
@@
 \end{center}
-    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR2 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2007 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2008 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12008}.}
+% \caption{Summary of the proportion of 2008 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12008}.}
 % \label{table:tab12008}
@@
 \end{center}
-    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2008 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22008}.}
+% \caption{Summary of the proportion of 2008 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22008}.}
 % \label{table:tab22008}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR2 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2008 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2009 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12009}.}
+% \caption{Summary of the proportion of 2009 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12009}.}
 % \label{table:tab12009}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2009 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22009}.}
+% \caption{Summary of the proportion of 2009 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22009}.}
 % \label{table:tab22009}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR2 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2009 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2010 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12010}.}
+% \caption{Summary of the proportion of 2010 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12010}.}
 % \label{table:tab12010}
@@
 \end{center}
-    \caption{Proportion of the New Zealand population born in 2010 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2010 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2010 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22010}.}
+% \caption{Summary of the proportion of 2010 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22010}.}
 % \label{table:tab22010}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2010 vaccinated against measles using MMR2 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2010 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2011 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12011}.}
+% \caption{Summary of the proportion of 2011 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12011}.}
 % \label{table:tab12011}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2011 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2011 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2011 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22011}.}
+% \caption{Summary of the proportion of 2011 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22011}.}
 % \label{table:tab22011}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2011 vaccinated against measles using MMR2 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2011 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2012 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12012}.}
+% \caption{Summary of the proportion of 2012 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12012}.}
 % \label{table:tab12012}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2012 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2012 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2012 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22012}.}
+% \caption{Summary of the proportion of 2012 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22012}.}
 % \label{table:tab22012}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2012 vaccinated against measles using MMR2 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2012 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12013}.}
+% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12013}.}
 % \label{table:tab12013}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2013 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2013 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22013}.}
+% \caption{Summary of the proportion of 2006 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22013}.}
 % \label{table:tab22013}
@@
 \end{center}
-\caption{Proportion of the New Zealand population born in 2013 vaccinated against measles using MMR2 in each 2013 area unit.
+\caption{Proportion of the New Zealand population born in 2013 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2014 birth cohort vaccinated against MMR1 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12014}.}
+% \caption{Summary of the proportion of 2014 birth cohort vaccinated against MMR1 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig12014}.}
 % \label{table:tab12014}
@@
  \end{center}
-    \caption{Proportion of the New Zealand population born in 2014 vaccinated against measles using MMR1 in each 2013 area unit.
+    \caption{Proportion of the New Zealand population born in 2014 vaccinated against measles using MMR1 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 % @
-% \caption{Summary of the proportion of 2014 birth cohort vaccinated against MMR2 across the 2013 area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22014}.}
+% \caption{Summary of the proportion of 2014 birth cohort vaccinated against MMR2 across the area units. Area units are aggregations of meshblock data. See \autoref{fig:fig22014}.}
 % \label{table:tab22014}
@@
  \end{center}
-\caption{Proportion of the New Zealand population born in 2014 vaccinated against measles using MMR2 in each 2013 area unit.
+\caption{Proportion of the New Zealand population born in 2014 vaccinated against measles using MMR2 in each area unit.
 Note no adjustment is made for vaccine efficacy.
@@
 
-Plotting the lower level data from the 2013 area unit data, available from the NIR database, shows that there is heterogeneity at the local level and an obvious temporal lag, as the proportion vaccinated declines with birth cohort (\autoref{fig:mmr_nir}), though recent vaccination levels are consistently high for MMR1.
+Plotting the lower level data from the area unit data, available from the NIR database, shows that there is heterogeneity at the local level and an obvious temporal lag, as the proportion vaccinated declines with birth cohort (\autoref{fig:mmr_nir}), though recent vaccination levels are consistently high for MMR1.
 For full details, see \autoref{fig:fig12006} (MMR1, 2006), \autoref{fig:fig22006} (MMR2, 2006), \autoref{fig:fig12007} (MMR1, 2007), \autoref{fig:fig22007} (MMR2, 2007), \autoref{fig:fig12008} (MMR1, 2008), \autoref{fig:fig22008} (MMR2, 2008), \autoref{fig:fig12009} (MMR1, 2009), \autoref{fig:fig22009} (MMR2, 2009), \autoref{fig:fig12010} (MMR1, 2010), \autoref{fig:fig22010} (MMR2, 2010), \autoref{fig:fig12011} (MMR1, 2011), \autoref{fig:fig22011} (MMR2, 2011), \autoref{fig:fig12012} (MMR1, 2012), \autoref{fig:fig22012} (MMR2, 2012), \autoref{fig:fig12013} (MMR1, 2013), \autoref{fig:fig22013} (MMR2, 2013), \autoref{fig:fig12014} (MMR1, 2014), \autoref{fig:fig22014} (MMR2, 2014), \autoref{table:tab_mmr_nir} and the supplementary files.
@@
 
-The correlation between the percentage vaccinated against MMR1 and MMR2 for the 2004 2013 area units is shown in figure \autoref{fig:cor_nir}.
+The correlation between the percentage vaccinated against MMR1 and MMR2 across the area units is shown in figure \autoref{fig:cor_nir}.
 
@@
      \end{center}
-     \caption{The correlation between measles-mumps-rubella (MMR) vaccination doses (MMR1 and MMR2) at the 2013 area unit level from the National Immunisation Register database for 2006--2014.}
+     \caption{The correlation between measles-mumps-rubella (MMR) vaccination doses (MMR1 and MMR2) at the area unit level from the National Immunisation Register database for 2006--2014.}
      \label{fig:cor_nir}

### 2015-04-24 draftfinalreport_v3.Rnw
update rest of risk analysis section

@@
 We use simple per capita rates for measles and the number of travellers to each country to score and map the risk of measles importation.
-We use 0--19 year old data for the immigration figures, as no further details were available and it was determined this was more appropriate than using all age classes because > 19 year olds are believe to be immune to measles more or less globally.
+We use 0--19 year old data for the immigration figures, as no further details were available and it was determined this was more appropriate than using all age classes because > 19 year olds are believed to be immune to measles more or less globally.
 These age category data were not available for New Zealand travellers, so we use all data on all travellers.
@@
 
-Globally, analyses of the most complete and recent year of data available, 2013, suggest measles incidence highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}).
-Immigration by 0--19 year old non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia \autoref{table:imtravel12} and \autoref{fig:immigration12}.
+Globally, analyses of the most complete and recent year of data available, 2013, suggest measles incidence is highest  and vaccination coverage is lowest in less developed nations (\autoref{table:incidence12} and \autoref{table:cover12}, and \autoref{fig:incidence12} and \autoref{fig:cover12}).
+Immigration by 0--19 year old non-New Zealanders (whether for work, pleasure, etc.) is dominated by people from Australia (\autoref{table:imtravel12} and \autoref{fig:immigration12}).
 A number of other countries then follow, including United Kingdom, China, Japan and the United States.
@@
 Though the precise interactions between these different risk factors are unknown, the most simple, a product of measles incidence in 2013 and immigration numbers of 0--19 year olds in 2013, suggest that those countries with greatest international travel, Australia, UK, China and South East Asian countries, pose the greatest risk to New Zealand for measles importation.
-Though immigration from and travel to some Asian countries is lower, some Asian countries also poses a high risk of measles importation to New Zealand.
-Note the South Pacific island do not appear to pose a greater risk, despite large numbers of travellers of New Zealand origin, perhaps because of low measles incidence in these nations.
+Though immigration from and travel to some Asian countries is lower, some Asian countries also pose a high risk of measles importation to New Zealand.
+Note that South Pacific islands do not appear to pose a greater risk, despite large numbers of travellers of New Zealand origin, perhaps because of low measles incidence in these nations.
 These data are shown in \autoref{table:risk12}.
-The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in \autoref{table:nzrisk12} and \autoref{table:imrisk12} and \autoref{fig:nzrisk12} and \autoref{fig:imrisk12}.
+The breakdown of risk of measles importation by New Zealander and non-New Zealander travellers is in \autoref{table:nzrisk12}, \autoref{table:imrisk12}, \autoref{fig:nzrisk12} and \autoref{fig:imrisk12}.
 The measles incidence and vaccination coverage for those countries identified as posing the greatest risk for measles importation to New Zealand are in \autoref{table:toprisk}.
@@
 The additional data we received from the Ministry from the NIR database allows us to look at vaccination coverage as a finer scale.
-There is consistently high vaccination cover by MMR1 (often exceeding 95\%, see \autoref{fig:mmr_nir}), with lower rates for MMR2, but apparently approaching 90\% (\autoref{fig:mmr_nir}).
-The heterogeneity discussed above, however, does not appear to be the same for each area unit, as there less correlation among the percent coverage for the two vaccines (\autoref{fig:cor_nir}) than might be expected, even at a yearly level (not shown).
+There is consistently high vaccination cover by MMR1 (often exceeding 95\%, see \autoref{fig:mmr_nir}), with lower rates for MMR2, approaching 90\% (\autoref{fig:mmr_nir}).
+The heterogeneity discussed above, however, does not appear to be the same for each area unit, as there is less correlation among the percent coverage for the two vaccines (\autoref{fig:cor_nir}) than might be expected, even at a yearly level (not shown).
 
@@
 We hope that the Ministry will find these informative, as they indicate which age classes may be the focus of vaccination efforts for each DHB.
-However, it is worth bearing in mind that the data here assume that the serosurvey results are appropriate for all DHB.
+However, it is worth bearing in mind that the data here assume that the serosurvey results are appropriate for all DHBs.
 
 Further analyses are required to see if the seasonal patterns in travel match seasonality in measles incidence in the countries from which most individuals come from when travelling to, or back to, New Zealand.
-However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where extra effort and vigilance is required for measles vaccination and control efforts.
+However, the strong seasonality in travel around the summer vacation in December suggests this may be a time where additional effort and vigilance is required for measles vaccination and control efforts.
 The area identified as greatest risk for measles importation, Australia, simply reflects the enormous amount of travel between Australia and New Zealand (\autoref{table:immigration12} and \autoref{fig:immigration12}).
@@
 \item European ethnicity comprise the largest number of measles cases.
-\item Pacific people are statistically more at risk per capita of measles infection.
+\item Pacific people are statistically more at risk per capita of measles infection, primarily 0--2 year olds.
 \item There is some statistical support for those living in better socio-economic situations being at greater risk of measles.
@@
 \item The majority of vaccine failures occur in those people which received single vaccinations around 1 year old.
-\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority predictably in DHBs with larger urban areas, though the \%  na\"{\i}ve is similar across DHBs.
+\item Distribution of numbers of na\"{\i}ve among New Zealand is uneven, with the majority in DHBs with larger urban areas, though the \%  na\"{\i}ve is similar across DHBs.
 \item There is a continued, and perhaps increasing, risk of measles importation due to travel and endemic measles elsewhere in the world.

### 2015-04-24 draftfinalreport_v3.Rnw
updates modelling section

@@
 We then estimated $R_t$ from the notification data (confirmed and probable cases) for each outbreak, defining outbreaks in the dataset given their reported outbreak numbers.
-The geographic locations of cases are shown in \autoref{fig:dhbcases}.
+The geographic location of cases are shown in \autoref{fig:dhbcases}.
 The outbreaks we used in our analyses are shown in \autoref{fig:outbreaks}.
@@
 To estimate the proportion of the population requiring vaccination utilising our estimates of $R_v$, we use the well-known equation for the final size of an epidemic in a homogeneously mixing susceptible population \citep{diekmann13}:
-
 \begin{equation} \label{eq:eqn1}
-\log\left(1-\Pe\right)+\Ro\Pe=0
+\log\left(1-\Pe\right)+\Ro\Pe=0,
 \end{equation}
-
 where $\Ro$ is the basic reproduction number and $\Pe$ is the proportion of the population infected over the course of the outbreak.
@@
 If a proportion $x_0$ of the population is susceptible following vaccination, then the  reproduction number under vaccination is $\Rr_V=x_0\Ro$, and the final size equation becomes
-
 \begin{equation} \label{eq:eqn2}
-\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0
+\log\left(1-\frac{\Pe}{x_0}\right)+\Ro\Pe=0.
 \end{equation}
@@
 Hence the relationship between the proportion initially susceptible and the proportion infected in an epidemic is
-
 \begin{equation} \label{eq:eqn3}
-x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}
+x_0=\frac{\Pe}{1-e^{-\Ro\Pe}}.
 \end{equation}
@@
 In order to prevent future epidemics, it is necessary that $\Rr_V<1$. Hence, the proportion of the population that must be vaccinated to prevent future outbreaks, $\Pv$ is
-
 \begin{equation} \label{eq:prop}
-\Pv = x_0-1/\Ro
+\Pv = x_0-1/\Ro.
 \end{equation}
@@
 This implies that outbreaks may persist within the population for a substantial period, even if $R_v$ values are not much greater than one.
-A caveat to this and other $R_v$ estimates is that the outbreaks may include some sporadic cases misclassified, but also exclude some sporadic cases as yet not associated with another outbreak.
+A caveat to this and other $R_v$ estimates is that the outbreaks may include some sporadic cases misclassified, but also exclude some sporadic cases as yet not associated with an outbreak.
 Undetected and unreported cases may lower the estimate.
@@
 Analysis of measles vaccination in various regions suggest that 85\% coverage at MMR1 and MMR2 could be sufficient to prevent future measles epidemics \citep{agur93, babad95, edmunds0, gay98, wallinga1}, whereas \citep{glass4} showed that high overall levels of measles vaccination can obscure pockets of poor coverage, resulting in localised regions with increased risk of infection making effective immunisation difficult.
-Future analyses at a smaller, more local level would be useful, but the lack of appropriate data matching between the NIR data and the census unit area data currently prevent these.
 
@@
 \item The reproduction number for measles in the partially immune New Zealand population is often close to and exceeding one, suggesting increased population level immunity is required to ensure prevention of measles persistence following importation.
- \item Additional vaccination levels to push $R_v$ below one among the currently na\"{\i}ve population in New Zealand range from 17\% to 34\% among DHBs, and 28\% at the national level (approximately 128,034 vaccinations).
+ \item Additional vaccination of the currently na\"{\i}ve population in New Zealand required to push $R_v$ below one ranges from 17\% to 34\% among DHBs, and is 28\% at the national level (approximately 128,034 vaccinations).
 \end{itemize}

### 2015-04-24 draftfinalreport_v3.Rnw
updates cost benefit

@@
 Though the global incidence of measles is declining, in recent years that decline has slowed (\autoref{fig:trendincidence}) and immigration rates to New Zealand have risen (\autoref{fig:travel}).
-This suggests that the risk of measles importation could increase, though further analyses are require to understand the interaction between these variables.
+This suggests that the risk of measles importation could stay at similar levels or increase, though further analyses are require to understand the interaction between these variables.
 Of note, however, is the clear seasonality in immigration and travel (\autoref{fig:travel}).
@@
 Of particular note there have been a number of outbreaks for which our estimated $R_v$ was well above one and thus we may expect outbreaks to occur if conditions remain the same.
-An important caveat to the analysis of the last outbreak is that because this outbreak that ended in August was ongoing when the data were collected, and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
+An important caveat to the analysis of the last outbreak is that because this outbreak that ended in August was ongoing when the data were collected (12 June 2014), and not in decline, $R_v$ is necessarily over one, and so the comparison with others must be cautious.
 However, this is not the only case for which the $R_v$ is greater than one (\autoref{fig:r0}).
@@
 To use the results from our modelling exercise to help inform the appropriate measles vaccination coverage, we use \autoref{eq:prop}.
-The proportion of the population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (\autoref{table:attack}).
+The proportion of the na\"{i}ve population requiring additional vaccination to make $R_v$ < 1 ranged from 17\% to 34\% at the DHB level, with a national average of approximately 28\% (\autoref{table:attack}).
 These additional vaccination numbers can be calculated in a number of different ways, as discussed in the benefit--cost section (\autoref{sub:cost_benefit}).
@@
 The incidence of cases was reduced by more than 50\% from 43 million in 1999 to approximately 20 million in 2005.
-Approximately 7.5 million deaths from measles were avoided from 2000--05 due vaccination \citep{wolfson7}.
+Approximately 7.5 million deaths from measles were avoided from 2000--05 due to vaccination \citep{wolfson7}.
 The World Health Organization (WHO) estimated 158,000 deaths from approximately 355,000 measles cases in 2011 \citep{who13}.
@@
 The estimated cost for a case ranged from US\$189--344\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$425--774} \citep{carabin3}; however, the average estimated cost of a typical hospital case ranges from US\$967--1,755\footnote{Jan 2001 \$ to US\$ rate 0.4445: \$2175--3948} \citep{carabin2}.
-\citep{stack11} estimated the economic benefits from cases averted due to measles vaccination.
+\citet{stack11} estimated the economic benefits from cases averted due to measles vaccination.
 They estimated that the expanded vaccination from 2005 to 2015 in 72 of the world's poorest countries could result in nearly US\$10 billion\footnote{Jan 2009 \$ to US\$ rate 0.5526: \$18.1billion} of costs averted between 2011 and 2020.
@@
 
-Wages lost due to measles are calculated for the period January 2008 - August 2014.
+Wages lost due to measles are calculated for the period January~2008--August~2014.
 We assume that the losses to the economy are the same, whether the the employee is able to obtain sick leave pay, as this is the general loss to the economy as work days are lost.
-Calculations are based on the assumptions that 5 days of work are lost for each case; however, individuals between 0-14 years of age are not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15-19 years of age was 41.9\%  (Statistics New Zealand, 2013).
-If the case is < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15-19 years of age.
+Calculations are based on the assumptions that 5 days of work are lost for each case; however, individuals between 0--14 years of age are not assumed to be employed and therefore did not suffer an income loss and the employment rate of teenagers 15--19 years of age was 41.9\%  (Statistics New Zealand, 2013).
+If the case is < 20 years of age, it was assumed there was an income loss of 5 days for the care giver, in addition to the wage loss of the case if 15--19 years of age.
 Case contact data was used to estimate further wage loss costs.
@@
 
-The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital from 2001--2014 (part year) was \$550,024 (\autoref{table:hosp}).
-The mean cost per case was \$1,877.
+The total cost for the 293 publicly funded hospital discharges with a measles primary diagnosis that spent 470 nights in hospital from 2000--2014 (part year) was \$550,024 (\autoref{table:hosp}).
+The mean cost per hospitalised case was \$1,877.
 The mean cost per day of stay in the hospital was \$1,170.
@@
 
-Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation, case age and year of case, and a less strong association with case gender (\autoref{table:regression_costs}).
+Regression analyses showed statistically significant associations between cost of hospitalisation and three variables: length of hospitalisation and year of case, and a less strong association with case age and case gender (\autoref{table:regression_costs}).
 Results showed the expected hospitalisation costs in 2000 of a female measles patient who did not stay overnight in the hospital was \$582.
 The cost was \$256 less if the case was a male.
-It increased of approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014.
+It increased approximately \$406 per night of hospitalisation and \$64 per year over the time period of 2000--2014.
 The cost of a case decreased with the age of the patient by approximately \$8 per year of case age.
@@
 
-Wages lost for the period January 2008 -- August 2014 for the 247 cases and care givers was estimated to be \$207,155.
+Wages lost for the period January~2008--August~2014 for the 247 cases and care givers was estimated to be \$207,155.
 This consisted of \$104,539 for the cases and \$102,616 for the care giver, but did not include wage losses for cases under 15 years of age.
-Overall, the cost per case from 2008--2014 was estimated to be \$2923.27 (\$839 in forgone wages, \$1,765 in management costs, and \$1,710 in hospital costs).
-This figure brings an approximate estimate of \$546650.78 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs for 17\%.
+Overall, the cost per case from 2008--2014 was estimated to be \$2923.27 (\$839 in forgone wages, \$1,765 in management costs, and \$1,877 in hospital costs for the 17\% of cases hospitalised).
+This figure brings an approximate estimate of \$546650.78 for 187 cases for the current outbreak in 2014 alone, which is comprised of earnings lost, case management and hospitalisation costs.
 
@@
 Thus the expected measles-related costs due to constant introduction of measles despite increased population immunity and the costs of the vaccination schemes are used to estimate the costs of additional vaccine programs for the B/C analysis.
-The financial savings from reduced measles cases are used to work out the benefits for the additional vaccination programs.
+The financial savings from reduced measles cases are used to estimate the benefits for the additional vaccination programs.
 The costs for the catch up vaccination programs are presumed to be over a single year.
@@
 Note vaccination and GP costs can be altered by the Ministry of Health in the spreadsheet provided.
+In addition, laboratory costs which were not included above may also be altered in the spreadsheet.
 
@@
 
-It is worth noting that vaccination strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (\autoref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (\autoref{fig:vaccstat}).
+In corcordance with current measles knowledge, strategies that target the very young (<1 year old) may be less effective, as our analyses of the vaccinated cases suggests a substantial proportion of vaccinated cases that were vaccinated (\autoref{fig:ageandvac}) were vaccinated with a single vaccine at a very young age (\autoref{fig:vaccstat}).
 Furthermore, it may be unnecessary to vaccinate very young, as it appears possible to reach the appropriate figure (28\% of currently na\"{\i}ve) by vaccinating all the currently na\"{\i}ve 2--17 year olds, so pre-school and school age children.
@@
 \begin{itemize}
-\item The mean wage losses per measles case is estimated to be \$839
-\item The mean cost per measles case attending hospital is estimated to be \$1,877
-\item Approximately 17\% of measles cases attend hospital
-\item The mean management cost per measles case for the current outbreak, estimated from three months of data is estimated to be \$1,765
+\item The mean wage losses per measles case is estimated to be \$839.
+\item The mean cost for measles cases that attend hospital is estimated to be \$1,877.
+\item Approximately 17\% of measles cases attend hospital.
+\item The mean management cost per measles case for the current outbreak, estimated from three months of data is estimated to be \$1,765.
 \item Using $R_0$ values estimated for measles in New Zealand prior to vaccination and the current na\"{\i}ve population sizes, the benefits of catch up vaccination strategies are clear (>1 B/C ratio).

### 2015-04-24 draftfinalreport_v3.Rnw
update comment on vaccien failure graphs

@@
 A further breakdown of those data suggest that majority of those `vaccine failures' were vaccinated around the first year of age (\autoref{fig:vaccstat}).
-Plotting the unvaccinated cases (\autoref{fig:yearandvac}) and vaccine failures (\autoref{fig:yearandvacf}) by year of birth to look at cohort effects appears to simply reflect the age of vaccine data (\autoref{fig:ageandvac}).
+The unvaccinated cases and vaccine failures by year of birth are shown in \autoref{fig:yearandvac} and \autoref{fig:yearandvacf}.
 

### 2015-04-24 draftfinalreport_v3.Rnw
drop figure 13

@@
 
-\begin{figure}
-     \begin{center}
-     \includegraphics[width=0.9\textwidth]{naive_map.pdf}
-     \end{center}
-     \caption{Numbers of na\"{\i}ve individuals per District Health Board, using national immunity data (\autoref{table:sero})}
-     \label{fig:naive_map}
-\end{figure}
-
 \subsection{Population immunity results}
@@
 However, the distribution of na\"{\i}ve at a national level shows that the recent MMR vaccination schemes are reducing the proportions of na\"{\i}ve population in the 3--5 year old age group to greater levels than in older young people  (\autoref{fig:naive}).
-Plotting the breakdown of these figures by DHB clearly shows that the greatest numbers of na\"{\i}ve people are in DHBs with larger urban areas (\autoref{fig:naive_map}).
 The distribution of the numbers of na\"{\i}ve and the total na\"{\i}ve populations per DHB, assuming national immunisation and immunity rates are representative, are given in the following figures in the appendix (\autoref{sec:app}):
@@
 However, the majority of these received only one vaccination, and were vaccinated when they were young (\autoref{fig:vaccstat}).
-The majority of na\"{\i}ve among the New Zealand population is clearly focused in the DHB with large urban areas (\autoref{fig:naive_map}).
 
@@
 The results of these modelling exercises suggest vaccination levels are close to eliminating the possibility of endemic measles transmission, as estimates of $R_v$ typically include 1 (\autoref{fig:r0}).
-However, the na\"{\i}ve population (\autoref{fig:naive} and \autoref{fig:naive_map}) and the higher $R_v$ for some of the 2009--2014 outbreaks (\autoref{fig:r0}) suggests that catch up vaccination is necessary (\autoref{table:attack}).
+However, the na\"{\i}ve population (\autoref{fig:naive}) and the higher $R_v$ for some of the 2009--2014 outbreaks (\autoref{fig:r0}) suggests that catch up vaccination is necessary (\autoref{table:attack}).
 The results of these analyses support other studies that suggest very high rates of vaccination are required to eliminate measles and prevent epidemics.
@@
 To estimate the benefits from additional vaccinations, as estimated from the above modelling section (\autoref{section:epidemic_modelling}), we carried out a number of analyses.
-Primarily, we used \autoref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (\autoref{fig:naive_map} and \autoref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (\autoref{section:epidemic_modelling}).
+Primarily, we used \autoref{eq:prop} to estimate the proportion of the na\"{\i}ve populations in each DHB (\autoref{table:attack}) and the national level requiring vaccination to reduce the $R_v$ to <1 (\autoref{section:epidemic_modelling}).
 We assume that the number of cases prevented by this is proportional to the outbreak size.
